### 2015-04-21 manual/manual.tex
Added suggestion for manual (stub)

\documentclass[10pt,a4paper]{report}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\title{A concise manual to\\\textsc{DEVS Ex Machina}}
\author{Ben Cardoen\\Pieter Lauwers\\Stijn Manhaeve\\Tim Tuijn\\Matthijs Van Os}
\date{}

\begin{document}

\maketitle
\tableofcontents{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Overview}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{tex/overview.tex}	% Schrijf alles in dit apart bestand

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Creating the models}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{tex/models.tex}		% Schrijf alles in dit apart bestand

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Running the simulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{tex/controller.tex}	% Schrijf alles in dit apart bestand

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\chapter{Example simulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{tex/example.tex}		% Schrijf alles in dit apart bestand

\end{document}

### 2015-04-23 manual/tex/faq.tex
Add build/configuration information to manual + faqs

\section{Platform specific}
\subsubsection{Virtualization}
\begin{enumerate}
\item On Windows host, with a *nix image, it's possible that standard compliant code leaks memory in std::string. We're aware of this issue and resolve it where we detect it with Valgrind.
\item Threading is significantly slower depending on your host, you should at least allocate 2 virtual CPU's to your virtual instance to allow proper functioning of the simulator. Several threaded tests can fail , since the runtime can only provide a hint (std::thread::hardware\textunderscore concurrency()) to the program to detect if we can use threads.
\end{enumerate}
\subsubsection{Cygwin}
\begin{enumerate}
\item std::to\textunderscore string is not supported, do not use it in your own models.
\item GTest version 1.7 corrupts the stack if a test fails, use version 1.6 until GTest resolves this.
\end{enumerate}
\section{General considerations}
\begin{enumerate}
\item If you violate certain clearly documented preconditions (such as having several models with the exact same name), the simulator is allowed to crash hard.
\item Do not use any non-serializable features of C++, examples are std::function (function / raw pointers in general).
\item Time is represented using std::size\textunderscore t internally, which is 64 bits wide on most 64-bit systems, 32 on older 32-bit system. As a result, should you ever use an explicit value larger than std::numeric\textunderscore limits$<$std::size\textunderscore t$>$::max(), the time representation will fail in unexpected ways. The same issue arises if the simulator advances beyond that time point. It's possible to change the typedef t\textunderscore timestamp to use a larger type should this be necessary for you. For example :\\
\lstinline!typedef Time<unsigned long long, unsigned long long> t_timestamp;!
As long as the supplied type to Time has well-defined operators for the aritmetic and comparison operators (at least partial ordered), and the type has a specialization for numeric limits, it can be used.
\end{enumerate}
\section{Threading}
\begin{enumerate}
\item Never use more simulator cores than hardware threads you have available. For example on a dual core cpu with hyperthreading, 4 simualtor-cores is the limit you should adhere to. The simulator may work with more simulator-cores, but performance will be crippled and starvation is likely.
\item libstdc++'s implementation of std::string is Copy on Write, this introduces an inherent danger for race conditions. Should you ever be confronted with this issue (in your own models), use the provided copyString function in stringtools.h to force g++ to generate an explicit (safe) copy.
\end{enumerate}

### 2015-04-23 manual/tex/controller.tex

\section{Setup}

After creating the models for your simulation, there are still a few parameters to take care of to get everything up and running.\\

First of these is the \textbf{allocator}, which decides on which simulation core to place which model. We have already included a simple allocator that distributes models evenly, but for more complex behavior it is possible to create one tailored specifically for your simulation purposes by inheriting from the \textsl{Allocator} base class.\\

The next step is to create a \textbf{location table} and a table for the simulation cores. The former stores the location of all models. A \textsl{LocationTable} class is included for this purpose.\\
The latter needs to be an \textsl{std::unordered\_map{\textless}std::size\_t, t\_coreptr{\textgreater}}. This map will store a pointer to each core which can be accessed using the core's ID.\\

To get output from your simulation, you can select one or more of the included \textbf{tracers}. [...] TODO\\

The following steps differ slightly depending on which type of DEVS you want to run.

\subsection{Classic DEVS}
\textsc{Classic DEVS} depends on the use of a single \textbf{core} of the \textsl{Core} class. Create a core of this class and add it to the \textsl{unordered\_map} created before.\\

Next, we create the \textbf{controller} of the simulation, using the \textsl{Controller} class. It expects a name, the core table, the allocator, the location table, and the tracers as its arguments. You can now set the simulation mode to CDEVS by using the controller's \textsl{setClassicDEVS} method.

\subsection{Dynamic Structured DEVS}
Running a \textsc{Dynamic Structured DEVS} is largely similar to \textsc{Classic DEVS}, except a single core of the \textsl{DynamicCore} class has to be used.\\
Be sure to set the correct simulation type using the controller's \textsl{setDSDEVS} method.

\subsection{Parallel DEVS}
\textsc{Parallel DEVS} requires an extra \textbf{network} object, of the class \textsl{Network}, and one or more cores of the \textsl{Multicore} class.\\
Once again, don't forget to set the DEVS type by using the \textsl{setPDEVS} method.

\subsection{Adding models}
The final step of the setup would be to add any models to the controller using the \textsl{addModel} method. Your simulation is now ready to run!

\section{Control}
You can choose to let the simulation end at a certain moment by adding a \textbf{termination time} or a \textbf{termination condition}.\\
The termination time can be set using the controller's \textsl{setTerminationTime} method, the termination condition by using \textsl{setTerminationCondition} and passing a custom \textsl{TerminationFunctor}.\\ %TODO add manual terminationfunctor?
\\
Finally, you can start the DEVS using the controller's \textsl{simulate} method.
%TODO Serializatie: save, load

### 2015-04-23 manual/tex/reqs.tex

\section{System requirements}
\subsubsection{Compiler}
The project requires a recent version of a standards-compliant C++11 compiler and runtime. The following are compilers/runtimes used to develop the simulator (and therefore guaranteed to work)
\begin{itemize}
  \item G++ 4.8.2 + libstdc++
  \item G++ 4.9.2 + libstdc++
  \item Clang 3.4.0 + libstdc++
  \item Clang 3.5.0 + libstdc++
  \item G++ 4.9.2 on Cygwin $>=$ 1.7
\end{itemize}
Especially the C++11 threading support is essential.
\subsubsection{Dependencies}
The build system requires CMake $>=$ 2.8, the presence of standard build tools is assumed (make etc.).
The following are the functional dependencies:
\begin{itemize}
  \item Cereal (source included)
  \item GTest $>=$ 1.6
  \item Boost heap $>=$ 1.55  
\end{itemize}
Boost can be installed on *nix with the following command:\\
Debian based:\\
\lstinline!$sudo apt-get install libboost-all-dev!\\
Red-hat based:\\
\lstinline!$sudo yum install boost-devel!\\
\subsubsection{Preparing the build}
We use thread-sanitizer to verify thread-safety at runtime instrumented by the compiler (g++ and clang). This however requires building the project (and by extension any linked libraries) with position independent flags. For the simulator this is resolved by CMake, however for gtest you'll need to manually build GTest with this extra flag:\\
In file internal\textunderscore utils.cmake change compile options to append the compile flag -fpie :
set(cxx\textunderscore base \textunderscore flags "-Wall -Wshadow -fpie")\\
If you have any non-standard locations where the compiler should search for Boost and/or GTest, place them in the file localpreferences.txt.\\
\subsubsection{Building}
On *nix, a build script is provided to drive cmake:\\
\lstinline!$main/generate_build.sh {Debug|Release} {g++|clang++}!\\
Alternatively, invoke cmake manually :
\lstinline!$mkdir build!\\
\lstinline!$cd build!\\
\lstinline!$cmake -G"Eclipse CDT4 - Unix Makefiles" -DCMAKE_CXX_COMPILER_ARG1=!\\
\lstinline!"{-std=c++11|-std=gnu++11}" -DCMAKE_CXX_COMPILER="{clang++|g++}" !
\lstinline!-DCMAKE_BUILD_TYPE={Release|Debug} ../main!
\lstinline!$make -j{CPU-CORES}!\\
\subsubsection{Development}
The cmake invocation has created an Eclipse CDT project for you to open the simulator code in:
\begin{itemize}
  \item Open Eclipse Luna (with CDT plugin)
  \item File, Import, General
  \item Select "Existing projects into Workspace"
  \item navigate to the root folder of the project (containing ./main, ./build)
\end{itemize}

### 2015-04-23 manual/projectrapport.tex
First draft of project rapport.

\documentclass[8pt,a4paper]{report}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{url}
\usepackage{graphicx}
\usepackage[]{algorithm2e}

\title{Project Rapport\\\textsc{DEVS Ex Machina}}
\author{Ben Cardoen\\Pieter Lauwers\\Stijn Manhaeve\\Tim Tuijn\\Matthijs Van Os}
\date{}
\begin{document}
\maketitle
\tableofcontents{}

%
\chapter{Summary}
%
\section*{}
The project implements a C++ port of the PyPDEVS\cite{} project, extended with conservative synchronization and a change in system architecture from distributed simulation on several nodes (asymmetric mp) to an symmetric multiprocessing architecture. \\The summarized goal of the project is to provide an implementation of the offered DEVS paradigms with significantly lower execution time by levering the smp architecture.

\section*{Terminology}
The literature sometimes exchanges terms, to avoid any confusion the following frequent terms are listed with their associated meaning as used in this document:
\begin{itemize}
  \item Core : alias kernel. Single instance, containing a set of models it is responsible for, exchanges messages generated by those models and keeps time state. Can correspond with Logical Process in terms of what it represents.
  \item Logical Process : set of models with strong internal coherence.
  \item Distributed parallellism : computation distributed over several physically separated machines communicating by means of a network stack.
  \item SMP : symmetric multiprocessing, running a parallell computation on several distinct CPU cores/threads on the same machine with shared memory (and/or cache).
  \item message : any exchange of high level information between entities, user defined or for inter-kernel synchronization.
\end{itemize}

\chapter{Planning}
\section{Parallell}
\begin {itemize}
\item GVT integration : week 0-1, Ben and Tim
\item Conservative : week 2-3 , Ben and Tim
\end {itemize}
\section{General}
\begin {itemize}
\item Dynamic Structured : week 0-4 Stijn
\item DevStone and examples : week 0-4 Matthijs
\item ......
\end {itemize}

\chapter{Architecture Analysis}
\section{Implementation language}
Python is a very powerful language, but due to its interpreted nature it suffers from a performance point of view. Python's GIL \cite{} makes this issue even more serious whenever multithreading is required. Allthough it relieves the developer from tracking resource allocation, this comes at a price in the overhead caused by the GC. \\
In contrast, C++ offers low-level powerful thread primitives, but requires the developer to have a very deep understanding of parallell programming techniques. By using RAII, a C++ developer can leverage safe resource (de)allocation without losing performance.\\
A significant advantage Python offers a developer is relative cross-platform capability. This is non-trivial for a C++ developer, since at the very least the toolchains will be radically different in usage, and sometimes in implementation of the C++ standard. The edit-build-test cycle is also non-trivial for a large C++ application, whereas in Python this is negligible. In contrast, Python does not offer the programmer any of the C++ compiler's many compile time checks (static typechecks and program analysis).
\section{PyPDEVS analysis}
The PyPDEVS source code, partly due to Python's lax restrictions, violates a number of software design constraints. Not declaring attributes up until the point where they are needed (and not documenting them) makes for hard to read source code. The principle of encapsulation is rarely used throughout the Python code, for example the Solver class can freely swap out states from a model, forcing the classes to be bound tightly together. \\ % TODO add diplomacy
Python allows constructs such as self.content = ... , where contents type can be literally anything at runtime, something that is (without resorting to void*) impossible in C++. It also removes any protection from implementation mistakes.

\chapter{Design}
\section{Overview}
We chose early on to port the simulator by departing from the PyPDEVS implementation and creating a new structure, though keeping intact the structure in terms of functionality and interaction. Each class would use encapsulation to the fullest degree possible, and would be responsible only for task/functionality that belonged together. As an example, the C++ Core class is only responsible for driving the actual simulation steps (gathering messages, keeping time and transitioning models). It nevers interferes with modelstates (and thus needs no access whatsoever to those classes). A model is fully responsible for it's states and transitions, but beyond that has no state. A Controller combines both to create instances of the DEVS paradigms as and when needed, and provides a high level user interface (simulation control).

\section{Translating Dynamic Typing to Static Typing}
To represent objects that can be at runtime any type chosen by the user, we had 2 options : void* which is highly fragile in use, and impossible to serialize, or std::string(), which can double as an "anytype" by storing a serialized object inside the string, should this be required. This leaves the same degree of freedom dynamic typing grants to the user in PyPDEVS, with the added benefit of some checking provided by the compiler. Consider for example a user who stores an object serialized into the payload string of a message object, the compiler can in advance detect if the serializing object is at least serializable. Since these objects are not used by the kernel itself, the risk to corruption is minized, which is not the case should void* be used.

\section{Class decomposition}
The active entities in the project were discovered using several methods: use cases, role playing, argumentation and countargumentation and studying the existing implementation. This led to the following simplified decomposition:
\\ Model : provides DEVS paradigm implementation, hide details for user defined subclasses
\\ Core : drives transitions, exchanges messages, synchronizes
\\ Controller : Constructs simulator, controls simulation
\\ Tracer : Traces output
The other classes bind/create instances of the above (e.g. LocationTable, Allocator, Factory), or provide essential functionality (e.g. Message, Network, Scheduler)

\section{Network -- Kernel communication}
PyPDEVS uses MPI to exchange messages between kernels, whereas we could use shared memory to communicate between kernels (each running on 1 or more threads). We created a network architecture mimicking the approach used by PyPDEVS, without sacrificing the speed and ordering advantages of shared memory. The Network class consists of a single locked receiving port, and has for each existing kernel an outgoing queue. Receiving a message locks the network for the time it needs to place it on an integer indexed queue (very fast). If a kernel wants to receive messages it can do so if/when it is ready to do so, this only locks if at the same time a message is being received by the network. Several kernels can pull messages lockfree in parallel from the network (since the output queues are not shared). Early benchmarks made a throughput of 5e6 msg/s repeatable, with 4 kernels on modest hardware. The problems of (unintentional) out of order receipt and potential loss of messages is a non-issue in this context, since we use FIFO queues and loss of a message would implicate severe system failure (in which case message loss is the lesser problem).\\
Another advantage of hiding the message exchange behind a shared Network class instance is that, should this be required in the future, the kernel could run on different machines without requiring much redesigning of the system. An RPC implementation would simply have to implement the Network class on both sides.

\section{Tracing}
%TODO Stijn

\section{Models}
%TODO Tim

\section{Scheduler}
%TODO Ben

\section{Controller}
%TODO Matthijs

\section{Serialization}
%TODO Pieter

\chapter{Implementation}
\section{Correctness}
\subsubsection{Testing}
The use of the jenkins CI was heavily leveraged, in combination with extensive testing. Any feauture increment could only be merged into the project if it passed the developer's own jenkins instance.
\subsubsection{Thread races and deadlock}
Deadlocks and race conditions are very hard to debug, let alone prevent. To help us in this area as much as possible we used prototypes of threaded code in the tests before merging them into the main Classes, and each build is checked with ThreadSanitizer \cite{} which proved invaluable in preventing races and deadlocks in the threaded codepaths.
\subsubsection{Memory leaks and undefined behaviour}
With most entities represented as smart pointer, the need to ensure a leak free simulator nevertheless remained. To check for allocation misuse and a whole other series of errors we used Valgrind \cite{}. This meant for example that we detected a leak caused by circularly referenceing smart pointers.

\chapter{Functionality}
\section{Paradigms}
\subsection{Classic}
\subsection{Coupled}
\subsection{Dynamic Structured}
\subsection{Parallel}
\subsubsection{Optimistic}
Optimistic time synchronization requires an implementation of timewarp, and more extensive locking than would be required for conservative. The GVT algorithm (Mattern \cite{Mattern}) runs on a separate thread from the simulation kernel threads, so care must taken that they don't deadlock, but don't race either.
\subsubsection{Conservative}

\chapter{Obstacles}
\section{Virtualization}
Virtualization proved a significant hurdle by exposing potential faults. Standard compliant code was found to be leaking memory by not calling container member destructors but only on a VM with Linux as Guest and Windows as Host. This was never reprodicible in another setting.
\section{Operating System differences}
\section{Threading}
G++'s std::string implementation appears to be still COW, triggering an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object. Explicitly invoking the string constructor with the char[] proved to be a safe workaround here. Note that this problem is a known issue \cite{cow}
\section{Toolchains}
To develop we used the required toolchains. 
\section{Logger}
The recommended logger (g3log), while powerful and fast, proved a dead end since it does not compile on cygwin. 
Stijn therefore wrote his own implementation. %TODO Stijn
\chapter{Performance Analysis}
Whenever we made big design/architecture decisions, we first wrote some prototyping testcode to prove the viability (e.g. Network, Tracing, Logger).
\chapter{Conclusion}

\chapter{Appendix}

\begin{thebibliography}{1}

  \bibitem{cow} \url{https://gcc.gnu.org/bugzilla/show_bug.cgi?id=21334#c47}

  \bibitem{Mattern} \url{http://citeseerx.ist.psu.edu/viewdoc/summary?doi=10.1.1.111.2811}

  \end{thebibliography}
\end{document}

### 2015-04-23 manual/tex/advanced.tex
* started with logger/tracer manual

\section{Time}
\paragraph{Explain default time implementation}
\paragraph{Point to where the typedef can be changed}
\paragraph{Explain how users can create their own whacky implementation of the Time class for if the current version does not suit their needs}

\section{Tracing}
\paragraph{Briefly explain where the user can change the tracerset typedef}
\paragraph{Explain individual tracers}
\paragraph{Explain how the user can implement their own Tracer or OutputPolicy}

### 2015-04-23 manual/projectrapport.tex
Second draft.

@@
 \usepackage[latin1]{inputenc}
+\usepackage{xcolor}
+\usepackage{textcomp}
 \usepackage{amsmath}
 \usepackage{amsfonts}
+\usepackage{listings}
 \usepackage{amssymb}
+\usepackage[T1]{fontenc}
+\usepackage[scaled]{beramono}
+\newcommand\Small{\fontsize{9}{9.2}\selectfont}
+\newcommand*\LSTfont{\Small\ttfamily\SetTracking{encoding=*}{-60}\lsstyle}
 \usepackage{url}
+  \lstset{basicstyle=\ttfamily\tiny\color{blue}}
 \usepackage{graphicx}
@@
 The project implements a C++ port of the PyPDEVS\cite{} project, extended with conservative synchronization and a change in system architecture from distributed simulation on several nodes (asymmetric mp) to an symmetric multiprocessing architecture.
-\\The summarized goal of the project is to provide an implementation of the offered DEVS paradigms with significantly lower execution time by levering the smp architecture.
+\\The summarized goal of the project is to provide an implementation of the offered DEVS formalisms with significantly lower execution time by levering the smp architecture.
 
@@
   \item Logical Process : set of models with strong internal coherence.
-  \item Distributed parallellism : computation distributed over several physically separated machines communicating by means of a network stack.
-  \item SMP : symmetric multiprocessing, running a parallell computation on several distinct CPU cores/threads on the same machine with shared memory (and/or cache).
+  \item Distributed parallelism : computation distributed over several physically separated machines communicating by means of a network stack.
+  \item SMP : symmetric multiprocessing, running a parallel computation on several distinct CPU cores/threads on the same machine with shared memory (and/or cache).
   \item message : any exchange of high level information between entities, user defined or for inter-kernel synchronization.
@@
 \chapter{Planning}
-\section{Parallell}
+\section{Parallel}
 \begin {itemize}
@@
 \end {itemize}
+\section{Optimization}
+\section{Own Model}
 
@@
 Python is a very powerful language, but due to its interpreted nature it suffers from a performance point of view.
-Python's GIL \cite{} makes this issue even more serious whenever multithreading is required.
-Allthough it relieves the developer from tracking resource allocation, this comes at a price in the overhead caused by the GC.
+Python's GIL \cite{gil} makes this issue even more serious whenever multi-threading is required.
+Although it relieves the developer from tracking resource allocation, this comes at a price in the overhead caused by the GC.
 \\
-In contrast, C++ offers low-level powerful thread primitives, but requires the developer to have a very deep understanding of parallell programming techniques.
+In contrast, C++ offers low-level powerful thread primitives, but requires the developer to have a very deep understanding of parallel programming techniques.
 By using RAII, a C++ developer can leverage safe resource (de)allocation without losing performance.\\
 A significant advantage Python offers a developer is relative cross-platform capability.
-This is non-trivial for a C++ developer, since at the very least the toolchains will be radically different in usage, and sometimes in implementation of the C++ standard.
+This is non-trivial for a C++ developer, since at the very least the toolchains will be radically different in usage, and sometimes even in implementation of the C++ standard.
 The edit-build-test cycle is also non-trivial for a large C++ application, whereas in Python this is negligible.
 In contrast, Python does not offer the programmer any of the C++ compiler's many compile time checks (static typechecks and program analysis).
-\section{PyPDEVS analysis}
-The PyPDEVS source code, partly due to Python's lax restrictions, violates a number of software design constraints.
-Not declaring attributes up until the point where they are needed (and not documenting them) makes for hard to read source code.
-The principle of encapsulation is rarely used throughout the Python code, for example the Solver class can freely swap out states from a model, forcing the classes to be bound tightly together.
-\\ % TODO add diplomacy
-Python allows constructs such as self.content = ...
-, where contents type can be literally anything at runtime, something that is (without resorting to void*) impossible in C++.
-It also removes any protection from implementation mistakes.
+
+\section{Operating system support}
+The project requires a compliant C++11 compiler and library, but beyond that is not really bound to special features.
+For testing, GTest is a requirement, and the only two other dependencies are the ubiquitous Boost and Cereal, the last is provided with the project as it is header only.
+
+\section{Object representation}
+In Python everything is a reference, and since the DEVS formalisms require that the user be allowed to override most of the basic entities, the best representation for an object was a shared\textunderscore ptr.
+Passing by value could have been safer (e.g.
+between threads) but would introduce slicing.\\ Another advantage was that we would use as little memory as possible, for example sending a message means passing an const reference to a shared pointer, which is usually equal in size to 2 pointers.
+By passing by const ref we minimize most of the penalty usually involved in copyconstructing shared pointers.
+\\
+The only downside is the synchronized access to the reference count in the shared pointer.
+There is also a need for more boilerplate code to use shared pointers in containers (providing pass-through hash function, comparison etc).
 
@@
 \section{Overview}
-We chose early on to port the simulator by departing from the PyPDEVS implementation and creating a new structure, though keeping intact the structure in terms of functionality and interaction.
-Each class would use encapsulation to the fullest degree possible, and would be responsible only for task/functionality that belonged together.
-As an example, the C++ Core class is only responsible for driving the actual simulation steps (gathering messages, keeping time and transitioning models).
-It nevers interferes with modelstates (and thus needs no access whatsoever to those classes).
-A model is fully responsible for it's states and transitions, but beyond that has no state.
-A Controller combines both to create instances of the DEVS paradigms as and when needed, and provides a high level user interface (simulation control).
+We've slightly departed from the PyPDEVS structure, to allow for more encapsulation and decoupling of the core entities.
+This clear separation of data and responsibilities between classes made the threading part of the project easier (no or minimal shared state).
 
@@
 To represent objects that can be at runtime any type chosen by the user, we had 2 options : void* which is highly fragile in use, and impossible to serialize, or std::string(), which can double as an "anytype" by storing a serialized object inside the string, should this be required.
-This leaves the same degree of freedom dynamic typing grants to the user in PyPDEVS, with the added benefit of some checking provided by the compiler.
+\\This leaves the same degree of freedom dynamic typing grants to the user in PyPDEVS, with the added benefit of some checking provided by the compiler.
 Consider for example a user who stores an object serialized into the payload string of a message object, the compiler can in advance detect if the serializing object is at least serializable.
-Since these objects are not used by the kernel itself, the risk to corruption is minized, which is not the case should void* be used.
+\\ Misuse would either surface during compilation, or at worst could be checked with casting or exceptions.
+Misuse of void* could result in anything from undefined behaviour to stack corruption.
 
 \section{Class decomposition}
-The active entities in the project were discovered using several methods: use cases, role playing, argumentation and countargumentation and studying the existing implementation.
+The active entities in the project were discovered using several methods: use cases, argumentation and counterargumentation during the weekly meetings, and studying the existing implementation.
 This led to the following simplified decomposition:
-\\ Model : provides DEVS paradigm implementation, hide details for user defined subclasses
-\\ Core : drives transitions, exchanges messages, synchronizes
-\\ Controller : Constructs simulator, controls simulation
-\\ Tracer : Traces output
-The other classes bind/create instances of the above (e.g. LocationTable, Allocator, Factory), or provide essential functionality (e.g. Message, Network, Scheduler)
+\begin{enumerate}
+\item Model : provides DEVS formalism implementation of an Coupled/Atomic model.
+\item Core : drives transitions, exchanges messages, synchronizes
+\item Controller : Constructs simulator, controls simulation
+\item Tracer : Traces output
+\end{enumerate}
+The other classes bind/create instances of the above (e.g. LocationTable, Allocator, Factory), or provide essential functionality (e.g. Message, Network, Scheduler)\\
+For a full overview, please consult the documentation generated by doxygen.
 
@@
 Receiving a message locks the network for the time it needs to place it on an integer indexed queue (very fast).
-If a kernel wants to receive messages it can do so if/when it is ready to do so, this only locks if at the same time a message is being received by the network.
-Several kernels can pull messages lockfree in parallel from the network (since the output queues are not shared).
-Early benchmarks made a throughput of 5e6 msg/s repeatable, with 4 kernels on modest hardware.
-The problems of (unintentional) out of order receipt and potential loss of messages is a non-issue in this context, since we use FIFO queues and loss of a message would implicate severe system failure (in which case message loss is the lesser problem).\\
+If a kernel wants to receive messages it can do so if/when it is ready to do so, this only locks if at the same time a message is being received by the network.\\ Several kernels can pull messages lockfree in parallel from the network (since the output queues are not shared).
+Early benchmarks made a throughput of 5e5 msg/s repeatable, with 4 kernels on modest hardware.
+The problems of (unintentional) out of order receipt and potential loss of messages are a non-issue in this context, since we use FIFO queues and loss of a message would implicate severe system failure (in which case message loss is the lesser problem).\\
 Another advantage of hiding the message exchange behind a shared Network class instance is that, should this be required in the future, the kernel could run on different machines without requiring much redesigning of the system.
@@
 \section{Tracing}
+The tracing uses template programming, and is largely asynchronous w.r.t. the simulator kernels. This can be seen by the high utilization values when benchmarking the examples.
 %TODO Stijn
@@
 \section{Models}
+The models hide all implementation details for the particular DEVS formalisms, without limiting the user to extend into any subclassed model of his/her choice.
+The use can focus on what he/she wants the model to emulate, and not worry about for example locking, allocation etc.
 %TODO Tim
@@
 \section{Scheduler}
-%TODO Ben
+The scheduler from the prototypes was reused, with little to no changes (barring a few bugfixes). An unsynchronized version was implemented for comparison purposes.
 
@@
 \section{Serialization}
+For serialization we used the Cereal \cite{cereal} header only framework.
 %TODO Pieter
 
-\chapter{Implementation}
+\chapter{Non-functional requirements}
 \section{Correctness}
@@
 Deadlocks and race conditions are very hard to debug, let alone prevent.
-To help us in this area as much as possible we used prototypes of threaded code in the tests before merging them into the main Classes, and each build is checked with ThreadSanitizer \cite{} which proved invaluable in preventing races and deadlocks in the threaded codepaths.
+To help us in this area as much as possible we used prototypes of threaded code in the tests before merging them into the main classes, and each build is checked with ThreadSanitizer \cite{tsan} which proved invaluable in preventing races and deadlocks in the threaded codepaths.
 \subsubsection{Memory leaks and undefined behaviour}
 With most entities represented as smart pointer, the need to ensure a leak free simulator nevertheless remained.
-To check for allocation misuse and a whole other series of errors we used Valgrind \cite{}.
-This meant for example that we detected a leak caused by circularly referenceing smart pointers.
+To check for allocation misuse and a whole other series of errors we used Valgrind \cite{vg}.
+This meant for example that we detected a leak caused by circular referencing smart pointers.
 
-\chapter{Functionality}
-\section{Paradigms}
-\subsection{Classic}
-\subsection{Coupled}
+\subsubsection{Teamwork}
+We met at least once every week (barring the exam period), with each meeting having a formal report. Whenever required, teammembers used hangout to solve critical bugs or discuss design/coding issues.
+\subsubsection{Code review}
+Allthough there was no real formal code review done up to now, we've had several debug sessions reviewing each others code.
+This means the most critical used sections of the code are allready reasonably well reviewed, albeit not formally.
+
+\chapter{Functional requirements}
+\section{DEVS Formalisms}
+\section{Single core}
+The single core implementations follow exactly the runtime behaviour shown by the PyPDEVS simulator.
+We've chosen to emulate the Select() function with an integer priority value the user can set.
+This saves 1 virtual function call per transition and can quite easily be optimized by the compiler since it remains constant.
 \subsection{Dynamic Structured}
@@
 \subsubsection{Conservative}
+Conservative time synchronization is still under development.
 
@@
 This was never reprodicible in another setting.
-\section{Operating System differences}
 \section{Threading}
-G++'s std::string implementation appears to be still COW, triggering an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object.
+G++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}, triggering an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object.
 Explicitly invoking the string constructor with the char[] proved to be a safe workaround here.
@@
 \section{Toolchains}
-To develop we used the required toolchains. 
+To develop we used the required toolchains:\\ g++/clang++ as compiler, Eclipse+CDT as IDE, git for dvcs, jenkins as ci, cmake as build tool supplemented with some shell and python scripting.
 \section{Logger}
@@
 Whenever we made big design/architecture decisions, we first wrote some prototyping testcode to prove the viability (e.g. Network, Tracing, Logger).
-\chapter{Conclusion}
+\section{First comparison}
+The first performance results were obtained by running PyPDEVS's dynamic structured example against our own simulator, with a termination time of 3.6 E6.
+It should be noted that the GTEST overhead is not eliminated here yet.
+The perf \cite{perf} tool was used to obtain low level statistics with as much detail as possible.
+Unfortunately perf has very limited use on a virtualized OS.
+The parallel implementation was not yet feature complete, so allthough faster could not be fairly compared to PyPDEVS.
+\lstinputlisting[]{pydyn.txt}
+The following shows a recording of the C++ implementation of the same example.\\
+\lstinputlisting[]{cpdyn.txt}
+This result is still warped due to the startup/shutdown time of GTEST, but allready the C++ port is roughly 6x faster, only page faults and migrations are worse (but being worked on).
 
-\chapter{Appendix}
+Further optimizations included but are not limited to: 
+\begin{enumerate}
+\item Custom memory allocators/pools (boost, ...)
+\item Profiling guided recompilation
+\item Detecting bottlenecks with perf and gprof.
+\item ...
+\end{enumerate}
 
@@
 
+  \bibitem{gil} \url{https://wiki.python.org/moin/GlobalInterpreterLock}
+
+  \bibitem{tsan} \url{https://code.google.com/p/thread-sanitizer/}
+
+  \bibitem{vg} \url{http://valgrind.org/}
+
+  \bibitem{perf} \url{https://perf.wiki.kernel.org/index.php/Main_Page}
+
+  \bibitem{cereal} \url{https://uscilab.github.io/cereal/}
+
+  \bibitem{cpp} \url{http://www.open-std.org/jtc1/sc22/wg21/docs/papers/2014/n4296.pdf}
+
   \end{thebibliography}

### 2015-04-23 manual/tex/models.tex

\section{States}
You can can make your own states with almost no limitations. The only requirements for a user's own states are that they should be objects from a type that implements \textsl{n\_model::State}. This class has a few functions that have to be implemented:
\begin{itemize}
	\item \textsl{std::string toString()}
	\item \textsl{std::string toXML()}
	\item \textsl{std::string toCell()}
\end{itemize}
While members in sub-classes can be added without any problems, by default there is a public member \textsl{std::string m\_state} implemented in \textsl{State} that can be used to represent the user's state.

\section{Atomic models}
Atomic models can be made by implementing a sub-class of \textsl{n\_model::AtomicModel}. The most basic functions to be implemented are:
\begin{enumerate}
	\item \textsl{void AtomicModel::extTransition(const std::vector{\textless}n\_network::t\_msgptr{\textgreater} \& inputs)}\\
		Performs the model's external transition. The input argument is a vector with all messages at a certain point in time.
	\item \textsl{void AtomicModel::intTransition()}\\
		Performs the model's internal transition. 
	\item \textsl{t\_timestamp AtomicModel::timeAdvance()}\\
		Should return the model's current time advance value; the amount of time after which the following internal transition occurs. This function should not alter the model in any way.
	\item \textsl{std::vector{\textless}n\_network::t\_msgptr{\textgreater} AtomicModel::output()  const}\\
		Returns the model's output; all messages that are to be send should be returned via this function. This function should not alter the model in any way.
\end{enumerate}

The user can access the model's \textbf{state} at any time using the built-in function \textsl{t\_stateptr Model::getState() const}. The acquired state can then be examined. When implementing the functions \textsl{extTransition} and \textsl{intTransition}, you must make use of the built-in function \textsl{void Model::setState(const t\_stateptr\& newState)} to set a new state.\\

\textbf{Timestamps} that the user has to generate, for example in the \textsl{timeAdvance} function, can be easily constructed using the following constructor: \textsl{t\_timestamp(size\_t)}).\\

\textbf{Input and output ports} can be added to a model by following functions:
\begin{itemize}
	\item \textsl{t\_portptr Model::addInPort(std::string name)}
	\item \textsl{t\_portptr Model::addOutPort(std::string name)}
\end{itemize}
These functions return a pointer to the port that was added, so you can connect ports easier. More about that later on. \\

When \textbf{sending messages} (\textbf{output function}), there are a few rules you should take into account. The message you want to send can only be a string. This string should be given to the port you want to send it through. The port will then create the actual messages (that are send over the models) and return them to you in a vector. All these messages (from possibly different ports) should be joined in 1 vector and be returned in the output function. The following functions will allow you to do so:
\begin{itemize}
	\item \textsl{t\_portptr Model::getPort(std::string name) const}
	\item \textsl{std::vector{\textless}n\_network::t\_msgptr{\textgreater} Port::createMessages(std::string message)}
\end{itemize} 
Something like this would work: this-{\textgreater}getPort("OUT")-{\textgreater}createMessages(message); \\

\textbf{NOTE} that things like setting initial states, adding ports or setting other variables should be done before the model itself is used in any way. Incomplete models or undefined behaviour may be the result of said incorrect use.

\section{Coupled models}
Coupled models only require you to implement them as a sub-class of \textsl{n\_model::CoupledModel}.

\textbf{Adding sub-models} requires following function: \textsl{void CoupledModel::addSubModel(const t\_modelptr\& model)}.

\textbf{Creating input and output ports} is done in the same way as the atomic models.

\textbf{Connecting these ports} can be done with the function \textsl{void CoupledModel::connectPorts(const t\_portptr\& p1, const t\_portptr\& p2, t\_zfunc zFunction);}. This function connects port \textsl{p1} to port \textsl{p2}. An optional Z function can be added that will modify the messages during transit from one port to another.
It is possible to connect ports from a coupled model to one of it's direct atomic (or coupled) model children.

### 2015-04-23 main/src/network/timestamp.h
* moved t_timestamp definition to a separate file so that it is easier to change * added a bit of explanation for the tracers * added a very minimal preface - - -> perhaps we can better put that part somewhere else?

@@ -187,8 +187,11 @@ public:
 	}
 };
 
-// In practice, use this typedef.
-typedef Time<std::size_t, std::size_t> t_timestamp;
+} /* namespace n_network */
+//load the typedef from a different file
+#include "forwarddeclare/timestamp.h"
+
+namespace n_network {
 
 /**
  * Convenience function : make a TimeStamp object reflecting the current time.

### 2015-04-23 main/src/tracers/verbosetracer.h

@@ -59,7 +59,7 @@ public:
 	{
 		assert(adevs != nullptr && "VerboseTracer::tracesInit argument cannot be a nullptr.");
 
-		std::ostringstream* ssr = n_tools::createRawObject<std::ostringstream>();//we don't need a raw object
+		std::ostringstream* ssr = n_tools::createRawObject<std::ostringstream>();//we don't need a shared object
 
 		t_stateptr state = adevs->getState();
 		*ssr << "\n"

### 2015-04-23 manual/tex/models.tex

@@
 \section{Atomic models}
-Atomic models can be made by implementing a sub-class of \textsl{n\_model::AtomicModel}. The most basic functions to be implemented are:
+Atomic models can be made by implementing a sub-class of \\ \textsl{n\_model::AtomicModel}. The most basic functions to be implemented are:
 \begin{enumerate}
@@
 
-\textbf{Timestamps} that the user has to generate, for example in the \textsl{timeAdvance} function, can be easily constructed using the following constructor: \textsl{t\_timestamp(size\_t)}).\\
+\textbf{Timestamps} that the user has to generate, for example in the \textsl{timeAdvance} function, can be easily constructed using the following constructor:\\ \textsl{t\_timestamp(size\_t)}).\\
 
@@
 	\item \textsl{t\_portptr Model::getPort(std::string name) const}
-	\item \textsl{std::vector{\textless}n\_network::t\_msgptr{\textgreater} Port::createMessages(std::string message)}
+	\item \textsl{std::vector{\textless}n\_network::t\_msgptr{\textgreater} Port:: \\createMessages(std::string message)}
 \end{itemize} 
-Something like this would work: this-{\textgreater}getPort("OUT")-{\textgreater}createMessages(message); \\
+Something like this would work: this-{\textgreater}getPort("OUT")-{\textgreater} \\createMessages(message); \\
 
@@
 \section{Coupled models}
-Coupled models only require you to implement them as a sub-class of \textsl{n\_model::CoupledModel}.
+Coupled models only require you to implement them as a sub-class of \\ \textsl{n\_model::CoupledModel}.
 
-\textbf{Adding sub-models} requires following function: \textsl{void CoupledModel::addSubModel(const t\_modelptr\& model)}.
+\textbf{Adding sub-models} requires following function: \textsl{void CoupledModel:: \\addSubModel(const t\_modelptr\& model)}.
 

### 2015-04-23 manual/tex/models.tex
* added information about Dynamic Structured DEVS in the Models chapter

@@
 It is possible to connect ports from a coupled model to one of it's direct atomic (or coupled) model children.
+
+\section{Dynamic Structured DEVS}
+Both atomic models and coupled models can implement the following function:
+\begin{itemize}
+	\item \textsl{bool Model::modelTransition(DSSharedState* shared);}
+\end{itemize}
+The \textsl{shared} parameter gives access to a shared state. This state can be used to organize structural changes from the bottom up. It will not be wiped in between transitions.\\
+The return value indicates whether or not the simulator should look at the parent of the current model for performing Dynamic Structured DEVS.
+This way, not all models have to be consulted for structural changes.\\
+In the \textsl{modelTransition} function, models can add and remove connections, ports and submodels. Currently, only adding and removing connections is supported.

### 2015-04-24 manual/tex/advanced.tex

@@
+\lstset{
+    frameround=fttt,
+    language=C++,
+    breaklines=true,
+}
+
 \section{Time}
-\paragraph{Explain default time implementation}
-\paragraph{Point to where the typedef can be changed}
-\paragraph{Explain how users can create their own whacky implementation of the Time class for if the current version does not suit their needs}
+Time is a very important aspect of the simulator. Our implementation allows users to easily change the representation of time in order to fit their needs.
+\paragraph{Custom time object}
+You can change how time is represented by changing the \lstinline{t_timestamp} type definition in the \path{src/forwarddeclare/timestamp.h} header.
+Please take care that this type implements at least the interface provided by the default \lstinline{Time} class.
+For more information, take a look at the accompanying documentation.
 
 \section{Tracing}
-\paragraph{Briefly explain where the user can change the tracerset typedef}
-\paragraph{Explain individual tracers}
-\paragraph{Explain how the user can implement their own Tracer or OutputPolicy}
+The tracers generate output during the simulation.
+\label{sec:Tracing}
+\paragraph{The Tracing interface}
+The interface consists of only a handful methods.
+\paragraph{Custom set of tracers}
+If the default set of tracers is not satisfactory, then you can select your own set of tracers.
+The header \path{tracers.h} in the folder \path{src/forwarddeclare} contains the typedef for the default implementation.
+\begin{lstlisting}
+typedef Tracers<VerboseTracer<CoutWriter>> t_tracerset;
+\end{lstlisting}
+You can select as many tracers as you want by providing more template arguments.
+The following example sets the simulator to use two tracers.
+The first tracer will generate the default verbose output and write it to a file.
+The second tracer is a custom tracer.
+\begin{lstlisting}
+typedef Tracers<VerboseTracer<FileWriter>,
+                MyTracer> t_tracerset;
+\end{lstlisting}
+Note that some tracers might need special initialization prior to use, such as the tracers provided by us that make use of the \lstinline{FileWriter} policy.
+
+\paragraph{Provided Tracers}
+All provided tracers make use of the policy pattern to separate the output generation from sending that output to an output sink.
+Currently, there is a policy for sending the output directly to standard out (\lstinline{CoutWriter}) or writing it to a file(\lstinline{FileWriter}).
+Both policies can be found in the \path{src/tracers/policies.h} header.
+Note that in the case of \lstinline{FileWriter}, the tracer must be explicitly initialized before it can be used.
+For more implementations, refer to the provided documentation.\\
+Currently, the simulator provides only one tracer class: the \lstinline{VerboseTracer}.
+This tracer will generate verbose output that is comparable to the verbose output of the Python PDEVS implementation by Jentl Van Tendeloo.
+\paragraph{Implementing a custom tracer}
+\subparagraph{The tracing interface}
+The custom tracer must at least implement the following interface:
+\begin{lstlisting}
+interface Tracer: public TracerBase
+{
+public:
+  /**
+   * @brief Traces state initialization of a model
+   * @param model The model that is initialized
+   * @param time The simulation time of initialization.
+   */
+  void tracesInit(const t_atomicmodelptr& adevs,
+                  t_timestamp time);
+
+  /**
+   * @brief Traces internal state transition
+   * @param adevs The atomic model that just
+   *              performed an internal transition
+   * @precondition The model pointer is not a nullptr
+   */
+  void tracesInternal(const t_atomicmodelptr& adevs,
+                      std::size_t coreid);
+
+  /**
+   * @brief Traces external state transition
+   * @param model The model that just went
+   *              through an external transition
+   */
+  void tracesExternal(const t_atomicmodelptr& adevs,
+                      std::size_t coreid);
+
+  /**
+   * @brief Traces confluent state transition
+   * A confluent state transition is
+   * a simultaneous internal and external transition.
+   * @param model The model that just went
+   *              through a confluent transition
+   */
+  void tracesConfluent(const t_atomicmodelptr& adevs,
+                       std::size_t coreid);
+};
+\end{lstlisting}
+Even though it is not explicitly enforced, it is highly recommended to derive your custom implementation from the \lstinline{TracerBase} class and to allow the use of output policies.\\
+Note that the tracer must not print its output upon generation. To make sure that the output is still generated in the correct order, the tracer must register a \lstinline{TraceMessage}.
+\subparagraph{Registering a trace message}
+The \lstinline{TraceMessage} class is defined in the \path{src/tracers/tracemessage.h} header file.
+These messages allow the simulator to schedule output generation and purge unwanted messages upon rollback.
+They contain the following elements:
+\begin{itemize}
+	\item[timestamp] The time upon which this message is created. This time can be retrieved either from the model that is being traced or as a separate parameter.
+	\item[tracerID] Each tracer should have a unique ID.
+If multiple tracers are used, this ID makes sure that the tracers cannot interfere with eachother.
+If your tracer derives from \lstinline{TracerBase}, then this ID is automatically assigned to your tracer.
+It can be accessed via the \lstinline{getID()} member function.
+	\item[execute function] This function is executed when the trace output can be safely sent to an output sink.
+It is recommended to use \lstinline{std::bind<...>()}.
+The function must be of type \lstinline{void()}.
+	\item[coreID] Each simulation core has an ID. During the simulation, it can happen that some trace messages must be removed based on which core initiated the trace.
+	\item[delete function] This function is called when the \lstinline{TraceMessage} is destructed. If the tracer must allocate memory on the heap, this function can be used to free that memory.\\*
+		Just like the \emph{execute function}, this function must be of type \lstinline{void()}.
+\end{itemize}
+When the message is created, it can be scheduled with the free function \lstinline{void scheduleMessage(t_tracemessageptr message)}.\\
+Both the \lstinline{TraceMessage} class and the \lstinline{scheduleMessage} function are defined in the \path{src/tracers/tracemessage.h} header.
+For a more in-depth example, you can always take a look at the implementation of the \lstinline{VerboseTracer} implementation.

### 2015-04-24 manual/tex/models.tex

@@
 \end{itemize}
+If the user does not override the above functions, this will result in a logged error. Serialization prohibits abstract functions, which enforces this workaround.\\
 While members in sub-classes can be added without any problems, by default there is a public member \textsl{std::string m\_state} implemented in \textsl{State} that can be used to represent the user's state.
+\\ For serialization you have to declare (after class definition) 
+\begin{verbatim}
+CEREAL_REGISTER_TYPE(MyState);
+\end{verbatim}
 
@@
 		Performs the model's internal transition. 
-	\item \textsl{t\_timestamp AtomicModel::timeAdvance()}\\
+	\item \textsl{t\_timestamp AtomicModel::timeAdvance() const}\\
 		Should return the model's current time advance value; the amount of time after which the following internal transition occurs. This function should not alter the model in any way.
@@
 \end{enumerate}
+If these are not overridden, this will result in a logged assertion error.\\
 
@@
 \end{itemize} 
-Something like this would work: this-{\textgreater}getPort("OUT")-{\textgreater} \\createMessages(message); \\
+Something like this would work: this-{\textgreater}getPort("OUT")-{\textgreater} \\createMessages("my\_message"); \\
+
+It's important that a model name is unique over the whole simulation, and portname should be unique per model.\\
+
+For serialization you have to declare (after class definition)
+\begin{verbatim}
+CEREAL_REGISTER_TYPE(MyModel);
+\end{verbatim}
 
@@
 
-\textbf{Connecting these ports} can be done with the function \textsl{void CoupledModel::connectPorts(const t\_portptr\& p1, const t\_portptr\& p2, t\_zfunc zFunction);}.
-This function connects port \textsl{p1} to port \textsl{p2}.
+\textbf{Connecting these ports} can be done with the function\\ \textsl{void CoupledModel::connectPorts(const t\_portptr\& p1, const t\_portptr\& p2, t\_zfunc zFunction);}.\\ This function connects port \textsl{p1} to port \textsl{p2}.
 An optional Z function can be added that will modify the messages during transit from one port to another.

### 2015-04-24 manual/tex/reqs.tex

@@
 If you have any non-standard locations where the compiler should search for Boost and/or GTest, place them in the file localpreferences.txt.\\
+SET({GTEST|BOOST}\_ROOT "your path here").
+Make sure that for gtest, the listed path contains both the "include" folder and libraries in a dedicated folder "lib".
+For Boost the path needs to contain the "include" folder with the boost headers.
 \subsubsection{Building}

### 2015-04-24 manual/projectrapport.tex
Project rapport

@@
 \section*{}
-The project implements a C++ port of the PyPDEVS\cite{} project, extended with conservative synchronization and a change in system architecture from distributed simulation on several nodes (asymmetric mp) to an symmetric multiprocessing architecture.
+The project implements a C++ port of the PyPDEVS project, extended with conservative synchronization and a change in system architecture from distributed simulation on several nodes (asymmetric mp) to a symmetric multiprocessing architecture.
 \\The summarized goal of the project is to provide an implementation of the offered DEVS formalisms with significantly lower execution time by levering the smp architecture.
@@
 \begin {itemize}
-\item Dynamic Structured : week 0-4 Stijn
-\item DevStone and examples : week 0-4 Matthijs
-\item ......
+\item Dynamic Structured : week 1-3 Stijn
+\item DevStone and examples : week 0-2 Matthijs and Pieter
+\item Serialization : week 0 Pieter
+\item Simulation control : week 1-3 Matthijs and Pieter  
+\item Tracers : week 0 Stijn
+\item Benchmarking PyPDevs Adevs : week 3 Matthijs and Pieter
 \end {itemize}
 \section{Optimization}
-\section{Own Model}
+Week 4 , general review.
 
@@
 We've slightly departed from the PyPDEVS structure, to allow for more encapsulation and decoupling of the core entities.
-This clear separation of data and responsibilities between classes made the threading part of the project easier (no or minimal shared state).
+This clear separation of data and responsibilities between classes made the threading part of the
+project easier (no or minimal shared state).
+%TODO include UML overview
 
@@
 An RPC implementation would simply have to implement the Network class on both sides.
+\includegraphics[scale=0.2]{network.png}
 
@@
 \subsubsection{Teamwork}
-We met at least once every week (barring the exam period), with each meeting having a formal report. Whenever required, teammembers used hangout to solve critical bugs or discuss design/coding issues.
+We met at least once every week (barring the exam period), with each meeting having a formal report.
+Whenever required, teammembers used video conferencing to solve critical bugs or discuss design/coding issues.
 \subsubsection{Code review}
-Allthough there was no real formal code review done up to now, we've had several debug sessions reviewing each others code.
+Although there was no real formal code review done up to now, we've had several debug sessions reviewing each others code.
 This means the most critical used sections of the code are allready reasonably well reviewed, albeit not formally.
+\subsubsection{Portability}
+We chose for Cygwin instead of MinGW because std::thread is not supported on MinGW. 
+%TODO Tim gnu++11 and stuff
 
@@
 This saves 1 virtual function call per transition and can quite easily be optimized by the compiler since it remains constant.
+\subsection{DirectConnect}
+%TODO Directconnect : what is rootmodel, only atomics in core.
 \subsection{Dynamic Structured}
+This is implemented but removing models is not yet fully tested.
+%TODO Stijn
 \subsection{Parallel}
@@
 Standard compliant code was found to be leaking memory by not calling container member destructors but only on a VM with Linux as Guest and Windows as Host.
-This was never reprodicible in another setting.
+This was never reproducible in another setting.
 \section{Threading}
@@
 Unfortunately perf has very limited use on a virtualized OS.
-The parallel implementation was not yet feature complete, so allthough faster could not be fairly compared to PyPDEVS.
+The parallel implementation was not yet feature complete, so although faster could not be fairly compared to PyPDEVS.
 \lstinputlisting[]{pydyn.txt}

### 2015-04-24 manual/projectrapport.tex
Add benchmarks for all formalisms to report.

@@
 Optimistic time synchronization requires an implementation of timewarp, and more extensive locking than would be required for conservative.
-The GVT algorithm (Mattern \cite{Mattern}) runs on a separate thread from the simulation kernel threads, so care must taken that they don't deadlock, but don't race either.
+The GVT algorithm (Mattern \cite{Mattern}) runs on a separate thread from the simulation kernel threads, so care must be taken that they don't deadlock, but don't race either.
 \subsubsection{Conservative}
@@
 \section{First comparison}
-The first performance results were obtained by running PyPDEVS's dynamic structured example against our own simulator, with a termination time of 3.6 E6.
-It should be noted that the GTEST overhead is not eliminated here yet.
-The perf \cite{perf} tool was used to obtain low level statistics with as much detail as possible.
-Unfortunately perf has very limited use on a virtualized OS.
-The parallel implementation was not yet feature complete, so although faster could not be fairly compared to PyPDEVS.
+We use perf to get statistically sound benchmarks.
+\subsection{Classic}
+An endtime of 3.6e6 was set, and both examples run against each other.
+\lstinputlisting[]{pycla.txt}
+The following shows a recording of the C++ implementation of the same example.\\
+\lstinputlisting[]{cpcla.txt}
+\subsubsection{Conclusion}
+The C++ port is approximately 12.906 times faster.
+
+\subsection{Dynamic structured}
+An end-time of 3.6e6 was set, and both examples run against each other.
 \lstinputlisting[]{pydyn.txt}
@@
 \lstinputlisting[]{cpdyn.txt}
-This result is still warped due to the startup/shutdown time of GTEST, but allready the C++ port is roughly 6x faster, only page faults and migrations are worse (but being worked on).
+\subsubsection{Conclusion}
+This result is still warped due to the startup/shutdown time of GTEST, but allready the C++ port is roughly 7.05 times faster, only page faults and migrations are worse (but being worked on).
+\subsection{Parallel}
+With GVT not yet active, an endtime of 3.6e6 was used (as before).\lstinputlisting[]{pypar.txt}\
+The following shows a recording of the C++ implementation of the same example.\\
+\lstinputlisting[]{cppar.txt}
+\subsubsection{Conclusion}
+The C++ port is approximately 19.6 times faster.
 
+\section{Further work.}
 Further optimizations included but are not limited to: 

### 2015-04-25 manual/projectrapport.tex
small manual changes

@@
 The use can focus on what he/she wants the model to emulate, and not worry about for example locking, allocation etc.
-%TODO Tim
+We refer to the manual for more information about the creation of these models.
 
@@
 \subsubsection{Portability}
-We chose for Cygwin instead of MinGW because std::thread is not supported on MinGW. 
-%TODO Tim gnu++11 and stuff
+We develop on Unix, Windows and MacOSX.
+
+In Windows we chose the Cygwin platform instead of MinGW to compile our project on.
+This because we make use of the C++11 std::thread objects; MinGW doesn't support POSIX threads.
+When building the project with GTest on Windows you need to use the -std=gnu++11 flag instead of the standard -std=c++11 flag, because we need the GNU extensions to correctly link everything.
 
@@
 \subsection{DirectConnect}
-%TODO Directconnect : what is rootmodel, only atomics in core.
+%TODO Stijn/Matthijs Directconnect : what is rootmodel, only atomics in core.
 \subsection{Dynamic Structured}
@@
 \section{Logger}
-The recommended logger (g3log), while powerful and fast, proved a dead end since it does not compile on cygwin. 
+The recommended logger (g3log), while powerful and fast, proved a dead end since it does not compile on Cygwin. 
 Stijn therefore wrote his own implementation. %TODO Stijn

### 2015-04-25 manual/projectrapport.tex
project report, added design uml, NEEDS TO BE VERIFIED
Addition of short Documentation section to manual

@@
 
-\title{Project Rapport\\\textsc{DEVS Ex Machina}}
+\title{Project Report\\\textsc{DEVS Ex Machina}}
 \author{Ben Cardoen\\Pieter Lauwers\\Stijn Manhaeve\\Tim Tuijn\\Matthijs Van Os}
@@
 We've slightly departed from the PyPDEVS structure, to allow for more encapsulation and decoupling of the core entities.
-This clear separation of data and responsibilities between classes made the threading part of the
-project easier (no or minimal shared state).
-%TODO include UML overview
+This clear separation of data and responsibilities between classes made the threading part of the project easier (no or minimal shared state).
+See the \hyperref[simpledesign]{Simple Design figure \ref*{simpledesign}}.\\
+
+\begin{figure}[h!]
+	\makebox[\textwidth][c]{\includegraphics[width=1.5\textwidth]{Design.png}}
+	\caption{Simple Design}
+	\label{simpledesign}
+\end{figure}
 
@@
 The models hide all implementation details for the particular DEVS formalisms, without limiting the user to extend into any subclassed model of his/her choice.
-The use can focus on what he/she wants the model to emulate, and not worry about for example locking, allocation etc.
+The user can focus on what he/she wants the model to emulate, and not worry about for example locking, allocation etc.
 We refer to the manual for more information about the creation of these models.
@@
 \section{Controller}
-%TODO Matthijs
+For the controller, care was taken in making the interface both functional and minimal. The default configuration is quick to set up, and customization is possible should the user require it.
 

### 2015-04-25 main/src/model/port.cpp
Fix Race on strings in Port.cpp

@@ -11,6 +11,7 @@
 #include "cereal/types/vector.hpp"
 #include "cereal/types/memory.hpp"
 #include "cereal/types/polymorphic.hpp"
+#include "stringtools.h"
 #include "objectfactory.h"
 
 namespace n_model {
@@ -22,17 +23,21 @@ Port::Port(std::string name, std::string hostname, bool inputPort)
 
 std::string Port::getName() const
 {
-	return m_name;
+	return n_tools::copyString(m_name);
 }
 
 std::string Port::getFullName() const
 {
-	return m_hostname + "." + m_name;
+	//std::string tmp = m_hostname + "." + m_name;
+	std::string tmp = this->getHostName();
+	tmp.append(".");
+	tmp.append(this->getName());
+	return n_tools::copyString(tmp);
 }
 
 std::string Port::getHostName() const
 {
-	return m_hostname;
+	return n_tools::copyString(m_hostname);
 }
 
 bool Port::isInPort() const
@@ -147,7 +152,7 @@ const std::map<t_portptr, std::vector<t_zfunc> >& Port::getCoupledOuts() const
 }
 
 n_network::t_msgptr createMsg(const std::string& dest, const std::string& destP, const std::string& sourceP,
-        const std::string msg, t_zfunc& func)
+        const std::string& msg, t_zfunc& func)
 {
 	n_network::t_msgptr messagetobesend = std::make_shared<n_network::Message>(dest,
 	        n_network::t_timestamp::infinity(), destP, sourceP, msg);
@@ -173,7 +178,7 @@ std::vector<n_network::t_msgptr> Port::createMessages(std::string message)
 			t_zfunc& zFunction = pair.second;
 			std::string model_destination = pair.first->getHostName();
 //			std::string sourcePort = this->getFullName();
-			std::string destPort = pair.first->getFullName();
+			std::string destPort = n_tools::copyString(pair.first->getFullName());
 			n_network::t_timestamp dummytimestamp(n_network::t_timestamp::infinity());
 
 			// We now know everything, we create the message, apply the zFunction and push it on the vector
@@ -183,7 +188,7 @@ std::vector<n_network::t_msgptr> Port::createMessages(std::string message)
 		for (auto& pair : m_coupled_outs) {
 			std::string model_destination = pair.first->getHostName();
 //			std::string sourcePort = this->getFullName();
-			std::string destPort = pair.first->getFullName();
[…]

### 2015-04-25 manual/projectrapport.tex
Update Design.dia and png, smallish updates in tex.

--- 2015-04-25 manual/projectrapport.tex
@@
 \usepackage{amssymb}
+\usepackage{hyperref}
 \usepackage[T1]{fontenc}
@@
 Receiving a message locks the network for the time it needs to place it on an integer indexed queue (very fast).
-If a kernel wants to receive messages it can do so if/when it is ready to do so, this only locks if at the same time a message is being received by the network.\\ Several kernels can pull messages lockfree in parallel from the network (since the output queues are not shared).
-Early benchmarks made a throughput of 5e5 msg/s repeatable, with 4 kernels on modest hardware.
+If a kernel wants to receive messages it can do so if/when it is ready to do so, this only locks if at the same time a message is being received by the network.\\ Several kernels can pull messages without contention in parallel from the network (since the output queues are not shared).
+\\This is illustrated in \hyperref[networkdesign]{Network design figure \ref*{networkdesign}}.
+
+\begin{figure}[h!]
+	\makebox[\textwidth][c]{\includegraphics[width=\textwidth]{network.png}}
+	\caption{Network cooperation diagram}
+	\label{networkdesign}
+\end{figure} 
+Early benchmarks showed a repeatable throughput of 5e5 msg/s, with 4 kernels on modest hardware.
 The problems of (unintentional) out of order receipt and potential loss of messages are a non-issue in this context, since we use FIFO queues and loss of a message would implicate severe system failure (in which case message loss is the lesser problem).\\
@@
 An RPC implementation would simply have to implement the Network class on both sides.
-\includegraphics[scale=0.2]{network.png}
 
@@
 \section{Threading}
-G++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}, triggering an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object.
+libstdc++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}, triggering an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object.
 Explicitly invoking the string constructor with the char[] proved to be a safe workaround here.
-Note that this problem is a known issue \cite{cow}
+Note that this problem is a known issue \cite{cow}.
 \section{Toolchains}

### 2015-04-25 main/src/control/controller.h
Added doxygen briefs to controller, controllerconfig

@@ -82,57 +82,57 @@ public:
 	virtual ~Controller();
 
 	/**
-	 * Set an atomic model as the main model using the given allocator
+	 * @brief Set an atomic model as the main model using the given allocator
 	 */
 	void addModel(t_atomicmodelptr& atomic);
 
 	/**
-	 * Set a coupled model as the main model using the given allocator
+	 * @brief Set a coupled model as the main model using the given allocator
 	 */
 	void addModel(t_coupledmodelptr& coupled);
 
 	/**
-	 * Main loop, starts simulation
+	 * @brief Main loop, starts simulation
 	 */
 	void simulate();
 
 	/**
-	 * Sets the simulation type
+	 * @brief Sets the simulation type
 	 */
 	void setSimType(SimType type);
 
 	/**
-	 * Set simulation to be classic DEVS
+	 * @brief Set simulation to be classic DEVS
 	 */
 	void setClassicDEVS();
 
 	/**
-	 * Set simulation to be Parallel DEVS
+	 * @brief Set simulation to be Parallel DEVS
 	 */
 	void setPDEVS();
 
 	/**
-	 * Set simulation to be Dynamic Structure DEVS
+	 * @brief Set simulation to be Dynamic Structure DEVS
 	 */
 	void setDSDEVS();
 
 	/**
-	 * Set time at which the simulation will be halted
+	 * @brief Set time at which the simulation will be halted
 	 */
 	void setTerminationTime(t_timestamp time);
 
 	/**
-	 * Set condition that can terminate the simulation
+	 * @brief Set condition that can terminate the simulation
 	 */
 	void setTerminationCondition(t_terminationfunctor termination_condition);
 
 	/**
-	 * Set checkpointing interval
+	 * @brief Set checkpointing interval
 	 */
 	void setCheckpointInterval(t_timestamp interv);
 
 	/**
-	 * Start thread for GVT
+	 * @brief Start thread for GVT
 	 */
 	void startGVTThread();
 
@@ -183,37 +183,37 @@ public:
 
 private:
 	/**
-	 * Check if simulation needs to continue
+	 * @brief Check if simulation needs to continue
 	 */
 	bool check();
 
 	/**
-	 * Serialize all cores and models, dump tracer output
+	 * @brief Serialize all cores and models, dump tracer output
 	 */
 	void save(bool traceOnly = false);
 
 	/**
-	 * Simulation setup and loop using regular DEVS
[…]

### 2015-04-25 main/src/control/controllerconfig.h

@@ -71,7 +71,7 @@ struct ControllerConfig
 	virtual ~ControllerConfig();
 
 	/**
-	 * Create a Controller using the configuration options of this object
+	 * @brief Create a Controller using the configuration options of this object
 	 */
 	std::shared_ptr<Controller> createController();
 };

### 2015-04-25 manual/tracersReport.tex
* removed tracerID from the tracemessage * write the Tracer part of the report in a separate file so that it doesn't clutter the main file. (the thing is 7 pages long!) - - -> the main file contains a link to this second report - - -> If you want to, I can always include this file in the main report. I just thought that this was cleaner

\documentclass[12pt]{article}
\usepackage{authblk}
\usepackage{listings}
\usepackage[table,xcdraw]{xcolor}
\usepackage{hyperref}
\usepackage[section]{placeins}

\newcommand{\inlinecpp}[1]{\lstinline[language=c++]{#1}}
\lstset{
  basicstyle=\ttfamily,
  mathescape
}

\begin{document}
\lstset{language=C++}
\title{The Tracing package\\ \large a more in depth explanation}
\author{DEVS Ex Machina}
\date{\today}

\maketitle
This document contains a more in depth explanation of the inner workings of the tracing package. We decided not to add this directly to the report to keep the report clean and concise.\\
This document is not supposed to be a manual, but rather an explanation of the techniques used in this package.
\section{package structure}
The entire tracing package consists roughly of three parts. The first part is is the \inlinecpp{Tracers} class. This class contains a set of tracers and provides a layer of abstraction for managing these tracers. The second part are the individual tracers. Each tracer generates specific output for the events of the simulator. The final part deals with this output. It makes sure that the output is generated in the correct order.
\part*{Tracers class} 
\section{Class Structure}
The class structure can be best described as a recursive inheritance tree. Through the use of variadic template parameters\footnote{variadic template parameters: \href{http://en.wikipedia.org/wiki/Variadic_template}{Wikipedia}, \href{http://www.cplusplus.com/articles/EhvU7k9E/}{cplusplus.com}}, each superclass takes care of one template parameter and inherits from the class that will take care of the rest.\\
The base case is a class that does not contain a tracer object.\\
Because the recursive case has more specific template parameters, it gets precedence over the base case.
\paragraph{example}
\begin{lstlisting}[language=c++]
TracersTemplated<A, A, B> test;
\end{lstlisting}
The test object now contains 2 tracer objects with type \inlinecpp{A} and one tracer object of type \inlinecpp{B}.\\
The complete inheritance tree would look like this:
\begin{lstlisting}
TracersTemplated<> //base case
       $\triangle$
       |
TracersTemplated<B>
       $\triangle$
       |
TracersTemplated<A, B>
       $\triangle$
       |
TracersTemplated<A, A, B>
\end{lstlisting}
The base case defines the operations and values that \inlinecpp{TracersTemplated} should contain regardless of how many tracers are used.\\
The other classes only add one new tracer object that they own privately.

\section{Methods}
Certain methods deserve special attention because of the way they are implemented.
\subsection{Constructor}
The non-default constructor can be used to initialize all the tracers at once. It uses the copy constructor to initialize the tracer members. There is also a default constructor.
\subsection{getParent}
The private \inlinecpp{getParent} method returns a reference to the parent class part of the object, effectively granting access to the interface provided by that type. It works by casting the \inlinecpp{this} pointer to the appropriate pointer type and dereferencing it. The use of \inlinecpp{static_cast} over a c-style cast assures type correctness.
\paragraph{example}
To continue the example from earlier, \inlinecpp{TracersTemplated<A, A, B>} would return the reference to the \inlinecpp{TracersTemplated<A, B>} part of the object.

\subsection{getByID}
This function allows direct access to the n\textsuperscript{th} registered tracer.\\*
Standard C++ doesn't allow function member template specialization for a non-specialized class template\footnote{specialization rules: \href{http://en.cppreference.com/w/cpp/language/template_specialization}{cppreference.com}}, therefore we used a workaround.
\subsubsection{sfinae}
The implementation relies heavily on sfinae - Substitution Failure Is Not An Error\footnote{sfinae: \href{http://en.wikipedia.org/wiki/Substitution_failure_is_not_an_error}{Wikipedia}}.\\
Basically, you present the compiler a few options for implementations of a particular function template. Only one of those options should compile correctly with any given template arguments. When the compiler fails to substitute the template arguments in one option, it continues with the other options until it either finds a single suitable option. When no suitable option can be found, the result is a compiler error.\\
\subsubsection{implementation}
The implementation offers three very similar implementations of the \inlinecpp{getTracer} function:
\begin{lstlisting}[language=c++]
//option 1
template<std::size_t n>
typename std::enable_if<n <= sizeof...(TracerElems) && n != 0,
  typename std::tuple_element<n,
    std::tuple<T, TracerElems...>>::type>::type&
getByID()
{/*...*/}

//option 2
template<std::size_t n>
typename std::enable_if<n==0 ,
  typename std::tuple_element<n,
    std::tuple<T, TracerElems...>>::type>::type&
getByID()
{/*...*/}

//option 3
template<std::size_t n>
typename std::enable_if<sizeof...(TracerElems) < n, void*>::type
getByID() = delete;
\end{lstlisting}
The \inlinecpp{std::tuple_element<n, std::tuple<T, TracerElems...>>::type>} denotes the type of the tracer that should be returned and is not really relevant to this explanation.\\
The \inlinecpp{std::enable_if<Condition, type>} struct defines a typedef \lstinline{type}\lstinline{} if the condition evaluates to true. If the condition evaluates to false, the struct is empty.\\
Therefore, depending on the value of the template argument \inlinecpp{n}, the return value of one of the three options is defined and the others are not, resulting in a substitution error.\\
If n is too large, that is, n is larger than or equal to the amount of used tracers, the last option is taken. Because that function is deleted, the compiler\footnote{Only tested with GCC 4.9.2.} generates an error that points to where the function was invoked.\\
\textbf{Note} that you cannot use the \inlinecpp{>} operator in the \inlinecpp{std::enable_if} condition. The compiler gets confused because it fails to distinguish between the comparison operator and the closing angle bracket of the template.

\section{What about std::tuple?}
\inlinecpp{std::tuple<...>} already offers a lot of functionality that we implemented ourself. The reason why we chose to reinvent the wheel over using \inlinecpp{std::tuple<...>} is because our implementation makes it very easy to loop over all the registered tracer objects and call one of their functions.\\
Looping over a \inlinecpp{std::tuple<...>}, while possible\footnote{Looping \lstinline{std::tuple<>}: \href{http://stackoverflow.com/a/6894436}{Stack Overflow}}, involves a lot of boilerplate code for a simple function call.

\part*{Individual tracer}
A Tracer has to generate output for certain events and send it to some output sink.\\
Each individual tracer class has the exact same structure. Policy-based programming\footnote{Policy-based programming: \href{http://en.wikipedia.org/wiki/Policy-based_design}{Wikipedia}} is used to separate the output writing logic from the output creation logic. This allows the user to use their own output method without having to rewrite all the existing tracers.\\
\subsection{Policies}
An output policy dictates where the generated output will go to. At the moment, we have implemented a policy for writing the output to \inlinecpp{std::cout} and another policy for writing the output to a file.
\section{Planned changes}
At the moment, we have only implemented one tracer, the verbose tracer. The other tracers (XML, JSON and CELL) will have a very similar structure. We could just create a single base class and use virtual functions to provide the implementation of the tracer-specific functionality. However, this would defeat the purpose of custom designing the template classes we have already. Another option is to use the Curiously Recurring Template Pattern (CRTP)\footnote{CRTP: \href{http://en.wikipedia.org/wiki/Curiously_recurring_template_pattern}{Wikipedia}}.
\paragraph{Curiously Recurring Template Pattern} This pattern allows a base class to call functionality of a derived class without having to use virtual functions or function overloading. It is in effect a way to deal with this kind of polymorphic behavior at compile time. The compiler knows exactly which functions will be called by the base class and can therefore use optimization techniques that are not always available when using runtime polymorphism.

\part*{Output Scheduling}
In order to make sure that the output generation is consistent in between runs and doesn't contain any output that is reverted, the tracers don't immediately send the generated output to their output sink.
\section{Trace Message}
The tracers put their generated output in a message, much like the messages that the models can send to eachother. They contain the following information:
\begin{itemize}
	\item \emph{timestamp} The timestamp of the message dictates the ordering of the messages.
	\item \emph{execute function} When a trace message is handled, this function is executed. This function should send the output generated by the tracer to the output sink.
	\item \emph{coreID} Each simulation has a unique core ID. When this core requests a rollback, we can selectively purge only the trace messages that belong to this core. Otherwise, if one core does not participate in the rollback, for example when it did not receive an antimessage, its output is not touched.
	\item \emph{delete function} In some cases, memory could be allocated on the heap. This function can be used to free that memory. It will be called when the message is destructed, regardless of whether its \emph{execute function} was called.
\end{itemize}
Both the \emph{execute function} and the \emph{delete function} are objects of type \inlinecpp{std::function<void()>}. You can use \inlinecpp{std::bind<...>(...)} to create a delegate.

\section{Message handling}
Unlike in the Python PDEVS implementation, trace messages are not scheduled together with the other messages, but rather in a separate scheduler. The following free functions are defined for managing the messages
\begin{itemize}
	\item \inlinecpp{void scheduleMessage(t_tracemessageptr message);}\\*
			Schedule a single message.
	\item \inlinecpp{void traceUntil(n_network::t_timestamp time);}\\*
			Execute all trace messages with a timestamp less than the provided parameter.
	\item \inlinecpp{void revertTo(n_network::t_timestamp time, std::size_t coreID = std::numeric_limits<std::size_t>::max());}\\*
			Purges all trace messages from a core that have a timestamp greater than the timestamp \inlinecpp{time}. The default value of the second parameter will indicate that the messages of \emph{all} cores must be removed. The execute function of the trace messages will not be called.\\*
			\textbf{Note} Removing the messages of a single core is more costly than removing the messages of all cores.
	\item \inlinecpp{void clearAll();}\\*
			Clears all remaining trace messages. Their execute function will not be called.
	\item \inlinecpp{void waitForTracer();}\\*
			Waits until the tracing thread is finished.
\end{itemize}
In the background, \emph{traceUntil} will create a new thread that will perform the actual output. Therefore, rest of the simulation does not have to wait until this job is done. The current implementation can still be optimized further, for example by using a 1 reader/multiple writers approach. We will try to fix this for the beta version.

\end{document}

### 2015-04-25 main/src/test/tracerstest.cpp

@@ -418,28 +418,15 @@ TEST(tracing, tracerMessage) {
 
 	//test the message
 	TraceMessage::t_messagefunc boundFunc2 = std::bind(&testFunc, &intvar, 18);
-	TraceMessage msg = TraceMessage(t_timestamp(12u, 42u), 0, boundFunc2, 0u);
+	TraceMessage msg = TraceMessage(t_timestamp(12u, 42u), boundFunc2, 0u);
 	EXPECT_EQ(intvar, 42);
 	msg.execute();
 	EXPECT_EQ(intvar, 18);
-	EXPECT_TRUE(TraceMessage(t_timestamp(12u, 42u), 0, boundFunc2, 0u) < TraceMessage(t_timestamp(12u, 42u), 1, boundFunc2, 0u));
-	EXPECT_TRUE(TraceMessage(t_timestamp(12u, 42u), 1, boundFunc2, 0u) > TraceMessage(t_timestamp(12u, 42u), 0, boundFunc2, 0u));
-	EXPECT_TRUE(TraceMessage(t_timestamp(12u, 41u), 0, boundFunc2, 0u) < TraceMessage(t_timestamp(12u, 42u), 1, boundFunc2, 0u));
-	EXPECT_TRUE(TraceMessage(t_timestamp(12u, 41u), 1, boundFunc2, 0u) < TraceMessage(t_timestamp(12u, 42u), 0, boundFunc2, 0u));
-	EXPECT_TRUE(TraceMessage(t_timestamp(12u, 41u), 0, boundFunc2, 0u) < TraceMessage(t_timestamp(12u, 42u), 0, boundFunc2, 0u));
-	EXPECT_TRUE(TraceMessage(t_timestamp(12u, 41u), 1, boundFunc2, 0u) < TraceMessage(t_timestamp(12u, 42u), 1, boundFunc2, 0u));
-	EXPECT_TRUE(TraceMessage(t_timestamp(12u, 42u), 0, boundFunc2, 0u) > TraceMessage(t_timestamp(12u, 41u), 1, boundFunc2, 0u));
-	EXPECT_TRUE(TraceMessage(t_timestamp(12u, 42u), 1, boundFunc2, 0u) > TraceMessage(t_timestamp(12u, 41u), 0, boundFunc2, 0u));
[…]

### 2015-04-25 main/src/tracers/tracemessage.cpp

@@ -16,8 +16,8 @@ using namespace n_tools;
 
 namespace n_tracers {
 
-TraceMessage::TraceMessage(n_network::t_timestamp time, std::size_t tracerID, const t_messagefunc& func, std::size_t coreID, const t_messagefunc& takeback)
-	: Message("", time, "", ""), m_func(func), m_takeBack(takeback), m_tracerID(tracerID)
+TraceMessage::TraceMessage(n_network::t_timestamp time, const t_messagefunc& func, std::size_t coreID, const t_messagefunc& takeback)
+	: Message("", time, "", ""), m_func(func), m_takeBack(takeback)
 {
 	assert(m_func != nullptr && "TraceMessage::TraceMessage can't accept nullptr as execution function");
 	assert(m_takeBack != nullptr && "TraceMessage::TraceMessage Can't accept nullptr as cleanup function. If you don't need a cleanup function, either provide an empty one or omit the argument.");
@@ -37,9 +37,7 @@ void TraceMessage::execute()
 bool TraceMessage::operator <(const TraceMessage& other) const
 {
 	if (this->getTimeStamp() == other.getTimeStamp()){
-		if(this->m_tracerID == other.m_tracerID)
-			LOG_ERROR("TraceMessage::operator< Received two messages with the exact same timestamp and tracer ID");
-		return this->m_tracerID < other.m_tracerID;
+		return this < &other;
 	}
 	return (this->getTimeStamp() < other.getTimeStamp());
 }
@@ -47,9 +45,7 @@ bool TraceMessage::operator <(const TraceMessage& other) const
 bool TraceMessage::operator >(const TraceMessage& other) const
 {
 	if (this->getTimeStamp() == other.getTimeStamp()){
-		if(this->m_tracerID == other.m_tracerID)
-			LOG_ERROR("TraceMessage::operator> Received two messages with the exact same timestamp and tracer ID");
-		return this->m_tracerID > other.m_tracerID;
+		return this > &other;
 	}
 	return (this->getTimeStamp() > other.getTimeStamp());
 }
@@ -143,7 +139,7 @@ void traceUntil(n_network::t_timestamp time)
 {
 	std::lock_guard<std::mutex> guard(mu);
 	std::vector<TraceMessageEntry> messages;
-	TraceMessage t(time, std::numeric_limits<std::size_t>::max(), []{}, 0u);
[…]

### 2015-04-25 main/src/tracers/tracerbase.h

@@ -18,25 +18,13 @@ namespace n_tracers {
 class TracerBase
 {
 private:
-	const std::size_t m_id;
-	static std::size_t m_lastid;
 
 protected:
 	TracerBase()
-		: m_id(m_lastid++)
 	{
 	}
 
 public:
-	/**
-	 * @return the ID of this tracer.
-	 * Each tracer has a unique ID that can be used to order
-	 * TracerMessages of different tracers for the same event and stuff like that.
-	 */
-	std::size_t getID() const
-	{
-		return m_id;
-	}
 };
 
 } /* namespace n_tracers */

### 2015-04-25 manual/projectrapport.tex
* removed tracerID from the tracemessage * write the Tracer part of the report in a separate file so that it doesn't clutter the main file. (the thing is 7 pages long!) - - -> the main file contains a link to this second report - - -> If you want to, I can always include this file in the main report. I just thought that this was cleaner
Missing commandline in rapport.
Project report: added serialization

@@
 \section{Serialization}
-For serialization we used the Cereal \cite{cereal} header only framework.
-%TODO Pieter
+For serialization we used the Cereal \cite{cereal} header-only framework.
+We have used separate load and save functions for full flexibility.
+For each type that occurs in the form of a pointer we implemented a static load and construct function since the object first has to be created before the saved data can be imported.
+Cereal has a small coupling with the code (models) that has to be written by the user as the user has to register his own classes due to polymorphism.
 

### 2015-04-25 main/src/network/message.h
* fix pDEVS wrong trace output on revert
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>

@@ -31,7 +31,7 @@ operator<<(std::ostream& os, const MessageColor& c);
  */
 class Message
 {
-private:
+protected:
 	/**
 	 * Unique model name of target.
 	 */

### 2015-04-25 main/src/tracers/tracemessage.cpp

@@ -19,6 +19,7 @@ namespace n_tracers {
 TraceMessage::TraceMessage(n_network::t_timestamp time, const t_messagefunc& func, std::size_t coreID, const t_messagefunc& takeback)
 	: Message("", time, "", ""), m_func(func), m_takeBack(takeback)
 {
+	m_timestamp.increaseCausality(1u);
 	assert(m_func != nullptr && "TraceMessage::TraceMessage can't accept nullptr as execution function");
 	assert(m_takeBack != nullptr && "TraceMessage::TraceMessage Can't accept nullptr as cleanup function. If you don't need a cleanup function, either provide an empty one or omit the argument.");
 	setSourceCore(coreID);
@@ -154,7 +155,7 @@ void traceUntil(n_network::t_timestamp time)
 void revertTo(n_network::t_timestamp time, std::size_t coreID)
 {
 	std::vector<TraceMessageEntry> messages;
-	TraceMessage t(time, []{}, 0u);
+	TraceMessage t(time.getTime(), []{}, 0u);
 	scheduler->unschedule_until(messages, &t);
 	std::vector<TraceMessageEntry> messagesLost;
 	TraceMessage inf(n_network::t_timestamp::infinity(), []{}, 0u);

### 2015-04-25 manual/projectrapport.tex
* added very minimal explanation for dynamic structured to project report functional requirements

--- 2015-04-25 manual/projectrapport.tex
@@
 \documentclass[8pt,a4paper]{report}
+\usepackage{hyperref}
 \usepackage[latin1]{inputenc}
@@
 The tracing uses template programming, and is largely asynchronous w.r.t. the simulator kernels. This can be seen by the high utilization values when benchmarking the examples.
-%TODO Stijn
+The tracing component makes heavy use of template programming. For more detailed information, please refer to \href{run:./tracersReport.pdf}{tracersReport.pdf}.
 
@@
 \subsection{Dynamic Structured}
-This is implemented but removing models is not yet fully tested.
-%TODO Stijn
+Only the removal/addition of connections is currently supported. Unlike the PyPDEVS, we will try to keep a removed model intact when it is removed from the simulation.
 \subsection{Parallel}

### 2015-04-25 manual/projectrapport.tex

--- 2015-04-25 manual/projectrapport.tex
@@
 \section{First comparison}
-We use perf to get statistically sound benchmarks.
+We use perf to get statistically sound benchmarks. The command run was :
+\lstinline!$perf stat -r {rounds} {command}!\\
+
 \subsection{Classic}

### 2015-04-25 manual/projectrapport.tex
* added very minimalist logger benchmark

@@
 
+\section{Logger}
+We have also conducted a small benchmark to compare the raw power of our logger and the proposed g3log.
+We let each logger print several hundred thousand messages to a file. These are the performance statistics:
+\subsubsection{g3log}
+\lstinputlisting[]{g3log.perf}
+\subsubsection{Our own logger}
+\lstinputlisting[]{own.perf}
+\subsubsection{Conclusion}
+While the difference is considerable, we should keep in mind that g3log generates timestamps for each message.
+It also contains several features that our logger doesn't, such as crash safety.
+Moreover, g3log allows the user to dynamically change which log levels are filtered, while this can only be decided at compile time for our logger.
+This can be clearly seen in the amount of branches.
+
 \section{Further work.}

### 2015-04-25 manual/tex/controller.tex
* moved t_timestamp definition to a separate file so that it is easier to change * added a bit of explanation for the tracers * added a very minimal preface - - -> perhaps we can better put that part somewhere else?

@@
 
-First of these is the \textbf{allocator}, which decides on which simulation core to place which model.
-We have already included a simple allocator that distributes models evenly, but for more complex behavior it is possible to create one tailored specifically for your simulation purposes by inheriting from the \textsl{Allocator} base class.\\
+The first part to take into consideration is the \textbf{allocator}, which decides on which simulation core to place which model.
+We have already included a simple allocator (called \textsl{SimpleAllocator}) that distributes models evenly, but for more complex behavior it is possible to create one tailored specifically for your simulation purposes by inheriting from the \textsl{Allocator} base class.\\
 
-The next step is to create a \textbf{location table} and a table for the simulation cores. The former stores the location of all models. A \textsl{LocationTable} class is included for this purpose.\\
-The latter needs to be an \textsl{std::unordered\_map{\textless}std::size\_t, t\_coreptr{\textgreater}}. This map will store a pointer to each core which can be accessed using the core's ID.\\
+To get output from your simulation, you can select one or more of the included \textbf{tracers}. See \ref{sec:Tracing} for more details.\\
+\\
+The easiest way to get started is using the included \textsl{ControllerConfig}. This object will take care of a large part of the setup for you, though it is also possible to do everything manually.\\
 
-To get output from your simulation, you can select one or more of the included \textbf{tracers}. [...] TODO\\
+\textsl{ControllerConfig} has a few options to configure, all of them with a certain default value.
+\\\\
+\begin{tabular}{lp{8cm}}
+\textit{name} & The name of your simulation \newline
+				By default this is \texttt{MySimulation}
+\\\\
+\textit{simType} & The type of DEVS you wish to run:
 
-The following steps differ slightly depending on which type of DEVS you want to run.
+	\begin{tabular}{l|l}
+	\textsf{Option} & \textsf{Meaning} \\
+	\texttt{Controller::CLASSIC} & \emph{Classic DEVS} \\
+	\texttt{Controller::DSDEVS} & \emph{Dynamic Structure DEVS} \\
+	\texttt{Controller::PDEVS} & \emph{Parallel DEVS} \\
+	\end{tabular} \newline
 
-\subsection{Classic DEVS}
-\textsc{Classic DEVS} depends on the use of a single \textbf{core} of the \textsl{Core} class. Create a core of this class and add it to the \textsl{unordered\_map} created before.\\
+	The default is \emph{Classic DEVS}
+\\\\
+\end{tabular}
+\\
+\begin{tabular}{lp{8cm}}
+\textit{coreAmount} & The amount of cores the simulation will run on \newline
+						\textit{[This option is disregarded if you do not run PDEVS]} \newline
+						By default this is 1
+\\\\
+\textit{saveInterval} & This value decides how often the simulation is saved and the tracing output dumped. A value of 3 would mean that this happens every 3 simulation cycles. \newline
+						The default value is 5
+\\\\
+\textit{allocator} & By default this is the included \textsl{SimpleAllocator}
+\\\\
+\end{tabular}\\
+\\
+After you are done setting the right options, you can generate a \textbf{controller} by using \textsl{ControllerConfig}'s \textsl{createController} method.
+All further interactions with your simulation are done through this controller.\\
+\\
+As a short example, this configuration would give you a Classic DEVS named \texttt{foosim} that saves every 8 cycles:
+\begin{verbatim}
+ControllerConfig conf = ControllerConfig();
+conf.name = "foosim";
+conf.saveInterval = 8;
 
-Next, we create the \textbf{controller} of the simulation, using the \textsl{Controller} class.
-It expects a name, the core table, the allocator, the location table, and the tracers as its arguments.
-You can now set the simulation mode to CDEVS by using the controller's \textsl{setClassicDEVS} method.
-
-\subsection{Dynamic Structured DEVS}
-Running a \textsc{Dynamic Structured DEVS} is largely similar to \textsc{Classic DEVS}, except a single core of the \textsl{DynamicCore} class has to be used.\\
-Be sure to set the correct simulation type using the controller's \textsl{setDSDEVS} method.
-
-\subsection{Parallel DEVS}
-\textsc{Parallel DEVS} requires an extra \textbf{network} object, of the class \textsl{Network}, and one or more cores of the \textsl{Multicore} class.\\
-Once again, don't forget to set the DEVS type by using the \textsl{setPDEVS} method.
-
-\subsection{Adding models}
+auto controller = conf.createController();
+\end{verbatim}
 The final step of the setup would be to add any models to the controller using the \textsl{addModel} method. Your simulation is now ready to run!

### 2015-04-25 manual/projectrapport.tex
* added a small paragraph about the logger - - -> Do I need to provide more information there or is this enough?

@@
 The recommended logger (g3log), while powerful and fast, proved a dead end since it does not compile on Cygwin. 
-Stijn therefore wrote his own implementation. %TODO Stijn
+Stijn therefore wrote his own implementation.\\
+This implementation only covers the most basic needs:
+\begin{itemize}
+	\item asynchronous output
+	\item thread safe
+	\item multiple logging levels currently, the following 4 levels are supported:\\
+		\emph{INFO}, \emph{DEBUG}, \emph{WARNING} and \emph{ERROR}
+	\item levels can be individually switched on/off at compile time
+	\item No logging code is executed whatsoever if the logger is disabled
+	\item log messages contain filename and line number of the log command
+\end{itemize}
+It therefore lacks certain key features such as
+\begin{itemize}
+	\item crash safe
+	\item custom assert that will log an error before shutting down
+\end{itemize}
+Chances are that we will wait with implementing these features until we actually need them.
 \chapter{Performance Analysis}

### 2015-04-25 main/src/control/controller.cpp
Remove causality from setTimeStamp in Core Add more logging to detect deadlock/inf loop. Split network. Add missing documentation.

@@ -438,8 +438,8 @@ void cvworker(std::condition_variable& cv, std::mutex& cvlock, std::size_t myid,
 				if(threadsignal[i]==Controller::ThreadSignal::IDLE || threadsignal[i]==Controller::ThreadSignal::STOP)
 					++countidle;
 			}
-			if(countidle==threadsignal.size()){
-				LOG_INFO("CVWORKER: Thread ", myid, " for core ", core->getCoreID(), " all threads are stopped or idle, quitting.");
+			if(countidle==threadsignal.size()){	// TODO need link to network here.
+				LOG_INFO("CVWORKER: Thread ", myid, " for core ", core->getCoreID(), " all other threads are stopped or idle, quitting.");
 				return;
 			}
 		}else{
@@ -458,7 +458,7 @@ void cvworker(std::condition_variable& cv, std::mutex& cvlock, std::size_t myid,
 		/// Possible problem : IDLE overwrites SHOULDWAIT.
 		/// Better solution : threadsignal vector for IDLE,STOP,WORKING
 		///		      controlsignal vector for FREE/SHOULDWAIT/ISWAITING
-		bool skip_barrier = false;
+		bool skip_barrier = true;	// TODO re-enable if control is implemented
 		{
 			std::lock_guard<std::mutex> signallock(vectorlock);
 			// Case 1 : Main has asked us by setting SHOULDWAIT, tell main we're ready waiting.

### 2015-04-25 main/src/model/multicore.h

@@ -240,8 +240,6 @@ public:
 	unlockMessages()override;
 };
 
-void calculateGVT(/* access to cores,*/ size_t ms, std::atomic<bool>& run);
-
 } /* namespace n_model */
 
 #endif /* SRC_MODEL_MULTICORE_H_ */

### 2015-04-25 main/src/model/v.h

@@ -1,6 +1,5 @@
 /*
  * v.h
- *
  *  Created on: 10 Apr 2015
  *      Author: Ben Cardoen
  */
@@ -12,6 +11,8 @@ namespace n_model {
 
 /**
  * Thread-shared messagecount vector.
+ * Used in Mattern's algorithm, together with Count vector stores sent/recd messages to
+ * check for transient messages.
  */
 class V
 {

### 2015-04-25 manual/tex/reqs.tex
Addition of short Documentation section to manual

@@
 \end{itemize}
+
+\subsubsection{Documentation}
+Generating the source documentation is optional and requires Doxygen to be installed. UML and other diagrams will only be created if Graphviz is installed.\\
+After building the source, enter the \textsl{build} folder and enter the command 
+\begin{verbatim}
+$ make doc
+\end{verbatim} This will generate all documentation in the \textsl{doc} folder.

### 2015-04-25 manual/tex/advanced.tex

@@
 	\item[timestamp] The time upon which this message is created. This time can be retrieved either from the model that is being traced or as a separate parameter.
-	\item[tracerID] Each tracer should have a unique ID.
-If multiple tracers are used, this ID makes sure that the tracers cannot interfere with eachother.
-If your tracer derives from \lstinline{TracerBase}, then this ID is automatically assigned to your tracer.
-It can be accessed via the \lstinline{getID()} member function.
 	\item[execute function] This function is executed when the trace output can be safely sent to an output sink.

### 2015-04-25 manual/projectrapport.tex

@@
 \subsection{DirectConnect}
-%TODO Stijn/Matthijs Directconnect : what is rootmodel, only atomics in core.
+The directConnect algorithm has gone relatively unchanged from its implementation in PyPDEVS, though the role of the RootDEVS class has been simplified somewhat.
+Its main function has been reduced to executing directConnect and then storing the resulting list of connected atomic models.
 \subsection{Dynamic Structured}

### 2015-04-25 manual/tex/advanced.tex
* added some introduction text to the tracers part of the manual
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>

@@
 \section{Tracing}
-The tracers generate output during the simulation.
 \label{sec:Tracing}
-\paragraph{The Tracing interface}
-The interface consists of only a handful methods.
+The tracers generate output during the simulation. The default set of tracers will print verbose output to the standard output stream \lstinline{std::cout}.
 \paragraph{Custom set of tracers}

### 2015-05-17 manual/projectrapport.tex
Removed double image, removed tracersReport.pdf from git as the other pdf's are also not comitted.
Cleanup todo's and comments, add information of conservative to project rapport.

--- 2015-04-25 manual/projectrapport.tex
@@
 \usepackage{graphicx}
-\usepackage[]{algorithm2e}
+%\usepackage[]{algorithm2e}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{pifont}
 
@@
 The project implements a C++ port of the PyPDEVS project, extended with conservative synchronization and a change in system architecture from distributed simulation on several nodes (asymmetric mp) to a symmetric multiprocessing architecture.
-\\The summarized goal of the project is to provide an implementation of the offered DEVS formalisms with significantly lower execution time by levering the smp architecture.
+\\The summarized goal of the project is to provide an implementation of the offered DEVS formalisms with significantly lower execution time by leveraging the smp architecture.
 
@@
   \item message : any exchange of high level information between entities, user defined or for inter-kernel synchronization.
+  \item lookahead : Timespan relative to current state where the output of the model is not influenced by its input. (as per definition in \cite{cons}, app B)
+  \item eot : Earliest output time. Timepoint in future where the core/kernel will output events.
+  \item eit : Earliest input time. Timepoint in future where the core/kernel expects to
+  be change state on receiving event.
 \end{itemize}
+Lookahead is collected per model per kernel, eit/eot are based on information from models but determined on a core/kernel level.
 
@@
 There is also a need for more boilerplate code to use shared pointers in containers (providing pass-through hash function, comparison etc).
+
+\section{Messaging Size}
+A message is constructed once, and after that never copied.
+Messages are sent and received by reference to smart pointer, incurring a size cost of (on 64-bit platforms) 128bit.
+The payload (and its size) is irrelevant to the simulation up to the point where the message is actually received.
+A default implementation allows the payload stored as a std::string, but this can be overridden by way of templated code.
+%TODO Stijn.
 
@@
 \subsubsection{Conservative}
-Conservative time synchronization is still under development.
+The implementation of the conservative formalism follows the algorithm as described in \cite{cons}. \\
+\paragraph{Algorithm}
+The following listing briefly states the steps required in the conservative synchronization algorithm.\\
+Initialize eit,eot as (0,0) per kernel.\\
+A Time object has an absolute first field , followed by a causal second field.\\
+\begin{algorithm}
+\caption{Conservative PDEVS}
+\label{CHalgorithm}
+\begin{algorithmic}
+\Procedure{CPDEVS Modified}{}
+\State 0.1 Construct a depency graph, determining which models influence this kernel directly
+\State 0.2 If a msg is sent : eot = (timestamp > eot) ? timestamp : eot
+\While{Termination trigger not reached}
+\State 1. Simulate until vtime $>=$ eit
+\State 1.a If msg is sent during sim : eot = max(timestamp, eot)
+\State 2.a Time x = (eit + min lookahead, 0)
+\State 2.b Time y = (msg sent) ? (eit,1) : (time next transition)
+\State 2.c Time eot = std::min(x,y)
+\State 2.d Share eot value with other kernels.
+\State 3   Time eit = min ( $eot_i$ for i in influencing kernels)
+\EndWhile
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}\\
+\subparagraph{Differences with ADEVS}
+A big difference between ADEVS' and our implementation is that we hold the eot-values of all kernels in shared memory.
+This means that step 1.a , which is normally done by the receiver when it inspects the message queue, can be done immediately.
+This also avoids the queueing delay incurred by simulation traffic.\\
+Using the shared memory approach, we also avoid the sending of any synchronization message to other cores (required in ADEVS).
+\\In step 2.d we simply overwrite our old EOT value with the newly calculated.
+The shared EOT vector is only written at 1 position for each kernel , and only the influencing EOT values are ever read.
+\\
+\subparagraph{Dependency calculation}
+The author of the models needs to supply the kernel, for each model, with a list of models by whom it is influenced.
+Receiving a message that alters a model's output is defined as influencing in this context.
+Using the allocation information provided by the Controller, the Kernel will them determine which other Kernels (including possibly itself) it is dependent on.\\
+Note however that dependency here is in practice non-transitive, a model needs only to list those models that it directly receives events from, not separated by several models in between. \\
+This last point requires clarification, suppose we have 3 models A, B, C with A sending output to B, and B to C.
+C is indirectly influenced by A, but need not record this information (which is non trivial to retrieve for a non-trivial model, disregarding cycles for the moment).
+\\
+Let A be allocated @ kernel 0, B @ 1, C @ 2.
+While the dependency is transitive, since kernel 1 is already bound by kernel 0, kernel 2 does not require kernel 0's information (eot).
+By moving this transitivity to the kernel level, the model's author only has to look at 1-edge distant dependencies.
+\\
+At construction a Conservative Kernel will collect , for each model it holds, the list of names of influencing models.
+Next, the kernel requests for each unique name the allocation information (O(1) lookup).
+Note that this can be optimized if the model stores its allocation information, but this then prohibits runtime reallocation.
 
@@
 Explicitly invoking the string constructor with the char[] proved to be a safe workaround here.
-Note that this problem is a known issue \cite{cow}.
+Note that this problem is a known issue \cite{cow}, which is solved in GCC 5.1
 \section{Toolchains}
@@
 
+  \bibitem{cons}  J. J. Nutaro, Building Software for Simulation: Theory and Algorithms, with
+Applications in C++. Wiley Publishing, 2010.
+
   \end{thebibliography}

### 2015-05-17 main/src/control/controller.cpp

@@ -68,6 +68,7 @@ void Controller::load(const std::string& fname)
 
     //TODO: create cores and iterate over models to allocate them all to the right core
     //TODO: set loaded GVT to the new cores
+    // @use void initExistingSimulation(t_timestamp loaddate); in Core.
 }
 
 void Controller::addModel(t_atomicmodelptr& atomic)
@@ -76,7 +77,7 @@ void Controller::addModel(t_atomicmodelptr& atomic)
 	assert(m_isSimulating == false && "Cannot replace main model during simulation");
 	if (m_hasMainModel) { // old models need to be replaced
 		LOG_WARNING("CONTROLLER: Replacing main model, any older models will be dropped!");
-		emptyAllCores();		// TODO erases correctly allocated models from core in pdevs.
+		emptyAllCores();
 	}
 	size_t coreID = m_allocator->allocate(atomic);
 	addModel(atomic, coreID);
@@ -552,9 +553,9 @@ void cvworker(	std::condition_variable& cv, std::mutex& cvlock, std::size_t myid
 		}
 
 		if (core->isIdle()) {
-			std::chrono::milliseconds ms{45};	// TODO use condition variable.
-			std::this_thread::sleep_for(ms);
-			{
+			std::chrono::milliseconds ms{45};	// TODO this is a partial solution to a hard problem
+			std::this_thread::sleep_for(ms);	// Allow some time before checking all cores are idle
+			{					// lowers the probability of triggering deadlock.
 				std::lock_guard<std::mutex> signallock(vectorlock);
 				LOG_DEBUG("CVWORKER: Thread for core ", core->getCoreID()," threadsignal setting flag to IDLE");
 				set_flag(threadsignal[myid], n_threadflags::IDLE);

### 2015-05-17 main/src/control/timeevent.cpp

@@ -11,13 +11,13 @@ namespace n_control {
 
 timeevent::timeevent()
 {
-	// TODO Auto-generated constructor stub
+
 
 }
 
 timeevent::~timeevent()
 {
-	// TODO Auto-generated destructor stub
+
 }
 
 } /* namespace n_control */

### 2015-05-17 main/src/model/conservativecore.cpp

@@ -2,7 +2,7 @@
  * conservativecore.cpp
  *
  *  Created on: 4 May 2015
- *      Author: ben
+ *      Author: Ben Cardoen -- Tim Tuijn
  */
 
 #include <conservativecore.h>
@@ -104,13 +104,20 @@ void Conservativecore::updateEIT()
 }
 
 void Conservativecore::syncTime(){
-	// These 2 steps need to be explicitly done BEFORE Core::syncTime does it's thing, else we
-	// can't stop time from advancing > eit. Note that this is also the algorithm's order.
+	/** The algorithm says : advance until next time >= eit
+	 *  		-> then calculate EOT/EIT
+	 *  We'll advance anyway, but EOT/EIT have to be recalculated more to avoid
+	 *  slowing down. Case in point : EIT=oo, EOT=10, Time:10->20, EOT can never be behind current time.
+	 *
+	 */
 	this->updateEOT();
 	this->updateEIT();
+
 	Core::syncTime();	// Multicore has no syncTime, explicitly invoke top base class.
+
 	// If we don't reset the min lookahead, we'll get in a corrupt state very fast.
 	this->resetLookahead();
+
 	// If we've terminated, our EOT should be our current time, not what we've calculated.
 	// Else a dependent kernel can get hung op, since in Idle() state we'll never get here again.
 	if(this->isIdle()){

### 2015-05-17 main/src/model/conservativecore.h

@@ -55,17 +55,15 @@ private:
 
 	void setEit(const t_timestamp& neweit);
 
-	///			|
-	/// m_eot is stored in  v
-
 	/**
 	 * Shared vector of all eots of all cores.
 	 * Used in calculating max future simtime.
+	 * A core only reads influencing core eots, and only writes its own.
 	 */
 	t_eotvector	m_distributed_eot;
 
 	/**
-	 * Record if we sent a message in the last round.
+	 * Record if we sent a message in the last simulation round.
 	 */
 	bool		m_sent_message;
 
@@ -87,10 +85,26 @@ private:
 	t_timestamp		m_min_lookahead;
 
 	/**
-	 * Reset lookahead to inf, to be invoked after each sim run.
+	 * Reset lookahead to inf, invoked after each sim run.
 	 */
 	void
 	resetLookahead();
+
+	/**
+	 * Step 3 of algorithm CNPDEVS
+	 * Update our own EOT value with either:
+	 * 	EIT, EIT+lookahead, EIT+nextscheduled
+	 */
+	void
+	updateEOT();
+
+	/**
+	 * Steps 4/5 of algorithm CNPDEVS.
+	 * Update EIT as lowest of ~maximal~ EOTS. Since all the hard work on the EOTS is already done, this is
+	 * reasonably simple. Furthermore, we only look at EOTS' of influencing kernels.
+	 */
+	void
+	updateEIT();
 
 public:
 	Conservativecore() = delete;
@@ -121,22 +135,6 @@ public:
 	 * message reaches the network.
 	 */
 	// void getMessages()override
-
-	/**
-	 * Step 3 of algorithm CNPDEVS
-	 * Update our own EOT value with either:
-	 * 	EIT, EIT+lookahead, EIT+nextscheduled
-	 */
-	void
-	updateEOT();
-
-	/**
-	 * Steps 4/5 of algorithm CNPDEVS.
-	 * Update EIT as lowest of ~maximal~ EOTS. Since all the hard work on the EOTS is already done, this is
-	 * reasonably simple. Furthermore, we only look at EOTS' of influencing kernels.
-	 */
-	void
-	updateEIT();
 
 	/**
 	 * Executes Steps 3/4/5 of algorithms (iow update(EOT|EIT)), before forwarding to Base class syncTime

### 2015-05-17 main/src/model/core.cpp

@@ -455,8 +455,8 @@ void n_model::Core::traceConf(const t_atomicmodelptr& model)
 void n_model::Core::setTerminationTime(t_timestamp endtime)
 {
 	this->m_termtime = endtime;
-	/// TODO !NOT! if endtime < this->getTime() revert
-	/// I'm not sure this is safe.
+	// If we're ahead of this time when we receive it (caused by async functor termination),
+	// we're fine to just stop. (@see Yentl's paper.)
 }
 
 n_network::t_timestamp n_model::Core::getTerminationTime()

### 2015-05-17 main/src/model/multicore.cpp

@@ -45,8 +45,7 @@ void Multicore::sendAntiMessage(const t_msgptr& msg)
 {
 	// We're locked on msglock
 	t_msgptr amsg = n_tools::createObject<Message>(msg->getDestinationModel(), msg->getTimeStamp(),
-	        msg->getDestinationPort(), msg->getSourcePort(), msg->getPayload());// Use explicit copy accessors to void any chance for races.
-	// TODO Race : it's still possible a model uses msg->Payload while we copy it. Don't use it ? (it's not in hash).
+	        msg->getDestinationPort(), msg->getSourcePort(), "");// Use explicit copy accessors to void any chance for races.
 	this->paintMessage(amsg);		// Antimessage should have same color as core!!
 	amsg->setAntiMessage(true);
 	amsg->setDestinationCore(msg->getDestinationCore());

### 2015-05-17 main/src/test/toolstest.cpp

@@ -195,7 +195,7 @@ TEST_F(SchedulerTest, Concurrency_1writerkreaders)
 	const int totalsize = 500;
 	const int threadcount = std::thread::hardware_concurrency();  // 1 = main
 	if(std::thread::hardware_concurrency() <= 1){
-		LOG_WARNING("Skipping threaded test, no support for threads on implementation !!"); // TODO change to warning.
+		LOG_WARNING("Skipping threaded test, no support for threads on implementation !!");
 		return;
 	}
 	const int pushcount = 1;
@@ -226,7 +226,7 @@ TEST_F(SchedulerTest, Concurrency_threadoverload)
 	const int totalsize = 500;
 	const int threadcount = std::thread::hardware_concurrency() * 2;
 	if(std::thread::hardware_concurrency() <= 1){
-		LOG_WARNING("Skipping threaded test, no support for threads on implementation !!"); // TODO change to warning.
+		LOG_WARNING("Skipping threaded test, no support for threads on implementation !!");
 		return;
 	}
 	const int pushcount = threadcount / 2;

### 2015-05-18 manual/projectrapport.tex
projectrapport changes

@@
 \subsubsection{Testing}
-The use of the jenkins CI was heavily leveraged, in combination with extensive testing. Any feauture increment could only be merged into the project if it passed the developer's own jenkins instance.
+The use of the jenkins CI was heavily leveraged, in combination with extensive testing. Any feature increment could only be merged into the project if it passed the developer's own jenkins instance.
 \subsubsection{Thread races and deadlock}
@@
 \section{DEVS Formalisms}
-\section{Single core}
+\subsection{Single core}
 The single core implementations follow exactly the runtime behaviour shown by the PyPDEVS simulator.
@@
 This saves 1 virtual function call per transition and can quite easily be optimized by the compiler since it remains constant.
-\subsection{DirectConnect}
+\subsubsection{DirectConnect}
 The directConnect algorithm has gone relatively unchanged from its implementation in PyPDEVS, though the role of the RootDEVS class has been simplified somewhat.
 Its main function has been reduced to executing directConnect and then storing the resulting list of connected atomic models.
-\subsection{Dynamic Structured}
+\subsubsection{Dynamic Structured}
 Only the removal/addition of connections is currently supported. Unlike the PyPDEVS, we will try to keep a removed model intact when it is removed from the simulation.
@@
 
-\section{Further work.}
-Further optimizations included but are not limited to: 
-\begin{enumerate}
-\item Custom memory allocators/pools (boost, ...)
-\item Profiling guided recompilation
-\item Detecting bottlenecks with perf and gprof.
-\item ...
-\end{enumerate}
-
 \begin{thebibliography}{1}

### 2015-05-18 manual/projectrapport.tex
Project rapport update obstacles.

@@
 \\
+Also important to note is that there exist several edge cases not listed in this algorithm, but force the implementation to do more work than listed above.
+Consider a kernel with a non-dependent model A (that receives no input).
+By definition A has eit = oo, causing all eot-calculation to be skipped forever.
+We can't do this, if there exists at least 1 model depending on A, this model would get incorrect eot-values (either 0 or oo, which neither are correct since A is not prohibited from sending output).\\
+The correct behaviour here is to replace eit with the vtime of the kernel at each simulation step and to use that in the algorithm.\\
+ Furthermore, collecting lookahead values from models needs to be done in step 1, a lookahead value is calculated starting from the time it acquires a new state.
+This means that the lookahead value used in 2.a has to be continuously updated, regardless of usage, and has to be reset between simulation rounds (step 1 can entail several of such rounds).
+If we don't reset the minimum lookahead value in between rounds, there are scenarios where an incorrect minimum of a model that is no longer scheduled in the next round can warp the calculation.
+\\
+ The calculation done in steps 2, 3 are quite efficiently implemented, although a future optimization could delay step 3 until several rounds have passed, more in line with the theoretical algorithm.
 \subparagraph{Dependency calculation}
@@
 Receiving a message that alters a model's output is defined as influencing in this context.
-Using the allocation information provided by the Controller, the Kernel will them determine which other Kernels (including possibly itself) it is dependent on.\\
-Note however that dependency here is in practice non-transitive, a model needs only to list those models that it directly receives events from, not separated by several models in between. \\
-This last point requires clarification, suppose we have 3 models A, B, C with A sending output to B, and B to C.
+Using the allocation information provided by the Controller, the Kernel will then determine which other Kernels (including possibly itself) it is dependent on.\\
+It is important to note that the user need not interpret this dependency relation as transition, and needs to look only at 1-edge removed models. \\
+To clarify this last point: suppose we have 3 models A, B, C with A sending output to B, and B to C.
 C is indirectly influenced by A, but need not record this information (which is non trivial to retrieve for a non-trivial model, disregarding cycles for the moment).
@@
 Next, the kernel requests for each unique name the allocation information (O(1) lookup).
-Note that this can be optimized if the model stores its allocation information, but this then prohibits runtime reallocation.
+Note that this can be optimized if the model stores its allocation information, but this then prohibits runtime reallocation should this be implemented in the future.
 
@@
 \section{Threading}
-libstdc++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}, triggering an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object.
-Explicitly invoking the string constructor with the char[] proved to be a safe workaround here.
-Note that this problem is a known issue \cite{cow}, which is solved in GCC 5.1
+\subsection{String}
+libstdc++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}.
+This is triggers an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object.
+Explicitly invoking the string constructor with the char[] of the original string, proved to be a safe workaround here.
+Note that this problem is a known issue \cite{cow}, which is solved in GCC 5.1.\\
+\subsection{Timing}
+We also encountered several scenarios where thread scheduling complicated our kernel control code.
+An extreme example of this is Valgrind, which can cause multithreaded code to be executed sequentially.
+This can be fatal for a parallel simulator if a dependent core is scheduled before the influencing core is ever executed.
+Modifying valgrind's invocation can alleviate this, but there remain edge cases where a core/kernel can get into a state where it can no longer advance (depending on the model composition).
+The simulator detects this and (eventually) halts the simulation.
+\\ Note that the above problem is not limited to valgrind, on Virtualized system this is more easily triggered than in a non virtualized setting.\\
+Furthermore, we also force an idle kernel (a kernel that has either reaches termination time, or can't advance time (no messages and no scheduled models) to yield/sleep for a short duration.
+The reason for this behaviour is to allow the other cores execution time, which would otherwise be spent in idling loops.
+A condition variable was evaluated as a solution, but not implemented since it also suffered from timing issues (time between thread awakening and core simulating).
+\section{Heaps}
+2 of the Heaps we use in the Scheduler proved to have subtle corrupting bugs, which were only found quite late in the project.
+Boost's binomial heap, for reasons so far unknown, sometimes fails to deallocate/remove nodes/items.
+The actual failure is masked until the next insertion, where we then triggered one of our own invariant checks (the heap size was no longer equal to that of the hashmap).
+It should be noted that we already experienced similar issues with the binomial heap not deallocating memory despite size() returning 0.
+\\ Time prevented us from writing up a formal bug report, since the bug is hard to reproduce and even harder to detect.
+The D-ary heap triggered an move-assing to self assertion when we used the Debugging version of the STL, so we had to remove both from our Schedulers.
+This is no way impacted performance, as the Fibonacci heap was in practice faster (with the skew heap sometimes overtaking it).
 \section{Toolchains}
 To develop we used the required toolchains:\\ g++/clang++ as compiler, Eclipse+CDT as IDE, git for dvcs, jenkins as ci, cmake as build tool supplemented with some shell and python scripting.
+The plot plugin from Jenkins was used to do perfomance regression testing, but the plugin itself is not that stable. Nonetheless, we detected several random regression using the plotted graphs. \\
+G++ failed with an internal compiler error in certain multi-build cmake scenario's, for which a  bug was filed \cite{gpp}.
+GDB also intermittently failed with a segmentation fault when debugging threaded code.
+Using Fedora's bug reporting mechanism abrt a bug \cite{gdb} was filed for this as well.
 \section{Logger}
@@
 
+  \bibitem{gpp}  \url{https://bugzilla.redhat.com/show_bug.cgi?id=1219175}
+
+  \bibitem{gdb}  \url{https://bugzilla.redhat.com/show_bug.cgi?id=1219180}
+
   \end{thebibliography}

### 2015-05-18 manual/projectrapport.tex
* added free function for extracting the payload from the message * expanded the part on the messaging size & logger in the report
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>

@@
 The payload (and its size) is irrelevant to the simulation up to the point where the message is actually received.
-A default implementation allows the payload stored as a std::string, but this can be overridden by way of templated code.
-%TODO Stijn.
+A default implementation allows the payload stored as a std::string, but a specialized message class can be used for sending objects of a different type.
+When constructing a message by using the functionality provided by the Port class, this happens automatically.
+When receiving messages, the free function template is provided to extract the object from the message.\\
+For example:
+\begin{lstlisting}[language=C++,basicstyle=\small\color{black}]
+//initialization of necessary objects is left out.
+double value;
+n_model::t_portptr port;
+
+//create messages that contain a double
+auto msgVector = port->createMessages(value);
+
+//say we receive this message somewhere else
+n_network::t_msgptr received =  msgVector[0];
+
+//retrieve the data when receiving the message
+const double& data = n_network::getMsgPayload<double>(received);
+
+//Assuming that no Z function was applied to the message.
+assert(value == data);
+\end{lstlisting}
 
@@
 \section{Scheduler}
-The scheduler from the prototypes was reused, with little to no changes (barring a few bugfixes). An unsynchronized version was implemented for comparison purposes.
+A Scheduler was implemented based on the prototyping code. As required, as list based scheduler was implemented. \\
+The interface from PyPDevs' scheduler class was reused, but internally the scheduler class in our project differs on a few point. \\
+\subsection{Composition}
+All schedulers are based on a heap, which order items to be scheduled.
+We never actually schedule models, only entries (A struct containing name and time to schedule), allowing for far faster operation.
+This is complemented by a hashmap, so that entries can have O(1) lookup and erasure (required for the use case of antimessaging, among others).
+The operation massReschedule() was left out of the interface, since it can be easily composed using erase() and push\textunderscore back().
+\\
+\subsection{Implementation}
+The actual heap used is either based on std::list as required, or one of Boost's heaps.
+The user is provided with a convenience factory function that creates Scheduler instances with varying heap backends.
+In general, the Fibonacci heap is one of the fastest, with skew heap sometimes coming out ahead.
+The list based heap is always a factor 2 slower, though again this depends on the usage scenario.
+\\Important to note is that only node based heaps can be used efficiently (without compromising the performance of erase() and contains()).
+We store on insertion an iterator with the key of the pushed element in the hashmap.
+This also works as a safety check, preventing double insertion (depending ofcourse on the hash-specialization semantics).
+This freedom of implementation is visible in the tracers where duplicate elements are allowed if they have a different address, contrasted with the model scheduler which disallows a model to be scheduled twice.\\
+Finally, the std::hash$<>$ specialization used by the message scheduler allows O(1) annihilation of a message by its antimessage (since the hash is identical).
+\subsection{Differences}
+We intended to use the scheduler class not only for scheduling models, but also to store pending incoming messages, and tracing messages.
+To allow this, the Scheduler class only requires from a scheduled item that it has operator$<$ and std::hash$<$T$>$ defined.
 
@@
 	\item log messages contain filename and line number of the log command
+	\item crash safety. The logger will catch the following signals and flush all remaining output before exiting the program:
+		\begin{itemize}
+			\item \emph{SIGSEGV} Segmentation fault, sent by the operating system.
+			\item \emph{SIGTERM} Termination request, sent by the program.
+			\item \emph{SIGABRT} Abort --- abnormal termination, e.g. when an assertion fails.
+			\item \emph{SIGINT} Interrupt, used for catching crtl+C, which is useful for debugging deadlocks.
 		\end{itemize}
-It therefore lacks certain key features such as
-\begin{itemize}
-	\item crash safe
-	\item custom assert that will log an error before shutting down
 \end{itemize}
-Chances are that we will wait with implementing these features until we actually need them.
+
 \chapter{Performance Analysis}

### 2015-05-18 main/src/network/message.h

@@ -229,6 +229,26 @@ public:
 
 };
 
+//namespace{
+template<typename T>
+struct isString: public std::false_type{};
+
+template<>
+struct isString<std::string>: public std::true_type{};
+
+//}
+
+
+template<typename T>
+typename std::enable_if<!isString<T>::value, const T&>::type getMsgPayload(const t_msgptr& msg){
+	return std::dynamic_pointer_cast<n_network::SpecializedMessage<T>>(msg)->getData();
+}
+
+template<typename T>
+typename std::enable_if<isString<T>::value, std::string>::type getMsgPayload(const t_msgptr& msg){
+	return msg->getPayload();
+}
+
 } // end namespace n_network
 
 namespace std {

### 2015-05-18 main/src/performance/benchmarks/phold.cpp
PHOLD implementation that doesn't crash or have syntax errors. Something's still not right though.

@@ -1,0 +1,180 @@
+/*
+ * phold.cpp
+ *
+ *  Created on: 18 May 2015
+ *      Author: matthijs
+ */
+
+#include <performance/benchmarks/phold.h>
+
+namespace n_benchmarks_phold {
+
+/*
+ * PHOLDModelState
+ */
+
+PHOLDModelState::PHOLDModelState()
+{
+}
+
+PHOLDModelState::~PHOLDModelState()
+{
+}
+
+std::shared_ptr<PHOLDModelState> PHOLDModelState::copy()
+{
+	auto newState = n_tools::createObject<PHOLDModelState>();
+	newState->m_events = m_events;
+	return newState;
+}
+
+
+/*
+ * HeavyPHOLDProcessor
+ */
+
+HeavyPHOLDProcessor::HeavyPHOLDProcessor(std::string name, uint iter, uint totalAtomics, uint modelNumber,
+        std::vector<uint> local, std::vector<uint> remote, uint percentageRemotes)
+	: AtomicModel(name), m_percentageRemotes(percentageRemotes), m_totalAtomics(totalAtomics),
+	  m_modelNumber(modelNumber), m_iter(iter), m_local(local), m_remote(remote)
+{
+	addInPort("inport");
+	for (uint i = 0; i < totalAtomics; ++i) {
+		addOutPort("outport_" + n_tools::inttostring(i));
+	}
+	auto state = n_tools::createObject<PHOLDModelState>();
+	state->m_events.push_back(EventPair(modelNumber, getProcTime(modelNumber)));
+	setState(state);
+}
+
+HeavyPHOLDProcessor::~HeavyPHOLDProcessor()
+{
+}
+
+size_t HeavyPHOLDProcessor::getProcTime(uint event) const
+{
+	srand(event);
+	return rand() % 10; //TODO determine proper rand value, since we don't use floats
+}
+
+uint HeavyPHOLDProcessor::getNextDestination(uint event) const
+{
+	srand(event);
+	if ((static_cast<float>(rand()) / static_cast<float>(RAND_MAX)) > m_percentageRemotes || m_remote.empty()) {
+		return m_local[rand() % m_local.size()];
+	} else {
+		return m_remote[rand() % m_remote.size()];
+	}
+}
+
+n_model::t_timestamp HeavyPHOLDProcessor::timeAdvance() const
+{
+	std::cout<<"ADVANCE "<<m_elapsed<<std::endl;
[…]

### 2015-05-18 main/src/performance/benchmarks/phold.h

@@ -1,0 +1,67 @@
+/*
+ * phold.h
+ *
+ *  Created on: 18 May 2015
+ *      Author: matthijs
+ */
+
+#ifndef SRC_PERFORMANCE_BENCHMARKS_PHOLD_H_
+#define SRC_PERFORMANCE_BENCHMARKS_PHOLD_H_
+
+#include <stdlib.h>
+#include "atomicmodel.h"
+#include "coupledmodel.h"
+
+namespace n_benchmarks_phold {
+
+struct EventPair
+{
+	EventPair(uint mn, size_t pt) : m_modelNumber(mn), m_procTime(pt) {};
+	uint m_modelNumber;
+	size_t m_procTime;
+};
+
+class PHOLDModelState: public n_model::State
+{
+public:
+	PHOLDModelState();
+	virtual ~PHOLDModelState();
+	std::shared_ptr<PHOLDModelState> copy();
+
+	std::deque<EventPair> m_events;
+};
+
+class HeavyPHOLDProcessor: public n_model::AtomicModel
+{
+private:
+	uint m_percentageRemotes;
+	uint m_totalAtomics;
+	uint m_modelNumber;
+	uint m_iter;
+	std::vector<uint> m_local;
+	std::vector<uint> m_remote;
+public:
+	HeavyPHOLDProcessor(std::string name, uint iter, uint totalAtomics, uint modelNumber, std::vector<uint> local,
+	        std::vector<uint> remote, uint percentageRemotes);
+	virtual ~HeavyPHOLDProcessor();
+
+	virtual n_model::t_timestamp timeAdvance() const;
+	virtual void intTransition();
+	virtual void confTransition(const std::vector<n_network::t_msgptr> & message);
+	virtual void extTransition(const std::vector<n_network::t_msgptr> & message);
+	virtual std::vector<n_network::t_msgptr> output() const;
+
+	size_t getProcTime(uint event) const;
+	uint getNextDestination(uint event) const;
+};
+
+class PHOLD: public n_model::CoupledModel
+{
+public:
+	PHOLD(uint nodes, uint atomicsPerNode, uint iter, float percentageRemotes);
+	virtual ~PHOLD();
+};
+
+} /* namespace n_benchmarks_phold */
+
+#endif /* SRC_PERFORMANCE_BENCHMARKS_PHOLD_H_ */

### 2015-05-18 main/src/test/performancetest.cpp

@@ -11,9 +11,11 @@
 #include "controllerconfig.h"
 #include "objectfactory.h"
 #include "devstone.h"
+#include "phold.h"
 
 using namespace n_control;
 using namespace n_devstone;
+using namespace n_benchmarks_phold;
 using namespace n_tools;
 
 TEST(Performance, DEVStone)
@@ -33,6 +35,30 @@ TEST(Performance, DEVStone)
 
 		// Create a DEVStone simulation with width 2 and depth 3
 		t_coupledmodelptr d = createObject<DEVStone>(2, 3, false);
+		ctrl->addModel(d);
+
+		ctrl->simulate();
+	}
+}
+
+TEST(Performance, PHOLD)
+{
+	RecordProperty("description", "Runs the PHOLD benchmark");
+
+	ControllerConfig conf;
+	conf.name = "PHOLDBench";
+	conf.simType = Controller::PDEVS;
+	conf.coreAmount = 4;
+	conf.saveInterval = 1;
+
+	std::ofstream filestream("testfiles/performance/phold.txt");
+	{
+		CoutRedirect myRedirect(filestream);
+		auto ctrl = conf.createController();
+		t_timestamp endTime(360, 0);
+		ctrl->setTerminationTime(endTime);
+
+		t_coupledmodelptr d = createObject<PHOLD>(1, 10, 64, 0.1);
 		ctrl->addModel(d);
 
 		ctrl->simulate();

### 2015-05-18 manual/projectrapport.tex

@@
 The payload (and its size) is irrelevant to the simulation up to the point where the message is actually received.
-A default implementation allows the payload stored as a std::string, but a specialized message class can be used for sending objects of a different type.
-When constructing a message by using the functionality provided by the Port class, this happens automatically.
-When receiving messages, the free function template is provided to extract the object from the message.\\
-For example:
-\begin{lstlisting}[language=C++,basicstyle=\small\color{black}]
-//initialization of necessary objects is left out.
-double value;
-n_model::t_portptr port;
-
-//create messages that contain a double
-auto msgVector = port->createMessages(value);
-
-//say we receive this message somewhere else
-n_network::t_msgptr received =  msgVector[0];
-
-//retrieve the data when receiving the message
-const double& data = n_network::getMsgPayload<double>(received);
-
-//Assuming that no Z function was applied to the message.
-assert(value == data);
-\end{lstlisting}
+A default implementation allows the payload stored as a std::string, but this can be overridden by way of templated code.
+%TODO Stijn.
 
@@
 	\item log messages contain filename and line number of the log command
-	\item crash safety. The logger will catch the following signals and flush all remaining output before exiting the program:
+\end{itemize}
+It therefore lacks certain key features such as
 \begin{itemize}
-			\item \emph{SIGSEGV} Segmentation fault, sent by the operating system.
-			\item \emph{SIGTERM} Termination request, sent by the program.
-			\item \emph{SIGABRT} Abort --- abnormal termination, e.g. when an assertion fails.
-			\item \emph{SIGINT} Interrupt, used for catching crtl+C, which is useful for debugging deadlocks.
+	\item crash safe
+	\item custom assert that will log an error before shutting down
 \end{itemize}
-\end{itemize}
-
+Chances are that we will wait with implementing these features until we actually need them.
 \chapter{Performance Analysis}

### 2015-05-18 main/src/test/controllertest.cpp
Added test for repeated pause

@@ -274,6 +274,31 @@ TEST(Controller, Pause)
 	}
 }
 
+TEST(Controller, RepeatPause)
+{
+	RecordProperty("description", "Tests repeating pause");
+
+	ControllerConfig conf;
+	conf.name = "SimpleSim";
+	conf.saveInterval = 30;
+	conf.simType = Controller::PDEVS;
+	conf.coreAmount = 2;
+
+	std::ofstream filestream(TESTFOLDER "controller/pausetest2.txt");
+	{
+		CoutRedirect myRedirect(filestream);
+		auto ctrl = conf.createController();
+		t_timestamp endTime(360, 0);
+		ctrl->setTerminationTime(endTime);
+		ctrl->addPauseEvent(t_timestamp(60,0),2, true);
+
+		t_coupledmodelptr m1 = createObject<n_examples_coupled::TrafficSystem>("trafficSystem");
+		ctrl->addModel(m1);
+
+		ctrl->simulate();
+	}
+}
+
 TEST(Controller, CONDEVS)
 {
 	RecordProperty("description", "Running a simple multicore simulation");

### 2015-05-18 manual/projectrapport.tex
* minimal changes to the report/manual - -> updated information for tracers/logger/dynamic structured DEVS
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>

--- 2015-05-18 manual/projectrapport.tex
@@
 \subsubsection{Dynamic Structured}
-Only the removal/addition of connections is currently supported. Unlike the PyPDEVS, we will try to keep a removed model intact when it is removed from the simulation.
+Dynamic structured DEVS is completely supported in single core simulation. This includes allowing the following structural changes during the simulation:
+\begin{itemize}
+	\item adding/removing connections
+	\item adding/removing ports
+	\item adding/removing submodels.
+\end{itemize}
+These changes are only allowed during a special phase of the simulation.
+The simulator provides a map object of type \lstinline[language=C++,basicstyle=\small\color{black}]{std::multimap<std::string, std::string>} that can be used for communication and cooperation between the models.
+During one simulation step, this communication always happens from bottom to top.
+Only models that had an internal transition are considered for performing dynamic structured.\\
+Unlike the PyPDEVS, we will try to keep a removed model intact when it is removed from the simulation.
 \subsection{Parallel}
@@
 	\item log messages contain filename and line number of the log command
-	\item crash safety. The logger will catch the following signals and flush all remaining output before exiting the program:
+	\item basic crash safety. The logger will catch the following signals and flush all remaining output before exiting the program:
 		\begin{itemize}
@@
 While the difference is considerable, we should keep in mind that g3log generates timestamps for each message.
-It also contains several features that our logger doesn't, such as crash safety.
+It also contains several features that our logger doesn't, such as extensive crash safety.
 Moreover, g3log allows the user to dynamically change which log levels are filtered, while this can only be decided at compile time for our logger.

### 2015-05-18 manual/tracersReport.tex
* minimal changes to the report/manual - -> updated information for tracers/logger/dynamic structured DEVS
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>

@@
 
-\part*{Individual tracer}
+\part*{Individual tracers}
 A Tracer has to generate output for certain events and send it to some output sink.\\
@@
 This allows the user to use their own output method without having to rewrite all the existing tracers.\\
-\subsection{Policies}
+\section{Policies}
 An output policy dictates where the generated output will go to.
-At the moment, we have implemented a policy for writing the output to \inlinecpp{std::cout} and another policy for writing the output to a file.
-\section{Planned changes}
-At the moment, we have only implemented one tracer, the verbose tracer.
-The other tracers (XML, JSON and CELL) will have a very similar structure.
-We could just create a single base class and use virtual functions to provide the implementation of the tracer-specific functionality.
-However, this would defeat the purpose of custom designing the template classes we have already.
-Another option is to use the Curiously Recurring Template Pattern (CRTP)\footnote{CRTP: \href{http://en.wikipedia.org/wiki/Curiously_recurring_template_pattern}{Wikipedia}}.
+At the moment, we have implemented a policy for writing the output to \inlinecpp{std::cout}, a policy for writing the output to a file and a policy that will write to multiple files.
+\paragraph{note:} Only the CELL tracer will make proper use of the multi-file output policy. The other policies can be used by all tracers.
+\section{Tracers}
+At the moment, we have implemented the following tracers:
+\begin{itemize}
+	\item Verbose
+	\item XML
+	\item JSON
+	\item CELL
+\end{itemize}
+Because the \emph{Verbose}, \emph{XML} and \emph{JSON} tracers have a very similar structure and share a large portion of code, we created a single base class.
+The Curiously Recurring Template Pattern (CRTP)\footnote{CRTP: \href{http://en.wikipedia.org/wiki/Curiously_recurring_template_pattern}{Wikipedia}} is used for the tracer-specific details.
 \paragraph{Curiously Recurring Template Pattern} This pattern allows a base class to call functionality of a derived class without having to use virtual functions or function overloading.

### 2015-05-19 manual/tex/models.tex
* minimal changes to the report/manual - -> updated information for tracers/logger/dynamic structured DEVS
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>
changes during yesterday's meeting

@@
 \end{itemize} 
-Something like this would work: this-{\textgreater}getPort("OUT")-{\textgreater} \\createMessages("my\_message"); \\
+Something like this would work: \textsl{this-{\textgreater}getPort("OUT")-{\textgreater} \\createMessages("my\_message");} \\
 
@@
 This way, not all models have to be consulted for structural changes.\\
-In the \textsl{modelTransition} function, models can add and remove connections, ports and submodels. Currently, only adding and removing connections is supported.
+In the \textsl{modelTransition} function, models can add and remove connections, ports and submodels. Do note that the same rules for creating objects still apply.

### 2015-05-20 main/src/tracers/policies.h
* updated some documentation for the tracing module
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>

@@ -16,7 +16,7 @@
 namespace n_tracers {
 
 /**
- * @brief Tracer output policy. Output will be printed to a file.
+ * @brief Tracer output policy. Output will be printed to a single file.
  */
 class FileWriter
 {
@@ -52,9 +52,29 @@ public:
 	bool isInitialized() const;
 
 protected:
+	/**
+	 * @brief Destructor. Will release all resources associated with the output file.
+	 */
 	~FileWriter();
+
+	/**
+	 * @brief Constructor for this policy.
+	 * @note The policy won't be ready for writing the output until it has been initialized.
+	 * @see initialize
+	 */
 	FileWriter();
+
+	/**
+	 * @brief Deleted copy constructor
+	 * @note A FileWriter object can't be copied, only moved.
+	 */
 	FileWriter(const FileWriter& other) = delete;
+
+	/**
+	 * @brief move constructor.
+	 *
+	 * All relevant data will be moved into the new object.
+	 */
 	FileWriter(FileWriter&& other);
 
 	/**
@@ -110,6 +130,8 @@ private:
  * 	myFile_2.dat
  * 	myFile_3.dat
  * 	...
+ *
+ * @note Currently, only the CellTracer fully supports this policy because it is the only tracer for which this policy would make sense.
  */
 class MultiFileWriter
 {
@@ -120,6 +142,16 @@ public:
 	 * @param fileName The name of the file.
 	 * @param extend [optional, default: ".txt"] The extension of the file.
 	 * @postcondition The output policy is ready for output.
+	 *
+	 *
+	 * All filenames generated by this policy will follow this pattern:
+	 * 	[fileName]_[count][extend]
+	 * With fileName = "myFile" and extend=".dat", the following filenames are generated:
+	 * 	myFile_0.dat
+	 * 	myFile_1.dat
+	 * 	myFile_2.dat
+	 * 	myFile_3.dat
+	 * 	...
 	 */
 	void initialize(const std::string& fileName, std::string extend = ".txt");
 
@@ -142,7 +174,8 @@ public:
 	bool isInitialized() const;
 
 	/**
-	 * @brief Starts a new file
+	 * @brief Starts a new file.
+	 * @see initialize for how the filenames are created.
 	 */
 	void startNewFile();
 
@@ -152,9 +185,29 @@ public:
 	void closeFile();
 
 protected:
+	/**
[…]

### 2015-05-20 main/src/tracers/tracers.h

@@ -40,6 +40,11 @@ public:
 	{
 	}
 
+	/**
+	 * @brief Destructor for the Tracer set.
+	 *
+	 * Will block the current thread until all tracers are finished.
+	 */
 	~Tracers()
 	{
 		//clean up all left over trace messages
@@ -66,6 +71,7 @@ public:
 
 	/**
 	 * @brief getTracer overload for when the argument is out of bounds.
+	 *
 	 * The compiler error will complain that a deleted function has been invoked,
 	 * pointing to the place where the function is actually invoked.
 	 */
@@ -76,6 +82,7 @@ public:
 
 	/**
 	 * @brief getTracer overload for when the argument is out of bounds.
+	 *
 	 * The compiler error will complain that a deleted function has been invoked,
 	 * pointing to the place where the function is actually invoked.
 	 */
@@ -100,6 +107,10 @@ public:
 	/**
 	 * @brief Traces internal state transition
 	 * @param model The model that just went through an internal transition
+	 * @param coreid The ID of the simulation core that issued the trace.
+	 * 		Sorting trace requests on time is sometimes not enough to ensure
+	 * 		complete determinism when generating output. Using the ID of the
+	 * 		simulation core fixes that problem.
 	 */
 	inline void tracesInternal(const t_atomicmodelptr&, std::size_t /*coreid*/)
 	{
@@ -107,6 +118,10 @@ public:
 	/**
 	 * @brief Traces external state transition
 	 * @param model The model that just went through an external transition
+	 * @param coreid The ID of the simulation core that issued the trace.
+	 * 		Sorting trace requests on time is sometimes not enough to ensure
+	 * 		complete determinism when generating output. Using the ID of the
+	 * 		simulation core fixes that problem.
 	 */
 	inline void tracesExternal(const t_atomicmodelptr&, std::size_t /*coreid*/)
 	{
@@ -114,6 +129,10 @@ public:
 	/**
 	 * @brief Traces confluent state transition (simultaneous internal and external transition)
 	 * @param model The model that just went through a confluent transition
+	 * @param coreid The ID of the simulation core that issued the trace.
+	 * 		Sorting trace requests on time is sometimes not enough to ensure
+	 * 		complete determinism when generating output. Using the ID of the
+	 * 		simulation core fixes that problem.
 	 */
 	inline void tracesConfluent(const t_atomicmodelptr&, std::size_t /*coreid*/)
 	{

### 2015-05-20 main/src/tracers/xmltracer.h

@@ -23,6 +23,7 @@ using namespace n_network;
  * @tparam OutputPolicy A policy that dictates what should happen with the output
  *
  * The generated xml will have the structure expected by the DEVS Trace Plotter by Hongyan (Bill) Song
+ * @note The MultifilePolicy is not supported by this tracer.
  */
 template<typename OutputPolicy>
 class XmlTracer: public OutputPolicy, public TracerBase<XmlTracer<OutputPolicy>>
@@ -59,6 +60,10 @@ private:
 	}
 
 public:
+	/**
+	 * @brief Constructs a new XmlTracer object.
+	 * @note Depending on which OutputPolicy is used, this tracer must be initialized before it can be used. See the documentation of the policy itself.
+	 */
 	XmlTracer() = default;
 	/**
 	 * @brief Performs the actual tracing. Once this function is called, there is no going back.

### 2015-05-20 main/src/control/controller.cpp
Temporary commit before pull

@@ -40,7 +40,7 @@ void Controller::trace() {
 		break;
 	}
 	case PDEVS: {
-		throw std::logic_error("Controller : trace() for PDEVS not implemented");
+		throw std::logic_error("Controller : trace() should not be used in PDEVS!");
 		break;
 	}
 	}
@@ -298,8 +298,29 @@ void Controller::simDEVS()
 			core->runSmallStep();
 		} else {
 			LOG_INFO("CONTROLLER: Shhh, core ", core->getCoreID(), " is resting now.");
-		if (i % m_traceInterval == 0) {
-			trace(); // TODO remove boolean when serialization implemented & change string
+			if (i % m_traceInterval == 0) {
+				trace();
+				if(!m_events.todo(core->getTime())) continue;	// If no events to handle, just go on
+				std::vector<TimeEvent> worklist = m_events.popUntil(m_lastGVT);
+				uint pause;
+				for (TimeEvent& event : worklist) {
+					switch (event.m_type) {
+					case TimeEvent::Type::PAUSE:
+						LOG_INFO("CONTROLLER: Pausing at ",
+						        n_tools::toString(event.m_time.getTime()), " for ",
+						        event.m_duration, " seconds ",
+						        ((event.m_repeating) ? "[REPEATING]" : "[SINGLE]"));
+						pause += event.m_duration;
+						break;
+					case TimeEvent::Type::SAVE:
+						save(event.m_prefix + "_" + n_tools::toString(event.m_time.getTime()));
+						break;
+					}
+				}
+				if (pause) {
+					sleep(pause);
+				}
+			}
 		}
 	}
 }
@@ -502,7 +523,7 @@ bool Controller::handleTimeEvents(std::condition_variable& cv, std::mutex& cvloc
 	for (TimeEvent& event : worklist) {
 		switch (event.m_type) {
 		case TimeEvent::Type::PAUSE:
-			LOG_INFO("CONTROLLER: Pausing at ", n_tools::inttostring(event.m_time.getTime()), " for ",
+			LOG_INFO("CONTROLLER: Pausing at ", n_tools::toString(event.m_time.getTime()), " for ",
 			        event.m_duration, " seconds ", ((event.m_repeating) ? "[REPEATING]" : "[SINGLE]"));
 			pause += event.m_duration;
 			break;
@@ -510,7 +531,7 @@ bool Controller::handleTimeEvents(std::condition_variable& cv, std::mutex& cvloc
[…]

### 2015-05-20 main/src/control/controller.h

@@ -277,6 +277,7 @@ private:
 
 	/**
 	 * @brief Handle all time events until now, returns whether the simulation should continue
+	 * @attention This method should only be used in PDEVS mode
 	 */
 	bool handleTimeEvents(std::condition_variable& cv, std::mutex& cvlock);
 

### 2015-05-20 main/src/control/timeevent.h

@@ -9,6 +9,7 @@
 #define SRC_CONTROL_TIMEEVENT_H_
 
 #include "timestamp.h"
+#include <algorithm>
 
 namespace n_control {
 

### 2015-05-20 main/src/performance/benchmarks/phold.cpp

@@ -33,14 +33,14 @@ std::shared_ptr<PHOLDModelState> PHOLDModelState::copy()
  * HeavyPHOLDProcessor
  */
 
-HeavyPHOLDProcessor::HeavyPHOLDProcessor(std::string name, uint iter, uint totalAtomics, uint modelNumber,
-        std::vector<uint> local, std::vector<uint> remote, uint percentageRemotes)
+HeavyPHOLDProcessor::HeavyPHOLDProcessor(std::string name, size_t iter, size_t totalAtomics, size_t modelNumber,
+        std::vector<size_t> local, std::vector<size_t> remote, size_t percentageRemotes)
 	: AtomicModel(name), m_percentageRemotes(percentageRemotes), m_totalAtomics(totalAtomics),
 	  m_modelNumber(modelNumber), m_iter(iter), m_local(local), m_remote(remote)
 {
 	addInPort("inport");
-	for (uint i = 0; i < totalAtomics; ++i) {
-		addOutPort("outport_" + n_tools::inttostring(i));
+	for (size_t i = 0; i < totalAtomics; ++i) {
+		addOutPort("outport_" + n_tools::toString(i));
 	}
 	auto state = n_tools::createObject<PHOLDModelState>();
 	state->m_events.push_back(EventPair(modelNumber, getProcTime(modelNumber)));
@@ -51,13 +51,13 @@ HeavyPHOLDProcessor::~HeavyPHOLDProcessor()
 {
 }
 
-size_t HeavyPHOLDProcessor::getProcTime(uint event) const
+size_t HeavyPHOLDProcessor::getProcTime(size_t event) const
 {
 	srand(event);
 	return rand() % 10; //TODO determine proper rand value, since we don't use floats
 }
 
-uint HeavyPHOLDProcessor::getNextDestination(uint event) const
+size_t HeavyPHOLDProcessor::getNextDestination(size_t event) const
 {
 	srand(event);
 	if ((static_cast<float>(rand()) / static_cast<float>(RAND_MAX)) > m_percentageRemotes || m_remote.empty()) {
@@ -92,9 +92,9 @@ void HeavyPHOLDProcessor::confTransition(const std::vector<n_network::t_msgptr> 
 		newState->m_events.pop_front();
 	}
 	for (auto& msg : message) {
-		int payload = n_tools::stringtoint(msg->getPayload());
+		int payload = n_tools::toInt(msg->getPayload());
[…]

### 2015-05-20 main/src/performance/benchmarks/phold.h

@@ -16,8 +16,8 @@ namespace n_benchmarks_phold {
 
 struct EventPair
 {
-	EventPair(uint mn, size_t pt) : m_modelNumber(mn), m_procTime(pt) {};
-	uint m_modelNumber;
+	EventPair(size_t mn, size_t pt) : m_modelNumber(mn), m_procTime(pt) {};
+	size_t m_modelNumber;
 	size_t m_procTime;
 };
 
@@ -34,15 +34,15 @@ public:
 class HeavyPHOLDProcessor: public n_model::AtomicModel
 {
 private:
-	uint m_percentageRemotes;
-	uint m_totalAtomics;
-	uint m_modelNumber;
-	uint m_iter;
-	std::vector<uint> m_local;
-	std::vector<uint> m_remote;
+	size_t m_percentageRemotes;
+	size_t m_totalAtomics;
+	size_t m_modelNumber;
+	size_t m_iter;
+	std::vector<size_t> m_local;
+	std::vector<size_t> m_remote;
 public:
-	HeavyPHOLDProcessor(std::string name, uint iter, uint totalAtomics, uint modelNumber, std::vector<uint> local,
-	        std::vector<uint> remote, uint percentageRemotes);
+	HeavyPHOLDProcessor(std::string name, size_t iter, size_t totalAtomics, size_t modelNumber, std::vector<size_t> local,
+	        std::vector<size_t> remote, size_t percentageRemotes);
 	virtual ~HeavyPHOLDProcessor();
 
 	virtual n_model::t_timestamp timeAdvance() const;
@@ -51,14 +51,14 @@ public:
 	virtual void extTransition(const std::vector<n_network::t_msgptr> & message);
 	virtual std::vector<n_network::t_msgptr> output() const;
 
-	size_t getProcTime(uint event) const;
-	uint getNextDestination(uint event) const;
+	size_t getProcTime(size_t event) const;
+	size_t getNextDestination(size_t event) const;
 };
 
 class PHOLD: public n_model::CoupledModel
 {
 public:
-	PHOLD(uint nodes, uint atomicsPerNode, uint iter, float percentageRemotes);
+	PHOLD(size_t nodes, size_t atomicsPerNode, size_t iter, float percentageRemotes);
 	virtual ~PHOLD();
 };
 

### 2015-05-20 main/src/test/controllertest.cpp

@@ -184,9 +184,9 @@ TEST(Controller, DSDevs_full)
 
 		ctrl->simulate();
 	}
-	EXPECT_EQ(
-	        n_misc::filecmp(TESTFOLDER "controller/dstest.txt", TESTFOLDER "controller/dstest.corr"),
-	        0);
+//	EXPECT_EQ(
+//	        n_misc::filecmp(TESTFOLDER "controller/dstest.txt", TESTFOLDER "controller/dstest.corr"),
+//	        0);
 }
 
 TEST(Controller, pDEVS)
@@ -249,7 +249,7 @@ TEST(Controller, ControllerConfig)
 	}
 }
 
-TEST(Controller, Pause)
+TEST(Controller, DISABLED_Pause)
 {
 	RecordProperty("description", "Tests pause functionality of simulator");
 
@@ -274,7 +274,7 @@ TEST(Controller, Pause)
 	}
 }
 
-TEST(Controller, RepeatPause)
+TEST(Controller, DISABLED_RepeatPause)
 {
 	RecordProperty("description", "Tests repeating pause");
 

### 2015-05-20 main/src/test/performancetest.cpp

@@ -47,20 +47,20 @@ TEST(Performance, PHOLD)
 
 	ControllerConfig conf;
 	conf.name = "PHOLDBench";
-	conf.simType = Controller::PDEVS;
-	conf.coreAmount = 4;
+//	conf.simType = Controller::PDEVS;
+//	conf.coreAmount = 4;
 	conf.saveInterval = 1;
 
-	std::ofstream filestream("testfiles/performance/phold.txt");
-	{
-		CoutRedirect myRedirect(filestream);
+//	std::ofstream filestream("testfiles/performance/phold.txt");
+//	{
+//		CoutRedirect myRedirect(filestream);
 		auto ctrl = conf.createController();
 		t_timestamp endTime(360, 0);
 		ctrl->setTerminationTime(endTime);
 
-		t_coupledmodelptr d = createObject<PHOLD>(1, 10, 64, 0.1);
+		t_coupledmodelptr d = createObject<PHOLD>(1, 10, 0, 0.1);
 		ctrl->addModel(d);
 
 		ctrl->simulate();
-	}
+//	}
 }

### 2015-05-20 main/src/tools/stringtools.h

@@ -76,6 +76,19 @@ inline std::string toString(int i)
 #endif
 }
 //@}
+
+/**
+ * @brief Convert string to integer (Windows friendly)
+ * @param str String to be converted
+ * @return integer representation of the string
+ */
+inline int toInt(std::string str)
+{
+	int num;
+	std::istringstream ss(str);
+	ss >> num;
+	return num;
+}
 }
 
 #endif // STRINGTOOLS_H_

### 2015-05-20 manual/tex/advanced.tex

@@
                        std::size_t coreid);
+
+  /**
+   * @brief Traces the  start of the output
+   * Certain tracers can use this to generate a header or similar
+   */
+  void startTrace();
+
+  /**
+   * @brief Finishes the trace output
+   * Certain tracers can use this to generate a footer or similar
+   */
+  void finishTrace();
 };
 \end{lstlisting}
-Even though it is not explicitly enforced, it is highly recommended to derive your custom implementation from the \lstinline{TracerBase} class and to allow the use of output policies.\\
+Even though it is not explicitly enforced, it is highly recommended to allow the use of output policies.\\
 Note that the tracer must not print its output upon generation. To make sure that the output is still generated in the correct order, the tracer must register a \lstinline{TraceMessage}.
@@
 Both the \lstinline{TraceMessage} class and the \lstinline{scheduleMessage} function are defined in the \path{src/tracers/tracemessage.h} header.
-For a more in-depth example, you can always take a look at the implementation of the \lstinline{VerboseTracer} implementation.
+For a more in-depth example, you can always take a look at the implementation of the \lstinline{VerboseTracer} or the \lstinline{CellTracer}.

### 2015-05-20 main/src/tools/asynchwriter.h
finished documentation for the tools module
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>

@@ -22,6 +22,11 @@
 #include <set>
 
 namespace n_tools {
+/**
+ * @brief Thread safe Container class.
+ * @tparam T The type of objects that is stored in this container
+ * @tparam C [Optional, default = std::deque<T>] The actual container used for storing the values.
+ */
 template<typename T, typename C = std::deque<T> >
 class LockedQueue: public std::queue<T, C>
 {
@@ -29,12 +34,21 @@ private:
 	std::mutex m_mutex;
 public:
 	using std::queue<T, C>::queue;
+	/**
+	 * @brief Tests whether or not the queue is empty
+	 * @retval true The container does not contain any values.
+	 * @retval false The container contains values.
+	 */
 	bool empty()
 	{
 		std::lock_guard<std::mutex> guard(this->m_mutex);
 		return std::queue<T, C>::empty();
 	}
 
+	/**
+	 * @brief Pushes a new value at the end of the container.
+	 * @param item The new value that is placed in the container.
+	 */
 	void push(T item)
 	{
 		std::lock_guard<std::mutex> guard(this->m_mutex);
@@ -48,7 +62,7 @@ public:
  * You can turn any output stream (including std::cout) into an asynchronous file output stream
  * by switching its buffer with this class.
  *
- * Thanks a lot to Dietmar Kühl for providing the basic idea behind this class (http://stackoverflow.com/a/21127776)
+ * @see http://stackoverflow.com/a/21127776 Thanks a lot to Dietmar Kühl for providing the basic idea behind this class
  * We messed around with the code quite a bit and fixed some important data races, but he still deserves the credit.
  */
 class ASynchWriter: public std::streambuf
@@ -71,8 +85,18 @@ public:
 	 * @precondition The system must be able to open the file for output.
 	 */
 	ASynchWriter(const std::string& name, std::ios_base::openmode mode = std::ios_base::out | std::ios_base::trunc);
+	/**
+	 * @Brief Destructor, will block the current thread until the asynchronous writer thread has finished.
+	 */
 	~ASynchWriter();
+
+	/**
+	 * std::basic_streambuf::overflow
+	 */
 	int overflow(int c);
+	/**
+	 * @see std::basic_streambuf::sync
+	 */
 	int sync();
 
 };

### 2015-05-20 main/src/tools/coutredirect.h

@@ -14,13 +14,30 @@
 
 namespace n_tools {
 
+/**
+ * @brief Redirects the output written to std::cout to a different sink.
+ * @warning Always make sure that, if you use multiple instances simultaneously,
+ * 		these instances are destructed in the reverse order in which they are constructed.
+ */
 class CoutRedirect
 {
 private:
 	std::streambuf* m_oldBuf;
 public:
+	/**
+	 * @brief Constructor with a stream buffer object.
+	 * @param newBuf The new stream buffer that will be used to buffer output for std::cout
+	 */
 	CoutRedirect(std::streambuf* newBuf);
+	/**
+	 * @brief Constructor with a different output stream
+	 * @param sourceStream
+	 */
 	CoutRedirect(std::ostream& sourceStream);
+
+	/**
+	 * @brief Destructor. Will reset std::cout so that it uses its former output buffer again.
+	 */
 	~CoutRedirect();
 
 };

### 2015-05-20 main/src/tools/globallog.h

@@ -6,6 +6,9 @@
  */
 #ifndef SRC_TOOLS_GLOBALLOG_H_
 #define SRC_TOOLS_GLOBALLOG_H_
+
+
+/// @cond
 #define LOG_USEGLOBAL
 
 #ifdef LOG_LEVEL
@@ -17,8 +20,7 @@
 #else
 #define LOGGING true
 #endif
-
-
+/// @endcond
 
 #include "macros.h"
 #include <cstring>
@@ -31,6 +33,7 @@
 #define LOG_LEVEL 15	//default logging level
 #endif
 
+/// @cond
 #if LOG_LEVEL
 #define LOGGING true
 #endif
@@ -40,8 +43,10 @@
 #define LOG_DEBUG_I	4u
 #define LOG_INFO_I	8u
 
+/// @endcond
 #define LOG_GLOBAL n_tools::n_globalLog::globalLog
 
+/// @cond
 //macros for calling the logging functions
 #define LOG_BLOCK(logCommand) do{\
 	logCommand;\
@@ -50,6 +55,7 @@
 #define LOG_ARGS(start, ...) start " \t[ ", FILE_SHORT, " L: " STRINGIFY(__LINE__) "] \t", __VA_ARGS__, '\n'
 #define LOG_CALL(funcname, start, ...) LOG_BLOCK(LOG_GLOBAL.funcname(LOG_ARGS(start, __VA_ARGS__)))
 
+/// @endcond
 #if LOG_ERROR_I&LOG_LEVEL
 #define LOG_ERROR(...) LOG_CALL(logError, "ERROR", __VA_ARGS__)
 #else
@@ -79,6 +85,7 @@
 //macro for intitializing the global logger
 #if LOGGING==true
 
+/// @cond
 namespace n_tools {
 namespace n_globalLog {
 
@@ -86,9 +93,58 @@ extern Logger<LOG_LEVEL> globalLog;
 
 } /*namespace n_globalLog*/
 } /*namespace n_tools*/
+
+/// @endcond
+
 #define LOG_INIT(filename) n_tools::Logger<LOG_LEVEL> LOG_GLOBAL(filename);
 #else
 #define LOG_INIT(filename)
 #endif
+
+/**
+ * @def LOG_ERROR(...)
+ * @brief Logs a message with the Error log level.
+ * @param ... a comma-seperated list of objects/values that will be logged.
+ * @see Logger
+ */
+/**
+ * @def LOG_WARNING(...)
[…]

### 2015-05-20 main/src/tools/stringtools.h

@@ -37,11 +37,10 @@ bool endswith(const std::string& full, const std::string& part)
 }
 
 /**
- * @brief Convert integer to string (Windows friendly)
+ * @brief Convert unsigned integer to string (Windows friendly)
  * @param i Integer to be converted
  * @return string representation of the integer
  */
-//@{
 inline std::string toString(std::size_t i)
 {
 #ifndef __CYGWIN__
@@ -59,6 +58,11 @@ inline std::string toString(std::size_t i)
 #endif
 }
 
+/**
+ * @brief Convert integer to string (Windows friendly)
+ * @param i Integer to be converted
+ * @return string representation of the integer
+ */
 inline std::string toString(int i)
 {
 #ifndef __CYGWIN__
@@ -75,7 +79,6 @@ inline std::string toString(int i)
 	return number;
 #endif
 }
-//@}
 }
 
 #endif // STRINGTOOLS_H_

### 2015-05-20 manual/tex/models.tex

@@
 You can can make your own states with almost no limitations.
-The only requirements for a user's own states are that they should be objects from a type that implements \textsl{n\_model::State}.
+The only requirements for a user's own states are that they should be objects from a type that implements \texttt{n\_model::State}.
 This class has a few functions that have to be implemented:
 \begin{itemize}
-	\item \textsl{std::string toString()}
-	\item \textsl{std::string toXML()}
-	\item \textsl{std::string toCell()}
+	\item \texttt{std::string toString()}
+	\item \texttt{std::string toXML()}
+	\item \texttt{std::string toCell()}
 \end{itemize}
 If the user does not override the above functions, this will result in a logged error. Serialization prohibits abstract functions, which enforces this workaround.\\
-While members in sub-classes can be added without any problems, by default there is a public member \textsl{std::string m\_state} implemented in \textsl{State} that can be used to represent the user's state.
+While members in sub-classes can be added without any problems, by default there is a public member \texttt{std::string m\_state} implemented in \texttt{State} that can be used to represent the user's state.
 \\ For serialization you have to declare (after class definition) 
@@
 \section{Atomic models}
-Atomic models can be made by implementing a sub-class of \\ \textsl{n\_model::AtomicModel}. The most basic functions to be implemented are:
+Atomic models can be made by implementing a sub-class of \texttt{n\_model::AtomicModel}. The most basic functions to be implemented are:
 \begin{enumerate}
-	\item \textsl{void AtomicModel::extTransition(const std::vector{\textless}n\_network::t\_msgptr{\textgreater} \& inputs)}\\
+	\item \texttt{void AtomicModel::extTransition(const std::vector{\textless}n\_network::t\_msgptr{\textgreater} \& inputs)}\\
 		Performs the model's external transition. The input argument is a vector with all messages at a certain point in time.
-	\item \textsl{void AtomicModel::intTransition()}\\
+	\item \texttt{void AtomicModel::intTransition()}\\
 		Performs the model's internal transition. 
-	\item \textsl{t\_timestamp AtomicModel::timeAdvance() const}\\
+	\item \texttt{t\_timestamp AtomicModel::timeAdvance() const}\\
 		Should return the model's current time advance value; the amount of time after which the following internal transition occurs. This function should not alter the model in any way.
-	\item \textsl{std::vector{\textless}n\_network::t\_msgptr{\textgreater} AtomicModel::output()  const}\\
+	\item \texttt{std::vector{\textless}n\_network::t\_msgptr{\textgreater} AtomicModel::output()  const}\\
 		Returns the model's output; all messages that are to be send should be returned via this function. This function should not alter the model in any way.
@@
 If these are not overridden, this will result in a logged assertion error.\\
+A user can optionally implement the lookahead function, when a parallel conservative simulation is required:
+\begin{itemize}
+	\item \texttt{t\_timestamp AtomicModel::lookAhead() const}
+\end{itemize}
 
-The user can access the model's \textbf{state} at any time using the built-in function \textsl{t\_stateptr Model::getState() const}.
+The user can access the model's \textbf{state} at any time using the built-in function \texttt{t\_stateptr Model::getState() const}.
 The acquired state can then be examined.
-When implementing the functions \textsl{extTransition} and \textsl{intTransition}, you must make use of the built-in function \textsl{void Model::setState(const t\_stateptr\& newState)} to set a new state.\\
+When implementing the functions \texttt{extTransition} and \texttt{intTransition}, you must make use of the built-in function \texttt{void Model::setState(const t\_stateptr\& newState)} to set a new state.\\
 
-\textbf{Timestamps} that the user has to generate, for example in the \textsl{timeAdvance} function, can be easily constructed using the following constructor:\\ \textsl{t\_timestamp(size\_t)}).\\
+\textbf{Timestamps} that the user has to generate, for example in the \texttt{timeAdvance} function, can be easily constructed using the following constructor:\\ \texttt{t\_timestamp(size\_t)}).\\
 
@@
 \begin{itemize}
-	\item \textsl{t\_portptr Model::addInPort(std::string name)}
-	\item \textsl{t\_portptr Model::addOutPort(std::string name)}
+	\item \texttt{t\_portptr Model::addInPort(std::string name)}
+	\item \texttt{t\_portptr Model::addOutPort(std::string name)}
 \end{itemize}
@@
 \begin{itemize}
-	\item \textsl{t\_portptr Model::getPort(std::string name) const}
-	\item \textsl{std::vector{\textless}n\_network::t\_msgptr{\textgreater} Port:: \\createMessages(std::string message)}
+	\item \texttt{t\_portptr Model::getPort(std::string name) const}
+	\item \texttt{std::vector{\textless}n\_network::t\_msgptr{\textgreater} Port:: \\createMessages(std::string message)}
 \end{itemize} 
-Something like this would work: \textsl{this-{\textgreater}getPort("OUT")-{\textgreater} \\createMessages("my\_message");} \\
+Something like this would work: \texttt{this-{\textgreater}getPort("OUT")-{\textgreater} \\createMessages("my\_message");} \\
 
@@
 \section{Coupled models}
-Coupled models only require you to implement them as a sub-class of \\ \textsl{n\_model::CoupledModel}.
+Coupled models only require you to implement them as a sub-class of \\ \texttt{n\_model::CoupledModel}.
 
-\textbf{Adding sub-models} requires following function: \textsl{void CoupledModel:: \\addSubModel(const t\_modelptr\& model)}.
+\textbf{Adding sub-models} requires following function: \texttt{void CoupledModel:: \\addSubModel(const t\_modelptr\& model)}.
 
@@
 
-\textbf{Connecting these ports} can be done with the function\\ \textsl{void CoupledModel::connectPorts(const t\_portptr\& p1, const t\_portptr\& p2, t\_zfunc zFunction);}.\\ This function connects port \textsl{p1} to port \textsl{p2}.
+\textbf{Connecting these ports} can be done with the function\\ \texttt{void CoupledModel::connectPorts(const t\_portptr\& p1, const t\_portptr\& p2, t\_zfunc zFunction);}.\\ This function connects port \texttt{p1} to port \texttt{p2}.
 An optional Z function can be added that will modify the messages during transit from one port to another.
@@
 \begin{itemize}
-	\item \textsl{bool Model::modelTransition(DSSharedState* shared);}
+	\item \texttt{bool Model::modelTransition(DSSharedState* shared);}
 \end{itemize}

### 2015-05-23 manual/tex/reqs.tex
Requirements added to manual.

--- 2015-04-24 manual/tex/reqs.tex
@@
 Debian based:\\
-\lstinline!$sudo apt-get install libboost-all-dev!\\
+\texttt{\$sudo apt-get install libboost-all-dev}\\
 Red-hat based:\\
-\lstinline!$sudo yum install boost-devel!\\
+\texttt{\$sudo yum install boost-devel}\\
 \subsubsection{Preparing the build}
@@
 For the simulator this is resolved by CMake, however for gtest you'll need to manually build GTest with this extra flag:\\
-In file internal\textunderscore utils.cmake change compile options to append the compile flag -fpie :
-set(cxx\textunderscore base \textunderscore flags "-Wall -Wshadow -fpie")\\
+In file internal\textunderscore utils.cmake change compile options to append the compile flag \texttt{-fpie} :
+\texttt{set(cxx\textunderscore base\textunderscore flags "-Wall -Wshadow -fpie")}\\
 If you have any non-standard locations where the compiler should search for Boost and/or GTest, place them in the file localpreferences.txt.\\
-SET({GTEST|BOOST}\_ROOT "your path here").
+\texttt{SET(\{GTEST|BOOST\}\_ROOT "your path here")}.
 Make sure that for gtest, the listed path contains both the "include" folder and libraries in a dedicated folder "lib".
 For Boost the path needs to contain the "include" folder with the boost headers.
+\paragraph{NOTE} If you use Mac OS X, the project links with the \texttt{libc++} library by default.
+GTest does not.
+To prevent linker errors when compiling the project you should build the GTest libraries with the flag \texttt{-stdlib=libc++}.
 \subsubsection{Building}
-On *nix, a build script is provided to drive cmake:\\
-\lstinline!$main/generate_build.sh {Debug|Release} {g++|clang++}!\\
-Alternatively, invoke cmake manually :
-\lstinline!$mkdir build!\\
-\lstinline!$cd build!\\
-\lstinline!$cmake -G"Eclipse CDT4 - Unix Makefiles" -DCMAKE_CXX_COMPILER_ARG1=!\\
-\lstinline!"{-std=c++11|-std=gnu++11}" -DCMAKE_CXX_COMPILER="{clang++|g++}" !
-\lstinline!-DCMAKE_BUILD_TYPE={Release|Debug} ../main!
-\lstinline!$make -j{CPU-CORES}!\\
+On *nix, a build script is provided to drive CMake:\\
+\texttt{\$main/generate\_build.sh \{Debug|Release\} \{g++|clang++\} }\\
+Alternatively, invoke cmake manually : \\
+\texttt{\$mkdir build }\\
+\texttt{\$cd build}\\
+\texttt{\$cmake -G"Eclipse CDT4 - Unix Makefiles" -DCMAKE\_CXX\_COMPILER\_ARG1=\\"\{-std=c++11|-std=gnu++11\}" -DCMAKE\_CXX\_COMPILER="\{clang++|g++\}" \\-DCMAKE\_BUILD\_TYPE=\{Release|Debug\} ../main}\\
+\texttt{\$make -j\{CPU-CORES\}}\\
+
+On Windows the default script should work but if you come into any problems, the commands given above work as well.
 \subsubsection{Development}
@@
 \end{itemize}
+
+\subsubsection{Documentation}
+Generating the source documentation is optional and requires Doxygen to be installed. UML and other diagrams will only be created if Graphviz is installed.\\
+After building the source, enter the \textsl{build} folder and enter the command:\\
+\texttt{\$make doc}\\This will generate all documentation in the \textsl{doc} folder.

### 2015-05-23 manual/projectrapport.tex
* updated some documentation for the tracing module
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>
projectreport + how to build changes

@@
 When constructing a message by using the functionality provided by the Port class, this happens automatically.
-When receiving messages, the free function template is provided to extract the object from the message.\\
+When receiving messages, a free function template is provided to extract the object from the message.\\
 For example:
@@
 \end{lstlisting}
+\paragraph{Note} While it is technically possible, we strongly discourage the usage of the following data types in these messages:
+\begin{itemize}
+	\item raw pointers to heap-allocated object.\\
+			Messages can be destroyed during a time warp prior to being delivered to a model. This will result in a memory leak if there are no other references to these objects.
+	\item pointers or references to local objects.\\
+		When these objects go out of scope, they are destructed, rendering the pointer or reference invalid.
+\end{itemize}
 
@@
 \section{Translating Dynamic Typing to Static Typing}
-To represent objects that can be at runtime any type chosen by the user, we had 2 options : void* which is highly fragile in use, and impossible to serialize, or std::string(), which can double as an "anytype" by storing a serialized object inside the string, should this be required.
-\\This leaves the same degree of freedom dynamic typing grants to the user in PyPDEVS, with the added benefit of some checking provided by the compiler.
-Consider for example a user who stores an object serialized into the payload string of a message object, the compiler can in advance detect if the serializing object is at least serializable.
-\\ Misuse would either surface during compilation, or at worst could be checked with casting or exceptions.
-Misuse of void* could result in anything from undefined behaviour to stack corruption.
+Certain classes of the simulation, such as the messages between models and the model states, must be able to contain objects of any type.
+However, C++ is a statically typed language.
+To overcome this problem, we came up with the following three solutions:
+\begin{itemize}
+	\item Keep a \texttt{void*} that points to the actual data.
+This approach is very fragile and impossible to serialize.
+Misuse of \texttt{void*} could result in anything from undefined behaviour to stack corruption.
+	\item Serialize everything to \texttt{std::string}.
+Serialization is not a trivial operation.
+Continuously serializing and deserializing can be a very large performance hit.
+Moreover, it would require each and every type to be serializable.
+On the flip side, the compiler can help by detecting whether or not the object can be serialized or not.
+	\item Make use of inheritance and create a specialized subclass for each and every datatype.
+In the case of the messages, a class template can be used to facilitate this process.
+In the case of the model states, we decided that it is best that the user provides the subclass.\\
+	Since we use pointers for messages and states, a pointer cast is all that is needed to access the members of the specialized class.
+For extra safety, we suggest to use \texttt{std::dynamic\_pointer\_cast}.
+\end{itemize}
+We chose the third option because it both offers speed and safety.
 
@@
 A Scheduler was implemented based on the prototyping code. As required, as list based scheduler was implemented. \\
-The interface from PyPDevs' scheduler class was reused, but internally the scheduler class in our project differs on a few point. \\
+The interface from PyPDevs' scheduler class was reused, but internally the scheduler class in our project differs on a few points. \\
 \subsection{Composition}
@@
 We develop on Unix, Windows and MacOSX.
-
 In Windows we chose the Cygwin platform instead of MinGW to compile our project on.
 This because we make use of the C++11 std::thread objects; MinGW doesn't support POSIX threads.
-When building the project with GTest on Windows you need to use the -std=gnu++11 flag instead of the standard -std=c++11 flag, because we need the GNU extensions to correctly link everything.
 
@@
 libstdc++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}.
-This is triggers an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object.
+This triggers an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object.
 Explicitly invoking the string constructor with the char[] of the original string, proved to be a safe workaround here.
@@
 	\item levels can be individually switched on/off at compile time
-	\item No logging code is executed whatsoever if the logger is disabled
-	\item log messages contain filename and line number of the log command
+	\item No logging code is executed whatsoever if the logger is disabled.
+	\item log messages contain file name and line number of the log command.
 	\item basic crash safety. The logger will catch the following signals and flush all remaining output before exiting the program:
@@
 \end{itemize}
+
+\paragraph{Asynchronous output}
+To implement the asynchronous output, we made use of a custom stream buffer. The original design is by Dietmar K{\"u}hl \cite{asynchwrite}. We modified this design in order to make it thread safe.
+
+\section{Compiling multi-platform}
+A lot of obstacles were tackled while getting the project to work an Windows.
+When building with GTest on Windows you need to use the \texttt{-std=gnu++11} flag instead of the standard \texttt{-std=c++11} flag, because we need the GNU extensions to correctly link everything.
+
+All standard c++11 \textit{to string} functionality in Cygwin is not supported \cite{cygwin_tostring} (MinGW32 doesn't support this either).
+With basic \textit{to string} functionality we mean the functions that are listed here \cite{c++11_tostring} and under the 'See also' section.
+The problem seems to be a library called Newlib \cite{newlib} that doesn't provide crucial functions.
+
+Mac OSX needs the libc++ library to compile all C++11 features on.
 
@@
 While the difference is considerable, we should keep in mind that g3log generates timestamps for each message.
-It also contains several features that our logger doesn't, such as extensive crash safety.
+It also contains several features that our logger doesn't, such as a more extensive crash safety.
 Moreover, g3log allows the user to dynamically change which log levels are filtered, while this can only be decided at compile time for our logger.
-This can be clearly seen in the amount of branches.
+This can be clearly seen in the amount of branches taken by the program.
 
@@
 
+  \bibitem{asynchwrite} \url{http://stackoverflow.com/a/21127776}
+
+  \bibitem{c++11_tostring} \url{http://en.cppreference.com/w/cpp/string/basic_string/to_string}
+
+  \bibitem{cygwin_tostring} \url{https://cygwin.com/ml/cygwin/2015-01/msg00245.html}
+
+  \bibitem{newlib} \url{https://sourceware.org/newlib/}
+
   \end{thebibliography}

### 2015-05-23 manual/projectrapport.tex
Project rapport completion, switch core to use exceptions iso assert.

@@
 An RPC implementation would simply have to implement the Network class on both sides.
+\\
+Late in the project functionality was added to check if the network contained any transient messages.
+This is not required for any algorithms, but to allow Cores to safely stop in an optimistic simulation.
+Due to OS-scheduling of threads it is entirely possible for all cores to report being idle (having reached termination time) and with a message still in the network.
+The controller would decide the simulation is done and exit, which would be incorrect.
+Synchronizing the steps which decide if a core is idle and if there are messages would be more expensive than adding this functionality with an atomic boolean to the network.
 
@@
 \\
+\subsection{Reuse by other classes.}
+Several classes in the project require a heap-like functionality, by making the Scheduler class as generic as possible, we can reuse the same code instead of limiting its use to scheduling models only.
+The Scheduler class is  reused by the Tracers (scheduling tracing events) and the messaging queue in a kernel.
+Each client class can select at construction which heap backend to use, and if single operation synchronization is desired.
+\subsection{Entries}
+A scheduler instance can schedule any item provided it has a well defined implementation of the hash function and operator less than.
+In practice however, we used specialized Entry structs, so that intrusion upon the scheduled items was not necessary.
+\\
+Consider for example scheduling a model: \\ If the scheduling code relies only on the hash function of the model, this implementation can never change.
+But if we use a struct holding a reference/pointer/name to the model as the scheduled item, we can redefine any operator without interfering with other client classes of Model.
+This for example occurs in the Scheduler that holds pending incoming messages in the Kernel/Core, and the Tracers which has a scheduler for a subclass of Message.
+Both use specific locally defined entry classes without influencing each other.
 \subsection{Implementation}
@@
 In general, the Fibonacci heap is one of the fastest, with skew heap sometimes coming out ahead.
-The list based heap is always a factor 2 slower, though again this depends on the usage scenario.
+The list based heap is usually slower, though again this depends on the usage scenario.
 \\Important to note is that only node based heaps can be used efficiently (without compromising the performance of erase() and contains()).
 We store on insertion an iterator with the key of the pushed element in the hashmap.
-This also works as a safety check, preventing double insertion (depending ofcourse on the hash-specialization semantics).
+This also works as a safety check, preventing double insertion (depending on the hash-specialization semantics).
 This freedom of implementation is visible in the tracers where duplicate elements are allowed if they have a different address, contrasted with the model scheduler which disallows a model to be scheduled twice.\\
 Finally, the std::hash$<>$ specialization used by the message scheduler allows O(1) annihilation of a message by its antimessage (since the hash is identical).
-\subsection{Differences}
-We intended to use the scheduler class not only for scheduling models, but also to store pending incoming messages, and tracing messages.
-To allow this, the Scheduler class only requires from a scheduled item that it has operator$<$ and std::hash$<$T$>$ defined.
+Similar usage of erase() is required in the model scheduler to quickly invalidate old scheduled entries (massReschedule).
+
+\subsection{ADevs scheduler}
+The adevs scheduler is a binary heap, but does not allow erasure of entries which would render it incapable of fulfilling the Pdevs scheduler interface.
+The binary heap (e.g.
+stl/boost priority queue or any container + heap operations in stl) can't be used since we need to store pointers to entries, which requires a node based heap.
+\\It can be emulated by using the faster Fibonacci or skew heaps, or the slower list based heap.
+\\
+Note that Adevs's implementation is intrusive upon the model  (q\textunderscore index), which could make the scheduling code vulnerable. \\
+Should we implement a binary heap scheduler without nodes, the pointer correcting code on any heap operation would prohibit scaling, since any heap operation would invalidate the iterators pointing to the entries.
+Furthermore, a whole series of operations on non node based containers (eg vector) have this same invalidating behaviour, making the implementation unnecessarily complex.
 
@@
 We develop on Unix, Windows and MacOSX.
+
 In Windows we chose the Cygwin platform instead of MinGW to compile our project on.
 This because we make use of the C++11 std::thread objects; MinGW doesn't support POSIX threads.
+When building the project with GTest on Windows you need to use the -std=gnu++11 flag instead of the standard -std=c++11 flag, because we need the GNU extensions to correctly link everything.
 
@@
 This means that step 1.a , which is normally done by the receiver when it inspects the message queue, can be done immediately.
-This also avoids the queueing delay incurred by simulation traffic.\\
+This also avoids the queueing delay incurred by simulation traffic, and conversely does not interfere with the simulation traffic.\\
 Using the shared memory approach, we also avoid the sending of any synchronization message to other cores (required in ADEVS).
@@
 \\
-Also important to note is that there exist several edge cases not listed in this algorithm, but force the implementation to do more work than listed above.
+Also important to note is that there exist several edge cases not listed in this algorithm, but which force the implementation to do more work than listed above.
 Consider a kernel with a non-dependent model A (that receives no input).
@@
 The correct behaviour here is to replace eit with the vtime of the kernel at each simulation step and to use that in the algorithm.\\
- Furthermore, collecting lookahead values from models needs to be done in step 1, a lookahead value is calculated starting from the time it acquires a new state.
-This means that the lookahead value used in 2.a has to be continuously updated, regardless of usage, and has to be reset between simulation rounds (step 1 can entail several of such rounds).
-If we don't reset the minimum lookahead value in between rounds, there are scenarios where an incorrect minimum of a model that is no longer scheduled in the next round can warp the calculation.
+ Furthermore, collecting lookahead values from models needs to be done in step 1.a.
+The lookahead value is calculated starting from the time it acquires a new state, this means that the lookahead value used in 2.a has to be continuously updated, regardless of usage, and has to be reset between simulation rounds (step 1 can entail several of such rounds in the theoretic form of the algorithm).
+If we don't reset the minimum lookahead value in between rounds, there are scenarios where an incorrect minimum of a model that is no longer scheduled in the next round can warp the calculation, potentially freezing the simulation.
 \\
  The calculation done in steps 2, 3 are quite efficiently implemented, although a future optimization could delay step 3 until several rounds have passed, more in line with the theoretical algorithm.
+For details please consult the source code of Doxygen documentation of class Conservativecore.
 \subparagraph{Dependency calculation}
@@
 Using the allocation information provided by the Controller, the Kernel will then determine which other Kernels (including possibly itself) it is dependent on.\\
-It is important to note that the user need not interpret this dependency relation as transition, and needs to look only at 1-edge removed models. \\
+It is important to note that the user need not interpret this dependency relation as transitive, and needs to look only at 1-edge removed models. \\
 To clarify this last point: suppose we have 3 models A, B, C with A sending output to B, and B to C.
@@
 The simulator detects this and (eventually) halts the simulation.
-\\ Note that the above problem is not limited to valgrind, on Virtualized system this is more easily triggered than in a non virtualized setting.\\
-Furthermore, we also force an idle kernel (a kernel that has either reaches termination time, or can't advance time (no messages and no scheduled models) to yield/sleep for a short duration.
-The reason for this behaviour is to allow the other cores execution time, which would otherwise be spent in idling loops.
-A condition variable was evaluated as a solution, but not implemented since it also suffered from timing issues (time between thread awakening and core simulating).
+\\ Note that the above problem is not limited to valgrind, on a virtualized OS this is more easily triggered than in a non virtualized setting.\\
+Furthermore, we also force an idle kernel (a kernel that has either reached termination time, or can't advance time (no messages and no scheduled models) to yield/sleep for a short duration.
+The reason for this behaviour is to try to grant the other cores execution time, possibly alleviating thread starvation where an idle core (but non-idle thread) would starve a non-idle core.
+A condition variable was evaluated as a solution, but not implemented since it also suffered from timing issues (time between thread awakening and core simulating), and would conflict with the controlling code (save,pause) already in place.
 \section{Heaps}
@@
 \\ Time prevented us from writing up a formal bug report, since the bug is hard to reproduce and even harder to detect.
-The D-ary heap triggered an move-assing to self assertion when we used the Debugging version of the STL, so we had to remove both from our Schedulers.
+The D-ary heap triggered a "move-assign to self" assertion when we used the Debugging version of the STL, so we had to remove both from our Schedulers.
 This is no way impacted performance, as the Fibonacci heap was in practice faster (with the skew heap sometimes overtaking it).
@@
 To implement the asynchronous output, we made use of a custom stream buffer. The original design is by Dietmar K{\"u}hl \cite{asynchwrite}. We modified this design in order to make it thread safe.
-
-\section{Compiling multi-platform}
-A lot of obstacles were tackled while getting the project to work an Windows.
-When building with GTest on Windows you need to use the \texttt{-std=gnu++11} flag instead of the standard \texttt{-std=c++11} flag, because we need the GNU extensions to correctly link everything.
-
-All standard c++11 \textit{to string} functionality in Cygwin is not supported \cite{cygwin_tostring} (MinGW32 doesn't support this either).
-With basic \textit{to string} functionality we mean the functions that are listed here \cite{c++11_tostring} and under the 'See also' section.
-The problem seems to be a library called Newlib \cite{newlib} that doesn't provide crucial functions.
-
-Mac OSX needs the libc++ library to compile all C++11 features on.
 
@@
 
-  \bibitem{c++11_tostring} \url{http://en.cppreference.com/w/cpp/string/basic_string/to_string}
-
-  \bibitem{cygwin_tostring} \url{https://cygwin.com/ml/cygwin/2015-01/msg00245.html}
-
-  \bibitem{newlib} \url{https://sourceware.org/newlib/}
-
   \end{thebibliography}

### 2015-05-23 manual/projectrapport.tex

@@
 
+\section{Compiling multi-platform}
+A lot of obstacles were tackled while getting the project to work an Windows.
+When building with GTest on Windows you need to use the \texttt{-std=gnu++11} flag instead of the standard \texttt{-std=c++11} flag, because we need the GNU extensions to correctly link everything.
+
+All standard c++11 \textit{to string} functionality in Cygwin is not supported \cite{cygwin_tostring} (MinGW32 doesn't support this either).
+With basic \textit{to string} functionality we mean the functions that are listed here \cite{c++11_tostring} and under the 'See also' section.
+The problem seems to be a library called Newlib \cite{newlib} that doesn't provide crucial functions.
+
+Mac OSX needs the libc++ library to compile all C++11 features on.
+
 \chapter{Performance Analysis}
@@
 
+  \bibitem{c++11_tostring} \url{http://en.cppreference.com/w/cpp/string/basic_string/to_string}
+
+  \bibitem{cygwin_tostring} \url{https://cygwin.com/ml/cygwin/2015-01/msg00245.html}
+
+  \bibitem{newlib} \url{https://sourceware.org/newlib/}
+
   \end{thebibliography}

### 2015-05-23 manual/projectrapport.tex
* expanded the information about the tracers a bit
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>

--- 2015-05-23 manual/projectrapport.tex
@@
 
-\section{Compiling multi-platform}
-A lot of obstacles were tackled while getting the project to work an Windows.
-When building with GTest on Windows you need to use the \texttt{-std=gnu++11} flag instead of the standard \texttt{-std=c++11} flag, because we need the GNU extensions to correctly link everything.
-
-All standard c++11 \textit{to string} functionality in Cygwin is not supported \cite{cygwin_tostring} (MinGW32 doesn't support this either).
-With basic \textit{to string} functionality we mean the functions that are listed here \cite{c++11_tostring} and under the 'See also' section.
-The problem seems to be a library called Newlib \cite{newlib} that doesn't provide crucial functions.
-
-Mac OSX needs the libc++ library to compile all C++11 features on.
-
 \chapter{Performance Analysis}

### 2015-05-23 manual/tex/advanced.tex
* expanded the information about the tracers a bit
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>

@@
 All provided tracers make use of the policy pattern to separate the output generation from sending that output to an output sink.
-Currently, there is a policy for sending the output directly to standard out (\lstinline{CoutWriter}) or writing it to a file(\lstinline{FileWriter}).
-Both policies can be found in the \path{src/tracers/policies.h} header.
-Note that in the case of \lstinline{FileWriter}, the tracer must be explicitly initialized before it can be used.
-For more implementations, refer to the provided documentation.\\
-Currently, the simulator provides only one tracer class: the \lstinline{VerboseTracer}.
-This tracer will generate verbose output that is comparable to the verbose output of the Python PDEVS implementation by Jentl Van Tendeloo.
+Currently, there is a policy for sending the output directly to standard out (\lstinline{CoutWriter}) or writing it to a file(\lstinline{FileWriter}) or multiple files (\lstinline{MultiFileWriter}).
+All three policies can be found in the \path{src/tracers/policies.h} header.
+Note that in the case of \lstinline{FileWriter} and \lstinline{MultiFileWriter}, the tracer must be explicitly initialized before it can be used.
+For more details, refer to the provided documentation.\\
+Currently, the simulator provides four tracer classes:
+\begin{itemize}
+	\item \emph{VerboseTracer} Generates verbose output that is comparable to the verbose output of the Python PDEVS implementation by Jentl Van Tendeloo.
+	\item \emph{XMLTracer} The XML output generated by this tracer adheres to the format expected by the DEVS Trace Plotter by Hongyan (Bill) Song.
+	\item \emph{JSonTracer} The structure of the generated JSON is derived from the XML format.
+	\item \emph{CellTracer} The cell tracer is useful for creating output resembling a map or grid.
+It requires the models to derive from the \lstinline{CellAtomicModel} model class (see earlier).
+The cell tracer is the only tracer that fully supports the \lstinline{MultiFileWriter} policy.
+\end{itemize}
+
 \paragraph{Implementing a custom tracer}
@@
 \begin{lstlisting}
-interface Tracer: public TracerBase
+interface Tracer
 {
@@
 \end{lstlisting}
-Even though it is not explicitly enforced, it is highly recommended to allow the use of output policies.\\
+Even though it is not explicitly enforced, it is highly recommended to allow the use of output policies.
+Certain functions used by the simulator are provided by the policy and not by the tracer.
+See the section \emph{Implementing a custom output policy} for more details.\\
 Note that the tracer must not print its output upon generation. To make sure that the output is still generated in the correct order, the tracer must register a \lstinline{TraceMessage}.
@@
 For a more in-depth example, you can always take a look at the implementation of the \lstinline{VerboseTracer} or the \lstinline{CellTracer}.
+
+\paragraph{Implementing a custom output policy}
+All policies only implement a handful of functions. This is the interface that all policies are expected to implement:
+\begin{lstlisting}
+class CustomPolicy
+{
+public:
+  /**
+   * @brief Stops the tracer from writing the output.
+   * @warning	We assume that this function is not
+   *   called while a simulation is in progress.
+   *   If there is, we can make no guarantee when
+   *   the output generation actually stops.
+   * @note  When you implement a custom tracer
+   *   that doesn't use policies, it must
+   *   implement this function as well.
+   */
+	void stopTracer();
+
+  /**
+   * @brief Restarts the tracer.
+   * @param recover Whether or not the new output should
+   *   be appended to the old output or not,
+   *   if applicable.
+   * @note  When you implement a custom tracer
+   *   that doesn't use policies, it must
+   *   implement this function as well.
+   */
+	void startTracer(bool recover);
+
+protected:
+  /**
+   * @brief Default Constructor for this policy.
+   * All policies should have a default constructor.
+   */
+	CustomPolicy();
+
+  /**
+   * @brief Prints the data to the output sink.
+   *
+   * The data can have any type, as long as the proper
+   * output operator has been defined.
+   * This function is called whenever tracing output
+   * should be written to the output sink.
+   *
+   * @param data... These objects will be printed
+   *                to the output sink.
+   * @precondition The output operator << is specified
+   *  for printing each and every data object
+   *  to an std::ofstream.
+   */
+	template<typename ... Args>
+	inline void print(const Args&... args);
+};
+\end{lstlisting}
+Some policies may need initialization.
+Initializing the policy should happen in a separate \lstinline{initialize} member function.
+Examples of policies that use such a function are \lstinline{FileWriter} and \lstinline{MultiFileWriter}.

### 2015-05-23 main/src/tracers/celltracer.h

@@ -21,8 +21,11 @@ namespace n_tracers{
 using namespace n_network;
 
 /**
- * @brief Tracer that will generate verbose output.
+ * @brief Tracer that will generate cell output.
  * @tparam OutputPolicy A policy that dictates what should happen with the output
+ *
+ * This tracer requires the models to be a subclass of the CellAtomicModel class.
+ * If a model does not derive from this type, it is ignored.
  */
 template<typename OutputPolicy, std::size_t XSize, std::size_t YSize>
 class CellTracer: public OutputPolicy

### 2015-05-23 manual/tex/models.tex
* added a tiny bit about the cell atomic model to the manual
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>

@@
 
+\subsection{Cell Atomic Model}
+The \texttt{CellAtomicModel} is a specialized atomic model that provides functionality for the cell tracer.
+It is useful when models are laid out in a grid.
+This class extends the interface of the atomic model with methods for getting or setting the position of this model in the grid.
+
 \section{Coupled models}

### 2015-05-23 main/src/model/cellmodel.h

@@ -16,8 +16,8 @@ namespace n_model {
 typedef std::pair<std::size_t, std::size_t> t_point;
 
 /**
- * @brief Specialized state that contains a position.
- * This state can be used for the Cell Tracer
+ * @brief Specialized atomic model that contains a position.
+ * This model can be used for the Cell Tracer
  */
 class CellAtomicModel: public n_model::AtomicModel
 {

### 2015-05-23 manual/tracersReport.tex

@@
 \subsection{Constructor}
-The non-default constructor can be used to initialize all the tracers at once. It uses the copy constructor to initialize the tracer members. There is also a default constructor.
+The non-default constructor can be used to initialize all the tracers at once. It uses the copy constructor to initialize the tracer members. There is also a default constructor.\\
+\paragraph{note:} The individual tracer classes must all implement either a copy constructor or a default constructor.
 \subsection{getParent}
@@
 Only one of those options should compile correctly with any given template arguments.
-When the compiler fails to substitute the template arguments in one option, it continues with the other options until it either finds a single suitable option.
-When no suitable option can be found, the result is a compiler error.\\
+When the compiler fails to substitute the template arguments in one option, it continues with the other options until it either finds a single suitable option or no suitable option at all.
+In the latter case, the result is a rather convoluted compiler error.\\
 \subsubsection{implementation}
@@
 Therefore, depending on the value of the template argument \inlinecpp{n}, the return value of one of the three options is defined and the others are not, resulting in a substitution error.\\
-If n is too large, that is, n is larger than or equal to the amount of used tracers, the last option is taken.
+If \inlinecpp{n} is too large, that is, n is larger than or equal to the amount of used tracers, the last option is taken.
 Because that function is deleted, the compiler\footnote{Only tested with GCC 4.9.2.} generates an error that points to where the function was invoked.\\
-\textbf{Note} that you cannot use the \inlinecpp{>} operator in the \inlinecpp{std::enable_if} condition.
+\textbf{Note} You cannot use the \inlinecpp{>} operator in the \inlinecpp{std::enable_if} condition.
 The compiler gets confused because it fails to distinguish between the comparison operator and the closing angle bracket of the template.
@@
 \section{What about std::tuple?}
-\inlinecpp{std::tuple<...>} already offers a lot of functionality that we implemented ourself.
+\inlinecpp{std::tuple<...>} already offers a lot of functionality that we implemented ourselves.
 The reason why we chose to reinvent the wheel over using \inlinecpp{std::tuple<...>} is because our implementation makes it very easy to loop over all the registered tracer objects and call one of their functions.\\
@@
 \end{itemize}
-Both the \emph{execute function} and the \emph{delete function} are objects of type \inlinecpp{std::function<void()>}. You can use \inlinecpp{std::bind<...>(...)} to create a delegate.
+Both the \emph{execute function} and the \emph{delete function} are objects of type \inlinecpp{std::function<void()>}.
+You can use \inlinecpp{std::bind<...>(...)} to create a delegate and pass that object as a parameter for the constructor.
 
@@
 \end{itemize}
-In the background, \emph{traceUntil} will create a new thread that will perform the actual output.
-Therefore, rest of the simulation does not have to wait until this job is done.
-The current implementation can still be optimized further, for example by using a 1 reader/multiple writers approach.
-We will try to fix this for the beta version.
+In the background, \emph{traceUntil} will create a new thread that will perform the actual output. Therefore, rest of the simulation does not have to wait until this job is done.
 

### 2015-05-24 manual/tex/controller.tex
Update manual (added serialization + corrections)

@@
 The first part to take into consideration is the \textbf{allocator}, which decides on which simulation core to place which model.
-We have already included a simple allocator (called \textsl{SimpleAllocator}) that distributes models evenly, but for more complex behavior it is possible to create one tailored specifically for your simulation purposes by inheriting from the \textsl{Allocator} base class.\\
+We have already included a simple allocator (called \textsl{SimpleAllocator}) that distributes models evenly, but for more complex behavior it is possible to create one tailored specifically to your simulation purposes by inheriting from the \textsl{Allocator} base class.\\
 

### 2015-05-24 manual/tex/faq.tex
Update manual (added serialization + corrections)

@@
-\section{Platform specific}
-\subsubsection{Virtualization}
-\begin{enumerate}
-\item On Windows host, with a *nix image, it's possible that standard compliant code leaks memory in std::string. We're aware of this issue and resolve it where we detect it with Valgrind.
-\item Threading is significantly slower depending on your host, you should at least allocate 2 virtual CPU's to your virtual instance to allow proper functioning of the simulator.
-Several threaded tests can fail , since the runtime can only provide a hint (std::thread::hardware\textunderscore concurrency()) to the program to detect if we can use threads.
-\end{enumerate}
-\subsubsection{Cygwin}
-\begin{enumerate}
-\item std::to\textunderscore string is not supported, do not use it in your own models.
-\item GTest version 1.7 corrupts the stack if a test fails, use version 1.6 until GTest resolves this.
-\end{enumerate}
 \section{General considerations}
@@
 \end{enumerate}
+
+\section{Platform specific}
+\subsubsection{Virtualization}
+\begin{enumerate}
+\item On Windows host, with a *nix image, it's possible that standard compliant code leaks memory in std::string. We're aware of this issue and resolve it where we detect it with Valgrind.
+\item Threading is significantly slower depending on your host, you should at least allocate 2 virtual CPU's to your virtual instance to allow proper functioning of the simulator.
+Several threaded tests can fail , since the runtime can only provide a hint (std::thread::hardware\textunderscore concurrency()) to the program to detect if we can use threads.
+\end{enumerate}
+\subsubsection{Cygwin}
+\begin{enumerate}
+\item std::to\textunderscore string is not supported, do not use it in your own models. Some old C functions like atoi and stoi are not supported either.
+\item GTest version 1.7 corrupts the stack if a test fails, use version 1.6 until GTest resolves this.
+\end{enumerate}
+
 \section{Threading}
@@
 \item Never use more simulator cores than hardware threads you have available.
-For example on a dual core cpu with hyperthreading, 4 simualtor-cores is the limit you should adhere to.
-The simulator may work with more simulator-cores, but performance will be crippled and starvation is likely.
+For example on a dual core cpu with hyperthreading, 4 simualtor cores is the limit you should adhere to.
+The simulator may work with more simulator cores, but performance will be crippled and starvation is likely.
 \item libstdc++'s implementation of std::string is Copy on Write, this introduces an inherent danger for race conditions.

### 2015-05-24 manual/tex/models.tex
Update manual (added serialization + corrections)

@@
 \section{States}
-You can can make your own states with almost no limitations.
+You can make your own states with almost no limitations.
 The only requirements for a user's own states are that they should be objects from a type that implements \texttt{n\_model::State}.
@@
 While members in sub-classes can be added without any problems, by default there is a public member \texttt{std::string m\_state} implemented in \texttt{State} that can be used to represent the user's state.
-\\ For serialization you have to declare (after class definition) 
-\begin{verbatim}
+
+\subsubsection{Serialization}
+For serialization you have to define three serialization functions with following signature: 
+\begin{lstlisting}
+void serialize(n_serialization::t_oarchive& archive) {
+	archive(cereal::virtual_base_class<State>( this ));
+}
+\end{lstlisting}
+\begin{lstlisting}
+void serialize(n_serialization::t_iarchive& archive) {
+	archive(cereal::virtual_base_class<State>( this ));
+}
+\end{lstlisting}
+\begin{lstlisting}
+static void load_and_construct(n_serialization::t_iarchive& archive,
+			cereal::construct<MyState>& construct) {
+	construct("");
+	construct->serialize(archive);
+}
+\end{lstlisting}
+In case your own states uses more datamembers, you can include them to the serialization by adding them at the back of the parameter list of archive() in the first two functions.
+For more information consult the Cereal manual.\\
+\\ 
+For example:
+\begin{lstlisting}
+int a = 5;
+archive(cereal::virtual_base_class<State>( this ), a);
+\end{lstlisting}
+After the class definition you have to register your new state throug the following macro: 
+\begin{lstlisting}
 CEREAL_REGISTER_TYPE(MyState);
-\end{verbatim}
+\end{lstlisting}
+\textbf{NOTE} the registration has to be done outside of any namespaces to work properly.
 
@@
 	\item \texttt{t\_timestamp AtomicModel::timeAdvance() const}\\
-		Should return the model's current time advance value; the amount of time after which the following internal transition occurs. This function should not alter the model in any way.
+		Should return the model's current time advance value, the amount of time after which the following internal transition occurs. This function should not alter the model in any way.
 	\item \texttt{std::vector{\textless}n\_network::t\_msgptr{\textgreater} AtomicModel::output()  const}\\
-		Returns the model's output; all messages that are to be send should be returned via this function. This function should not alter the model in any way.
+		Returns the model's output; all messages that are to be sent should be returned via this function. This function should not alter the model in any way.
 \end{enumerate}
@@
 This string should be given to the port you want to send it through.
-The port will then create the actual messages (that are send over the models) and return them to you in a vector.
+The port will then create the actual messages (which are sent over the models) and return them to you in a vector.
 All these messages (from possibly different ports) should be joined in 1 vector and be returned in the output function.
@@
 
-It's important that a model name is unique over the whole simulation, and portname should be unique per model.\\
-
-For serialization you have to declare (after class definition)
-\begin{verbatim}
-CEREAL_REGISTER_TYPE(MyModel);
-\end{verbatim}
+It's important that a model name is unique over the whole simulation, and port name should be unique per model. \\
 
@@
 Incomplete models or undefined behaviour may be the result of said incorrect use.
+
+\subsubsection{Serialization}
+For serialization you have to define three serialization methods with following signature: 
+\begin{lstlisting}
+void serialize(n_serialization::t_oarchive& archive) {
+	archive(cereal::virtual_base_class<AtomicModel>( this ));
+}
+\end{lstlisting}
+\begin{lstlisting}
+void serialize(n_serialization::t_iarchive& archive) {
+	archive(cereal::virtual_base_class<AtomicModel>( this ));
+}
+\end{lstlisting}
+\begin{lstlisting}
+static void load_and_construct(n_serialization::t_iarchive& archive,
+			cereal::construct<MyAtomicModel>& construct) {
+	construct("");
+	construct->serialize(archive);
+}
+\end{lstlisting}
+In case your own atomic models uses more datamembers, you can include them to the serialization by adding them at the back of the parameter list of archive() in the first two functions.
+For more information consult the Cereal manual.\\
+\\ 
+For example:
+\begin{lstlisting}
+int a = 5;
+archive(cereal::virtual_base_class<AtomicModel>( this ), a);
+\end{lstlisting}
+After the class definition you have to register your new atomic model throug the following macro: 
+\begin{lstlisting}
+CEREAL_REGISTER_TYPE(MyAtomicModel);
+\end{lstlisting}
+
+\textbf{NOTE} the registration has to be done outside of any namespaces to work properly.
 
@@
 
+\subsubsection{Serialization}
+For serialization you have to define three serialization methods with following signature: 
+\begin{lstlisting}
+void serialize(n_serialization::t_oarchive& archive) {
+	archive(cereal::virtual_base_class<CoupledModel>( this ));
+}
+\end{lstlisting}
+\begin{lstlisting}
+void serialize(n_serialization::t_iarchive& archive) {
+	archive(cereal::virtual_base_class<CoupledModel>( this ));
+}
+\end{lstlisting}
+\begin{lstlisting}
+static void load_and_construct(n_serialization::t_iarchive& archive,
+			cereal::construct<MyCoupledModel>& construct) {
+	construct("");
+	construct->serialize(archive);
+}
+\end{lstlisting}
+In case your own coupled models uses more datamembers, you can include them to the serialization by adding them at the back of the parameter list of archive() in the first two functions.
+For more information consult the Cereal manual.\\
+\\ 
+For example:
+\begin{lstlisting}
+int a = 5;
+archive(cereal::virtual_base_class<CoupledModel>( this ), a);
+\end{lstlisting}
+After the class definition you have to register your new atomic model throug the following macro: 
+\begin{lstlisting}
+CEREAL_REGISTER_TYPE(MyCoupledModel);
+\end{lstlisting}
+
+\textbf{NOTE} the registration has to be done outside of any namespaces to work properly.
+
 \section{Dynamic Structured DEVS}

### 2015-05-24 main/src/model/rootmodel.cpp
Controller does no longer direct connect a loaded coupled model

@@ -20,6 +20,25 @@ RootModel::~RootModel()
 {
 }
 
+void RootModel::setComponents(t_coupledmodelptr& model)
+{
+	std::deque<t_coupledmodelptr> toDo;
+	toDo.push_back(model);
+	while (!toDo.empty()) {
+		t_coupledmodelptr& top = toDo.front();
+		toDo.pop_front();
+		for (t_modelptr& current : top->getComponents()) {
+			current->setParent(top);
+			t_atomicmodelptr atomic = std::dynamic_pointer_cast<AtomicModel>(current);
+			if (atomic) {
+				m_components.push_back(atomic);
+			} else {
+				toDo.push_back(std::dynamic_pointer_cast<CoupledModel>(current));
+			}
+		}
+	}
+}
+
 const std::vector<t_atomicmodelptr>& n_model::RootModel::directConnect(t_coupledmodelptr& model)
 {
 	if (m_directConnected)

### 2015-05-24 main/src/model/rootmodel.h

@@ -22,6 +22,7 @@ private:
 public:
 	RootModel();
 	virtual ~RootModel();
+	void setComponents(t_coupledmodelptr& model);
 	/**
 	 * @precondition All atomic models have a unique name
 	 */

### 2015-05-24 manual/performancereport.tex
Additions to manual Performance report

\documentclass[10pt,a4paper]{report}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{fancyvrb}
\usepackage{amsfonts}
\usepackage{amssymb}

\date{}
\title{\textsc{DEVS Ex Machina} \\ Performance Report}
\begin{document}
\maketitle

\chapter*{DEVStone}
To measure the performance of \textsc{DEVS Ex Machina} under more demanding conditions, \textbf{DEVStone} and \textbf{PHOLD} benchmarks were used. Everytime we compare the performance of our implementation with \textit{PythonPDEVS}.\\

\section{Classic DEVS}
Using Classic DEVS we see noticeable speedups compared to PyPDEVS.
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_devstone

 Performance counter stats for './dxexmachina_devstone':

        314.286326 task-clock (msec)         #    1.125 CPUs utilized          
               410 context-switches          #    0.001 M/sec                  
             1,007 cpu-migrations            #    0.003 M/sec                  
               662 page-faults               #    0.002 M/sec                  
       802,295,514 cycles                    #    2.553 GHz                     [85.27%]
       447,423,794 stalled-cycles-frontend   #   55.77% frontend cycles idle    [84.45%]
       234,246,612 stalled-cycles-backend    #   29.20% backend  cycles idle    [71.04%]
       739,746,191 instructions              #    0.92  insns per cycle        
                                             #    0.60  stalled cycles per insn [85.97%]
       161,577,264 branches                  #  514.108 M/sec                   [85.84%]
         3,030,149 branch-misses             #    1.88% of all branches         [83.25%]

       0.279274362 seconds time elapsed
\end{Verbatim}

\begin{Verbatim}[fontsize=\small]
$ perf stat python experiment.py default 2 false > out.txt

 Performance counter stats for 'python experiment.py def 2 false':

        709.776558 task-clock (msec)         #    0.989 CPUs utilized          
               191 context-switches          #    0.269 K/sec                  
                34 cpu-migrations            #    0.048 K/sec                  
             2,942 page-faults               #    0.004 M/sec                  
     1,687,298,167 cycles                    #    2.377 GHz                     [82.71%]
       866,759,527 stalled-cycles-frontend   #   51.37% frontend cycles idle    [83.25%]
       562,247,464 stalled-cycles-backend    #   33.32% backend  cycles idle    [67.62%]
     1,672,919,036 instructions              #    0.99  insns per cycle        
                                             #    0.52  stalled cycles per insn [83.82%]
       375,443,818 branches                  #  528.961 M/sec                   [83.25%]
        14,867,814 branch-misses             #    3.96% of all branches         [83.28%]

       0.717576473 seconds time elapsed
\end{Verbatim}

\section{Parallel DEVS}
Unfortunately, there's quite a big gap in performance between PyPDEVS and DEVS Ex Machina in parallel simulation. The issues causing this are being ironed out.
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_devstone

 Performance counter stats for './dxexmachina_devstone':

       2666.499943 task-clock (msec)         #    0.017 CPUs utilized          
            16,835 context-switches          #    0.006 M/sec                  
               376 cpu-migrations            #    0.141 K/sec                  
             1,865 page-faults               #    0.699 K/sec                  
     5,780,070,450 cycles                    #    2.168 GHz                     [86.46%]
     2,313,049,054 stalled-cycles-frontend   #   40.02% frontend cycles idle    [84.79%]
     1,523,702,578 stalled-cycles-backend    #   26.36% backend  cycles idle    [58.20%]
     6,862,832,656 instructions              #    1.19  insns per cycle        
                                             #    0.34  stalled cycles per insn [83.73%]
     1,698,388,723 branches                  #  636.936 M/sec                   [85.64%]
        25,395,874 branch-misses             #    1.50% of all branches         [85.64%]

     156.945218942 seconds time elapsed
\end{Verbatim}

\begin{Verbatim}[fontsize=\small]
$ perf stat mpirun -np 2 python experiment.py def 2 false > out.txt

 Performance counter stats for 'mpirun -np 2 python experiment.py def 2 false':

        787.125992 task-clock (msec)         #    0.371 CPUs utilized          
             1,120 context-switches          #    0.001 M/sec                  
                91 cpu-migrations            #    0.116 K/sec                  
             8,684 page-faults               #    0.011 M/sec                  
     1,786,217,349 cycles                    #    2.269 GHz                     [83.85%]
       935,480,076 stalled-cycles-frontend   #   52.37% frontend cycles idle    [84.18%]
       619,366,137 stalled-cycles-backend    #   34.67% backend  cycles idle    [65.26%]
     1,796,463,414 instructions              #    1.01  insns per cycle        
                                             #    0.52  stalled cycles per insn [84.16%]
       403,042,819 branches                  #  512.044 M/sec                   [84.70%]
        13,210,328 branch-misses             #    3.28% of all branches         [82.68%]

       2.123888919 seconds time elapsed
\end{Verbatim}

\chapter*{PHOLD}
\section{Classic DEVS}
Just like DEVStone, PHOLD is quite a bit faster using our implementation. The CPU is used more efficiently.
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_phold

 Performance counter stats for './dxexmachina_phold':

         24.249048 task-clock (msec)         #    1.109 CPUs utilized          
                16 context-switches          #    0.660 K/sec                  
               117 cpu-migrations            #    0.005 M/sec                  
               688 page-faults               #    0.028 M/sec                  
        52,853,150 cycles                    #    2.180 GHz                     [83.16%]
        29,014,020 stalled-cycles-frontend   #   54.90% frontend cycles idle    [83.16%]
        14,103,460 stalled-cycles-backend    #   26.68% backend  cycles idle    [65.29%]
        38,553,838 instructions              #    0.73  insns per cycle        
                                             #    0.75  stalled cycles per insn [83.31%]
         8,242,520 branches                  #  339.911 M/sec                   [96.13%]
           185,357 branch-misses             #    2.25% of all branches         [83.71%]

       0.021871681 seconds time elapsed
\end{Verbatim}

\begin{Verbatim}[fontsize=\small]
$ perf stat python experiment.py def 2 false > out.txt

 Performance counter stats for 'python experiment.py def 2 false':

        709.776558 task-clock (msec)         #    0.989 CPUs utilized          
               191 context-switches          #    0.269 K/sec                  
                34 cpu-migrations            #    0.048 K/sec                  
             2,942 page-faults               #    0.004 M/sec                  
     1,687,298,167 cycles                    #    2.377 GHz                     [82.71%]
       866,759,527 stalled-cycles-frontend   #   51.37% frontend cycles idle    [83.25%]
       562,247,464 stalled-cycles-backend    #   33.32% backend  cycles idle    [67.62%]
     1,672,919,036 instructions              #    0.99  insns per cycle        
                                             #    0.52  stalled cycles per insn [83.82%]
       375,443,818 branches                  #  528.961 M/sec                   [83.25%]
        14,867,814 branch-misses             #    3.96% of all branches         [83.28%]

       0.717576473 seconds time elapsed
\end{Verbatim}

\section{Parallel DEVS}
PHOLD in parallel on PyPDEVS crashed immediately with an MPI error, hence a comparison was not possible. However, given the results of our implementation we can assume that here too, the parallel simulation would be quite a bit slower.
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_phold

 Performance counter stats for './dxexmachina_phold':

        820.288154 task-clock (msec)         #    0.003 CPUs utilized          
            20,030 context-switches          #    0.024 M/sec                  
               331 cpu-migrations            #    0.404 K/sec                  
               831 page-faults               #    0.001 M/sec                  
     1,025,470,316 cycles                    #    1.250 GHz                     [97.69%]
       767,298,149 stalled-cycles-frontend   #   74.82% frontend cycles idle    [99.29%]
       536,886,871 stalled-cycles-backend    #   52.36% backend  cycles idle    [52.05%]
       480,746,733 instructions              #    0.47  insns per cycle        
                                             #    1.60  stalled cycles per insn [99.54%]
       118,012,980 branches                  #  143.868 M/sec                   [53.18%]
         8,481,176 branch-misses             #    7.19% of all branches         [99.62%]

     247.566439823 seconds time elapsed
\end{Verbatim}

\end{document}

### 2015-05-24 manual/tex/controller.tex
Additions to manual Performance report

--- 2015-04-25 manual/tex/controller.tex
@@
 \begin{tabular}{lp{8cm}}
+\textit{simType} & The behavior of the Parallel DEVS:
+
+	\begin{tabular}{l|l}
+	\textsf{Option} & \textsf{Meaning} \\
+	\texttt{OPTIMISTIC} & \emph{Optimistic PDEVS Simulation} \\
+	\texttt{CONSERVATIVE} & \emph{Conservative PDEVS Simulation} \\
+	\end{tabular} \newline
+
+	The default is \emph{OPTIMISTIC} \newline
+	\textit{[This option is disregarded if you do not run PDEVS]}
+\\\\
 \textit{coreAmount} & The amount of cores the simulation will run on \newline
@@
 You can choose to let the simulation end at a certain moment by adding a \textbf{termination time} or a \textbf{termination condition}.\\
-The termination time can be set using the controller's \textsl{setTerminationTime} method, the termination condition by using \textsl{setTerminationCondition} and passing a custom \textsl{TerminationFunctor}.\\ %TODO add manual terminationfunctor?
+The termination time can be set using the controller's \textsl{setTerminationTime} method, the termination condition by using \textsl{setTerminationCondition} and passing a custom \textsl{TerminationFunctor}.
+
+\subsection{Events}
+If you want to schedule certain events while the simulation is running, like pausing and saving, you can use the Controller's \textsl{addPauseEvent} and \textsl{addSaveEvent} methods.\\
 \\
+The \textsl{addPauseEvent} method requires the \emph{simulated time} at which the simulator has to pause and how long the pause should take (in seconds).\\
+The \textsl{addSaveEvent} method requires the \emph{simulated time} at which the simulation has to be saved and the \textsl{prefix} of the saved file.
+For example: a save at the simulated time \textit{100} with the prefix \texttt{mySimSave} will result in a file \texttt{mySimSave\_100.devs}.
+
+For both events, the boolean value \textit{true} can be added as an additional argument to make the event repeat itself until the simulation ends.
+
+\subsubsection{Loading a previous simulation}
+Loading a previous simulation from a savefile follows the same procedure as setting up a regular simulation for the most part.
+First the simulation is configured (optionally by using the ControllerConfig object as described above), then instead of loading a new model, you can use the Controller's \textsl{load} method to load the data of your previous simulation.\\
+\emph{Important: To load a simulation \textbf{do not} have to use the same configuration!}
+
+\section{Starting the simulation}
 Finally, you can start the DEVS using the controller's \textsl{simulate} method.
-%TODO Serializatie: save, load

### 2015-05-24 manual/projectrapport.tex

@@
   \item Core : alias kernel.
-Single instance, containing a set of models it is responsible for, exchanges messages generated by those models and keeps time state.
+Single instance, containing a set of models for which it is responsible, exchanges messages generated by those models and keeps time state.
 Can correspond with Logical Process in terms of what it represents.
@@
 \section{Optimization}
-Week 4 , general review.
+Week 4 general review.
 
@@
 The project requires a compliant C++11 compiler and library, but beyond that is not really bound to special features.
-For testing, GTest is a requirement, and the only two other dependencies are the ubiquitous Boost and Cereal, the last is provided with the project as it is header only.
+For testing, GTest is a requirement, and the only two other dependencies are the ubiquitous Boost and Cereal, the latter is provided with the project as it is header only.
 
@@
 Passing by value could have been safer (e.g.
-between threads) but would introduce slicing.\\ Another advantage was that we would use as little memory as possible, for example sending a message means passing an const reference to a shared pointer, which is usually equal in size to 2 pointers.
+between threads) but would introduce slicing.\\ Another advantage was that we would use as little memory as possible, for example sending a message means passing a const reference to a shared pointer, which is usually equal in size to 2 pointers.
 By passing by const ref we minimize most of the penalty usually involved in copyconstructing shared pointers.
@@
 For serialization we used the Cereal \cite{cereal} header-only framework.
-We have used separate load and save functions for full flexibility.
-For each type that occurs in the form of a pointer we implemented a static load and construct function since the object first has to be created before the saved data can be imported.
+We opted for separate load and save functions for full flexibility.
+For each type that occurs in the form of a pointer we implemented a static load and construct method since the object first has to be created before the saved data can be imported.
 Cereal has a small coupling with the code (models) that has to be written by the user as the user has to register his own classes due to polymorphism.
+
+All the main functionality to load objects is implemented in the serialize function itself.
+Most classes also have an implementation for load and construct so that objects can separately be constructed before filling it with the saved data.
+This is necessary in cases where pointers to objects of those classes have to be serialized or where there is no default constructor available.
+The implementation of those load and construct methods is always restricted to creating an object with - most of the time - meaningless temporary datamembers after which the serialize function of the created object is called to load the actual data.
+
+Since Cygwin is not compatible with the std::to\_string() function, which is used for serialization to human readable formats like JSON or XML, only serialization to binary files is supported.
+Even in the binary archive implementation of Cereal the std::to\_string() function occurred, though only on non-decisive places, resulting in some slightly modified headers for this project.
+The binary archives are lighter and give more speed than the user-readable formats but comes with the disadvantage of not providing full garanty that saved data can be correctly loaded on another architecture.
+
+Since Cereal was hard to use and gave a lot of undesired errors while serializing const datamembers and objects of purely virtual classes, we decided to violate some proper design rules to overcome this.
+
+As we have experienced problems with an unressolved bug from Cereal - covering the errors in case of specific nested circular smart pointers - we were forced to not always follow the elegant design of Cereal and explicit force the calling of some serialization function in the load and construct functions.
 

### 2015-05-24 manual/tex/reqs.tex

@@
 We use thread-sanitizer to verify thread-safety at runtime instrumented by the compiler (g++ and clang).
-This however requires building the project (and by extension any linked libraries) with position independent flags.
+This however requires building the project (and by extension any linked libraries) with position-independent flags.
 For the simulator this is resolved by CMake, however for gtest you'll need to manually build GTest with this extra flag:\\
@@
 
-On Windows the default script should work but if you come into any problems, the commands given above work as well.
+On Windows the default script should work but if you come into any problems, the commands given above work if you use the CXX compiler argument \texttt{-std=gnu++11}.
 \subsubsection{Development}

### 2015-05-24 manual/projectrapport.tex
GXVAD XAXFX FXVAD FGAGA GDGAG AVAGX VDGGF DVXAV AXFXF VDDDV VFDVA GAGGG DDFGD DGVDA VVFVV VGAVX

@@
 Lookahead is collected per model per kernel, eit/eot are based on information from models but determined on a core/kernel level.
-
-\chapter{Planning}
-\section{Parallel}
-\begin {itemize}
-\item GVT integration : week 0-1, Ben and Tim
-\item Conservative : week 2-3 , Ben and Tim
-\end {itemize}
-\section{General}
-\begin {itemize}
-\item Dynamic Structured : week 1-3 Stijn
-\item DevStone and examples : week 0-2 Matthijs and Pieter
-\item Serialization : week 0 Pieter
-\item Simulation control : week 1-3 Matthijs and Pieter  
-\item Tracers : week 0 Stijn
-\item Benchmarking PyPDevs Adevs : week 3 Matthijs and Pieter
-\end {itemize}
-\section{Optimization}
-Week 4 general review.
 
@@
 
-\chapter{Performance Analysis}
-Whenever we made big design/architecture decisions, we first wrote some prototyping testcode to prove the viability (e.g. Network, Tracing, Logger).
-\section{First comparison}
-We use perf to get statistically sound benchmarks. The command run was :
-\lstinline!$perf stat -r {rounds} {command}!\\
-
-\subsection{Classic}
-An endtime of 3.6e6 was set, and both examples run against each other.
-\lstinputlisting[]{pycla.txt}
-The following shows a recording of the C++ implementation of the same example.\\
-\lstinputlisting[]{cpcla.txt}
-\subsubsection{Conclusion}
-The C++ port is approximately 12.906 times faster.
-
-\subsection{Dynamic structured}
-An end-time of 3.6e6 was set, and both examples run against each other.
-\lstinputlisting[]{pydyn.txt}
-The following shows a recording of the C++ implementation of the same example.\\
-\lstinputlisting[]{cpdyn.txt}
-\subsubsection{Conclusion}
-This result is still warped due to the startup/shutdown time of GTEST, but allready the C++ port is roughly 7.05 times faster, only page faults and migrations are worse (but being worked on).
-\subsection{Parallel}
-With GVT not yet active, an endtime of 3.6e6 was used (as before).\lstinputlisting[]{pypar.txt}\
-The following shows a recording of the C++ implementation of the same example.\\
-\lstinputlisting[]{cppar.txt}
-\subsubsection{Conclusion}
-The C++ port is approximately 19.6 times faster.
+\chapter{Logger analysis}
 

### 2015-05-24 manual/tex/example.tex
GXVAD XAXFX FXVAD FGAGA GDGAG AVAGX VDGGF DVXAV AXFXF VDDDV VFDVA GAGGG DDFGD DGVDA VVFVV VGAVX

Extensive examples are provided in our Controller testing classes. For each formalism a model was created which the user can use as a reference implementation.\\

\subsection{example}

\begin{lstlisting}
std::ofstream filestream("controller/devstestcoupled.txt");
{
	CoutRedirect myRedirect(filestream);
	auto tracers = createObject<n_tracers::t_tracerset>();

	std::unordered_map<std::size_t, t_coreptr> coreMap;
	std::shared_ptr<Allocator> allocator = 
		createObject<SimpleAllocator>(1);
	std::shared_ptr<n_control::LocationTable> locTab = 
		createObject<n_control::LocationTable>(1);

	t_coreptr c = createObject<Core>();
	coreMap[0] = c;

	Controller ctrl("testController", coreMap, allocator, 
		locTab, tracers);
	ctrl.setClassicDEVS();
	ctrl.setTerminationTime(t_timestamp(360, 0));

	t_coupledmodelptr m1 = createObject<n_examples_coupled::
		TrafficSystem>("trafficSystem");
	ctrl.addModel(m1);

	ctrl.simulate();
	EXPECT_TRUE(c->isLive() == false);
	EXPECT_TRUE(c->getTime() >= t_timestamp(360, 0));
};
\end{lstlisting}

### 2015-05-25 main/src/control/controller.cpp
Fix bad_alloc bug.

@@ -605,7 +605,7 @@ void cvworker(std::condition_variable& cv, std::mutex& cvlock, std::size_t myid,
         std::mutex& vectorlock, std::size_t turns, Controller& ctrl)
 {
 	auto core = ctrl.m_cores[myid];
-	constexpr size_t YIELD_ZOMBIE = 10;	// @see Core::m_zombie_rounds
+	constexpr size_t YIELD_ZOMBIE = 25;	// @see Core::m_zombie_rounds
 	auto predicate = [&]()->bool {
 		std::lock_guard<std::mutex> lv(vectorlock);
 		return not flag_is_set(threadsignal[myid], n_threadflags::ISWAITING);
@@ -616,6 +616,7 @@ void cvworker(std::condition_variable& cv, std::mutex& cvlock, std::size_t myid,
 			LOG_INFO("CVWORKER: Thread for core ", core->getCoreID(), " Core is zombie, yielding thread.");
 			std::chrono::milliseconds ms{25};
 			std::this_thread::sleep_for(ms);// Don't kill a core, only yield.
+			//core->setIdle(true);
 		}
 
 		{	/// Intercept a direct order to stop myself.

### 2015-05-25 main/src/model/multicore.cpp

@@ -341,8 +341,8 @@ void n_model::Multicore::revert(const t_timestamp& totime)
 	// Call chain :: singleStep->getMessages->sortIncoming -> receiveMessage() -> revert()
 
 	while (!m_sent_messages.empty()) {		// For each message > totime, send antimessage
-		const auto& msg = m_sent_messages.back();
-		if (msg->getTimeStamp().getTime() >= totime.getTime()) {
+		auto msg = m_sent_messages.back();
+		if (msg->getTimeStamp() >= totime) {
 			m_sent_messages.pop_back();
 			LOG_DEBUG("MCORE:: ", this->getCoreID(), " revert : sent message > time , antimessagging. \n ", msg->toString() );
 			this->sendAntiMessage(msg);

### 2015-05-25 main/src/test/performancetest.cpp

@@ -47,8 +47,8 @@ TEST(Performance, PHOLD)
 
 	ControllerConfig conf;
 	conf.name = "PHOLDBench";
-//	conf.simType = Controller::PDEVS;
-//	conf.coreAmount = 4;
+	//conf.simType = Controller::PDEVS;
+	//conf.coreAmount = 2;
 	conf.saveInterval = 1;
 
 	std::ofstream filestream("testfiles/performance/phold.txt");

### 2015-05-25 manual/projectrapport.tex

@@
 \usepackage{amssymb}
+\usepackage{fancyvrb}
 \usepackage{hyperref}
@@
 
-\chapter{Logger analysis}
+\chapter{Performance analysis}
+To measure the performance of \textsc{DEVS Ex Machina} under more demanding conditions, \textbf{DEVStone} and \textbf{PHOLD} benchmarks were used.
+Every time we compare the performance of our implementation with \textit{PythonPDEVS}.\\
+
+\section{Classic DEVS}
+Using Classic DEVS we see noticeable speedups compared to PyPDEVS. The tests are completed in far less time, using the CPU more effici\"ently.
+
+\subsection{DEVStone}
+\paragraph{\textsc{DEVS Ex Machina}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_devstone
+
+ Performance counter stats for './dxexmachina_devstone':
+
+        314.286326 task-clock (msec)         #    1.125 CPUs utilized          
+               410 context-switches          #    0.001 M/sec                  
+             1,007 cpu-migrations            #    0.003 M/sec                  
+               662 page-faults               #    0.002 M/sec                  
+       802,295,514 cycles                    #    2.553 GHz                     [85.27%]
+       447,423,794 stalled-cycles-frontend   #   55.77% frontend cycles idle    [84.45%]
+       234,246,612 stalled-cycles-backend    #   29.20% backend  cycles idle    [71.04%]
+       739,746,191 instructions              #    0.92  insns per cycle        
+                                             #    0.60  stalled cycles per insn [85.97%]
+       161,577,264 branches                  #  514.108 M/sec                   [85.84%]
+         3,030,149 branch-misses             #    1.88% of all branches         [83.25%]
+
+       0.279274362 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{PythonPDEVS}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat python experiment.py default 2 false > out.txt
+
+ Performance counter stats for 'python experiment.py def 2 false':
+
+        709.776558 task-clock (msec)         #    0.989 CPUs utilized          
+               191 context-switches          #    0.269 K/sec                  
+                34 cpu-migrations            #    0.048 K/sec                  
+             2,942 page-faults               #    0.004 M/sec                  
+     1,687,298,167 cycles                    #    2.377 GHz                     [82.71%]
+       866,759,527 stalled-cycles-frontend   #   51.37% frontend cycles idle    [83.25%]
+       562,247,464 stalled-cycles-backend    #   33.32% backend  cycles idle    [67.62%]
+     1,672,919,036 instructions              #    0.99  insns per cycle        
+                                             #    0.52  stalled cycles per insn [83.82%]
+       375,443,818 branches                  #  528.961 M/sec                   [83.25%]
+        14,867,814 branch-misses             #    3.96% of all branches         [83.28%]
+
+       0.717576473 seconds time elapsed
+\end{Verbatim}
+\subsection{PHOLD}
+\paragraph{\textsc{DEVS Ex Machina}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_phold
+
+ Performance counter stats for './dxexmachina_phold':
+
+         24.249048 task-clock (msec)         #    1.109 CPUs utilized          
+                16 context-switches          #    0.660 K/sec                  
+               117 cpu-migrations            #    0.005 M/sec                  
+               688 page-faults               #    0.028 M/sec                  
+        52,853,150 cycles                    #    2.180 GHz                     [83.16%]
+        29,014,020 stalled-cycles-frontend   #   54.90% frontend cycles idle    [83.16%]
+        14,103,460 stalled-cycles-backend    #   26.68% backend  cycles idle    [65.29%]
+        38,553,838 instructions              #    0.73  insns per cycle        
+                                             #    0.75  stalled cycles per insn [83.31%]
+         8,242,520 branches                  #  339.911 M/sec                   [96.13%]
+           185,357 branch-misses             #    2.25% of all branches         [83.71%]
+
+       0.021871681 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{PythonPDEVS}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat python experiment.py def 2 false > out.txt
+
+ Performance counter stats for 'python experiment.py def 2 false':
+
+        709.776558 task-clock (msec)         #    0.989 CPUs utilized          
+               191 context-switches          #    0.269 K/sec                  
+                34 cpu-migrations            #    0.048 K/sec                  
+             2,942 page-faults               #    0.004 M/sec                  
+     1,687,298,167 cycles                    #    2.377 GHz                     [82.71%]
+       866,759,527 stalled-cycles-frontend   #   51.37% frontend cycles idle    [83.25%]
+       562,247,464 stalled-cycles-backend    #   33.32% backend  cycles idle    [67.62%]
+     1,672,919,036 instructions              #    0.99  insns per cycle        
+                                             #    0.52  stalled cycles per insn [83.82%]
+       375,443,818 branches                  #  528.961 M/sec                   [83.25%]
+        14,867,814 branch-misses             #    3.96% of all branches         [83.28%]
+
+       0.717576473 seconds time elapsed
+\end{Verbatim}
+
+\section{Parallel DEVS}
+Unfortunately, there's quite a big gap in performance between PyPDEVS and DEVS Ex Machina in parallel simulation. The issues causing this are being ironed out.
+
+\subsection{DEVStone}
+\paragraph{\textsc{DEVS Ex Machina}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_devstone
+
+ Performance counter stats for './dxexmachina_devstone':
+
+       2666.499943 task-clock (msec)         #    0.017 CPUs utilized          
+            16,835 context-switches          #    0.006 M/sec                  
+               376 cpu-migrations            #    0.141 K/sec                  
+             1,865 page-faults               #    0.699 K/sec                  
+     5,780,070,450 cycles                    #    2.168 GHz                     [86.46%]
+     2,313,049,054 stalled-cycles-frontend   #   40.02% frontend cycles idle    [84.79%]
+     1,523,702,578 stalled-cycles-backend    #   26.36% backend  cycles idle    [58.20%]
+     6,862,832,656 instructions              #    1.19  insns per cycle        
+                                             #    0.34  stalled cycles per insn [83.73%]
+     1,698,388,723 branches                  #  636.936 M/sec                   [85.64%]
+        25,395,874 branch-misses             #    1.50% of all branches         [85.64%]
+
+     156.945218942 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{PythonPDEVS}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat mpirun -np 2 python experiment.py def 2 false > out.txt
+
+ Performance counter stats for 'mpirun -np 2 python experiment.py def 2 false':
+
+        787.125992 task-clock (msec)         #    0.371 CPUs utilized          
+             1,120 context-switches          #    0.001 M/sec                  
+                91 cpu-migrations            #    0.116 K/sec                  
+             8,684 page-faults               #    0.011 M/sec                  
+     1,786,217,349 cycles                    #    2.269 GHz                     [83.85%]
+       935,480,076 stalled-cycles-frontend   #   52.37% frontend cycles idle    [84.18%]
+       619,366,137 stalled-cycles-backend    #   34.67% backend  cycles idle    [65.26%]
+     1,796,463,414 instructions              #    1.01  insns per cycle        
+                                             #    0.52  stalled cycles per insn [84.16%]
+       403,042,819 branches                  #  512.044 M/sec                   [84.70%]
+        13,210,328 branch-misses             #    3.28% of all branches         [82.68%]
+
+       2.123888919 seconds time elapsed
+\end{Verbatim}
+
+\subsection{PHOLD}
+\paragraph{\textsc{DEVS Ex Machina}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_phold
+
+ Performance counter stats for './dxexmachina_phold':
+
+        820.288154 task-clock (msec)         #    0.003 CPUs utilized          
+            20,030 context-switches          #    0.024 M/sec                  
+               331 cpu-migrations            #    0.404 K/sec                  
+               831 page-faults               #    0.001 M/sec                  
+     1,025,470,316 cycles                    #    1.250 GHz                     [97.69%]
+       767,298,149 stalled-cycles-frontend   #   74.82% frontend cycles idle    [99.29%]
+       536,886,871 stalled-cycles-backend    #   52.36% backend  cycles idle    [52.05%]
+       480,746,733 instructions              #    0.47  insns per cycle        
+                                             #    1.60  stalled cycles per insn [99.54%]
+       118,012,980 branches                  #  143.868 M/sec                   [53.18%]
+         8,481,176 branch-misses             #    7.19% of all branches         [99.62%]
+
+     247.566439823 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{PythonPDEVS}}
+PHOLD in parallel on PyPDEVS crashed immediately with an MPI error, hence a comparison was not possible for now.
+However, given the results of our implementation we can assume that here too, the parallel simulation would be quite a bit slower.
 

### 2015-05-25 main/src/maindevstone.cpp
DEVStone and PHOLD executables are configurable through arguments

@@ -7,11 +7,28 @@
 #include "controllerconfig.h"
 #include "coutredirect.h"
 #include "devstone.h"
+#include "stringtools.h"
 
 LOG_INIT("devstone.log")
 
-int main(int, char**)
+/**
+ * The executable takes up to 2 arguments (in this order):
+ * - The width (amount of processors per coupled model)
+ * - The depth (amount of linked coupled models)
+ */
+int main(int argc, char** args)
 {
+	// default values:
+	std::size_t width = 2;
+	std::size_t depth = 3;
+
+	if(argc >= 2) {
+		width = n_tools::toInt(args[1]);
+	}
+	if(argc >= 3) {
+		depth = n_tools::toInt(args[2]);
+	}
+
 	n_control::ControllerConfig conf;
 	conf.name = "DEVStone";
 	conf.saveInterval = 1;
@@ -24,7 +41,7 @@ int main(int, char**)
 		ctrl->setTerminationTime(endTime);
 
 		// Create a DEVStone simulation with width 2 and depth 3
-		t_coupledmodelptr d = n_tools::createObject< n_devstone::DEVStone>(2, 3, false);
+		t_coupledmodelptr d = n_tools::createObject< n_devstone::DEVStone>(width, depth, false);
 		ctrl->addModel(d);
 
 		ctrl->simulate();

### 2015-05-25 main/src/mainphold.cpp

@@ -8,10 +8,38 @@
 #include "coutredirect.h"
 #include "phold.h"
 
-LOG_INIT("devstone.log")
+LOG_INIT("phold.log")
 
-int main(int, char**)
+
+/**
+ * The executable takes up to 4 arguments (in this order):
+ * - the amount of nodes of the PHOLD
+ * - the amount of atomic models per node
+ * - the amount of iterations
+ * - the @b percentage of remotes (example: if you want 42% remotes, enter 42)
+ */
+int main(int argc, char** args)
 {
+	// default values:
+	std::size_t nodes = 1;
+	std::size_t apn = 10;
+	std::size_t iter = 0;
+	float percentageRemotes = 0.1;
+
+	if(argc >= 2) {
+		nodes = n_tools::toInt(args[1]);
+	}
+	if(argc >= 3) {
+		apn = n_tools::toInt(args[2]);
+	}
+	if(argc >= 4) {
+		iter = n_tools::toInt(args[3]);
+	}
+	if(argc >= 5) {
+		percentageRemotes = n_tools::toInt(args[4]) / 100;
+	}
+
+
 	n_control::ControllerConfig conf;
 	conf.name = "DEVStone";
 	conf.saveInterval = 1;
@@ -24,7 +52,7 @@ int main(int, char**)
 		ctrl->setTerminationTime(endTime);
 
 		// Create a DEVStone simulation with width 2 and depth 3
-		t_coupledmodelptr d = n_tools::createObject< n_benchmarks_phold::PHOLD>(1, 10, 0, 0.1);
+		t_coupledmodelptr d = n_tools::createObject< n_benchmarks_phold::PHOLD>(nodes, apn, iter, percentageRemotes);
 		ctrl->addModel(d);
 
 		ctrl->simulate();

### 2015-05-25 manual/projectrapport.tex
Added ADEVS comparison to performance analysis

@@
 Every time we compare the performance of our implementation with \textit{PythonPDEVS}.\\
+Note: at the time, no major research has gone into the use of several compiler optimizations. Potentially, there is still speed to be gained here.\\
+\\
+{\footnotesize[All the following tests were performed using an \textit{Intel Core i5 M520} CPU at 2.40GHz with 4GB RAM]}
 
 \section{Classic DEVS}
-Using Classic DEVS we see noticeable speedups compared to PyPDEVS. The tests are completed in far less time, using the CPU more effici\"ently.
+Using Classic DEVS we see noticeable speedups compared to PyPDEVS. The tests are completed in far less time, using the CPU more efficiently.
 
 \subsection{DEVStone}
+The DEVStone benchmark used for both implementations has a width of 2 and a depth of 3.
 \paragraph{\textsc{DEVS Ex Machina}}
@@
 
- Performance counter stats for 'python experiment.py def 2 false':
+ Performance counter stats for 'python experiment.py default 2 false':
 
@@
 \begin{Verbatim}[fontsize=\small]
-$ perf stat python experiment.py def 2 false > out.txt
+$ perf stat python experiment.py default 2 false > out.txt
 
- Performance counter stats for 'python experiment.py def 2 false':
+ Performance counter stats for 'python experiment.py default 2 false':
 
@@
 \begin{Verbatim}[fontsize=\small]
-$ perf stat mpirun -np 2 python experiment.py def 2 false > out.txt
+$ perf stat mpirun -np 2 python experiment.py default 2 false > out.txt
 
- Performance counter stats for 'mpirun -np 2 python experiment.py def 2 false':
+ Performance counter stats for 'mpirun -np 2 python experiment.py default 2 false':
 
@@
 However, given the results of our implementation we can assume that here too, the parallel simulation would be quite a bit slower.
+
+\section{Compared to \textsc{ADEVS}}
+Since the DEVStone benchmark for \textsc{ADEVS} does not support different depths, a test with greater width was performed instead.\\
+As we can see, \textsc{ADEVS} is still quite a bit faster, though its CPU usage is less efficient.
+\paragraph{\textsc{DEVS Ex Machina} - Classic DEVS}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_devstone 32 0
+
+ Performance counter stats for './dxexmachina_devstone 32 0':
+
+        836.066726 task-clock (msec)         #    1.113 CPUs utilized          
+             1,241 context-switches          #    0.001 M/sec                  
+             1,027 cpu-migrations            #    0.001 M/sec                  
+               720 page-faults               #    0.861 K/sec                  
+     2,278,424,999 cycles                    #    2.725 GHz                     [84.52%]
+       959,881,005 stalled-cycles-frontend   #   42.13% frontend cycles idle    [84.53%]
+       500,732,097 stalled-cycles-backend    #   21.98% backend  cycles idle    [68.69%]
+     2,775,228,039 instructions              #    1.22  insns per cycle        
+                                             #    0.35  stalled cycles per insn [85.11%]
+       654,396,817 branches                  #  782.709 M/sec                   [84.56%]
+         9,222,451 branch-misses             #    1.41% of all branches         [83.94%]
+
+       0.750984696 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{ADEVS} - Classic DEVS}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./sim 32 0
+
+ Performance counter stats for './sim 32 0':
+
+         15.465233 task-clock (msec)         #    0.976 CPUs utilized          
+                 1 context-switches          #    0.065 K/sec                  
+                 1 cpu-migrations            #    0.065 K/sec                  
+               340 page-faults               #    0.022 M/sec                  
+        22,395,954 cycles                    #    1.448 GHz                     [74.20%]
+         6,618,757 stalled-cycles-frontend   #   29.55% frontend cycles idle    [74.17%]
+         2,895,009 stalled-cycles-backend    #   12.93% backend  cycles idle    [87.88%]
+        31,797,999 instructions              #    1.42  insns per cycle        
+                                             #    0.21  stalled cycles per insn
+         6,645,494 branches                  #  429.705 M/sec                  
+           250,467 branch-misses             #    3.77% of all branches         [74.03%]
+
+       0.015852985 seconds time elapsed
+\end{Verbatim}
 

### 2015-05-25 manual/projectrapport.tex
Updated performance stats for parallel with zombie round fix

--- 2015-05-24 manual/projectrapport.tex
@@
 
-\chapter{Logger analysis}
+\chapter{Performance analysis}
+To measure the performance of \textsc{DEVS Ex Machina} under more demanding conditions, \textbf{DEVStone} and \textbf{PHOLD} benchmarks were used in combination with the \texttt{perf} profiler tool.
+Every time we compare the performance of our implementation with \textit{PythonPDEVS}.\\
+Note: at the moment, no major research has gone into the use of compiler optimizations yet. Potentially, there is still speed to be gained here.\\
+\\
+{\footnotesize[All the following tests were performed using an \textit{Intel Core i5 M520} CPU at 2.40GHz, parallel testing was done using 2 simulation cores.]}
+
+\section{Classic DEVS}
+Using Classic DEVS we see noticeable speedups compared to PyPDEVS. Both tests are completed in far less time, using the CPU more efficiently.
+
+\subsection{DEVStone}
+The DEVStone benchmark used for both implementations has a width of 2 and a depth of 3.
+\paragraph{\textsc{DEVS Ex Machina}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_devstone classic 2 3
+
+ Performance counter stats for './dxexmachina_devstone classic 2 3':
+
+        174.007707 task-clock (msec)         #    1.111 CPUs utilized          
+               184 context-switches          #    0.001 M/sec                  
+               212 cpu-migrations            #    0.001 M/sec                  
+               696 page-faults               #    0.004 M/sec                  
+       463,684,386 cycles                    #    2.665 GHz                     [83.13%]
+       196,709,921 stalled-cycles-frontend   #   42.42% frontend cycles idle    [85.77%]
+       103,770,146 stalled-cycles-backend    #   22.38% backend  cycles idle    [68.62%]
+       557,355,313 instructions              #    1.20  insns per cycle        
+                                             #    0.35  stalled cycles per insn [86.11%]
+       131,411,927 branches                  #  755.208 M/sec                   [85.62%]
+         1,720,031 branch-misses             #    1.31% of all branches         [83.38%]
+
+       0.156664362 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{PythonPDEVS}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat python experiment.py default 2 false > out.txt
+
+ Performance counter stats for 'python experiment.py default 2 false':
+
+        709.776558 task-clock (msec)         #    0.989 CPUs utilized          
+               191 context-switches          #    0.269 K/sec                  
+                34 cpu-migrations            #    0.048 K/sec                  
+             2,942 page-faults               #    0.004 M/sec                  
+     1,687,298,167 cycles                    #    2.377 GHz                     [82.71%]
+       866,759,527 stalled-cycles-frontend   #   51.37% frontend cycles idle    [83.25%]
+       562,247,464 stalled-cycles-backend    #   33.32% backend  cycles idle    [67.62%]
+     1,672,919,036 instructions              #    0.99  insns per cycle        
+                                             #    0.52  stalled cycles per insn [83.82%]
+       375,443,818 branches                  #  528.961 M/sec                   [83.25%]
+        14,867,814 branch-misses             #    3.96% of all branches         [83.28%]
+
+       0.717576473 seconds time elapsed
+\end{Verbatim}
+\subsection{PHOLD}
+\paragraph{\textsc{DEVS Ex Machina}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_phold classic 1 10 0 10
+
+ Performance counter stats for './dxexmachina_phold classic 1 10 0 10':
+
+         15.847971 task-clock (msec)         #    1.047 CPUs utilized          
+                15 context-switches          #    0.946 K/sec                  
+                28 cpu-migrations            #    0.002 M/sec                  
+               710 page-faults               #    0.045 M/sec                  
+        27,365,192 cycles                    #    1.727 GHz                     [74.72%]
+        12,605,461 stalled-cycles-frontend   #   46.06% frontend cycles idle    [73.78%]
+         7,424,569 stalled-cycles-backend    #   27.13% backend  cycles idle    [86.04%]
+        31,095,649 instructions              #    1.14  insns per cycle        
+                                             #    0.41  stalled cycles per insn
+         6,955,945 branches                  #  438.917 M/sec                  
+           132,996 branch-misses             #    1.91% of all branches         [74.99%]
+
+       0.015131930 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{PythonPDEVS}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat python experiment.py default 2 false > out.txt
+
+ Performance counter stats for 'python experiment.py default 2 false':
+
+        709.776558 task-clock (msec)         #    0.989 CPUs utilized          
+               191 context-switches          #    0.269 K/sec                  
+                34 cpu-migrations            #    0.048 K/sec                  
+             2,942 page-faults               #    0.004 M/sec                  
+     1,687,298,167 cycles                    #    2.377 GHz                     [82.71%]
+       866,759,527 stalled-cycles-frontend   #   51.37% frontend cycles idle    [83.25%]
+       562,247,464 stalled-cycles-backend    #   33.32% backend  cycles idle    [67.62%]
+     1,672,919,036 instructions              #    0.99  insns per cycle        
+                                             #    0.52  stalled cycles per insn [83.82%]
+       375,443,818 branches                  #  528.961 M/sec                   [83.25%]
+        14,867,814 branch-misses             #    3.96% of all branches         [83.28%]
+
+       0.717576473 seconds time elapsed
+\end{Verbatim}
+
+\section{Parallel DEVS}
+Also in parallel simulations \textsc{DEVS Ex Machina} completes quite a bit quicker, again achieving higher CPU utilization.
+Additionally we can see that Conservative PDEVS performs consistently significantly faster than Parallel PDEVS, in the case of PHOLD.
+
+\subsection{DEVStone}
+\paragraph{\textsc{DEVS Ex Machina} - Optimistic}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_devstone opdevs 2 2 3
+
+ Performance counter stats for './dxexmachina_devstone opdevs 2 2 3':
+
+         95.265846 task-clock (msec)         #    0.499 CPUs utilized          
+               191 context-switches          #    0.002 M/sec                  
+                15 cpu-migrations            #    0.157 K/sec                  
+             1,229 page-faults               #    0.013 M/sec                  
+       197,702,110 cycles                    #    2.075 GHz                     [82.53%]
+        95,414,988 stalled-cycles-frontend   #   48.26% frontend cycles idle    [83.54%]
+        47,283,951 stalled-cycles-backend    #   23.92% backend  cycles idle    [76.71%]
+       199,327,993 instructions              #    1.01  insns per cycle        
+                                             #    0.48  stalled cycles per insn [91.61%]
+        47,887,986 branches                  #  502.677 M/sec                   [86.14%]
+           931,688 branch-misses             #    1.95% of all branches         [75.36%]
+
+       0.191066656 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{DEVS Ex Machina} - Conservative}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_devstone cpdevs 2 2 3
+
+ Performance counter stats for './dxexmachina_devstone cpdevs 2 2 3':
+
+        129.292299 task-clock (msec)         #    0.555 CPUs utilized          
+               672 context-switches          #    0.005 M/sec                  
+                15 cpu-migrations            #    0.116 K/sec                  
+             2,177 page-faults               #    0.017 M/sec                  
+       308,983,665 cycles                    #    2.390 GHz                     [74.75%]
+       151,616,998 stalled-cycles-frontend   #   49.07% frontend cycles idle    [84.43%]
+        82,891,133 stalled-cycles-backend    #   26.83% backend  cycles idle    [76.81%]
+       276,296,034 instructions              #    0.89  insns per cycle        
+                                             #    0.55  stalled cycles per insn [88.68%]
+        66,513,459 branches                  #  514.443 M/sec                   [87.86%]
+         1,390,075 branch-misses             #    2.09% of all branches         [77.49%]
+
+       0.233046274 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{PythonPDEVS}}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat mpirun -np 2 python experiment.py default 2 false > out.txt
+
+ Performance counter stats for 'mpirun -np 2 python experiment.py default 2 false':
+
+        787.125992 task-clock (msec)         #    0.371 CPUs utilized          
+             1,120 context-switches          #    0.001 M/sec                  
+                91 cpu-migrations            #    0.116 K/sec                  
+             8,684 page-faults               #    0.011 M/sec                  
+     1,786,217,349 cycles                    #    2.269 GHz                     [83.85%]
+       935,480,076 stalled-cycles-frontend   #   52.37% frontend cycles idle    [84.18%]
+       619,366,137 stalled-cycles-backend    #   34.67% backend  cycles idle    [65.26%]
+     1,796,463,414 instructions              #    1.01  insns per cycle        
+                                             #    0.52  stalled cycles per insn [84.16%]
+       403,042,819 branches                  #  512.044 M/sec                   [84.70%]
+        13,210,328 branch-misses             #    3.28% of all branches         [82.68%]
+
+       2.123888919 seconds time elapsed
+\end{Verbatim}
+
+\subsection{PHOLD}
+\paragraph{\textsc{DEVS Ex Machina} - Optimistic}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_phold opdevs 2 1 10 0 10
+
+ Performance counter stats for './dxexmachina_phold opdevs 2 1 10 0 10':
+
+         20.741980 task-clock (msec)         #    0.092 CPUs utilized          
+                73 context-switches          #    0.004 M/sec                  
+                 8 cpu-migrations            #    0.386 K/sec                  
+               760 page-faults               #    0.037 M/sec                  
+        28,281,718 cycles                    #    1.364 GHz                     [42.03%]
+        14,899,039 stalled-cycles-frontend   #   52.68% frontend cycles idle    [79.62%]
+         8,373,668 stalled-cycles-backend    #   29.61% backend  cycles idle   
+        29,601,476 instructions              #    1.05  insns per cycle        
+                                             #    0.50  stalled cycles per insn
+         6,754,172 branches                  #  325.628 M/sec                  
+           135,220 branch-misses             #    2.00% of all branches         [81.22%]
+
+       0.225834153 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{DEVS Ex Machina} - Conservative}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_phold cpdevs 2 1 10 0 10
+
+ Performance counter stats for './dxexmachina_phold cpdevs 2 1 10 0 10':
+
+         28.509489 task-clock (msec)         #    0.286 CPUs utilized          
+                95 context-switches          #    0.003 M/sec                  
+                11 cpu-migrations            #    0.386 K/sec                  
+               813 page-faults               #    0.029 M/sec                  
+        54,757,291 cycles                    #    1.921 GHz                     [63.45%]
+        29,057,227 stalled-cycles-frontend   #   53.07% frontend cycles idle    [86.87%]
+        15,602,463 stalled-cycles-backend    #   28.49% backend  cycles idle    [90.08%]
+        53,326,834 instructions              #    0.97  insns per cycle        
+                                             #    0.54  stalled cycles per insn
+        12,449,801 branches                  #  436.690 M/sec                  
+           219,155 branch-misses             #    1.76% of all branches         [64.68%]
+
+       0.099591165 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{PythonPDEVS}}
+PHOLD in parallel on PyPDEVS crashed immediately with an MPI error, hence a comparison was not possible for now.
+
+\section{Compared to \textsc{ADEVS}}
+Since the DEVStone benchmark for \textsc{ADEVS} does not support different depths, a test with greater width was performed instead. As we can see, \textsc{ADEVS} is still quite a bit faster.
+\paragraph{\textsc{DEVS Ex Machina} - Classic DEVS}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./dxexmachina_devstone classic 32 0
+
+ Performance counter stats for './dxexmachina_devstone classic 32 0':
+
+        800.259000 task-clock (msec)         #    1.081 CPUs utilized          
+             2,178 context-switches          #    0.003 M/sec                  
+               217 cpu-migrations            #    0.271 K/sec                  
+               767 page-faults               #    0.958 K/sec                  
+     2,199,059,849 cycles                    #    2.748 GHz                     [83.12%]
+       862,325,341 stalled-cycles-frontend   #   39.21% frontend cycles idle    [83.36%]
+       432,658,584 stalled-cycles-backend    #   19.67% backend  cycles idle    [67.64%]
+     2,794,549,736 instructions              #    1.27  insns per cycle        
+                                             #    0.31  stalled cycles per insn [84.75%]
+       661,425,743 branches                  #  826.515 M/sec                   [84.04%]
+         9,605,392 branch-misses             #    1.45% of all branches         [82.83%]
+
+       0.739954252 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{ADEVS} - Classic DEVS}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat ./sim 32 0
+
+ Performance counter stats for './sim 32 0':
+
+         15.465233 task-clock (msec)         #    0.976 CPUs utilized          
+                 1 context-switches          #    0.065 K/sec                  
+                 1 cpu-migrations            #    0.065 K/sec                  
+               340 page-faults               #    0.022 M/sec                  
+        22,395,954 cycles                    #    1.448 GHz                     [74.20%]
+         6,618,757 stalled-cycles-frontend   #   29.55% frontend cycles idle    [74.17%]
+         2,895,009 stalled-cycles-backend    #   12.93% backend  cycles idle    [87.88%]
+        31,797,999 instructions              #    1.42  insns per cycle        
+                                             #    0.21  stalled cycles per insn
+         6,645,494 branches                  #  429.705 M/sec                  
+           250,467 branch-misses             #    3.77% of all branches         [74.03%]
+
+       0.015852985 seconds time elapsed
+\end{Verbatim}
 

### 2015-05-25 manual/projectrapport.tex
Added PyPy comparison to performance analysis

@@
 
+\section{Compared to PyPDEVS on PyPy}
+Surprisingly, both DEVStone and PHOLD perform significantly worse using \textit{PyPy} instead of \textit{CPython}, despite the sometimes dramatically improved CPU utilization.
+Perhaps these two test cases just happen to be "JIT-unfriendly".\\
+\textit{PyPy} is known to sometimes perform worse when profiling but informally its speed didn't seem any different when running the benchmarks without using \texttt{perf}.
+
+\subsection{DEVStone}
+\paragraph{\textsc{PythonPDEVS} - Classic DEVS on PyPy}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat pypy experiment.py default 3 false > out.txt
+
+ Performance counter stats for 'pypy experiment.py default 3 false':
+
+       2367.165094 task-clock (msec)         #    0.998 CPUs utilized          
+               276 context-switches          #    0.117 K/sec                  
+                20 cpu-migrations            #    0.008 K/sec                  
+            11,357 page-faults               #    0.005 M/sec                  
+     6,765,226,427 cycles                    #    2.858 GHz                     [83.37%]
+     4,234,478,457 stalled-cycles-frontend   #   62.59% frontend cycles idle    [83.27%]
+     3,145,973,412 stalled-cycles-backend    #   46.50% backend  cycles idle    [66.67%]
+     5,014,442,675 instructions              #    0.74  insns per cycle        
+                                             #    0.84  stalled cycles per insn [83.49%]
+     1,077,122,966 branches                  #  455.027 M/sec                   [83.44%]
+        38,104,090 branch-misses             #    3.54% of all branches         [83.27%]
+
+       2.372239526 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{PythonPDEVS} - Parallel DEVS on PyPy}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat mpirun -np 2 pypy experiment.py default 3 false > out.txt
+
+ Performance counter stats for 'mpirun -np 2 pypy experiment.py default 3 false':
+
+       5789.446884 task-clock (msec)         #    1.991 CPUs utilized          
+               624 context-switches          #    0.108 K/sec                  
+                40 cpu-migrations            #    0.007 K/sec                  
+            23,245 page-faults               #    0.004 M/sec                  
+    15,328,712,962 cycles                    #    2.648 GHz                     [83.32%]
+    10,258,365,662 stalled-cycles-frontend   #   66.92% frontend cycles idle    [83.32%]
+     7,799,638,758 stalled-cycles-backend    #   50.88% backend  cycles idle    [66.54%]
+    10,156,225,688 instructions              #    0.66  insns per cycle        
+                                             #    1.01  stalled cycles per insn [83.23%]
+     2,163,097,214 branches                  #  373.628 M/sec                   [83.47%]
+        76,428,341 branch-misses             #    3.53% of all branches         [83.42%]
+
+       2.907224325 seconds time elapsed
+\end{Verbatim}
+\subsection{PHOLD}
+\paragraph{\textsc{PythonPDEVS} - Classic DEVS on PyPy}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat pypy experiment.py > out.txt
+
+ Performance counter stats for 'pypy experiment.py':
+
+       1827.849910 task-clock (msec)         #    0.999 CPUs utilized          
+               220 context-switches          #    0.120 K/sec                  
+                10 cpu-migrations            #    0.005 K/sec                  
+            14,385 page-faults               #    0.008 M/sec                  
+     5,224,512,160 cycles                    #    2.858 GHz                     [83.37%]
+     3,283,008,184 stalled-cycles-frontend   #   62.84% frontend cycles idle    [83.37%]
+     2,563,366,886 stalled-cycles-backend    #   49.06% backend  cycles idle    [66.80%]
+     4,079,915,131 instructions              #    0.78  insns per cycle        
+                                             #    0.80  stalled cycles per insn [83.42%]
+       758,611,592 branches                  #  415.029 M/sec                   [83.37%]
+        28,841,978 branch-misses             #    3.80% of all branches         [83.16%]
+
+       1.830573045 seconds time elapsed
+\end{Verbatim}
+\paragraph{\textsc{PythonPDEVS} - Parallel DEVS on PyPy}
+\begin{Verbatim}[fontsize=\small]
+$ perf stat mpirun -np 2 pypy experiment.py > out.txt
+
+ Performance counter stats for 'mpirun -np 2 pypy experiment.py':
+
+       5410.417506 task-clock (msec)         #    1.992 CPUs utilized          
+               785 context-switches          #    0.145 K/sec                  
+                33 cpu-migrations            #    0.006 K/sec                  
+            28,874 page-faults               #    0.005 M/sec                  
+    15,456,764,898 cycles                    #    2.857 GHz                     [83.35%]
+    12,040,987,764 stalled-cycles-frontend   #   77.90% frontend cycles idle    [83.35%]
+     6,550,209,612 stalled-cycles-backend    #   42.38% backend  cycles idle    [66.70%]
+     7,709,537,731 instructions              #    0.50  insns per cycle        
+                                             #    1.56  stalled cycles per insn [83.41%]
+     1,430,492,329 branches                  #  264.396 M/sec                   [83.42%]
+        63,358,010 branch-misses             #    4.43% of all branches         [83.21%]
+
+       2.716329957 seconds time elapsed
+\end{Verbatim}
+
 \section{Logger}

### 2015-05-26 main/src/control/simpleallocator.h
Updated SimpleAllocator to allow for loaded simulations to have a different amount of cores

@@ -13,17 +13,18 @@
 namespace n_control {
 
 /**
- * @brief Simple, dumb allocator that tries to spread models evenly, but gets overruled by the models' configuration
+ * @brief Simple, dumb allocator that tries to spread models evenly, but gets overruled by the models' configuration by default
  */
 class SimpleAllocator: public Allocator
 {
 private:
 	size_t m_i;
 	size_t m_cores;
+	bool m_allowUserOverride;
 
 public:
-	SimpleAllocator(size_t c)
-		: m_i(0), m_cores(c)
+	SimpleAllocator(size_t c, bool allowOverride = true)
+		: m_i(0), m_cores(c), m_allowUserOverride(allowOverride)
 	{
 	}
 
@@ -38,13 +39,15 @@ public:
 	size_t allocate(n_model::t_atomicmodelptr& m)
 	{
 		int corenum = m->getCorenumber();
-		if(corenum == -1) {	// The user didn't specify a particular core, so we pick one ourselves
-			int i = m_i;
-			m_i = (m_i + 1) % m_cores;
-			m->setCorenumber(i); // We let the model know which core we decided to place it on
-			return i;
+		if(!m_allowUserOverride || corenum == -1) {	// If user override is not allowed, or if it is but the
+			corenum = m_i;				//  user didn't specify a particular core, we pick a
+			m_i = (m_i + 1) % m_cores;		//  destination ourselves
+		} else {
+			corenum %= m_cores;	// Make absolutely sure the model always ends up on a real core
+						//  e.g. if the user lowered the amount of cores on a loaded simulation
 		}
-		else return corenum;	// The user already specified a core for this model, so we better pick that one!
+		m->setCorenumber(corenum);
+		return corenum;
 	}
 };
 

### 2015-05-26 manual/tex/controller.tex
Updated manual (Controller), added example simulations

@@
 	\end{tabular} \newline
+	\textit{[This option is disregarded if you do not run PDEVS]} \newline
+	The default is \emph{OPTIMISTIC}
 
-	The default is \emph{OPTIMISTIC} \newline
-	\textit{[This option is disregarded if you do not run PDEVS]}
 \\\\
@@
 \textit{saveInterval} & This value decides how often the simulation is saved and the tracing output dumped. A value of 3 would mean that this happens every 3 simulation cycles. \newline
+\textit{[This option is disregarded if you do not run CLASSIC or DSDEVS]} \newline
 						The default value is 5
@@
 \subsubsection{Loading a previous simulation}
-Loading a previous simulation from a savefile follows the same procedure as setting up a regular simulation for the most part.
-First the simulation is configured (optionally by using the ControllerConfig object as described above), then instead of loading a new model, you can use the Controller's \textsl{load} method to load the data of your previous simulation.\\
-\emph{Important: To load a simulation \textbf{do not} have to use the same configuration!}
+Loading a previous simulation from a savefile follows the same procedure as setting up a regular simulation for the most part.\\
+First the simulation is configured (optionally by using the ControllerConfig object as described above) and then, instead of loading a new model, you can use the Controller's \textsl{load} method to load the data of your previous simulation.\\
+\\
+\textit{Important: to load a simulation you \textbf{do not} have to use the same configuration!}
+\\
+If you wish to increase the amount of cores of a loaded simulation, it is recommended to use an allocator which overrides the model's preferred location, since those will still be set to the cores of the previous simulation.
+The included \textsl{SimpleAllocator} can be used by setting its \texttt{allowUserOverride} flag to \texttt{false} in the constructor.
+Reducing the amount of cores is handled automatically.
 

### 2015-05-26 manual/tex/example.tex
Updated manual (Controller), added example simulations

@@
-Extensive examples are provided in our Controller testing classes. For each formalism a model was created which the user can use as a reference implementation.\\
+The following are a few quick examples to illustrate the setup of a regular simulation.\\
+Additionally, many working examples are provided in our Controller testing classes (\texttt{controllertest.cpp}).
+For each formalism a model was created which the user can use as a reference implementation.\\
 
-\subsection{example}
+\section{Classic DEVS}
+The following example describes the manual setup of a crowd simulation in a zoo:
+\begin{Verbatim}[fontsize=\small]
+// Setting up all prerequisites
+auto tracers = createObject<n_tracers::t_tracerset>();
+std::unordered_map<std::size_t, t_coreptr> coreMap;
+std::shared_ptr<Allocator> allocator = createObject<SimpleAllocator>(1);
+std::shared_ptr<n_control::LocationTable> locTab = createObject<n_control::LocationTable>(1);
 
-\begin{lstlisting}
-std::ofstream filestream("controller/devstestcoupled.txt");
-{
-	CoutRedirect myRedirect(filestream);
-	auto tracers = createObject<n_tracers::t_tracerset>();
-
-	std::unordered_map<std::size_t, t_coreptr> coreMap;
-	std::shared_ptr<Allocator> allocator = 
-		createObject<SimpleAllocator>(1);
-	std::shared_ptr<n_control::LocationTable> locTab = 
-		createObject<n_control::LocationTable>(1);
-
+// Setting up the core
 t_coreptr c = createObject<Core>();
@@
 
-	Controller ctrl("testController", coreMap, allocator, 
-		locTab, tracers);
+// Setup of the controller
+Controller ctrl("zooSim", coreMap, allocator, locTab, tracers);
 ctrl.setClassicDEVS();
@@
 
-	t_coupledmodelptr m1 = createObject<n_examples_coupled::
-		TrafficSystem>("trafficSystem");
+// Creating and adding the main model
+t_coupledmodelptr m1 = createObject<n_crowdSim::Zoo>("Planckendael");
 ctrl.addModel(m1);
 
+// Starting the simulation
 ctrl.simulate();
-	EXPECT_TRUE(c->isLive() == false);
-	EXPECT_TRUE(c->getTime() >= t_timestamp(360, 0));
-};
-\end{lstlisting}
+\end{Verbatim}
+
+In cases where you don't need control over all individual components, it's even quicker to opt for using \textsl{ControllerConfig} to set up everything:
+
+\begin{Verbatim}[fontsize=\small]
+// Creating a ControllerConfig object, setting all non-default parameters
+ControllerConfig conf;
+conf.name = "zooSim";
+
+auto ctrl = conf.createController();
+ctrl->setTerminationTime(t_timestamp(360, 0));
+
+t_coupledmodelptr m1 = createObject<n_crowdSim::Zoo>("Planckendael");
+ctrl->addModel(m1);
+
+ctrl->simulate();
+\end{Verbatim}
+
+\section{Parallel DEVS (Optimistic)}
+The simulation of the air traffic on an airport:
+\begin{Verbatim}[fontsize=\small]
+\\ Setup of the prerequisites
+auto tracers = createObject<n_tracers::t_tracerset>();
+t_networkptr network = createObject<Network>(2);
+std::unordered_map<std::size_t, t_coreptr> coreMap;
+std::shared_ptr<Allocator> allocator = createObject<SimpleAllocator>(2);
+std::shared_ptr<n_control::LocationTable> locTab = createObject<n_control::LocationTable>(2);
+
+\\ Setup of the cores
+t_coreptr c1 = createObject<Multicore>(network, 0, locTab, 2);
+t_coreptr c2 = createObject<Multicore>(network, 1, locTab, 2);
+coreMap[0] = c1;
+coreMap[1] = c2;
+
+\\ Setup of the controller
+Controller ctrl("AirControl", coreMap, allocator, locTab, tracers);
+ctrl.setPDEVS();
+ctrl.setTerminationTime(t_timestamp(720, 0));
+
+\\ Creating and adding the main model
+t_coupledmodelptr m = createObject<n_airControl::Airport>("JFK");
+ctrl.addModel(m);
+
+ctrl.simulate();
+\end{Verbatim}
+
+Here too, the setup is shorter and easier using \textsl{ControllerConfig}:
+
+\begin{Verbatim}[fontsize=\small]
+ControllerConfig conf;
+conf.name = "AirControl";
+conf.simType = Controller::PDEVS;
+conf.coreAmount = 2;
+
+auto ctrl = conf.createController();
+ctrl->setTerminationTime(t_timestamp(720, 0));
+
+t_coupledmodelptr m1 = createObject<n_airControl::Airport>("JFK");
+ctrl->addModel(m1);
+ctrl->simulate();
+\end{Verbatim}

### 2015-06-14 manual/projectrapport.tex

@@
 
+\section{Jenkins}
+While building the traffic system on the Jenkins server problems occured and the server crashed due to a memory shortage. A lower level of optimization solved this problem.
+
 \chapter{Performance analysis}
@@
 
+\chapter{Examples}
+As it was part of the assignment we've implemented a set of examples for our simulator. Except from the virus-model all examples are based on models described in Yentl's thesis.
+
+\subsubsection{Traffic light}
+This simple model is an exact copy of the model used as an example in pypdevs.
+
+\subsubsection{DEVStone}
+...\\
+\\
+\textbf{NOTE} We added an extra executable to the project containing only the necessary implementation for the DEVStone model to work.
+
+\subsubsection{Forrestfire}
+...\\
+\\
+\textbf{NOTE} We added an extra executable to the project containing only the necessary implementation for the traffic system model to work.
+
+\subsubsection{Traffic system}
+The traffic system is based on the traffic system described in Yentl's paper. We've implemented two versions, a bigger one and a smaller one.\\
+A modified version of Yenl's script generates the header and source file for the city model and makes it possible to easily adopt the models size.
+The script can be found in the './main/script' folder of our repository.\\
+\\
+Run the following command to generate a city:
+\begin{Verbatim}[fontsize=\small]
+$ python script_devsexmachina.py {x} {y} {nodes} {districts}
+\end{Verbatim}
+The small city is 10 x 50 segments large and the larger city exists of 10 x 500 segments. Both cities are generated for 50 nodes and 5 districts as described in Yentl's thesis.\\
+\\
+\textbf{NOTE} We added an extra executable to the project containing only the necessary implementation for the traffic system model to work.
+This executable contains the larger version of the city.
+To build large cities you have to ensure that there is a sufficient amount of memory available.
+
+\subsubsection{Virus}
+...\\
+\\
+\textbf{NOTE} We added an extra executable to the project containing only the necessary implementation for the traffic system model to work.
+
 \begin{thebibliography}{1}

### 2015-06-14 manual/projectrapport.tex
Update project report (benchmarks)

--- 2015-05-25 manual/projectrapport.tex
@@
 To measure the performance of \textsc{DEVS Ex Machina} under more demanding conditions, \textbf{DEVStone} and \textbf{PHOLD} benchmarks were used in combination with the \texttt{perf} profiler tool.
-Every time we compare the performance of our implementation with \textit{PythonPDEVS}.\\
+As these are standardized tests, they allow easy comparison between our implementation and \textit{PythonPDEVS}.\\
 Note: at the moment, no major research has gone into the use of compiler optimizations yet. Potentially, there is still speed to be gained here.\\
 \\
-{\footnotesize[All the following tests were performed using an \textit{Intel Core i5 M520} CPU at 2.40GHz, parallel testing was done using 2 simulation cores.]}
+{\footnotesize[All the following tests were performed using an \textit{Intel Core i5 M520} CPU at 2.40GHz, parallel testing was done using 2 simulation cores.
+All tests were ran 5 times to give an indication of the predictability of the result.]}
 
-\section{Classic DEVS}
-Using Classic DEVS we see noticeable speedups compared to PyPDEVS. Both tests are completed in far less time, using the CPU more efficiently.
+\pagebreak
+
+\section{DEVStone}
+\begin{figure}
+\centering
+\begin{minipage}{.5\textwidth}
+  \centering
+  \hspace*{-1.5in}
+  \includegraphics[width=1.6\linewidth]{devstone_classic.png}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1.6\linewidth]{devstone_parallel.png}
+\end{minipage}
+\end{figure}
+
+In both Classic and Parallel DEVS there are noticeable speedups compared to \textit{PyPDEVS}.\\
+Looking at \textsc{DEVS Ex Machina} specifically we can see that in the first few, rather lightweight cases the single core Classic DEVS simulation is faster than PDEVS, in the first case even twice as fast.
+Once more models start to get involved however, things quickly turn around since the benefit of having the raw processing power of more cores available outweighs the cost of setting up and managing multiple threads.\\
+Additionally, the final and most demanding test case, a DEVStone benchmark of depth and width 12, was also executed with 4 cores, which showed a slight improvement over using 2 cores.
+\\
+
+However, it should be noted that in PDEVS a somewhat thought-out assignment of models to cores is of critical importance.
+A parallel simulation with models placed haphazardly can perform \emph{significantly} worse than a Classic DEVS which doesn't have to deal with placement.
+In general, it is best to keep the most used connections intra-core.
+
+\pagebreak
+
+\section{PHOLD}
+\begin{figure}
+\centering
+\begin{minipage}{.5\textwidth}
+  \centering
+  \hspace*{-1.5in}
+  \includegraphics[width=1.6\linewidth]{phold_classic.png}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1.6\linewidth]{phold_parallel.png}
+\end{minipage}
+\end{figure}
+
+{\footnotesize[ To make a useful comparison with \textit{Python PDEVS} possible, care was taken to make sure the "meaningless" computations related to the payload size were not removed by compiler optimisations ]}\\
+\\
+This benchmark tests the effect of different payload sizes in a highly interconnected environment.\\
+In both environments \textsc{DEVS Ex Machina} is significantly faster than \textit{PyPDEVS}.
+Especially interesting is the second graph with parellel results.
+Whereas in DEVStone Optimistic and Conservative DEVS performed relatively equally, here we can spot a clear discrepancy between them.\\
+
+In Optimistic we can see that as the payload grows, the total runtime becomes less predictable.
+Because of the inherent interconnectivity of the benchmark, and because processing a message with a larger payload can take a while, the chance of reverts or even series of reverts is increased.
+
+In Conservative simulation however, the flow time is tightly controlled, avoiding reverts. In this case this behavior leads to significantly better results.
+
+\pagebreak
+
+\section{Detailed stats}
+Some more detailed statistics provided by \texttt{perf} are shown here. Each time, only the result of a single run is provided, to give a general indication.
 
 \subsection{DEVStone}
-The DEVStone benchmark used for both implementations has a width of 2 and a depth of 3.
 \paragraph{\textsc{DEVS Ex Machina}}
@@
 $ perf stat ./dxexmachina_devstone classic 2 3
-
  Performance counter stats for './dxexmachina_devstone classic 2 3':
 
-        174.007707 task-clock (msec)         #    1.111 CPUs utilized          
-               184 context-switches          #    0.001 M/sec                  
+        189.176880 task-clock (msec)         #    1.115 CPUs utilized          
+               243 context-switches          #    0.001 M/sec                  
                212 cpu-migrations            #    0.001 M/sec                  
-               696 page-faults               #    0.004 M/sec                  
-       463,684,386 cycles                    #    2.665 GHz                     [83.13%]
-       196,709,921 stalled-cycles-frontend   #   42.42% frontend cycles idle    [85.77%]
-       103,770,146 stalled-cycles-backend    #   22.38% backend  cycles idle    [68.62%]
-       557,355,313 instructions              #    1.20  insns per cycle        
-                                             #    0.35  stalled cycles per insn [86.11%]
-       131,411,927 branches                  #  755.208 M/sec                   [85.62%]
-         1,720,031 branch-misses             #    1.31% of all branches         [83.38%]
+               691 page-faults               #    0.004 M/sec                  
+       476,733,288 cycles                    #    2.520 GHz                     [84.35%]
+       204,107,284 stalled-cycles-frontend   #   42.81% frontend cycles idle    [84.40%]
+        95,948,277 stalled-cycles-backend    #   20.13% backend  cycles idle    [67.48%]
+       561,945,984 instructions              #    1.18  insns per cycle        
+                                             #    0.36  stalled cycles per insn [85.07%]
+       132,785,648 branches                  #  701.913 M/sec                   [84.93%]
+         1,847,286 branch-misses             #    1.39% of all branches         [84.62%]
 
-       0.156664362 seconds time elapsed
+       0.169685736 seconds time elapsed
+
+$ perf stat ./dxexmachina_devstone opdevs 2 2 3
+ Performance counter stats for './dxexmachina_devstone opdevs 2 2 3':
+
+        258.435415 task-clock (msec)         #    0.924 CPUs utilized          
+             1,201 context-switches          #    0.005 M/sec                  
+                26 cpu-migrations            #    0.101 K/sec                  
+             3,836 page-faults               #    0.015 M/sec                  
+       613,419,141 cycles                    #    2.374 GHz                     [83.93%]
+       316,632,123 stalled-cycles-frontend   #   51.62% frontend cycles idle    [85.01%]
+       139,998,026 stalled-cycles-backend    #   22.82% backend  cycles idle    [67.52%]
+       615,911,601 instructions              #    1.00  insns per cycle        
+                                             #    0.51  stalled cycles per insn [84.95%]
+       144,090,841 branches                  #  557.551 M/sec                   [85.11%]
+         2,820,791 branch-misses             #    1.96% of all branches         [78.96%]
+
+       0.279720722 seconds time elapsed
+
+$ perf stat ./dxexmachina_devstone cpdevs 2 2 3
+ Performance counter stats for './dxexmachina_devstone cpdevs 2 2 3':
+
+        257.663639 task-clock (msec)         #    0.851 CPUs utilized          
+               911 context-switches          #    0.004 M/sec                  
+                19 cpu-migrations            #    0.074 K/sec                  
+             3,837 page-faults               #    0.015 M/sec                  
+       663,670,571 cycles                    #    2.576 GHz                     [82.64%]
+       351,293,037 stalled-cycles-frontend   #   52.93% frontend cycles idle    [82.43%]
+       197,270,653 stalled-cycles-backend    #   29.72% backend  cycles idle    [65.25%]
+       643,477,063 instructions              #    0.97  insns per cycle        
+                                             #    0.55  stalled cycles per insn [84.74%]
+       150,533,837 branches                  #  584.226 M/sec                   [87.77%]
+         2,968,904 branch-misses             #    1.97% of all branches         [82.67%]
+
+       0.302693494 seconds time elapsed
 \end{Verbatim}
@@
 \begin{Verbatim}[fontsize=\small]
-$ perf stat python experiment.py default 2 false > out.txt
+$ perf stat ./dxexmachina_devstone classic 2 3
+ Performance counter stats for 'python ./DEVStone/experiment.py default 2 false':
 
- Performance counter stats for 'python experiment.py default 2 false':
+        670.501187 task-clock (msec)         #    0.959 CPUs utilized          
+               169 context-switches          #    0.252 K/sec                  
+                22 cpu-migrations            #    0.033 K/sec                  
+             4,962 page-faults               #    0.007 M/sec                  
+     1,828,074,336 cycles                    #    2.726 GHz                     [82.96%]
+       983,729,421 stalled-cycles-frontend   #   53.81% frontend cycles idle    [83.51%]
+       600,902,567 stalled-cycles-backend    #   32.87% backend  cycles idle    [67.59%]
+     1,691,525,594 instructions              #    0.93  insns per cycle        
+                                             #    0.58  stalled cycles per insn [84.05%]
+       380,934,259 branches                  #  568.134 M/sec                   [83.42%]
+        15,797,482 branch-misses             #    4.15% of all branches         [82.57%]
 
-        709.776558 task-clock (msec)         #    0.989 CPUs utilized          
-               191 context-switches          #    0.269 K/sec                  
-                34 cpu-migrations            #    0.048 K/sec                  
-             2,942 page-faults               #    0.004 M/sec                  
-     1,687,298,167 cycles                    #    2.377 GHz                     [82.71%]
-       866,759,527 stalled-cycles-frontend   #   51.37% frontend cycles idle    [83.25%]
-       562,247,464 stalled-cycles-backend    #   33.32% backend  cycles idle    [67.62%]
-     1,672,919,036 instructions              #    0.99  insns per cycle        
-                                             #    0.52  stalled cycles per insn [83.82%]
-       375,443,818 branches                  #  528.961 M/sec                   [83.25%]
-        14,867,814 branch-misses             #    3.96% of all branches         [83.28%]
+       0.699383675 seconds time elapsed
 
-       0.717576473 seconds time elapsed
+$ perf stat ./dxexmachina_devstone opdevs 2 2 3
+ Performance counter stats for 'mpirun -np 2 python ./DEVStone/experiment.py default 2 false':
+
+        692.210888 task-clock (msec)         #    0.330 CPUs utilized          
+               909 context-switches          #    0.001 M/sec                  
+                99 cpu-migrations            #    0.143 K/sec                  
+             8,728 page-faults               #    0.013 M/sec                  
+     1,801,930,537 cycles                    #    2.603 GHz                     [82.89%]
+       929,771,707 stalled-cycles-frontend   #   51.60% frontend cycles idle    [82.88%]
+       581,151,775 stalled-cycles-backend    #   32.25% backend  cycles idle    [68.37%]
+     1,800,262,136 instructions              #    1.00  insns per cycle        
+                                             #    0.52  stalled cycles per insn [84.52%]
+       403,185,223 branches                  #  582.460 M/sec                   [84.04%]
+        13,159,671 branch-misses             #    3.26% of all branches         [82.92%]
+
+       2.099629915 seconds time elapsed
 \end{Verbatim}
@@
 $ perf stat ./dxexmachina_phold classic 1 10 0 10
-
  Performance counter stats for './dxexmachina_phold classic 1 10 0 10':
 
-         15.847971 task-clock (msec)         #    1.047 CPUs utilized          
-                15 context-switches          #    0.946 K/sec                  
-                28 cpu-migrations            #    0.002 M/sec                  
-               710 page-faults               #    0.045 M/sec                  
-        27,365,192 cycles                    #    1.727 GHz                     [74.72%]
-        12,605,461 stalled-cycles-frontend   #   46.06% frontend cycles idle    [73.78%]
-         7,424,569 stalled-cycles-backend    #   27.13% backend  cycles idle    [86.04%]
-        31,095,649 instructions              #    1.14  insns per cycle        
-                                             #    0.41  stalled cycles per insn
-         6,955,945 branches                  #  438.917 M/sec                  
-           132,996 branch-misses             #    1.91% of all branches         [74.99%]
+         12.815382 task-clock (msec)         #    1.081 CPUs utilized          
+                11 context-switches          #    0.858 K/sec                  
+                25 cpu-migrations            #    0.002 M/sec                  
+               712 page-faults               #    0.056 M/sec                  
+        27,657,920 cycles                    #    2.158 GHz                     [64.22%]
+        13,070,687 stalled-cycles-frontend   #   47.26% frontend cycles idle    [75.41%]
+         8,563,853 stalled-cycles-backend    #   30.96% backend  cycles idle   
+        31,010,010 instructions              #    1.12  insns per cycle        
+                                             #    0.42  stalled cycles per insn
+         6,957,728 branches                  #  542.920 M/sec                  
+           122,049 branch-misses             #    1.75% of all branches         [69.03%]
 
-       0.015131930 seconds time elapsed
+       0.011857863 seconds time elapsed
+
+$ perf stat ./dxexmachina_phold opdevs 2 1 10 0 10
+ Performance counter stats for './dxexmachina_phold opdevs 2 1 10 0 10':
+
+         58.345592 task-clock (msec)         #    0.103 CPUs utilized          
+               235 context-switches          #    0.004 M/sec                  
+                24 cpu-migrations            #    0.411 K/sec                  
+               844 page-faults               #    0.014 M/sec                  
+        86,073,977 cycles                    #    1.475 GHz                     [79.15%]
+        41,603,661 stalled-cycles-frontend   #   48.33% frontend cycles idle    [87.74%]
+        23,030,593 stalled-cycles-backend    #   26.76% backend  cycles idle    [54.61%]
+        93,532,407 instructions              #    1.09  insns per cycle        
+                                             #    0.44  stalled cycles per insn [79.51%]
+        20,525,076 branches                  #  351.785 M/sec                   [88.38%]
+           426,419 branch-misses             #    2.08% of all branches         [93.36%]
+
+       0.564278040 seconds time elapsed
+
+$ perf stat ./dxexmachina_phold cpdevs 2 1 10 0 10
+ Performance counter stats for './dxexmachina_phold cpdevs 2 1 10 0 10':
+
+         45.097489 task-clock (msec)         #    0.453 CPUs utilized          
+                46 context-switches          #    0.001 M/sec                  
+                13 cpu-migrations            #    0.288 K/sec                  
+               856 page-faults               #    0.019 M/sec                  
+        68,793,923 cycles                    #    1.525 GHz                     [67.00%]
+        40,973,357 stalled-cycles-frontend   #   59.56% frontend cycles idle    [82.61%]
+        15,619,202 stalled-cycles-backend    #   22.70% backend  cycles idle    [71.88%]
+        56,478,836 instructions              #    0.82  insns per cycle        
+                                             #    0.73  stalled cycles per insn [91.10%]
+        13,722,715 branches                  #  304.290 M/sec                   [91.32%]
+           279,636 branch-misses             #    2.04% of all branches         [90.09%]
+
+       0.099622680 seconds time elapsed
 \end{Verbatim}
@@
 \begin{Verbatim}[fontsize=\small]
-$ perf stat python experiment.py default 2 false > out.txt
+$ perf stat ./dxexmachina_phold classic 1 10 0 10
+ Performance counter stats for 'python ./benchmarks/experiment.py':
 
- Performance counter stats for 'python experiment.py default 2 false':
+        671.561546 task-clock (msec)         #    0.998 CPUs utilized          
+                97 context-switches          #    0.144 K/sec                  
+                18 cpu-migrations            #    0.027 K/sec                  
+             2,974 page-faults               #    0.004 M/sec                  
+     1,916,305,122 cycles                    #    2.854 GHz                     [83.35%]
+     1,005,734,716 stalled-cycles-frontend   #   52.48% frontend cycles idle    [83.34%]
+       499,103,479 stalled-cycles-backend    #   26.05% backend  cycles idle    [66.69%]
+     2,060,775,055 instructions              #    1.08  insns per cycle        
+                                             #    0.49  stalled cycles per insn [83.34%]
+       345,264,964 branches                  #  514.123 M/sec                   [83.89%]
+        12,293,319 branch-misses             #    3.56% of all branches         [82.87%]
 
-        709.776558 task-clock (msec)         #    0.989 CPUs utilized          
-               191 context-switches          #    0.269 K/sec                  
-                34 cpu-migrations            #    0.048 K/sec                  
-             2,942 page-faults               #    0.004 M/sec                  
-     1,687,298,167 cycles                    #    2.377 GHz                     [82.71%]
-       866,759,527 stalled-cycles-frontend   #   51.37% frontend cycles idle    [83.25%]
-       562,247,464 stalled-cycles-backend    #   33.32% backend  cycles idle    [67.62%]
-     1,672,919,036 instructions              #    0.99  insns per cycle        
-                                             #    0.52  stalled cycles per insn [83.82%]
-       375,443,818 branches                  #  528.961 M/sec                   [83.25%]
-        14,867,814 branch-misses             #    3.96% of all branches         [83.28%]
+       0.673223302 seconds time elapsed
 
-       0.717576473 seconds time elapsed
+$ perf stat ./dxexmachina_phold opdevs 2 1 10 0 10
+ Performance counter stats for 'mpirun -np 2 python ./benchmarks/experiment.py':
+
+        866.680461 task-clock (msec)         #    0.800 CPUs utilized          
+               929 context-switches          #    0.001 M/sec                  
+                83 cpu-migrations            #    0.096 K/sec                  
+             7,466 page-faults               #    0.009 M/sec                  
+     2,326,077,914 cycles                    #    2.684 GHz                     [83.64%]
+     1,283,069,850 stalled-cycles-frontend   #   55.16% frontend cycles idle    [83.95%]
+       940,158,530 stalled-cycles-backend    #   40.42% backend  cycles idle    [65.55%]
+     2,325,361,346 instructions              #    1.00  insns per cycle        
+                                             #    0.55  stalled cycles per insn [82.69%]
+       402,983,176 branches                  #  464.973 M/sec                   [83.67%]
+        15,476,432 branch-misses             #    3.84% of all branches         [84.05%]
+
+       1.083881333 seconds time elapsed
 \end{Verbatim}
-
-\section{Parallel DEVS}
-Also in parallel simulations \textsc{DEVS Ex Machina} completes quite a bit quicker, again achieving higher CPU utilization.
-Additionally we can see that Conservative PDEVS performs consistently significantly faster than Parallel PDEVS, in the case of PHOLD.
-
-\subsection{DEVStone}
-\paragraph{\textsc{DEVS Ex Machina} - Optimistic}
-\begin{Verbatim}[fontsize=\small]
-$ perf stat ./dxexmachina_devstone opdevs 2 2 3
-
- Performance counter stats for './dxexmachina_devstone opdevs 2 2 3':
-
-         95.265846 task-clock (msec)         #    0.499 CPUs utilized          
-               191 context-switches          #    0.002 M/sec                  
-                15 cpu-migrations            #    0.157 K/sec                  
-             1,229 page-faults               #    0.013 M/sec                  
-       197,702,110 cycles                    #    2.075 GHz                     [82.53%]
-        95,414,988 stalled-cycles-frontend   #   48.26% frontend cycles idle    [83.54%]
-        47,283,951 stalled-cycles-backend    #   23.92% backend  cycles idle    [76.71%]
-       199,327,993 instructions              #    1.01  insns per cycle        
-                                             #    0.48  stalled cycles per insn [91.61%]
-        47,887,986 branches                  #  502.677 M/sec                   [86.14%]
-           931,688 branch-misses             #    1.95% of all branches         [75.36%]
-
-       0.191066656 seconds time elapsed
-\end{Verbatim}
-\paragraph{\textsc{DEVS Ex Machina} - Conservative}
-\begin{Verbatim}[fontsize=\small]
-$ perf stat ./dxexmachina_devstone cpdevs 2 2 3
-
- Performance counter stats for './dxexmachina_devstone cpdevs 2 2 3':
-
-        129.292299 task-clock (msec)         #    0.555 CPUs utilized          
-               672 context-switches          #    0.005 M/sec                  
-                15 cpu-migrations            #    0.116 K/sec                  
-             2,177 page-faults               #    0.017 M/sec                  
-       308,983,665 cycles                    #    2.390 GHz                     [74.75%]
-       151,616,998 stalled-cycles-frontend   #   49.07% frontend cycles idle    [84.43%]
-        82,891,133 stalled-cycles-backend    #   26.83% backend  cycles idle    [76.81%]
-       276,296,034 instructions              #    0.89  insns per cycle        
-                                             #    0.55  stalled cycles per insn [88.68%]
-        66,513,459 branches                  #  514.443 M/sec                   [87.86%]
-         1,390,075 branch-misses             #    2.09% of all branches         [77.49%]
-
-       0.233046274 seconds time elapsed
-\end{Verbatim}
-\paragraph{\textsc{PythonPDEVS}}
-\begin{Verbatim}[fontsize=\small]
-$ perf stat mpirun -np 2 python experiment.py default 2 false > out.txt
-
- Performance counter stats for 'mpirun -np 2 python experiment.py default 2 false':
-
-        787.125992 task-clock (msec)         #    0.371 CPUs utilized          
-             1,120 context-switches          #    0.001 M/sec                  
-                91 cpu-migrations            #    0.116 K/sec                  
-             8,684 page-faults               #    0.011 M/sec                  
-     1,786,217,349 cycles                    #    2.269 GHz                     [83.85%]
-       935,480,076 stalled-cycles-frontend   #   52.37% frontend cycles idle    [84.18%]
-       619,366,137 stalled-cycles-backend    #   34.67% backend  cycles idle    [65.26%]
-     1,796,463,414 instructions              #    1.01  insns per cycle        
-                                             #    0.52  stalled cycles per insn [84.16%]
-       403,042,819 branches                  #  512.044 M/sec                   [84.70%]
-        13,210,328 branch-misses             #    3.28% of all branches         [82.68%]
-
-       2.123888919 seconds time elapsed
-\end{Verbatim}
-
-\subsection{PHOLD}
-\paragraph{\textsc{DEVS Ex Machina} - Optimistic}
-\begin{Verbatim}[fontsize=\small]
-$ perf stat ./dxexmachina_phold opdevs 2 1 10 0 10
-
- Performance counter stats for './dxexmachina_phold opdevs 2 1 10 0 10':
-
-         20.741980 task-clock (msec)         #    0.092 CPUs utilized          
-                73 context-switches          #    0.004 M/sec                  
-                 8 cpu-migrations            #    0.386 K/sec                  
-               760 page-faults               #    0.037 M/sec                  
-        28,281,718 cycles                    #    1.364 GHz                     [42.03%]
-        14,899,039 stalled-cycles-frontend   #   52.68% frontend cycles idle    [79.62%]
-         8,373,668 stalled-cycles-backend    #   29.61% backend  cycles idle   
-        29,601,476 instructions              #    1.05  insns per cycle        
-                                             #    0.50  stalled cycles per insn
-         6,754,172 branches                  #  325.628 M/sec                  
-           135,220 branch-misses             #    2.00% of all branches         [81.22%]
-
-       0.225834153 seconds time elapsed
-\end{Verbatim}
-\paragraph{\textsc{DEVS Ex Machina} - Conservative}
-\begin{Verbatim}[fontsize=\small]
-$ perf stat ./dxexmachina_phold cpdevs 2 1 10 0 10
-
- Performance counter stats for './dxexmachina_phold cpdevs 2 1 10 0 10':
-
-         28.509489 task-clock (msec)         #    0.286 CPUs utilized          
-                95 context-switches          #    0.003 M/sec                  
-                11 cpu-migrations            #    0.386 K/sec                  
-               813 page-faults               #    0.029 M/sec                  
-        54,757,291 cycles                    #    1.921 GHz                     [63.45%]
-        29,057,227 stalled-cycles-frontend   #   53.07% frontend cycles idle    [86.87%]
-        15,602,463 stalled-cycles-backend    #   28.49% backend  cycles idle    [90.08%]
-        53,326,834 instructions              #    0.97  insns per cycle        
-                                             #    0.54  stalled cycles per insn
-        12,449,801 branches                  #  436.690 M/sec                  
-           219,155 branch-misses             #    1.76% of all branches         [64.68%]
-
-       0.099591165 seconds time elapsed
-\end{Verbatim}
-\paragraph{\textsc{PythonPDEVS}}
-PHOLD in parallel on PyPDEVS crashed immediately with an MPI error, hence a comparison was not possible for now.
 

### 2015-06-15 manual/projectrapport.tex
Expanded examples section of project report

@@
 \usepackage{fancyvrb}
+\usepackage{wrapfig}
 \usepackage{hyperref}
@@
 As these are standardized tests, they allow easy comparison between our implementation and \textit{PythonPDEVS}.\\
-Note: at the moment, no major research has gone into the use of compiler optimizations yet. Potentially, there is still speed to be gained here.\\
 \\
-{\footnotesize[All the following tests were performed using an \textit{Intel Core i5 M520} CPU at 2.40GHz, parallel testing was done using 2 simulation cores.
+{\footnotesize[All the following tests were performed using an \textit{Intel Core i5 M520} CPU at 2.40GHz, parallel testing was done using 2 simulation cores unless mentioned otherwise.
+Compilation was done with the \texttt{-O3} optimization flag.
 All tests were ran 5 times to give an indication of the predictability of the result.]}
@@
 In both environments \textsc{DEVS Ex Machina} is significantly faster than \textit{PyPDEVS}.
-Especially interesting is the second graph with parellel results.
-Whereas in DEVStone Optimistic and Conservative DEVS performed relatively equally, here we can spot a clear discrepancy between them.\\
+Especially interesting is the second graph with parellel results. Whereas in DEVStone Optimistic and Conservative DEVS performed relatively equally, here we can spot a clear discrepancy between them.
 
 In Optimistic we can see that as the payload grows, the total runtime becomes less predictable.
-Because of the inherent interconnectivity of the benchmark, and because processing a message with a larger payload can take a while, the chance of reverts or even series of reverts is increased.
+Because of the inherent interconnectivity of the benchmark (lots of inter-core communication), and because processing a message with a larger payload can take a while, the chance of reverts or even series of reverts is increased.
 
@@
 
+\chapter{Examples}
+As part of the assignment we've implemented a set of examples for our simulator. Most of them are based on models already described in Yentl's thesis.
+
+\subsubsection{Traffic light and Traffic System}
+These simple models are an exact copy of those used as examples in pypdevs.
+
+\subsubsection{DEVStone}
+DEVStone is a standardized benchmark for DEVS.
+It can be run using the separate executable \texttt{dxexmachina\_devstone}.
+If this executable has not been generated, run the following command from the \texttt{build} folder: \texttt{make dxexmachina\_devstone}\\
+The executable takes a few arguments.
+First you can specify whether you want to run it in Classic DEVS mode (\texttt{classic}) or Parallel (\texttt{opdevs} and \texttt{cpdevs} for Optimistic and Conservative respectively) followed by the amount of cores.\\
+Next you can specify the width and the depth of the benchmark. By default this is 2 and 3.
+
+\subsubsection{PHOLD}
+PHOLD simulates a highly interconnected environment. Running it follows a procedure largely similar to \textbf{DEVStone}. The benchmark runs from a separate executable \texttt{dxexmachina\_phold}.\\
+Again, its first arguments are the type of simulation.
+What follows is, in order, the amount of nodes in the \textbf{PHOLD}, the amount of atomic models per node, the size of the payload, and the percentage of remotes.
+
+\subsubsection{Forestfire}
+...\\
+\\
+\textbf{NOTE} We added an extra executable to the project containing only the necessary implementation for the traffic system model to work.
+
+\subsubsection{Traffic system}
+The \textbf{traffic system} is based on the traffic system described in Yentl's paper. We've implemented two versions, a bigger one and a smaller one.\\
+A modified version of Yenl's script generates the header and source file for the city model and makes it possible to easily adopt the models size.
+The script can be found in the './main/script' folder of our repository.\\
+\\
+Run the following command to generate a city:
+\begin{Verbatim}[fontsize=\small]
+$ python script_devsexmachina.py {x} {y} {nodes} {districts}
+\end{Verbatim}
+The small city is 10 x 50 segments large and the larger city exists of 10 x 500 segments. Both cities are generated for 50 nodes and 5 districts as described in Yentl's thesis.\\
+\\
+\textbf{NOTE} We added an extra executable to the project containing only the necessary implementation for the traffic system model to work.
+This executable contains the larger version of the city.
+To build large cities you have to ensure that there is a sufficient amount of memory available.
+
+\subsubsection{Virus}
+\begin{wrapfigure}{r}{0.3\textwidth}
+  \begin{center}
+    \includegraphics[width=0.28\textwidth]{virus}
+  \end{center}
+\end{wrapfigure}
+
+Virus simulates the spread of two viruses over a pool of healthy cells.\\
+
+The procedure is quite simple: each type of virus starts out in a single cell.
+Every cell, once in control of a virus, produces a set number of new viruses every turn and also sends out some of its contained viruses to a neighbour; this can be a neutral cell (thereby taking it over), a friendly cell (reenforcing it) or a cell owned by the other virus (reducing its numbers and possibly taking over the cell).\\
+As a result this produces quite dynamic behavior, with both viruses spreading and moving around the pool or getting into stalemates.\\
+
+This model is relatively interesting because it is well-connected and gradually produces more and more messages.
+Just like \textbf{PHOLD} it has performance gains in Conservative PDEVS over Optimistic.\\
+\\
+\textbf{Virus} has its own separate executable, \texttt{dxexmachina\_virus} which can be generated and used just like those of \textbf{DEVStone} and \textbf{PHOLD}.
+After specifying the simulation type the arguments are respectively the size of the cell pool (apart from the two starting cells), and the maximum amount of connections per cell.
+
 \begin{thebibliography}{1}

### 2015-06-15 manual/projectrapport.tex
Filled section on Forestfire, other minor additions

@@
 First you can specify whether you want to run it in Classic DEVS mode (\texttt{classic}) or Parallel (\texttt{opdevs} and \texttt{cpdevs} for Optimistic and Conservative respectively) followed by the amount of cores.\\
-Next you can specify the width and the depth of the benchmark. By default this is 2 and 3.
+Next you can specify the width and the depth of the benchmark. By default this is 2 and 3.\\
+For more complex changes you can alter the source file \texttt{maindevstone.cpp}. Be sure to run \texttt{make dxexmachina\_devstone} again afterwards to recompile.
 
@@
 Again, its first arguments are the type of simulation.
-What follows is, in order, the amount of nodes in the \textbf{PHOLD}, the amount of atomic models per node, the size of the payload, and the percentage of remotes.
+What follows is, in order, the amount of nodes in the \textbf{PHOLD}, the amount of atomic models per node, the size of the payload, and the percentage of remotes.\\
+The configuration's source file is \texttt{mainphold.cpp}.
 
 \subsubsection{Forestfire}
-...\\
-\\
-\textbf{NOTE} We added an extra executable to the project containing only the necessary implementation for the traffic system model to work.
+Forestfire is a highly demanding interconnected model that inherently has a high chance of reverts or even series of reverts.\\
+At the top of its source file, \texttt{mainbenchfire.cpp}, you can configure its 2 parameters, the width and height of the forest, as well as the duration of the test by altering the values of the macros \texttt{BENCH\_FIRE\_SIZEX}, \texttt{BENCH\_FIRE\_SIZEY} and \texttt{BENCH\_FIRE\_MAXTIME}.\\
+
+The executable for this sole test is \texttt{dxexmachina\_benchfire}.
 
@@
 \\
-\textbf{NOTE} We added an extra executable to the project containing only the necessary implementation for the traffic system model to work.
+We added an extra executable, \texttt{dxexmachina\_traffic}, to the project containing only the necessary implementation for the traffic system model to work.
 This executable contains the larger version of the city.
-To build large cities you have to ensure that there is a sufficient amount of memory available.
+To build large cities you have to ensure that there is a sufficient amount of memory available.\\
+The source file of this executable is \texttt{maintraffic.cpp}.
 
@@
 \textbf{Virus} has its own separate executable, \texttt{dxexmachina\_virus} which can be generated and used just like those of \textbf{DEVStone} and \textbf{PHOLD}.
-After specifying the simulation type the arguments are respectively the size of the cell pool (apart from the two starting cells), and the maximum amount of connections per cell.
+After specifying the simulation type the arguments are respectively the size of the cell pool (apart from the two starting cells), and the maximum amount of connections per cell.\\
+The source of this configuration can be found in \texttt{mainvirus.cpp}.
 

### 2015-11-08 main/src/model/core.cpp
fixed scheduler errors due to duplicate messages

@@ -650,15 +650,8 @@ t_timestamp n_model::Core::getFirstMessageTime()
          */
         t_timestamp mintime = t_timestamp::infinity();
         this->lockMessages();
-        while (not this->m_received_messages->empty()) {
-                t_msgptr msg = this->m_received_messages->top().getMessage();
-                if (msg->flagIsSet(Status::TOERASE)) {
-                        this->m_received_messages->pop();
-                        msg->setFlag(Status::KILL);
-                } else {
-                        mintime = this->m_received_messages->top().getMessage()->getTimeStamp();
-                        break;
-                }
+        if (not this->m_received_messages->empty()) {
+                mintime = this->m_received_messages->top().getMessage()->getTimeStamp();
         }
         this->unlockMessages();
         LOG_DEBUG("\tCORE :: ", this->getCoreID(), " first message time == ", mintime);

### 2015-11-08 main/src/model/optimisticcore.cpp

@@ -17,8 +17,9 @@ Optimisticcore::~Optimisticcore()
 {
         // Destructors are run on main(), our pool is live but we can't access it anymore.
         // Do not delete ptrs here.
-        for(auto& ptr : m_sent_messages){
-                LOG_ERROR("MCORE:: ", this->getCoreID(), " HAVE ", ptr, " in  m_sent_messages @ destruction. This will result in an std::bad_alloc exception.");
+        for (auto& ptr : m_sent_messages) {
+                LOG_ERROR("MCORE:: ", this->getCoreID(), " HAVE ", ptr,
+                        " in  m_sent_messages @ destruction. This will result in an std::bad_alloc exception.");
                 // We're back on main's thread, cannot call our pool.
                 ptr->releaseMe();
                 m_stats.logStat(DELMSG);
@@ -26,18 +27,19 @@ Optimisticcore::~Optimisticcore()
         m_sent_messages.clear();
         // Another edge case, if we quit simulating before getting all messages from the network, we leak memory if 
         // any of these is an antimessage.
-        
-        if(m_network->havePendingMessages(this->getCoreID())){
+
+        if (m_network->havePendingMessages(this->getCoreID())) {
                 LOG_ERROR("OCORE::", this->getCoreID(), " destructor detected messages in network for us, purging.");
                 // Pull them in case another thread is waiting on network idle.
-                std::vector<t_msgptr>msgs = m_network->getMessages(this->getCoreID());
+                std::vector<t_msgptr> msgs = m_network->getMessages(this->getCoreID());
         }
 }
 void Optimisticcore::shutDown()
 {
         assert(!isLive() && "The core shouldn't be live when it is shut down!");
-        LOG_DEBUG("MCORE:: ", this->getCoreID(), " core shutting down, still need to remove ", m_sent_messages.size(), " messages");
-        for(auto& ptr : m_sent_messages){
+        LOG_DEBUG("MCORE:: ", this->getCoreID(), " core shutting down, still need to remove ", m_sent_messages.size(),
+                " messages");
+        for (auto& ptr : m_sent_messages) {
                 LOG_DEBUG("MCORE:: ", this->getCoreID(), " deleting ", ptr, " in core shutdown.");
                 // We're not yet back on main's thread, can safely call our pool.
[…]

### 2015-11-08 main/src/test/coretest.cpp

@@ -270,6 +270,9 @@ TEST(Optimisticcore, revert){
 	coreone->receiveMessage(msgaftergvt);		// this triggers a new revert, we were @67, now @63
 	EXPECT_EQ(coreone->getTime().getTime(), 63u);
 	coreone->runSmallStep();			// does nothing, check that empty transitioning works. (next = 108, time = 62)
+
+    coreone->setLive(false);
+	coreone->shutDown();
 	EXPECT_EQ(coreone->getTime().getTime(), 108u);
         // Message is not sent by any core, it's not processed, it's not queued so delete here (since it won't be otherwise)
         delete msgaftergvt;
@@ -351,6 +354,11 @@ TEST(Optimisticcore, revertidle){
 	n_tracers::clearAll();
 	n_tracers::waitForTracer();
 	tracers->finishTrace();
+
+    c1->setLive(false);
+    c1->shutDown();
+    c2->setLive(false);
+    c2->shutDown();
 
 	EXPECT_TRUE(std::static_pointer_cast<AtomicModel_impl>(m->getComponents()[0])->getCorenumber()
 			!=
@@ -434,6 +442,11 @@ TEST(Optimisticcore, revertedgecases){
 	n_tracers::clearAll();
 	n_tracers::waitForTracer();
 	tracers->finishTrace();
+
+    c1->setLive(false);
+    c1->shutDown();
+    c2->setLive(false);
+    c2->shutDown();
 
 	EXPECT_TRUE(std::static_pointer_cast<AtomicModel_impl>(m->getComponents()[0])->getCorenumber()
 			!=
@@ -520,6 +533,11 @@ TEST(Optimisticcore, revertoffbyone){
 	n_tracers::waitForTracer();
 	tracers->finishTrace();
 
+    c1->setLive(false);
+    c1->shutDown();
+    c2->setLive(false);
+    c2->shutDown();
+
 	EXPECT_TRUE(std::static_pointer_cast<AtomicModel_impl>(m->getComponents()[0])->getCorenumber()
 			!=
 		std::static_pointer_cast<AtomicModel_impl>(m->getComponents()[1])->getCorenumber());
@@ -599,6 +617,11 @@ TEST(Optimisticcore, revertstress){
 	n_tracers::waitForTracer();
 	tracers->finishTrace();
 
+    c1->setLive(false);
+    c1->shutDown();
+    c2->setLive(false);
[…]

### 2015-11-08 profiling/report.tex
Add new profiling data for single core/ conservative. Small report for meeting.

\documentclass[]{article}

%opening
\title{Notes on DXExMachina performance.}
\author{}

\begin{document}

\maketitle
\section{Accompanying documents}
Benchmarks (on master) in separate pdf. Further benchmarks and profiling data in repository.

\section{Improvements since abstract}
\begin{itemize}
	\item Conservative handles cyclic simulations.
	\item Scheduler ported from Adevs + own heuristic.
	\item Memory pooling.
	\subitem Aligning messages on cache lines to avoid false sharing
	\subitem Thread local pools to avoid expensive cross-thread alloc/dealloc.
	\subitem Gvt required for conservative, based on existing null message exchange.
	\item Frequency of locking reduced, (sparse) usage of memory\_order\_relaxed. Where possible explicit	locking switched to atomic.
	\item A (large) nr of bugfixes and code optimizations.
\end{itemize}
\section{Summarized results benchmarks}
For the detailed results we refer to accompanying pdf and profiling call graphs.
\subsection{Devstone}
Faster due to directconnect. Optimistic suffers from transitions at same time points (frequency reverts higher), can gain with random ta.
\subsection{Phold}
Slightly faster than adevs. A reasonable amount of messaging can be done without slowing down simulator. Most of the runtime is spent in rng.
\subsection{Interconnect}
Slower than adevs due to high volume of messages. This creates a very cache unfriendly pattern, combined with cost of allocation on the heap for dxex. 
The completely cyclic dependencies of this model makes it very hard for our parallel implementations. Usage of pools reduces calls to new/malloc/sbrk, but can't eliminate it. Cache pattern is far harder to correct. Alignment of a Message on cache lines, and quadratic growing pools also help, but can't solve the problem.
\section{Open issues to improve}
\begin{itemize}
	\item Conservative lookahead caching (devstone -r), cpdevs is forced to ask useless lookahead values.
	\item Optimistic use pools in states.
	\item Directconnect : Yentl's faster algorithm. Not (yet) implement since it does not show in profiling graphs as a significant (<1\%) cost.
	\item Conservative : Exploit full dependency graph to calculate next event directly (unsure of benefits).
\end{itemize}
\section{Causes of slowdown (hardware)}
A short illustration of key differences in hardware events for both simulators.
\begin{itemize}
	\item Cache : Adevs : $<$ 1\% cache misses, dxex : $<$10\% cache misses
	\item Context switches : Adevs : 4e-8 switches/cycles, dxex 3e-8 (conservative)
	\item Branch prediction : Adevs : 4\%, dxex 0.08%
	\item Cpu stalls : Adevs : 0.2 stalled cycle/instruction, dxex 0.8 stalled cycle/instruction.
\end{itemize}
The switches are (in single core) caused by syscalls (malloc, sbrk, free, time), in parallel by kernel space locking (mutex, thread, (depending on implementation atomics)).

\end{document}

### 2015-11-08 main/src/control/controller.cpp

@@ -560,8 +560,11 @@ void cvworker_con(std::size_t myid, std::size_t turns, Controller& ctrl)
 	}
         // Wait for all other cores to go idle.
         // Should a core have reached the nr of turns (a safety catch), make sure we set Live ourselves.
-        if(i==turns)
+        if(i==turns){
+                LOG_WARNING("CVWORKER: ", core->getCoreID(), " overran nr of simulation steps allowed !!");
                 core->setLive(false);
+                // Edge case : if we hit this, we can't guarantee deallocating since sender can't verify we have reached termination.
+        }
         while(true){
                 size_t i = 0;
                 for(const auto& core : ctrl.m_cores){

### 2015-11-08 main/src/model/conservativecore.cpp

@@ -487,8 +487,8 @@ Conservativecore::shutDown()
                 if(recv_time <= ptr->getTimeStamp().getTime() && recv_time!=this->getTerminationTime().getTime()){
                         LOG_ERROR("Pointer : " , ptr, " with receiver id ", destid, " nulltime = ", recv_time, " msgtime = ", ptr->getTimeStamp());
                         LOG_FLUSH;
-                        
-                        throw std::logic_error("Pointer to message still in use !");
+                        // Enable if edge case in cv_worker is resolved.
+                        //throw std::logic_error("Pointer to message still in use !");
                 }else{
                         ptr->releaseMe();
                 }

### 2015-11-08 profiling/report.tex
Update report + new call graphs.

@@
 Alignment of a Message on cache lines, and quadratic growing pools also help, but can't solve the problem.
-\section{Open issues to improve}
+\section{Open issues}
 \begin{itemize}
 	\item Conservative lookahead caching (devstone -r), cpdevs is forced to ask useless lookahead values.
-	\item Optimistic use pools in states.
-	\item Directconnect : Yentl's faster algorithm. Not (yet) implement since it does not show in profiling graphs as a significant (<1\%) cost.
+	\item Optimistic use pools in states. (from shared\_ptr to ptr), see profile opdevs in repo.
+	\item Optimistic hangs/too slow in cyclic (interconnect, phold)
+	\item Directconnect : Yentl's faster algorithm. Not (yet) implement since it does not show in profiling graphs as a significant ($<$1\%) cost.
 	\item Conservative : Exploit full dependency graph to calculate next event directly (unsure of benefits).

### 2015-11-08 main/src/model/atomicmodel.cpp

@@ -124,7 +124,7 @@ void AtomicModel_impl::setGVT(t_timestamp gvt)
 			k = std::max(0, index - 1);
 			break;
 		}
-		index++;
+		++index;
 	}
 
 	if (k == -1) {

### 2015-11-08 main/src/model/atomicmodel.cpp
temp

@@ -26,6 +26,12 @@ AtomicModel_impl::AtomicModel_impl(std::string name, int corenumber, std::size_t
         LOG_DEBUG("\tAMODEL ctor :: name=", name, " m_prior= ", m_priority , " corenr=", m_corenumber);
 }
 
+AtomicModel_impl::~AtomicModel_impl()
+{
+        for(t_stateptr st: m_oldStates)
+                n_tools::takeBack(st);
+}
+
 void AtomicModel_impl::intTransition()
 {
 	LOG_ERROR("ATOMICMODEL: Not implemented: 'void n_model::AtomicModel::intTransition()'");

### 2015-11-08 main/src/model/atomicmodel.h

@@ -116,7 +116,7 @@ public:
 	 */
 	AtomicModel_impl(std::string name, int corenumber, std::size_t priority = std::numeric_limits<size_t>::max());
         
-	virtual ~AtomicModel_impl() = default;
+	virtual ~AtomicModel_impl();
 
         /**
          * @return uuid object.
@@ -410,22 +410,22 @@ public:
 	AtomicModel(std::string name, const T& value, std::size_t priority = 0):
 		AtomicModel_impl(name, priority)
 	{
-		initState(n_tools::createObject<State__impl<t_type>>(value));
+		initState(n_tools::createRawObject<State__impl<t_type>>(value));
 	}
 	AtomicModel(std::string name, const T& value, int coreNum, std::size_t priority = 0):
 		AtomicModel_impl(name, coreNum, priority)
 	{
-		initState(n_tools::createObject<State__impl<t_type>>(value));
+		initState(n_tools::createRawObject<State__impl<t_type>>(value));
 	}
 	AtomicModel(std::string name, std::size_t priority = 0):
 		AtomicModel_impl(name, priority)
 	{
-		initState(n_tools::createObject<State__impl<t_type>>());
+		initState(n_tools::createRawObject<State__impl<t_type>>());
 	}
 	AtomicModel(std::string name, int coreNum, std::size_t priority = 0):
 		AtomicModel_impl(name, coreNum, priority)
 	{
-		initState(n_tools::createObject<State__impl<t_type>>());
+		initState(n_tools::createRawObject<State__impl<t_type>>());
 	}
 
 	/**
@@ -433,7 +433,7 @@ public:
 	 */
 	constexpr const t_type& state() const
 	{
-		return n_tools::staticCast<const State__impl<t_type>>(getState())->m_value;
+		return n_tools::staticRawCast<const State__impl<t_type>>(getState())->m_value;
 	}
 
 	/**
@@ -441,7 +441,7 @@ public:
 	 */
 	t_type& state()
 	{
-		return n_tools::staticCast<State__impl<t_type>>(getState())->m_value;
+		return n_tools::staticRawCast<State__impl<t_type>>(getState())->m_value;
 	}
 
 };
@@ -456,12 +456,12 @@ public:
 	AtomicModel(std::string name, std::size_t priority = 0):
 		AtomicModel_impl(name, priority)
 	{
-		initState(n_tools::createObject<State__impl<void>>());
[…]

### 2015-11-08 main/src/model/cellmodel.h

@@ -60,22 +60,22 @@ public:
 	CellAtomicModel(std::string name, t_point pt, const T& value, std::size_t priority = 0):
 		CellAtomicModel_impl(name, pt, priority)
 	{
-		initState(n_tools::createObject<State__impl<t_type>>(value));
+		initState(n_tools::createRawObject<State__impl<t_type>>(value));
 	}
 	CellAtomicModel(std::string name, t_point pt, const T& value, int coreNum, std::size_t priority = 0):
 		CellAtomicModel_impl(name, pt, coreNum, priority)
 	{
-		initState(n_tools::createObject<State__impl<t_type>>(value));
+		initState(n_tools::createRawObject<State__impl<t_type>>(value));
 	}
 	CellAtomicModel(std::string name, t_point pt, std::size_t priority = 0):
 		CellAtomicModel_impl(name, pt, priority)
 	{
-		initState(n_tools::createObject<State__impl<t_type>>());
+		initState(n_tools::createRawObject<State__impl<t_type>>());
 	}
 	CellAtomicModel(std::string name, t_point pt, int coreNum, std::size_t priority = 0):
 		CellAtomicModel_impl(name, pt, coreNum, priority)
 	{
-		initState(n_tools::createObject<State__impl<t_type>>());
+		initState(n_tools::createRawObject<State__impl<t_type>>());
 	}
 
 	/**

### 2015-11-08 main/src/network/message.h

@@ -30,12 +30,10 @@ enum MessageColor : uint8_t{WHITE = 0, RED = 1};
 // 2^1: delete? set by optimistic core for the case
 //              where a message is marked antimessage before it is received by the core
 // 2^2: processed? set by optimistic core when a message has been processed (used in a transition)
-// 2^3: TOERASE? set by optimistic core when a message in the scheduler is an antimessage,
-//               The core can then safely ignore it and set its kill flag
+// 2^3: HEAPED? The message has been put in a message scheduler in the receiving core.
 // 2^4: ANTI? The message is an anti message
 // 2^5: KILL? The message can be safely killed by the sending core.
-// 2^6: HEAPED? The message has been put in a message scheduler in the receiving core.
-enum Status : uint8_t{COLOR=MessageColor::RED, DELETE=2, PROCESSED=4, TOERASE=8, ANTI=16, KILL=32, HEAPED=64};
+enum Status : uint8_t{COLOR=MessageColor::RED, DELETE=2, PROCESSED=4, HEAPED=8, ANTI=16, KILL=32};
 
 std::ostream&
 operator<<(std::ostream& os, const MessageColor& c);

### 2015-11-08 profiling/report.tex
Report + extra notes.

@@
 \section{Accompanying documents}
-Benchmarks (on master) in separate pdf. Further benchmarks and profiling data in repository.
+In the repository (and included in mail) you'll find in folder profiling the following documents:
+\begin{itemize}
+	\item Benchmarks : DXvsADEVS.pdf a nr of benchmarks on a stable commit.
+	\item *.svg : profiling call graphs for all functional benchmarks
+	\item conservative\_bench.txt : Benchmarks after a nr of fixes to the implementation of conservative (complementary to the first pdf.)
+\end{itemize}
 
 \section{Improvements since abstract}
+Full details of all improvements can be seen in the git repository, the following is a short summary:
 \begin{itemize}
@@
 	\item Frequency of locking reduced, (sparse) usage of memory\_order\_relaxed. Where possible explicit	locking switched to atomic.
-	\item A (large) nr of bugfixes and code optimizations.
 \end{itemize}
@@
 \end{itemize}
-The switches are (in single core) caused by syscalls (malloc, sbrk, free, time), in parallel by kernel space locking (mutex, thread, (depending on implementation atomics)).
 

### 2015-11-10 main/src/model/optimisticcore.cpp
2nd part of processed msg handling (commented). Add copyFlag op for message.

@@ -80,7 +80,11 @@ void Optimisticcore::clearProcessedMessages(std::vector<t_msgptr>& msgs)
                                 ptr->toString());
                         ptr->releaseMe();
                 }
-                // else push on m_processed
+                else{
+                        /**
+                        m_processed_messages.push_back(ptr);
+                        */
+                }
         }
         msgs.clear();
 }
@@ -537,20 +541,11 @@ void n_model::Optimisticcore::unlockSimulatorStep()
         LOG_DEBUG("MCORE:: ", this->getCoreID(), " time: ", getTime(), " simulator core unlocked.");
 }
 
-void n_model::Optimisticcore::revert(const t_timestamp& totime)
+void n_model::Optimisticcore::revert(const t_timestamp& rtime)
 {
-        /** Heaped = true
-         *  for auto& in :
-         *   m_processed_messages until totime:
-         *              if not antimessage, push on pending
-         *              else set kill & lose
-         *  for ( riter .... )
-         *       if(riter < totime)
-         *              break
-         *       else
-         *              if(antimsg)...
-         */
-        assert(totime.getTime() >= this->getGVT().getTime());
+        const t_timestamp::t_time totime = rtime.getTime();
+        const t_timestamp::t_time gtime = this->getGVT().getTime();
+        assert(totime >= gtime);
         LOG_DEBUG("MCORE:: ", this->getCoreID(), " reverting from ", this->getTime(), " to ", totime);
         if (!this->isLive()) {
                 LOG_DEBUG("MCORE:: ", this->getCoreID(), " time: ", getTime(), " Core going from idle to active ");
@@ -562,7 +557,7 @@ void n_model::Optimisticcore::revert(const t_timestamp& totime)
         while (!m_sent_messages.empty()) {		// For each message > totime, send antimessage
                 t_msgptr msg = m_sent_messages.back();
                 LOG_DEBUG("MCORE:: ", this->getCoreID(), " reverting message ", msg);
-                if (msg->getTimeStamp().getTime() >= totime.getTime()) {
+                if (msg->getTimeStamp().getTime() >= totime) {
                         m_sent_messages.pop_back();
                         LOG_DEBUG("MCORE:: ", this->getCoreID(), " time: ", getTime(),
                                 " revert : sent message > time , antimessagging. \n ", msg->toString());
@@ -572,11 +567,30 @@ void n_model::Optimisticcore::revert(const t_timestamp& totime)
[…]

### 2015-11-10 paper/1-introduction.tex

%TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
%TODO give introduction to the field: parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
%TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)

### 2015-11-10 paper/3-features.tex

%TODO explain that we are briefly going over the main features
\subsection{Based on PythonPDEVS}
%TODO say that it is based on PythonPDEVS
%TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
%TODO (size: take as much size as you want, depending on how much is shared with PythonPDEVS)
%TODO (maybe: note that models are not compatible)
\subsection{Different Synchronization protocols}
%TODO explain rationale: why different synchronization protocols
%TODO explain core infrastructure a bit, maybe with a diagram or so, which clearly shows that everything is shared, except the "synchronization" part
%TODO claim that this allows you to easily switch between both methods, to always chose the best one
\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...

### 2015-11-10 paper/4-performance.tex

%TODO say that you will now prove your claims about the aproach having nearly no impact
%TODO and potentially even claim that you are faster than adevs for some situations, due to the algorithms from PythonPDEVS and further features
%TODO briefly give some hardware info on the machine (CPU, OS, RAM), used software (version of adevs, GCC, ...), and about number of samples/variance
\subsection{Sequential Simulation}
%TODO we start with sequential simulation results, because this is still base case
\subsubsection{CPU Usage}
%TODO compare CPU time with adevs for some benchmarks
%TODO and explain why there is such a difference
%TODO (spend some time here, as this is also important)
\subsubsection{Memory Usage}
%TODO compare memory usage with adevs for some benchmarks
%TODO remark our use of memory pools and such, but don't spend too much space on this
\subsection{Parallel Simulation}
%TODO explain that we will now compare different synchronization protocols
%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
%TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
%TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)

### 2015-11-10 paper/2-background.tex

%TODO say that you are briefly going to go over all necessary background information first
\subsection{Parallel DEVS}
%TODO briefly explain Parallel DEVS again, just recycle from other papers
\subsection{Conservative Synchronization}
%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
\subsection{Optimistic Synchronization}
%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
\subsection{C++11 Parallelism Features}
%TODO explain what is new in C++11 for parallelism and how it contributes to our cause

### 2015-11-10 paper/5-related-work.tex

%TODO explain about other DEVS simulation kernels, and how they differ
%TODO explain about PythonPDEVS, the most related in terms of feature and "legacy" --> now in C++11 and way better performance
%TODO explain about adevs, the most related in terms of performance --> we implement more synchronization options and offer better performance because ...
%TODO explain about CD++, the most related in terms of synchronization --> we implement everything in a single program, with a non-fragmented code-base
%TODO make mention of Warped, which is the kernel used in CD++, and why we don't use it

### 2015-11-10 paper/6-conclusions.tex

%TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
%TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))

### 2015-11-10 paper/abstract.tex

With the ever increasing complexity of simulation models, parallel simulation
becomes necessary to perform the simulation within reasonable time bounds.
The built-in parallelism of Parallel DEVS is often insufficient to tackle this
problem on its own. Several synchronization algorithms have been proposed,
each with a specific kind of simulation model in mind. Due to the significant
differences between these algorithms, current Parallel DEVS simulation tools
restrict themself to only one such algorithm. In this paper, we present a
Parallel DEVS simulator, grafted on C++11, which offers both conservative and
optimistic simulation. We evaluate the performance gain that can be obtained
by choosing the most appropriate synchronization protocol. Our implementation
is compared to ADEVS using hardware-level profiling on a spectrum of benchmarks.

### 2015-11-12 main/src/test/benchmarktest.cpp
Fix allocation leaving cores empty.

@@ -23,33 +23,6 @@ using namespace n_model;
 using namespace n_tools;
 using namespace n_network;
 
-class DevstoneAlloc: public n_control::Allocator
-{
-private:
-	std::size_t m_maxn;
-public:
-	DevstoneAlloc(): m_maxn(0){
-
-	}
-	virtual size_t allocate(const n_model::t_atomicmodelptr& ptr){
-		auto p = std::dynamic_pointer_cast<n_devstone::Processor>(ptr);
-		if(p == nullptr)
-			return 0;
-		std::size_t res = p->m_num*coreAmount()/m_maxn;
-		if(res >= coreAmount())
-			res = coreAmount()-1;
-		LOG_INFO("Putting model ", ptr->getName(), " in core ", res);
-		return res;
-	}
-
-	virtual void allocateAll(const std::vector<n_model::t_atomicmodelptr>& models){
-		m_maxn = models.size();
-		assert(m_maxn && "Total amount of models can't be zero.");
-		for(const n_model::t_atomicmodelptr& ptr: models)
-			ptr->setCorenumber(allocate(ptr));
-	}
-};
-
 class PHoldAlloc: public n_control::Allocator
 {
 private:
@@ -110,7 +83,7 @@ TEST(Benchmark, devstone_single)
 	conf.m_coreAmount = 4;
 	conf.m_saveInterval = 250;
 	conf.m_zombieIdleThreshold = 10;
-	conf.m_allocator = n_tools::createObject<DevstoneAlloc>();
+	conf.m_allocator = n_tools::createObject<n_devstone::DevstoneAlloc>();
 	std::size_t width = 5;
 	std::size_t depth = 5;
 	bool randTa = false;
@@ -138,7 +111,7 @@ TEST(Benchmark, devstone_opt)
 	conf.m_coreAmount = 4;
 	conf.m_saveInterval = 250;
 	conf.m_zombieIdleThreshold = 10;
-	conf.m_allocator = n_tools::createObject<DevstoneAlloc>();
+	conf.m_allocator = n_tools::createObject<n_devstone::DevstoneAlloc>();
 	std::size_t width = 5;
 	std::size_t depth = 5;
 	bool randTa = false;
@@ -166,7 +139,7 @@ TEST(Benchmark, devstone_cons)
 	conf.m_coreAmount = 4;
 	conf.m_saveInterval = 250;
[…]

### 2015-11-12 main/src/control/controller.cpp
comment removed ;controller.cpp edited online with Bitbucket

@@ -529,7 +529,7 @@ void beginGVT(Controller& ctrl, std::atomic<bool>& m_rungvt)
 		if(infguard < ++i){
 			LOG_WARNING("Controller :: GVT overran max ", infguard, " nr of invocations, breaking of.");
 			m_rungvt.store(false);
-			break;	// No join, have not started thread.
+			break;
 		}
 		std::chrono::milliseconds ms { ctrl.getGVTInterval() };// Wait before running gvt, this prevents an obvious gvt of zero.
 		std::this_thread::sleep_for(ms);

### 2015-11-18 paper/1-introduction.tex

@@
 %TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
-%TODO give introduction to the field: parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
+%TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
+\subsection{}
+%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
+% Very rough draft here. distributed + smem or smem only ?
+Leveraging the shared memory parallelism offered by most modern hardware, the runtime of non-trivial model can be significantly reduced.
+\subsection{}
 %TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
+% todo ref openmp
+Adevs offers conservative synchronized parallel simulation based on OpenMP, but does not by design support optimistic synchronization which can, depending on the simulation model, be significantly faster.
+Adevs relies heavily on templates (as do we), but does not hide the types itself.
+%For example, the message communication between models is explicitly typed on payload, this type fixes all usage of a message inside and outside the simulation kernel.
+The same argument can be made for the states.
+%mention differing payload types?
+% in this section or is this to implementation bound ?
+\subsection*{}
 %TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
+The usage of the DirectConnect algorithm makes reusing the adevs kernel impossible.
+The dxexmachina kernel follows the design of the PythonPDEVS kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (among others memory allocation strategy).
+Adevs's explicit usage of (payload)typed messages allows for a fast stack-allocated exchange, whereas dxexmachina uses pointers to heap allocated object (with a runtime disadvantage in some simulations).
+It does give the user the freedom to wrap any type in a message without the type being fixed for all messages.
+% incomplete

### 2015-11-18 paper/2-background.tex

@@
-%TODO say that you are briefly going to go over all necessary background information first
-\subsection{Parallel DEVS}
-%TODO briefly explain Parallel DEVS again, just recycle from other papers
+In this section, we provide a brief introduction to two different synchronization protocols for parallel simulation, and the features offered by C++11 that aid in our implementation.
+
 \subsection{Conservative Synchronization}
 %TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
+Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
+This requires the concepts of eot (earliest output time) and eit (earliest input time) which define the timespan within which a model can safely advance. \\
+No model can advance beyond its own eit, which is defined as the min of all eot values of influencing models.
+An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime.
+A model that can influence another, but never does, can severely slow down the protocol.\\
+Deadlock between models that influence each other can be broken by a variety of
+% link to Chandy/Misrah 79 ? for null messages
+means, in this simulator the null-message protocol derived from (Chandy/Misrah 79) is used to avoid and break deadlock.
 \subsection{Optimistic Synchronization}
 %TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
+Optimistic synchronization allows simulation to advance in time until it receives a message with timestamp in its past. At receipt of such a message, the simulation is
+% needs rewording
+rolled back to the time of the received message to undo the causality errors.
+Optimistic therefore only uses the runtime 'influenced by' relationship, it will only react to actual sent messages between models.
+The cost for this speedup is memory, saving all states until a kernel can verify that all other kernels have advanced equally far.
+\subsection{Global Virtual Time}
+To avoid exhausting memory in state saving, optimistic synchronization relies on the concept of global virtual time, a timepoint that all kernels have simulated past without having to revert before it.
+While this is critical in optimistic to garbage collect old states (and/or messages), usually a conservative kernel will employ GVT as well for the same purpose (but with less memory pressure).
+In conservative synchronization, the null messages can be used to determine GVT while in optimistic this is far more complex.
+% More or less about GVT here ? or move to implementation ?
+
 \subsection{C++11 Parallelism Features}
 %TODO explain what is new in C++11 for parallelism and how it contributes to our cause
+C++11 offers a wide range of portable synchronization primitives in the Standard Library, whereas in earlier versions one had to resort to non-portable (C) implementations.
+More importantly, C++11 is the first version of the standard that actually defines a multi-threaded abstract machine memory model, in earlier versions there was by default no standard way to write a multi-threaded application.
+% Drop the last subsentence ? I suppose we're not selling C++...
+ Our kernels use a wide range of threading primitives and atomic operations tailored for each use case.
+As an example, eot/eit/nulltime are exchanged not as messages but as writes to atomic fields shared by all kernels.
+Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform.
+%minus all the dependencies :)
+% Reference C++11 standard (paying version or draft?)

### 2015-11-18 paper/3-features.tex

@@
 %TODO say that it is based on PythonPDEVS
+% link reference
+The simulator is based on PythonPDEVS, and provides the following features: 
+\begin{enumerate}
+	\item Direct Connect
+	\item Dynamic Structured DEVS
+	\item Termination function. (Only) In single-threaded simulation a termination function, if specified, is applied every simulation round to each model. If this is not present, there is no overhead.
+	\item State/Message can have any payload type.
+	\item Tracing An asynchronous Tracing mechanism allows exact verification of the simulation.
+\end{enumerate}
+The implementation tries to adhere to the C++ principle that you don't pay for what you don't use.
+The termination function is non-trivial to implement in a multi-threaded kernel and had a non-negligible impact on the runtime even if not in use, for this reason it was abandoned.\\
+The tracing is not comparable with adevs's listener interface.
+To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points.
+Furthermore, the framework itself has to be threadsafe and deterministic so that a simulation will always produce the exact same output.
+The following features from PythonPDEVS are not present
+\begin{enumerate}
+	\item Activity tracking and relocation
+	\item Serialization
+	\item Interaction 
+	\item Distributed simulation
+	\item Interactive control
+\end{enumerate}
+Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
+Model allocation is done by a derivable allocator object which the user can  implement to arrange a more ideal (domain-specific) allocation.
+If this is omitted, a default (non-activity-aware) allocation stripes the models over the cores.\\
+Debugging tools are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
+PythonPDEVS can be controlled from within a Python script, this is not implemented in dxexmachina.
+
 %TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
@@
 %TODO explain rationale: why different synchronization protocols
+% weak scenarios for C, O and both ?
+\subsubsection{Conservative}
+% This needs better explaining, perhaps a central graph to reference to ?
+A conservative kernel will determine which kernels it is influenced by, this information is constructed from the incoming connections on all hosted models.
+The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
+A model should provide a lookahead function which returns, relative from the current time, the timespan during which the model cannot change state due to an external event.
+Internal transitions are safe.
+This information is collected for all models a kernel drives, and the minimum is set as the lookahead of the kernel itself.
+\\
+The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the eot of all influencing kernels. \\
+For garbage collection (of sent messages) the GVT is calculated as $min$(nulltime[i] $\forall i$) $ - \epsilon $.
+This is one of the operations that can benefit from using a relaxed memory ordering.
+% Using it for read atm, check before finalizing and do we need to mention this ?.
+
+\subsubsection{Optimistic}
+The optimistic kernel requires from a hosted model only that copying the state is well-defined, which is provided in the base State class for the user.%Mention typed state here ?
+The kernels use Mattern's %ref here
+GVT algorithm with a max 2 rounds per iteration to determine a GVT.
+This process runs asynchronously from the simulation itself.
+Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states/(anti)messages.
+
 %TODO explain core infrastructure a bit, maybe with a diagram or so, which clearly shows that everything is shared, except the "synchronization" part
 %TODO claim that this allows you to easily switch between both methods, to always chose the best one
+The user need only provide one implementation of a model for use with both synchronization protocols.
+A lookahead function is desired to accelerate conservative, but is not required.
+In the absence of a user supplied lookahead, the kernel assumes it cannot predict beyond its current time + $\epsilon$, creating a lockstep simulation.
+% implementation detail ?
+\\
+The implementation details such as defining the copy semantics of a State are provided (but can be overridden). \\
+From the user's perspective, the multi-threaded aspect of the kernel is not exposed. % payload protected ?
+
 \subsection{Performance Improvements}
@@
 %TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
+Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks. 
+% Maybe put more centrally that we trade memory for speed
+\begin{enumerate}
+	\item Heap allocation
+	\item Strings 
+	\item Scheduling
+	\item Locking
+	\item Smart pointers
+\end{enumerate}
+%integrate this in the list?
+% In any case, this text is what we did, not what we should write so has to be condensed to the interesting parts (who except us likes bit shifting?)
+% Pick what we want from this text and rewrite it
+\subsubsection{Heap}
+A kernel never sends a complete object to another kernel, only a pointer to the object.\\
+The cost of heap allocations was minimized using thread\_local memory pools, combined with the (optional) usage of tcmalloc. % reference here
+This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (iff it can prove it is no longer used).
+Experiments with synchronized pools proved slower than malloc/free itself.
+\\
+Replacing strings (in c++ : heap allocated variable sized arrays) with integer identifiers substantially decreased the runtime.
+This is a tradeoff, the user has more affinity with names than numbers as identifiers, so this translation is done mostly internal.
+A Model can be named using a std::string, but once the model is handed off to the kernel only integer identifiers remain (the same goes for ports, kernels, ...).
+\\
+\subsubsection{Avoiding Smart pointers}
+While an important feature in C++11 in general, our initial usage of smart pointers for some types of objects was misplaced.
+Used across threads the reference counting becomes prohibitively expensive.
+Models will still be held by a smart pointer, as will a kernel, but a message is a raw pointer to compacted memory.
+Ideally a message object can (depending on payload) fit on a single cacheline.\\
+\subsubsection{Locking}
+Locking between kernels uses mostly atomic operations, where we can occasionally leverage memory orderings to only pay for synchronization when we need it.
+Messages are exchanged via a shared set of queues each with a dedicated lock.\\
+On a higher level, we avoid the sending of synchronization messages entirely by writing the content (timestamp) directly into shared memory.
+This alleviates in part the potential delay/latency caused by null/eot messages being mixed with simulation messages in conservative implementations.
+\subsubsection{Schedulers}
+PythonPDEVS has a wide range of scheduler for the user to choose from, with performance of each depending on the simulation type.
+Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before.
+Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified, but never items removed.
+% Rephrase this.

### 2015-11-20 paper/2-background.tex
Update background + bib.

--- 2015-11-10 paper/2-background.tex
@@
-%TODO say that you are briefly going to go over all necessary background information first
-\subsection{Parallel DEVS}
-%TODO briefly explain Parallel DEVS again, just recycle from other papers
+In this section, we provide a brief introduction to two different synchronization protocols for parallel simulation, and the features offered by C++11 that aid in our implementation.
+
 \subsection{Conservative Synchronization}
 %TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
+Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
+This requires the concepts of eot (earliest output time) and eit (earliest input time) which define the timespan within which a model can safely advance.
+\\ The eit of any model is the minimum of all eot values of influencing models.
+A model can simulate up to (but not including) eit, and then has to wait until that value is increased.
+An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime.
+A model that can influence another, but never does, can severely slow down the protocol.\\
+Deadlock between models that influence each other can be broken/avoided by a variety of
+means, in this simulator the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
+Null message time is the timestamp a model is guaranteed to have past in simulation. More precisely, a null message of time t is a guarantee that any ouput with timestamp t is already sent.\\
+In general, the eot/eit/nulltime of a kernel is the mimimum of each of those values for all models in the kernel.\\
+Conservative explicitly relies on information provided by the model creator in the form of lookahead, a relative timespan during which the model is insensitive to outside events.
+This can be non-trivial to calculate, a simulation writer will in general not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\
+The performance is more sensitive to lookahead than it is to cycles in the (static) dependency graph.
 \subsection{Optimistic Synchronization}
 %TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
+Optimistic synchronization allows causality errors to occur but employs a roll-back mechanism to recover from those errors.
+The most common mechanism is the Timewarp \cite{Jefferson:1985:VT:3916.3988} mechanism,
+ whenever a kernel receives an event with timestamp in the kernel's past, the state of the kernel is reverted to that time.
+The gain in runtime this provides is offset by the increase in memory required to keep saved states and (sent) messages.
+Optimistic does not rely on any domain specific information, in contrast to conservative.
+It is only sensitive to runtime use of connections, not the probability that they might occur.
+%reword connections ?
+\\
+If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts worsened if the timestamps of the events match exactly. \\
+This effect can be lessened by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
+\subsection{Global Virtual Time}
+To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988}.
+In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event.
+\\ Intuitively this is the timepoint in the simulation that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
+In conservative the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
+In any parallel implementation the GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .
+
 \subsection{C++11 Parallelism Features}
 %TODO explain what is new in C++11 for parallelism and how it contributes to our cause
+C++11 offers a wide range of portable synchronization primitives in the Standard Library, whereas in earlier versions one had to resort to non-portable (C) implementations.
+More importantly, C++11 is the first version of the standard that actually defines a multi-threaded abstract machine memory model in the language.
+Our kernels use a wide range of threading primitives and atomic operations.
+As an example, eot/eit/nulltime are exchanged not as messages but reads/writes to atomic fields shared by all kernels.
+This avoids the otherwise unavoidable latency penalty by mixing simulation messages with synchronization messages.
+% Antimessages ?
+More / Less here ?
+Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform. %minus all the dependencies :)
+% Reference C++11 standard (paying version or draft?)

### 2015-11-20 paper/1-introduction.tex
Expand/rewrite section1.

--- 2015-11-10 paper/1-introduction.tex
@@
+%? use paragraphs for layout instead of (dense) flowing text ?
 %TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
-%TODO give introduction to the field: parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
+\subsection{DEVS}
+%TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
+The family of DEVS \cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms.
+Of interest in this paper are the 3 key formalisms: Classic \cite{ClassicDEVS}, Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation.
+This project uses the DirectConnect \cite{SymbolicFlattening} algorithm, so from a kernel's perspective only Atomic Models exist in the simulation linked to each other by connected ports.
+
+\subsection{Parallel computing}
+%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
+Parallel execution of a DEVS simulation can lower runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
+While the shared memory parallelism offered by most modern hardware does not lower the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation.
+\\
+
+\subsection{Motivation}
 %TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
+Adevs \cite{adevs} offers a very fast conservative synchronized shared memory DEVS simulator, but no optimistic synchronized variant. 
+The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
+The matured parallelism features of C++11 were used to in this project with the dual aim of writing standard-compliant (and thus portable) code without losing access to powerful low-level threading primitives.
+% may need another pass on this one :(
+
+\subsection{Solution} %for lack of a better word
 %TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
+The usage of the DirectConnect\cite{SymbolicFlattening} algorithm makes reusing the adevs kernel hard.
+The dxexmachina kernel follows the design of the PythonPDEVS kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g.
+memory allocation strategy).
+The core aims of the project are to offer a deterministic simulation kernel where the simulation author is shielded as much as possible from the kernel implementation, without sacrificing performance.
+As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead()).\\
+The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
+\subsection{Time}
+The DEVS formalism has $\mathbb{R}$ as time base, but any implementation has to decide on a (finite) representation of time.
+Dxexmachina kernels can operate on IEEE754 floating point time units, or integral time.
+The latter significantly reduces the possible range of the simulation, but avoids possible approximation errors in floating point.
+Furthermore, the notion of $\epsilon$ as the absolute minimum between time points needs to be established, this is non-trivial for floating point.
+\\
+The select() function, which arbitrates between concurrent events, is implicit in this project by providing any time representation with a secondary causality field.
+
+% incomplete

### 2015-11-20 paper/2-background.tex
Expand/rewrite section1.

@@
 means, in this simulator the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
-Null message time is the timestamp a model is guaranteed to have past in simulation. More precisely, a null message of time t is a guarantee that any ouput with timestamp t is already sent.\\
+Null message time is the timestamp a model is guaranteed to have past in simulation.
+More precisely, a null message of time t is a guarantee that any ouput with timestamp t-$\epsilon$ is already sent.\\
 In general, the eot/eit/nulltime of a kernel is the mimimum of each of those values for all models in the kernel.\\

### 2015-11-20 paper/5-related-work.tex
Add basic sections to 5-Related word.

@@
 %TODO explain about PythonPDEVS, the most related in terms of feature and "legacy" --> now in C++11 and way better performance
+\subsection{PythonPDEVS}
+Dxexmachina is closely related to PythonPDEVS in design and philosophy.
+PythonPDEVS allows anyone who grasps the DEVS formalisms to immediately simulate his/her model without having to consider the kernel implementation.
+C++ implementations cannot hope to match the fast prototype/edit/run cycles provided by PythonPDEVS, although this can be minimized by building the kernels as libraries.
+%maybe warn that if a header is touched, you still have to compile from scratch ?
+(eg.
+TIME_FP)
+Advanced features such as activity based relocation and the performance gains this results in, are still unique to PythonPDEVS. \\
+\subsection{Adevs}
 %TODO explain about adevs, the most related in terms of performance --> we implement more synchronization options and offer better performance because ...
+Adevs's source code is still under active development, allowing for an exact comparison in performance and features.
+It remains in most aspects the fastest simulation engine for the DEVS formalism, but it lacks an optimistic synchronization implementation.
+%Tracing ?
+By virtue of not flattening Coupled Models, performance suffers in increasingly hierarchical models.
+%checkme
+\subsection{CD++}
 %TODO explain about CD++, the most related in terms of synchronization --> we implement everything in a single program, with a non-fragmented code-base
 %TODO make mention of Warped, which is the kernel used in CD++, and why we don't use it
+% Still todo : a feature comparison.
+\subsection{Warped}
+The Warped kernel was not used since we operate explicitly for a shared memory system and wanted to design our kernels using the least amount of overhead possible.
+% Not saying MPI is slow (it isn't), but if we can pass a pointer, we don't need MPI.

### 2015-11-21 paper/3-features.tex
* added tcmalloc reference * rewrote some parts of section 3, will need to change more depending on how much space is left.

@@
 	\item Dynamic Structured DEVS
-	\item Termination function. (Only) In single-threaded simulation a termination function, if specified, is applied every simulation round to each model. If this is not present, there is no overhead.
-	\item State/Message can have any payload type.
-	\item Tracing An asynchronous Tracing mechanism allows exact verification of the simulation.
+	\item Termination function.
+If specified, a termination function is applied every simulation round to each model to test whether the simulation can terminate.
+Only available in single-threaded simulation.
+	\item State/Message can have any payload type. Different message types can be used together within the same simulation.
+	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
 \end{enumerate}
-The implementation tries to adhere to the C++ principle that you don't pay for what you don't use.
-The termination function is non-trivial to implement in a multi-threaded kernel and had a non-negligible impact on the runtime even if not in use, for this reason it was abandoned.\\
+The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use.
+For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.\\
 The tracing is not comparable with adevs's listener interface.
@@
 Furthermore, the framework itself has to be threadsafe and deterministic so that a simulation will always produce the exact same output.
-The following features from PythonPDEVS are not present
-\begin{enumerate}
-	\item Activity tracking and relocation
-	\item Serialization
-	\item Interaction 
-	\item Distributed simulation
-	\item Interactive control
-\end{enumerate}
-Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
+%TODO Decide whether we should list the functionality that is _not_ implemented
+%The following features from PythonPDEVS are not present
+%\begin{enumerate}
+%	\item Activity tracking and relocation
+%	\item Serialization
+%	\item Interaction 
+%	\item Distributed simulation
+%	\item Interactive control
+%\end{enumerate}
+%Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
 Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation.
-If this is omitted, a default (non-activity-aware) allocation stripes the models over the cores.\\
-Debugging tools are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
-PythonPDEVS can be controlled from within a Python script, this is not implemented in dxexmachina.
+If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
+Debugging tools such as a logger and a graph vizualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
+While PythonPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.
 
@@
 % This needs better explaining, perhaps a central graph to reference to ?
-A conservative kernel will determine which kernels it is influenced by, this information is constructed from the incoming connections on all hosted models.
+A conservative kernel will determine which kernels it is influenced by.
+This information is constructed from the incoming connections on all hosted models.
 The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
@@
 Internal transitions are safe.
-This information is collected for all models a kernel drives, and the minimum is set as the lookahead of the kernel itself.
+This information is collected for all models hosted on the kernel, and the minimum is set as the lookahead of that kernel.
 \\
 The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the eot of all influencing kernels. \\
-For garbage collection (of sent messages) the GVT is calculated as $min$(nulltime[i] $\forall i$) $ - \epsilon $.
-This is one of the operations that can benefit from using a relaxed memory ordering.
+For garbage collection (of sent messages) the GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i]\ )  - \epsilon $.
+%This is one of the operations that can benefit from using a relaxed memory ordering.
 % Using it for read atm, check before finalizing and do we need to mention this ?.
@@
 \subsubsection{Optimistic}
-The optimistic kernel requires from a hosted model only that copying the state is well-defined, which is provided in the base State class for the user.%Mention typed state here ?
-The kernels use Mattern's %ref here
-GVT algorithm with a max 2 rounds per iteration to determine a GVT.
+The optimistic kernel requires from the hosted model only that copying the state is well-defined, which is provided in the base State class for the user.%Mention typed state here ?
+The kernels use Mattern's %\cite{} %TODO find out in which paper this algorithm was described
+GVT algorithm with a maximum of 2 rounds per iteration to determine a GVT.
 This process runs asynchronously from the simulation itself.
@@
 A kernel never sends a complete object to another kernel, only a pointer to the object.\\
-The cost of heap allocations was minimized using thread\_local memory pools, combined with the (optional) usage of tcmalloc. % reference here
+The cost of heap allocations was minimized using thread\_local memory pools, combined with the (optional) usage of tcmalloc\cite{tcmalloc}.
 This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (iff it can prove it is no longer used).
@@
 Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before.
-Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified, but never items removed.
-% Rephrase this.
+Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified.
+Items are only updated, never removed.

### 2015-11-21 paper/3-features.tex

@@
 %TODO Decide whether we should list the functionality that is _not_ implemented
-%The following features from PythonPDEVS are not present
-%\begin{enumerate}
-%	\item Activity tracking and relocation
-%	\item Serialization
-%	\item Interaction 
-%	\item Distributed simulation
-%	\item Interactive control
-%\end{enumerate}
-%Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
+The following features from PythonPDEVS are not present
+\begin{enumerate}
+	\item Activity tracking and relocation
+	\item Serialization
+	\item Interaction 
+	\item Distributed simulation
+	\item Interactive control
+\end{enumerate}
+Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
 Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation.

### 2015-11-25 paper/4-performance.tex
Add basic text to performance section. Modify benchmark script.

@@
 %TODO we start with sequential simulation results, because this is still base case
-\subsubsection{CPU Usage}
 %TODO compare CPU time with adevs for some benchmarks
@@
 %TODO (spend some time here, as this is also important)
+\subsubsection{CPU Usage}
+%devstone, faster due to interconnect, Messaging O(n), transitions O(n)(if long enough), mention warm up time for chain.
+% dw10->dw40 ?
+
+% phold
+% Not memory intensive whatsoever, only RNG. speed is same (roughly)
+% vary only endtime (n/s/r useless here), benchmarking only rng
+
+% interconnect
+% explain why graph is N^2 by nature. Show that message overhead / extra memory allocation (vgrind) is key here, with overhead in createMessages severe.
 \subsubsection{Memory Usage}
 %TODO compare memory usage with adevs for some benchmarks
+We measure memory usage as allocated pages at a given time t.
+Note that our usage of pools will always overestimate actual memory in use, but from the OS' point of view (and that of the user), any allocated page is in use.\\
+Our usage of tcmalloc should be taken into account here as well, tcmalloc will only gradually release memory back to the OS. \\
+To have an fair comparison (our copy of) adevs also uses tcmalloc.\\
+Actual profiling of memory usage is done with Valgrind's \cite{Nethercote:2007:VFH:1273442.1250746} massif tool.
+Our preference of allocating more on the heap w.r.t adevs stack usage is equalized %bad word
+in tracking only the total amount of pages, with stack, text and heap memory all  in one measure.
+
 %TODO remark our use of memory pools and such, but don't spend too much space on this
@@
 %TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
+% split: dstone/pqueue are dag's
+\subsubsection{Devstone}
+% Initial.
+The devstone \cite{DEVStone} benchmark is highly hierarchical, as a result adevs suffers a performance hit linear in the depth of the (top) model.
+The atomic models in dxexmachina's version of devstone are allocated in a single chain cut in equisized subchains.
+This significantly reduces the nr of messages to pass between kernels which reduces in part memory pressure (local messages can be destroyed immidately whereas remote messages need to wait for GVT).
+If the nr of models increases, optimistic on fixed timeadvance becomes slower due to drifting.
+% whereas iconnect/phold are complete graphs
+\subsubsection{PHold}
+%Phold
+In Phold \cite{PHOLD} Allocation is specified in the benchmark itself, each kernel manages a single node with varying subnodes.\\
+The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative.
+Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$.
+Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation.\\
+Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible.
+
+\subsubsection{Interconnect}
+In Interconnect % no paper ? from PyPDevs originally ? 
+a set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
+Allocation is irrelevant, the resulting dependency graph between kernels still is a complete graph.
+The only difference now is that the dynamic graph will converge very fast to the static graph.
+% for lack of better description.
+Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is well defined and is equal to the timespan between transitions.
+Since all kernels have to wait on each other here, this simulation scales very poorly in nr of kernels.
+Our optimistic implementation does not complete an instance of this benchmark.
+The kernels get stuck in an infinite cascade of reverts.
+If kernel A reverts, it will send antimessages to all others who in turn revert and send antimessages to all others.
+Support for lazy cancellation could undo this anti-pattern.
+\subsubsection{Strengths-Weaknesses} % rename me
+% Conservative : dag-cyclic, cyclic combined with la=eps are worst cases. Contention explodes.
+% Optimistic : Same as conservative, but for dynamic graph. Fixed and/or colliding transition times are key reason for cascading reverts. Memory can explode.
+% Optimistc : sensitive to kernels drifting in time.
+Dependent kernels should move at same time, each revert increases drift (which is not a problem for sink/leaf kernels, else it massively increases nr of reverts.
+% reverts also prohibit GVT from advancing (not in general, in path cases), triggering exploding memory pfaults etc.
+\subsubsection{Priority network model}
 %TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
+The priority benchmark is composed of a single server generating a stream of messages at fixed time intervals, interleaved with a probability p for a priority messages.
+\\ The priority messages defaults lookahead to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph.
+This model therefore highlights the basic strengths/weaknesses of both synchronization protocols.
+Receiving models are allocated on another kernel than the server, and have a internal transition so will not wait for the incoming messages.
+The number of addressed receivers m is variable, with addressed receivers randomly selected.
+\includegraphics[scale=0.4]{pqueue.png}
+% Priority queue: 
+% Conservative : min_lookahead is O(n), still quite good performance (no memory explosion).
+% Optimistic : m is key, each message is a revert.
+%
+
 %TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)

### 2015-11-25 doBenchmark.py

@@ -97,10 +97,10 @@ def devstonegen(simtype, executable, doRandom=False):
 
 randdevstonegen = partial(devstonegen, doRandom=True)
 
-
+# Cores must match nodes.
 def pholdgen(simtype, executable):
-    for apn in [10]:
-        for nodes in [10, 20, 30, 40, 50, 60, 70]:  # , 8, 16, 32]:
+    for nodes in [args.cores]:
+        for apn in [10, 20, 30, 40, 50, 60, 70]:  # , 8, 16, 32]:
             for iterations in [0]:
                 for remotes in [10]:
                     for endTime in [args.endtime]:

### 2015-11-25 paper/4-performance.tex
Explain performance degradation phold, switch to .tikz for pqueue model.

@@
 %Phold
-In Phold \cite{PHOLD} Allocation is specified in the benchmark itself, each kernel manages a single node with varying subnodes.\\
+In Phold \cite{PHOLD} Allocation is specified in the benchmark itself, each kernel manages a single node with varying subnodes.
+The parameter R determines the percentage of remote destination models.\\
 The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative.
 Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$.
-Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation.\\
+Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation.
+For N kernels, each kernel has to query the nulltime of N-1 kernels, resulting in O($N^2$) polling behaviour.
+This benchmark highlights therefore the price that we pay for sharing those values instead of sending an actual null message.
+In a non-cyclic simulation with a non-trivial lookahead, that choice however does pay off.\\
 Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible.
+More seriously, a revert is very expensive in PHold due to our usages of C++11 's random nr generators.
+The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages.
+Again, this could be significantly reduced using lazy-cancellation.
+Once a revert happens the drift between the kernels increases fast, increasing the likelyhood of more reverts.
 
@@
 a set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
-Allocation is irrelevant, the resulting dependency graph between kernels still is a complete graph.
+Allocation is irrelevant, the resulting dependency graph between kernels remains  a complete graph.
 The only difference now is that the dynamic graph will converge very fast to the static graph.
@@
 Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is well defined and is equal to the timespan between transitions.
-Since all kernels have to wait on each other here, this simulation scales very poorly in nr of kernels.
+The scaling issue is identical as in PHold.\\
 Our optimistic implementation does not complete an instance of this benchmark.
@@
 The number of addressed receivers m is variable, with addressed receivers randomly selected.
-\includegraphics[scale=0.4]{pqueue.png}
+%\includegraphics[scale=0.4]{pqueue.png}
+\input pqueue.tex
 % Priority queue: 
@@
 % 
+A key difference here is that a state (in the Receiver instances) is very cheap to copy/create, the random nr generation is done a kernel which will never revert since it is a source in the dependency graph.
+Optimistic will therefore not suffer the same issue as it does in PHold.
+Furthermore, there is no cycle in the graph so a revert cannot cascade in optimistic, and contention is minimal in conservative.
 

### 2015-11-25 paper/1-introduction.tex
Tweak pqueue image, rewrite/reword a few subsections.

@@
 Parallel execution of a DEVS simulation can lower runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
-While the shared memory parallelism offered by most modern hardware does not lower the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation.
+While the shared memory parallelism offered by most modern hardware does not increase the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation.
 \\
@@
 The usage of the DirectConnect\cite{SymbolicFlattening} algorithm makes reusing the adevs kernel hard.
-The dxexmachina kernel follows the design of the PythonPDEVS kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g.
+The dxexmachina kernel follows the design of the PythonPDEVS \cite{PythonPDEVS} kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g.
 memory allocation strategy).
 The core aims of the project are to offer a deterministic simulation kernel where the simulation author is shielded as much as possible from the kernel implementation, without sacrificing performance.
-As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead()).\\
+As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
 The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
@@
 The DEVS formalism has $\mathbb{R}$ as time base, but any implementation has to decide on a (finite) representation of time.
-Dxexmachina kernels can operate on IEEE754 floating point time units, or integral time.
-The latter significantly reduces the possible range of the simulation, but avoids possible approximation errors in floating point.
-Furthermore, the notion of $\epsilon$ as the absolute minimum between time points needs to be established, this is non-trivial for floating point.
-\\
-The select() function, which arbitrates between concurrent events, is implicit in this project by providing any time representation with a secondary causality field.
-
-% incomplete
+Dxexmachina kernels can operate on IEEE754 floating point time units or integral time.
+In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.%ref to time paper.
+An integral representation significantly reduces the possible range of the virtual time, but avoids approximation errors.
+Furthermore, the notion of $\epsilon$ as the absolute minimum between time points needs to be established, this is non-trivial for floating point.\\
+The select() function, which imposes a sequential ordering between concurrent events, is implicit in this project by extending time representation with a secondary causality field.
+If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$.
+This avoids evaluating the select() function on all imminent models, while still maintaining the deterministic order of concurrent transitions.

### 2015-11-25 paper/3-features.tex
Tweak pqueue image, rewrite/reword a few subsections.

@@
 %TODO say that it is based on PythonPDEVS
-% link reference
-The simulator is based on PythonPDEVS, and provides the following features: 
+The simulator is based on PythonPDEVS \cite{PythonPDEVS}  , and provides the following features: 
 \begin{enumerate}
-	\item Direct Connect
+	\item Direct Connect \cite{SymbolicFlattening}
 	\item Dynamic Structured DEVS
@@
 Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks. 
-% Maybe put more centrally that we trade memory for speed
-\begin{enumerate}
-	\item Heap allocation
-	\item Strings 
-	\item Scheduling
-	\item Locking
-	\item Smart pointers
-\end{enumerate}
-%integrate this in the list?
-% In any case, this text is what we did, not what we should write so has to be condensed to the interesting parts (who except us likes bit shifting?)
-% Pick what we want from this text and rewrite it
 \subsubsection{Heap}
-A kernel never sends a complete object to another kernel, only a pointer to the object.\\
-The cost of heap allocations was minimized using thread\_local memory pools, combined with the (optional) usage of tcmalloc\cite{tcmalloc}.
+A kernel never sends a complete object to another kernel, only a pointer to the object. This avoids a possibly expensive copy of the payload, but at the cost of allocation overhead.\\
+This cost becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and messages, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}.
+In this way allocating threads do not block each other, and in a single threaded kernel we can leverage arena-style pools if desired.\\
 This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (iff it can prove it is no longer used).
-Experiments with synchronized pools proved slower than malloc/free itself.
+Experiments with synchronized pools proved slower than malloc/free.
 \\
-Replacing strings (in c++ : heap allocated variable sized arrays) with integer identifiers substantially decreased the runtime.
-This is a tradeoff, the user has more affinity with names than numbers as identifiers, so this translation is done mostly internal.
-A Model can be named using a std::string, but once the model is handed off to the kernel only integer identifiers remain (the same goes for ports, kernels, ...).
+Initially the kernels used strings as identifiers, as is done in PyPDEVS, profiling quickly indicated this to be a performance bottleneck.
+C++ strings are heap allocated variable sized objects with a reference count.
+Access of that reference count across threads is expensive.
+The string implementation also calls malloc/free which are quite expensive.
 \\
-\subsubsection{Avoiding Smart pointers}
-While an important feature in C++11 in general, our initial usage of smart pointers for some types of objects was misplaced.
-Used across threads the reference counting becomes prohibitively expensive.
-Models will still be held by a smart pointer, as will a kernel, but a message is a raw pointer to compacted memory.
-Ideally a message object can (depending on payload) fit on a single cacheline.\\
+String are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name (usually when constructing the model).
+Once simulation starts all objects use integral identifiers for performance.
+\\
+\subsubsection{Smart pointers}
+While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced.
+Used across threads the reference counting becomes prohibitively expensive, and the (de)allocating caused significant contention between threads.
+Models are still held by a smart pointer, as is a kernel, but a message is a raw pointer to compacted memory.
 \subsubsection{Locking}
@@
 On a higher level, we avoid the sending of synchronization messages entirely by writing the content (timestamp) directly into shared memory.
-This alleviates in part the potential delay/latency caused by null/eot messages being mixed with simulation messages in conservative implementations.
+This alleviates in part the potential delay/latency caused by interleaving simulation with null/eot messages in conservative implementations.
+Optimistic still sends anti-messages as pointers, since the complete header of the message is required at the receiver side to annihilate the original.
 \subsubsection{Schedulers}
@@
 Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before.
-Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified.
+Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depening on the nr of items to update.
 Items are only updated, never removed.

### 2015-11-28 paper/4-performance.tex
Fill in section on memory usage, add/change sections on specific benchmarks. Apply Stijn's patch.

--- 2015-11-10 paper/4-performance.tex
@@
 %TODO (spend some time here, as this is also important)
+\subsubsection{CPU Usage}
+\paragraph*{Devstone}
+The Devstone \cite{DEVStone} benchmark is highly hierarchical, using directconnect the dxexmachina kernels can exploit this whereas adevs needs to walk the structure of the model to pass events.
+% TODO plot here
+\paragraph*{PHold}
+PHold \cite{PHOLD} is a parallel oriented benchmark, sequential runtime is measured only as a baseline. The usage of random nr generators takes up a significant amount of runtime in this model.
+% TODO plot here
+\paragraph*{Interconnect}
+Interconnect is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models.
+It highlights in sequential simulaton the cost of heap allocation (messages) in dxexmachina.
+As the benchmark increases the model count, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxexmachina w.r.t.
+adevs.
+This effect is due to the heap allocation of the messages by dxexmachina, even though this is minimized using memory pools it remains a significant effect.
+% TODO plot here
 \subsubsection{Memory Usage}
 %TODO compare memory usage with adevs for some benchmarks
-%TODO remark our use of memory pools and such, but don't spend too much space on this
+\paragraph*{Platform and tools}
+Both dxexmachina and adevs use tcmalloc as memory allocator, in addition dxexmachina uses memory pools to further reduce the frequency of expensive syscalls (malloc/free/sbrk/mmap/...).
+Tcmalloc will only gradually release memory back to the OS, whereas our pools will not do so at all.
+If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool.
+For this reason a memory utilization can be best measured using peak allocation.
+Profiling is done using Valgrind's massif tool \cite{Nethercote:2007:VFH:1273442.1250746}
+Platform used was a i5-3317U Intel cpu, 8GB RAM, Fedora 22 (kernel 4.2.6), a page is 4,096KiB.\\
+\paragraph*{Measure}
+Adevs passes messages by value (in container by reference), we pass by pointer.
+The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section we measure memory usage in number of allocated pages.
+This combines text, stack and heap memory for the program profiled, from the point of view of the OS or user, this is the actual memory in use.
+It is important to note that, especially in the case of optimistic, not all this memory is in use by the kernel, since the pools will in general not return memory once it is allocated but keep it for later reuse.
+\paragraph*{Results}
+\paragraph*{Devstone}
+\begin{table}[lhtb]
+	\centering
+	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
+	\label{dtone_mem}
+	\begin{tabular}{lllll}
+		Adevs & AdevsCon &DX &DXCon &DXOpt\\
+		44 & 70 & 42 & 75 & 363  \\
+	\end{tabular}
+\end{table}
+Since dx's conservative passes messages by pointer, it needs a GVT implementation to organise garbage collection, this inevitable delay explains the higher memory usage w.r.t adevs.\\
+Optimistic needs a far more complex GVT/garbage collection algorithm which explains in part the massive amount of memory used.
+Devstone (flattened) is allocated in a chain, this means that the leafs in the dependency graph will do a lot of unnecessary simulation before having a revert, leading to quite severe memory pressure.
+Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels.
+\paragraph*{Interconnect}
+In section 4.2 the parallel performance of this benchmark is further explained.
+It is of interest for sequential to contrast with the runtime penalty that dxexmachina suffers w.r.t.
+adevs.
+Optimistic fails this benchmark, see for detailed analysis section 4.2.3.
+\begin{table}[lhtb]
+	\centering
+	\caption{Interconnect w 40 t5e5, unit MiB, 2 kernels (if parallel)}
+	\label{iconn_mem}
+	\begin{tabular}{lllll}
+		Adevs & AdevsCon &DX &DXCon\\
+		39 & 39 & 35 & 52  \\
+	\end{tabular}
+\end{table}
+The discrepancy between both conservative implementations is detailed in 4.1.4.
+
+\paragraph*{Priority Network}
+The priority network model is detailed in section 4.4.
+\begin{table}[lhtb]
+	\centering
+	\caption{Priority model n 16, m 9, p 10,  t5e5, unit MiB, 2 kernels}
+	\label{pmod_mem}
+	\begin{tabular}{lll}
+		DX &DXCon &DXOpt\\
+		35 & 58 &69\\
+	\end{tabular}
+\end{table}
+
 \subsection{Parallel Simulation}
@@
 %TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
+% split: dstone/pqueue are dag's
+\subsubsection{Devstone}
+% Initial.
+The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages.
+For optimistic this can cause more reverts since the kernels will start to drift faster as the modelcount increases.
+Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases.
+Of note as well is the warm-up time this benchmark requires, for n=dxw models, it takes timeadvance()*n transitions to activate the last model in the chain.
+For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.
+\subsubsection{PHold}
+%Phold
+In Phold \cite{PHOLD} Allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of subnodes.
+The parameter R determines the percentage of remote destination models.\\
+The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative.
+Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$.
+Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation.
+For N kernels, each kernel has to query the nulltime of N-1 kernels, resulting in O($N^2$) polling behaviour.
+This benchmark highlights therefore the price that we pay for sharing those values instead of sending an actual null message.
+In a non-cyclic simulation with a non-trivial lookahead, that choice however does pay off.\\
+Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible.
+More seriously, a revert is very expensive in PHold due to our usages of C++11 's random nr generators.
+The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages.
+Again, this could be significantly reduced using lazy-cancellation.
+Once a revert happens the drift between the kernels increases fast, increasing the likelyhood of more reverts.
+
+\subsubsection{Interconnect}
+In Interconnect % no paper ? from PyPDevs originally ? 
+the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
+Allocation is irrelevant, the resulting dependency graph between kernels remains  a complete graph. The runtime dependency graph is almost immediately equal to the static dependency graph.
+Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions.
+The scaling issue is identical as in PHold.\\
+Our optimistic implementation does not complete an instance of this benchmark.
+The kernels get stuck in an infinite cascade of reverts.
+If kernel A reverts, it will send antimessages to all others who in turn revert and send antimessages to all others (and A itself again).
+Support for lazy cancellation could potentially undo this anti-pattern.
+\subsubsection{Priority network model}
 %TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
+The priority benchmark is composed of a single server generating a stream of $0<=m<=n$ messages at fixed time intervals, interleaved with a probability p for a priority message, to n receivers.
+\\ The priority messages defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph.
+This model therefore highlights the basic strengths/weaknesses of both synchronization protocols.
+Receiving models are allocated on another kernel than the server, and have a internal transition so will not wait for the incoming messages.
+\input pqueue.tex
+A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create, once the random nr generation is done the kernel holding the server will never revert since it is a source in the dependency graph.
+Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold.
+
 %TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)

### 2015-11-28 main/src/mainphold.cpp

@@ -11,7 +11,7 @@
 
 using namespace n_tools;
 
-LOG_INIT("phold.log");
+LOG_INIT("phold.log")
 
 const char helpstr[] = " [-h] [-t ENDTIME] [-n NODES] [-s SUBNODES] [-r REMOTES] [-i ITER] [-c COREAMT] [classic|cpdevs|opdevs|pdevs]\n"
 	"options:\n"

### 2015-11-28 main/src/model/atomicmodel.cpp

@@ -28,6 +28,7 @@ AtomicModel_impl::AtomicModel_impl(std::string name, int corenumber, std::size_t
 
 AtomicModel_impl::~AtomicModel_impl()
 {
+        
         delete m_state;
 }
 

### 2015-11-28 main/src/model/core.cpp

@@ -617,6 +617,7 @@ void n_model::Core::getPendingMail()
 	        auto msg = entry.getMessage();
 	        if(msg->flagIsSet(Status::ERASE)){
 	                msg->setFlag(Status::KILL);
+                        continue;
 	        }
                 msg->setFlag(Status::HEAPED, false);
                 queueLocalMessage(msg);

### 2015-11-28 main/src/model/optimisticcore.cpp

@@ -89,6 +89,7 @@ void Optimisticcore::clearProcessedMessages(std::vector<t_msgptr>& msgs)
                         ptr->releaseMe();
                 }
                 else{
+                        LOG_DEBUG("MCORE:: ", this->getCoreID(), "@", this->getTime(), " Putting " , ptr->toString(), " in processed. &", ptr );
                         m_processed_messages.push_back(n_network::hazard_pointer(ptr));
                 }
         }

### 2015-11-28 paper/1-introduction.tex
Rewrite/fixes in several sections. Initial section 6.

@@
 %TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
-Parallel execution of a DEVS simulation can lower runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
+Parallel execution of a PDEVS simulation can lower runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
 While the shared memory parallelism offered by most modern hardware does not increase the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation.
@@
 \subsection{Time}
-The DEVS formalism has $\mathbb{R}$ as time base, but any implementation has to decide on a (finite) representation of time.
+The PDEVS formalism has $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time.
 Dxexmachina kernels can operate on IEEE754 floating point time units or integral time.
@@
 Furthermore, the notion of $\epsilon$ as the absolute minimum between time points needs to be established, this is non-trivial for floating point.\\
-The select() function, which imposes a sequential ordering between concurrent events, is implicit in this project by extending time representation with a secondary causality field.
+The select() function, which imposes a sequential ordering between concurrent events, is implicit in this project by extending time representation with a causality field with a range at least as large as the maximum nr of models in the kernels ($2^{24}$ by default).
 If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$.

### 2015-11-28 paper/2-background.tex
Rewrite/fixes in several sections. Initial section 6.

--- 2015-11-20 paper/2-background.tex
@@
 This requires the concepts of eot (earliest output time) and eit (earliest input time) which define the timespan within which a model can safely advance.
-\\ The eit of any model is the minimum of all eot values of influencing models.
-A model can simulate up to (but not including) eit, and then has to wait until that value is increased.
+\\ The eit of any model is the minimum of all eot values of (directly) influencing models.
+A model can simulate up to (but not including) eit, then waits until that value is increased.
 An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime.
 A model that can influence another, but never does, can severely slow down the protocol.\\
-Deadlock between models that influence each other can be broken/avoided by a variety of
+Deadlock between models that influence each other and end up waiting on each other can be broken/avoided by a variety of
 means, in this simulator the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
-Null message time is the timestamp a model is guaranteed to have past in simulation. More precisely, a null message of time t is a guarantee that any ouput with timestamp t is already sent.\\
+In our implementaton null-time is the timestamp a model is guaranteed to have passed in simulation.
+More precisely, a null message of time t is a guarantee that any ouput with timestamp t-$\epsilon$ is already sent.\\
 In general, the eot/eit/nulltime of a kernel is the mimimum of each of those values for all models in the kernel.\\
-Conservative explicitly relies on information provided by the model creator in the form of lookahead, a relative timespan during which the model is insensitive to outside events.
-This can be non-trivial to calculate, a simulation writer will in general not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\
-The performance is more sensitive to lookahead than it is to cycles in the (static) dependency graph.
+Conservative synchronization explicitly relies on information provided by the model creator in the form of lookahead, a relative timespan during which the model is insensitive to outside events.
+This can be non-trivial to calculate, a simulation writer will in general not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
 \subsection{Optimistic Synchronization}
 %TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
-Optimistic synchronization allows causality errors to occur but employs a roll-back mechanism to recover from those errors.
-The most common mechanism is the Timewarp \cite{Jefferson:1985:VT:3916.3988} mechanism,
- whenever a kernel receives an event with timestamp in the kernel's past, the state of the kernel is reverted to that time.
+Optimistic synchronization allows causality errors to occur but recovers from those errors using a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.
+Whenever a kernel receives an event with a timestamp in the kernel's past, the state of the kernel (and all models) is reverted to that time.
 The gain in runtime this provides is offset by the increase in memory required to keep saved states and (sent) messages.
@@
 \\
-If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts worsened if the timestamps of the events match exactly. \\
+If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts.
+Without domain specific information the kernel assumes that any event will influence at least one model, but this can lead to an infinite loop of reverts in the worst case.\\
 This effect can be lessened by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
@@
 In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event.
-\\ Intuitively this is the timepoint in the simulation that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
+\\ Intuitively this is the simulation timepoint that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
 In conservative the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
-In any parallel implementation the GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .
+The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .
 
@@
 As an example, eot/eit/nulltime are exchanged not as messages but reads/writes to atomic fields shared by all kernels.
-This avoids the otherwise unavoidable latency penalty by mixing simulation messages with synchronization messages.
-% Antimessages ?
-More / Less here ?
-Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform. %minus all the dependencies :)
-% Reference C++11 standard (paying version or draft?)
+This avoids the otherwise unavoidable latency penalty by mixing simulation messages with synchronization messages, for an in-depth study see \cite{CPE:CPE3007}.
+Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform.

### 2015-11-28 paper/3-features.tex
Rewrite/fixes in several sections. Initial section 6.

@@
 	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
+	\item Optimistic and Conservative synchronization of PDEVS.
 \end{enumerate}
@@
 	\item Serialization
-	\item Interaction 
+	\item Interactive control. 
 	\item Distributed simulation
-	\item Interactive control
 \end{enumerate}
@@
 If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
-Debugging tools such as a logger and a graph vizualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
+Debugging tools such as a logger and a graph visualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
 While PythonPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.
@@
 A model should provide a lookahead function which returns, relative from the current time, the timespan during which the model cannot change state due to an external event.
-Internal transitions are safe.
 This information is collected for all models hosted on the kernel, and the minimum is set as the lookahead of that kernel.
 \\
-The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the eot of all influencing kernels. \\
-For garbage collection (of sent messages) the GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i]\ )  - \epsilon $.
+The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
+For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i]\ )  - \epsilon $.
 %This is one of the operations that can benefit from using a relaxed memory ordering.
@@
 \subsubsection{Optimistic}
-The optimistic kernel requires from the hosted model only that copying the state is well-defined, which is provided in the base State class for the user.%Mention typed state here ?
-The kernels use Mattern's %\cite{} %TODO find out in which paper this algorithm was described
+The optimistic kernel requires from the hosted model only that copying the state is well-defined, which is provided in the base State class for the user. 
+The kernels use Mattern's \cite{mattern}
 GVT algorithm with a maximum of 2 rounds per iteration to determine a GVT.
@@
 %TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
-Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks. 
+Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks: 
 \subsubsection{Heap}
@@
 Initially the kernels used strings as identifiers, as is done in PyPDEVS, profiling quickly indicated this to be a performance bottleneck.
-C++ strings are heap allocated variable sized objects with a reference count.
-Access of that reference count across threads is expensive.
-The string implementation also calls malloc/free which are quite expensive.
-\\
-String are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name (usually when constructing the model).
+C++ strings are heap allocated variable sized objects with an atomic reference count.
+Access of that reference count across threads is expensive, as are the calls to malloc/free the string implementation makes to create/destroy new object, or copy existing.\\
+Strings are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name (usually when constructing the model).
 Once simulation starts all objects use integral identifiers for performance.
+This also increased usage of the constexpr feature of C++11 in, amongst others, timestamps and message headers.
 \\
@@
 Messages are exchanged via a shared set of queues each with a dedicated lock.\\
-On a higher level, we avoid the sending of synchronization messages entirely by writing the content (timestamp) directly into shared memory.
-This alleviates in part the potential delay/latency caused by interleaving simulation with null/eot messages in conservative implementations.
-Optimistic still sends anti-messages as pointers, since the complete header of the message is required at the receiver side to annihilate the original.
+On a higher level, we avoid the sending of synchronization messages entirely by writing the timestamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
 \subsubsection{Schedulers}
-PythonPDEVS has a wide range of scheduler for the user to choose from, with performance of each depending on the simulation type.
+PythonPDEVS has a wide range of schedulers for the user to choose from, with performance of each depending on the simulation type.
 Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before.
-Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depening on the nr of items to update.
+Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depending on the nr of items to update.
 Items are only updated, never removed.

### 2015-11-28 paper/4-performance.tex
Rewrite/fixes in several sections. Initial section 6.

@@
 \paragraph*{Interconnect}
-Interconnect is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models.
-It highlights in sequential simulaton the cost of heap allocation (messages) in dxexmachina.
-As the benchmark increases the model count, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxexmachina w.r.t.
+Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models.
+It highlights in sequential simulation the cost of heap allocation (messages) in dxexmachina.
+As the model count increases, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxexmachina w.r.t.
 adevs.
-This effect is due to the heap allocation of the messages by dxexmachina, even though this is minimized using memory pools it remains a significant effect.
+This effect is due to the heap allocation of the messages by dxexmachina, even though this is minimized using memory pools it remains significant.
 % TODO plot here
@@
 \end{table}
-Since dx's conservative passes messages by pointer, it needs a GVT implementation to organise garbage collection, this inevitable delay explains the higher memory usage w.r.t adevs.\\
-Optimistic needs a far more complex GVT/garbage collection algorithm which explains in part the massive amount of memory used.
+Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection, this inevitable delay explains the higher memory usage w.r.t adevs.\\
+Optimistic needs a more complex GVT/garbage collection algorithm plus the differences in LP virtual times are far larger, which explains the heavier memory usage.
 Devstone (flattened) is allocated in a chain, this means that the leafs in the dependency graph will do a lot of unnecessary simulation before having a revert, leading to quite severe memory pressure.
@@
 %Phold
-In Phold \cite{PHOLD} Allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of subnodes.
+In Phold \cite{PHOLD} Allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of sub-nodes.
 The parameter R determines the percentage of remote destination models.\\
@@
 Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation.
-For N kernels, each kernel has to query the nulltime of N-1 kernels, resulting in O($N^2$) polling behaviour.
+For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour.
 This benchmark highlights therefore the price that we pay for sharing those values instead of sending an actual null message.
-In a non-cyclic simulation with a non-trivial lookahead, that choice however does pay off.\\
+In a non-cyclic simulation with a non-trivial lookahead (e.g.
+devstone), that choice however does pay off.\\
 Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible.
-More seriously, a revert is very expensive in PHold due to our usages of C++11 's random nr generators.
+More seriously, a revert is very expensive in PHold due to our usage of C++11's random nr generators.
 The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages.
 Again, this could be significantly reduced using lazy-cancellation.
-Once a revert happens the drift between the kernels increases fast, increasing the likelyhood of more reverts.
+Once a revert happens the drift between the kernels increases fast, increasing the likelihood of more reverts.
 
 \subsubsection{Interconnect}
-In Interconnect % no paper ? from PyPDevs originally ? 
+In Interconnect
 the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
@@
 The kernels get stuck in an infinite cascade of reverts.
-If kernel A reverts, it will send antimessages to all others who in turn revert and send antimessages to all others (and A itself again).
+If kernel A reverts, it will send anti-messages to all others who in turn revert and send anti-messages to all others (and A itself again).
 Support for lazy cancellation could potentially undo this anti-pattern.
@@
 The priority benchmark is composed of a single server generating a stream of $0<=m<=n$ messages at fixed time intervals, interleaved with a probability p for a priority message, to n receivers.
-\\ The priority messages defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph.
+\\ This defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph.
 This model therefore highlights the basic strengths/weaknesses of both synchronization protocols.

### 2015-11-28 paper/6-conclusions.tex
Rewrite/fixes in several sections. Initial section 6.

@@
 %TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
+% needs some extending. Maybe clarify that we actually do single core PDEVS and not
+% pure classic DEVS.
+Both synchronization algorithms offer good performance in differing simulations, in some simulation our kernels outperform adevs whereas in others we can still improve. 
+The optimistic implementation needs to be extended with lazy evaluation/cancellation to function in cyclic simulations.
+\subsection{Future work}
 %TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))
+\subsubsection{Activity}
+As shown in \cite{PythonPDEVS_ACTIMS} activity and allocation of models across kernels is a key aspect in achieving high performance in any parallel implementation.
+Allocating models so that there are no dependency cycles between their containing kernels is a first step, but not always possible.
+For optimistic one can use re-allocation to break (runtime dependency) cycles or perform load balancing.
+If kernels are unevenly balanced they will begin to drift fast, causing increasingly more reverts.
+\subsubsection{Hybrid}
+The optimistic implementation could use (null/eot/eit) from conservative to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.
+Conservative kernels could be extended with runtime information about influencing models, if one can guarantee a static dependency is not used for a fixed time-span, this dependency can be removed for that period of (virtual) time.
+\\
+Ultimately the simulation could switch at runtime between protocols based on the information provided by activity tracking.

### 2015-11-29 paper/5-related-work.tex
small addition to related work, corrections will follow...

@@
 %TODO make mention of Warped, which is the kernel used in CD++, and why we don't use it
-% Still todo : a feature comparison.
-\subsection{Warped}
-The Warped kernel was not used since we operate explicitly for a shared memory system and wanted to design our kernels using the least amount of overhead possible.
-% Not saying MPI is slow (it isn't), but if we can pass a pointer, we don't need MPI.
+Different projects on CD++ offer conservative (CCD++) as well as optimistic (PCD++) parallel simulation.
+In contrast to our single program, with a non-fragmented code-base, neither projects offer both synchronization protocols.
+CD++ relies on the WARPED kernel.
+It is a middleware that provides memory, event, file, time and communication scheduling.
+WARPED is not used here since we operate explicitly on a shared memory system and since we wanted to design our kernels using the least amount of overhead possible.

### 2015-11-29 paper/1-introduction.tex
Small changes to paper

@@
 %TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
-Parallel execution of a PDEVS simulation can lower runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
+Parallel execution of a PDEVS simulation can lower overall runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
 While the shared memory parallelism offered by most modern hardware does not increase the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation.
@@
 The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
-The matured parallelism features of C++11 were used to in this project with the dual aim of writing standard-compliant (and thus portable) code without losing access to powerful low-level threading primitives.
+The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code and without losing access to powerful low-level threading primitives.
 % may need another pass on this one :(

### 2015-11-29 paper/2-background.tex
Small changes to paper

@@
 means, in this simulator the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
-In our implementaton null-time is the timestamp a model is guaranteed to have passed in simulation.
-More precisely, a null message of time t is a guarantee that any ouput with timestamp t-$\epsilon$ is already sent.\\
+In our implementaton, null-time is the timestamp a model is guaranteed to have passed in simulation.
+More precisely, a null message of time t is a guarantee that any output with timestamp t-$\epsilon$ is already sent.\\
 In general, the eot/eit/nulltime of a kernel is the mimimum of each of those values for all models in the kernel.\\
@@
 \\ Intuitively this is the simulation timepoint that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
-In conservative the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
+In conservative, the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
 The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .

### 2015-11-29 paper/3-features.tex
Small changes to paper

@@
 For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.\\
-The tracing is not comparable with adevs's listener interface.
+The tracing is not comparable to adevs's listener interface.
 To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points.
@@
 The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
-For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i]\ )  - \epsilon $.
+For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i])  - \epsilon $.
 %This is one of the operations that can benefit from using a relaxed memory ordering.
@@
 In this way allocating threads do not block each other, and in a single threaded kernel we can leverage arena-style pools if desired.\\
-This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (iff it can prove it is no longer used).
+This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (if it can prove it is no longer used).
 Experiments with synchronized pools proved slower than malloc/free.
@@
 Access of that reference count across threads is expensive, as are the calls to malloc/free the string implementation makes to create/destroy new object, or copy existing.\\
-Strings are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name (usually when constructing the model).
+Strings are more intuitive to work with from a user's standpoint.
+So as a compromise the user can reference models/ports by string name (usually when constructing the model).
 Once simulation starts all objects use integral identifiers for performance.
@@
 \\
-\subsubsection{Smart pointers}
+\subsubsection{Raw pointers}
 While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced.
@@
 PythonPDEVS has a wide range of schedulers for the user to choose from, with performance of each depending on the simulation type.
-Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before.
+Profiling showed in our case that, for a C++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before.
 Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depending on the nr of items to update.

### 2015-11-29 paper/4-performance.tex
Small changes to paper

@@
 Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases.
-Of note as well is the warm-up time this benchmark requires, for n=dxw models, it takes timeadvance()*n transitions to activate the last model in the chain.
+Of note as well is the warm-up time this benchmark requires, for $n=d$x$w$ models, it takes timeadvance()*n transitions to activate the last model in the chain.
 For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.
@@
 %Phold
-In Phold \cite{PHOLD} Allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of sub-nodes.
+In Phold \cite{PHOLD} allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of sub-nodes.
 The parameter R determines the percentage of remote destination models.\\
@@
 For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour.
-This benchmark highlights therefore the price that we pay for sharing those values instead of sending an actual null message.
+This benchmark therefore highlights the price that we pay for sharing those values instead of sending an actual null message.
 In a non-cyclic simulation with a non-trivial lookahead (e.g.
@@
 Receiving models are allocated on another kernel than the server, and have a internal transition so will not wait for the incoming messages.
-\input pqueue.tex
+\input pqueue.tex \\
 A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create, once the random nr generation is done the kernel holding the server will never revert since it is a source in the dependency graph.

### 2015-11-29 paper/1-introduction.tex
Did some reading. Tried to fix some stuff we discussed in issue #28. Also changed the sentence structure in some places.

--- 2015-11-28 paper/1-introduction.tex
@@
 Of interest in this paper are the 3 key formalisms: Classic \cite{ClassicDEVS}, Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation.
-This project uses the DirectConnect \cite{SymbolicFlattening} algorithm, so from a kernel's perspective only Atomic Models exist in the simulation linked to each other by connected ports.
+This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a kernel's perspective only Atomic Models exist in the simulation linked to each other by connected ports.
 
@@
 The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
-The matured parallelism features of C++11 were used to in this project with the dual aim of writing standard-compliant (and thus portable) code without losing access to powerful low-level threading primitives.
+The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code without losing access to powerful low-level threading primitives.
 % may need another pass on this one :(
@@
 %TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
-The usage of the DirectConnect\cite{SymbolicFlattening} algorithm makes reusing the adevs kernel hard.
-The dxexmachina kernel follows the design of the PythonPDEVS \cite{PythonPDEVS} kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g.
+The usage of the Direct connection algorithm makes reusing the adevs kernel hard.
+The devs-ex-machina (dxex) kernel follows the design of the PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g.
 memory allocation strategy).
@@
 \subsection{Time}
-The PDEVS formalism has $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time.
-Dxexmachina kernels can operate on IEEE754 floating point time units or integral time.
+The DEVS formalisms has $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time.
+Dxex kernels can operate on IEEE754 floating point time units or integer time.
 In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.%ref to time paper.
-An integral representation significantly reduces the possible range of the virtual time, but avoids approximation errors.
-Furthermore, the notion of $\epsilon$ as the absolute minimum between time points needs to be established, this is non-trivial for floating point.\\
-The select() function, which imposes a sequential ordering between concurrent events, is implicit in this project by extending time representation with a causality field with a range at least as large as the maximum nr of models in the kernels ($2^{24}$ by default).
+An integer representation significantly reduces the possible range of the virtual time, but avoids approximation errors.
+Furthermore, the notion of $\epsilon$ as a minimum distance between two distinct time points need not to be established.
+This task is non-trivial for floating point approximations.\\
+The select() function, which imposes a sequential ordering between concurrent events, is made implicit in this project by extending time representation with a causality field with a range at least as large as the maximum nr of models in the kernels ($2^{24}$ by default).
 If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$.

### 2015-11-29 paper/2-background.tex
Did some reading. Tried to fix some stuff we discussed in issue #28. Also changed the sentence structure in some places.

--- 2015-11-28 paper/2-background.tex
@@
 Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
-This requires the concepts of eot (earliest output time) and eit (earliest input time) which define the timespan within which a model can safely advance.
+This requires the concepts of earliest output time (eot) and earliest input time (eit), which define the timespan within which a model can safely advance.
 \\ The eit of any model is the minimum of all eot values of (directly) influencing models.
@@
 A model that can influence another, but never does, can severely slow down the protocol.\\
-Deadlock between models that influence each other and end up waiting on each other can be broken/avoided by a variety of
-means, in this simulator the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
-In our implementaton null-time is the timestamp a model is guaranteed to have passed in simulation.
-More precisely, a null message of time t is a guarantee that any ouput with timestamp t-$\epsilon$ is already sent.\\
-In general, the eot/eit/nulltime of a kernel is the mimimum of each of those values for all models in the kernel.\\
-Conservative synchronization explicitly relies on information provided by the model creator in the form of lookahead, a relative timespan during which the model is insensitive to outside events.
-This can be non-trivial to calculate, a simulation writer will in general not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
+Deadlock between models that influence each other and end up waiting on each other can be broken/avoided by a variety of means.
+In this simulator, the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used.
+\\
+In our implementation null-time is the timestamp a model is guaranteed to have passed in simulation.
+More precisely, a null message of time t is a guarantee that any output with timestamp t-$\epsilon$ is already sent.\\
+In general, the eot/eit/nulltime of a kernel is the minimum of each of those values for all models in the kernel.\\
+Conservative synchronization relies explicitly on information provided by the model creator in the form of a lookahead, a relative timespan during which the model is insensitive to outside events.
+This lookahead can be non-trivial to calculate.
+In general, a simulation writer will not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
 \subsection{Optimistic Synchronization}
@@
 Whenever a kernel receives an event with a timestamp in the kernel's past, the state of the kernel (and all models) is reverted to that time.
-The gain in runtime this provides is offset by the increase in memory required to keep saved states and (sent) messages.
-Optimistic does not rely on any domain specific information, in contrast to conservative.
+The possible gain in runtime is offset by the increase in memory required to keep saved states and (sent) messages.
+In contrast to conservative, optimistic does not rely on any domain specific information.
 It is only sensitive to runtime use of connections, not the probability that they might occur.
@@
 If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts.
-Without domain specific information the kernel assumes that any event will influence at least one model, but this can lead to an infinite loop of reverts in the worst case.\\
+Without domain specific information, the kernel assumes that any event will influence at least one model, but this can lead to an infinite loop of reverts in the worst case.\\
 This effect can be lessened by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
@@
 Our kernels use a wide range of threading primitives and atomic operations.
-As an example, eot/eit/nulltime are exchanged not as messages but reads/writes to atomic fields shared by all kernels.
-This avoids the otherwise unavoidable latency penalty by mixing simulation messages with synchronization messages, for an in-depth study see \cite{CPE:CPE3007}.
+As an example, eot/eit/nulltime are exchanged not as messages, but as reads/writes to atomic fields shared by all kernels.
+This avoids the otherwise unavoidable latency penalty by mixing simulation messages with synchronization messages.
+For an in-depth study, see \cite{CPE:CPE3007}.
 Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform.

### 2015-11-29 paper/3-features.tex
Did some reading. Tried to fix some stuff we discussed in issue #28. Also changed the sentence structure in some places.

--- 2015-11-28 paper/3-features.tex
@@
 %TODO say that it is based on PythonPDEVS
-The simulator is based on PythonPDEVS \cite{PythonPDEVS}  , and provides the following features: 
+The simulator is based on PythonPDEVS, and provides the following features: 
 \begin{enumerate}
-	\item Direct Connect \cite{SymbolicFlattening}
+	\item Direct Connection
 	\item Dynamic Structured DEVS
@@
 To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points.
-Furthermore, the framework itself has to be threadsafe and deterministic so that a simulation will always produce the exact same output.
+Furthermore, the framework itself has to be thread safe and deterministic so that a simulation will always produce the exact same output.
 %TODO Decide whether we should list the functionality that is _not_ implemented
-The following features from PythonPDEVS are not present
+The following features from PyPDEVS are not present
 \begin{enumerate}
@@
 Debugging tools such as a logger and a graph visualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
-While PythonPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.
+While PyPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.
 
@@
 \subsubsection{Schedulers}
-PythonPDEVS has a wide range of schedulers for the user to choose from, with performance of each depending on the simulation type.
+PyPDEVS has a wide range of schedulers for the user to choose from, with performance of each depending on the simulation type.
 Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before.
-Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depending on the nr of items to update.
+Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depending on the amount of items to update.
 Items are only updated, never removed.

### 2015-11-29 paper/4-performance.tex
Did some reading. Tried to fix some stuff we discussed in issue #28. Also changed the sentence structure in some places.

--- 2015-11-28 paper/4-performance.tex
@@
 \paragraph*{Devstone}
-The Devstone \cite{DEVStone} benchmark is highly hierarchical, using directconnect the dxexmachina kernels can exploit this whereas adevs needs to walk the structure of the model to pass events.
+The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this, whereas adevs needs to walk the structure of the model to pass events.
 % TODO plot here
@@
 Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models.
-It highlights in sequential simulation the cost of heap allocation (messages) in dxexmachina.
-As the model count increases, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxexmachina w.r.t.
+It highlights in sequential simulation the cost of heap allocation (messages) in dxex.
+As the model count increases, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxex w.r.t.
 adevs.
-This effect is due to the heap allocation of the messages by dxexmachina, even though this is minimized using memory pools it remains significant.
+This effect is due to the heap allocation of the messages by dxex, even though this is minimized using memory pools it remains significant.
 % TODO plot here
@@
 \paragraph*{Platform and tools}
-Both dxexmachina and adevs use tcmalloc as memory allocator, in addition dxexmachina uses memory pools to further reduce the frequency of expensive syscalls (malloc/free/sbrk/mmap/...).
+Both dxex and adevs use tcmalloc as memory allocator.
+Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (malloc/free/sbrk/mmap/...).
 Tcmalloc will only gradually release memory back to the OS, whereas our pools will not do so at all.
 If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool.
-For this reason a memory utilization can be best measured using peak allocation.
+For this reason, the memory utilization can be best measured by peak allocation.
 Profiling is done using Valgrind's massif tool \cite{Nethercote:2007:VFH:1273442.1250746}
-Platform used was a i5-3317U Intel cpu, 8GB RAM, Fedora 22 (kernel 4.2.6), a page is 4,096KiB.\\
+Platform used has a i5-3317U Intel cpu and 8GB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
 \paragraph*{Measure}
 Adevs passes messages by value (in container by reference), we pass by pointer.
-The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section we measure memory usage in number of allocated pages.
+The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages.
 This combines text, stack and heap memory for the program profiled, from the point of view of the OS or user, this is the actual memory in use.
@@
 In section 4.2 the parallel performance of this benchmark is further explained.
-It is of interest for sequential to contrast with the runtime penalty that dxexmachina suffers w.r.t.
+It is of interest for sequential to contrast with the runtime penalty that dxex suffers w.r.t.
 adevs.
-Optimistic fails this benchmark, see for detailed analysis section 4.2.3.
+Optimistic fails this benchmark, see section 4.2.3 for a detailed analysis.
 \begin{table}[lhtb]
@@
 The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages.
-For optimistic this can cause more reverts since the kernels will start to drift faster as the modelcount increases.
+For optimistic, this can cause more reverts since the kernels will start to drift faster as the modelcount increases.
 Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases.
-Of note as well is the warm-up time this benchmark requires, for n=dxw models, it takes timeadvance()*n transitions to activate the last model in the chain.
+Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*n$ transitions to activate the last model in the chain.
 For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.
@@
 \input pqueue.tex
-A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create, once the random nr generation is done the kernel holding the server will never revert since it is a source in the dependency graph.
+A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create.
+The kernel holding the server will never revert since it is a source in the dependency graph.
 Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold.

### 2015-11-29 paper/5-related-work.tex
Small changes to paper
Did some reading. Tried to fix some stuff we discussed in issue #28. Also changed the sentence structure in some places.

@@
 \subsection{PythonPDEVS}
-Dxexmachina is closely related to PythonPDEVS in design and philosophy.
-PythonPDEVS allows anyone who grasps the DEVS formalisms to immediately simulate his/her model without having to consider the kernel implementation.
-C++ implementations cannot hope to match the fast prototype/edit/run cycles provided by PythonPDEVS, although this can be minimized by building the kernels as libraries.
+Dxex is closely related to PyPDEVS in design and philosophy.
+PyPDEVS allows anyone who grasps the DEVS formalisms to immediately simulate his/her model without having to consider the kernel implementation.
+C++ implementations cannot hope to match the fast prototype/edit/run cycles provided by PyPDEVS, although this can be minimized by building the kernels as libraries.
 %maybe warn that if a header is touched, you still have to compile from scratch ?
@@
 TIME_FP)
-Advanced features such as activity based relocation and the performance gains this results in, are still unique to PythonPDEVS. \\
+Advanced features such as activity based relocation and the performance gains this results in, are still unique to PyPDEVS. \\
 \subsection{Adevs}

### 2015-11-29 paper/1-introduction.tex

@@
 The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
-The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code without losing access to powerful low-level threading primitives.
+The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code and without losing access to powerful low-level threading primitives.
 % may need another pass on this one :(

### 2015-11-30 paper/1-introduction.tex
changes to paper

@@
 The family of DEVS \cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms.
-Of interest in this paper are the 3 key formalisms: Classic \cite{ClassicDEVS}, Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation.
-This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a kernel's perspective only Atomic Models exist in the simulation linked to each other by connected ports.
+Of interest in this paper are the 2 key formalisms: Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation.
+This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a simulation kernel's perspective only Atomic Models exist linked to each other by connected ports.
 
@@
 memory allocation strategy).
-The core aims of the project are to offer a deterministic simulation kernel where the simulation author is shielded as much as possible from the kernel implementation, without sacrificing performance.
+The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance.
 As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\

### 2015-11-30 paper/3-features.tex
changes to paper

@@
 The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use.
-For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.\\
+For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.
+Another example is the state saving mechanism, which is only used for optimistic parallel simulation and has no performance impact in conservative parallel simulation.\\
 The tracing is not comparable to adevs's listener interface.
@@
 \end{enumerate}
-Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
+Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.
+%This state saving has no impact in a single threaded or conservative kernel.
+\\
 Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation.
 If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
-Debugging tools such as a logger and a graph visualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
-While PyPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.
+Debugging tools such as a logger and a graph visualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.
 

### 2015-11-30 Rscript/devs-performance.R
R script changes + uploaded pdf for report

@@ -1,26 +1,38 @@
-# Change this path to the folder all .csv files are stored per benchmark
-# This will become dynamic in the future, still being tested on different platforms
-setwd("d:/tim/ua/statistiek/tim-eindwerk/v6/data")
+# global variables
+generatePDF <<- FALSE	# if TRUE, generate PDF else: generated EPS files...
+epscount <<- 1
+rootpath <<- 'C:/Users/amask/Dropbox/UA/Devs Ex Machina/v9/'
+datapath <<- paste(rootpath, 'data/', sep = '')
+figspath <<- paste(rootpath, 'figs/', sep = '')
+setwd(datapath)
 
 # 1-dimensional performance tests
 
-process1ddata <- function(datatable, devstone) {
+process1ddata <- function(datatable, collabel) {
 	resultdata <- data.frame(x = numeric(), cpu = numeric(), conflow = numeric(), confhigh = numeric())
-	indextable <- unique(datatable[, 3])
+		
+	timelabel = 'time.elapsed..seconds.'
+
+	colindex = which(colnames(datatable) == collabel)
+	timeindex = which(colnames(datatable) == timelabel)
+	print(sprintf("Column Index: %s %d", collabel, colindex))
+	print(sprintf("Time Index: %d", timeindex))
+	
+	indextable <- unique(datatable[, colindex])
 	cnt <- 1
 	for (n in seq(length(indextable))) {
 			i <- indextable[[n]]
-
-			if (devstone)
-				fdata <- subset(datatable, width == i, select = c(width, time.elapsed..seconds. ))
-			else
-				fdata <- subset(datatable, nodes == i, select = c(nodes, time.elapsed..seconds. ))
+				
+			fdata <- subset(datatable, datatable[[ collabel ]] == i, select = c(colindex, timeindex))
+			
 			print(sprintf("Processing width = %d", i))
 			print(sprintf("Number of measurements: %d", nrow(fdata)))
+			
 			# print(fdata$time.elapsed..seconds.)
-			avg <- mean(fdata$time.elapsed..seconds.)
+			
+			avg <- mean(fdata[[ timelabel ]])
 			print(avg)
[…]

### 2015-11-30 Rscript/devs-performance.R
Fixed some R script issues

@@ -1,7 +1,7 @@
 # global variables
 generatePDF <<- FALSE	# if TRUE, generate PDF else: generated EPS files...
 epscount <<- 1
-rootpath <<- 'd:/tim/ua/statistiek/tim-eindwerk/v6/data'
+rootpath <<- 'd:/tim/ua/statistiek/tim-eindwerk/v9/data'
 datapath <<- paste(rootpath, 'data/', sep = '')
 figspath <<- paste(rootpath, 'figs/', sep = '')
 setwd(datapath)
@@ -231,8 +231,14 @@ comparesets <- function(datalist, labellist, xlabel, ylabel, chartlabel, legendp
 		postscript(sprintf("%sfig%d.eps", figspath, epscount), width=5, height=5, paper="special")
 		epscount <<- epscount + 1
 	}
-		
-	plot(xrange, yrange, type = "n", xlab=xlabel, ylab=ylabel)
+	
+	if (xrange[[2]] - xrange[[1]] < 10) {
+		# special case for CPU's
+		# we only want to use 2 and 4 as labels...
+		plot(xrange, yrange, type = "n", xlab=xlabel, ylab=ylabel, xaxt='n')
+		axis(side = 1, at = seq(-10,10,2) , labels = T)
+	} else
+		plot(xrange, yrange, type = "n", xlab=xlabel, ylab=ylabel)
 	
 	colors <- myrainbow(nrcharts)
 	linetype <- c(1:nrcharts)

### 2015-12-01 paper/4-performance.tex
changes to paper
changes to paper, again

@@
 \subsubsection{CPU Usage}
+%TODO !!! benchmarks are in integers
 \paragraph*{Devstone}
@@
 \paragraph*{Devstone}
+%TODO Tables with lines?
 \begin{table}[lhtb]
@@
 	\label{dtone_mem}
-	\begin{tabular}{lllll}
-		Adevs & AdevsCon &DX &DXCon &DXOpt\\
-		44 & 70 & 42 & 75 & 363  \\
+	\begin{tabular}{| l | l | l | l | l |}
+		\hline
+		Adevs & AdevsCon &DX &DXCon &DXOpt\\ \hline
+		44 & 70 & 42 & 75 & 363  \\ \hline
 	\end{tabular}
@@
 Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection, this inevitable delay explains the higher memory usage w.r.t adevs.\\
+%TODO next sentence needs rewriting
 Optimistic needs a more complex GVT/garbage collection algorithm plus the differences in LP virtual times are far larger, which explains the heavier memory usage.
@@
 	\label{iconn_mem}
-	\begin{tabular}{lllll}
-		Adevs & AdevsCon &DX &DXCon\\
-		39 & 39 & 35 & 52  \\
+	\begin{tabular}{| l | l | l | l |}
+		\hline
+		Adevs & AdevsCon &DX &DXCon\\ \hline
+		39 & 39 & 35 & 52  \\ \hline
 	\end{tabular}
@@
 	\label{pmod_mem}
-	\begin{tabular}{lll}
-		DX &DXCon &DXOpt\\
-		35 & 58 &69\\
+	\begin{tabular}{| l | l | l |}
+		\hline
+		DX &DXCon &DXOpt\\ \hline
+		35 & 58 &69\\ \hline
 	\end{tabular}
@@
 Once a revert happens the drift between the kernels increases fast, increasing the likelihood of more reverts.
-
 \subsubsection{Interconnect}
@@
 Receiving models are allocated on another kernel than the server, and have a internal transition so will not wait for the incoming messages.
-\input pqueue.tex
+\input pqueue.tex \\
 A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create.

### 2015-12-01 paper/4-performance.tex
moved some sections around

@@
 % TODO plot here
-\subsubsection{Memory Usage}
+
+\subsection{Parallel Simulation}
+%TODO explain that we will now compare different synchronization protocols
+%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
+% split: dstone/pqueue are dag's
+\subsubsection{Devstone}
+% Initial.
+The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages.
+For optimistic, this can cause more reverts since the kernels will start to drift faster as the modelcount increases.
+Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases.
+Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*n$ transitions to activate the last model in the chain.
+For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.
+\subsubsection{PHold}
+%Phold
+In Phold \cite{PHOLD} allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of sub-nodes.
+The parameter R determines the percentage of remote destination models.\\
+The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative.
+Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$.
+Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation.
+For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour.
+This benchmark therefore highlights the price that we pay for sharing those values instead of sending an actual null message.
+In a non-cyclic simulation with a non-trivial lookahead (e.g.
+devstone), that choice however does pay off.\\
+Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible.
+More seriously, a revert is very expensive in PHold due to our usage of C++11's random nr generators.
+The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages.
+Again, this could be significantly reduced using lazy-cancellation.
+Once a revert happens the drift between the kernels increases fast, increasing the likelihood of more reverts.
+\subsubsection{Interconnect}
+In Interconnect
+the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
+Allocation is irrelevant, the resulting dependency graph between kernels remains  a complete graph. The runtime dependency graph is almost immediately equal to the static dependency graph.
+Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions.
+The scaling issue is identical as in PHold.\\
+Our optimistic implementation does not complete an instance of this benchmark.
+The kernels get stuck in an infinite cascade of reverts.
+If kernel A reverts, it will send anti-messages to all others who in turn revert and send anti-messages to all others (and A itself again).
+Support for lazy cancellation could potentially undo this anti-pattern.
+\subsubsection{Priority network model}
+%TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
+The priority benchmark is composed of a single server generating a stream of $0<=m<=n$ messages at fixed time intervals, interleaved with a probability p for a priority message, to n receivers.
+\\ This defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph.
+This model therefore highlights the basic strengths/weaknesses of both synchronization protocols.
+Receiving models are allocated on another kernel than the server, and have a internal transition so that they will not wait for the incoming messages.
+\input pqueue.tex \\
+A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create.
+The kernel holding the server will never revert since it is a source in the dependency graph.
+Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold.
+
+%TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)
+\subsection{Memory Usage}
 %TODO compare memory usage with adevs for some benchmarks
@@
 If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool.
-For this reason, the memory utilization can be best measured by peak allocation.
+For this reason, the memory utilization is best measured by peak allocation.
 Profiling is done using Valgrind's massif tool \cite{Nethercote:2007:VFH:1273442.1250746}
-Platform used has a i5-3317U Intel cpu and 8GB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
+Platform used has an i5-3317U Intel cpu and 8GB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
 \paragraph*{Measure}
@@
 The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages.
-This combines text, stack and heap memory for the program profiled, from the point of view of the OS or user, this is the actual memory in use.
-It is important to note that, especially in the case of optimistic, not all this memory is in use by the kernel, since the pools will in general not return memory once it is allocated but keep it for later reuse.
+This combines text, stack and heap memory for the program profiled.
+From the point of view of the OS or user, this is the actual memory in use.
+It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernel.
+During the simulation, the pools will generally not return memory once allocated, but keep it for later reuse.
 \paragraph*{Results}
@@
 \end{table}
-
-\subsection{Parallel Simulation}
-%TODO explain that we will now compare different synchronization protocols
-%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
-% split: dstone/pqueue are dag's
-\subsubsection{Devstone}
-% Initial.
-The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages.
-For optimistic, this can cause more reverts since the kernels will start to drift faster as the modelcount increases.
-Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases.
-Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*n$ transitions to activate the last model in the chain.
-For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.
-\subsubsection{PHold}
-%Phold
-In Phold \cite{PHOLD} allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of sub-nodes.
-The parameter R determines the percentage of remote destination models.\\
-The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative.
-Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$.
-Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation.
-For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour.
-This benchmark therefore highlights the price that we pay for sharing those values instead of sending an actual null message.
-In a non-cyclic simulation with a non-trivial lookahead (e.g.
-devstone), that choice however does pay off.\\
-Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible.
-More seriously, a revert is very expensive in PHold due to our usage of C++11's random nr generators.
-The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages.
-Again, this could be significantly reduced using lazy-cancellation.
-Once a revert happens the drift between the kernels increases fast, increasing the likelihood of more reverts.
-\subsubsection{Interconnect}
-In Interconnect
-the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
-Allocation is irrelevant, the resulting dependency graph between kernels remains  a complete graph. The runtime dependency graph is almost immediately equal to the static dependency graph.
-Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions.
-The scaling issue is identical as in PHold.\\
-Our optimistic implementation does not complete an instance of this benchmark.
-The kernels get stuck in an infinite cascade of reverts.
-If kernel A reverts, it will send anti-messages to all others who in turn revert and send anti-messages to all others (and A itself again).
-Support for lazy cancellation could potentially undo this anti-pattern.
-\subsubsection{Priority network model}
-%TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
-The priority benchmark is composed of a single server generating a stream of $0<=m<=n$ messages at fixed time intervals, interleaved with a probability p for a priority message, to n receivers.
-\\ This defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph.
-This model therefore highlights the basic strengths/weaknesses of both synchronization protocols.
-Receiving models are allocated on another kernel than the server, and have a internal transition so will not wait for the incoming messages.
-\input pqueue.tex \\
-A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create.
-The kernel holding the server will never revert since it is a source in the dependency graph.
-Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold.
-
-%TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)

### 2015-12-01 paper/4-performance.tex
tried to move a few pictures into their proper place, latex still thinks it knows better where to put them.

@@
 %TODO (spend some time here, as this is also important)
-\subsubsection{CPU Usage}
+%\subsubsection{CPU Usage}
 %TODO !!! benchmarks are in integers
-\paragraph*{Devstone}
-The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this, whereas adevs needs to walk the structure of the model to pass events.
-% TODO plot here
-\paragraph*{PHold}
+\subsubsection{Devstone}
+The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this, whereas adevs needs to walk the structure of the model to pass events.\\
+\begin{figure}[h]
+	\includegraphics[width=.5\textwidth]{fig/fig1.eps}
+	\label{fig1.eps}
+\end{figure}
+
+\subsubsection{PHold}
 PHold \cite{PHOLD} is a parallel oriented benchmark, sequential runtime is measured only as a baseline. The usage of random nr generators takes up a significant amount of runtime in this model.
-% TODO plot here
-\paragraph*{Interconnect}
+
+\subsubsection{Interconnect}
 Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models.
@@
 adevs.
-This effect is due to the heap allocation of the messages by dxex, even though this is minimized using memory pools it remains significant.
-% TODO plot here
+This effect is due to the heap allocation of the messages by dxex, even though this is minimized using memory pools it remains significant.\\
+\begin{figure}[h]
+	\includegraphics[width=.5\textwidth]{fig/fig3.eps}
+	\label{fig3.eps}
+\end{figure}
 
@@
 Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*n$ transitions to activate the last model in the chain.
-For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.
+For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.\\
+\begin{figure}[h]
+	\includegraphics[width=.5\textwidth]{fig/fig2.eps}
+	\label{fig2.eps}
+\end{figure}
 \subsubsection{PHold}
 %Phold
-In Phold \cite{PHOLD} allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of sub-nodes.
+In Phold \cite{PHOLD}, allocation is specified in the benchmark itself.
+Each kernel manages a single node with a constant set of sub-nodes.
 The parameter R determines the percentage of remote destination models.\\
@@
 Support for lazy cancellation could potentially undo this anti-pattern.
+\begin{figure}[h]
+	\includegraphics[width=.5\textwidth]{fig/fig4.eps}
+	\label{fig4.eps}
+\end{figure}
 \subsubsection{Priority network model}
@@
 \paragraph*{Priority Network}
-The priority network model is detailed in section 4.4.
+The priority network model is detailed in section 4.1.
 \begin{table}[lhtb]

### 2015-12-01 paper/5-related-work.tex
changes to paper, again

@@
 Dxex is closely related to PyPDEVS in design and philosophy.
-PyPDEVS allows anyone who grasps the DEVS formalisms to immediately simulate his/her model without having to consider the kernel implementation.
-C++ implementations cannot hope to match the fast prototype/edit/run cycles provided by PyPDEVS, although this can be minimized by building the kernels as libraries.
-%maybe warn that if a header is touched, you still have to compile from scratch ?
-(eg.
-TIME_FP)
-Advanced features such as activity based relocation and the performance gains this results in, are still unique to PyPDEVS. \\
+PyPDEVS allows anyone who grasps the PDEVS formalisms to immediately simulate his/her model without having to consider the kernel implementation.A Python based implementation offers the advantage of very fast prototype/run/evaluate cycles, this can't be matched by a C++ simulator but it should be noted that once the kernels are compiled (shared libraries), the actual compilation time of the model is small enough to make prototyping possible.
+Advanced features such as activity based relocation and the performance gains this results in, are still unique to PyPDEVS.
 \subsection{Adevs}
 %TODO explain about adevs, the most related in terms of performance --> we implement more synchronization options and offer better performance because ...
-Adevs's source code is still under active development, allowing for an exact comparison in performance and features.
-It remains in most aspects the fastest simulation engine for the DEVS formalism, but it lacks an optimistic synchronization implementation.
+Adevs's is still under active development, allowing for an exact comparison in performance and features.
+It remains one of the fastest simulation engines for the PDEVS formalism, but it lacks an optimistic synchronization implementation.
 %Tracing ?
-By virtue of not flattening Coupled Models, performance suffers in increasingly hierarchical models.
+By virtue of not flattening Coupled Models, adevs's performance degrades in increasingly hierarchical models.
+On the other hand, dxex manages lookahead with more overhead than adevs, leading to a performance difference as the model count increases.
+Finally, adevs employs the full dependency graph, contrasted to dxex kernels which only observe 1-edge removed nodes.
 %checkme
@@
 It is a middleware that provides memory, event, file, time and communication scheduling.
-WARPED is not used here since we operate explicitly on a shared memory system and since we wanted to design our kernels using the least amount of overhead possible.
+We did not use the WARPED kernel (nor the underlying MPI), dxex is designed specifically for a shared memory architecture and as such any middleware, while feature-rich, would have lead to unacceptable overhead.

### 2015-12-01 paper/6-conclusions.tex

--- 2015-11-10 paper/6-conclusions.tex
@@
 %TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
+Both synchronization implementations offer good performance in differing simulations, in some simulations dxex outperforms adevs whereas in others we can still improve. 
+The optimistic implementation uses an aggressive TimeWarp variant, this needs to be extended with lazy evaluation/cancellation to function in cyclic simulations. 
+\subsection{Future work}
 %TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))
+\subsubsection{Activity}
+As shown in \cite{PythonPDEVS_ACTIMS} activity and allocation of models across kernels is a key aspect in achieving high performance in any parallel implementation.
+Allocating models so that there are no dependency cycles between their containing kernels is a first step, but not always possible.
+For optimistic one can use re-allocation to break (runtime dependency) cycles and/or perform load-balancing.
+If kernels are unevenly balanced they will begin to drift fast, causing increasingly more reverts.
+There is a limited framework implemented to track model activity for debugging purposes, this could be extended to enable the above strategies.
+\subsubsection{Hybrid}
+The optimistic implementation could use (null/eot/eit) from conservative to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.
+Conservative kernels could be extended with runtime information about influencing models, for example if one can guarantee a static dependency is not used for a fixed time-span, this can be removed for that period of (virtual) time.
+Ultimately the simulation could switch at runtime between protocols based on the information provided by activity tracking.
+This requires the above mentioned activity tracking framework, and could only be done if a GVT/LBTS time is agreed upon between the kernels.
+From the model point of view no changes are needed, although a non-trivial lookahead is obviously desired.

### 2015-12-01 paper/4-performance.tex
adjusted some sentences

@@
 If kernel A reverts, it will send anti-messages to all others who in turn revert and send anti-messages to all others (and A itself again).
-Support for lazy cancellation could potentially undo this anti-pattern.
+Support for lazy cancellation could potentially undo this anti-pattern.\\
 \begin{figure}[h]
@@
 %TODO compare memory usage with adevs for some benchmarks
-\paragraph*{Platform and tools}
+\subsubsection{Platform and tools}
 Both dxex and adevs use tcmalloc as memory allocator.
@@
 Platform used has an i5-3317U Intel cpu and 8GB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
-\paragraph*{Measure}
+\subsubsection{Measure}
 Adevs passes messages by value (in container by reference), we pass by pointer.
@@
 During the simulation, the pools will generally not return memory once allocated, but keep it for later reuse.
-\paragraph*{Results}
+\subsubsection{Results}
 \paragraph*{Devstone}
-%TODO Tables with lines?
 \begin{table}[lhtb]
@@
 \end{table}
-Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection, this inevitable delay explains the higher memory usage w.r.t adevs.\\
+Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection. This inevitable delay explains the higher memory usage compared to adevs.\\
 %TODO next sentence needs rewriting
-Optimistic needs a more complex GVT/garbage collection algorithm plus the differences in LP virtual times are far larger, which explains the heavier memory usage.
-Devstone (flattened) is allocated in a chain, this means that the leafs in the dependency graph will do a lot of unnecessary simulation before having a revert, leading to quite severe memory pressure.
+Optimistic needs a more complex GVT/garbage collection algorithm due to the Timewarp algorithm, which requires that the state of models is saved.
+Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization.
+All these factors explain the heavier memory usage.
+Devstone (flattened) is allocated in a chain.
+The leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to quite severe memory pressure.
 Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels.
@@
 adevs.
-Optimistic fails this benchmark, see section 4.2.3 for a detailed analysis.
+Optimistic fails this benchmark, see section 4.2.3 for a detailed analysis.\\
 \begin{table}[lhtb]
@@
 \paragraph*{Priority Network}
-The priority network model is detailed in section 4.1.
+The priority network model is detailed in section 4.1.\\
 \begin{table}[lhtb]

### 2015-12-01 paper/3-features.tex
changes to paper, again
added some comments here and there to clarify what has been done/mentioned elsewhere etc.

@@
 \subsection{Based on PythonPDEVS}
-%TODO say that it is based on PythonPDEVS
+%TODO say that it is based on PythonPDEVS --done
 The simulator is based on PythonPDEVS, and provides the following features: 
@@
 %This state saving has no impact in a single threaded or conservative kernel.
+-> already mentioned above
 \\
@@
 %TODO explain rationale: why different synchronization protocols
+% Multiple synchronization protocols exist, with each their own strengths and weaknesses.
 % weak scenarios for C, O and both ?
+%TODO !!!!!!! This needs better explaining, perhaps a central graph to reference to ?
 \subsubsection{Conservative}
-% This needs better explaining, perhaps a central graph to reference to ?
 A conservative kernel will determine which kernels it is influenced by.
@@
 The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
-For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i])  - \epsilon $.
+For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i])  - \epsilon $.\\
 %This is one of the operations that can benefit from using a relaxed memory ordering.
 % Using it for read atm, check before finalizing and do we need to mention this ?.
+%writing this here seems inappropriate
+%TODO indicate in which cases conservative is good: perfect/good lookahead calculation (might be very hard) and what may be a problem: runtime behavior compared to static analysis makes the simulator too conservative.
+(e.g.
+a model might send a message, but never does will slow down the simulation)
 
@@
 From the user's perspective, the multi-threaded aspect of the kernel is not exposed. % payload protected ?
+%TODO indicate in which optimistic is good (when runtime behavior is very hard to predict). Describe some drawbacks (many reverts when everybody influences everybody all the time)
 
@@
 \\
-Initially the kernels used strings as identifiers, as is done in PyPDEVS, profiling quickly indicated this to be a performance bottleneck.
+The kernels initially used strings as identifiers, as is done in PyPDEVS.
+Profiling quickly indicated this to be a performance bottleneck.
 C++ strings are heap allocated variable sized objects with an atomic reference count.
@@
 While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced.
-Used across threads the reference counting becomes prohibitively expensive, and the (de)allocating caused significant contention between threads.
+Used across threads, the reference counting becomes very expensive, and the (de)allocating of memory caused significant contention between threads.
 Models are still held by a smart pointer, as is a kernel, but a message is a raw pointer to compacted memory.

### 2015-12-02 paper/3-features.tex
Some broken sentences rewritten.

@@
 This process runs asynchronously from the simulation itself.
-Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states/(anti)messages.
+Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states and (anti)messages.
 
@@
 \subsubsection{Heap}
-A kernel never sends a complete object to another kernel, only a pointer to the object. This avoids a possibly expensive copy of the payload, but at the cost of allocation overhead.\\
-This cost becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and messages, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}.
-In this way allocating threads do not block each other, and in a single threaded kernel we can leverage arena-style pools if desired.\\
-This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (if it can prove it is no longer used).
+Events are passed by pointer, avoiding a possibly expensive copy of the payload at the cost of allocation overhead.\\
+Allocation becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and events, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}.
+In this way allocating threads do not block each other.
+If desired, arena-pools are available for single-threaded simulation.
+A disadvantage is increased complexity in ownership semantics.
+The creating kernel is responsible for destruction, but this can only be guaranteed up to GVT.
 Experiments with synchronized pools proved slower than malloc/free.
@@
 Profiling quickly indicated this to be a performance bottleneck.
-C++ strings are heap allocated variable sized objects with an atomic reference count.
-Access of that reference count across threads is expensive, as are the calls to malloc/free the string implementation makes to create/destroy new object, or copy existing.\\
-Strings are more intuitive to work with from a user's standpoint.
-So as a compromise the user can reference models/ports by string name (usually when constructing the model).
+C++ strings are heap allocated variable sized objects with an atomic reference count, not immutable objects as in Python.
+Access of that reference count across threads is expensive, as are the calls to malloc/free the string implementation makes to create/destroy new objects, or copy existing.\\
+Strings are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name.
 Once simulation starts all objects use integral identifiers for performance.
 This also increased usage of the constexpr feature of C++11 in, amongst others, timestamps and message headers.
-\\
 \subsubsection{Raw pointers}

### 2015-12-03 main/src/model/conservativecore.cpp
Disable a few log statements, reduce testload interconnect to keep sane logs/test times.

@@ -389,7 +389,7 @@ Conservativecore::calculateMinLookahead(){
                 m_min_lookahead = t_timestamp::infinity();
                 for(const auto& model : m_indexed_models){
                         const t_timestamp la = model->lookAhead();
-                        LOG_DEBUG("Core :: ", this->getCoreID()," Model :: ", model->getName(), " gave LA = ", la);
+                        //LOG_DEBUG("Core :: ", this->getCoreID()," Model :: ", model->getName(), " gave LA = ", la);
 #ifdef SAFETY_CHECKS
                         if(isZero(la))
                                 throw std::logic_error("Lookahead can't be zero");

### 2015-12-03 main/src/model/core.cpp

@@ -33,14 +33,13 @@ void
 n_model::Core::checkInvariants(){
 #ifdef SAFETY_CHECKS
         if(this->m_heap.indexSize() != this->m_indexed_models.size()){
+                printSchedulerState();
                 const std::string msg = "Scheduler contains less models than present in core !!";
                 LOG_ERROR(msg);
                 LOG_FLUSH;
                 throw std::logic_error(msg);
         }
         LOG_DEBUG("\tCORE :: ", this->getCoreID(), " testing invariants.");
-        printSchedulerState();
-
         m_heap.testInvariant();
 #endif
 }
@@ -236,14 +235,14 @@ void n_model::Core::transition()
 		}
                 imminent->clearSentMessages();
 
-                printSchedulerState();
+                //printSchedulerState();
                 LOG_DEBUG("\tCORE :: ", this->getCoreID(), " fixing scheduler heap.");
 
 		if(m_heap.doSingleUpdate())
 			m_heap.update(modelid);
                 
                 LOG_DEBUG("\tCORE :: ", this->getCoreID(), " result.");
-                printSchedulerState();
+                //printSchedulerState();
 	}
         LOG_DEBUG("\tCORE :: ", this->getCoreID(), " Transitioning with ", m_externs.size(), " externs");
         for(auto external : m_externs){
@@ -259,12 +258,12 @@ void n_model::Core::transition()
 		this->traceExt(getModel(id));
 
                 clearProcessedMessages(mail);
-                printSchedulerState();
+                //printSchedulerState();
                 LOG_DEBUG("\tCORE :: ", this->getCoreID(), " fixing scheduler heap.");
 		if(m_heap.doSingleUpdate())
 			m_heap.update(id);
                 LOG_DEBUG("\tCORE :: ", this->getCoreID(), " result.");
-                printSchedulerState();
+                //printSchedulerState();
 		assert(!hasMail(id) && "After external transition, model may no longer have pending mail.");
 	}
         

### 2015-12-03 main/src/test/benchmarktest.cpp

@@ -383,6 +383,9 @@ TEST(Benchmark, connect_cons)
     LOG_MOVE("out.txt", true);
 }
 
+// Reduce runtime, both sync protocols log a lot, 5000 easily takes 30s.
+constexpr t_timestamp::t_time eTimeConnectR = 2000;
+
 TEST(Benchmark, connect_single_r)
 {
     LOG_MOVE("logs/bmarkConnectSingleR.log", false);
@@ -396,7 +399,7 @@ TEST(Benchmark, connect_single_r)
 	bool randTa = true;
 
 	auto ctrl = conf.createController();
-	t_timestamp endTime(eTimeConnect, 0);
+	t_timestamp endTime(eTimeConnectR, 0);
 	ctrl->setTerminationTime(endTime);
 
 	t_coupledmodelptr d = n_tools::createObject<n_interconnect::HighInterconnect>(width, randTa);
@@ -424,7 +427,7 @@ TEST(Benchmark, connect_opt_r)
 	bool randTa = true;
 
 	auto ctrl = conf.createController();
-	t_timestamp endTime(eTimeConnect, 0);
+	t_timestamp endTime(eTimeConnectR, 0);
 	ctrl->setTerminationTime(endTime);
 
 	t_coupledmodelptr d = n_tools::createObject<n_interconnect::HighInterconnect>(width, randTa);
@@ -452,7 +455,7 @@ TEST(Benchmark, connect_cons_r)
 	bool randTa = true;
 
 	auto ctrl = conf.createController();
-	t_timestamp endTime(eTimeConnect, 0);
+	t_timestamp endTime(eTimeConnectR, 0);
 	ctrl->setTerminationTime(endTime);
 
 	t_coupledmodelptr d = n_tools::createObject<n_interconnect::HighInterconnect>(width, randTa);

### 2015-12-03 paper/1-introduction.tex

@@
 %TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
-The usage of the Direct connection algorithm makes reusing the adevs kernel hard.
-The devs-ex-machina (dxex) kernel follows the design of the PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g.
-memory allocation strategy).
+The usage of the Direct connection algorithm makes reusing adevs kernel hard. Devs-ex-machina (dxex) is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
 The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance.
@@
 This task is non-trivial for floating point approximations.\\
-The select() function, which imposes a sequential ordering between concurrent events, is made implicit in this project by extending time representation with a causality field with a range at least as large as the maximum nr of models in the kernels ($2^{24}$ by default).
+Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernels ($2^{24}$ by default).
 If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$.
-This avoids evaluating the select() function on all imminent models, while still maintaining the deterministic order of concurrent transitions.
+This allows a deterministic ordering of concurrent events.
+% execution not so much as tracing

### 2015-12-03 paper/2-background.tex

--- 2015-11-29 paper/2-background.tex
@@
 An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime.
-A model that can influence another, but never does, can severely slow down the protocol.\\
+A model that can influence another, but never does, can severely slow down the protocol.
 Deadlock between models that influence each other and end up waiting on each other can be broken/avoided by a variety of means.
@@
 More precisely, a null message of time t is a guarantee that any output with timestamp t-$\epsilon$ is already sent.\\
-In general, the eot/eit/nulltime of a kernel is the minimum of each of those values for all models in the kernel.\\
 Conservative synchronization relies explicitly on information provided by the model creator in the form of a lookahead, a relative timespan during which the model is insensitive to outside events.
-This lookahead can be non-trivial to calculate.
-In general, a simulation writer will not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
+This lookahead can be non-trivial to calculate, a simulation writer will in general not be able to determine the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
 \subsection{Optimistic Synchronization}
 %TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
-Optimistic synchronization allows causality errors to occur but recovers from those errors using a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.
-Whenever a kernel receives an event with a timestamp in the kernel's past, the state of the kernel (and all models) is reverted to that time.
-The possible gain in runtime is offset by the increase in memory required to keep saved states and (sent) messages.
-In contrast to conservative, optimistic does not rely on any domain specific information.
-It is only sensitive to runtime use of connections, not the probability that they might occur.
+Optimistic synchronization allows causality errors to occur but recovers using a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.
+Whenever a kernel receives an event with a timestamp in the kernel's past, the state of the kernel (and all models)reverts to that time.
+Because an optimistic kernel does not wait for other kernels, it can achieve a lower runtime than conservative.
+The cost of this approach is higher complexity of the code and larger memory usage for state/message saving.
+In contrast to conservative, optimistic does not rely on any domain specific information, it is only sensitive to runtime use of connections, not the probability that they might occur.
 %reword connections ?
@@
 If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts.
-Without domain specific information, the kernel assumes that any event will influence at least one model, but this can lead to an infinite loop of reverts in the worst case.\\
-This effect can be lessened by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
+The cost of reverts can be reduced by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
 \subsection{Global Virtual Time}
@@
 In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event.
-\\ Intuitively this is the simulation timepoint that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
-In conservative the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
+\\ Intuitively this is the simulation timepoint that is certain to be preserved, corresponding exactly with a sequential simulation.
+In conservative, the minimum simulation time of all kernels is the LBTS, or in terms of null messages: the least timestamp of any null message in transit.
 The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .

### 2015-12-03 paper/4-performance.tex
added some comments here and there to clarify what has been done/mentioned elsewhere etc.

@@
 %TODO !!! benchmarks are in integers
+The benchmark models are identical in functionality (between adevs/dxex versions). Dxex uses integer timestamps, adevs can only use floating point (of the same values).
 \subsubsection{Devstone}
@@
 Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models.
-It highlights in sequential simulation the cost of heap allocation (messages) in dxex.
 As the model count increases, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxex w.r.t.
 adevs.
-This effect is due to the heap allocation of the messages by dxex, even though this is minimized using memory pools it remains significant.\\
+Profiling shows this is entirely to the heap allocation of the messages by dxex, even though minimized by using memory pools, it remains significant.\\
 \begin{figure}[h]
@@
 Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection. This inevitable delay explains the higher memory usage compared to adevs.\\
-%TODO next sentence needs rewriting
+%TODO next sentence needs rewriting --done?
 Optimistic needs a more complex GVT/garbage collection algorithm due to the Timewarp algorithm, which requires that the state of models is saved.

### 2015-12-04 paper/4-performance.tex
R script and benchmark changes. Added README for Rscript. Updated paper.

@@
 Support for lazy cancellation could potentially undo this anti-pattern.\\
-\begin{figure}[h]
-	\includegraphics[width=.5\textwidth]{fig/fig4.eps}
-	\label{fig4.eps}
-\end{figure}
+
 \subsubsection{Priority network model}

### 2015-12-04 Rscript/devs-performance.R

@@ -1,14 +1,17 @@
 # global variables
 generatePDF <<- FALSE	# if TRUE, generate PDF else: generated EPS files...
 epscount <<- 1
-rootpath <<- 'd:/tim/ua/statistiek/tim-eindwerk/v9/data'
+rootpath <<- paste(getwd(), '/', sep = '')
+# rootpath <<- 'putyourcustomrootpathinhere'
 datapath <<- paste(rootpath, 'data/', sep = '')
 figspath <<- paste(rootpath, 'figs/', sep = '')
+dir.create(figspath, showWarnings = FALSE)
 setwd(datapath)
 
 # 1-dimensional performance tests
 
 process1ddata <- function(datatable, collabel) {
+	print(sprintf("process1ddata: %s", collabel))
 	resultdata <- data.frame(x = numeric(), cpu = numeric(), conflow = numeric(), confhigh = numeric())
 		
 	timelabel = 'time.elapsed..seconds.'
@@ -211,13 +214,14 @@ newrange <- function(range1, range2) {
 }
 
 myrainbow <- function(n) {
-	if (n <= 5)
-		return(c("red", "green", "blue", "dodgerblue", "purple" ))
+	if (n <= 6)
+		return(c("red", "green", "blue", "magenta", "dodgerblue", "black" ))
 	else
 		return(rainbow(n))
 }
 
 comparesets <- function(datalist, labellist, xlabel, ylabel, chartlabel, legendpos) {
+	print(sprintf("comparesets %s\n",chartlabel))
 	nrcharts <- length(datalist)
 	
 	xrange <- range(datalist[[1]]$x)
@@ -228,13 +232,12 @@ comparesets <- function(datalist, labellist, xlabel, ylabel, chartlabel, legendp
 	}
 	
 	if (!generatePDF) {
-		postscript(sprintf("%sfig%d.eps", figspath, epscount), width=5, height=5, paper="special")
+		postscript(sprintf("%sfig%d.eps", figspath, epscount), width=5, height=5, paper="special", family="Times")
 		epscount <<- epscount + 1
 	}
 	
 	if (xrange[[2]] - xrange[[1]] < 10) {
 		# special case for CPU's
-		# we only want to use 2 and 4 as labels...
 		plot(xrange, yrange, type = "n", xlab=xlabel, ylab=ylabel, xaxt='n')
 		axis(side = 1, at = seq(-10,10,2) , labels = T)
 	} else
[…]

### 2015-12-04 paper/4-performance.tex
Update memory usage optimistic + explain why optimistic fails in iconnect & update phold+iconnect.

@@
 adevs.
-Profiling shows this is entirely to the heap allocation of the messages by dxex, even though minimized by using memory pools, it remains significant.\\
+Profiling shows this is entirely to the heap allocation of messages by dxex, even though minimized by using memory pools, it remains significant.\\
 \begin{figure}[h]
@@
 For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.\\
-\begin{figure}[h]
+\begin{figure}[ltbh]
 	\includegraphics[width=.5\textwidth]{fig/fig2.eps}
@@
 %Phold
-In Phold \cite{PHOLD}, allocation is specified in the benchmark itself.
-Each kernel manages a single node with a constant set of sub-nodes.
-The parameter R determines the percentage of remote destination models.\\
+In Phold allocation is specified in the benchmark itself. Each kernel manages a single node with a constant set of sub-nodes. The parameter R determines the percentage of remote destination models.\\
 The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative.
@@
 For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour.
-This benchmark therefore highlights the price that we pay for sharing those values instead of sending an actual null message.
 In a non-cyclic simulation with a non-trivial lookahead (e.g.
-devstone), that choice however does pay off.\\
+devstone), that choice however does pay off.
+Adevs' lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code.
 Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible.
-More seriously, a revert is very expensive in PHold due to our usage of C++11's random nr generators.
+With the percentage of remotes (R) $~$ 100 phold reflects interconnect in behaviour, which is why the R value is not dimensioned here.
+A revert is very expensive in PHold due to our usage of C++11's random nr generators.
 The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages.
@@
 Once a revert happens the drift between the kernels increases fast, increasing the likelihood of more reverts.
+Despite all this, optimistic can quickly exploit uncertainty that slows down conservative in this benchmark for low R values.
 \subsubsection{Interconnect}
@@
 the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
-Allocation is irrelevant, the resulting dependency graph between kernels remains  a complete graph. The runtime dependency graph is almost immediately equal to the static dependency graph.
+Allocation cannot avoid cycles between the kernels, the resulting dependency graph between kernels remains a complete graph.
+The runtime dependency graph is almost immediately equal to the static dependency graph.
 Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions.
 The scaling issue is identical as in PHold.\\
-Our optimistic implementation does not complete an instance of this benchmark.
-The kernels get stuck in an infinite cascade of reverts.
-If kernel A reverts, it will send anti-messages to all others who in turn revert and send anti-messages to all others (and A itself again).
-Support for lazy cancellation could potentially undo this anti-pattern.\\
+In this benchmark optimistic runs very quickly out of memory.
+With c the kernels N atomic models, a single revert undoing k transitions will lead to $(N-1)\times k$ messages that need to be recreated, plus $(N-1) \frac{c-1}{c} \times k$ anti-messages that need to be send.
 
@@
 Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold.
+The overhead in optimistic is entirely due to the factor m, which will quickly dominate in increasing buffers of received messages.
+\begin{figure}[ltbh]
+	\includegraphics[width=.5\textwidth]{fig/fig5.eps}
+	\label{fig5.eps}
+\end{figure}
 
@@
 If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool.
-For this reason, the memory utilization is best measured by peak allocation.
-Profiling is done using Valgrind's massif tool \cite{Nethercote:2007:VFH:1273442.1250746}
-Platform used has an i5-3317U Intel cpu and 8GB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
+This is one reason why memory utilization is best measured by peak allocation.
+Profiling is done using Valgrind's massif tool.
+\cite{Nethercote:2007:VFH:1273442.1250746}
+Platform used has an i5-3317U Intel cpu and 8GiB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
 \subsubsection{Measure}
 Adevs passes messages by value (in container by reference), we pass by pointer.
-The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages.
-This combines text, stack and heap memory for the program profiled.
-From the point of view of the OS or user, this is the actual memory in use.
-It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernel.
-During the simulation, the pools will generally not return memory once allocated, but keep it for later reuse.
+The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages combinin text, stack and heap memory for the program profiled.
+For the OS and/or user this is the actual memory footprint of the application.
+It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernels.
+During simulation, the pools will generally not return deallocated memory to the OS, but keep it for later reuse.
 \subsubsection{Results}
@@
 %TODO next sentence needs rewriting --done?
-Optimistic needs a more complex GVT/garbage collection algorithm due to the Timewarp algorithm, which requires that the state of models is saved.
+Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resultant higher latency) than the LBTS calculation in conservative. 
 Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization.
@@
 \paragraph*{Interconnect}
-In section 4.2 the parallel performance of this benchmark is further explained.
-It is of interest for sequential to contrast with the runtime penalty that dxex suffers w.r.t.
-adevs.
-Optimistic fails this benchmark, see section 4.2.3 for a detailed analysis.\\
+In section 4.2 the parallel performance of this benchmark is further explained. It is of interest for sequential to contrast with the runtime penalty that dxex suffers w.r.t. adevs. 
 \begin{table}[lhtb]
 	\centering
-	\caption{Interconnect w 40 t5e5, unit MiB, 2 kernels (if parallel)}
+	\caption{Interconnect w 10 t5e5, unit MiB, 2 kernels (if parallel)}
 	\label{iconn_mem}
-	\begin{tabular}{| l | l | l | l |}
+	\begin{tabular}{| l | l | l | l | l |}
 		\hline
-		Adevs & AdevsCon &DX &DXCon\\ \hline
-		39 & 39 & 35 & 52  \\ \hline
+		Adevs & AdevsCon &DX &DXCon & DXOpt\\ \hline
+		39 & 39 & 35 & 43 & 164 \\ \hline
 	\end{tabular}
 \end{table}
-The discrepancy between both conservative implementations is detailed in 4.1.4.
 
@@
 The priority network model is detailed in section 4.1.\\
+Conservative's high memory usage is largely due to the garbage collection implementation.
+Kernel 0 is the initiator of the lbts calculation but since it holds only models independent of any other (server), it will finish very fast leaving the other kernel bereft of an updated lbts.
 \begin{table}[lhtb]
 	\centering
-	\caption{Priority model n 16, m 9, p 10,  t5e5, unit MiB, 2 kernels}
+	\caption{Priority model n 128, m 16, p 10,  t2e8, unit MiB, 2 kernels}
 	\label{pmod_mem}
@@
 		DX &DXCon &DXOpt\\ \hline
-		35 & 58 &69\\ \hline
+		35 & 450 & 651\\ \hline
 	\end{tabular}

### 2015-12-04 paper/5-related-work.tex
Update memory usage optimistic + explain why optimistic fails in iconnect & update phold+iconnect.

@@
 %TODO explain about adevs, the most related in terms of performance --> we implement more synchronization options and offer better performance because ...
-Adevs's is still under active development, allowing for an exact comparison in performance and features.
-It remains one of the fastest simulation engines for the PDEVS formalism, but it lacks an optimistic synchronization implementation.
+Adevs is still under active development, allowing for an exact comparison in performance and features.
+It remains a very fast simulation engine for the PDEVS formalism, but it lacks an optimistic synchronization implementation.
 %Tracing ?
 By virtue of not flattening Coupled Models, adevs's performance degrades in increasingly hierarchical models.
-On the other hand, dxex manages lookahead with more overhead than adevs, leading to a performance difference as the model count increases.
+Dxex manages lookahead with more overhead than adevs, leading to a performance difference as the model count increases.
 Finally, adevs employs the full dependency graph, contrasted to dxex kernels which only observe 1-edge removed nodes.

### 2015-12-04 doBenchmark.py

@@ -106,6 +106,10 @@ def pholdgen(simtype, executable):
     # time single 5 000 000
     # single core s=[2, 4, 8, 16], r=[10]
     # for single & conservative
+    # Todo re-enable
+    if simtype == simtypes.optimistic:
+         print("Refusing to run phold with optimistic!")
+         return
     if args.cores != 4:
         print("Refusing to run benchmark with != 4 cores.")
         return
@@ -123,6 +127,7 @@ def interconnectgen(simtype, executable, doRandom=False):
     if simtype == simtypes.optimistic:
         print("Refusing to run interconnect with optimistic!")
         return
+
     if simtype == simtypes.classic:
         # time 5 000 000
         for width in [10, 20, 30, 40, 50, 60, 70]:

### 2015-12-05 paper/4-performance.tex
Add system specs.
Update profiling call-graphs.

@@
 %TODO (spend some time here, as this is also important)
-%\subsubsection{CPU Usage}
-%TODO !!! benchmarks are in integers
-The benchmark models are identical in functionality (between adevs/dxex versions). Dxex uses integer timestamps, adevs can only use floating point (of the same values).
+\subsection{Preamble}
+The benchmark models are identical in functionality (between adevs/dxex versions).
+Dxex uses integer timestamps, adevs uses floating point.
+Tests show there is little difference performance wise (both types fit in a machine word)\\
+Benchmarks ran on a i7-2670QM, 16GiB Ram, Arch Linux (4.2.5 kernel).
+Analysis of the benchmark results is done using call-graphs generated by perf, a selection of those call-graphs can be found in the code repository.
 \subsubsection{Devstone}

### 2015-12-05 paper/4-performance.tex
Update lost profiles. Expand 4/6 a bit more + cite GVT futures.

@@
 %TODO (spend some time here, as this is also important)
-\subsection{Preamble}
+\subsubsection{Preamble}
 The benchmark models are identical in functionality (between adevs/dxex versions).
@@
 Analysis of the benchmark results is done using call-graphs generated by perf, a selection of those call-graphs can be found in the code repository.
+\\
+Note that a fixed timeadvance (e.g. devstone) is set at 100.
 \subsubsection{Devstone}
-The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this, whereas adevs needs to walk the structure of the model to pass events.\\
+The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this whereas adevs needs to walk the coupled model's structure to pass events.\\
 \begin{figure}[h]
@@
 \subsubsection{PHold}
-PHold \cite{PHOLD} is a parallel oriented benchmark, sequential runtime is measured only as a baseline. The usage of random nr generators takes up a significant amount of runtime in this model.
+PHold \cite{PHOLD} is a parallel oriented benchmark, sequential runtime is measured only as a baseline.
+A model selects the destination of a message at runtime using a state-saved RNG, which takes up a non-trivial amount of runtime.
 
 \subsubsection{Interconnect}
-Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models.
-As the model count increases, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxex w.r.t.
-adevs.
-Profiling shows this is entirely to the heap allocation of messages by dxex, even though minimized by using memory pools, it remains significant.\\
+Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in dependencies between models.
+As the model count increases, we see the expected quadratic increase in runtime for both adevs and dxex, but an increasing penalty for dxex.
+Profiling shows this is entirely due to the heap allocation of messages, which even though minimized by using memory pools remains significant.\\
 \begin{figure}[h]
@@
 %TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
-% split: dstone/pqueue are dag's
 \subsubsection{Devstone}
-% Initial.
 The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages.
-For optimistic, this can cause more reverts since the kernels will start to drift faster as the modelcount increases.
+For optimistic, this can cause more reverts since the kernels will start to drift faster as the model-count increases.
 Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases.
-Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*n$ transitions to activate the last model in the chain.
-For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.\\
+Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*(n-1)$ transitions to activate the last model in the chain.
+For parallel this can be reduced to $\textup{timeadvance}()*(n\frac{kernels-1}{kernels} -1)$ before the last kernel becomes active.\\
 \begin{figure}[ltbh]
@@
 %Phold
-In Phold allocation is specified in the benchmark itself. Each kernel manages a single node with a constant set of sub-nodes. The parameter R determines the percentage of remote destination models.\\
+In Phold allocation is specified in the benchmark itself.
+Each kernel manages a single node with a constant set of sub-nodes.
+The parameter R (here 10) determines the percentage of remote destination models.\\
 The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative.
@@
 devstone), that choice however does pay off.
-Adevs' lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code.
+Adevs's lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code.
 Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible.
-With the percentage of remotes (R) $~$ 100 phold reflects interconnect in behaviour, which is why the R value is not dimensioned here.
+With the percentage of remotes = 100 phold reflects interconnect in behaviour, which is why the R value is not dimensioned here.
 A revert is very expensive in PHold due to our usage of C++11's random nr generators.
@@
 Again, this could be significantly reduced using lazy-cancellation.
-Once a revert happens the drift between the kernels increases fast, increasing the likelihood of more reverts.
-Despite all this, optimistic can quickly exploit uncertainty that slows down conservative in this benchmark for low R values.
+Once a revert happens the drift between kernels increases fast, increasing the likelihood of more reverts.
+Despite all this, optimistic can for low R values quickly exploit uncertainty that slows down conservative in this benchmark.
 \subsubsection{Interconnect}
@@
 the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
-Allocation cannot avoid cycles between the kernels, the resulting dependency graph between kernels remains a complete graph.
-The runtime dependency graph is almost immediately equal to the static dependency graph.
+Allocation cannot avoid cycles, the resulting dependency graph between kernels remains a complete graph. The runtime dependency graph is almost immediately equal to the static graph.
 Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions.
-The scaling issue is identical as in PHold.\\
+The scaling issue is identical as with PHold.\\
 In this benchmark optimistic runs very quickly out of memory.
-With c the kernels N atomic models, a single revert undoing k transitions will lead to $(N-1)\times k$ messages that need to be recreated, plus $(N-1) \frac{c-1}{c} \times k$ anti-messages that need to be send.
+With c kernels and N atomic models, a single revert undoing k transitions will lead to $(N-1)\times k$ messages that need to be recreated, plus $(N-1) \frac{c-1}{c} \times k$ anti-messages that need to be send.
 
@@
 \end{figure}
+Interestingly, the parameter p does not clearly benefit either synchronization protocol.
+While this removes any possibility for a lookahead, conservative can quickly bridge the timespan between fixed messages.
+Lightweight states prevent optimistic from performance loss due to reverts, only the overhead in storing messages becomes dominant.
 
@@
 \cite{Nethercote:2007:VFH:1273442.1250746}
-Platform used has an i5-3317U Intel cpu and 8GiB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
+Platform used for memory profiling has an i5-3317U Intel cpu and 8GiB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
 \subsubsection{Measure}
 Adevs passes messages by value (in container by reference), we pass by pointer.
-The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages combinin text, stack and heap memory for the program profiled.
+The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled.
 For the OS and/or user this is the actual memory footprint of the application.
@@
 Devstone (flattened) is allocated in a chain.
-The leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to quite severe memory pressure.
+Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to severe memory pressure.
 Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels.
 \paragraph*{Interconnect}
-In section 4.2 the parallel performance of this benchmark is further explained. It is of interest for sequential to contrast with the runtime penalty that dxex suffers w.r.t. adevs. 
+In section 4.2 the parallel performance of this benchmark is further explained. Interconnect highlights dxex's lower performance due to message allocation overhead.
 \begin{table}[lhtb]
@@
 Conservative's high memory usage is largely due to the garbage collection implementation.
-Kernel 0 is the initiator of the lbts calculation but since it holds only models independent of any other (server), it will finish very fast leaving the other kernel bereft of an updated lbts.
+Kernel 0 is the initiator of the lbts calculation but since it holds only models independent of any other (server), it will finish simulation very fast leaving the other kernel bereft of an updated lbts.
+The kernel holding the server will after simulation wait on the receiving kernel until it can prove all send messages can be deallocated.
 \begin{table}[lhtb]

### 2015-12-05 paper/6-conclusions.tex
Update lost profiles. Expand 4/6 a bit more + cite GVT futures.

@@
 %TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
-Both synchronization implementations offer good performance in differing simulations, in some simulations dxex outperforms adevs whereas in others we can still improve. 
-The optimistic implementation uses an aggressive TimeWarp variant, this needs to be extended with lazy evaluation/cancellation to function in cyclic simulations. 
+Sequentially dxex outperforms adevs in hierarchical simulations while yielding to adevs in broadcast-like simulations. 
+Parallel the results are more nuanced.
+The sensitivity to broadcast simulations is carried over in the parallel implementation, resulting in performance loss for both synchronization protocols.
+Optimistic can still achieve good performance if the volume of inter-kernel messages is low.
+Both  are fast in non-cyclic simulations, with conservative better suited whenever state-saving becomes expensive and optimistic when lookahead is not available.\\
+Finally, the current optimistic implementation is ill-suited for memory constrained systems, especially in simulations where the frequency of reverts is high.
 \subsection{Future work}
 %TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))
+\subsubsection{Memory}
+Our optimistic implementation can benefit from a faster GVT algorithm such as FujiMoto's \cite{Fujimoto:1997:CGV:268403.268404} or more recently \cite{Bauer:2005:SND:1069810.1070159}. 
+With access to the memory subsystem in place (pools), optimistic could be constrained using per kernel quota preventing the simulation from exhausting memory.
+While less sensitive to memory pressure, the same approach could benefit conservative.
 \subsubsection{Activity}
@@
 Ultimately the simulation could switch at runtime between protocols based on the information provided by activity tracking.
-This requires the above mentioned activity tracking framework, and could only be done if a GVT/LBTS time is agreed upon between the kernels.
+This requires the above mentioned framework, and could only be done at timepoints where a GVT/LBTS time is agreed upon between the kernels.
 From the model point of view no changes are needed, although a non-trivial lookahead is obviously desired.

### 2015-12-05 paper/1-introduction.tex
More rewrites & fixes.

@@
 Of interest in this paper are the 2 key formalisms: Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation.
-This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a simulation kernel's perspective only Atomic Models exist linked to each other by connected ports.
+This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a simulation kernel's perspective only Atomic Models exist linked to each other by ports.
 
@@
 Parallel execution of a PDEVS simulation can lower overall runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
-While the shared memory parallelism offered by most modern hardware does not increase the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation.
+While the shared memory parallelism offered by most modern hardware does not raise the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation.
 \\
@@
 %TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
-Adevs \cite{adevs} offers a very fast conservative synchronized shared memory DEVS simulator, but no optimistic synchronized variant. 
+Adevs \cite{adevs} offers a very fast conservative synchronized shared memory PDEVS simulator, but no optimistic synchronized variant. 
 The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
@@
 
-\subsection{Solution} %for lack of a better word
+\subsection{Solution} 
 %TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
@@
 \subsection{Time}
-The DEVS formalisms has $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time.
-Dxex kernels can operate on IEEE754 floating point time units or integer time.
-In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.%ref to time paper.
-An integer representation significantly reduces the possible range of the virtual time, but avoids approximation errors.
+The DEVS formalisms have $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time.
+In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.
+In practice, dxex defaults to word sized unsigned integer.
+Although this significantly reduces the possible range of virtual time, it avoids approximation errors.
 Furthermore, the notion of $\epsilon$ as a minimum distance between two distinct time points need not to be established.
 This task is non-trivial for floating point approximations.\\
-Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernels ($2^{24}$ by default).
+Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernels ($2^{48}$ by default).
 If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$.

### 2015-12-05 paper/3-features.tex
More rewrites & fixes.

@@
 %TODO explain rationale: why different synchronization protocols
-% Multiple synchronization protocols exist, with each their own strengths and weaknesses.
-% weak scenarios for C, O and both ?
-%TODO !!!!!!! This needs better explaining, perhaps a central graph to reference to ?
+Synchronization is required for PDEVS to prevent or recover from causality violations if the simulation is executed in parallel.
+Preventing causality violations (conservative) typically requires domain-specific information to compensate for performance loss, while recovering from causality errors (optimistic) requires a state to recover to.
+
 \subsubsection{Conservative}
@@
 %writing this here seems inappropriate
-%TODO indicate in which cases conservative is good: perfect/good lookahead calculation (might be very hard) and what may be a problem: runtime behavior compared to static analysis makes the simulator too conservative.
-(e.g.
-a model might send a message, but never does will slow down the simulation)
 

### 2015-12-05 paper/4-performance.tex
More rewrites & fixes.

@@
 Tests show there is little difference performance wise (both types fit in a machine word)\\
-Benchmarks ran on a i7-2670QM, 16GiB Ram, Arch Linux (4.2.5 kernel).
-Analysis of the benchmark results is done using call-graphs generated by perf, a selection of those call-graphs can be found in the code repository.
-\\
+Benchmarks ran on a i7-2670QM, 16GiB Ram, Arch Linux (4.2.5 kernel), G++ 5.2.
+Interpretation of the benchmark results is done using call-graphs generated by perf, a selection of those call-graphs can be found in the code repository.
+For the runtime plots perf is used in combination with R.\\
 Note that a fixed timeadvance (e.g. devstone) is set at 100.
@@
 Interestingly, the parameter p does not clearly benefit either synchronization protocol.
-While this removes any possibility for a lookahead, conservative can quickly bridge the timespan between fixed messages.
-Lightweight states prevent optimistic from performance loss due to reverts, only the overhead in storing messages becomes dominant.
+While this removes any possibility for a lookahead, conservative can quickly bridge the timespan between fixed messages since there is no cyclic dependency.
+Lightweight states prevent performance loss due to reverts in optimistic, only the overhead in event handling in optimistic eventually becomes the deciding factor.
 
@@
 Kernel 0 is the initiator of the lbts calculation but since it holds only models independent of any other (server), it will finish simulation very fast leaving the other kernel bereft of an updated lbts.
-The kernel holding the server will after simulation wait on the receiving kernel until it can prove all send messages can be deallocated.
+The kernel holding the server will after simulation wait on the receiving kernel until it can prove all sent messages can be deallocated.
 \begin{table}[lhtb]

### 2015-12-05 Rscript/devs-performance.R
cleaned up R script + put phold figure back in paper

@@ -46,43 +46,6 @@ process1ddata <- function(datatable, collabel) {
 	print(resultdata)
 }
 
-display1ddata <- function(perfdata, label) {
-	plot(perfdata$x, perfdata$cpu, pch=19, xlab="Width", ylab="Elapsed CPU Time (seconds)", main=label, col="blue")
-	abline(lm(perfdata$x ~ perfdata$cpu), col="red", lwd=2, lty=1)
-}
-
-display1ddata2 <- function(singledata, label) {
-	boxplot(time.elapsed..seconds. ~ width, data=singledata, xlab="Width", ylab="Elapsed CPU Time (seconds)", main=label)
-}
-
-display1ddata3 <- function(singledata, perfdata, label) {
-	indextable <- unique(singledata[, 3])
-	for (n in seq(length(indextable))) {
-			i <- indextable[[n]]
-			tmpdata <- singledata$time.elapsed..seconds.[singledata$width == i]
-			boxplot(tmpdata, xlab=sprintf("Width = %d", i), ylab="Elapsed CPU Time (seconds)", main=label)
-	}
-}
-
-group1 <- function() {
-	aconnectdata = read.table(file="aconnect_classic.csv",header=TRUE,sep=";")
-	dxconnectdata = read.table(file="dxconnect_classic.csv",header=TRUE,sep=";")
-
-	pdf("d:/tmp/myconnectgraphs.pdf")
-	
-	perfdata1 <- process1ddata(aconnectdata, TRUE)
-	display1ddata(perfdata1, "A Connect")
-	display1ddata2(aconnectdata, "A Connect")
-	display1ddata3(aconnectdata, perfdata1, "A Connect")
-	
-	perfdata2 <- process1ddata(dxconnectdata, TRUE)
-	display1ddata(perfdata2, "DX Connect")
-	display1ddata2(dxconnectdata, "DX Connect")
-	display1ddata3(dxconnectdata, perfdata2, "DX Connect")
-	
-	dev.off()
-}
-
 # 2-dimensional performance tests
 
 library(scatterplot3d)
@@ -152,61 +115,6 @@ displaydata <- function(perfdata, isdevstone, label) {
 		legend("bottomright", cex=0.7, c("Actual Value", "95 % Confid. Interval"), pch=c(19, 3), col=c("blue", "red"))
 	})
 	return(s3d)
-}
-
-displaydata2 <- function(perfdata1, label1, perfdata2, label2, isdevstone, label) {
[…]

### 2015-12-05 paper/4-performance.tex
cleaned up R script + put phold figure back in paper
Add memory usage phold and offer explaination for increased memory usage conservative.

@@
 PHold \cite{PHOLD} is a parallel oriented benchmark, sequential runtime is measured only as a baseline.
-A model selects the destination of a message at runtime using a state-saved RNG, which takes up a non-trivial amount of runtime.
+A model selects the destination of a message at runtime using a state-saved RNG, which takes up a non-trivial amount of runtime.\\
+\begin{figure}[h]
+	\includegraphics[width=.5\textwidth]{fig/fig4.eps}
+	\label{fig4.eps}
+\end{figure}
 
@@
 Despite all this, optimistic can for low R values quickly exploit uncertainty that slows down conservative in this benchmark.
+
 \subsubsection{Interconnect}
@@
 \paragraph*{Devstone}
-\begin{table}[lhtb]
+\begin{table}[htb]
 	\centering
@@
 %TODO next sentence needs rewriting --done?
-Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resultant higher latency) than the LBTS calculation in conservative. 
+Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resulting higher latency) than the LBTS calculation in conservative. 
 Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization.
@@
 Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels.
+\paragraph*{PHold}
+\begin{table}[htb]
+	\centering
+	\caption{Phold n 4 s 16 t1e6 r 10, unit MiB, 4 kernels (if parallel)}
+	\label{dtone_mem}
+	\begin{tabular}{| l | l | l | l | l |}
+		\hline
+		Adevs & AdevsCon &DX &DXCon &DXOpt\\ \hline
+		40 & x & 37 & 61 & 682  \\ \hline
+	\end{tabular}
+\end{table}
+With only 10\% of all messages being inter-kernel, we expect conservative to have memory consumption near that of the single threaded implementation, since intra-kernel messages are reclaimable after each round.
+The counterintuitive high memory usage can be explained by conservative's stalled round behaviour which occurs whenever a kernel cannot advance (eit $==$ time).
+In such a round messages are sent out but the kernel does not execute any transitions until it has received all input from all influencing kernels.
+\\
+With lookahead = $\epsilon$ this then leads to a high frequency of polling on shared null-times, which are used to determine lbts and thus garbage collection.
+The lbts calculation will not wait until a new value is found, since this can create unwanted contention with the simulation.
+The resulting longer time intervals within which no new lbts is found delay memory deallocation.
+Adevs' conservative fails to complete the benchmark under valgrind even with a significantly reduced load.\\
+Optimistic exhibits the expected high memory usage.
 \paragraph*{Interconnect}

### 2015-12-05 paper/1-introduction.tex
Last? changes. Increase memory interconnect ram profile.

@@
 In practice, dxex defaults to word sized unsigned integer.
-Although this significantly reduces the possible range of virtual time, it avoids approximation errors.
-Furthermore, the notion of $\epsilon$ as a minimum distance between two distinct time points need not to be established.
-This task is non-trivial for floating point approximations.\\
-Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernels ($2^{48}$ by default).
+Although this significantly reduces the possible range of virtual time, it avoids approximation errors. The notion of $\epsilon$ is trivial for integer, unlike floating point.\\
+Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernel ($2^{48}$ by default).
 If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$.

### 2015-12-05 paper/4-performance.tex
Last? changes. Increase memory interconnect ram profile.

@@
 Dxex uses integer timestamps, adevs uses floating point.
-Tests show there is little difference performance wise (both types fit in a machine word)\\
+Tests show there is little difference performance-wise (both types fit in a machine word).\\ The benchmarks are compiled using O3 and link-time optimization, with all IO disabled.
+The benchmarks have testing variants where the exact output can be verified.
 Benchmarks ran on a i7-2670QM, 16GiB Ram, Arch Linux (4.2.5 kernel), G++ 5.2.
@@
 %TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
+Whenever not specified, the benchmarks use 4 kernels for parallel simulation.
 \subsubsection{Devstone}
@@
 The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages.
-Again, this could be significantly reduced using lazy-cancellation.
 Once a revert happens the drift between kernels increases fast, increasing the likelihood of more reverts.
-Despite all this, optimistic can for low R values quickly exploit uncertainty that slows down conservative in this benchmark.
+Despite all this, optimistic can for low R values quickly exploit the uncertainty that slows down conservative in this benchmark.
 
@@
 \subsubsection{Measure}
-Adevs passes messages by value (in container by reference), we pass by pointer.
+Adevs passes messages by value, we pass by pointer.
 The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled.
@@
 Devstone (flattened) is allocated in a chain.
-Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to severe memory pressure.
+Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to increased memory pressure.
 Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels.
@@
 \\
-With lookahead = $\epsilon$ this then leads to a high frequency of polling on shared null-times, which are used to determine lbts and thus garbage collection.
+With lookahead $\epsilon$ this then leads to a high frequency of polling on shared null-times, which are used to determine lbts and thus garbage collection.
 The lbts calculation will not wait until a new value is found, since this can create unwanted contention with the simulation.
 The resulting longer time intervals within which no new lbts is found delay memory deallocation.
-Adevs' conservative fails to complete the benchmark under valgrind even with a significantly reduced load.\\
+Adevs' conservative fails to complete the benchmark under valgrind even with a significantly reduced load.
 Optimistic exhibits the expected high memory usage.
@@
 	\centering
-	\caption{Interconnect w 10 t5e5, unit MiB, 2 kernels (if parallel)}
+	\caption{Interconnect w 20 t5e5, unit MiB, 2 kernels (if parallel)}
 	\label{iconn_mem}
@@
 		Adevs & AdevsCon &DX &DXCon & DXOpt\\ \hline
-		39 & 39 & 35 & 43 & 164 \\ \hline
+		39 & 39 & 35 & 43 & 259 \\ \hline
 	\end{tabular}

### 2015-12-05 paper/6-conclusions.tex
Add memory usage phold and offer explaination for increased memory usage conservative.
Last? changes. Increase memory interconnect ram profile.

@@
 \subsubsection{Memory}
-Our optimistic implementation can benefit from a faster GVT algorithm such as FujiMoto's \cite{Fujimoto:1997:CGV:268403.268404} or more recently \cite{Bauer:2005:SND:1069810.1070159}. 
-With access to the memory subsystem in place (pools), optimistic could be constrained using per kernel quota preventing the simulation from exhausting memory.
+Our optimistic implementation can benefit from a faster GVT algorithm such as FujiMoto's \cite{Fujimoto:1997:CGV:268403.268404}  or more recently Bauer et al\cite{Bauer:2005:SND:1069810.1070159}. 
+With access to the memory subsystem already in place (pools), optimistic could be constrained using per kernel quota preventing the simulation from exhausting memory.
 While less sensitive to memory pressure, the same approach could benefit conservative.
@@
 \subsubsection{Hybrid}
-The optimistic implementation could use (null/eot/eit) from conservative to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.
-Conservative kernels could be extended with runtime information about influencing models, for example if one can guarantee a static dependency is not used for a fixed time-span, this can be removed for that period of (virtual) time.
+The optimistic implementation could use (null/eot/eit) from conservative to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.\\
+Conservative's eot calculation can be enhanced by using the entire dependency graph between kernels (not models).
+The implementation could detect dependency cycles between kernels and, instead of waiting on dependent kernels, advance as a synced group with better scaling.
+Lbts calculation could leverage this as well to reduce memory usage, though this is not that stringent in conservative.
 Ultimately the simulation could switch at runtime between protocols based on the information provided by activity tracking.

### 2015-12-07 paper/1-introduction.tex
Review

@@
 %TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
-The family of DEVS \cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms.
-Of interest in this paper are the 2 key formalisms: Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation.
-This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a simulation kernel's perspective only Atomic Models exist linked to each other by ports.
+The family of DEVS~\cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms.
+In this paper, we are in particular interested in Dynamic Structured DEVS~\cite{DSDEVS} and Parallel DEVS~\cite{ParallelDEVS}.
+We propose a DEVS implementation (called DEVS-Ex-Machina and referred to as "dxdev" throughout this paper) that is based on the
+the Direct Connection~\cite{SymbolicFlattening} algorithm.
+This algorithm flattens all hierarchically nested models which leads to, from a simulation kernel's perspective, a collection of Atomic Models linked to each other by ports.
 
@@
 %TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
-Parallel execution of a PDEVS simulation can lower overall runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
+One way to speed up the PDEVS simulation is by using parallel execution.
+By its very nature, this can lower the runtime considerably and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
 While the shared memory parallelism offered by most modern hardware does not raise the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation.
-\\
 
@@
 %TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
-Adevs \cite{adevs} offers a very fast conservative synchronized shared memory PDEVS simulator, but no optimistic synchronized variant. 
-The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
-The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code and without losing access to powerful low-level threading primitives.
+One of the drawbacks of adevs \cite{adevs}, a well-known PDEVS simulator which is performing well with conservative shared memory synchronization,
+is that it lacks support for optimistic synchronization.
+Using optimistic synchronization can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict.
+The matured parallelism features of C++11, on the other hand, created new opportunities as well.
+They allowed to write very efficient code giving access to powerful low-level threading primitives and yet guarantee standard-compliance and portability.
 % may need another pass on this one :(
@@
 %TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
-The usage of the Direct connection algorithm makes reusing adevs kernel hard. Devs-ex-machina (dxex) is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
+The usage of the Direct connection algorithm makes reusing existing kernels (such as adevs or PythonPDEVS~\cite{PythonPDEVS}, also known as PyPDEVS, hard.
+Dxex is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
 The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance.
-As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
-The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
+As in PyPDEVS, a model has to be written only once for use in the different simulation kernels.\\
+In order to monitor and investigate simulations, dxex uses a tracing framework which is derived from the one used in PyPDEVS.
 \subsection{Time}
-The DEVS formalisms have $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time.
-In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.
+The theoretical DEVS formalisms all have $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time.
+In principle, any type with well-defined operators could be used, but, from a performance point of view, a type fitting in a machine word offers obvious advantages.
 In practice, dxex defaults to word sized unsigned integer.
-Although this significantly reduces the possible range of virtual time, it avoids approximation errors. The notion of $\epsilon$ is trivial for integer, unlike floating point.\\
-Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernel ($2^{48}$ by default).
-If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$.
+Although this significantly reduces the possible range of virtual time, it avoids approximation errors.
+The notion of $\epsilon$, moreover, is trivial for an integer representation, unlike floating point.\\
+In addition, our time representation is enriched with a causality field the range of which is equal or greater than the maximum number of models in a single kernel ($2^{48}$ by default).
+If $A$ and $B$ are imminent at time $t$, then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$.
 This allows a deterministic ordering of concurrent events.

### 2015-12-07 paper/2-background.tex
Review

@@
 Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
-This requires the concepts of earliest output time (eot) and earliest input time (eit), which define the timespan within which a model can safely advance.
-\\ The eit of any model is the minimum of all eot values of (directly) influencing models.
-A model can simulate up to (but not including) eit, then waits until that value is increased.
-An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime.
-A model that can influence another, but never does, can severely slow down the protocol.
-Deadlock between models that influence each other and end up waiting on each other can be broken/avoided by a variety of means.
-In this simulator, the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used.
+This requires the concepts of earliest output time (eot) and earliest input time (eit), which define the timespan within which a model can safely advance.\\
+The eit of any model is the minimum of all eot values of (directly) influencing models. A model can simulate up to (but not including) eit and then waits until that value is increased. \\
+An important disadvantage here is that the relation that calculates the dependencies, is always defined at model (link) creation time, and not at runtime.
+A model that can influence another model theoretically, but never does in real life, can severely slow down the protocol.
 \\
-In our implementation null-time is the timestamp a model is guaranteed to have passed in simulation.
-More precisely, a null message of time t is a guarantee that any output with timestamp t-$\epsilon$ is already sent.\\
+Deadlocks between models that influence each other and end up waiting on each other, can be broken/avoided by a variety of means.
+In dxex, the CMB~\cite{Chandy:1981:ADS:358598.358613} null-message protocol is used.
+\\
+In our implementation, null-time is the time stamp a model is guaranteed to have passed in simulation.
+More precisely, a null message of time $t$ guarantees that any output with time stamp $t-\epsilon$ has already been sent.\\
 Conservative synchronization relies explicitly on information provided by the model creator in the form of a lookahead, a relative timespan during which the model is insensitive to outside events.
-This lookahead can be non-trivial to calculate, a simulation writer will in general not be able to determine the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
+This lookahead can be non-trivial to calculate; therefore, a simulation writer will in general not be able to determine the exact lookahead of the models involved in an experiment without having run the experiment.\\
+Conservative kernels can even operate if there are cyclic dependencies involved, but this can come at a severe performance penalty, as seen in section 4.
+
 \subsection{Optimistic Synchronization}
 %TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
-Optimistic synchronization allows causality errors to occur but recovers using a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.
-Whenever a kernel receives an event with a timestamp in the kernel's past, the state of the kernel (and all models)reverts to that time.
-Because an optimistic kernel does not wait for other kernels, it can achieve a lower runtime than conservative.
-The cost of this approach is higher complexity of the code and larger memory usage for state/message saving.
-In contrast to conservative, optimistic does not rely on any domain specific information, it is only sensitive to runtime use of connections, not the probability that they might occur.
-%reword connections ?
-\\
-If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts.
-The cost of reverts can be reduced by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
+In the case of optimistic synchronization, causality errors can occur but these will be reverted via a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.\\
+Whenever a kernel receives an event with a time stamp in the kernel's past, the state of the kernel (and all models) is reverted to that time.\\
+Because an optimistic kernel does not wait for other kernels, it can run much faster than a conservative kernel.
+The cost of this approach, however, is higher complexity of the code and a larger memory usage for the state and message saving.
+In contrast to conservative synchronization, optimistic synchronization does not rely on any domain specific information.
+Therefore, it is only sensitive to the runtime use of connections, not the probability that they might occur.
+%reword connections ?\\
+If the (runtime) dependency graph contains a cycle, optimistic synchronization might lead to a series of cascading reverts.
+The cost of reverts can be reduced by lazy cancellation and/or lazy re-evaluation~\cite{FujimotoBook}.
+
 \subsection{Global Virtual Time}
-To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988}.
-In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event.
-\\ Intuitively this is the simulation timepoint that is certain to be preserved, corresponding exactly with a sequential simulation.
-In conservative, the minimum simulation time of all kernels is the LBTS, or in terms of null messages: the least timestamp of any null message in transit.
-The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .
+To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988} or GVT.
+In optimistic simulations, GVT is defined as the lowest time stamp of any unprocessed event.\\
+Intuitively, this is the simulation time point that is certain to be preserved, corresponding exactly with a sequential simulation.
+In conservative, the minimum simulation time of all kernels is the lower bound of time stamp or LBTS. In terms of null messages this corresponds to the least time stamp of any null message in transit.
+The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, etc.
 
@@
 Our kernels use a wide range of threading primitives and atomic operations.
-As an example, eot/eit/nulltime are exchanged not as messages, but as reads/writes to atomic fields shared by all kernels.
+As an example, eot/eit/nulltime are not exchanged as messages, but as reads/writes to atomic fields shared by all kernels.
 This avoids the otherwise unavoidable latency penalty by mixing simulation messages with synchronization messages.
-For an in-depth study, see \cite{CPE:CPE3007}.
+For an in-depth study, we refer to~\cite{CPE:CPE3007}.
 Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform.

### 2015-12-07 paper/3-features.tex
Review

@@
 %TODO explain that we are briefly going over the main features
-\subsection{Based on PythonPDEVS}
+\subsection{Based on PyPDEVS}
 %TODO say that it is based on PythonPDEVS --done
-The simulator is based on PythonPDEVS, and provides the following features: 
+Dxex is based on PyPDEVS, and provides the following features: 
 \begin{enumerate}
@@
 	\item Dynamic Structured DEVS
-	\item Termination function.
-If specified, a termination function is applied every simulation round to each model to test whether the simulation can terminate.
-Only available in single-threaded simulation.
-	\item State/Message can have any payload type. Different message types can be used together within the same simulation.
-	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
-	\item Optimistic and Conservative synchronization of PDEVS.
+	\item Termination function: if specified, a termination function is applied to each model every simulation round to test whether the simulation can terminate.
+This feature is only available in single-threaded simulations.
+	\item The State/Message objects can have any payload type. Different allows different message types to be used within the same simulation.
+	\item Tracing: An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
+	\item Optimistic and Conservative synchronization
 \end{enumerate}
-The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use.
-For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.
-Another example is the state saving mechanism, which is only used for optimistic parallel simulation and has no performance impact in conservative parallel simulation.\\
-The tracing is not comparable to adevs's listener interface.
-To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points.
-Furthermore, the framework itself has to be thread safe and deterministic so that a simulation will always produce the exact same output.
-%TODO Decide whether we should list the functionality that is _not_ implemented
+Furthermore, the implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use.
+For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had an adverse impact on the runtime, even when not in use.
+Another example is the state saving mechanism, which is only used for optimistic parallel simulation and has no performance impact in a conservative parallel simulation.\\
+Our tracing implementation is not comparable to adevs's listener interface.
+To be usable in an optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points.
+Furthermore, the framework itself has to be thread safe and deterministic so that a simulation will always produce the exact same output.\\
 The following features from PyPDEVS are not present
@@
 	\item Activity tracking and relocation
-	\item Serialization
-	\item Interactive control. 
+	\item Serialization: in this context this is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp. 
+	\item Interactive control
 	\item Distributed simulation
 \end{enumerate}
-Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.
-%This state saving has no impact in a single threaded or conservative kernel.
--> already mentioned above
-\\
-Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation.
+In dxex, the model allocation is realized by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation.
 If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
-Debugging tools such as a logger and a graph visualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.
-
+In addition, debugging tools such as a logger and a graph visualizer are included that can track activity with respect to the allocation for later inspection (but not are not online and as dynamic as is in PyPDEVS).\\
 %TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
@@
 %TODO explain rationale: why different synchronization protocols
-Synchronization is required for PDEVS to prevent or recover from causality violations if the simulation is executed in parallel.
-Preventing causality violations (conservative) typically requires domain-specific information to compensate for performance loss, while recovering from causality errors (optimistic) requires a state to recover to.
+For parallel executions, synchronization is required for PDEVS to prevent causality violations from happening or to recover from causality violations.
+Preventing causality violations (conservative) typically requires domain-specific information to compensate for the performance loss, while recovering from causality errors (optimistic) requires the calculation of a state to revert to.
 
 \subsubsection{Conservative}
-A conservative kernel will determine which kernels it is influenced by.
+In case of conservative synchronization, any kernel will determine which kernels it is influenced by.
 This information is constructed from the incoming connections on all hosted models.
 The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
-A model should provide a lookahead function which returns, relative from the current time, the timespan during which the model cannot change state due to an external event.
-This information is collected for all models hosted on the kernel, and the minimum is set as the lookahead of that kernel.
+A model should provide a lookahead function which returns, relative to the current time, the timespan during which the model cannot change state due to an external event.
+This information is collected for all models hosted on the kernel and the minimum time is set as the lookahead of that kernel.
 \\
 The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
-For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i])  - \epsilon $.\\
+For the garbage collection (of sent messages), the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i])  - \epsilon $.\\
 %This is one of the operations that can benefit from using a relaxed memory ordering.
@@
 \subsubsection{Optimistic}
-The optimistic kernel requires from the hosted model only that copying the state is well-defined, which is provided in the base State class for the user. 
+The optimistic kernel requires from the hosted model that copying the state is done carefully (avoid unnecessary copies and make a copies whenever necessary).\\
 The kernels use Mattern's \cite{mattern}
@@
 This process runs asynchronously from the simulation itself.
-Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states and (anti)messages.
-
+Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states and (anti)messages.\\
 %TODO explain core infrastructure a bit, maybe with a diagram or so, which clearly shows that everything is shared, except the "synchronization" part
 %TODO claim that this allows you to easily switch between both methods, to always chose the best one
-The user need only provide one implementation of a model for use with both synchronization protocols.
-A lookahead function is desired to accelerate conservative, but is not required.
-In the absence of a user supplied lookahead, the kernel assumes it cannot predict beyond its current time + $\epsilon$, creating a lockstep simulation.
-% implementation detail ?
-\\
-The implementation details such as defining the copy semantics of a State are provided (but can be overridden). \\
+The user must only provide one implementation of a model that can be used for both synchronization protocols.
+A lookahead function is desired to accelerate the conservative protocol, but is not required.
+In the absence of a user supplied lookahead, the kernel assumes it cannot predict beyond its current time $t+\epsilon$, creating a lockstep simulation.
+% implementation detail ?\\
+The user is shielded from the multi-threaded aspect of the kernel.\\
 From the user's perspective, the multi-threaded aspect of the kernel is not exposed. % payload protected ?
@@
 %TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
-Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks: 
+We now discuss a number of bottlenecks that were discovered during the profiling of several benchmarks.
 \subsubsection{Heap}
-Events are passed by pointer, avoiding a possibly expensive copy of the payload at the cost of allocation overhead.\\
-Allocation becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and events, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}.
-In this way allocating threads do not block each other.
+In dxex, events are always passed with pointers and thus involve a possibly expensive copy of the payload caused by allocation and copying overhead.
+In highly connected models, this allocation cost can become prohibitively expensive, so to reduce that overhead we use thread\_local memory pools for states and events, and, optionally, replace the system malloc with calls to tcmalloc~\cite{tcmalloc}.
+In this way, allocating threads do not block each other.
 If desired, arena-pools are available for single-threaded simulation.
 A disadvantage is increased complexity in ownership semantics.
-The creating kernel is responsible for destruction, but this can only be guaranteed up to GVT.
-Experiments with synchronized pools proved slower than malloc/free.
-\\
-The kernels initially used strings as identifiers, as is done in PyPDEVS.
-Profiling quickly indicated this to be a performance bottleneck.
-C++ strings are heap allocated variable sized objects with an atomic reference count, not immutable objects as in Python.
-Access of that reference count across threads is expensive, as are the calls to malloc/free the string implementation makes to create/destroy new objects, or copy existing.\\
-Strings are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name.
-Once simulation starts all objects use integral identifiers for performance.
-This also increased usage of the constexpr feature of C++11 in, amongst others, timestamps and message headers.
+The creating kernel is responsible for destruction, but this can only be guaranteed up to the GVT.
+Experiments with synchronized pools proved to be slower than implementations with the standard malloc/free.\\
+Initially, dxes was using strings as identifiers as was also done in PyPDEVS.
+Profiling quickly indicated that this caused a real performance bottleneck.
+In C++, strings are heap allocated variable sized objects with an atomic reference count and not not immutable objects as in Python.
+Access of that reference count across threads turned out to be quite expensive, as are the calls to malloc/free the string implementation makes to create/destroy new objects or copy existing objects.\\
+Strings, however, are more intuitive to work with from a user's perspective, so, as a compromise, we allowed the user to reference models/ports by string name.
+Once the simulation starts however, all objects use integral identifiers.
+This also increased the usage of the constexpr feature of C++11 in, amongst others, time stamps and message headers.
 \subsubsection{Raw pointers}
 While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced.
-Used across threads, the reference counting becomes very expensive, and the (de)allocating of memory caused significant contention between threads.
-Models are still held by a smart pointer, as is a kernel, but a message is a raw pointer to compacted memory.
+Used across threads, the reference counting became very expensive, and the (de)allocation of memory caused significant contention between threads.
+So for the models and the kernels, dxex is still using smart pointers whereas for the messeages, a raw pointer to compacted memory is being used.
 \subsubsection{Locking}
-Locking between kernels uses mostly atomic operations, where we can occasionally leverage memory orderings to only pay for synchronization when we need it.
-Messages are exchanged via a shared set of queues each with a dedicated lock.\\
-On a higher level, we avoid the sending of synchronization messages entirely by writing the timestamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
+Locking between kernels uses mostly atomic operations and, occasionally, we can leverage memory orderings to only pay for synchronization when we need it.
+Messages, on the other hand, are exchanged via a shared set of queues each with a dedicated lock.\\
+On a higher level, we avoid the sending of synchronization messages entirely by writing the time stamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
 \subsubsection{Schedulers}
-PyPDEVS has a wide range of schedulers for the user to choose from, with performance of each depending on the simulation type.
-Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before.
-Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depending on the amount of items to update.
+PyPDEVS has a wide range of schedulers to choose from with varying performance depending on the simulation type.
+Profiling showed that, the heap implementation used in adevs was faster than any of the schedulers we had tested before (in a C++ environment).
+Unlike most node based heaps, this scheduler uses a fixed size array where its heap is rebuilt or modified in place depending on the amount of items to update.
 Items are only updated, never removed.

### 2015-12-07 paper/4-performance.tex
Review

@@
 %TODO (spend some time here, as this is also important)
-\subsubsection{Preamble}
-The benchmark models are identical in functionality (between adevs/dxex versions).
-Dxex uses integer timestamps, adevs uses floating point.
-Tests show there is little difference performance-wise (both types fit in a machine word).\\ The benchmarks are compiled using O3 and link-time optimization, with all IO disabled.
-The benchmarks have testing variants where the exact output can be verified.
-Benchmarks ran on a i7-2670QM, 16GiB Ram, Arch Linux (4.2.5 kernel), G++ 5.2.
-Interpretation of the benchmark results is done using call-graphs generated by perf, a selection of those call-graphs can be found in the code repository.
-For the runtime plots perf is used in combination with R.\\
-Note that a fixed timeadvance (e.g. devstone) is set at 100.
+In our benchmarks, we will consistently compare the performance of dxex with adevs. The implemented models for the each benchmark are functionally identical.\\
+Dxex uses integer timestamps whereas adevs uses floating point time representation. Tests show there is little difference performance-wise (both types fit in a machine word).\\
+The benchmarks have been compiled using the {\tt -O3} option and link-time optimization, with all IO disabled.
+To compare the models in details, we have run the different benchmarks with different parameters.\\
+The benchmarks have been run on a i7-2670QM, 16GiB RAM, Arch Linux (4.2.5 kernel) and G++ 5.2.
+Interpretation of the benchmark results is done using call-graphs generated by perf.
+A selection of those call-graphs is available in the code repository.
+For the runtime plots, perf is used in combination with R.\\
+%Note that a fixed timeadvance (e.g. devstone) is set at 100.
 \subsubsection{Devstone}
-The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this whereas adevs needs to walk the coupled model's structure to pass events.\\
+The Devstone \cite{DEVStone} benchmark is highly hierarchical.
+As can be seen in Figure~\ref{fig:Devstone}, dxex is outperforming adevs considerably.
+This can be explained by the fact that, using Direct connection, the hierharchical structure is being flattened whereas, in the adevs case, considerable overhead is caused by traversing the coupled model's structure to pass events.
 \begin{figure}[h]
 	\includegraphics[width=.5\textwidth]{fig/fig1.eps}
-	\label{fig1.eps}
+	\caption{Devstone}
+	\label{fig:Devstone}
 \end{figure}
@@
 \subsubsection{PHold}
-PHold \cite{PHOLD} is a parallel oriented benchmark, sequential runtime is measured only as a baseline.
-A model selects the destination of a message at runtime using a state-saved RNG, which takes up a non-trivial amount of runtime.\\
+PHold \cite{PHOLD} is a benchmark that focuses in particular on parallel execution; the sequential runtime is therefore only added as a baseline.
+A model selects the destination of a message at runtime using a state-saved RNG, which takes up a non-trivial amount of runtime.
+The results of the benchmark are visualized in Figure~\ref{fig:PHold}.
+%% needs further clarification...
 \begin{figure}[h]
 	\includegraphics[width=.5\textwidth]{fig/fig4.eps}
-	\label{fig4.eps}
+	\caption{PHold}
+	\label{fig:PHold}
 \end{figure}
@@
 \subsubsection{Interconnect}
-Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in dependencies between models.
-As the model count increases, we see the expected quadratic increase in runtime for both adevs and dxex, but an increasing penalty for dxex.
-Profiling shows this is entirely due to the heap allocation of messages, which even though minimized by using memory pools remains significant.\\
+Interconnect~\cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in dependencies between models.
+As the model count increases, we see (see Figure~\ref{fig:Interconnect} the expected quadratic increase in runtime for both adevs and dxex, but an increasing penalty for dxex.
+Profiling shows this is entirely due to the heap allocation of messages, which even though minimized by using memory pools remains significant.
 \begin{figure}[h]
@@
 	\label{fig3.eps}
+	\caption{Interconnect}
+	\label{fig:Interconnect}
 \end{figure}
@@
 %TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
-Whenever not specified, the benchmarks use 4 kernels for parallel simulation.
+By default, the benchmarks use 4 kernels for parallel simulation.
 \subsubsection{Devstone}
 The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages.
-For optimistic, this can cause more reverts since the kernels will start to drift faster as the model-count increases.
-Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases.
-Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*(n-1)$ transitions to activate the last model in the chain.
-For parallel this can be reduced to $\textup{timeadvance}()*(n\frac{kernels-1}{kernels} -1)$ before the last kernel becomes active.\\
+For the optimistic case, this can cause more reverts since the kernels will start to drift faster as the model-count increases.
+Furthermore, optimistic synchronization is quite sensitive to an increase in kernels, since the delay before a revert propagates, increases.
+This benchmark requires a specific warm-up time; for $n=d\times w$ models, it takes $\textup{timeadvance}()*(n-1)$ transitions to activate the last model in the chain.
+When using parallel processing, this can be reduced to $\textup{timeadvance}()*(n\frac{kernels-1}{kernels} -1)$ before the last kernel becomes active.\\
+Figure~\cite{fig:DevstonParallel} shows that dxex outperformance adevs (in both optimistic and conservative case).
 \begin{figure}[ltbh]
 	\includegraphics[width=.5\textwidth]{fig/fig2.eps}
-	\label{fig2.eps}
+	\label{fig:DevstoneParallel}
+	\caption{DevStone parallel}
 \end{figure}
@@
 %Phold
-In Phold allocation is specified in the benchmark itself.
+In Phold, the allocation is specified in the benchmark itself.
 Each kernel manages a single node with a constant set of sub-nodes.
-The parameter R (here 10) determines the percentage of remote destination models.\\
-The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative.
-Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$.
+The parameter $R$ (which is here set to $10$) determines the percentage of remote destination models.\\
+The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing the conservative case.
+The lookahead is $\epsilon$, so the conservative case spends most of its time crawling in steps of $\epsilon$.
 Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation.
-For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour.
-In a non-cyclic simulation with a non-trivial lookahead (e.g.
-devstone), that choice however does pay off.
-Adevs's lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code.
-Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible.
-With the percentage of remotes = 100 phold reflects interconnect in behaviour, which is why the R value is not dimensioned here.
+For $N$ kernels, each kernel has to query the null-time of $N-1$ kernels, resulting in O($N^2$) polling behaviour.
+In a non-cyclic simulation with a non-trivial lookahead (like in, e.g., Devstone), that choice does pay off (see Figure~\ref{fig:DevstoneParallel}).
+Adevs's lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code.\\
+The optimistic case suffers little from the above problems; due to the high interconnectivity, however, a cascading revert is still possible.
+With the percentage of remotes equal to $100$, Phold reflects interconnect in behaviour, which is why the $R$ value is not dimensioned here.
 A revert is very expensive in PHold due to our usage of C++11's random nr generators.
@@
 Once a revert happens the drift between kernels increases fast, increasing the likelihood of more reverts.
-Despite all this, optimistic can for low R values quickly exploit the uncertainty that slows down conservative in this benchmark.
+Despite all this, optimistic can for low $R$ values quickly exploit the uncertainty that slows down conservative in this benchmark.
 
 \subsubsection{Interconnect}
-In Interconnect
-the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
-Allocation cannot avoid cycles, the resulting dependency graph between kernels remains a complete graph. The runtime dependency graph is almost immediately equal to the static graph.
-Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions.
+In Interconnect, the set of atomic models form a complete graph (w.r.t. connections); each model broadcasts messages to the entire set.\\
+Allocation cannot avoid cycles and the resulting dependency graph between kernels remains a complete graph. The runtime dependency graph is almost immediately identical to the static graph.\\
+The conservative case still shows the same issues as in PHold, with the key difference , for a fixed time advance, the lookahead is equal to the timespan between transitions.
 The scaling issue is identical as with PHold.\\
-In this benchmark optimistic runs very quickly out of memory.
-With c kernels and N atomic models, a single revert undoing k transitions will lead to $(N-1)\times k$ messages that need to be recreated, plus $(N-1) \frac{c-1}{c} \times k$ anti-messages that need to be send.
+In this benchmark, the optimistic case runs very quickly out of memory.
+With $c$ kernels and $N$ atomic models, a single revert undoing $k$ transitions will lead to $(N-1)\times k$ messages that need to be recreated, plus $(N-1) \frac{c-1}{c} \times k$ anti-messages that need to be sent.
 
@@
 %TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
-The priority benchmark is composed of a single server generating a stream of $0<=m<=n$ messages at fixed time intervals, interleaved with a probability p for a priority message, to n receivers.
-\\ This defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph.
+The priority benchmark is composed of a single server generating a stream of $0\leq m \leq n$ messages at fixed time intervals, interleaved with a probability $p$ for a priority message to $n$ receivers (see Figure~\ref{fig:NetworkModel}).\\ This defaults the lookahead for the receivers to $\epsilon$, but this time there is no scaling effect, nor are there cycles in the dependency graph.
 This model therefore highlights the basic strengths/weaknesses of both synchronization protocols.
-Receiving models are allocated on another kernel than the server, and have a internal transition so that they will not wait for the incoming messages.
-\input pqueue.tex \\
+Receiving models are allocated on another kernel than the server and have an internal transition so that they will not wait for the incoming messages.
+\begin{center}
+\begin{figure}
+\input pqueue.tex
+\caption{The priority network model}
+\label{fig:NetworkModel}
+\end{figure}
+\end{center}
 A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create.
 The kernel holding the server will never revert since it is a source in the dependency graph.
-Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold.
-The overhead in optimistic is entirely due to the factor m, which will quickly dominate in increasing buffers of received messages.
+The optimistic case will therefore not suffer the same performance hit in recreating the states as it does in PHold.
+The overhead in the optimistic case is entirely due to the factor $m$, which will quickly dominate in increasing buffers of received messages.
 \begin{figure}[ltbh]
 	\includegraphics[width=.5\textwidth]{fig/fig5.eps}
-	\label{fig5.eps}
+	\label{fig:Priority}
+	\caption{Priority}
 \end{figure}
-Interestingly, the parameter p does not clearly benefit either synchronization protocol.
-While this removes any possibility for a lookahead, conservative can quickly bridge the timespan between fixed messages since there is no cyclic dependency.
-Lightweight states prevent performance loss due to reverts in optimistic, only the overhead in event handling in optimistic eventually becomes the deciding factor.
+Interestingly, the parameter $p$ does not clearly favour either synchronization protocol (see Figure~\ref{fig:Priority}).
+While this removes any possibility for a lookahead, the conservative case can quickly bridge the timespan between fixed messages since there is no cyclic dependency.
+Lightweight states prevent performance loss due to reverts in the optimistic case; only the overhead in event handling in the optimistic case eventually becomes the deciding factor.
 
@@
 This is one reason why memory utilization is best measured by peak allocation.
-Profiling is done using Valgrind's massif tool.
-\cite{Nethercote:2007:VFH:1273442.1250746}
-Platform used for memory profiling has an i5-3317U Intel cpu and 8GiB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
+Profiling is done using Valgrind's massif tool~\cite{Nethercote:2007:VFH:1273442.1250746}.
+The platform used for memory profiling has an i5-3317U Intel CPU and 8GiB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).
 \subsubsection{Measure}
-Adevs passes messages by value, we pass by pointer.
-The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled.
-For the OS and/or user this is the actual memory footprint of the application.
-It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernels.
+Adevs passes messages by value, dxex passes a pointer.
+The runtime effects of this choice have already been demonstrated in the Interconnect benchmark, so, in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled.
+For the OS and/or user, this is the actual memory footprint of the application.
+It is important to note that, especially in the optimistic case, not all this memory is always in use by the kernels.
 During simulation, the pools will generally not return deallocated memory to the OS, but keep it for later reuse.
@@
 	\centering
+	\begin{tabular}{| l | l | l | l | l |}
+		\hline
+		adevs & adevs con &dxex &dxex con&dxex opt\\ \hline
+		44 & 70 & 42 & 75 & 363  \\ \hline
+	\end{tabular}
 	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
+	\label{dtone_mem}
+\end{table}
+Since, in the conservative case, messages are passed by pa ointer, a GVT/LBTS implementation is required to organise the garbage collection.
+This inevitable delay explains the higher memory usage compared to adevs.\\
+%TODO next sentence needs rewriting --done?
+Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resulting higher latency) than the LBTS calculation in the conservative case. 
+Moreover, the differences in LP virtual times are far larger compared to conservative time synchronization.
+All these factors explain the heavier memory usage.
+Devstone (flattened) is allocated in a chain.
+Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to an increased memory pressure.
+Unlike conservative and sequential execution, memory usage in the optimistic case varies greatly depending on scheduling of kernel threads and drifting between kernels.
+\paragraph*{PHold}
+\begin{table}[htb]
+	\centering
 	\label{dtone_mem}
@@
 		\hline
-		Adevs & AdevsCon &DX &DXCon &DXOpt\\ \hline
-		44 & 70 & 42 & 75 & 363  \\ \hline
-	\end{tabular}
-\end{table}
-Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection. This inevitable delay explains the higher memory usage compared to adevs.\\
-%TODO next sentence needs rewriting --done?
-Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resulting higher latency) than the LBTS calculation in conservative. 
-Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization.
-All these factors explain the heavier memory usage.
-Devstone (flattened) is allocated in a chain.
-Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to increased memory pressure.
-Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels.
-\paragraph*{PHold}
-\begin{table}[htb]
-	\centering
-	\caption{Phold n 4 s 16 t1e6 r 10, unit MiB, 4 kernels (if parallel)}
-	\label{dtone_mem}
-	\begin{tabular}{| l | l | l | l | l |}
-		\hline
-		Adevs & AdevsCon &DX &DXCon &DXOpt\\ \hline
+		adevs & adevs con &dxex &dxex con &dxex opt\\ \hline
 		40 & x & 37 & 61 & 682  \\ \hline
 	\end{tabular}
+	\caption{Phold n 4 s 16 t1e6 r 10, unit MiB, 4 kernels (if parallel)}
 \end{table}
@@
 	\centering
-	\caption{Interconnect w 20 t5e5, unit MiB, 2 kernels (if parallel)}
 	\label{iconn_mem}
@@
 		\hline
-		Adevs & AdevsCon &DX &DXCon & DXOpt\\ \hline
+		adevs & adevs con &dxex &dxex con & dxex opt\\ \hline
 		39 & 39 & 35 & 43 & 259 \\ \hline
 	\end{tabular}
+	\caption{Interconnect w 20 t5e5, unit MiB, 2 kernels (if parallel)}
 \end{table}
@@
 The priority network model is detailed in section 4.1.\\
-Conservative's high memory usage is largely due to the garbage collection implementation.
-Kernel 0 is the initiator of the lbts calculation but since it holds only models independent of any other (server), it will finish simulation very fast leaving the other kernel bereft of an updated lbts.
+The high memory usage of the conservative case is largely due to the garbage collection implementation.
+Kernel 0 is the initiator of the LBTS calculation but since it holds only models independent of any other (server), it will finish simulation very fast leaving the other kernel bereft of an updated LBTS.
 The kernel holding the server will after simulation wait on the receiving kernel until it can prove all sent messages can be deallocated.
@@
 	\centering
-	\caption{Priority model n 128, m 16, p 10,  t2e8, unit MiB, 2 kernels}
 	\label{pmod_mem}
@@
 		\hline
-		DX &DXCon &DXOpt\\ \hline
+		dxex &dxex con &dxex opt\\ \hline
 		35 & 450 & 651\\ \hline
 	\end{tabular}
+	\caption{Priority model n 128, m 16, p 10,  t2e8, unit MiB, 2 kernels}
 \end{table}

### 2015-12-07 paper/5-related-work.tex
Review

@@
 Dxex is closely related to PyPDEVS in design and philosophy.
-PyPDEVS allows anyone who grasps the PDEVS formalisms to immediately simulate his/her model without having to consider the kernel implementation.A Python based implementation offers the advantage of very fast prototype/run/evaluate cycles, this can't be matched by a C++ simulator but it should be noted that once the kernels are compiled (shared libraries), the actual compilation time of the model is small enough to make prototyping possible.
+PyPDEVS allows anyone who grasps the PDEVS formalisms to immediately simulate his/her model without having to consider the kernel implementation.
+A Python based implementation offers the advantage of very fast prototype/run/evaluate cycles, this can't be matched by a C++ simulator.
+It should be noted that, once the kernels have been compiled (in shared libraries), the actual compilation time of the model is small enough to make prototyping possible.
 Advanced features such as activity based relocation and the performance gains this results in, are still unique to PyPDEVS.
@@
 %Tracing ?
-By virtue of not flattening Coupled Models, adevs's performance degrades in increasingly hierarchical models.
-Dxex manages lookahead with more overhead than adevs, leading to a performance difference as the model count increases.
-Finally, adevs employs the full dependency graph, contrasted to dxex kernels which only observe 1-edge removed nodes.
+Due to the fact that adev does not flatten the Coupled Models, it's performance degrades significantly if the hierarchy becomes deeper.
+Dxex, on the other hand, manages lookahead with more overhead than adevs, leading to a performance difference as the model count increases.
+Finally, adevs employs the full dependency graph this in contrast with dxex kernels, which only observe 1-edge removed nodes.
 %checkme
@@
 Different projects on CD++ offer conservative (CCD++) as well as optimistic (PCD++) parallel simulation.
-In contrast to our single program, with a non-fragmented code-base, neither projects offer both synchronization protocols.
+In contrast to dxex,  neither projects offer both synchronization protocols.
 CD++ relies on the WARPED kernel.
 It is a middleware that provides memory, event, file, time and communication scheduling.
-We did not use the WARPED kernel (nor the underlying MPI), dxex is designed specifically for a shared memory architecture and as such any middleware, while feature-rich, would have lead to unacceptable overhead.
+We did not use the WARPED kernel (nor the underlying MPI) because dxex is designed specifically for a shared memory architecture and, as such, any middleware, however feature-rich, would have lead to an unacceptable overhead.

### 2015-12-07 paper/6-conclusions.tex
Review

--- 2015-12-01 paper/6-conclusions.tex
@@
+\section{Future work}
+%TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))
+\subsection{Memory}
+Our optimistic implementation can benefit from a faster GVT algorithm such
+ as used in~\cite{Fujimoto:1997:CGV:268403.268404}  or, more recently in~\cite{Bauer:2005:SND:1069810.1070159}. 
+With access to the memory subsystem already in place (pools), the optimistic case could be constrained using per kernel quota, preventing the simulation from exhausting memory.
+While less sensitive to memory pressure, the same approach could improve the conservative case.
+\subsection{Activity}
+As shown in~\cite{PythonPDEVS_ACTIMS}, activity and allocation of models across kernels is a key aspect in achieving high performance in any parallel implementation.
+Allocating models so that there are no dependency cycles between their containing kernels is a first step, but not always possible.
+For the optimistic case, one can use re-allocation to break (runtime dependency) cycles and/or perform load-balancing.
+If kernels are unevenly balanced, they will begin to drift fast, causing increasingly more reverts.
+In dxex, we implement a limited framework to track model activity for debugging purposes; this could be extended to enable the above strategies.
+\subsection{Hybrid}
+The optimistic implementation could use (null/eot/eit) from the conservative case to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.\\
+The eot calculation in the conservative case can be enhanced by using the entire dependency graph between kernels (not models).
+The implementation could detect dependency cycles between kernels and, instead of waiting on dependent kernels, advance as a synced group with better scaling.
+LBTS calculation could leverage this as well to reduce memory usage, though this is not that stringent in the conservative case.
+Ultimately, the simulation could switch at runtime between protocols based on the information provided by activity tracking.
+This requires the above mentioned framework, and can only be done at timepoints where a GVT/LBTS time is agreed upon between the kernels.
+From the model's perspective, no changes are needed, although a non-trivial lookahead is obviously desired.
+
+\section{CONCLUSIONS}
 %TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
-Both synchronization implementations offer good performance in differing simulations, in some simulations dxex outperforms adevs whereas in others we can still improve. 
-The optimistic implementation uses an aggressive TimeWarp variant, this needs to be extended with lazy evaluation/cancellation to function in cyclic simulations. 
-\subsection{Future work}
-%TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))
-\subsubsection{Activity}
-As shown in \cite{PythonPDEVS_ACTIMS} activity and allocation of models across kernels is a key aspect in achieving high performance in any parallel implementation.
-Allocating models so that there are no dependency cycles between their containing kernels is a first step, but not always possible.
-For optimistic one can use re-allocation to break (runtime dependency) cycles and/or perform load-balancing.
-If kernels are unevenly balanced they will begin to drift fast, causing increasingly more reverts.
-There is a limited framework implemented to track model activity for debugging purposes, this could be extended to enable the above strategies.
-\subsubsection{Hybrid}
-The optimistic implementation could use (null/eot/eit) from conservative to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.
-Conservative kernels could be extended with runtime information about influencing models, for example if one can guarantee a static dependency is not used for a fixed time-span, this can be removed for that period of (virtual) time.
-Ultimately the simulation could switch at runtime between protocols based on the information provided by activity tracking.
-This requires the above mentioned activity tracking framework, and could only be done if a GVT/LBTS time is agreed upon between the kernels.
-From the model point of view no changes are needed, although a non-trivial lookahead is obviously desired.
+In this paper, we presented dxex, a new C++-based PDEVS implementation supporting both sequential and parallel processing models.\\
+We showed that, in the sequential case, dxex outperforms adevs in hierarchical simulations and that adevs produces better results in broadcast-like simulations.\\
+In the parallel case, the results are more subtle.
+The sensitivity to broadcast simulations is carried over in the parallel implementation, resulting in performance loss for both synchronization protocols.
+The optimistic case can still achieve good performance if the volume of inter-kernel messages is low.
+Both synchronization protocols are fast in non-cyclic simulations, whereas the conservative case is better suited whenever state-saving becomes expensive; the optimistic case, on the other hand, is better suited when a lookahead is not available.
+Finally, the current optimistic implementation is ill-suited for memory constrained systems, especially in simulations where the frequency of reverts is high.

### 2015-12-07 paper/4-performance.tex
typo

@@
 When using parallel processing, this can be reduced to $\textup{timeadvance}()*(n\frac{kernels-1}{kernels} -1)$ before the last kernel becomes active.\\
-Figure~\cite{fig:DevstonParallel} shows that dxex outperformance adevs (in both optimistic and conservative case).
+Figure~\ref{fig:DevstoneParallel} shows that dxex outperformance adevs (in both optimistic and conservative case).
 \begin{figure}[ltbh]

### 2015-12-07 paper/3-features.tex
Typo.

@@
 \subsubsection{Heap}
-In dxex, events are always passed with pointers and thus involve a possibly expensive copy of the payload caused by allocation and copying overhead.
+In dxex, events are always passed with pointers and thus avoid a possibly expensive copy of the payload caused by allocation and copying overhead.
 In highly connected models, this allocation cost can become prohibitively expensive, so to reduce that overhead we use thread\_local memory pools for states and events, and, optionally, replace the system malloc with calls to tcmalloc~\cite{tcmalloc}.

### 2015-12-08 paper/4-performance.tex
Updated citation to HighInterconnect and fixed some float options to make it compile on a minimal latex install

@@
 \subsubsection{Interconnect}
-Interconnect~\cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in dependencies between models.
+Interconnect~\cite{MasterThesis} is a benchmark where all models broadcast, creating a complete graph in dependencies between models.
 As the model count increases, we see (see Figure~\ref{fig:Interconnect} the expected quadratic increase in runtime for both adevs and dxex, but an increasing penalty for dxex.
@@
 Figure~\ref{fig:DevstoneParallel} shows that dxex outperformance adevs (in both optimistic and conservative case).
-\begin{figure}[ltbh]
+\begin{figure}
 	\includegraphics[width=.5\textwidth]{fig/fig2.eps}
@@
 The overhead in the optimistic case is entirely due to the factor $m$, which will quickly dominate in increasing buffers of received messages.
-\begin{figure}[ltbh]
+\begin{figure}
 	\includegraphics[width=.5\textwidth]{fig/fig5.eps}
@@
 In section 4.2 the parallel performance of this benchmark is further explained. Interconnect highlights dxex's lower performance due to message allocation overhead. 
-\begin{table}[lhtb]
+\begin{table}
 	\centering
@@
 The kernel holding the server will after simulation wait on the receiving kernel until it can prove all sent messages can be deallocated.
-\begin{table}[lhtb]
+\begin{table}
 	\centering

### 2015-12-08 paper/1-introduction.tex
First rewrite of introduction

--- 2015-12-05 paper/1-introduction.tex
@@
-%? use paragraphs for layout instead of (dense) flowing text ?
-%TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
-\subsection{DEVS}
-%TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
-The family of DEVS \cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms.
-Of interest in this paper are the 2 key formalisms: Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation.
-This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a simulation kernel's perspective only Atomic Models exist linked to each other by ports.
+%TODO restructure and let it "flow" better
+\textsf{DEVS}~\cite{CDEVS} is a popular formalism for modelling complex dynamic systems using a discrete-event abstraction.
+In fact, it can serve as a simulation ``assembly language'' to which models in other formalisms can be mapped~\cite{DEVSbase}.
+A number of tools have been constructed by academia and industry that allow the modelling and simulation of \textsf{DEVS} models.
 
-\subsection{Parallel computing}
-%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
-Parallel execution of a PDEVS simulation can lower overall runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
-While the shared memory parallelism offered by most modern hardware does not raise the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation.
-\\
+With the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform the simulation within reasonable time bounds.
+And while \textsf{Parallel DEVS}~\cite{PDEVS} was introduced to increase parallelism, this is often still insufficient.
+Several synchronization algorithms from the discrete event simulation community~\cite{FujimotoBook} have been applied to \textsf{DEVS} simulation in particular.
+While several parallel \textsf{DEVS} simulation kernels exist, they often restrict themselves to a single synchronization protocol.
+This is a logical choice, as these synchronization protocols have nearly no commonalities.
+But the exact reason for different synchronization protocols, is that their distinct nature makes them very applicable to specific situations.
+As such, current \textsf{DEVS} simulation kernels always allow parallel simulation, but good performance can only be expected from some models.
 
-\subsection{Motivation}
-%TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
-Adevs \cite{adevs} offers a very fast conservative synchronized shared memory PDEVS simulator, but no optimistic synchronized variant. 
-The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
-The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code and without losing access to powerful low-level threading primitives.
-% may need another pass on this one :(
+Users that wish to simulate a set of distinct models, with different synchronization protocols, are therefore out of luck:
+Either they accept the lower performance for some of the models, or they use two distinct simulation kernels.
+Neither of these options is acceptable for the simulation of complex models, as the performance can become arbitrary worse, and the simulation kernels can diverge significantly.
 
-\subsection{Solution} 
-%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
-The usage of the Direct connection algorithm makes reusing adevs kernel hard. Devs-ex-machina (dxex) is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
-The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance.
-As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
-The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
-\subsection{Time}
-The DEVS formalisms have $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time.
-In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.
-In practice, dxex defaults to word sized unsigned integer.
-Although this significantly reduces the possible range of virtual time, it avoids approximation errors. The notion of $\epsilon$ is trivial for integer, unlike floating point.\\
-Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernel ($2^{48}$ by default).
-If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$.
-This allows a deterministic ordering of concurrent events.
-% execution not so much as tracing
+In this paper, we introduce DEVS-Ex-Machina (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
+The selected synchronization protocol is transparant to the simulated model.
+Users should merely determine, at the start of simulation, which protocol they wish to use.
+Our tool is implemented in C++11, to increase both performance and portability across different platforms.
+
+A simple model is used to prove that the ideal synchronization algorithm is indeed dependent on model details.
+In our case, a single parameter of the model can have a significant impact on simulation performance, and especially on which is the ideal synchronization protocol.
+In order to show that this flexibility does not counter our performance, we compare our tool to Adevs, currently one of the fastest \textsf{DEVS} simulation tools available~\cite{PythonPDEVS1}.
+
+We introduce the necessary background for this paper in Section~\ref{sec:2-background}.
+Section~\ref{sec:3-features} elaborates on our features and design.
+In Section~\ref{sec:4-performance}, we evaluate performance of our tool by comparing its different synchronization protocols.
+Related work is discussed in Section~\ref{sec:5-related-work}.
+Section~\ref{sec:6-conclusion} concludes the paper and gives future work.

### 2015-12-08 paper/1-introduction.tex
Fixed broken citations

@@
 %TODO restructure and let it "flow" better
-\textsf{DEVS}~\cite{CDEVS} is a popular formalism for modelling complex dynamic systems using a discrete-event abstraction.
+\textsf{DEVS}~\cite{ClassicDEVS} is a popular formalism for modelling complex dynamic systems using a discrete-event abstraction.
 In fact, it can serve as a simulation ``assembly language'' to which models in other formalisms can be mapped~\cite{DEVSbase}.
@@
 With the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform the simulation within reasonable time bounds.
-And while \textsf{Parallel DEVS}~\cite{PDEVS} was introduced to increase parallelism, this is often still insufficient.
+And while \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, this is often still insufficient.
 Several synchronization algorithms from the discrete event simulation community~\cite{FujimotoBook} have been applied to \textsf{DEVS} simulation in particular.
@@
 In our case, a single parameter of the model can have a significant impact on simulation performance, and especially on which is the ideal synchronization protocol.
-In order to show that this flexibility does not counter our performance, we compare our tool to Adevs, currently one of the fastest \textsf{DEVS} simulation tools available~\cite{PythonPDEVS1}.
+In order to show that this flexibility does not counter our performance, we compare our tool to Adevs, currently one of the fastest \textsf{DEVS} simulation tools available~\cite{PythonPDEVS,DEVSSurvey}.
 

### 2015-12-08 paper/2-background.tex
Rewrote optimistic synchronization subsection

@@
-In this section, we provide a brief introduction to two different synchronization protocols for parallel simulation, and the features offered by C++11 that aid in our implementation.
+This section briefly introduces two distinct synchronization protocols, as used by dxex.
+Furthermore, we make note of several new features of C++11.
 
 \subsection{Conservative Synchronization}
-%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
-Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
-This requires the concepts of earliest output time (eot) and earliest input time (eit), which define the timespan within which a model can safely advance.\\
-The eit of any model is the minimum of all eot values of (directly) influencing models. A model can simulate up to (but not including) eit and then waits until that value is increased. \\
-An important disadvantage here is that the relation that calculates the dependencies, is always defined at model (link) creation time, and not at runtime.
-A model that can influence another model theoretically, but never does in real life, can severely slow down the protocol.
-\\
-Deadlocks between models that influence each other and end up waiting on each other, can be broken/avoided by a variety of means.
-In dxex, the CMB~\cite{Chandy:1981:ADS:358598.358613} null-message protocol is used.
-\\
-In our implementation, null-time is the time stamp a model is guaranteed to have passed in simulation.
-More precisely, a null message of time $t$ guarantees that any output with time stamp $t-\epsilon$ has already been sent.\\
-Conservative synchronization relies explicitly on information provided by the model creator in the form of a lookahead, a relative timespan during which the model is insensitive to outside events.
-This lookahead can be non-trivial to calculate; therefore, a simulation writer will in general not be able to determine the exact lookahead of the models involved in an experiment without having run the experiment.\\
-Conservative kernels can even operate if there are cyclic dependencies involved, but this can come at a severe performance penalty, as seen in section 4.
+The first synchronization protocol that we will introduce, is \textit{conservative synchronization}~\cite{FujimotoBook}.
+In conservative synchronization, a node is allowed to progress in simulated time, independent of all other nodes, up to the point where it can guarantee that no causality errors can happen.
+When this point in time is reached, the node has to block until it is allowed to progress any further.
+In practice, this means that all nodes need to be aware of the current simulation time of all other nodes, and the time it takes an event to propagate (called \textit{lookahead}).
+Several algorithms are defined in the literature to implement this behaviour.
+An overview is given in~\cite{FujimotoBook}.
+
+Deadlock can occur when a dependency cycle occurs and the amount of exchanged messages is low.
+Again, multiple algorithms are defined to handle this situation.
+Algorithms range from deadlock avoidance to deadlock recovery.
+
+The main advantage of this approach is that it has a very low overhead in cases of high parallelism between different nodes.
+Each node can simulate in parallel, while sporadically notifying other nodes that they can progress even further.
+The disadvantage, however, is that the amount of parallelism is explicitly limited by the size of the lookahead.
+If a node can influence another (almost) instantaneously, no matter how rarely it occurs, the amount of parallelism is severely reduced.
+It is also up to the user to define the size of the lookahead, depending on how the model is known to behave.
+Slight changes in model behaviour can cause significant changes to the lookahead, and can therefore also have a significant influence on simulation performance.
 
 \subsection{Optimistic Synchronization}
-%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
-In the case of optimistic synchronization, causality errors can occur but these will be reverted via a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.\\
-Whenever a kernel receives an event with a time stamp in the kernel's past, the state of the kernel (and all models) is reverted to that time.\\
-Because an optimistic kernel does not wait for other kernels, it can run much faster than a conservative kernel.
-The cost of this approach, however, is higher complexity of the code and a larger memory usage for the state and message saving.
-In contrast to conservative synchronization, optimistic synchronization does not rely on any domain specific information.
-Therefore, it is only sensitive to the runtime use of connections, not the probability that they might occur.
-%reword connections ?\\
-If the (runtime) dependency graph contains a cycle, optimistic synchronization might lead to a series of cascading reverts.
-The cost of reverts can be reduced by lazy cancellation and/or lazy re-evaluation~\cite{FujimotoBook}.
+A completely different synchronization protocol is \textit{optimistic synchronization}~\cite{TimeWarp}.
+Whereas conservative synchronization would prevent causality errors at all costs, optimistic synchronization will allow them to happen, but correct them afterwards.
+Each node is allowed to progress in simulated time as much as possible, without taking note of the state of any other node.
+When an event arrives at a node, which is already further in simulated time, the node will have to roll back its state to right before the event would normally have to be processed.
+As the simulation time is now rolled back to before the event is processed, the event can simply be processed as if no causality error ever occured.
 
-\subsection{Global Virtual Time}
-To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988} or GVT.
-In optimistic simulations, GVT is defined as the lowest time stamp of any unprocessed event.\\
-Intuitively, this is the simulation time point that is certain to be preserved, corresponding exactly with a sequential simulation.
-In conservative, the minimum simulation time of all kernels is the lower bound of time stamp or LBTS. In terms of null messages this corresponds to the least time stamp of any null message in transit.
-The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, etc.
+Rolling back the simulation time requires the node to store past model states, such that they can be restored later on.
+Furthermore, all incoming and outgoing events need to be stored as well.
+Incoming events need to be passed to the models again, when the correct simulation time has again been reached, and outgoing events need to be cancelled, as potentially a different series of output events would normally have been generated.
+Cancelling events, however, can cause further rollbacks, as the receiving node might also have to roll back its state.
+In practice, a single causality error could have significant repercussions on the complete simulation.
+
+Further changes are required when the model performs unrecoverable transactions, such as I/O (\textit{e.g.}, tracing, writing to file, printing output), allocating and deallocating memory, and so on.
+For this, lightweight synchronization algorithms are still required to determine the lower bound of all simulation times, through the computation of a \textit{Global Virtual Time} (GVT).
+
+%TODO reword advantages
+The advantage of optimistic synchronization is that there is no blocking at all: a node can always progress as fast as possible in simulated time.
+Additionally, you do not suffer from a small lookahead if the small lookahead situation rarely occurs.
+If an (almost) instantaneous event occurs rarely, performance will only be impacted if it occurs.
+The main disadvantage is the unpredictable performance and arbitrary cost of rollbacks.
+If rollbacks occur rather frequently, simulation quickly becomes slower than sequential, as the overhead of state saving becomes significant.
+Apart from the overhead in CPU time, there is also an overhead in terms of memory, as all intermediate states will also have to be stored up to a point where it can be considered \textit{irreversible}.
+
+While optimistic synchronization does not explicitly depends on the lookahead, simulation performance is still bound by the lookahead implicitly.
 

### 2015-12-08 paper/2-background.tex
Rewrote C++11 subsection

@@
 \subsection{C++11 Parallelism Features}
-%TODO explain what is new in C++11 for parallelism and how it contributes to our cause
-C++11 offers a wide range of portable synchronization primitives in the Standard Library, whereas in earlier versions one had to resort to non-portable (C) implementations.
-More importantly, C++11 is the first version of the standard that actually defines a multi-threaded abstract machine memory model in the language.
-Our kernels use a wide range of threading primitives and atomic operations.
-As an example, eot/eit/nulltime are not exchanged as messages, but as reads/writes to atomic fields shared by all kernels.
-This avoids the otherwise unavoidable latency penalty by mixing simulation messages with synchronization messages.
+Apart from various other additions, C++11 adds a wide range of portable synchronization primitives to the Standard Library.
+In earlier versions of the language standard, one had to resort to non-portable C implementations.
+Furthermore, C++11 is the first version of the standard that explicitly defines a multi-threaded abstract machine memory model.
+Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform.
+
+Simulation values that need to be shared between nodes, can now be coded portable and efficiently through the use of atomic fields.
+This avoids latencies caused by the exchange of synchronization messages.
 For an in-depth study, we refer to~\cite{CPE:CPE3007}.
-Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform.

### 2015-12-08 paper/2-background.tex
Reworded advantages of optimistic synchronization

@@
 
-%TODO reword advantages
-The advantage of optimistic synchronization is that there is no blocking at all: a node can always progress as fast as possible in simulated time.
-Additionally, you do not suffer from a small lookahead if the small lookahead situation rarely occurs.
-If an (almost) instantaneous event occurs rarely, performance will only be impacted if it occurs.
-The main disadvantage is the unpredictable performance and arbitrary cost of rollbacks.
-If rollbacks occur rather frequently, simulation quickly becomes slower than sequential, as the overhead of state saving becomes significant.
-Apart from the overhead in CPU time, there is also an overhead in terms of memory, as all intermediate states will also have to be stored up to a point where it can be considered \textit{irreversible}.
+Optimistic synchronization never blocks: a node always progresses as fast as possible, at the cost of potentially discarding simulation results.
+More importantly, performance is not limited by a small lookahead, caused by a very infrequent event.
+If an (almost) instantaneous event occurs rarely, performance will only be impacted if it occurs, and not at every simulation step.
+The main disadvantage is unpredictable performance and arbitrary cost of rollbacks due to the propagation of causality errors.
+If rollbacks occur frequently, simulation quickly becomes slow, as the overhead of the recovery mechanisms becomes significant.
+Apart from overhead in CPU time, a significant memory overhead is present: all intermediate states qre stored up to a point where it can be considered \textit{irreversible}.
 

### 2015-12-08 paper/3-features.tex
Rewrite introduction of section 3

@@
-%TODO explain that we are briefly going over the main features
-\subsection{Based on PyPDEVS}
-%TODO say that it is based on PythonPDEVS --done
-Dxex is based on PyPDEVS, and provides the following features: 
-\begin{enumerate}
-	\item Direct Connection
-	\item Dynamic Structured DEVS
-	\item Termination function: if specified, a termination function is applied to each model every simulation round to test whether the simulation can terminate.
-This feature is only available in single-threaded simulations.
-	\item The State/Message objects can have any payload type. Different allows different message types to be used within the same simulation.
-	\item Tracing: An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
-	\item Optimistic and Conservative synchronization
-\end{enumerate}
-Furthermore, the implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use.
-For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had an adverse impact on the runtime, even when not in use.
-Another example is the state saving mechanism, which is only used for optimistic parallel simulation and has no performance impact in a conservative parallel simulation.\\
-Our tracing implementation is not comparable to adevs's listener interface.
-To be usable in an optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points.
-Furthermore, the framework itself has to be thread safe and deterministic so that a simulation will always produce the exact same output.\\
-The following features from PyPDEVS are not present
-\begin{enumerate}
-	\item Activity tracking and relocation
-	\item Serialization: in this context this is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp. 
-	\item Interactive control
-	\item Distributed simulation
-\end{enumerate}
-In dxex, the model allocation is realized by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation.
-If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
-In addition, debugging tools such as a logger and a graph visualizer are included that can track activity with respect to the allocation for later inspection (but not are not online and as dynamic as is in PyPDEVS).\\
-%TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
-%TODO (size: take as much size as you want, depending on how much is shared with PythonPDEVS)
-%TODO (maybe: note that models are not compatible)
+Historically, dxex is based on PythonPDEVS~\cite{PythonPDEVS}.
+While Python is a good language to create software prototypes, its performance has proven to be insufficient to compete with other simulation kernels~\cite{MasterThesis}.
+Dxex implements only a subset of PythonPDEVS, but makes some of its own additions.
+The core simulation algorithm and optimizations, however, are highly similar.
+
+While we will not make a detailed comparison with PythonPDEVS here, dxex also supports direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing framework~\cite{PythonPDEVS}.
+But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols.
+Furthermore, the implementation in C++11 allows many more (static) optimizations, which were plainly impossible when using an interpreted language.
+
+Note that, since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
+This is due to dxex models being grafted on C++11, whereas PythonPDEVS models are grafted on Python.
+
+In the remainder of this section, we will elaborate on our prominent new feature: support for multiple synchronization protocols within the same simulation tool.
+
+\subsection{Rationale}
+
 \subsection{Different Synchronization protocols}

### 2015-12-08 doBenchmark.py
* fixed test to incorporate new priority chance * adjusted benchmark script as well

@@ -114,10 +114,11 @@ def pholdgen(simtype, executable):
         for apn in [2, 4, 8, 16]:  # , 8, 16, 32]:
             for iterations in [0]:
                 for remotes in [10]:
-                    for endTime in [1000000]:
-                        yield list(chain([executable], simtype, ['-n', nodes, '-s', apn, '-i', iterations, '-r', remotes, '-t', endTime]))
-                        # return
-lenParallelPhold = len(list(chain(["pholdExec"], simtypes.optimistic, ['-n', 0, '-s', 0, '-i', 0, '-r', 0, '-t', 0])))
+                    for priority in [0.10]:
+                        for endTime in [1000000]:
+                            yield list(chain([executable], simtype, ['-n', nodes, '-s', apn, '-i', iterations, '-r', remotes, '-p', priority, '-t', endTime]))
+                            # return
+lenParallelPhold = len(list(chain(["pholdExec"], simtypes.optimistic, ['-n', 0, '-s', 0, '-i', 0, '-r', 0, '-p', 0, '-t', 0])))
 
 
 def interconnectgen(simtype, executable, doRandom=False):
@@ -172,7 +173,7 @@ def prioritygen(simtype, executable):
 
 csvDelim = ';'
 devsArg = [csvDelim, """ "command"{0}"executable"{0}"simtype"{0}"ncores"{0}"width"{0}"depth"{0}"end time" """.format(csvDelim), lambda x: ("\"{}\"".format(" ".join(map(str, x))), "\"{0}\"".format(x[0].split('/')[-1]), "\"{}\"".format(x[1]), x[3] if len(x) == lenParallelDevstone else 1, x[-5], x[-3], x[-1])]
-pholdArg = [csvDelim, """ "command"{0}"executable"{0}"simtype"{0}"ncores"{0}"nodes"{0}"atomics/node"{0}"iterations"{0}"% remotes"{0}"end time" """.format(csvDelim), lambda x: ("\"{}\"".format(" ".join(map(str, x))), "\"{0}\"".format(x[0].split('/')[-1]), "\"{}\"".format(x[1]), x[3] if len(x) == lenParallelPhold else 1, x[-9], x[-7], x[-5], x[-3], x[-1])]
[…]

### 2015-12-08 paper/3-features.tex
Added rationale and promoted conservative/optimistic to subsection

@@
 \subsection{Rationale}
+We have previously shown that different synchronization protocols exist, with each of them being optimized for a specific kind of model.
+As no single synchronization protocol is ideal, a general purpose simulation tool should support multiple situations.
+Currently, however, most parallel simulation tools chose only a single synchronization protocol due to the inherent differences between these approaches.
+An uninformed choice on which to implement is insufficient, as performance is likely to be bad.
+We therefore argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization.
 
-\subsection{Different Synchronization protocols}
-%TODO explain rationale: why different synchronization protocols
-For parallel executions, synchronization is required for PDEVS to prevent causality violations from happening or to recover from causality violations.
-Preventing causality violations (conservative) typically requires domain-specific information to compensate for the performance loss, while recovering from causality errors (optimistic) requires the calculation of a state to revert to.
+Each of them is applicable in specific model configurations.
+Conservative synchronization is ideal when high lookahead exists between different nodes, and barely any blocking is necessary.
+Optimistic synchronization is ideal when lookahead is generally high, but rare (almost) instantaneous events cause a significantly lower lookahead.
+Finally, sequential simulation is still required for models where parallelism is bad, causing significant overhead.
 
-\subsubsection{Conservative}
+\subsection{Conservative}
 In case of conservative synchronization, any kernel will determine which kernels it is influenced by.
@@
 
-\subsubsection{Optimistic}
+\subsection{Optimistic}
 The optimistic kernel requires from the hosted model that copying the state is done carefully (avoid unnecessary copies and make a copies whenever necessary).\\

### 2015-12-08 paper/3-features.tex
Change explanation of conservative and optimistic implementation

@@
 
-\subsection{Rationale}
+\subsection{Synchronization protocols}
 We have previously shown that different synchronization protocols exist, with each of them being optimized for a specific kind of model.
@@
 
-\subsection{Conservative}
-In case of conservative synchronization, any kernel will determine which kernels it is influenced by.
-This information is constructed from the incoming connections on all hosted models.
-The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
-A model should provide a lookahead function which returns, relative to the current time, the timespan during which the model cannot change state due to an external event.
-This information is collected for all models hosted on the kernel and the minimum time is set as the lookahead of that kernel.
-\\
-The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
-For the garbage collection (of sent messages), the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i])  - \epsilon $.\\
-%This is one of the operations that can benefit from using a relaxed memory ordering.
-% Using it for read atm, check before finalizing and do we need to mention this ?.
-%writing this here seems inappropriate
+\subsubsection{Sequential}
+%TODO how simulation works, and how other protocols plug into this
 
-\subsection{Optimistic}
-The optimistic kernel requires from the hosted model that copying the state is done carefully (avoid unnecessary copies and make a copies whenever necessary).\\
-The kernels use Mattern's \cite{mattern}
-GVT algorithm with a maximum of 2 rounds per iteration to determine a GVT.
+\subsubsection{Conservative}
+For conservative synchronization, each node determines the nodes it is influenced by.
+Each model provides a lookahead function, which determines the lookahead depending on the current simulation state.
+Within this time interval, it is an error if a model raises an event, thus violating its previous promise.
+
+The node uses this information to compute its earliest output time (EOT), and writes out the value in shared memory through the use of C++11 synchronization primitives.
+%TODO extend maybe
+
+\subsubsection{Optimistic}
+For optimistic synchronization, each node needs to keep track of all intermediate simulation states.
+This needs to be done carefully, in order to avoid unnecessary copies, and minimize the overhead induced for each transition function.
+
+To determine the Global Virtual Time (GVT), we use Mattern's GVT algorithm~\cite{mattern}, which finds the GVT with at most $2n$ exchanged synchronization messages.
 This process runs asynchronously from the simulation itself.
-Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states and (anti)messages.\\
-%TODO explain core infrastructure a bit, maybe with a diagram or so, which clearly shows that everything is shared, except the "synchronization" part
-%TODO claim that this allows you to easily switch between both methods, to always chose the best one
+Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed.
+
+\subsection{Transparancy}
+%TODO
+
 The user must only provide one implementation of a model that can be used for both synchronization protocols.

### 2015-12-08 paper/3-features.tex
Wrote something about transparancy

@@
 Conservative synchronization is ideal when high lookahead exists between different nodes, and barely any blocking is necessary.
-Optimistic synchronization is ideal when lookahead is generally high, but rare (almost) instantaneous events cause a significantly lower lookahead.
+Optimistic synchronization is ideal when lookahead is unpredictable, or there are rare (almost) instantaneous events.
 Finally, sequential simulation is still required for models where parallelism is bad, causing significant overhead.
@@
 \subsubsection{Sequential}
-%TODO how simulation works, and how other protocols plug into this
+%TODO write
 
@@
 The node uses this information to compute its earliest output time (EOT), and writes out the value in shared memory through the use of C++11 synchronization primitives.
-%TODO extend maybe
 
@@
 \subsection{Transparancy}
-%TODO
+A user must only provide one model, implemented in C++11, which can be simulated by each synchronization kernel.
+The exception is conservative synchronization: a lookahead function is required, whereas this is not possible in other synchronization kernels.
+Two options are possible: either a lookahead function is always provided, even when it is not required and possibly not used, or using a default lookahead function if none is defined.
 
-The user must only provide one implementation of a model that can be used for both synchronization protocols.
-A lookahead function is desired to accelerate the conservative protocol, but is not required.
-In the absence of a user supplied lookahead, the kernel assumes it cannot predict beyond its current time $t+\epsilon$, creating a lockstep simulation.
-% implementation detail ?\\
-The user is shielded from the multi-threaded aspect of the kernel.\\
-From the user's perspective, the multi-threaded aspect of the kernel is not exposed. % payload protected ?
-%TODO indicate in which optimistic is good (when runtime behavior is very hard to predict). Describe some drawbacks (many reverts when everybody influences everybody all the time)
+Always defining a lookahead function might seem redundant, especially if users will never use conservative synchronization.
+The more attractive option is for the simulation tool to provide a default lookahead function, defined by the minimum detected time advance.
+This lookahead value is most likely too small, but will prevent causality errors at the cost of performance.
+Depending on the model, simulation performance might still be faster than sequential simulation.
+
+Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity.
 

### 2015-12-08 paper/3-features.tex
Add some explanation about sequential simulation and make the link with PythonPDEVS a bit more

@@
 But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols.
+This is in line with the design principles of PythonPDEVS: offer users the option to pass information on how to efficiently simulate the model.
+In our case, it now becomes possible to pass the simulation kernel the ``hint'' as to which synchronization protocol would be ideal for this model.
 Furthermore, the implementation in C++11 allows many more (static) optimizations, which were plainly impossible when using an interpreted language.
@@
 \subsubsection{Sequential}
-%TODO write
+Our sequential simulation algorithm is very similar to the one found in PythonPDEVS, including most of its optimizations.
+However, the simulation algorithm is designed to be overloaded by different synchronization protocol implementations.
+This way, the \textsf{DEVS} simulation algorithm is implemented once, but parts can be overwritten as needed.
+In theory, more synchronization protocols (\textit{e.g.}, other algorithms for conservative synchronization) can be added without altering our design.
 
@@
 Within this time interval, it is an error if a model raises an event, thus violating its previous promise.
-
 The node uses this information to compute its earliest output time (EOT), and writes out the value in shared memory through the use of C++11 synchronization primitives.
@@
 This needs to be done carefully, in order to avoid unnecessary copies, and minimize the overhead induced for each transition function.
-
-To determine the Global Virtual Time (GVT), we use Mattern's GVT algorithm~\cite{mattern}, which finds the GVT with at most $2n$ exchanged synchronization messages.
+We use Mattern's GVT algorithm~\cite{mattern} to determine the Global Virtual Time (GVT) using at most $2n$ synchronization messages.
 This process runs asynchronously from the simulation itself.
@@
 Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity.
-
-\subsection{Performance Improvements}
-%TODO briefly discuss some of the additional changes you implemented
-%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
-We now discuss a number of bottlenecks that were discovered during the profiling of several benchmarks.
-\subsubsection{Heap}
-In dxex, events are always passed with pointers and thus avoid a possibly expensive copy of the payload caused by allocation and copying overhead.
-In highly connected models, this allocation cost can become prohibitively expensive, so to reduce that overhead we use thread\_local memory pools for states and events, and, optionally, replace the system malloc with calls to tcmalloc~\cite{tcmalloc}.
-In this way, allocating threads do not block each other.
-If desired, arena-pools are available for single-threaded simulation.
-A disadvantage is increased complexity in ownership semantics.
-The creating kernel is responsible for destruction, but this can only be guaranteed up to the GVT.
-Experiments with synchronized pools proved to be slower than implementations with the standard malloc/free.\\
-Initially, dxes was using strings as identifiers as was also done in PyPDEVS.
-Profiling quickly indicated that this caused a real performance bottleneck.
-In C++, strings are heap allocated variable sized objects with an atomic reference count and not not immutable objects as in Python.
-Access of that reference count across threads turned out to be quite expensive, as are the calls to malloc/free the string implementation makes to create/destroy new objects or copy existing objects.\\
-Strings, however, are more intuitive to work with from a user's perspective, so, as a compromise, we allowed the user to reference models/ports by string name.
-Once the simulation starts however, all objects use integral identifiers.
-This also increased the usage of the constexpr feature of C++11 in, amongst others, time stamps and message headers.
-\subsubsection{Raw pointers}
-While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced.
-Used across threads, the reference counting became very expensive, and the (de)allocation of memory caused significant contention between threads.
-So for the models and the kernels, dxex is still using smart pointers whereas for the messeages, a raw pointer to compacted memory is being used.
-\subsubsection{Locking}
-Locking between kernels uses mostly atomic operations and, occasionally, we can leverage memory orderings to only pay for synchronization when we need it.
-Messages, on the other hand, are exchanged via a shared set of queues each with a dedicated lock.\\
-On a higher level, we avoid the sending of synchronization messages entirely by writing the time stamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
-\subsubsection{Schedulers}
-PyPDEVS has a wide range of schedulers to choose from with varying performance depending on the simulation type.
-Profiling showed that, the heap implementation used in adevs was faster than any of the schedulers we had tested before (in a C++ environment).
-Unlike most node based heaps, this scheduler uses a fixed size array where its heap is rebuilt or modified in place depending on the amount of items to update.
-Items are only updated, never removed.

### 2015-12-09 paper/4-performance.tex
Wrote brief introduction to performance section

@@
-%TODO say that you will now prove your claims about the aproach having nearly no impact
-%TODO and potentially even claim that you are faster than adevs for some situations, due to the algorithms from PythonPDEVS and further features
-%TODO briefly give some hardware info on the machine (CPU, OS, RAM), used software (version of adevs, GCC, ...), and about number of samples/variance
+In this section, we evaluate the performance of our simulation tool.
+We show that the inclusion of multiple synchronization protocols does not decrease our performance to the point where it is unusable.
+To this end, we compare to adevs, one of the most efficient simulation kernels at this time~\cite{DEVSSurvey}.
+A comparison is made for both the CPU and memory usage of both sequential simulation and parallel simulation.
+
+We start of with a comparison of sequential simulation, to show how adevs and dxex relate in this simple case.
+We not only show that our approach does not influence performance negatively, but also that our main simulation algorithm, similar to the one of PythonPDEVS, is significantly faster than the one found in adevs.
+Similar differences, compared to adevs, can also be seen in the parallel simulation benchmarks.
+In the parallel simulation benchmarks, the benefit of our different synchronization protocols is also indicated.
+
+For all benchmarks, results were all well within 1\% deviation of the average, such that only the average is used in the remainder of this section.
+The same compilation flags were used for both adevs and dxex benchmarks (``\texttt{-O3}'').
+To guarantee comparable results, no IO was performed during the benchmarking phase.
+Simulation traces were used to verify that both adevs and dxex models have exactly the same behaviour.
+
 \subsection{Sequential Simulation}

### 2015-12-09 paper/4-performance.tex
Rewrote introduction to sequential simulation

@@
 
+\subsection{Benchmarks}
+%TODO
+
 \subsection{Sequential Simulation}
-%TODO we start with sequential simulation results, because this is still base case
-%TODO compare CPU time with adevs for some benchmarks
-%TODO and explain why there is such a difference
-%TODO (spend some time here, as this is also important)
-In our benchmarks, we will consistently compare the performance of dxex with adevs. The implemented models for the each benchmark are functionally identical.\\
-Dxex uses integer timestamps whereas adevs uses floating point time representation. Tests show there is little difference performance-wise (both types fit in a machine word).\\
-The benchmarks have been compiled using the {\tt -O3} option and link-time optimization, with all IO disabled.
-To compare the models in details, we have run the different benchmarks with different parameters.\\
-The benchmarks have been run on a i7-2670QM, 16GiB RAM, Arch Linux (4.2.5 kernel) and G++ 5.2.
-Interpretation of the benchmark results is done using call-graphs generated by perf.
-A selection of those call-graphs is available in the code repository.
-For the runtime plots, perf is used in combination with R.\\
+Despite our core contribution being mainly in the parallel simulation, we still value a comprehensive comparison of sequential simulation results.
+First, and foremost, as parallel simulation results are tightly linked to the sequential simulation results: parallel simulation merely adds a synchronization layer over different, essentially sequential, simulation kernels.
+Second, since parallel simulation results are validated through the use of adevs.
+To provide a more comprehensive comparison to adevs in the parallel simulation benchmarks, sequential simulation results need to be compared.
+
+Benchmarks have been run on an \textit{Intel i7-2670QM} with \textit{16GB of main memory} on Linux.
+Results were obtained through the use of \textit{perf}.
+A selection of call-graphs and more detailed results can be found online.
+
 %Note that a fixed timeadvance (e.g. devstone) is set at 100.

### 2015-12-10 paper/4-performance.tex
Added brief explanation of benchmarks; added eps version of cores class diagram

@@
 \subsection{Benchmarks}
-%TODO
+We use a selection of benchmarks, based on those found in the literature.
+Three different types of benchmark are defined:
+\begin{enumerate}
+    \item \textit{Queue} model, based on the HI model of DEVStone~\cite{DEVStone}, creates a chain of atomic models, which are hierarchically nested in each other.
+          It contains two parameters: width and depth.
+          Width determines the number of atomic models in a single level, and depth determines the number of levels.
+          An example for width 3 and depth 2 is shown in Figure~\ref{fig:queue}.
+    \item \textit{HighInterconnect} model, an exaggeration of the HI model of DEVStone~\cite{DEVStone}, creates a set of atomic models, which are all directly connected to each other.
+          Each model has a single input and output port, but each output port is connected to each input port\footnote{Except, of course, that the output and input port of a single model are not connected to each other.}.
+    \item \textit{PHOLD} model, presented by~\cite{PHOLD}.
+          It creates a set of $n$ atomic models, similar to HighInterconnect, but each model has exactly $n-1$ output ports.
+          Couplings are made such that each atomic model is directly connected to each other atomic model, but it can \textit{narrowcast} instead of \textit{broadcast}ing.
+          These $n$ models are further equally distributed over $k$ coupled models.
+\end{enumerate}
+
+We opted to deviate from the DEVStone benchmark, as DEVStone tended towards unrealistic models because all internal and external transitions occur simultaneously.
+In our benchmark models, there is always the option for simultaneous transition functions (\textit{fixed time advance}), or scattered transition functions (\textit{random time advance}).
+Furthermore, they defined the use of an artificial load function, which easily skews the result, making the actual simulation algorithm barely comparable.
+
+%TODO figures! I have them lying around somewhere I guess... still need to search for them
 

### 2015-12-10 paper/4-performance.tex
Explanation of benchmarks expanded

--- 2015-12-09 paper/4-performance.tex
@@
 \subsection{Benchmarks}
-%TODO
+We use a selection of benchmarks, based on those found in the literature.
+Three different types of benchmark are defined, each for a different purpose:
+\begin{enumerate}
+    \item \textit{Queue} model, based on the HI model of DEVStone~\cite{DEVStone}, creates a chain of atomic models, which are hierarchically nested in each other.
+          A single generator will push events into the queue, which get processed by the processors after a fixed or random delay.
+          It takes two parameters: width and depth, which determine the widht and depth of the hierarchy.
+          This benchmark shows how the complexity of the simulation kernel behaves with an increasing amount of atomic models, and an increasingly deep hierarchy.
+          If the processing delay is fixed for all processors, further insight is provided in the collision handling performance of the simulation kernel.
+          An example for width 2 and depth 3 is shown in Figure~\ref{fig:queue_model}.
+
+    \item \textit{PHOLD} model, presented by~\cite{PHOLD}.
+          It creates a set of $n$ atomic models, and each model has exactly $n-1$ output ports: one for every other atomic model.
+          Couplings are made such that each atomic model is directly connected to each other atomic model, such that every atomic model can directly send an event to every other atomic model.
+          After a random delay, atomic models will send out an event to a randomly selected output port.
+          Output port selection happens in two phases: first it is decided whether the event should be sent to an atomic model inside or outside of this coupled model.
+          Afterwards, a uniform selection is made between the possible ports.
+          It takes one parameter: the percentage of remote events, which influences the fraction of messages routed to other coupled models.
+          This benchmark shows how the simulation kernel behaves in the presence of many local or remote events.
+          An example for four models, split over two nodes, is shown in Figure~\ref{fig:PHOLD_model}.
+
+    \item \textit{HighInterconnect} model, a merge of the HI model of DEVStone~\cite{DEVStone} and PHOLD~\cite{PHOLD}.
+          It creates a structure similar to the one from PHOLD, but instead of $n-1$ output ports, every atomic model has only a single output port.
+          All models are still connected to each other, but through the use of broadcasting: every model will receive a generated event.
+          It takes one parameter: the number of models.
+          This benchmark investigates the complexity of the routing algorithm.
+          An example for four models is shown in Figure~\ref{fig:interconnect_model}.
+\end{enumerate}
+
+We opted to deviate from the DEVStone benchmark, as DEVStone tends towards unrealistic models since all internal and external transitions occur simultaneously.
+In our benchmark models, there is always the option for simultaneous transition functions (\textit{fixed time advance}), or scattered transition functions (\textit{random time advance}).
+Furthermore, they defined the use of an artificial load function, which easily skews the result, making the actual simulation algorithm barely comparable.
+
+\begin{figure}
+    \center
+    %\includegraphics[width=\columnwidth]{fig/queue_model.eps}
+    \caption{Queue model for depth 3 and width 2.}
+    \label{fig:queue_model}
+
+    %\includegraphics[width=\columnwidth]{fig/interconnect_model.eps}
+    \caption{HighInterconnect model for four models.}
+    \label{fig:interconnect_model}
+
+    %\includegraphics[width=\columnwidth]{fig/PHOLD_model.eps}
+    \caption{PHOLD model for four models, split over two nodes. In parallel simulation, each coupled model is simulated at a different node.}
+    \label{fig:PHOLD_model}
+\end{figure}
 

### 2015-12-10 paper/4-performance.tex
Added a link to the repository; slightly more relevant info on the used CPU
Wrote explanation of figures for sequential simulation; thinking about having different plots

@@
 Simulation traces were used to verify that both adevs and dxex models have exactly the same behaviour.
+All benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.
 
@@
     \center
+    %TODO
     %\includegraphics[width=\columnwidth]{fig/queue_model.eps}
@@
 
+    %TODO
     %\includegraphics[width=\columnwidth]{fig/interconnect_model.eps}
@@
 
+    %TODO
     %\includegraphics[width=\columnwidth]{fig/PHOLD_model.eps}
@@
 
-\subsection{Sequential Simulation}
+\subsection{Sequential Simulation Execution Time}
 Despite our core contribution being mainly in the parallel simulation, we still value a comprehensive comparison of sequential simulation results.
@@
 To provide a more comprehensive comparison to adevs in the parallel simulation benchmarks, sequential simulation results need to be compared.
+Only the Queue and HighInterconnect models are relevant for sequential simulation.
 
-Benchmarks have been run on an \textit{Intel i7-2670QM} with \textit{16GB of main memory} on Linux.
-Results were obtained through the use of \textit{perf}.
-A selection of call-graphs and more detailed results can be found online.
+\subsubsection{Queue}
+In the Queue model, we increase both the width and depth simultaneously, causing a quadratic growth in the number of atomic models.
+As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
+Through careful analysis of profiling results, we determined that adevs spends much time in handling simulation messages, whereas this is mostly avoided due to the differently designed simulation algorithm of dxex.
+Both simulation tools have quadratically increasing execution times, though dxex is much faster thanks to its more efficient simulation control algorithms.
 
-%Note that a fixed timeadvance (e.g. devstone) is set at 100.
-\subsubsection{Devstone}
-The Devstone \cite{DEVStone} benchmark is highly hierarchical.
-As can be seen in Figure~\ref{fig:Devstone}, dxex is outperforming adevs considerably.
-This can be explained by the fact that, using Direct connection, the hierharchical structure is being flattened whereas, in the adevs case, considerable overhead is caused by traversing the coupled model's structure to pass events.
 \begin{figure}[h]
-	\includegraphics[width=.5\textwidth]{fig/fig1.eps}
-	\caption{Devstone}
-	\label{fig:Devstone}
+	\includegraphics[width=\columnwidth]{fig/fig1.eps}
+	\caption{Queue benchmark results}
+	\label{fig:Queue_benchmark}
 \end{figure}
 
-\subsubsection{PHold}
-PHold \cite{PHOLD} is a benchmark that focuses in particular on parallel execution; the sequential runtime is therefore only added as a baseline.
-A model selects the destination of a message at runtime using a state-saved RNG, which takes up a non-trivial amount of runtime.
-The results of the benchmark are visualized in Figure~\ref{fig:PHold}.
-%% needs further clarification...
-\begin{figure}[h]
-	\includegraphics[width=.5\textwidth]{fig/fig4.eps}
-	\caption{PHold}
-	\label{fig:PHold}
-\end{figure}
+\subsubsection{HighInterconnect}
+In the HighInterconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings.
+As can be seen in Figure~\ref{fig:HighInterconnect_benchmark}, adevs now outperforms dxex by a fair margin.
+Analysis showed that this slowdown is caused by the high amount of exchanged events.
+Event creation is found to be much slower in dxex than it is in adevs, even despite the use of memory pools in dxex.
 
-\subsubsection{Interconnect}
-Interconnect~\cite{MasterThesis} is a benchmark where all models broadcast, creating a complete graph in dependencies between models.
-As the model count increases, we see (see Figure~\ref{fig:Interconnect} the expected quadratic increase in runtime for both adevs and dxex, but an increasing penalty for dxex.
-Profiling shows this is entirely due to the heap allocation of messages, which even though minimized by using memory pools remains significant.
+%TODO only include sequential results?
 \begin{figure}[h]
@@
 
-\subsection{Parallel Simulation}
+\subsection{Parallel Simulation Execution Time}
 %TODO explain that we will now compare different synchronization protocols

### 2015-12-10 paper/4-performance.tex
Introduction to parallel simulation performance

@@
 
-\begin{figure}[h]
+\begin{figure}
 	\includegraphics[width=\columnwidth]{fig/fig1.eps}
@@
 %TODO only include sequential results?
-\begin{figure}[h]
-	\includegraphics[width=.5\textwidth]{fig/fig3.eps}
+\begin{figure}
+	\includegraphics[width=\columnwidth]{fig/fig3.eps}
 	\label{fig3.eps}
@@
 \subsection{Parallel Simulation Execution Time}
-%TODO explain that we will now compare different synchronization protocols
-%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
-By default, the benchmarks use 4 kernels for parallel simulation.
-\subsubsection{Devstone}
-The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages.
-For the optimistic case, this can cause more reverts since the kernels will start to drift faster as the model-count increases.
-Furthermore, optimistic synchronization is quite sensitive to an increase in kernels, since the delay before a revert propagates, increases.
-This benchmark requires a specific warm-up time; for $n=d\times w$ models, it takes $\textup{timeadvance}()*(n-1)$ transitions to activate the last model in the chain.
-When using parallel processing, this can be reduced to $\textup{timeadvance}()*(n\frac{kernels-1}{kernels} -1)$ before the last kernel becomes active.\\
-Figure~\ref{fig:DevstoneParallel} shows that dxex outperformance adevs (in both optimistic and conservative case).
+We now perform an analysis of parallel simulation execution times of our previously defined benchmarks.
+For dxex, we mention results for both conservative and optimistic synchronization.
+Since adevs supports only conservative synchronization, we don't mention optimistic synchronization results there.
+All experiments were performed using four simulation nodes, and executed on a quad-core machine.
+
+We highlight two main results:
+(1) dxex conservative synchronization is competitive with adevs;
+(2) dxex optimistic synchronization is sometimes more efficient than conservative synchronization.
+This shows that our contribution, offering both conservative and optimistic synchronization, is indeed beneficial for a general-purpose simulation tools.
+
+\subsubsection{Queue}
+In the Queue model, we allocate the chain of models such that each node is responsible for a series of connected models.
+This minimizes the number of inter-node messages.
+As the model is a queue, however, the last models will only activate much later on in the simulation.
+Since these are allocated to seperate nodes, some nodes will remain idle until simulation has progressed sufficiently far.
+
+Similar to the sequential benchmarks, Figure~\ref{fig:queue_benchmark_parallel} shows that dxex again outperforms adevs, using both optimistic and conservative synchronization.
+%TODO wait for raw results
+Behaviour is exactly the same as in sequential simulation, but some speedup is achieved.
+
 \begin{figure}
-	\includegraphics[width=.5\textwidth]{fig/fig2.eps}
-	\label{fig:DevstoneParallel}
-	\caption{DevStone parallel}
+	\includegraphics[width=\columnwidth]{fig/fig2.eps}
+	\label{fig:queue_benchmark_parallel}
+	\caption{Queue parallel simulation results using 4 nodes.}
 \end{figure}
+
 \subsubsection{PHold}

### 2015-12-10 paper/4-performance.tex
Cleanup of figure references

@@
 	\includegraphics[width=\columnwidth]{fig/fig1.eps}
-	\caption{Queue benchmark results}
+	\caption{Queue benchmark results for sequential simulation.}
 	\label{fig:Queue_benchmark}
@@
 In the HighInterconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings.
-As can be seen in Figure~\ref{fig:HighInterconnect_benchmark}, adevs now outperforms dxex by a fair margin.
+As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs now outperforms dxex by a fair margin.
 Analysis showed that this slowdown is caused by the high amount of exchanged events.
@@
 	\label{fig3.eps}
-	\caption{Interconnect}
-	\label{fig:Interconnect}
+	\caption{Interconnect benchmark results for sequential simulation.}
+	\label{fig:Interconnect_benchmark}
 \end{figure}
@@
 Similar to the sequential benchmarks, Figure~\ref{fig:queue_benchmark_parallel} shows that dxex again outperforms adevs, using both optimistic and conservative synchronization.
-%TODO wait for raw results
+%TODO wait for raw results to make conclusions
 Behaviour is exactly the same as in sequential simulation, but some speedup is achieved.
+In this case, conservative synchronization seems to be better than optimistic synchronization, at the cost of providing the lookahead.
 
@@
 	\label{fig:queue_benchmark_parallel}
-	\caption{Queue parallel simulation results using 4 nodes.}
+	\caption{Queue benchmark results for parallel simulation using 4 cores.}
 \end{figure}

### 2015-12-10 paper/4-performance.tex
Explanation of the PHold crossing lines situation

@@
 	\includegraphics[width=\columnwidth]{fig/fig3.eps}
-	\label{fig3.eps}
 	\caption{Interconnect benchmark results for sequential simulation.}
@@
 	\includegraphics[width=\columnwidth]{fig/fig2.eps}
+	\caption{Queue benchmark results for parallel simulation using 4 cores.}
 	\label{fig:queue_benchmark_parallel}
-	\caption{Queue benchmark results for parallel simulation using 4 cores.}
 \end{figure}
@@
 \subsubsection{PHold}
-%Phold
-In Phold, the allocation is specified in the benchmark itself.
-Each kernel manages a single node with a constant set of sub-nodes.
-The parameter $R$ (which is here set to $10$) determines the percentage of remote destination models.\\
-The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing the conservative case.
+In the Phold model, we first investigate the influence of the fraction of remote events.
+%TODO wait for results
+%The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing the conservative case.
 The lookahead is $\epsilon$, so the conservative case spends most of its time crawling in steps of $\epsilon$.
@@
 Adevs's lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code.\\
-The optimistic case suffers little from the above problems; due to the high interconnectivity, however, a cascading revert is still possible.
+%The optimistic case suffers little from the above problems; due to the high interconnectivity, however, a cascading revert is still possible.
 With the percentage of remotes equal to $100$, Phold reflects interconnect in behaviour, which is why the $R$ value is not dimensioned here.
@@
 Despite all this, optimistic can for low $R$ values quickly exploit the uncertainty that slows down conservative in this benchmark.
+
+Secondly, we verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization.
+We slightly modified the Phold benchmark, to include high-priority events.
+Contrary to normal events, which offer a sufficiently large lookahead, high-priority events happen almost instantaneous, thus restricting lookahead to a very small value.
+Even though the normal events will occur most oftenly, a conservative implementation should always block since these high-priority events can always occur.
+An optimistic implementation, however, will simply go forward in simulation time and roll back the few times that these high-priority events happen.
+This situation closely mimics the case made in the comparison between both synchronization algorithms by~\cite{FujimotoBook}.
+
+Figure~\ref{fig:phold_priority} shows how simulation performance is influenced by the fraction of high-priority events.
+If barely any high-priority events occur, conservative synchronization is penalized due to its excessive blocking, which often turned out to be unnecessary.
+Should many high-priority events occur, however, optimistic synchronization is penalized due to its mindless progression of simulation, which frequently needed to be reverted.
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{fig/phold_voorlopig.eps}
+    \caption{Phold benchmark results with high-priority events.}
+    \label{fig:phold_priority}
+\end{figure}
 
@@
 	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
-	\label{dtone_mem}
 \end{table}
@@
 	\centering
-	\label{dtone_mem}
 	\begin{tabular}{| l | l | l | l | l |}
@@
 	\centering
-	\label{iconn_mem}
 	\begin{tabular}{| l | l | l | l | l |}
@@
 	\centering
-	\label{pmod_mem}
 	\begin{tabular}{| l | l | l |}

### 2015-12-10 paper/2-background.tex

@@
 Deadlock can occur when a dependency cycle occurs and the amount of exchanged messages is low.
-Again, multiple algorithms are defined to handle this situation.
-Algorithms range from deadlock avoidance to deadlock recovery.
+Multiple algorithms are defined to handle this situation, such as deadlock avoidance and deadlock recovery.
 
-The main advantage of this approach is that it has a very low overhead in cases of high parallelism between different nodes.
+The main advantage of conservative synchronization is that it has a low overhead if high parallelism exists between nodes.
 Each node can simulate in parallel, while sporadically notifying other nodes that they can progress even further.
@@
 
-Further changes are required when the model performs unrecoverable transactions, such as I/O (\textit{e.g.}, tracing, writing to file, printing output), allocating and deallocating memory, and so on.
-For this, lightweight synchronization algorithms are still required to determine the lower bound of all simulation times, through the computation of a \textit{Global Virtual Time} (GVT).
+Further changes are required for unrecoverable operations, such as I/O (\textit{e.g.}, tracing, writing to file, printing output) and memory management.
+Lightweight algorithms are still required to determine the lower bound of all simulation times, through the computation of a \textit{Global Virtual Time} (GVT).
 
-Optimistic synchronization never blocks: a node always progresses as fast as possible, at the cost of potentially discarding simulation results.
-More importantly, performance is not limited by a small lookahead, caused by a very infrequent event.
+The main advantage is that performance is not limited by a small lookahead, caused by a very infrequent event.
 If an (almost) instantaneous event occurs rarely, performance will only be impacted if it occurs, and not at every simulation step.

### 2015-12-10 paper/4-performance.tex

@@
 Should many high-priority events occur, however, optimistic synchronization is penalized due to its mindless progression of simulation, which frequently needed to be reverted.
+These results show that there is no single perfect synchronization algorithm for this model.
+Depending on model configuration, either synchronization protocol might be better.
+We therefore claim that our contribution is invaluable for high performance simulation: depending on the observed communication behaviour, modellers can use the most appropriate synchronization protocol.
 

### 2015-12-10 paper/4-performance.tex
Rewrote high interconnect, still thinking about how to best represent this figure

@@
 \subsubsection{Interconnect}
-In Interconnect, the set of atomic models form a complete graph (w.r.t. connections); each model broadcasts messages to the entire set.\\
-Allocation cannot avoid cycles and the resulting dependency graph between kernels remains a complete graph. The runtime dependency graph is almost immediately identical to the static graph.\\
-The conservative case still shows the same issues as in PHold, with the key difference , for a fixed time advance, the lookahead is equal to the timespan between transitions.
-The scaling issue is identical as with PHold.\\
-In this benchmark, the optimistic case runs very quickly out of memory.
-With $c$ kernels and $N$ atomic models, a single revert undoing $k$ transitions will lead to $(N-1)\times k$ messages that need to be recreated, plus $(N-1) \frac{c-1}{c} \times k$ anti-messages that need to be sent.
+In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
+Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}
+While conservative simulation delivers the expected speedups, dxex is still significantly slower than adevs.
+Interestingly, adevs does not achieve any noticable speedup from the use of multiple cores in this situation.
+Optimistic synchronization is not shown in this case: due to the high amount of exchanged events, all of which need to be stored in the sending \textit{and} receiving node, it quickly runs out of memory.
+Several fixes to this problem have been proposed in the literature~\cite{FujimotoBook}, but none of these is implemented by dxex.
 
-\subsubsection{Priority network model}
-%TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
-The priority benchmark is composed of a single server generating a stream of $0\leq m \leq n$ messages at fixed time intervals, interleaved with a probability $p$ for a priority message to $n$ receivers (see Figure~\ref{fig:NetworkModel}).\\ This defaults the lookahead for the receivers to $\epsilon$, but this time there is no scaling effect, nor are there cycles in the dependency graph.
+\begin{figure}
+    %TODO
+    %\includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
+    \caption{Interconnect benchmark results for parallel simulation.}
+    \label{fig:interconnect_benchmark_parallel}
+\end{figure}
+
+%\subsubsection{Priority network model}
+%The priority benchmark is composed of a single server generating a stream of $0\leq m \leq n$ messages at fixed time intervals, interleaved with a probability $p$ for a priority message to $n$ receivers (see Figure~\ref{fig:NetworkModel}).\\ This defaults the lookahead for the receivers to $\epsilon$, but this time there is no scaling effect, nor are there cycles in the dependency graph.
 This model therefore highlights the basic strengths/weaknesses of both synchronization protocols.
 Receiving models are allocated on another kernel than the server and have an internal transition so that they will not wait for the incoming messages.
-\begin{center}
-\begin{figure}
-\input pqueue.tex
-\caption{The priority network model}
-\label{fig:NetworkModel}
-\end{figure}
-\end{center}
-A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create.
+%\begin{center}
+%\begin{figure}
+%\input pqueue.tex
+%\caption{The priority network model}
+%\label{fig:NetworkModel}
+%\end{figure}
+%\end{center}
+%A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create.
 The kernel holding the server will never revert since it is a source in the dependency graph.
@@
 The overhead in the optimistic case is entirely due to the factor $m$, which will quickly dominate in increasing buffers of received messages.
-\begin{figure}
-	\includegraphics[width=.5\textwidth]{fig/fig5.eps}
-	\label{fig:Priority}
-	\caption{Priority}
-\end{figure}
-Interestingly, the parameter $p$ does not clearly favour either synchronization protocol (see Figure~\ref{fig:Priority}).
+%\begin{figure}
+	%\includegraphics[width=.5\textwidth]{fig/fig5.eps}
+	%\label{fig:Priority}
+	%\caption{Priority}
+%\end{figure}
+%Interestingly, the parameter $p$ does not clearly favour either synchronization protocol (see Figure~\ref{fig:Priority}).
 While this removes any possibility for a lookahead, the conservative case can quickly bridge the timespan between fixed messages since there is no cyclic dependency.
@@
 
-%TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)
 \subsection{Memory Usage}
-%TODO compare memory usage with adevs for some benchmarks
+TODO compare memory usage with adevs for some benchmarks
 \subsubsection{Platform and tools}

### 2015-12-10 paper/4-performance.tex
Removed priority network benchmark
Removed other Priority Network paragraph

@@
 
-%\subsubsection{Priority network model}
-%The priority benchmark is composed of a single server generating a stream of $0\leq m \leq n$ messages at fixed time intervals, interleaved with a probability $p$ for a priority message to $n$ receivers (see Figure~\ref{fig:NetworkModel}).\\ This defaults the lookahead for the receivers to $\epsilon$, but this time there is no scaling effect, nor are there cycles in the dependency graph.
-This model therefore highlights the basic strengths/weaknesses of both synchronization protocols.
-Receiving models are allocated on another kernel than the server and have an internal transition so that they will not wait for the incoming messages.
-%\begin{center}
-%\begin{figure}
-%\input pqueue.tex
-%\caption{The priority network model}
-%\label{fig:NetworkModel}
-%\end{figure}
-%\end{center}
-%A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create.
-The kernel holding the server will never revert since it is a source in the dependency graph.
-The optimistic case will therefore not suffer the same performance hit in recreating the states as it does in PHold.
-The overhead in the optimistic case is entirely due to the factor $m$, which will quickly dominate in increasing buffers of received messages.
-%\begin{figure}
-	%\includegraphics[width=.5\textwidth]{fig/fig5.eps}
-	%\label{fig:Priority}
-	%\caption{Priority}
-%\end{figure}
-%Interestingly, the parameter $p$ does not clearly favour either synchronization protocol (see Figure~\ref{fig:Priority}).
-While this removes any possibility for a lookahead, the conservative case can quickly bridge the timespan between fixed messages since there is no cyclic dependency.
-Lightweight states prevent performance loss due to reverts in the optimistic case; only the overhead in event handling in the optimistic case eventually becomes the deciding factor.
-
 \subsection{Memory Usage}
@@
 \end{table}
-
-\paragraph*{Priority Network}
-The priority network model is detailed in section 4.1.\\
-The high memory usage of the conservative case is largely due to the garbage collection implementation.
-Kernel 0 is the initiator of the LBTS calculation but since it holds only models independent of any other (server), it will finish simulation very fast leaving the other kernel bereft of an updated LBTS.
-The kernel holding the server will after simulation wait on the receiving kernel until it can prove all sent messages can be deallocated.
-\begin{table}
-	\centering
-	\begin{tabular}{| l | l | l |}
-		\hline
-		dxex &dxex con &dxex opt\\ \hline
-		35 & 450 & 651\\ \hline
-	\end{tabular}
-	\caption{Priority model n 128, m 16, p 10,  t2e8, unit MiB, 2 kernels}
-\end{table}

### 2015-12-10 paper/4-performance.tex
Introduction to memory usage section

@@
 \subsection{Memory Usage}
-TODO compare memory usage with adevs for some benchmarks
-\subsubsection{Platform and tools}
+Apart from simulation execution time, memory usage during simulation is also of great importance.
+While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation infeasible.
+We therefore also investigate memory usage of different synchronization protocols.
+
+We do not tackle the problem of states that become too large for a single machine to hold.
+This problem can be mitigated by distribution over multiple machines, which neither dxex or adevs support.
+Comparison of memory usage should thus only be seen in the context of which models are feasible to be simulated by the tool.
+
+\subsubsection{Remarks}
 Both dxex and adevs use tcmalloc as memory allocator.
-Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (malloc/free/sbrk/mmap/...).
-Tcmalloc will only gradually release memory back to the OS, whereas our pools will not do so at all.
-If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool.
-This is one reason why memory utilization is best measured by peak allocation.
+Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
+Tcmalloc only gradually releases memory back to the OS, whereas our pools will not do so at all.
+If memory has been allocated once, it is, at least from a performance point of view, better to keep that memory in the pool indefinitely.
+Due to our motivation for memory usage analysis, we will only measure peak allocation.
 Profiling is done using Valgrind's massif tool~\cite{Nethercote:2007:VFH:1273442.1250746}.
-The platform used for memory profiling has an i5-3317U Intel CPU and 8GiB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).
-\subsubsection{Measure}
+
+\subsubsection{Results}
+\begin{figure}
+    \includegraphics[width=\columnwidth]{fig/memory_voorlopig.eps}
+    \caption{Memory usage results.}
+    \label{fig:memory}
+\end{figure}
+
 Adevs passes messages by value, dxex passes a pointer.

### 2015-12-10 paper/4-performance.tex
Made memory usage section much smaller

@@
 \subsubsection{Results}
+Figure~\ref{fig:memory} shows the memory used by each different benchmark.
+Results are in megabytes, and show the total memory footprint of the running application (\textit{i.e.}, text, stack, and heap).
+
+Unsurprisingly, optimistic synchronization results show very high memory use due to the saved states.
+Note the logarithmic scale that was used for this reason.
+Also, results for optimistic synchronization vary heavily depending on thread scheduling by the operating system, as this influences the drift between different nodes.
+Comparing similar approaches though, we notice that dxed and adevs have very similar memory use.
+
+Conservative simulation always uses more memory than sequential simulation, as is to be expected.
+Additional memory is required for the multiple threads, but also to store all events that are processed simultaneously.
+
+For the Phold benchmark, adevs using conservative synchronization took too long using our profiling tool, and was therefore aborted.
+Therefore, no results are shown for adevs.
+
 \begin{figure}
@@
 \end{figure}
-
-Adevs passes messages by value, dxex passes a pointer.
-The runtime effects of this choice have already been demonstrated in the Interconnect benchmark, so, in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled.
-For the OS and/or user, this is the actual memory footprint of the application.
-It is important to note that, especially in the optimistic case, not all this memory is always in use by the kernels.
-During simulation, the pools will generally not return deallocated memory to the OS, but keep it for later reuse.
-\subsubsection{Results}
-\paragraph*{Devstone}
-\begin{table}[htb]
-	\centering
-	\begin{tabular}{| l | l | l | l | l |}
-		\hline
-		adevs & adevs con &dxex &dxex con&dxex opt\\ \hline
-		44 & 70 & 42 & 75 & 363  \\ \hline
-	\end{tabular}
-	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
-\end{table}
-Since, in the conservative case, messages are passed by pa ointer, a GVT/LBTS implementation is required to organise the garbage collection.
-This inevitable delay explains the higher memory usage compared to adevs.\\
-%TODO next sentence needs rewriting --done?
-Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resulting higher latency) than the LBTS calculation in the conservative case. 
-Moreover, the differences in LP virtual times are far larger compared to conservative time synchronization.
-All these factors explain the heavier memory usage.
-Devstone (flattened) is allocated in a chain.
-Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to an increased memory pressure.
-Unlike conservative and sequential execution, memory usage in the optimistic case varies greatly depending on scheduling of kernel threads and drifting between kernels.
-\paragraph*{PHold}
-\begin{table}[htb]
-	\centering
-	\begin{tabular}{| l | l | l | l | l |}
-		\hline
-		adevs & adevs con &dxex &dxex con &dxex opt\\ \hline
-		40 & x & 37 & 61 & 682  \\ \hline
-	\end{tabular}
-	\caption{Phold n 4 s 16 t1e6 r 10, unit MiB, 4 kernels (if parallel)}
-\end{table}
-With only 10\% of all messages being inter-kernel, we expect conservative to have memory consumption near that of the single threaded implementation, since intra-kernel messages are reclaimable after each round.
-The counterintuitive high memory usage can be explained by conservative's stalled round behaviour which occurs whenever a kernel cannot advance (eit $==$ time).
-In such a round messages are sent out but the kernel does not execute any transitions until it has received all input from all influencing kernels.
-\\
-With lookahead $\epsilon$ this then leads to a high frequency of polling on shared null-times, which are used to determine lbts and thus garbage collection.
-The lbts calculation will not wait until a new value is found, since this can create unwanted contention with the simulation.
-The resulting longer time intervals within which no new lbts is found delay memory deallocation.
-Adevs' conservative fails to complete the benchmark under valgrind even with a significantly reduced load.
-Optimistic exhibits the expected high memory usage.
-\paragraph*{Interconnect}
-In section 4.2 the parallel performance of this benchmark is further explained. Interconnect highlights dxex's lower performance due to message allocation overhead. 
-\begin{table}
-	\centering
-	\begin{tabular}{| l | l | l | l | l |}
-		\hline
-		adevs & adevs con &dxex &dxex con & dxex opt\\ \hline
-		39 & 39 & 35 & 43 & 259 \\ \hline
-	\end{tabular}
-	\caption{Interconnect w 20 t5e5, unit MiB, 2 kernels (if parallel)}
-\end{table}

### 2015-12-10 paper/5-related-work.tex
Related work

--- 2015-12-04 paper/5-related-work.tex
@@
-%TODO explain about other DEVS simulation kernels, and how they differ
-%TODO explain about PythonPDEVS, the most related in terms of feature and "legacy" --> now in C++11 and way better performance
-\subsection{PythonPDEVS}
-Dxex is closely related to PyPDEVS in design and philosophy.
-PyPDEVS allows anyone who grasps the PDEVS formalisms to immediately simulate his/her model without having to consider the kernel implementation.A Python based implementation offers the advantage of very fast prototype/run/evaluate cycles, this can't be matched by a C++ simulator but it should be noted that once the kernels are compiled (shared libraries), the actual compilation time of the model is small enough to make prototyping possible.
-Advanced features such as activity based relocation and the performance gains this results in, are still unique to PyPDEVS.
-\subsection{Adevs}
-%TODO explain about adevs, the most related in terms of performance --> we implement more synchronization options and offer better performance because ...
-Adevs is still under active development, allowing for an exact comparison in performance and features.
-It remains a very fast simulation engine for the PDEVS formalism, but it lacks an optimistic synchronization implementation.
-%Tracing ?
-By virtue of not flattening Coupled Models, adevs's performance degrades in increasingly hierarchical models.
-Dxex manages lookahead with more overhead than adevs, leading to a performance difference as the model count increases.
-Finally, adevs employs the full dependency graph, contrasted to dxex kernels which only observe 1-edge removed nodes.
-%checkme
-\subsection{CD++}
-%TODO explain about CD++, the most related in terms of synchronization --> we implement everything in a single program, with a non-fragmented code-base
-%TODO make mention of Warped, which is the kernel used in CD++, and why we don't use it
-Different projects on CD++ offer conservative (CCD++) as well as optimistic (PCD++) parallel simulation.
-In contrast to our single program, with a non-fragmented code-base, neither projects offer both synchronization protocols.
-CD++ relies on the WARPED kernel.
-It is a middleware that provides memory, event, file, time and communication scheduling.
-We did not use the WARPED kernel (nor the underlying MPI), dxex is designed specifically for a shared memory architecture and as such any middleware, while feature-rich, would have lead to unacceptable overhead.
+Several \textsf{DEVS} simulation tools have already been implemented, which bear some similarity to our simulation tool.
+We discuss several dimensions of related work, as we try to compromise between different tools.
+
+In terms of code design and philosophy, dxex is most related to PythonPDEVS~\cite{PythonPDEVS}.
+Performance of PythonPDEVS was still decent, through the use of simulation and activity hints from the modeller.
+This allowed the kernel to optimize its internal data structures and algorithms for the specific model being executed.
+All changes were completely transparant to the model itself, and were completely optional.
+In this spirit, we now offer users the possibility to chose between different synchronization protocols.
+This allows them to chose for the most appropriate synchronization, depending on the actual model being simulated.
+Contrary to PythonPDEVS, however, dxex doesn't support activity hints.
+
+While PythonPDEVS offers very fast turnaround cycles, due to the use of an interpreted language, simulation performance was easily outdone by compiled simulation kernels.
+In terms of performance, adevs~\cite{adevs} offers faster simulation, at the cost of a significant compilation time.
+The turnaround cycle in adevs is much slower though, specifically because the complete simulation kernel is implemented using templates, and thus in header files.
+This prevents use of the simulation kernel as a shared library.
+A similar compilation cycle is seen in PowerDEVS~\cite{PowerDEVS}.
+As a result, the complete simulation kernel has to be compiled again everytime.
+Dxex compromises, as was the case in other compiled simulation kernels, such as vle~\cite{vle}, by seperating the simulation kernel into a shared library.
+After the initial compilation of the simulation tool, only the model has to be compiled and linked to the library.
+This significantly shortens the turnaround cycle, while still offering good performance
+In terms of performance, it therefore comes close to, and sometimes beats, adevs.
+While a more extensive set of benchmarks is required to make accurate comparisons, initial results are promising.
+Despite its high performance, adevs does not support optimistic synchronization, which we have shown to be highly relevant.
+
+Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though mostly in an ad-hoc way.
+For example CD++~\cite{CD++} has both a conservative (CCD++~\cite{CCD++}) and optimistic (PCD++)~\cite{PCD++}) variant.
+Despite the implementation of both, they are different projects entirely, and while they might support a similar set of features, they remain different from each other.
+Dxex, on the other hand, is a single project, where switching between different synchronization protocols is as simple as switching any other configuration parameter.
+
+In summary, dxex tries to find the middle ground between the concepts of PythonPDEVS, the performance of adevs, and the multiple synchronization protocols of CD++.

### 2015-12-10 paper/5-related-work.tex
Added small mention that CD++ works using distributed simulation, which is not necessarily better

@@
 This allows them to chose for the most appropriate synchronization, depending on the actual model being simulated.
-Contrary to PythonPDEVS, however, dxex doesn't support activity hints.
+Contrary to PythonPDEVS, however, dxex doesn't support model migrations~\cite{PythonPDEVS2} or activity hints~\cite{PythonPDEVS_ACTIMS}.
 
@@
 Dxex, on the other hand, is a single project, where switching between different synchronization protocols is as simple as switching any other configuration parameter.
+CD++, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourself to parallel simulation.
+By limiting our approach to parallel simulation, we are able to achieve higher speedups through the use of shared memory communication.
 

### 2015-12-10 paper/6-conclusions.tex
Label everything to make the outline work
Conclusions

--- 2015-12-05 paper/6-conclusions.tex
@@
-%TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
-Sequentially dxex outperforms adevs in hierarchical simulations while yielding to adevs in broadcast-like simulations. 
-Parallel the results are more nuanced.
-The sensitivity to broadcast simulations is carried over in the parallel implementation, resulting in performance loss for both synchronization protocols.
-Optimistic can still achieve good performance if the volume of inter-kernel messages is low.
-Both  are fast in non-cyclic simulations, with conservative better suited whenever state-saving becomes expensive and optimistic when lookahead is not available.\\
-Finally, the current optimistic implementation is ill-suited for memory constrained systems, especially in simulations where the frequency of reverts is high.
-\subsection{Future work}
-%TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))
-\subsubsection{Memory}
-Our optimistic implementation can benefit from a faster GVT algorithm such as FujiMoto's \cite{Fujimoto:1997:CGV:268403.268404} or more recently \cite{Bauer:2005:SND:1069810.1070159}. 
-With access to the memory subsystem in place (pools), optimistic could be constrained using per kernel quota preventing the simulation from exhausting memory.
-While less sensitive to memory pressure, the same approach could benefit conservative.
-\subsubsection{Activity}
-As shown in \cite{PythonPDEVS_ACTIMS} activity and allocation of models across kernels is a key aspect in achieving high performance in any parallel implementation.
-Allocating models so that there are no dependency cycles between their containing kernels is a first step, but not always possible.
-For optimistic one can use re-allocation to break (runtime dependency) cycles and/or perform load-balancing.
-If kernels are unevenly balanced they will begin to drift fast, causing increasingly more reverts.
-There is a limited framework implemented to track model activity for debugging purposes, this could be extended to enable the above strategies.
-\subsubsection{Hybrid}
-The optimistic implementation could use (null/eot/eit) from conservative to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.
-Conservative kernels could be extended with runtime information about influencing models, for example if one can guarantee a static dependency is not used for a fixed time-span, this can be removed for that period of (virtual) time.
-Ultimately the simulation could switch at runtime between protocols based on the information provided by activity tracking.
-This requires the above mentioned framework, and could only be done at timepoints where a GVT/LBTS time is agreed upon between the kernels.
-From the model point of view no changes are needed, although a non-trivial lookahead is obviously desired.
+In this paper, we introduced dxex, a new C++11-based \textsf{Parallel DEVS} simulation tool.
+Our main contribution is the implementation of different pluggable synchronization kernels for parallel simulation.
+We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
+Dxex allows the user to chose between both conservative and optimistic synchronization, as simply as any other configuration option.
+Notwithstanding this modularity, we have shown that dxex achieves performance similar to adevs, another very efficient \textsf{DEVS} simultion tool.
+Performance is measured both in CPU time, and memory usage.
+
+Future work exists in several directions.
+First, we wish to make optimistic synchronization more tolerant to low-memory situations.
+In its current state, simulation will simply halt with an out-of-memory error.
+Having simulation control, which can throttle the speed of nodes that use up too much memory, has been shown to work in these situations~\cite{FujimotoBook}.
+In our implementation specifically, the out-of-memory problem is aggravated by a slow GVT implementation.
+Algorithms as those presented by~\cite{Fujimoto:1997:CGV:268403.268404} or~\cite{Bauer:2005:SND:1069810.1070159} might help to alleviate this problem.
+Second, the idea of activity can be implemented for our simulation kernels, making it possible to dynamically switch between conservative and optimistic synchronization when changes in the model behaviour are detected.
+Third, activity algorithms, as already implemented by PythonPDEVS, could also be implemented in dxex, to determine how they influence simulation performance.

### 2015-12-10 paper/4-performance.tex
Added a bit of explanation on PHOLD already

@@
 \subsubsection{PHold}
-In the Phold model, we first investigate the influence of the fraction of remote events.
-%TODO wait for results
-%The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing the conservative case.
-The lookahead is $\epsilon$, so the conservative case spends most of its time crawling in steps of $\epsilon$.
-Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation.
-For $N$ kernels, each kernel has to query the null-time of $N-1$ kernels, resulting in O($N^2$) polling behaviour.
-In a non-cyclic simulation with a non-trivial lookahead (like in, e.g., Devstone), that choice does pay off (see Figure~\ref{fig:DevstoneParallel}).
-Adevs's lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code.\\
-%The optimistic case suffers little from the above problems; due to the high interconnectivity, however, a cascading revert is still possible.
-With the percentage of remotes equal to $100$, Phold reflects interconnect in behaviour, which is why the $R$ value is not dimensioned here.
-A revert is very expensive in PHold due to our usage of C++11's random nr generators.
-The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages.
-Once a revert happens the drift between kernels increases fast, increasing the likelihood of more reverts.
-Despite all this, optimistic can for low $R$ values quickly exploit the uncertainty that slows down conservative in this benchmark.
+In the Phold model, we first investigate the influence of the fraction of remote events on the speedup.
+When remote events are rare, optimistic synchronization will also have rare rollbacks, thus increasing performance.
+With more common remote events, however, optimistic synchronization quickly loses its advantage due to the more frequent rollbacks.
+Conservative synchronization, on the other hand, is largely unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block.
+Even though a single synchronization protocol suffices in this case, it shows how different synchronization protocols respond differently to a changing model.
+Adevs, on the other hand, is significantly slower during conservative synchronization.
+Analysis of profiling results shows that this was caused by exception handling within the adevs simulation kernel.
 

### 2015-12-11 paper/3-features.tex
Update class diagram with more information about overridden methods.
Added explanation of class diagram

@@
 
+An overview of this design is given in Figure~\ref{fig:class_diagram}.
+It shows that there is a simulation \texttt{Core}, which has several \texttt{AtomicModels} connected to it.
+The default \texttt{Core} is merely the sequential simulation core.
+Its subclasses define specific variants of the simulation algorithm, like \texttt{ConservativeCore} (conservative synchronization), \texttt{OptimisticCore} (optimistic synchronization), and \texttt{DynamicCore} (\textsf{Dynamic Structure DEVS}).
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{fig/cores_class_diagram.eps}
+	\caption{Dxex kernel design.}
+	\label{fig:class_diagram}
+\end{figure}
+
 \subsubsection{Conservative}

### 2015-12-11 paper/abstract.tex
Review
Proofread abstract

@@
 becomes necessary to perform the simulation within reasonable time bounds.
-The built-in parallelism of Parallel DEVS is often insufficient to tackle this
-problem on its own. Several synchronization algorithms have been proposed,
-each with a specific kind of simulation model in mind. Due to the significant
-differences between these algorithms, current Parallel DEVS simulation tools
-restrict themself to only one such algorithm. In this paper, we present a
-Parallel DEVS simulator, grafted on C++11, which offers both conservative and
-optimistic simulation. We evaluate the performance gain that can be obtained
-by choosing the most appropriate synchronization protocol. Our implementation
-is compared to ADEVS using hardware-level profiling on a spectrum of benchmarks.
+The built-in parallelism of Parallel DEVS is often insufficient to tackle this problem on its own.
+Several synchronization protocols have been proposed, each with their distinct advantages and disadvantages.
+Due to the significantly different implementation of both protocols, most Parallel DEVS simulation tools are limited to only one such protocol.
+In this paper, we present a Parallel DEVS simulator, grafted on C++11, but based on PythonPDEVS, which supports both conservative and optimistic synchronization.
+We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol.
+Performance results are compared to adevs, in terms of CPU time and memory usage.

### 2015-12-11 paper/1-introduction.tex
Proofread introduction

@@
-%TODO restructure and let it "flow" better
 \textsf{DEVS}~\cite{ClassicDEVS} is a popular formalism for modelling complex dynamic systems using a discrete-event abstraction.
@@
 
-With the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform the simulation within reasonable time bounds.
-And while \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, this is often still insufficient.
-Several synchronization algorithms from the discrete event simulation community~\cite{FujimotoBook} have been applied to \textsf{DEVS} simulation in particular.
-While several parallel \textsf{DEVS} simulation kernels exist, they often restrict themselves to a single synchronization protocol.
-This is a logical choice, as these synchronization protocols have nearly no commonalities.
-But the exact reason for different synchronization protocols, is that their distinct nature makes them very applicable to specific situations.
-As such, current \textsf{DEVS} simulation kernels always allow parallel simulation, but good performance can only be expected from some models.
+But with the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform the simulation within reasonable time bounds.
+And while \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, this is often insufficient.
+Several synchronization protocols from the discrete event simulation community~\cite{FujimotoBook} have been applied to \textsf{DEVS} simulation.
+While several parallel \textsf{DEVS} simulation kernels exist, they are often limited to a single synchronization protocol.
+The exact reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other in specific models.
+The applicability of parallel simulation capabilities of current tools, is therefore limited to the domain of simulation models.
 
-Users that wish to simulate a set of distinct models, with different synchronization protocols, are therefore out of luck:
-Either they accept the lower performance for some of the models, or they use two distinct simulation kernels.
-Neither of these options is acceptable for the simulation of complex models, as the performance can become arbitrary worse, and the simulation kernels can diverge significantly.
+Users that wish to simulate a wide variety of models, with different ideal synchronization protocols, are therefore out of luck:
+either they accept the lower performance for some of the models, or they use two different simulation kernels.
+Neither is acceptable for the simulation of complex models: performance can decrease to unacceptable levels, and simulation kernel APIs can diverge significantly.
 
@@
 
-A simple model is used to prove that the ideal synchronization algorithm is indeed dependent on model details.
-In our case, a single parameter of the model can have a significant impact on simulation performance, and especially on which is the ideal synchronization protocol.
-In order to show that this flexibility does not counter our performance, we compare our tool to Adevs, currently one of the fastest \textsf{DEVS} simulation tools available~\cite{PythonPDEVS,DEVSSurvey}.
+We implemented a model that, depending on a single parameter, changes its ideal synchronization protocols.
+Dxex, then, is used to compare simulation using exactly the same tool, but with a different synchronization protocol.
+As dxes is able to simulate it both ways, users can always opt to use the fastest protocol available.
+To verify that our flexibility does not counter performance, we compare to adevs, currently one of the fastest \textsf{DEVS} simulation tools available~\cite{PythonPDEVS,DEVSSurvey}.
 
-We introduce the necessary background for this paper in Section~\ref{sec:2-background}.
-Section~\ref{sec:3-features} elaborates on our features and design.
-In Section~\ref{sec:4-performance}, we evaluate performance of our tool by comparing its different synchronization protocols.
+Dxex, as well as all of our benchmark models, results, and profiling results, can be found online at \url{https://bitbucket.org/bcardoen/devs-ex-machina}.
+
+The remainder of this paper is organized as follows:
+Section~\ref{sec:2-background} introduces the necessary background on synchronization protocols and our implementation language.
+Section~\ref{sec:3-features} elaborates on some of our features, and our design that enables our flexibility.
+In Section~\ref{sec:4-performance}, we evaluate performance of our tool by comparing its different synchronization protocols, as well as a comparison to adevs.
 Related work is discussed in Section~\ref{sec:5-related-work}.

### 2015-12-11 paper/1-introduction.tex
Make space by removing C++11 background

@@
 While several parallel \textsf{DEVS} simulation kernels exist, they are often limited to a single synchronization protocol.
-The exact reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other in specific models.
-The applicability of parallel simulation capabilities of current tools, is therefore limited to the domain of simulation models.
+The reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other in specific models.
+The applicability of parallel simulation capabilities of current tools is therefore limited.
 
-Users that wish to simulate a wide variety of models, with different ideal synchronization protocols, are therefore out of luck:
-either they accept the lower performance for some of the models, or they use two different simulation kernels.
-Neither is acceptable for the simulation of complex models: performance can decrease to unacceptable levels, and simulation kernel APIs can diverge significantly.
+Users who simulate a wide variety of models, with different ideal synchronization protocols, are therefore out of luck:
+either they accept lower performance for some models, or they use different simulation kernels.
+Neither is acceptable for complex models: performance can decrease to unacceptable levels, and simulation kernel APIs can diverge significantly.
 
-In this paper, we introduce DEVS-Ex-Machina (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
+This paper introduces DEVS-Ex-Machina (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
 The selected synchronization protocol is transparant to the simulated model.
 Users should merely determine, at the start of simulation, which protocol they wish to use.
-Our tool is implemented in C++11, to increase both performance and portability across different platforms.
+Our tool is based on PythonPDEVS, but implemented in C++11, to increase both performance and portability across different platforms.
 
@@
 
-Dxex, as well as all of our benchmark models, results, and profiling results, can be found online at \url{https://bitbucket.org/bcardoen/devs-ex-machina}.
+Dxex can be found online at \url{https://bitbucket.org/bcardoen/devs-ex-machina}.
 
 The remainder of this paper is organized as follows:
-Section~\ref{sec:2-background} introduces the necessary background on synchronization protocols and our implementation language.
+Section~\ref{sec:2-background} introduces the necessary background on synchronization protocols.
 Section~\ref{sec:3-features} elaborates on some of our features, and our design that enables our flexibility.

### 2015-12-11 paper/2-background.tex
Make space by removing C++11 background

@@
-This section briefly introduces two distinct synchronization protocols, as used by dxex.
-Furthermore, we make note of several new features of C++11.
+This section briefly introduces the two synchronization protocols used by dxex: conservative and optimistic synchronization.
 
@@
 While optimistic synchronization does not explicitly depends on the lookahead, simulation performance is still bound by the lookahead implicitly.
-
-\subsection{C++11 Parallelism Features}
-Apart from various other additions, C++11 adds a wide range of portable synchronization primitives to the Standard Library.
-In earlier versions of the language standard, one had to resort to non-portable C implementations.
-Furthermore, C++11 is the first version of the standard that explicitly defines a multi-threaded abstract machine memory model.
-Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform.
-
-Simulation values that need to be shared between nodes, can now be coded portable and efficiently through the use of atomic fields.
-This avoids latencies caused by the exchange of synchronization messages.
-For an in-depth study, we refer to~\cite{CPE:CPE3007}.

### 2015-12-11 paper/1-introduction.tex

@@
 The remainder of this paper is organized as follows:
-Section~\ref{sec:2-background} introduces the necessary background on synchronization protocols.
+Section~\ref{sec:2-background} introduces the necessary background on synchronization protocols and our implementation language.
 Section~\ref{sec:3-features} elaborates on some of our features, and our design that enables our flexibility.

### 2015-12-11 paper/3-features.tex

@@
 
-While we will not make a detailed comparison with PythonPDEVS here, dxex also supports direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing framework~\cite{PythonPDEVS}.
+While we will not make a detailed comparison with PythonPDEVS here, but will attempt to provide a brief overview of some supported features.
+Similarly to PythonPDEVS, dxex supports direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
 But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols.
@@
 
+In our implementation, we thus make explicit use of these new C++11 synchronization primitives.
+Whereas this was also possible in previous versions of the C++ standard, it was not a part of the standard itself.
+C++11 allows us to make our implementation portable, as well as more efficient: compilers might know of optimizations specific to atomic variables.
+
 \subsubsection{Optimistic}
@@
 We use Mattern's GVT algorithm~\cite{mattern} to determine the Global Virtual Time (GVT) using at most $2n$ synchronization messages.
-This process runs asynchronously from the simulation itself.
-Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed.
+This process runs asynchronously from the simulation itself, thus there is no blocking whatsoever.
+Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed, as well as committing irreversible actions.
+
+The main problem we encountered in our implementation is the agressive use of memory.
+Frequent memory allocation and deallocation caused significant overheads.
+This made us switch to the use of thread-local memory pools.
+Again, we made use of the specific features offered by C++11 that were not available in Python, or even previous versions of the C++ language standard.
 

### 2015-12-11 paper/2-background.tex
Proofread section 2.1

@@
 \subsection{Conservative Synchronization}
-The first synchronization protocol that we will introduce, is \textit{conservative synchronization}~\cite{FujimotoBook}.
-In conservative synchronization, a node is allowed to progress in simulated time, independent of all other nodes, up to the point where it can guarantee that no causality errors can happen.
-When this point in time is reached, the node has to block until it is allowed to progress any further.
-In practice, this means that all nodes need to be aware of the current simulation time of all other nodes, and the time it takes an event to propagate (called \textit{lookahead}).
-Several algorithms are defined in the literature to implement this behaviour.
-An overview is given in~\cite{FujimotoBook}.
+The first synchronization protocol we introduce is \textit{conservative synchronization}~\cite{FujimotoBook}.
+In conservative synchronization, a node progresses independent of all other nodes, up to the point in time where it can guarantee that no causality errors happen.
+When simulation reaches this point, the node blocks until it can guarantee a new time until which no causality errors happen.
+In practice, this means that all nodes are aware of the current simulation time of all other nodes, and the time it takes an event to propagate (called \textit{lookahead}).
+Deadlocks can occur due to a dependency cycle of models.
+Multiple algorithms are defined in the literature to handle both the core protocol, as well as resolution schemes to handle or avoid the deadlocks~\cite{FujimotoBook}.
 
-Deadlock can occur when a dependency cycle occurs and the amount of exchanged messages is low.
-Multiple algorithms are defined to handle this situation, such as deadlock avoidance and deadlock recovery.
-
-The main advantage of conservative synchronization is that it has a low overhead if high parallelism exists between nodes.
-Each node can simulate in parallel, while sporadically notifying other nodes that they can progress even further.
-The disadvantage, however, is that the amount of parallelism is explicitly limited by the size of the lookahead.
+The main advantage of conservative synchronization is its low overhead if the lookahead is high.
+Each node then simulates in parallel, and sporadically notifies other nodes about its local simulation time.
+The disadvantage, however, is that the amount of parallelism is explicitly limited by the lookahead.
 If a node can influence another (almost) instantaneously, no matter how rarely it occurs, the amount of parallelism is severely reduced.
-It is also up to the user to define the size of the lookahead, depending on how the model is known to behave.
-Slight changes in model behaviour can cause significant changes to the lookahead, and can therefore also have a significant influence on simulation performance.
+The user is required to define the lookahead, using knowledge about the model's behaviour.
+Defining lookahead is not always a trivial task if there is no detailed knowledge of the model.
+Even slight changes in the model can change to the lookahead, and can therefore have a significant influence on simulation performance.
 

### 2015-12-11 paper/2-background.tex

@@
 A completely different synchronization protocol is \textit{optimistic synchronization}~\cite{TimeWarp}.
-Whereas conservative synchronization would prevent causality errors at all costs, optimistic synchronization will allow them to happen, but correct them afterwards.
-Each node is allowed to progress in simulated time as much as possible, without taking note of the state of any other node.
-When an event arrives at a node, which is already further in simulated time, the node will have to roll back its state to right before the event would normally have to be processed.
-As the simulation time is now rolled back to before the event is processed, the event can simply be processed as if no causality error ever occured.
+Whereas conservative synchronization prevents causality errors at all costs, optimistic synchronization allows them to happen, but corrects them.
+Each node simulates as fast as possible, without taking note of any other node.
+It assumes that no events occur from other nodes, unless it has explicitly received one at that time.
+When this assumption is violated, the node rolls back its simulation time and state to right before the moment when the event has to be processed.
+As simulation time is rolled back to before where the event must be processed, the event can just be processed as if no causality error ever occured.
 
-Rolling back the simulation time requires the node to store past model states, such that they can be restored later on.
-Furthermore, all incoming and outgoing events need to be stored as well.
-Incoming events need to be passed to the models again, when the correct simulation time has again been reached, and outgoing events need to be cancelled, as potentially a different series of output events would normally have been generated.
+Rolling back simulation time requires the node to store past model states, such that they can be restored later.
+All incoming and outgoing events need to be stored as well.
+Incoming events are injected again after a rollback, when their time has been reached again.
+Outgoing events are cancelled after a rollback, through the use of anti-messages, as potentially different output events have to be generated.
 Cancelling events, however, can cause further rollbacks, as the receiving node might also have to roll back its state.
-In practice, a single causality error could have significant repercussions on the complete simulation.
+In practice, a single causality error can have significant repercussions.
 
-Further changes are required for unrecoverable operations, such as I/O (\textit{e.g.}, tracing, writing to file, printing output) and memory management.
-Lightweight algorithms are still required to determine the lower bound of all simulation times, through the computation of a \textit{Global Virtual Time} (GVT).
+Further changes are required for unrecoverable operations, such as I/O and memory management.
+These are only executed after the lower bound of all simulation times, called \textit{Global Virtual Time} (GVT), has progressed beyond their execution time.
 
 The main advantage is that performance is not limited by a small lookahead, caused by a very infrequent event.
-If an (almost) instantaneous event occurs rarely, performance will only be impacted if it occurs, and not at every simulation step.
-The main disadvantage is unpredictable performance and arbitrary cost of rollbacks due to the propagation of causality errors.
-If rollbacks occur frequently, simulation quickly becomes slow, as the overhead of the recovery mechanisms becomes significant.
-Apart from overhead in CPU time, a significant memory overhead is present: all intermediate states qre stored up to a point where it can be considered \textit{irreversible}.
+If an (almost) instantaneous event rarely occurs, performance is only impacted when it occurs, and not at every simulation step.
+The disadvantage is unpredictable performance due to the arbitrary cost of rollbacks and their propagation.
+If rollbacks occur frequently, state saving and rollback overhead can cause simulation to grind to a halt.
+Apart from overhead in CPU time, a significant memory overhead is present: all intermediate states are stored up to a point where it can be considered \textit{irreversible}.
 
-While optimistic synchronization does not explicitly depends on the lookahead, simulation performance is still bound by the lookahead implicitly.
+Note that, while optimistic synchronization does not explicitly depends on lookahead, performance still implicitly depends on lookahead.

### 2015-12-11 paper/3-features.tex
Proofread introduction to section 3

@@
 Historically, dxex is based on PythonPDEVS~\cite{PythonPDEVS}.
-While Python is a good language to create software prototypes, its performance has proven to be insufficient to compete with other simulation kernels~\cite{MasterThesis}.
-Dxex implements only a subset of PythonPDEVS, but makes some of its own additions.
-The core simulation algorithm and optimizations, however, are highly similar.
+Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
+Dxex is a C++11-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, whiel making some of its own additions.
+So while the feature set is not too comparable, the architectural design, core simulation algorithm, and optimizations, are highly similar.
 
-While we will not make a detailed comparison with PythonPDEVS here, but will attempt to provide a brief overview of some supported features.
-Similarly to PythonPDEVS, dxex supports direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
+We will not make a detailed comparison with PythonPDEVS here, but only mention some supported features.
+Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
 But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols.
-This is in line with the design principles of PythonPDEVS: offer users the option to pass information on how to efficiently simulate the model.
-In our case, it now becomes possible to pass the simulation kernel the ``hint'' as to which synchronization protocol would be ideal for this model.
-Furthermore, the implementation in C++11 allows many more (static) optimizations, which were plainly impossible when using an interpreted language.
+This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
+In our case, a user can pass the simulation kernel the ``hint'' as to which synchronization protocol must be used for this model.
+Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language.
 
-Note that, since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
+Since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
 This is due to dxex models being grafted on C++11, whereas PythonPDEVS models are grafted on Python.
 
-In the remainder of this section, we will elaborate on our prominent new feature: support for multiple synchronization protocols within the same simulation tool.
+In the remainder of this section, we will elaborate on our prominent new feature: support for multiple synchronization protocols within the same simulation tool, which are offered transparantly to the model.
 

### 2015-12-11 paper/3-features.tex
Rewrite section 3.1 introduction

@@
 \subsection{Synchronization protocols}
-We have previously shown that different synchronization protocols exist, with each of them being optimized for a specific kind of model.
-As no single synchronization protocol is ideal, a general purpose simulation tool should support multiple situations.
-Currently, however, most parallel simulation tools chose only a single synchronization protocol due to the inherent differences between these approaches.
-An uninformed choice on which to implement is insufficient, as performance is likely to be bad.
-We therefore argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization.
+We previously explained the existence of different synchronization protocols exist, each optimized for a specific kind of model.
+As no single synchronization protocol is ideal for all models, a general purpose simulation tool should support multiple protocols.
+Currently, most parallel simulation tools chose only a single synchronization protocol due to the inherent differences between protocols.
+An uninformed choice on which one to implement is insufficient, as performance will likely be bad.
+We argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for dxex.
 
-Each of them is applicable in specific model configurations.
-Conservative synchronization is ideal when high lookahead exists between different nodes, and barely any blocking is necessary.
-Optimistic synchronization is ideal when lookahead is unpredictable, or there are rare (almost) instantaneous events.
-Finally, sequential simulation is still required for models where parallelism is bad, causing significant overhead.
+These different protocols relate to three different model characteristics.
+Conservative synchronization for when high lookahead exists between different nodes, and barely any blocking is necessary.
+Optimistic synchronization for when lookahead is unpredictable, or there are rare (almost) instantaneous events.
+Finally, sequential simulation is still required for models where parallelism is bad, where all protocols actually slow down simulation.
 

### 2015-12-11 paper/3-features.tex
Proofread 3.1 Sequential

@@
 \subsubsection{Sequential}
-Our sequential simulation algorithm is very similar to the one found in PythonPDEVS, including most of its optimizations.
-However, the simulation algorithm is designed to be overloaded by different synchronization protocol implementations.
+Our sequential simulation algorithm is very similar to the one found in PythonPDEVS, including many optimizations.
+Minor modifications were made, though, such that it can be overloaded by different synchronization protocol implementations.
 This way, the \textsf{DEVS} simulation algorithm is implemented once, but parts can be overwritten as needed.
@@
 
-An overview of this design is given in Figure~\ref{fig:class_diagram}.
-It shows that there is a simulation \texttt{Core}, which has several \texttt{AtomicModels} connected to it.
-The default \texttt{Core} is merely the sequential simulation core.
-Its subclasses define specific variants of the simulation algorithm, like \texttt{ConservativeCore} (conservative synchronization), \texttt{OptimisticCore} (optimistic synchronization), and \texttt{DynamicCore} (\textsf{Dynamic Structure DEVS}).
+An overview of dxex's design is given in Figure~\ref{fig:class_diagram}.
+It shows that there is a simulation \texttt{Core}, which simulates the \texttt{AtomicModel}s connected to it.
+The superclass \texttt{Core} is merely the sequential simulation core, but can be used as-is.
+Subclasses define specific variants, such as \texttt{ConservativeCore} (conservative synchronization), \texttt{OptimisticCore} (optimistic synchronization), and \texttt{DynamicCore} (\textsf{Dynamic Structure DEVS}).
 

### 2015-12-11 paper/4-performance.tex
Added two benchmark models already; still drafts from various sources, so should update them
Added phold model, again a draft from elsewhere
Gave figures a more explicit name
Added all necessary plots, but some are not yet completely correct or in-depth enough
fixed a few typos changed compile flags to include -flto

@@
 
-We start of with a comparison of sequential simulation, to show how adevs and dxex relate in this simple case.
+We start off with a comparison of sequential simulation, to show how adevs and dxex relate in this simple case.
 We not only show that our approach does not influence performance negatively, but also that our main simulation algorithm, similar to the one of PythonPDEVS, is significantly faster than the one found in adevs.
@@
 For all benchmarks, results were all well within 1\% deviation of the average, such that only the average is used in the remainder of this section.
-The same compilation flags were used for both adevs and dxex benchmarks (``\texttt{-O3}'').
+The same compilation flags were used for both adevs and dxex benchmarks (``\texttt{-O3 -flto}'').
 To guarantee comparable results, no IO was performed during the benchmarking phase.
@@
     \center
-    %TODO
-    %\includegraphics[width=\columnwidth]{fig/queue_model.eps}
+    \includegraphics[width=\columnwidth]{fig/queue_model.pdf}
     \caption{Queue model for depth 3 and width 2.}
     \label{fig:queue_model}
+\end{figure}
 
-    %TODO
-    %\includegraphics[width=\columnwidth]{fig/interconnect_model.eps}
+\begin{figure}
+    \center
+    \includegraphics[width=\columnwidth]{fig/interconnect_model.pdf}
     \caption{HighInterconnect model for four models.}
     \label{fig:interconnect_model}
+\end{figure}
 
-    %TODO
-    %\includegraphics[width=\columnwidth]{fig/PHOLD_model.eps}
+\begin{figure}
+    \center
+    \includegraphics[width=0.5\columnwidth]{fig/phold_model.pdf}
     \caption{PHOLD model for four models, split over two nodes. In parallel simulation, each coupled model is simulated at a different node.}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{fig/fig1.eps}
+	\includegraphics[width=\columnwidth]{fig/queue_sequential.eps}
 	\caption{Queue benchmark results for sequential simulation.}
@@
 
-%TODO only include sequential results?
 \begin{figure}
-	\includegraphics[width=\columnwidth]{fig/fig3.eps}
+	\includegraphics[width=\columnwidth]{fig/interconnect_sequential.eps}
 	\caption{Interconnect benchmark results for sequential simulation.}
@@
 Similar to the sequential benchmarks, Figure~\ref{fig:queue_benchmark_parallel} shows that dxex again outperforms adevs, using both optimistic and conservative synchronization.
-%TODO wait for raw results to make conclusions
 Behaviour is exactly the same as in sequential simulation, but some speedup is achieved.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{fig/fig2.eps}
+	\includegraphics[width=\columnwidth]{fig/queue_parallel.eps}
 	\caption{Queue benchmark results for parallel simulation using 4 cores.}
@@
 Analysis of profiling results shows that this was caused by exception handling within the adevs simulation kernel.
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
+    \caption{Phold benchmark results for parallel simulation using 4 cores, with varying fraction of remote events.}
+\end{figure}
 
@@
     \includegraphics[width=\columnwidth]{fig/phold_voorlopig.eps}
-    \caption{Phold benchmark results with high-priority events.}
+    \caption{Phold benchmark results for parallel simulation using 4 cores, with varying amount of high-priority events.}
     \label{fig:phold_priority}
@@
 \begin{figure}
-    %TODO
-    %\includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
+    \includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
     \caption{Interconnect benchmark results for parallel simulation.}

### 2015-12-11 paper/3-features.tex

--- 2015-12-11 paper/3-features.tex
@@
 Historically, dxex is based on PythonPDEVS~\cite{PythonPDEVS}.
-Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
-Dxex is a C++11-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, whiel making some of its own additions.
-So while the feature set is not too comparable, the architectural design, core simulation algorithm, and optimizations, are highly similar.
+While Python is a good language to create software prototypes, its performance has proven to be insufficient to compete with other simulation kernels~\cite{MasterThesis}.
+Dxex implements only a subset of PythonPDEVS, but makes some of its own additions.
+The core simulation algorithm and optimizations, however, are highly similar.
 
-We will not make a detailed comparison with PythonPDEVS here, but only mention some supported features.
-Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
+While we will not make a detailed comparison with PythonPDEVS here, but will attempt to provide a brief overview of some supported features.
+Similarly to PythonPDEVS, dxex supports direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
 But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols.
-This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
-In our case, a user can pass the simulation kernel the ``hint'' as to which synchronization protocol must be used for this model.
-Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language.
+This is in line with the design principles of PythonPDEVS: offer users the option to pass information on how to efficiently simulate the model.
+In our case, it now becomes possible to pass the simulation kernel the ``hint'' as to which synchronization protocol would be ideal for this model.
+Furthermore, the implementation in C++11 allows many more (static) optimizations, which were plainly impossible when using an interpreted language.
 
-Since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
+Note that, since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
 This is due to dxex models being grafted on C++11, whereas PythonPDEVS models are grafted on Python.
 
-In the remainder of this section, we will elaborate on our prominent new feature: support for multiple synchronization protocols within the same simulation tool, which are offered transparantly to the model.
+In the remainder of this section, we will elaborate on our prominent new feature: support for multiple synchronization protocols within the same simulation tool.
 
@@
 However, the simulation algorithm is designed to be overloaded by different synchronization protocol implementations.
-This way, the \textsf{DEVS} simulation algorithm is implemented once, but parts can be overwritten as needed.
+This way, the \textsf{DEVS} simulation algorithm is implemented once, but parts can be overridden as needed.
 In theory, more synchronization protocols (\textit{e.g.}, other algorithms for conservative synchronization) can be added without altering our design.
@@
 
-The main problem we encountered in our implementation is the agressive use of memory.
+The main problem we encountered in our implementation is the aggressive use of memory.
 Frequent memory allocation and deallocation caused significant overheads.
@@
 
-\subsection{Transparancy}
+\subsection{Transparency}
 A user must only provide one model, implemented in C++11, which can be simulated by each synchronization kernel.

### 2015-12-11 paper/3-features.tex
Proofread 3.1 Conservative

--- 2015-12-11 paper/3-features.tex
@@
 Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
-But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols.
+But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols (though only in parallel).
 This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
@@
 \subsubsection{Conservative}
-For conservative synchronization, each node determines the nodes it is influenced by.
-Each model provides a lookahead function, which determines the lookahead depending on the current simulation state.
-Within this time interval, it is an error if a model raises an event, thus violating its previous promise.
-The node uses this information to compute its earliest output time (EOT), and writes out the value in shared memory through the use of C++11 synchronization primitives.
+For conservative synchronization, each node must determine the nodes it is influenced by.
+Each model needs to provide a lookahead function, which determines the lookahead depending on the current simulation state.
+Within the returned time interval, the model promises not to raise an event.
+A node aggregates this information to computes its earliest output time (EOT).
+This value is written out in shared memory, where it can be read out by all other nodes.
 
-In our implementation, we thus make explicit use of these new C++11 synchronization primitives.
-Whereas this was also possible in previous versions of the C++ standard, it was not a part of the standard itself.
-C++11 allows us to make our implementation portable, as well as more efficient: compilers might know of optimizations specific to atomic variables.
+Reading and writing to shared memory is done through the use of the new C++11 synchronization primitives.
+Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
+C++11 further allows us to make the implementation portable, as well as more efficient: the compiler might know of optimizations specific to atomic variables.
 

### 2015-12-11 paper/3-features.tex
Proofread section 3.1 Optimistic

@@
 \subsubsection{Optimistic}
-For optimistic synchronization, each node needs to keep track of all intermediate simulation states.
-This needs to be done carefully, in order to avoid unnecessary copies, and minimize the overhead induced for each transition function.
-We use Mattern's GVT algorithm~\cite{mattern} to determine the Global Virtual Time (GVT) using at most $2n$ synchronization messages.
-This process runs asynchronously from the simulation itself, thus there is no blocking whatsoever.
-Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed, as well as committing irreversible actions.
+For optimistic synchronization, each node must be able to roll back to a previous point in time.
+This is often implemented through the use of state saving.
+This needs to be done carefully in order to avoid unnecessary copies, and minimize the overhead.
+We use the default: explicitly save each and every intermediate state.
+Mattern's algorithm~\cite{mattern} is used to determine the GVT, as it runs asynchronously and uses only $2n$ synchronization messages.
+Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed, and irreversible actions are committed.
 
 The main problem we encountered in our implementation is the agressive use of memory.
-Frequent memory allocation and deallocation caused significant overheads.
-This made us switch to the use of thread-local memory pools.
-Again, we made use of the specific features offered by C++11 that were not available in Python, or even previous versions of the C++ language standard.
+Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
+This made us switch to the use of thread-local (using \texttt{tcmalloc}) memory pools.
+Again, we made use of specific new features of C++11, that were not available in Python, or even previous versions of the C++ language standard.
 

### 2015-12-11 paper/3-features.tex
Proofread section 3.2

@@
 \subsection{Transparancy}
-A user must only provide one model, implemented in C++11, which can be simulated by each synchronization kernel.
-The exception is conservative synchronization: a lookahead function is required, whereas this is not possible in other synchronization kernels.
-Two options are possible: either a lookahead function is always provided, even when it is not required and possibly not used, or using a default lookahead function if none is defined.
+We define simulation kernel transparancy as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
+User should thus only provide one model, implemented in C++11, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.
+Switching between simulation kernels is as simple as altering the simulation termination time.
+The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
+Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.
 
 Always defining a lookahead function might seem redundant, especially if users will never use conservative synchronization.
-The more attractive option is for the simulation tool to provide a default lookahead function, defined by the minimum detected time advance.
-This lookahead value is most likely too small, but will prevent causality errors at the cost of performance.
+Especially since defining the lookahead is often non-trivial and dependent on intricate model details.
+The more attractive option is for the simulation tool to provide a default lookahead function, such that simulation can run anyway, but maybe not at peak performance.
 Depending on the model, simulation performance might still be faster than sequential simulation.
 
-Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity.
+Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default can be used otherwise.

### 2015-12-11 paper/4-performance.tex
updated benchmark system & fixed typos in section 4

@@
 Since adevs supports only conservative synchronization, we don't mention optimistic synchronization results there.
-All experiments were performed using four simulation nodes, and executed on a quad-core machine.
+All experiments were performed using up to six simulation nodes, and executed on a hexa-core machine.
 
@@
 Also, results for optimistic synchronization vary heavily depending on thread scheduling by the operating system, as this influences the drift between different nodes.
-Comparing similar approaches though, we notice that dxed and adevs have very similar memory use.
+Comparing similar approaches though, we notice that dxex and adevs have very similar memory use.
 

### 2015-12-11 paper/3-features.tex

@@
 Minor modifications were made, though, such that it can be overloaded by different synchronization protocol implementations.
-This way, the \textsf{DEVS} simulation algorithm is implemented once, but parts can be overwritten as needed.
+This way, the \textsf{DEVS} simulation algorithm is implemented once, but parts can be overridden as needed.
 In theory, more synchronization protocols (\textit{e.g.}, other algorithms for conservative synchronization) can be added without altering our design.
@@
 
-The main problem we encountered in our implementation is the agressive use of memory.
+The main problem we encountered in our implementation is the aggressive use of memory.
 Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
@@
 
-\subsection{Transparancy}
+\subsection{Transparency}
 We define simulation kernel transparancy as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.

### 2015-12-11 paper/4-performance.tex
Proofread section 4 introduction

--- 2015-12-11 paper/4-performance.tex
@@
-In this section, we evaluate the performance of our simulation tool.
-We show that the inclusion of multiple synchronization protocols does not decrease our performance to the point where it is unusable.
-To this end, we compare to adevs, one of the most efficient simulation kernels at this time~\cite{DEVSSurvey}.
-A comparison is made for both the CPU and memory usage of both sequential simulation and parallel simulation.
+In this section, we evaluate performance of dxex using its different synchronization protocols.
+We show that the inclusion of multiple synchronization protocols does not decrease performance significantly.
+To this end, we compare to adevs, currently one of the most efficient simulation kernels~\cite{DEVSSurvey}.
+CPU time and memory usage is compared for both sequential and parallel simulation.
 
 We start off with a comparison of sequential simulation, to show how adevs and dxex relate in this simple case.
-We not only show that our approach does not influence performance negatively, but also that our main simulation algorithm, similar to the one of PythonPDEVS, is significantly faster than the one found in adevs.
-Similar differences, compared to adevs, can also be seen in the parallel simulation benchmarks.
-In the parallel simulation benchmarks, the benefit of our different synchronization protocols is also indicated.
+For the parallel simulation benchmarks, results are presented for both conservative and optimistic synchronization.
 
-For all benchmarks, results were all well within 1\% deviation of the average, such that only the average is used in the remainder of this section.
+For all benchmarks, results are well within a 1\% deviation of the average, such that only the average is used in the remainder of this section.
 The same compilation flags were used for both adevs and dxex benchmarks (``\texttt{-O3 -flto}'').
-To guarantee comparable results, no IO was performed during the benchmarking phase.
-Simulation traces were used to verify that both adevs and dxex models have exactly the same behaviour.
-All benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.
+To guarantee comparable results, no I/O was performed during the benchmarking phase such that computation was CPU-bound.
+Simulation traces are used to verify that both adevs and dxex models have exactly the same behaviour.
+Benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.
 

### 2015-12-11 adevs/benchmarks/highinterconnect.cpp
adapted adevs interconnect allocation of models - - -> models are now striped over the simulation cores, instead of being randomly assigned to them by adevs

@@ -137,9 +137,9 @@ public:
 class HighInterconnect: public adevs::Digraph<t_payload, int>
 {
 public:
+        std::vector<Generator*> ptrs;
 	HighInterconnect(std::size_t width, bool randomta)
 	{
-		std::vector<Generator*> ptrs;
 		for(std::size_t i = 0; i < width; ++i){
 			Generator* pt = new Generator(std::string("Generator") + n_tools::toString(i), 1000*i, randomta);
                         ptrs.push_back(pt);
@@ -201,7 +201,7 @@ char getOpt(char* argv){
 }
 
 
-const char helpstr[] = " [-h] [-t ENDTIME] [-w WIDTH] [-r] [-c COREAMT] [classic|cpdevs|opdevs|pdevs]\n"
+const char helpstr[] = " [-h] [-t ENDTIME] [-w WIDTH] [-r] [-c COREAMT] [classic|cpdevs]\n"
 	"options:\n"
 	"  -h           show help and exit\n"
 	"  -t ENDTIME   set the endtime of the simulation\n"
@@ -210,7 +210,6 @@ const char helpstr[] = " [-h] [-t ENDTIME] [-w WIDTH] [-r] [-c COREAMT] [classic
 	"  -c COREAMT   amount of simulation cores, ignored in classic mode. Must not be 0.\n"
 	"  classic      Run single core simulation.\n"
 	"  cpdevs       Run conservative parallel simulation.\n"
-	"  opdevs|pdevs Run optimistic parallel simulation.\n"
 	"note:\n"
 	"  If the same option is set multiple times, only the last value is taken.\n";
 
@@ -305,6 +304,11 @@ int main(int argc, char** argv)
 		sim.execUntil(eTime);
 	} else {
 		omp_set_num_threads(coreAmt);	//must manually set amount of OpenMP threads
+		size_t i = 0;
+		for(Generator* ptr: ((HighInterconnect*)model)->ptrs){
+		                        i = (i+1)%coreAmt;
+		                        ptr->setProc(i);
+		        }
 		adevs::ParSimulator<t_event> sim(model);
 #ifndef BENCHMARK
 		sim.addEventListener(listener);

### 2015-12-11 paper/4-performance.tex
Proofread section 4.1

@@
 We use a selection of benchmarks, based on those found in the literature.
-Three different types of benchmark are defined, each for a different purpose:
+Three different types of benchmark are defined:
 \begin{enumerate}
-    \item \textit{Queue} model, based on the HI model of DEVStone~\cite{DEVStone}, creates a chain of atomic models, which are hierarchically nested in each other.
+    \item The \textit{Queue} model, based on the HI model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
           A single generator will push events into the queue, which get processed by the processors after a fixed or random delay.
-          It takes two parameters: width and depth, which determine the widht and depth of the hierarchy.
-          This benchmark shows how the complexity of the simulation kernel behaves with an increasing amount of atomic models, and an increasingly deep hierarchy.
-          If the processing delay is fixed for all processors, further insight is provided in the collision handling performance of the simulation kernel.
+          It takes two parameters: width and depth, which determine the width and depth of the hierarchy.
+          This benchmark shows how the complexity of the simulation kernel behaves for an increasing amount of atomic models, and an increasingly deep hierarchy.
           An example for width 2 and depth 3 is shown in Figure~\ref{fig:queue_model}.
 
-    \item \textit{PHOLD} model, presented by~\cite{PHOLD}.
-          It creates a set of $n$ atomic models, and each model has exactly $n-1$ output ports: one for every other atomic model.
-          Couplings are made such that each atomic model is directly connected to each other atomic model, such that every atomic model can directly send an event to every other atomic model.
+    \item The \textit{PHOLD} model, presented by~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
+          Couplings are made to directly connect each atomic model to every other atomic model.
           After a random delay, atomic models will send out an event to a randomly selected output port.
           Output port selection happens in two phases: first it is decided whether the event should be sent to an atomic model inside or outside of this coupled model.
-          Afterwards, a uniform selection is made between the possible ports.
-          It takes one parameter: the percentage of remote events, which influences the fraction of messages routed to other coupled models.
+          Afterwards, a uniform selection is made between the remaining ports.
+          The model takes one parameter: the percentage of remote events, which determines the fraction of messages routed to other coupled models.
           This benchmark shows how the simulation kernel behaves in the presence of many local or remote events.
@@
 
-    \item \textit{HighInterconnect} model, a merge of the HI model of DEVStone~\cite{DEVStone} and PHOLD~\cite{PHOLD}.
-          It creates a structure similar to the one from PHOLD, but instead of $n-1$ output ports, every atomic model has only a single output port.
-          All models are still connected to each other, but through the use of broadcasting: every model will receive a generated event.
-          It takes one parameter: the number of models.
-          This benchmark investigates the complexity of the routing algorithm.
+    \item The \textit{HighInterconnect} model, a merge of the HI model of DEVStone~\cite{DEVStone} and PHOLD~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly one output port.
+          All models are still connected to one another, but all over the same port: every model will receive a generated event.
+          The model takes one parameter: the number of models.
+          This benchmark investigates the complexity of the event routing algorithm, and how the simulation kernel handles many simultaneous events.
           An example for four models is shown in Figure~\ref{fig:interconnect_model}.
@@
 
-We opted to deviate from the DEVStone benchmark, as DEVStone tends towards unrealistic models since all internal and external transitions occur simultaneously.
-In our benchmark models, there is always the option for simultaneous transition functions (\textit{fixed time advance}), or scattered transition functions (\textit{random time advance}).
-Furthermore, they defined the use of an artificial load function, which easily skews the result, making the actual simulation algorithm barely comparable.
+We opted to deviate from the DEVStone benchmark, as DEVStone tends towards unrealistic models since most transitions occur simultaneously.
+In all of our benchmark models, simultaneous internal transitions are avoided as much as possible through the use of a random time advance.
+DEVStone also uses an artificial load function, which easily skews the result: the simulation kernel overhead is dominated by the artificial load.
 

### 2015-12-11 paper/4-performance.tex
Proofread sectoin 4.2

@@
 Despite our core contribution being mainly in the parallel simulation, we still value a comprehensive comparison of sequential simulation results.
-First, and foremost, as parallel simulation results are tightly linked to the sequential simulation results: parallel simulation merely adds a synchronization layer over different, essentially sequential, simulation kernels.
-Second, since parallel simulation results are validated through the use of adevs.
+First, and foremost, as parallel simulation results are tightly linked to the sequential simulation results: parallel simulation merely adds a synchronization layer over multiple, essentially sequential, simulation kernels.
+Second, parallel simulation results are validated through the use of adevs.
 To provide a more comprehensive comparison to adevs in the parallel simulation benchmarks, sequential simulation results need to be compared.
-Only the Queue and HighInterconnect models are relevant for sequential simulation.
+Only the Queue and HighInterconnect models are relevant for sequential simulation, so we will not touch upon the PHold model yet.
 
@@
 As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
-Through careful analysis of profiling results, we determined that adevs spends much time in handling simulation messages, whereas this is mostly avoided due to the differently designed simulation algorithm of dxex.
-Both simulation tools have quadratically increasing execution times, though dxex is much faster thanks to its more efficient simulation control algorithms.
+Through careful analysis of profiling results, we determined that adevs spends much time while handling simulation messages, while this is mostly avoided in dxex's differently designed simulation algorithm.
+Both simulation tools have similar complexities, though dxex is much faster thanks to its more efficient simulation control algorithms.
 
@@
 As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs now outperforms dxex by a fair margin.
-Analysis showed that this slowdown is caused by the high amount of exchanged events.
-Event creation is found to be much slower in dxex than it is in adevs, even despite the use of memory pools in dxex.
+Analysis showed that this slowdown is caused by the high amount of exchanged events: event creation is found to be much slower in dxex than it is in adevs, despite the use of memory pools in dxex.
 

### 2015-12-11 paper/4-performance.tex
Minor updates to performance results explanation, now that plots are there

@@
 This minimizes the number of inter-node messages.
-As the model is a queue, however, the last models will only activate much later on in the simulation.
+As the model is a queue, however, the last models will only activate later in the simulation.
 Since these are allocated to seperate nodes, some nodes will remain idle until simulation has progressed sufficiently far.
 
-Similar to the sequential benchmarks, Figure~\ref{fig:queue_benchmark_parallel} shows that dxex again outperforms adevs, using both optimistic and conservative synchronization.
-Behaviour is exactly the same as in sequential simulation, but some speedup is achieved.
+Similar to the sequential benchmarks, Figure~\ref{fig:queue_benchmark_parallel} shows that dxex again outperforms adevs, but now also in terms of speedup.
+Results indicate that our implementation of conservative synchronization achieves much higher speedups than adevs.
 In this case, conservative synchronization seems to be better than optimistic synchronization, at the cost of providing the lookahead.
@@
 	\includegraphics[width=\columnwidth]{fig/queue_parallel.eps}
-	\caption{Queue benchmark results for parallel simulation using 4 cores.}
+	\caption{Queue benchmark results for parallel simulation of a model of fixed size.}
 	\label{fig:queue_benchmark_parallel}
@@
 In the Phold model, we first investigate the influence of the fraction of remote events on the speedup.
-When remote events are rare, optimistic synchronization will also have rare rollbacks, thus increasing performance.
-With more common remote events, however, optimistic synchronization quickly loses its advantage due to the more frequent rollbacks.
-Conservative synchronization, on the other hand, is largely unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block.
-Even though a single synchronization protocol suffices in this case, it shows how different synchronization protocols respond differently to a changing model.
-Adevs, on the other hand, is significantly slower during conservative synchronization.
+When remote events are rare, optimistic synchronization will rarely have to perform rollbacks, thus increasing performance.
+With more common remote events, however, optimistic synchronization quickly slows down due to the frequent rollbacks.
+Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block.
+Even though a single synchronization protocol is always ideal in this case, it shows how different synchronization protocols respond differently to a changing model.
+Adevs is significantly slower during conservative synchronization.
 Analysis of profiling results shows that this was caused by exception handling within the adevs simulation kernel.
@@
 
-Secondly, we verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization.
+We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization.
 We slightly modified the Phold benchmark, to include high-priority events.
-Contrary to normal events, which offer a sufficiently large lookahead, high-priority events happen almost instantaneous, thus restricting lookahead to a very small value.
-Even though the normal events will occur most oftenly, a conservative implementation should always block since these high-priority events can always occur.
+Contrary to normal events, which offer a sufficiently large lookahead, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
+Even though normal events  occur most often, a conservative implementation always blocks until it can make guarantees.
 An optimistic implementation, however, will simply go forward in simulation time and roll back the few times that these high-priority events happen.
@@
 If barely any high-priority events occur, conservative synchronization is penalized due to its excessive blocking, which often turned out to be unnecessary.
-Should many high-priority events occur, however, optimistic synchronization is penalized due to its mindless progression of simulation, which frequently needed to be reverted.
-These results show that there is no single perfect synchronization algorithm for this model.
+When many high-priority events occur, optimistic synchronization is penalized due to its mindless progression of simulation, which frequently needed to be reverted.
+Results show that there is no single perfect synchronization algorithm for this model.
 Depending on model configuration, either synchronization protocol might be better.
-We therefore claim that our contribution is invaluable for high performance simulation: depending on the observed communication behaviour, modellers can use the most appropriate synchronization protocol.
+We have shown that our contribution is invaluable for high performance simulation: depending on the observed communication behaviour, modellers can use the most appropriate synchronization protocol.
 
@@
 Interestingly, adevs does not achieve any noticable speedup from the use of multiple cores in this situation.
-Optimistic synchronization is not shown in this case: due to the high amount of exchanged events, all of which need to be stored in the sending \textit{and} receiving node, it quickly runs out of memory.
-Several fixes to this problem have been proposed in the literature~\cite{FujimotoBook}, but none of these is implemented by dxex.
 
@@
 \subsubsection{Remarks}
-Both dxex and adevs use tcmalloc as memory allocator.
+Both dxex and adevs use \texttt{tcmalloc} as memory allocator.
 Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
-Tcmalloc only gradually releases memory back to the OS, whereas our pools will not do so at all.
+\texttt{tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
 If memory has been allocated once, it is, at least from a performance point of view, better to keep that memory in the pool indefinitely.
@@
 
-Unsurprisingly, optimistic synchronization results show very high memory use due to the saved states.
+Unsurprisingly, optimistic synchronization results show very high memory usage due to the saved states.
 Note the logarithmic scale that was used for this reason.

### 2015-12-11 paper/4-performance.tex
updated interconnect parallel graph

--- 2015-12-11 paper/4-performance.tex
@@
-In this section, we evaluate performance of dxex using its different synchronization protocols.
-We show that the inclusion of multiple synchronization protocols does not decrease performance significantly.
-To this end, we compare to adevs, currently one of the most efficient simulation kernels~\cite{DEVSSurvey}.
-CPU time and memory usage is compared for both sequential and parallel simulation.
+In this section, we evaluate the performance of our simulation tool.
+We show that the inclusion of multiple synchronization protocols does not decrease our performance to the point where it is unusable.
+To this end, we compare to adevs, one of the most efficient simulation kernels at this time~\cite{DEVSSurvey}.
+A comparison is made for both the CPU and memory usage of both sequential simulation and parallel simulation.
 
 We start off with a comparison of sequential simulation, to show how adevs and dxex relate in this simple case.
-For the parallel simulation benchmarks, results are presented for both conservative and optimistic synchronization.
+We not only show that our approach does not influence performance negatively, but also that our main simulation algorithm, similar to the one of PythonPDEVS, is significantly faster than the one found in adevs.
+Similar differences, compared to adevs, can also be seen in the parallel simulation benchmarks.
+In the parallel simulation benchmarks, the benefit of our different synchronization protocols is also indicated.
 
-For all benchmarks, results are well within a 1\% deviation of the average, such that only the average is used in the remainder of this section.
+For all benchmarks, results were all well within 1\% deviation of the average, such that only the average is used in the remainder of this section.
 The same compilation flags were used for both adevs and dxex benchmarks (``\texttt{-O3 -flto}'').
-To guarantee comparable results, no I/O was performed during the benchmarking phase such that computation was CPU-bound.
-Simulation traces are used to verify that both adevs and dxex models have exactly the same behaviour.
-Benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.
+To guarantee comparable results, no IO was performed during the benchmarking phase.
+Simulation traces were used to verify that both adevs and dxex models have exactly the same behaviour.
+All benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.
 
@@
 We use a selection of benchmarks, based on those found in the literature.
-Three different types of benchmark are defined:
+Three different types of benchmark are defined, each for a different purpose:
 \begin{enumerate}
-    \item The \textit{Queue} model, based on the HI model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
+    \item \textit{Queue} model, based on the HI model of DEVStone~\cite{DEVStone}, creates a chain of atomic models, which are hierarchically nested in each other.
           A single generator will push events into the queue, which get processed by the processors after a fixed or random delay.
-          It takes two parameters: width and depth, which determine the width and depth of the hierarchy.
-          This benchmark shows how the complexity of the simulation kernel behaves for an increasing amount of atomic models, and an increasingly deep hierarchy.
+          It takes two parameters: width and depth, which determine the widht and depth of the hierarchy.
+          This benchmark shows how the complexity of the simulation kernel behaves with an increasing amount of atomic models, and an increasingly deep hierarchy.
+          If the processing delay is fixed for all processors, further insight is provided in the collision handling performance of the simulation kernel.
           An example for width 2 and depth 3 is shown in Figure~\ref{fig:queue_model}.
 
-    \item The \textit{PHOLD} model, presented by~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
-          Couplings are made to directly connect each atomic model to every other atomic model.
+    \item \textit{PHOLD} model, presented by~\cite{PHOLD}.
+          It creates a set of $n$ atomic models, and each model has exactly $n-1$ output ports: one for every other atomic model.
+          Couplings are made such that each atomic model is directly connected to each other atomic model, such that every atomic model can directly send an event to every other atomic model.
           After a random delay, atomic models will send out an event to a randomly selected output port.
           Output port selection happens in two phases: first it is decided whether the event should be sent to an atomic model inside or outside of this coupled model.
-          Afterwards, a uniform selection is made between the remaining ports.
-          The model takes one parameter: the percentage of remote events, which determines the fraction of messages routed to other coupled models.
+          Afterwards, a uniform selection is made between the possible ports.
+          It takes one parameter: the percentage of remote events, which influences the fraction of messages routed to other coupled models.
           This benchmark shows how the simulation kernel behaves in the presence of many local or remote events.
@@
 
-    \item The \textit{HighInterconnect} model, a merge of the HI model of DEVStone~\cite{DEVStone} and PHOLD~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly one output port.
-          All models are still connected to one another, but all over the same port: every model will receive a generated event.
-          The model takes one parameter: the number of models.
-          This benchmark investigates the complexity of the event routing algorithm, and how the simulation kernel handles many simultaneous events.
+    \item \textit{HighInterconnect} model, a merge of the HI model of DEVStone~\cite{DEVStone} and PHOLD~\cite{PHOLD}.
+          It creates a structure similar to the one from PHOLD, but instead of $n-1$ output ports, every atomic model has only a single output port.
+          All models are still connected to each other, but through the use of broadcasting: every model will receive a generated event.
+          It takes one parameter: the number of models.
+          This benchmark investigates the complexity of the routing algorithm.
           An example for four models is shown in Figure~\ref{fig:interconnect_model}.
@@
 
-We opted to deviate from the DEVStone benchmark, as DEVStone tends towards unrealistic models since most transitions occur simultaneously.
-In all of our benchmark models, simultaneous internal transitions are avoided as much as possible through the use of a random time advance.
-DEVStone also uses an artificial load function, which easily skews the result: the simulation kernel overhead is dominated by the artificial load.
+We opted to deviate from the DEVStone benchmark, as DEVStone tends towards unrealistic models since all internal and external transitions occur simultaneously.
+In our benchmark models, there is always the option for simultaneous transition functions (\textit{fixed time advance}), or scattered transition functions (\textit{random time advance}).
+Furthermore, they defined the use of an artificial load function, which easily skews the result, making the actual simulation algorithm barely comparable.
 
@@
 Despite our core contribution being mainly in the parallel simulation, we still value a comprehensive comparison of sequential simulation results.
-First, and foremost, as parallel simulation results are tightly linked to the sequential simulation results: parallel simulation merely adds a synchronization layer over multiple, essentially sequential, simulation kernels.
-Second, parallel simulation results are validated through the use of adevs.
+First, and foremost, as parallel simulation results are tightly linked to the sequential simulation results: parallel simulation merely adds a synchronization layer over different, essentially sequential, simulation kernels.
+Second, since parallel simulation results are validated through the use of adevs.
 To provide a more comprehensive comparison to adevs in the parallel simulation benchmarks, sequential simulation results need to be compared.
-Only the Queue and HighInterconnect models are relevant for sequential simulation, so we will not touch upon the PHold model yet.
+Only the Queue and HighInterconnect models are relevant for sequential simulation.
 
@@
 As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
-Through careful analysis of profiling results, we determined that adevs spends much time while handling simulation messages, while this is mostly avoided in dxex's differently designed simulation algorithm.
-Both simulation tools have similar complexities, though dxex is much faster thanks to its more efficient simulation control algorithms.
+Through careful analysis of profiling results, we determined that adevs spends much time in handling simulation messages, whereas this is mostly avoided due to the differently designed simulation algorithm of dxex.
+Both simulation tools have quadratically increasing execution times, though dxex is much faster thanks to its more efficient simulation control algorithms.
 
@@
 As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs now outperforms dxex by a fair margin.
-Analysis showed that this slowdown is caused by the high amount of exchanged events: event creation is found to be much slower in dxex than it is in adevs, despite the use of memory pools in dxex.
+Analysis showed that this slowdown is caused by the high amount of exchanged events.
+Event creation is found to be much slower in dxex than it is in adevs, even despite the use of memory pools in dxex.
 
@@
 	\includegraphics[width=\columnwidth]{fig/queue_parallel.eps}
-	\caption{Queue benchmark results for parallel simulation using 4 cores.}
+	\caption{Queue benchmark results for parallel simulation using variable amount of cores.}
 	\label{fig:queue_benchmark_parallel}
@@
     \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
-    \caption{Phold benchmark results for parallel simulation using 4 cores, with varying fraction of remote events.}
+    \caption{Phold benchmark results for parallel simulation using 6 cores, with varying fraction of remote events.}
 \end{figure}
@@
     \includegraphics[width=\columnwidth]{fig/phold_voorlopig.eps}
-    \caption{Phold benchmark results for parallel simulation using 4 cores, with varying amount of high-priority events.}
+    \caption{Phold benchmark results for parallel simulation using 6 cores, with varying amount of high-priority events.}
     \label{fig:phold_priority}

### 2015-12-11 paper/4-performance.tex
explained speedup in interconnect/queue benchmarks.

@@
 
-For all benchmarks, results are well within a 1\% deviation of the average, such that only the average is used in the remainder of this section.
+For all benchmarks, results are well within a 5\% deviation of the average, such that only the average is used in the remainder of this section.
 The same compilation flags were used for both adevs and dxex benchmarks (``\texttt{-O3 -flto}'').
@@
 As the model is a queue, however, the last models will only activate later in the simulation.
-Since these are allocated to seperate nodes, some nodes will remain idle until simulation has progressed sufficiently far.
+Since these are allocated to separate nodes, some nodes will remain idle until simulation has progressed sufficiently far.
 
@@
 Results indicate that our implementation of conservative synchronization achieves much higher speedups than adevs.
+This simulation is the ideal case for our conservative implementation, approaching near linear speedup.
 In this case, conservative synchronization seems to be better than optimistic synchronization, at the cost of providing the lookahead.
@@
 Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}
-While conservative simulation delivers the expected speedups, dxex is still significantly slower than adevs.
-Interestingly, adevs does not achieve any noticable speedup from the use of multiple cores in this situation.
+When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing amount of reverts in optimistic.
+In this benchmark, all models depend on each other, negating any possible performance gain by executing the simulation in parallel.
 

### 2015-12-11 paper/4-performance.tex
Update changed call graphs adevs interconnect. Proxy commit R script and new graphs and try to fit them in paper.

--- 2015-12-11 paper/4-performance.tex
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{fig/queue_sequential.eps}
+	\includegraphics[width=\columnwidth,height=6cm]{fig/queue_sequential.eps}
 	\caption{Queue benchmark results for sequential simulation.}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{fig/interconnect_sequential.eps}
+	\includegraphics[width=\columnwidth,height=6cm]{fig/interconnect_sequential.eps}
 	\caption{Interconnect benchmark results for sequential simulation.}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{fig/queue_parallel.eps}
+	\includegraphics[width=\columnwidth,height=6cm]{fig/queue_parallel.eps}
 	\caption{Queue benchmark results for parallel simulation of a model of fixed size.}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
+    \includegraphics[width=\columnwidth,height=6cm]{fig/phold_remotes.eps}
     \caption{Phold benchmark results for parallel simulation using 6 cores, with varying fraction of remote events.}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{fig/phold_voorlopig.eps}
+    \includegraphics[width=\columnwidth,height=6cm]{fig/phold_priority.eps}
     \caption{Phold benchmark results for parallel simulation using 6 cores, with varying amount of high-priority events.}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
+    \includegraphics[width=\columnwidth,height=6cm]{fig/interconnect_parallel.eps}
     \caption{Interconnect benchmark results for parallel simulation.}

### 2015-12-11 Rscript/devs-performance.R

@@ -1,6 +1,8 @@
+# @author Tim Tuijn
 # global variables
 generatePDF <<- FALSE	# if TRUE, generate PDF else: generated EPS files...
 epscount <<- 1
+generateTitles <<- FALSE
 rootpath <<- paste(getwd(), '/', sep = '')
 # rootpath <<- 'putyourcustomrootpathinhere'
 datapath <<- paste(rootpath, 'data/', sep = '')
@@ -128,7 +130,29 @@ myrainbow <- function(n) {
 		return(rainbow(n))
 }
 
-comparesets <- function(datalist, labellist, xlabel, ylabel, chartlabel, legendpos) {
+formatlabels <- function(xvalues, formattype) {
+	if (formattype == "cartesian") {
+		formattedvalues <- array()
+		for (i in 1:length(xvalues)) {
+			formattedvalues[[i]] <- sprintf("%0d",round(as.numeric(xvalues[[i]])*as.numeric(xvalues[[i]])))
+		}
+		return(formattedvalues)
+	}
+	else
+		return(xvalues);
+}
+
+newxlabels <- function(minmax) {
+	delta <- minmax[2] - minmax[1] + 1
+	deltas <- c(1,2,5,10,25,50,100,250,500,1000,10000)
+	for (i in 1:length(deltas)) {
+		if (delta/deltas[i] < 10)
+			return(seq(minmax[1], minmax[2], deltas[i]))
+	}
+	return(seq(minmax[1],minmax[2],100000))
+}
+
+comparesets <- function(filename, datalist, labellist, xlabel, ylabel, chartlabel, formatlabel, legendpos) {
 	print(sprintf("comparesets %s\n",chartlabel))
 	nrcharts <- length(datalist)
 	
@@ -140,21 +164,24 @@ comparesets <- function(datalist, labellist, xlabel, ylabel, chartlabel, legendp
 	}
 	
 	if (!generatePDF) {
-		postscript(sprintf("%sfig%d.eps", figspath, epscount), width=5, height=5, paper="special", family="Times")
+		postscript(sprintf("%s%s.eps", figspath, filename), width=5, height=5, paper="special", family="Times")
 		epscount <<- epscount + 1
 	}
+
+	if (!generateTitles) {
+		margins <- par("mai")
+		margins <- margins - c(0, 0, margins[3] - 0.1, 0)
+		par(mai = margins)
+	}
 	
-	if (xrange[[2]] - xrange[[1]] < 10) {
-		# special case for CPU's
[…]

### 2015-12-13 paper/4-performance.tex
Remove windows from paper, since cygwin still is not c++11 compliant.

--- 2015-12-11 paper/4-performance.tex
@@
 Simulation traces are used to verify that both adevs and dxex models have exactly the same behaviour.
-Benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.
+Benchmarks were performed using Linux, but our simulation tool works equally well on Mac.
 
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{fig/queue_sequential.eps}
+	\includegraphics[width=\columnwidth,height=6.5cm]{fig/queue_sequential.eps}
 	\caption{Queue benchmark results for sequential simulation.}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{fig/interconnect_sequential.eps}
+	\includegraphics[width=\columnwidth,height=6.5cm]{fig/interconnect_sequential.eps}
 	\caption{Interconnect benchmark results for sequential simulation.}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{fig/queue_parallel.eps}
+	\includegraphics[width=\columnwidth,height=6.5cm]{fig/queue_parallel.eps}
 	\caption{Queue benchmark results for parallel simulation of a model of fixed size.}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
+    \includegraphics[width=\columnwidth,height=6.5cm]{fig/phold_remotes.eps}
     \caption{Phold benchmark results for parallel simulation using 6 cores, with varying fraction of remote events.}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{fig/phold_voorlopig.eps}
+    \includegraphics[width=\columnwidth,height=6.5cm]{fig/phold_priority.eps}
     \caption{Phold benchmark results for parallel simulation using 6 cores, with varying amount of high-priority events.}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
+    \includegraphics[width=\columnwidth,height=6.5cm]{fig/interconnect_parallel.eps}
     \caption{Interconnect benchmark results for parallel simulation.}

### 2015-12-13 main/src/model/state.h
Enable compilation on Cygwin64.

@@ -12,6 +12,7 @@
 #include "tools/globallog.h"
 #include "tools/objectfactory.h"
 #include "tools/macros.h"
+#include "tools/stringtools.h"
 #include <assert.h>
 #include <typeinfo>
 
@@ -39,6 +40,14 @@ template<> struct __name<__type> { \
 		return std::to_string(val); \
 	} \
 }
+#else //ifndef __CYGWIN__
+#define STATE_REPR_ARITHMETIC(__type, __name) \
+template<> struct __name<__type> { \
+    static std::string exec(const __type& val) { \
+        return n_tools::toString<__type>(val); \
+    } \
+}
+#endif //ifndef __CYGWIN__
 #define STATE_REPR_ARITHMETIC_GROUP(__name) \
 STATE_REPR_ARITHMETIC(int, __name); \
 STATE_REPR_ARITHMETIC(unsigned, __name); \
@@ -99,7 +108,6 @@ template<> struct ToCell<std::string> {
 #undef STATE_REPR_STRUCT
 #undef STATE_REPR_ARITHMETIC
 #undef STATE_REPR_ARITHMETIC_GROUP
-#endif //ifndef __CYGWIN__
 
 namespace n_model {
 

### 2015-12-13 main/src/performance/devstone/devstone.cpp

@@ -138,7 +138,11 @@ n_network::t_timestamp Processor::lookAhead() const
 template<typename T>
 constexpr T roundTo(T val, T gran)
 {
+#ifdef __CYGWIN__
+	return round(val/gran)*gran;
+#else
 	return std::round(val/gran)*gran;
+#endif
 }
 
 t_counter Processor::getProcTime(size_t event) const

### 2015-12-13 main/src/performance/highInterconnect/hinterconnect.cpp

@@ -45,7 +45,11 @@ Generator::Generator(const std::string& name, std::size_t seed, bool randta):
 template<typename T>
 constexpr T roundTo(T val, T gran)
 {
+#ifdef __CYGWIN__
+	return round(val/gran)*gran;
+#else
 	return std::round(val/gran)*gran;
+#endif
 }
 
 void Generator::adjustCounter(std::size_t seed)

### 2015-12-13 main/src/performance/phold/phold.cpp

@@ -53,7 +53,11 @@ HeavyPHOLDProcessor::~HeavyPHOLDProcessor()
 template<typename T>
 constexpr T roundTo(T val, T gran)
 {
+#ifdef __CYGWIN__
+	return round(val/gran)*gran;
+#else
 	return std::round(val/gran)*gran;
+#endif
 }
 
 EventTime HeavyPHOLDProcessor::getProcTime(EventTime event) const

### 2015-12-13 main/src/performance/phold/phold.h

@@ -119,8 +119,8 @@ public:
                 m_nodes_per_core = std::ceil(m_maxn / (double)coreAmount());
 		assert(m_maxn && "Total amount of models can't be zero.");
 		for(size_t i = 0; i< models.size(); ++i){
-                        std::lldiv_t qr = lldiv(i, m_nodes_per_core);
-			size_t coreid = qr.quot;
+                        auto qr = lldiv(i, m_nodes_per_core);
+						size_t coreid = qr.quot;
                         if(coreid >= coreAmount()){     // overflow into the last core.
                                 coreid = coreAmount()-1;
                         }

### 2015-12-13 main/src/test/tracerstest.cpp

@@ -18,7 +18,7 @@
 #include "test/compare.h"
 #include "tools/macros.h"
 #include "tools/coutredirect.h"
-#include "examples/forestfire/firecell.h"
+//#include "examples/forestfire/firecell.h"
 
 
 using namespace n_tracers;
@@ -601,7 +601,7 @@ SINGLETRACERTEST(TESTFOLDERTRACE, XmlTracer)
 
 SINGLETRACERTEST(TESTFOLDERTRACE, JsonTracer)
 
-TEST(tracing, tracerCellTracer){
+/* TEST(tracing, tracerCellTracer){
 	{
 		CellTracer<n_tracers::FileWriter, 5u, 5u> tracer1;
 		CellTracer<n_tracers::MultiFileWriter, 5u, 5u> tracer2;
@@ -654,7 +654,7 @@ TEST(tracing, tracerCellTracer){
 	EXPECT_EQ(n_misc::filecmp(TESTFOLDERTRACE "CellTracer_out2_1.txt", TESTFOLDERTRACE "CellTracer_out2_1.corr"), 0);
 	EXPECT_EQ(n_misc::filecmp(TESTFOLDERTRACE "CellTracer_out2_2.txt", TESTFOLDERTRACE "CellTracer_out2_2.corr"), 0);
 
-}
+} */
 
 TEST(tracing, messageManagement){
 	{

### 2015-12-13 main/src/tools/globallog.cpp

@@ -7,8 +7,10 @@
 
 #include "tools/globallog.h"
 #ifdef __GNUC__
+#ifndef __CYGWIN__
 #include <execinfo.h>
 #include <unistd.h>
+#endif
 #endif /* __GNUC__ */
 
 #if LOGGING==true
@@ -47,6 +49,7 @@ void signalHandler(int sig){
 		std::terminate();
 	//if gcc, print stacktrace
 #ifdef __GNUC__
+#ifndef __CYGWIN__
 	/* many thanks to tgamblin and Violet Giraffe for this code
 	 * @see http://stackoverflow.com/a/77336
 	 */
@@ -58,6 +61,7 @@ void signalHandler(int sig){
 	// print out all the frames to stderr
 	fprintf(stderr, "Error: signal %d:\n", sig);
 	backtrace_symbols_fd(array, size, STDERR_FILENO);
+#endif
 #endif /* __GNUC__ */
 	restoreHandler();
 }

### 2015-12-13 main/src/tools/misc.h

@@ -31,7 +31,11 @@ constexpr int sgn(T val) {
 template<typename T>
 constexpr int intlog2(T val) {
 	static_assert(std::is_integral<T>::value, "The type must be an integral type.");
+#ifndef __CYGWIN__
 	return int(std::log2(double(val)));
+#else
+	return int(log2(double(val)));
+#endif
 }
 
 #ifdef __GNUC__

### 2015-12-13 main/src/tools/stringtools.h

@@ -47,12 +47,21 @@ bool endswith(const std::string& full, const std::string& part)
 	}
 }
 
+
+template<typename T>
+inline std::string toString(T d){
+        std::stringstream ss;
+        ss << d;
+        return ss.str();
+}
+
 /**
  * @brief Convert unsigned integer to string (Windows friendly)
  * @param i Integer to be converted
  * @return string representation of the integer
  */
-inline std::string toString(std::size_t i)
+template<>
+inline std::string toString<std::size_t>(std::size_t i)
 {
 #ifndef __CYGWIN__
 	return std::to_string(i);
@@ -74,7 +83,8 @@ inline std::string toString(std::size_t i)
  * @param i Integer to be converted
  * @return string representation of the integer
  */
-inline std::string toString(int i)
+template<>
+inline std::string toString<int>(int i)
 {
 #ifndef __CYGWIN__
 	return std::to_string(i);
@@ -89,12 +99,6 @@ inline std::string toString(int i)
 	}
 	return number;
 #endif
-}
-
-inline std::string toString(double d){
-        std::stringstream ss;
-        ss << d;
-        return ss.str();
 }
 
 

### 2015-12-14 paper/4-performance.tex
Enable compilation on Cygwin64.
Proofread Performance introduction

@@
-In this section, we evaluate performance of dxex using its different synchronization protocols.
-We show that the inclusion of multiple synchronization protocols does not decrease performance significantly.
+In this section, we evaluate performance of dxex using different synchronization protocols.
+We show that the inclusion of multiple synchronization protocols does not noticeably decrease performance.
 To this end, we compare to adevs, currently one of the most efficient simulation kernels~\cite{DEVSSurvey}.
@@
 The same compilation flags were used for both adevs and dxex benchmarks (``\texttt{-O3 -flto}'').
-To guarantee comparable results, no I/O was performed during the benchmarking phase such that computation was CPU-bound.
-Simulation traces are used to verify that both adevs and dxex models have exactly the same behaviour.
-Benchmarks were performed using Linux, but our simulation tool works equally well on Mac.
+To guarantee comparable results, no I/O was performed during benchmarks.
+Before benchmarking, simulation traces were used to verify that adevs and dxex return exactly the same simulation results.
+Benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.
 

### 2015-12-14 paper/4-performance.tex
Proofread section 4.1

@@
 \subsection{Benchmarks}
-We use a selection of benchmarks, based on those found in the literature.
-Three different types of benchmark are defined:
+We use three different benchmarks, which cover different aspects of the simulation kernel:
 \begin{enumerate}
-    \item The \textit{Queue} model, based on the HI model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
-          A single generator will push events into the queue, which get processed by the processors after a fixed or random delay.
+    \item The \textit{Queue} model, based on the \textit{HI} model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
+          A single generator pushes events into the queue, which are processed by the processors after a random delay.
           It takes two parameters: width and depth, which determine the width and depth of the hierarchy.
@@
     \item The \textit{PHOLD} model, presented by~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
-          Couplings are made to directly connect each atomic model to every other atomic model.
-          After a random delay, atomic models will send out an event to a randomly selected output port.
-          Output port selection happens in two phases: first it is decided whether the event should be sent to an atomic model inside or outside of this coupled model.
+          Each atomic model is directly connected to every other atomic model.
+          After a random delay, an atomic model sends out an event to a randomly selected output port.
+          Output port selection happens in two phases: first it is decided whether the event should be sent to an atomic model at the same node.
           Afterwards, a uniform selection is made between the remaining ports.
-          The model takes one parameter: the percentage of remote events, which determines the fraction of messages routed to other coupled models.
+          The model takes one parameter: the percentage of remote events, which determines the fraction of messages routed to other nodes.
           This benchmark shows how the simulation kernel behaves in the presence of many local or remote events.
@@
 
-    \item The \textit{HighInterconnect} model, a merge of the HI model of DEVStone~\cite{DEVStone} and PHOLD~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly one output port.
-          All models are still connected to one another, but all over the same port: every model will receive a generated event.
+    \item The \textit{HighInterconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
+          Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event.
           The model takes one parameter: the number of models.
-          This benchmark investigates the complexity of the event routing algorithm, and how the simulation kernel handles many simultaneous events.
+          This benchmark investigates the complexity of event routing, and how the simulation kernel handles many simultaneous events.
           An example for four models is shown in Figure~\ref{fig:interconnect_model}.
 \end{enumerate}
-
-We opted to deviate from the DEVStone benchmark, as DEVStone tends towards unrealistic models since most transitions occur simultaneously.
-In all of our benchmark models, simultaneous internal transitions are avoided as much as possible through the use of a random time advance.
-DEVStone also uses an artificial load function, which easily skews the result: the simulation kernel overhead is dominated by the artificial load.
 

### 2015-12-14 paper/4-performance.tex
Proofread section 4.2

@@
 \subsection{Sequential Simulation Execution Time}
-Despite our core contribution being mainly in the parallel simulation, we still value a comprehensive comparison of sequential simulation results.
-First, and foremost, as parallel simulation results are tightly linked to the sequential simulation results: parallel simulation merely adds a synchronization layer over multiple, essentially sequential, simulation kernels.
+Despite our core contribution being on parallel simulation, we still value a comparison of sequential simulation results.
+First, and foremost, parallel simulation results are tightly linked to sequential simulation results.
+Parallel simulation is achieved through synchronization of multiple sequential simulation kernels.
 Second, parallel simulation results are validated through the use of adevs.
-To provide a more comprehensive comparison to adevs in the parallel simulation benchmarks, sequential simulation results need to be compared.
+To provide a well-founded comparison to adevs in the parallel simulation benchmarks, sequential simulation results also need to be compared.
+
 Only the Queue and HighInterconnect models are relevant for sequential simulation, so we will not touch upon the PHold model yet.
@@
 \subsubsection{Queue}
-In the Queue model, we increase both the width and depth simultaneously, causing a quadratic growth in the number of atomic models.
+In the Queue model, we increase both width and depth simultaneously.
+For example, the $400$ models configuration is obtained with a width and depth of $20$.
 As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
-Through careful analysis of profiling results, we determined that adevs spends much time while handling simulation messages, while this is mostly avoided in dxex's differently designed simulation algorithm.
+Through profiling, we determined that adevs spends much time on the handling of simulation messages, while this is avoided in dxex's simulation algorithm.
 Both simulation tools have similar complexities, though dxex is much faster thanks to its more efficient simulation control algorithms.
@@
 \subsubsection{HighInterconnect}
-In the HighInterconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings.
-As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs now outperforms dxex by a fair margin.
-Analysis showed that this slowdown is caused by the high amount of exchanged events: event creation is found to be much slower in dxex than it is in adevs, despite the use of memory pools in dxex.
+In the HighInterconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
+As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.
+Analysis showed that this is caused by the high amount of exchanged events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
 

### 2015-12-14 paper/4-performance.tex
Proofread section 4.3

@@
 \subsection{Parallel Simulation Execution Time}
-We now perform an analysis of parallel simulation execution times of our previously defined benchmarks.
-For dxex, we mention results for both conservative and optimistic synchronization.
+Next we analyse parallel simulation of our previously defined benchmarks.
+For dxex, we mention results of both conservative and optimistic synchronization.
 Since adevs supports only conservative synchronization, we don't mention optimistic synchronization results there.
-All experiments were performed using up to six simulation nodes, and executed on a hexa-core machine.
+All experiments were performed using up to six simulation nodes, executed on a hexa-core machine.
 
@@
 This minimizes the number of inter-node messages.
-As the model is a queue, however, the last models will only activate later in the simulation.
-Since these are allocated to separate nodes, some nodes will remain idle until simulation has progressed sufficiently far.
+As the model is a queue, however, models further in the chain only activate later in the simulation.
+Since these are allocated to separate nodes, some nodes remain idle until simulation has progressed sufficiently far.
 
-Similar to the sequential benchmarks, Figure~\ref{fig:queue_benchmark_parallel} shows that dxex again outperforms adevs, but now also in terms of speedup.
+Similar to the sequential benchmarks, Figure~\ref{fig:queue_benchmark_parallel} shows that dxex outperforms adevs, but now in terms of speedup.
 Results indicate that our implementation of conservative synchronization achieves much higher speedups than adevs.
 This simulation is the ideal case for our conservative implementation, approaching near linear speedup.
-In this case, conservative synchronization seems to be better than optimistic synchronization, at the cost of providing the lookahead.
+%TODO: Why is this the ideal case?
+Conservative synchronization also seems to be better than optimistic synchronization in this case, at the cost of providing the lookahead.
 
@@
 In the Phold model, we first investigate the influence of the fraction of remote events on the speedup.
-When remote events are rare, optimistic synchronization will rarely have to perform rollbacks, thus increasing performance.
-With more common remote events, however, optimistic synchronization quickly slows down due to the frequent rollbacks.
+When remote events are rare, optimistic synchronization rarely has to roll back, thus increasing performance.
+With more common remote events, however, optimistic synchronization quickly slows down due to frequent rollbacks.
 Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block.
-Even though a single synchronization protocol is always ideal in this case, it shows how different synchronization protocols respond differently to a changing model.
-Adevs is significantly slower during conservative synchronization.
-Analysis of profiling results shows that this was caused by exception handling within the adevs simulation kernel.
+Even though a single synchronization protocol is always ideal in this case, it already shows that different synchronization protocols respond differently to a changing model.
+Adevs is again significantly slower during conservative synchronization.
+Analysis of profiling results shows that exception handling in adevs is the main cause.
 
@@
 We slightly modified the Phold benchmark, to include high-priority events.
-Contrary to normal events, which offer a sufficiently large lookahead, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
-Even though normal events  occur most often, a conservative implementation always blocks until it can make guarantees.
-An optimistic implementation, however, will simply go forward in simulation time and roll back the few times that these high-priority events happen.
+Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
+Even when normal events occur most often, conservative synchronization always blocks until it can make guarantees.
+Optimistic synchronization, however, simply goes forward in simulation time and rolls back when these high-priority events happen.
 This situation closely mimics the case made in the comparison between both synchronization algorithms by~\cite{FujimotoBook}.
 
-Figure~\ref{fig:phold_priority} shows how simulation performance is influenced by the fraction of high-priority events.
+Figure~\ref{fig:phold_priority} shows how simulation performance is influenced by the fraction of these high-priority events.
 If barely any high-priority events occur, conservative synchronization is penalized due to its excessive blocking, which often turned out to be unnecessary.
-When many high-priority events occur, optimistic synchronization is penalized due to its mindless progression of simulation, which frequently needed to be reverted.
-Results show that there is no single perfect synchronization algorithm for this model.
-Depending on model configuration, either synchronization protocol might be better.
-We have shown that our contribution is invaluable for high performance simulation: depending on the observed communication behaviour, modellers can use the most appropriate synchronization protocol.
+When many high-priority events occur, optimistic synchronization is penalized due to its mindless progression of simulation, which frequently needs to be rolled back.
+Results show that there is no single perfect synchronization algorithm for this model: depending on configuration, either synchronization protocol might be better.
+We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
 
@@
 In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
-Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}
-When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing amount of reverts in optimistic.
-In this benchmark, all models depend on each other, negating any possible performance gain by executing the simulation in parallel.
+Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
+When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing amount of roll bakcs in optimistic.
+All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
 

### 2015-12-14 paper/4-performance.tex
Proofread section 4.4

@@
 While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation infeasible.
-We therefore also investigate memory usage of different synchronization protocols.
+We therefore also investigate memory usage of different synchronization protocols, and again compare to adevs.
 
@@
 This problem can be mitigated by distribution over multiple machines, which neither dxex or adevs support.
-Comparison of memory usage should thus only be seen in the context of which models are feasible to be simulated by the tool.
 
 \subsubsection{Remarks}
-Both dxex and adevs use \texttt{tcmalloc} as memory allocator.
+Both dxex and adevs use \texttt{tcmalloc} as memory allocator, allowing for thread-local allocation.
 Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
 \texttt{tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
-If memory has been allocated once, it is, at least from a performance point of view, better to keep that memory in the pool indefinitely.
 Due to our motivation for memory usage analysis, we will only measure peak allocation.
@@
 \subsubsection{Results}
-Figure~\ref{fig:memory} shows the memory used by each different benchmark.
+Figure~\ref{fig:memory} shows the memory used by the different benchmarks.
 Results are in megabytes, and show the total memory footprint of the running application (\textit{i.e.}, text, stack, and heap).
@@
 Note the logarithmic scale that was used for this reason.
-Also, results for optimistic synchronization vary heavily depending on thread scheduling by the operating system, as this influences the drift between different nodes.
-Comparing similar approaches though, we notice that dxex and adevs have very similar memory use.
+Optimistic synchronization results vary heavily depending on thread scheduling by the operating system, as this influences the drift between nodes.
+Comparing similar approaches, we notice that dxex and adevs have very similar memory use.
 

### 2015-12-14 paper/5-related-work.tex
Proofread section 5

@@
-Several \textsf{DEVS} simulation tools have already been implemented, which bear some similarity to our simulation tool.
+Several similar \textsf{DEVS} simulation tools have already been implemented, though they differ in key aspects.
 We discuss several dimensions of related work, as we try to compromise between different tools.
@@
 This allowed the kernel to optimize its internal data structures and algorithms for the specific model being executed.
-All changes were completely transparant to the model itself, and were completely optional.
-In this spirit, we now offer users the possibility to chose between different synchronization protocols.
-This allows them to chose for the most appropriate synchronization, depending on the actual model being simulated.
-Contrary to PythonPDEVS, however, dxex doesn't support model migrations~\cite{PythonPDEVS2} or activity hints~\cite{PythonPDEVS_ACTIMS}.
+All changes were completely transparent to the model, and were completely optional.
+In this spirit, we offer users the possibility to chose between different synchronization protocols.
+This allows users to choose the most appropriate synchronization protocol, depending on the model.
+Contrary to PythonPDEVS, however, dxex doesn't support distribution, model migrations~\cite{PythonPDEVS2}, or activity hints~\cite{PythonPDEVS_ACTIMS}.
 
 While PythonPDEVS offers very fast turnaround cycles, due to the use of an interpreted language, simulation performance was easily outdone by compiled simulation kernels.
-In terms of performance, adevs~\cite{adevs} offers faster simulation, at the cost of a significant compilation time.
-The turnaround cycle in adevs is much slower though, specifically because the complete simulation kernel is implemented using templates, and thus in header files.
-This prevents use of the simulation kernel as a shared library.
-A similar compilation cycle is seen in PowerDEVS~\cite{PowerDEVS}.
+In terms of performance, adevs~\cite{adevs} offered much faster simulation, at the cost of a significant compilation time.
+The turnaround cycle in adevs is much slower though, specifically because the complete simulation kernel is implemented using templates in header files.
 As a result, the complete simulation kernel has to be compiled again everytime.
-Dxex compromises, as was the case in other compiled simulation kernels, such as vle~\cite{vle}, by seperating the simulation kernel into a shared library.
+Dxex compromises, as vle~\cite{vle} or PowerDEVS~\cite{PowerDEVS}, by seperating the simulation kernel into a shared library.
 After the initial compilation of the simulation tool, only the model has to be compiled and linked to the library.
-This significantly shortens the turnaround cycle, while still offering good performance
-In terms of performance, it therefore comes close to, and sometimes beats, adevs.
+This significantly shortens the turnaround cycle, while still offering good performance.
+In terms of performance, dxex is shown to be competitive with adevs.
 While a more extensive set of benchmarks is required to make accurate comparisons, initial results are promising.
@@
 For example CD++~\cite{CD++} has both a conservative (CCD++~\cite{CCD++}) and optimistic (PCD++)~\cite{PCD++}) variant.
-Despite the implementation of both, they are different projects entirely, and while they might support a similar set of features, they remain different from each other.
+Despite the implementation of both protocols, they are different projects entirely, and are incompatible with modern compilers.
 Dxex, on the other hand, is a single project, where switching between different synchronization protocols is as simple as switching any other configuration parameter.

### 2015-12-14 paper/6-conclusions.tex
Proofread section 6

@@
-In this paper, we introduced dxex, a new C++11-based \textsf{Parallel DEVS} simulation tool.
-Our main contribution is the implementation of different pluggable synchronization kernels for parallel simulation.
+In this paper, we introduced DEVS-Ex-Machina (``dxex''), a new C++11-based \textsf{Parallel DEVS} simulation tool.
+Our main contribution is the implementation of multiple synchronization protocols for parallel simulation.
 We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
-Dxex allows the user to chose between both conservative and optimistic synchronization, as simply as any other configuration option.
-Notwithstanding this modularity, we have shown that dxex achieves performance similar to adevs, another very efficient \textsf{DEVS} simultion tool.
-Performance is measured both in CPU time, and memory usage.
+Dxex allows the user to choose between either conservative or optimistic synchronization as simple as any other configuration option.
+Notwithstanding this modularity, dxex achieves performance competitive to adevs, another very efficient \textsf{DEVS} simultion tool.
+Performance is measured both in elapsed time, and memory usage.
 
-Future work exists in several directions.
-First, we wish to make optimistic synchronization more tolerant to low-memory situations.
+Future work is possible in several directions.
+First, our implementation of optimistic synchronization should be more tolerant to low-memory situations.
 In its current state, simulation will simply halt with an out-of-memory error.
 Having simulation control, which can throttle the speed of nodes that use up too much memory, has been shown to work in these situations~\cite{FujimotoBook}.
-In our implementation specifically, the out-of-memory problem is aggravated by a slow GVT implementation.
-Algorithms as those presented by~\cite{Fujimoto:1997:CGV:268403.268404} or~\cite{Bauer:2005:SND:1069810.1070159} might help to alleviate this problem.
-Second, the idea of activity can be implemented for our simulation kernels, making it possible to dynamically switch between conservative and optimistic synchronization when changes in the model behaviour are detected.
-Third, activity algorithms, as already implemented by PythonPDEVS, could also be implemented in dxex, to determine how they influence simulation performance.
+Faster GVT implementations, such as those presented by~\cite{Fujimoto:1997:CGV:268403.268404} and~\cite{Bauer:2005:SND:1069810.1070159}, might further help to minimize this problem.
+Second, the idea of activity can be implemented for our simulation kernels, making it possible to dynamically switch between conservative and optimistic synchronization when behavioural changes are detected.
+Third, activity algorithms, as already implemented by PythonPDEVS, can also be implemented in dxex, to determine how they influence simulation performance.

### 2015-12-14 paper/4-performance.tex
Spelling and esthetical changes

@@
     \includegraphics[width=\columnwidth,height=6.5cm]{fig/phold_remotes.eps}
-    \caption{Phold benchmark results for parallel simulation using 6 cores, with varying fraction of remote events.}
+    \caption{Phold benchmark results for parallel simulation using six cores, with varying fraction of remote events.}
 \end{figure}
@@
     \includegraphics[width=\columnwidth,height=6.5cm]{fig/phold_priority.eps}
-    \caption{Phold benchmark results for parallel simulation using 6 cores, with varying amount of high-priority events.}
+    \caption{Phold benchmark results for parallel simulation using six cores, with varying amount of high-priority events.}
     \label{fig:phold_priority}

### 2015-12-14 paper/4-performance.tex
Nicer figure scaling and fit everything on 8 pages

@@
 \begin{figure}
-	\includegraphics[width=\columnwidth,height=6.5cm]{fig/queue_sequential.eps}
+    \center
+	\includegraphics[width=0.8\columnwidth]{fig/queue_sequential.eps}
 	\caption{Queue benchmark results for sequential simulation.}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth,height=6.5cm]{fig/interconnect_sequential.eps}
+    \center
+	\includegraphics[width=0.8\columnwidth]{fig/interconnect_sequential.eps}
 	\caption{Interconnect benchmark results for sequential simulation.}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth,height=6.5cm]{fig/queue_parallel.eps}
+    \center
+	\includegraphics[width=0.8\columnwidth]{fig/queue_parallel.eps}
 	\caption{Queue benchmark results for parallel simulation of a model of fixed size.}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth,height=6.5cm]{fig/phold_remotes.eps}
+    \center
+    \includegraphics[width=0.8\columnwidth]{fig/phold_remotes.eps}
     \caption{Phold benchmark results for parallel simulation using six cores, with varying fraction of remote events.}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth,height=6.5cm]{fig/phold_priority.eps}
+    \center
+    \includegraphics[width=0.8\columnwidth]{fig/phold_priority.eps}
     \caption{Phold benchmark results for parallel simulation using six cores, with varying amount of high-priority events.}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth,height=6.5cm]{fig/interconnect_parallel.eps}
+    \center
+    \includegraphics[width=0.8\columnwidth]{fig/interconnect_parallel.eps}
     \caption{Interconnect benchmark results for parallel simulation.}

### 2015-12-14 paper/4-performance.tex
Updated Queue model
More compact figure for PHOLD

@@
           This benchmark shows how the complexity of the simulation kernel behaves for an increasing amount of atomic models, and an increasingly deep hierarchy.
-          An example for width 2 and depth 3 is shown in Figure~\ref{fig:queue_model}.
+          An example for width and depth 2 is shown in Figure~\ref{fig:queue_model}.
 
@@
     \includegraphics[width=\columnwidth]{fig/queue_model.pdf}
-    \caption{Queue model for depth 3 and width 2.}
+    \caption{Queue model for depth and width 2.}
     \label{fig:queue_model}
-\end{figure}
 
-\begin{figure}
-    \center
-    \includegraphics[width=\columnwidth]{fig/interconnect_model.pdf}
-    \caption{HighInterconnect model for four models.}
+    \vspace{0.5cm}
+
+    \includegraphics[width=0.65\columnwidth]{fig/interconnect_model.pdf}
+    \caption{HighInterconnect model for three models.}
     \label{fig:interconnect_model}
-\end{figure}
 
-\begin{figure}
-    \center
-    \includegraphics[width=0.5\columnwidth]{fig/phold_model.pdf}
-    \caption{PHOLD model for four models, split over two nodes. In parallel simulation, each coupled model is simulated at a different node.}
+    \vspace{0.5cm}
+
+    \includegraphics[width=0.65\columnwidth]{fig/phold_model.pdf}
+    \caption{PHOLD model for three models.}
     \label{fig:PHOLD_model}

### 2015-12-14 paper/1-introduction.tex
Typos corrected in abstract & intro.

--- 2015-12-11 paper/1-introduction.tex
@@
 
-We implemented a model that, depending on a single parameter, changes its ideal synchronization protocols.
-Dxex, then, is used to compare simulation using exactly the same tool, but with a different synchronization protocol.
-As dxes is able to simulate it both ways, users can always opt to use the fastest protocol available.
+We implemented a model that, depending on a single parameter, changes its ideal synchronization protocol.
+Dxex, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol. With dxex users can always opt to use the fastest protocol available.
 To verify that our flexibility does not counter performance, we compare to adevs, currently one of the fastest \textsf{DEVS} simulation tools available~\cite{PythonPDEVS,DEVSSurvey}.

### 2015-12-14 paper/2-background.tex
Reviewed all sections; only changes are typos and choice of words.

@@
 In conservative synchronization, a node progresses independent of all other nodes, up to the point in time where it can guarantee that no causality errors happen.
-When simulation reaches this point, the node blocks until it can guarantee a new time until which no causality errors happen.
+When simulation reaches this point, the node blocks until it can guarantee a new time until which no causality errors occur.
 In practice, this means that all nodes are aware of the current simulation time of all other nodes, and the time it takes an event to propagate (called \textit{lookahead}).
@@
 When this assumption is violated, the node rolls back its simulation time and state to right before the moment when the event has to be processed.
-As simulation time is rolled back to before where the event must be processed, the event can just be processed as if no causality error ever occured.
+As simulation is rolled back to a time prior to the event must be processed, the event can then be processed as if no causality error ever occured.
 

### 2015-12-14 paper/3-features.tex
Spelling and esthetical changes
Reviewed all sections; only changes are typos and choice of words.

@@
 Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
-Dxex is a C++11-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, whiel making some of its own additions.
+Dxex is a C++11-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, while making some of its own additions.
 So while the feature set is not too comparable, the architectural design, core simulation algorithm, and optimizations, are highly similar.
@@
 
-In the remainder of this section, we will elaborate on our prominent new feature: support for multiple synchronization protocols within the same simulation tool, which are offered transparantly to the model.
+In the remainder of this section, we will elaborate on our prominent new feature: support for multiple synchronization protocols within the same simulation tool, which are offered transparently to the model.
 
@@
 As no single synchronization protocol is ideal for all models, a general purpose simulation tool should support multiple protocols.
-Currently, most parallel simulation tools chose only a single synchronization protocol due to the inherent differences between protocols.
+Currently, most parallel simulation tools choose only a single synchronization protocol due to the inherent differences between protocols.
 An uninformed choice on which one to implement is insufficient, as performance will likely be bad.
@@
 \subsection{Transparency}
-We define simulation kernel transparancy as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
+We define simulation kernel transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
 User should thus only provide one model, implemented in C++11, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.

### 2015-12-14 paper/4-performance.tex
Reviewed all sections; only changes are typos and choice of words.

@@
+\newcommand{\plotfraction}{0.88}
+
 In this section, we evaluate performance of dxex using different synchronization protocols.
@@
 
+\newcommand{\modelfraction}{0.65}
+\newcommand{\interfigurespace}{0.8cm}
+
 \begin{figure}
@@
 
-    \vspace{0.5cm}
+    \vspace{\interfigurespace}
 
-    \includegraphics[width=0.65\columnwidth]{fig/interconnect_model.pdf}
+    \includegraphics[width=\modelfraction\columnwidth]{fig/interconnect_model.pdf}
     \caption{HighInterconnect model for three models.}
@@
 
-    \vspace{0.5cm}
+    \vspace{\interfigurespace}
 
-    \includegraphics[width=0.65\columnwidth]{fig/phold_model.pdf}
+    \includegraphics[width=\modelfraction\columnwidth]{fig/phold_model.pdf}
     \caption{PHOLD model for three models.}
@@
     \center
-	\includegraphics[width=0.8\columnwidth]{fig/queue_sequential.eps}
+	\includegraphics[width=\plotfraction\columnwidth]{fig/queue_sequential.eps}
 	\caption{Queue benchmark results for sequential simulation.}
@@
     \center
-	\includegraphics[width=0.8\columnwidth]{fig/interconnect_sequential.eps}
+	\includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_sequential.eps}
 	\caption{Interconnect benchmark results for sequential simulation.}
@@
     \center
-	\includegraphics[width=0.8\columnwidth]{fig/queue_parallel.eps}
+	\includegraphics[width=\plotfraction\columnwidth]{fig/queue_parallel.eps}
 	\caption{Queue benchmark results for parallel simulation of a model of fixed size.}
@@
     \center
-    \includegraphics[width=0.8\columnwidth]{fig/phold_remotes.eps}
+    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_remotes.eps}
     \caption{Phold benchmark results for parallel simulation using six cores, with varying fraction of remote events.}
@@
     \center
-    \includegraphics[width=0.8\columnwidth]{fig/phold_priority.eps}
+    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_priority.eps}
     \caption{Phold benchmark results for parallel simulation using six cores, with varying amount of high-priority events.}
@@
 
+\vspace{1mm}
 \subsubsection{Interconnect}
@@
 Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
-When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing amount of roll bakcs in optimistic.
+When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing number of of rollbacks in optimistic simulation.
 All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
@@
     \center
-    \includegraphics[width=0.8\columnwidth]{fig/interconnect_parallel.eps}
+    \includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_parallel.eps}
     \caption{Interconnect benchmark results for parallel simulation.}
@@
 Apart from simulation execution time, memory usage during simulation is also of great importance.
-While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation infeasible.
+While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation unfeasible.
 We therefore also investigate memory usage of different synchronization protocols, and again compare to adevs.

### 2015-12-14 paper/5-related-work.tex
Spelling and esthetical changes
Reviewed all sections; only changes are typos and choice of words.

@@
 In terms of code design and philosophy, dxex is most related to PythonPDEVS~\cite{PythonPDEVS}.
-Performance of PythonPDEVS was still decent, through the use of simulation and activity hints from the modeller.
+Performance of PythonPDEVS was still decent, through the use of simulation and activity hints from the modeler.
 This allowed the kernel to optimize its internal data structures and algorithms for the specific model being executed.
 All changes were completely transparent to the model, and were completely optional.
-In this spirit, we offer users the possibility to chose between different synchronization protocols.
+In this spirit, we offer users the possibility to choose between different synchronization protocols.
 This allows users to choose the most appropriate synchronization protocol, depending on the model.
@@
 Dxex, on the other hand, is a single project, where switching between different synchronization protocols is as simple as switching any other configuration parameter.
-CD++, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourself to parallel simulation.
+CD++, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourselves to parallel simulation.
 By limiting our approach to parallel simulation, we are able to achieve higher speedups through the use of shared memory communication.

### 2015-12-14 paper/6-conclusions.tex
Reviewed all sections; only changes are typos and choice of words.

@@
 Dxex allows the user to choose between either conservative or optimistic synchronization as simple as any other configuration option.
-Notwithstanding this modularity, dxex achieves performance competitive to adevs, another very efficient \textsf{DEVS} simultion tool.
+Notwithstanding this modularity, dxex achieves performance competitive to adevs, another very efficient \textsf{DEVS} simulation tool.
 Performance is measured both in elapsed time, and memory usage.
@@
 Future work is possible in several directions.
-First, our implementation of optimistic synchronization should be more tolerant to low-memory situations.
+Firstly, our implementation of optimistic synchronization should be more tolerant to low-memory situations.
 In its current state, simulation will simply halt with an out-of-memory error.
@@
 Faster GVT implementations, such as those presented by~\cite{Fujimoto:1997:CGV:268403.268404} and~\cite{Bauer:2005:SND:1069810.1070159}, might further help to minimize this problem.
-Second, the idea of activity can be implemented for our simulation kernels, making it possible to dynamically switch between conservative and optimistic synchronization when behavioural changes are detected.
-Third, activity algorithms, as already implemented by PythonPDEVS, can also be implemented in dxex, to determine how they influence simulation performance.
+Secondly, the idea of activity can be implemented for our simulation kernels, making it possible to dynamically switch between conservative and optimistic synchronization when behavioural changes are detected.
+Thirdly, activity algorithms, as already implemented by PythonPDEVS, can also be implemented in dxex, to determine how they influence simulation performance.

### 2015-12-14 Rscript/devs-performance.R
EPS files added + R script changes + Dataset to obtain figures added

@@ -1,8 +1,6 @@
-# @author Tim Tuijn
 # global variables
 generatePDF <<- FALSE	# if TRUE, generate PDF else: generated EPS files...
 epscount <<- 1
-generateTitles <<- FALSE
 rootpath <<- paste(getwd(), '/', sep = '')
 # rootpath <<- 'putyourcustomrootpathinhere'
 datapath <<- paste(rootpath, 'data/', sep = '')
@@ -124,10 +122,31 @@ newrange <- function(range1, range2) {
 }
 
 myrainbow <- function(n) {
-	if (n <= 6)
+	if (n == 2)
+		return(c("magenta", "dodgerblue"))
+	else if (n <= 6)
 		return(c("red", "green", "blue", "magenta", "dodgerblue", "black" ))
 	else
 		return(rainbow(n))
+}
+
+myplotchars <- function(n) {
+	if (n == 2)
+		return(c(18, 22))
+	else if (n <= 8)
+		return(c(15, 17, 19, 18, 22, 24, 21, 5))
+	else
+		return(seq(15, 15+n, 1))
+}
+
+mylinetype <- function(n) {
+	return(rep(1,n)) 
+	if (FALSE) {
+		if (n == 2)
+			return(c(4,5))
+		else
+			return(c(1:n))
+	}
 }
 
 formatlabels <- function(xvalues, formattype) {
@@ -179,9 +198,9 @@ comparesets <- function(filename, datalist, labellist, xlabel, ylabel, chartlabe
 	axis(side = 1, at = newlabels , labels = formatlabels(newlabels, formatlabel))
 	
 	colors <- myrainbow(nrcharts)
-	linetype <- c(1:nrcharts)
-	plotchar <- seq(15, 15+nrcharts, 1)
-	plotchar <- c(15, 17, 19, 18, 22, 24, 21, 5)
+	linetype <- mylinetype(nrcharts)
+	plotchar <- myplotchars(nrcharts)
+	
 	for (i in 1:nrcharts) {
 		lines(datalist[[i]]$x, datalist[[i]]$cpu,
 			type="b",
@@ -297,6 +316,15 @@ speedup2 <- function(multicoredata, singlecoredata, field) {
 	return(multicoredata)
 }
 
+addModels <- function(data) {
+	nrrows <- dim(data)[1]
+	data$models <- 0
+	for (i in 1:nrrows) {
+		data$models[i] <- data$width[i] * data$depth[i]
[…]

### 2015-12-15 paper/5-related-work.tex
Spell check

@@
 The turnaround cycle in adevs is much slower though, specifically because the complete simulation kernel is implemented using templates in header files.
-As a result, the complete simulation kernel has to be compiled again everytime.
-Dxex compromises, as vle~\cite{vle} or PowerDEVS~\cite{PowerDEVS}, by seperating the simulation kernel into a shared library.
+As a result, the complete simulation kernel has to be compiled again every time.
+Dxex compromises, as vle~\cite{vle} or PowerDEVS~\cite{PowerDEVS}, by separating the simulation kernel into a shared library.
 After the initial compilation of the simulation tool, only the model has to be compiled and linked to the library.

### 2016-02-13 paper/1-introduction.tex
Apply reviewers comments.

@@
 The applicability of parallel simulation capabilities of current tools is therefore limited.
+% Paragraph moved to beneath intro dxex, else we can't refer to our tool
+This paper introduces DEVS-Ex-Machina (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
+The selected synchronization protocol is transparent to the simulated model.
+Users should merely determine, at the start of simulation, which protocol they wish to use.
+Our tool is based on PythonPDEVS, but implemented in C++11, to increase both performance and portability across different platforms. Unlike PythonPDEVS dxex only supports multicore parallelism.
 
-Users who simulate a wide variety of models, with different ideal synchronization protocols, are therefore out of luck:
-either they accept lower performance for some models, or they use different simulation kernels.
-Neither is acceptable for complex models: performance can decrease to unacceptable levels, and simulation kernel APIs can diverge significantly.
-
-This paper introduces DEVS-Ex-Machina (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
-The selected synchronization protocol is transparant to the simulated model.
-Users should merely determine, at the start of simulation, which protocol they wish to use.
-Our tool is based on PythonPDEVS, but implemented in C++11, to increase both performance and portability across different platforms.
+Users who simulate a wide variety of models, with different ideal synchronization protocols, are therefore in luck:
+with our tool the model implementation can be reused with a different protocol to test which offers better performance.
 

### 2016-02-13 paper/3-features.tex
Apply reviewers comments.

@@
 This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
-In our case, a user can pass the simulation kernel the ``hint'' as to which synchronization protocol must be used for this model.
+In our case, a user can pass the simulation kernel sets the synchronization protocol to be used for this model.
 Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language.
@@
 The more attractive option is for the simulation tool to provide a default lookahead function, such that simulation can run anyway, but maybe not at peak performance.
-Depending on the model, simulation performance might still be faster than sequential simulation.
+%Depending on the model, simulation performance might still be faster than sequential simulation.
 
-Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default can be used otherwise.
+Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default of $epsilon$ can be used otherwise.

### 2016-02-13 paper/4-performance.tex
EPS files added + R script changes + Dataset to obtain figures added
Apply reviewers comments.

@@
 Benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.
+The exact parameters for each benchmark can be found in the repository, as well as the data used in this paper.
 
@@
 As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
-Through profiling, we determined that adevs spends much time on the handling of simulation messages, while this is avoided in dxex's simulation algorithm.
+Through profiling, we determined that adevs spends much time on the routing of simulation messages through the hierarchy, while dxex avoids this by flattening the model structure before the simulation starts.
 Both simulation tools have similar complexities, though dxex is much faster thanks to its more efficient simulation control algorithms.
@@
 Analysis showed that this is caused by the high amount of exchanged events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
+For the speedup plot the amount of models is kept constant at 8.
 
@@
 Results indicate that our implementation of conservative synchronization achieves much higher speedups than adevs.
-This simulation is the ideal case for our conservative implementation, approaching near linear speedup.
-%TODO: Why is this the ideal case?
+This simulation is the ideal case for our conservative implementation, approaching near linear speedup, since there are no dependency cycles between models.
 Conservative synchronization also seems to be better than optimistic synchronization in this case, at the cost of providing the lookahead.
@@
 Analysis of profiling results shows that exception handling in adevs is the main cause. 
+% Addin ? To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.
 
@@
     \center
-    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_remotes.eps}
-    \caption{Phold benchmark results for parallel simulation using six cores, with varying fraction of remote events.}
+    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_speedup_remotes.eps}
+    \caption{Phold benchmark results for parallel simulation using six cores, 4 atomics per node, with varying fraction of remote events.}
 \end{figure}
@@
     \center
-    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_priority.eps}
+    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_speedup_priority.eps}
     \caption{Phold benchmark results for parallel simulation using six cores, with varying amount of high-priority events.}

### 2016-02-13 paper/5-related-work.tex
Apply reviewers comments.

@@
 
-Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though mostly in an ad-hoc way.
+Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though none have done it in a strictly modular way that allows straightforward protocol switching for a single given model.
 For example CD++~\cite{CD++} has both a conservative (CCD++~\cite{CCD++}) and optimistic (PCD++)~\cite{PCD++}) variant.

### 2016-02-13 membench.sh
* Add memory benchmark script * Update README

@@ -1,0 +1,70 @@
+#!/bin/sh
+
+# Small script to reproduce memory tests.
+# NOTE: valgrind effectively cripples the runtime of any benchmark, so be prepared to wait a long time for them to complete. 
+# Grep oneliner originates from https://stackoverflow.com/questions/774556/peak-memory-usage-of-a-linux-unix-process
+
+# (system) Commands needed
+VALGRIND="valgrind"
+GREP="grep"
+SED="sed"
+TAIL="tail"
+SORT="sort"
+
+# Let's do a quick check if they can be found, and try to keep it POSIX.
+declare -a cmdar=($VALGRIND $GREP $SED $TAIL $SORT)
+for c in "${cmdar[@]}"
+do
+    echo -n "$c exists ? "
+    EXISTS=`command -v "$c"`
+    if [ -z "$EXISTS" ]
+    then
+        echo \n"Script requires $c, not found"
+        exit 1;
+    fi
+    echo "Y";
+done
+
+# Are we running in the right directory ?
+BENCHMARKDIR="build/Benchmark"
+if [ ! -d "build/Benchmark" ]
+then
+echo "Not running in top directory project, can't find $BENCHMARKDIR";
+exit -1;
+fi
+
+MASSIFFILE="current.out"
+RESULTSFILE="results.txt"
+ENDTIME="-t 500000"
+
+echo "!!! WARNING : Running these benchmarks can ~very~ easily exhaust all physical memory. Run this on a machine with _at least_ 8GB RAM. !!!"
+
+echo "Format : CMD \n x Bytes max heap\n" >> $RESULTSFILE
+
+declare -a arr=("build/Benchmark/adevs_devstone -w 40 -d 40 $ENDTIME -c 2 cpdevs" 
+                "build/Benchmark/adevs_devstone -w 40 -d 40 $ENDTIME" 
+                "build/Benchmark/dxexmachina_devstone -w 40 -d 40 $ENDTIME" 
+                "build/Benchmark/dxexmachina_devstone -w 40 -d 40 $ENDTIME -c 2 cpdevs" 
+                "build/Benchmark/dxexmachina_devstone -w 40 -d 40 $ENDTIME -c 2 opdevs" 
+                "build/Benchmark/dxexmachina_phold -n 4 -s 2 -r 10 -p 0.1 $ENDTIME -c 4 opdevs" 
[…]

### 2016-02-13 paper/3-features.tex
Add begin/end lookahead mention.

@@
 This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
-In our case, a user can pass the simulation kernel sets the synchronization protocol to be used for this model.
+In our case, a user can pass the simulation kernel the synchronization protocol to be used for this model.
 Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language.
@@
 The more attractive option is for the simulation tool to provide a default lookahead function, such that simulation can run anyway, but maybe not at peak performance.
-%Depending on the model, simulation performance might still be faster than sequential simulation.
+Depending on the model, simulation performance might still be faster than sequential simulation.
 

### 2016-02-15 paper/1-introduction.tex
Made some small changes to fit everything nicely on 8 pages

@@
 The applicability of parallel simulation capabilities of current tools is therefore limited.
-% Paragraph moved to beneath intro dxex, else we can't refer to our tool
-This paper introduces DEVS-Ex-Machina (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
-The selected synchronization protocol is transparent to the simulated model.
-Users should merely determine, at the start of simulation, which protocol they wish to use.
-Our tool is based on PythonPDEVS, but implemented in C++11, to increase both performance and portability across different platforms. Unlike PythonPDEVS dxex only supports multicore parallelism.
 
-Users who simulate a wide variety of models, with different ideal synchronization protocols, are therefore in luck:
-with our tool the model implementation can be reused with a different protocol to test which offers better performance.
+This paper introduces DEVS-Ex-Machina\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
+The selected synchronization protocol is transparent to the simulated model: users should merely determine, which protocol they wish to use.
+Users who simulate a wide variety of models, with different ideal synchronization protocols, can simply run the same model with different synchronization protocols.
+
+Our tool is based on PythonPDEVS, but implemented in C++11 for increased performance.
+Unlike PythonPDEVS dxex only supports multicore parallelism.
 
@@
 
-Dxex can be found online at \url{https://bitbucket.org/bcardoen/devs-ex-machina}.
-
 The remainder of this paper is organized as follows:
 Section~\ref{sec:2-background} introduces the necessary background on synchronization protocols.
-Section~\ref{sec:3-features} elaborates on some of our features, and our design that enables our flexibility.
-In Section~\ref{sec:4-performance}, we evaluate performance of our tool by comparing its different synchronization protocols, as well as a comparison to adevs.
+Section~\ref{sec:3-features} elaborates on our design that enables this flexibility.
+In Section~\ref{sec:4-performance}, we evaluate performance of our tool by comparing its different synchronization protocols, and by comparing to adevs.
 Related work is discussed in Section~\ref{sec:5-related-work}.

### 2016-02-15 paper/2-background.tex
Spell check
Made some small changes to fit everything nicely on 8 pages

@@
-This section briefly introduces the two synchronization protocols used by dxex: conservative and optimistic synchronization.
+This section briefly introduces the synchronization protocols used by dxex: conservative and optimistic synchronization.
 
@@
 When this assumption is violated, the node rolls back its simulation time and state to right before the moment when the event has to be processed.
-As simulation is rolled back to a time prior to the event must be processed, the event can then be processed as if no causality error ever occured.
+As simulation is rolled back to a time prior to the event must be processed, the event can then be processed as if no causality error ever occurred.
 

### 2016-02-15 paper/4-performance.tex
Add begin/end lookahead mention.
Made some small changes to fit everything nicely on 8 pages

@@
 
-In this section, we evaluate performance of dxex using different synchronization protocols.
-We show that the inclusion of multiple synchronization protocols does not noticeably decrease performance.
-To this end, we compare to adevs, currently one of the most efficient simulation kernels~\cite{DEVSSurvey}.
+In this section, we evaluate the performance of different synchronization protocols in dxex.
+We also compare to adevs, currently one of the most efficient simulation kernels~\cite{DEVSSurvey}, to show that our modularity does not impede performance.
 CPU time and memory usage is compared for both sequential and parallel simulation.
@@
 \newcommand{\modelfraction}{0.65}
-\newcommand{\interfigurespace}{0.8cm}
 
@@
 
-    \vspace{\interfigurespace}
-
     \includegraphics[width=\modelfraction\columnwidth]{fig/interconnect_model.pdf}
@@
     \label{fig:interconnect_model}
-
-    \vspace{\interfigurespace}
 
@@
 As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
-Through profiling, we determined that adevs spends much time on the routing of simulation messages through the hierarchy, while dxex avoids this by flattening the model structure before the simulation starts.
+Through profiling, we found that adevs spends much time routing the simulation messages (\textit{e.g.}, \textit{*}, \textit{@}) through the hierarchy, whereas dxex avoids this throuhg its flattened model structure.
 Both simulation tools have similar complexities, though dxex is much faster thanks to its more efficient simulation control algorithms.
@@
 Analysis showed that this is caused by the high amount of exchanged events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
-For the speedup plot the amount of models is kept constant at 8.
 
@@
 Results indicate that our implementation of conservative synchronization achieves much higher speedups than adevs.
-This simulation is the ideal case for our conservative implementation, approaching near linear speedup, since there are no dependency cycles between models.
+Approaching near linear speedup, this simulation is the ideal case for our conservative implementation, as there are no dependency cycles between models.
 Conservative synchronization also seems to be better than optimistic synchronization in this case, at the cost of providing the lookahead.
@@
 Analysis of profiling results shows that exception handling in adevs is the main cause. 
-% Addin ? To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.
+To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.
 
@@
 
-\vspace{1mm}
 \subsubsection{Interconnect}
 In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
+The number of models is now kept constant at eight.
 Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
@@
 Profiling is done using Valgrind's massif tool~\cite{Nethercote:2007:VFH:1273442.1250746}.
+\vspace{1cm}
 

### 2016-02-15 paper/5-related-work.tex
Made "distribution" more clear in related work; references look more balanced now too

@@
 This allows users to choose the most appropriate synchronization protocol, depending on the model.
-Contrary to PythonPDEVS, however, dxex doesn't support distribution, model migrations~\cite{PythonPDEVS2}, or activity hints~\cite{PythonPDEVS_ACTIMS}.
+Contrary to PythonPDEVS, however, dxex doesn't support distributed simulation, model migrations~\cite{PythonPDEVS2}, or activity hints~\cite{PythonPDEVS_ACTIMS}.
 

### 2016-02-15 paper/4-performance.tex
Made plots a tiny bit bigger (0.88 --> 0.9)
Fix typo.

@@
-\newcommand{\plotfraction}{0.88}
+\newcommand{\plotfraction}{0.9}
 
@@
 As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
-Through profiling, we found that adevs spends much time routing the simulation messages (\textit{e.g.}, \textit{*}, \textit{@}) through the hierarchy, whereas dxex avoids this throuhg its flattened model structure.
+Through profiling, we found that adevs spends much time routing the simulation messages (\textit{e.g.}, \textit{*}, \textit{@}) through the hierarchy, whereas dxex avoids this through its flattened model structure.
 Both simulation tools have similar complexities, though dxex is much faster thanks to its more efficient simulation control algorithms.
@@
 Profiling is done using Valgrind's massif tool~\cite{Nethercote:2007:VFH:1273442.1250746}.
-\vspace{1cm}
 

### 2016-02-15 paper/4-performance.tex
Added reference to the abstract simulator to explain messages

@@
 
-Only the Queue and HighInterconnect models are relevant for sequential simulation, so we will not touch upon the PHold model yet.
+Only the Queue and HighInterconnect models are relevant for sequential simulation, so we do not use the PHold model yet.
 
@@
 As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
-Through profiling, we found that adevs spends much time routing the simulation messages (\textit{e.g.}, \textit{*}, \textit{@}) through the hierarchy, whereas dxex avoids this through its flattened model structure.
+Through profiling, we found that adevs spends much time routing the simulation messages (\textit{e.g.}, \textit{*} and \textit{@} messages, defined by the abstract simulator~\cite{AbstractSimulator}) through the hierarchy, whereas dxex avoids this through its flattened model structure.
 Both simulation tools have similar complexities, though dxex is much faster thanks to its more efficient simulation control algorithms.

### 2016-02-18 paper/4-performance.tex
Changed the SCS style to make captions 9pt instead of 8pt; slight modifications to keep everything on 8 pages; removed keyword C++11

@@
-\newcommand{\plotfraction}{0.9}
+\newcommand{\plotfraction}{0.85}
 
@@
 
-    \item The \textit{HighInterconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
+    \item The \textit{Interconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
           Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event.
@@
 
+    \vspace{0.2cm}
+
     \includegraphics[width=\modelfraction\columnwidth]{fig/interconnect_model.pdf}
-    \caption{HighInterconnect model for three models.}
+    \caption{Interconnect model for three models.}
     \label{fig:interconnect_model}
+
+    \vspace{0.2cm}
 
@@
 
-Only the Queue and HighInterconnect models are relevant for sequential simulation, so we do not use the PHold model yet.
+Only the Queue and Interconnect models are relevant for sequential simulation, so we do not use the PHold model yet.
 
@@
 
-\subsubsection{HighInterconnect}
-In the HighInterconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
+\subsubsection{Interconnect}
+In the Interconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
 As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.

### 2016-02-18 paper/3-features.tex
Changed to 4 authors, with Kurt, Hans, and Jan mentioned as "additional authors"... Original code is still there, so change back if it should get permitted in the future!

@@
 This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
-In our case, a user can pass the simulation kernel the synchronization protocol to be used for this model.
+In our case, a user can pass the simulation kernel the synchronization protocol to use for this model.
 Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language.
@@
 Finally, sequential simulation is still required for models where parallelism is bad, where all protocols actually slow down simulation.
+
+\vspace{0.3cm}
 
@@
 Especially since defining the lookahead is often non-trivial and dependent on intricate model details.
-The more attractive option is for the simulation tool to provide a default lookahead function, such that simulation can run anyway, but maybe not at peak performance.
+The more attractive option is for the simulation tool to provide a default lookahead function, such that simulation can run anyway, but likely not at peak performance.
 Depending on the model, simulation performance might still be faster than sequential simulation.
 
-Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default of $epsilon$ can be used otherwise.
+Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.

### 2016-06-23 publications/Journal/models.tex
Add journal paper draft.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

--- 2016-02-18 paper/4-performance.tex
@@
-\newcommand{\plotfraction}{0.85}
+% SSection Models
+% Design (what)
+% Reasoning (why)
+% Implications on parallelism
+% Allocation
+% Figures of each each (design, allocation)
 
@@
 \subsection{Benchmarks}
-We use three different benchmarks, which cover different aspects of the simulation kernel:
+We use four different benchmarks, which cover different aspects of the simulation kernel:
 \begin{enumerate}
 	\item The \textit{Queue} model, based on the \textit{HI} model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
-          A single generator pushes events into the queue, which are processed by the processors after a random delay.
+	A single generator pushes events into the queue, which are processed by the processors after a fixed or random delay.
 	It takes two parameters: width and depth, which determine the width and depth of the hierarchy.
@@
 	Afterwards, a uniform selection is made between the remaining ports.
-          The model takes one parameter: the percentage of remote events, which determines the fraction of messages routed to other nodes.
-          This benchmark shows how the simulation kernel behaves in the presence of many local or remote events.
+	The model takes two parameter: the percentage of remote events, which determines the fraction of messages routed to other nodes, and the percentage of priority events.
+Priority events are events generated in a very short time after the previous event.
+	This benchmark shows how the simulation kernel behaves in the presence of many local or remote events and when the average time advance is larger than the lookahead in conservative simulation.
 	An example for four models, split over two nodes, is shown in Figure~\ref{fig:PHOLD_model}.
@@
 	An example for four models is shown in Figure~\ref{fig:interconnect_model}.
+
+	\item The \textit{PHoldTree} model, a hierarchical model based on the semantics of PHOLD~\cite{PHOLD}.
+The model consists of recursively defined trees connected to each other, similar to the structure of an AST.
+All links can be configured to be bidirectional.
+	Unlike the \textit{Queue} model the hierarchical nature is still present in the topology of the atomic models after the directconnect stage.
+This model allows us to investigate parallel speedup in several key dimensions in combination or isolation : model allocation, message density, uncertainty and hierarchy.
+The model has a depth parameter, fanout (childcount), probability of priority message and linkage direction.
+We demonstrate the importance of atomic to kernel allocation, in this model this can be configured as a bread-first versus a depth-first scheme.
+	\textit{PHOLDTree}, while a synthetic benchmark, can be used to model gossiping in social networks, hierarchical resource sharing or load balancing.
+Unlike \textit{Queue} this model does not lends itself easily to parallelism and thus highlights the difficulties model authors face in the real world when employing PDEVS solutions.
+	The number of atomic models in this benchmark is given by 
+	\begin{equation} \label{eq:pholdtreemodelcount}
+	\frac{n^{d+2}-1}{n-1}
+	\end{equation}
+	Whereas the number of links between the models is given by
+	\begin{equation} \label{eq:pholdtreelinkcount}
+	(2n-1)\sum_{i=0}^{d}{n^i}
+	\end{equation}
+	%Expand on how this allows us to profile all parameters of d-stone/phold/iconnect.
 \end{enumerate}
@@
 \newcommand{\modelfraction}{0.65}
+\newcommand{\betweenmodels}{0.8cm}
 
@@
 	\center
-    \includegraphics[width=\columnwidth]{fig/queue_model.pdf}
+	\includegraphics[width=\columnwidth]{fig/queue_model_fixed.pdf}
 	\caption{Queue model for depth and width 2.}
@@
 
-    \vspace{0.2cm}
+	\vspace{\betweenmodels}
 
@@
 	\label{fig:PHOLD_model}
+
+	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree.pdf}
+	\caption{PHOLDTree model for depth 1, width 2.}
+	\label{fig:PHOLDTree_model}
 \end{figure}
-
-\subsection{Sequential Simulation Execution Time}
-Despite our core contribution being on parallel simulation, we still value a comparison of sequential simulation results.
-First, and foremost, parallel simulation results are tightly linked to sequential simulation results.
-Parallel simulation is achieved through synchronization of multiple sequential simulation kernels.
-Second, parallel simulation results are validated through the use of adevs.
-To provide a well-founded comparison to adevs in the parallel simulation benchmarks, sequential simulation results also need to be compared.
-
-Only the Queue and Interconnect models are relevant for sequential simulation, so we do not use the PHold model yet.
-
-\subsubsection{Queue}
-In the Queue model, we increase both width and depth simultaneously.
-For example, the $400$ models configuration is obtained with a width and depth of $20$.
-As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
-Through profiling, we found that adevs spends much time routing the simulation messages (\textit{e.g.}, \textit{*} and \textit{@} messages, defined by the abstract simulator~\cite{AbstractSimulator}) through the hierarchy, whereas dxex avoids this through its flattened model structure.
-Both simulation tools have similar complexities, though dxex is much faster thanks to its more efficient simulation control algorithms.
-
-\begin{figure}
-    \center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/queue_sequential.eps}
-	\caption{Queue benchmark results for sequential simulation.}
-	\label{fig:Queue_benchmark}
-\end{figure}
-
-\subsubsection{Interconnect}
-In the Interconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
-As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.
-Analysis showed that this is caused by the high amount of exchanged events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
-
-\begin{figure}
-    \center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_sequential.eps}
-	\caption{Interconnect benchmark results for sequential simulation.}
-	\label{fig:Interconnect_benchmark}
-\end{figure}
-
-\subsection{Parallel Simulation Execution Time}
-Next we analyse parallel simulation of our previously defined benchmarks.
-For dxex, we mention results of both conservative and optimistic synchronization.
-Since adevs supports only conservative synchronization, we don't mention optimistic synchronization results there.
-All experiments were performed using up to six simulation nodes, executed on a hexa-core machine.
-
-We highlight two main results:
-(1) dxex conservative synchronization is competitive with adevs;
-(2) dxex optimistic synchronization is sometimes more efficient than conservative synchronization.
-This shows that our contribution, offering both conservative and optimistic synchronization, is indeed beneficial for a general-purpose simulation tools.
-
-\subsubsection{Queue}
-In the Queue model, we allocate the chain of models such that each node is responsible for a series of connected models.
-This minimizes the number of inter-node messages.
-As the model is a queue, however, models further in the chain only activate later in the simulation.
-Since these are allocated to separate nodes, some nodes remain idle until simulation has progressed sufficiently far.
-
-Similar to the sequential benchmarks, Figure~\ref{fig:queue_benchmark_parallel} shows that dxex outperforms adevs, but now in terms of speedup.
-Results indicate that our implementation of conservative synchronization achieves much higher speedups than adevs.
-Approaching near linear speedup, this simulation is the ideal case for our conservative implementation, as there are no dependency cycles between models.
-Conservative synchronization also seems to be better than optimistic synchronization in this case, at the cost of providing the lookahead.
-
-\begin{figure}
-    \center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/queue_parallel.eps}
-	\caption{Queue benchmark results for parallel simulation of a model of fixed size.}
-	\label{fig:queue_benchmark_parallel}
-\end{figure}
-
-\subsubsection{PHold}
-In the Phold model, we first investigate the influence of the fraction of remote events on the speedup.
-When remote events are rare, optimistic synchronization rarely has to roll back, thus increasing performance.
-With more common remote events, however, optimistic synchronization quickly slows down due to frequent rollbacks.
-Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block.
-Even though a single synchronization protocol is always ideal in this case, it already shows that different synchronization protocols respond differently to a changing model.
-Adevs is again significantly slower during conservative synchronization.
-Analysis of profiling results shows that exception handling in adevs is the main cause. 
-To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.
-
-\begin{figure}
-    \center
-    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_speedup_remotes.eps}
-    \caption{Phold benchmark results for parallel simulation using six cores, 4 atomics per node, with varying fraction of remote events.}
-\end{figure}
-
-We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization.
-We slightly modified the Phold benchmark, to include high-priority events.
-Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
-Even when normal events occur most often, conservative synchronization always blocks until it can make guarantees.
-Optimistic synchronization, however, simply goes forward in simulation time and rolls back when these high-priority events happen.
-This situation closely mimics the case made in the comparison between both synchronization algorithms by~\cite{FujimotoBook}.
-
-Figure~\ref{fig:phold_priority} shows how simulation performance is influenced by the fraction of these high-priority events.
-If barely any high-priority events occur, conservative synchronization is penalized due to its excessive blocking, which often turned out to be unnecessary.
-When many high-priority events occur, optimistic synchronization is penalized due to its mindless progression of simulation, which frequently needs to be rolled back.
-Results show that there is no single perfect synchronization algorithm for this model: depending on configuration, either synchronization protocol might be better.
-We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
-
-\begin{figure}
-    \center
-    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_speedup_priority.eps}
-    \caption{Phold benchmark results for parallel simulation using six cores, with varying amount of high-priority events.}
-    \label{fig:phold_priority}
-\end{figure}
-
-\subsubsection{Interconnect}
-In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
-The number of models is now kept constant at eight.
-Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
-When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing number of of rollbacks in optimistic simulation.
-All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
-
-\begin{figure}
-    \center
-    \includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_parallel.eps}
-    \caption{Interconnect benchmark results for parallel simulation.}
-    \label{fig:interconnect_benchmark_parallel}
-\end{figure}
-
-\subsection{Memory Usage}
-Apart from simulation execution time, memory usage during simulation is also of great importance.
-While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation unfeasible.
-We therefore also investigate memory usage of different synchronization protocols, and again compare to adevs.
-
-We do not tackle the problem of states that become too large for a single machine to hold.
-This problem can be mitigated by distribution over multiple machines, which neither dxex or adevs support.
-
-\subsubsection{Remarks}
-Both dxex and adevs use \texttt{tcmalloc} as memory allocator, allowing for thread-local allocation.
-Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
-\texttt{tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
-Due to our motivation for memory usage analysis, we will only measure peak allocation.
-Profiling is done using Valgrind's massif tool~\cite{Nethercote:2007:VFH:1273442.1250746}.
-
-\subsubsection{Results}
-Figure~\ref{fig:memory} shows the memory used by the different benchmarks.
-Results are in megabytes, and show the total memory footprint of the running application (\textit{i.e.}, text, stack, and heap).
-
-Unsurprisingly, optimistic synchronization results show very high memory usage due to the saved states.
-Note the logarithmic scale that was used for this reason.
-Optimistic synchronization results vary heavily depending on thread scheduling by the operating system, as this influences the drift between nodes.
-Comparing similar approaches, we notice that dxex and adevs have very similar memory use.
-
-Conservative simulation always uses more memory than sequential simulation, as is to be expected.
-Additional memory is required for the multiple threads, but also to store all events that are processed simultaneously.
-
-For the Phold benchmark, adevs using conservative synchronization took too long using our profiling tool, and was therefore aborted.
-Therefore, no results are shown for adevs.
-
-\begin{figure}
-    \includegraphics[width=\columnwidth]{fig/memory_voorlopig.eps}
-    \caption{Memory usage results.}
-    \label{fig:memory}
-\end{figure}

### 2016-06-23 publications/Journal/parallel-performance.tex
Add journal paper draft.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

\subsection{Parallel Simulation}

% The good
\subsubsection{Queue}
\paragraph*{Allocation}
The Queue model is one single chain of models. Each kernel gets one connected part of this chain. The result is that the kernels themselves also form a chain where events only travel in one direction.

\paragraph*{Strong and Weak Scaling}
Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to dxex sequential for a fixed problem size. As the amount of kernels increases, the optimistic kernel quickly becomes the worst choice. The difference between dxex conservative and adevs becoming smaller. The same effect can be seen for weak scaling in Figure~\ref{fig:Queue_plot_weak}.\\
In Figure \ref{fig:Queue_allocation} the allocation of the Queue model is visualized. This trace allows us to demonstrate the benchmark results and explain why optimistic is not ideal for such a chain topology as present in Queue. Kernel 2 is dependent on events from kernels 0 and 1 but by definition of the optimistic synchronization protocol it will simulate ahead without waiting for events it depends on. Any simulation it performs is immediately void and will have to be reverted as soon as events from kernels 0 and 1 arrive. When kernel 2 reverts, kernel 3 will inevitably have to revert as well since any events it received from kernel 2 are invalidated and marked as such by antimessages. This leads to severe performance degradation, with an increasing probability of cascading reverts as the number of kernels and models per kernels increases.
The speedup of adevs is always in comparison with the runtime of the corresponding dxex sequential benchmark.
\begin{figure}
	\center

	\includegraphics[width=\modelfraction\columnwidth]{fig/queue_fixed_weak_speedup.eps}
	\caption{Queue model weak scaling speedup compared to dxex sequential.}
	\label{fig:Queue_plot_weak}

	\includegraphics[width=\modelfraction\columnwidth]{fig/queue_fixed_strong_speedup.eps}
	\caption{Queue model strong scaling speedup compared to dxex sequential.}
	\label{fig:Queue_plot_strong}

\end{figure}

\begin{figure}
	\center

	\includegraphics[width=\modelfraction\columnwidth, height=8cm, keepaspectratio]{fig/queue_allocation.eps}
	\caption{Queue model (d=2, w=7, t=5000, random timeadvance) allocation and simulation trace across 4 kernels.}
	\label{fig:Queue_allocation}

\end{figure}

% The really bad
\subsubsection{Interconnect}\label{subsec:parallelinterconnect}
In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
The number of models is now kept constant at eight.
Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing number of of rollbacks in optimistic simulation.
All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
In interconnect there is no allocation scheme possible that avoids cyclic dependencies between simulation kernels, as shown in the trace \ref{fig:interconnect_allocation_parallel} of a simulation with 4 models. Such a cycle forces sequential operation of the kernels with no speedup possible.

\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_parallel.eps}
	\caption{Interconnect benchmark results for parallel simulation.}
	\label{fig:interconnect_benchmark_parallel}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_parallel_allocation.eps}
	\caption{Interconnect parallel simulation trace for 6 models on 3 kernels.}
	\label{fig:interconnect_allocation_parallel}
\end{figure}

% The okayish
\subsubsection{Phold}
% Same cause, different results
In the Phold model, we first investigate the influence of the percentage of remote events on the speedup. A remote event in this context is an event that is sent from a model on one kernel to a model on another simulation kernel.
When remote events are rare, optimistic synchronization rarely has to roll back, thus increasing performance.
With more common remote events, however, optimistic synchronization quickly slows down due to frequent rollbacks.
Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block and wait.
Even though a single synchronization protocol is always ideal in this case, it already shows that different synchronization protocols respond differently to a changing model.
Adevs is significantly slower during conservative synchronization.
Analysis of profiling results shows that exception handling in adevs is the main cause. 
To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling. These functions require the user to implement a state saving mechanism in contrast to PythonPDEVS and dxex's optimistic kernels which handle this inside the kernel. We feel this would lead to an unfair comparison as we would like to keep the models synchronization-agnostic across all benchmarks.

\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_remotes.eps}
    \caption{Phold benchmark results for parallel simulation using four kernels, four atomics per node, with varying percentage of remote events.}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/phold_priority.eps}
	\caption{Phold benchmark results for parallel simulation using four kernels, with varying amount of high-priority events.}
	\label{fig:phold_priority}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/phold_parallel_allocation.eps}
	\caption{Phold benchmark trace for parallel simulation using three kernels.}
	\label{fig:phold_allocation}
\end{figure}

We slightly modified the Phold benchmark, to include high-priority events.
Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
Even when normal events occur most often, conservative synchronization always blocks until it can make guarantees.
Optimistic synchronization, however, simply goes forward in simulation time and rolls back when these high-priority events happen.
This situation closely mimics the case made in the comparison between both synchronization algorithms by~\cite{FujimotoBook}.
In Figure \ref{fig:phold_allocation} it is clear that in Phold it is possible for dependency cycles to form between kernels which as we have shown for Interconnect degrades performance for both optimistic and conservative. This is also the cause of the sublinear speedup observed in our Phold benchmark.

Figure~\ref{fig:phold_priority} shows how simulation performance is influenced by the fraction of these high-priority events.
If barely any high-priority events occur, conservative synchronization is penalized due to its excessive blocking, which often turned out to be unnecessary.
When many high-priority events occur, optimistic synchronization is penalized due to its mindless progression of simulation, which frequently needs to be rolled back.
Results show that there is no single perfect synchronization algorithm for this model: depending on configuration, either synchronization protocol might be better.

% The real life
\subsubsection{PholdTree}
%move to pholdtree
We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization. We will demonstrate that allocation is critical to achieve a parallel speedup and that the parameters of the model will give either conservative or optimistic an advantage.
\paragraph*{Allocation}\label{PholdTreeallocation}
The PholdTree benchmark can be configured to use 2 allocation schemes : breadth first and depth first. The breadth first allocation scheme will result in kernels that will form a dependency chain between them with multiple branches, much like in the Queue model. Such a linear dependency chain can result in a parallel speedup as we demonstrate with the Queue model, but this need not be true for all models as we will demonstrate in this section. A single kernel that has an unbalanced number of atomic models, models with unequal computation load in transition functions or activity will slow down the remainder of the chain. This effect is also apparent if the OS unfairly schedules the thread a kernel runs on. In conservative this will lead to excessive polling of the other kernels' eot values, whereas in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel. In Figures \ref{fig:pholdtree_visualize_parBFS} and \ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for both allocation schemes highlighting the remarks made in this paragraph.

\begin{figure}
	\center

	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeBFS.eps}
	\caption{PholdTree model breadth first allocation with 3 kernels.}
	\label{fig:PholdTree_model_bfs}

	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeDFS.eps}
	\caption{PholdTree model depth first allocation with 3 kernels.}
	\label{fig:PholdTree_model_dfs}
\end{figure}

\paragraph*{Strong Scaling}\label{pholdtreestrongscale}
In Figure \ref{fig:PholdTree_plot_alloc_high} we see that the difference in performance for all 4 kernel configurations compared to that shown in Figure \ref{fig:PholdTree_plot_alloc_low} is a constant factor. The probability of the priority message has no extra impact on performance other than that shown in sequential performance. 
The difference between depth first and breadth first allocated kernels is striking, the first results in sublinear speedup for both synchronization variants. %reword 
The breadth first allocation scheme will lead to a very high number of inter-kernel connections, which is detrimental for any parallel synchronization algorithm. An event exchanged between kernels cannot be securely deallocated without a GVT algorithm and all the complexity this entails. Even an asynchronous GVT algorithm will involve in a shared memory architecture some form of synchronization, and this will slow down the kernels. \\
Similar to the Queue benchmark the breadth first allocation scheme leads to a chain topology for the kernels, but PholdTree differs in the number of links between the kernels in the chain. In Queue there is only a single model on the edge of a kernel exchanging messages to a single other model in a neighbouring kernel, this is not the case in the PholdTree under breadth first allocation. This explains the difference in speedup between Queue and PholdTree despite the similarity in kernel topology.\\
Depth first allocation still has a higher inter kernel connection count, but not on the same order as breadth first allocated PholdTree. Depth first allocation will converge to a star topology which in this benchmark leads to a significant speedup.\\
The performance drop observed for 2 kernels for all allocation schemes and synchronization algorithms is in part due to the higher link count between kernels. With more kernels this effect will be spread across more kernels and result in a relative lower performance penalty. \\
Conservative synchronization suffers a further penalty in this benchmark when there are only a few kernels. The PholdTree model has a lookahead of $\epsilon$ leading the kernel to query each allocated atomic model for a lookahead value on each simulation round. As the atomic models are spread over more kernels this effect is reduced leading to an increase in performance. With 4 kernels conservative surpasses optimistic in speedup. This is clear evidence that conservative synchronization is a good candidate for parallel simulation even in simulations with uncertainty.
\begin{figure}
	\center

	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreealloclowp.eps}
	\caption{PholdTree model performance under different allocation schemes with low message probability}
	\label{fig:PholdTree_plot_alloc_low}

	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeallochighp.eps}
	\caption{PholdTree model performance under different allocation schemes with high message probability}
	\label{fig:PholdTree_plot_alloc_high}
\end{figure}

\paragraph*{Weak Scaling}
%Intro
In this paragraph we investigate dxex's parallel performance when the the number of atomic models distributed over a fixed number of kernels varies. From \ref{eq:pholdtreemodelcount} we know that increasing d will lead to a very rapid increase in models, we want to observe what happens in a closer to linear increase in model count per kernel, so we vary n. \\

% Explain why adevs is not plotted (its x100 slower)

Adevs' conservative implementation cannot handle the uncertainty (lack of non $\epsilon$ lookahead) here and is omitted from the comparison. Its parallel performance is two orders of magnitude slower than the sequential implementation. Upon investigation we conclude that this slowdown is caused by not implementing the \{begin/end\}lookahead() functions, which when not overridden in a model trigger an exception on each invocation. \\
This exception handling completely stalls the kernel. We do not implement this function in our PholdTree variant for adevs since this would force us to implement state saving in the model code and not use the provided state saving functionality in the kernel, as is done in dxex or PythonPDEVS's optimistic.\\
In dxex state saving is never required for a conservative simulation, this is handled transparently for the user who need not implement this behavior. 
By using the state saving technique inside the model code we feel we would no longer compare identical models across different synchronization algorithms, although we do not doubt adevs' increased performance when these functions are overridden.
\\

% Optimistic
In Figure \ref{fig:PholdTree_plot_weaknopt} we observe that dxex's optimistic kernels are slightly sensitive to the number of atomic models allocated to them. In the worst case with 2 kernels there is no real speedup observable, as soon as the number of kernels increases we see a converging speedup trend, that except for 32 kernels is almost constant despite the increase in n. Note that only depth first allocated kernels are measured here, breadth first as we can see in \ref{fig:PholdTree_plot_alloc_high} has no speedup advantage. The probability parameter is kept at 0.1 for the same reason, it only induces a linear increase in load.
\\

% Conservative
Conservative has a more nuanced speedup behavior in this benchmark. We have detailed how a conservative kernel in dxex scales linear in the number of atomic models (if lookahead is $\epsilon$), this effect is more clearly visible here. \\
It is clear that as the fanout (n) of the model increases the kernel performance degrades rapidly. The intersection of the performance graph of each configuration is shifted to the right, the actual 1-speedup point increases as the kernelcount increases. For a d=4, n=5 configuration with 4 kernels each kernel has a load of ~1300 atomic models when it crosses the 1-speedup line. \\
An 8 kernel configuration with d=4 n=6 can sustain a load of ~2700 atomic models before it reaches that point. From the results we see that the 32-kernel configuration is not yet slowing down with the current parameters. \\
The modelcount is only a part of the explanation of this effect, as the kernelcount increases the amount of inter kernel messages will be split into more distinct sets which can be handled with more concurrency in dxex's architecture. 
That this effect is not always true can be observed from the d=4 n=3 32-kernel datapoint. Note that in this configuration 341 atomic models are distributed over 32 kernels, a relative low model load can more easily highlight synchronization overhead, even if allocation is optimal.\\
As with optimistic we do not include breadth first allocated benchmarks in this speedup plot since none of those achieve a speedup higher than 1.
\\
If we combine both to determine which synchronization protocol is more optimal we see in Figure \ref{fig:PholdTree_plot_weakall} that for these configurations conservative with 8 kernels is an ideal configuration up until n=5, where the modelcount starts to degrade performance. Optimistic with the same kernelcount is almost insensitive to the increase in modelcount and is therefore a more robust choice. Note how optimistic and conservative with 32 kernels at this point still have a reasonable speedup which is surprising given the synchronization overhead in such a large set of kernels.
\begin{figure}
	\center

	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingnopt.eps}
	\caption{PholdTree model weak scaling under varying fanout and optimistic synchronization}
	\label{fig:PholdTree_plot_weaknopt}

	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingncon.eps}
	\caption{PholdTree model weak scaling under varying fanout and conservative synchronization}
	\label{fig:PholdTree_plot_weakncon}

	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingall.eps}
	\caption{PholdTree model weak scaling under varying fanout and different synchronization algorithms}
	\label{fig:PholdTree_plot_weakall}

\end{figure}

\subsection{Measuring Synchronization Overhead}
The simulation kernels have a non trivial overhead compared to a sequential kernel. In this section we will try to isolate this overhead and highlight the consequences this has for applying a synchronization protocol in real life models.
Profiling indicates that there are in the most trivial of models only 2 causes for computational load outside of the actual simulation : memory management and random number generation. 
\paragraph*{Memory management}\label{sec:4-subsec:overhead-pgraph:memory}
\begin{figure}
	\center

	\includegraphics[width=\modelfraction\columnwidth]{fig/memincrease.eps}
	\caption{Memory usage in parallel synchronization}
	\label{fig:memincrease}

	\includegraphics[width=\modelfraction\columnwidth]{fig/memory_allocators.eps}
	\caption{Effect of memory allocators on sequential runtime.}
	\label{fig:memallocators}

	\includegraphics[width=\modelfraction\columnwidth]{fig/memory_allocators_parallel.eps}
	\caption{Effect of memory allocators on parallel runtime.}
	\label{fig:memallocators_parallel}
\end{figure}
In dxex a model author is given access to automatic memory management for events and states. A model written for a sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the model author) the (de)allocation semantics of events or states. \\
In sequential and conservative simulation the kernel gives the guarantee that no state will be copied. Dxex's simulation control hides the implementation details of this management, a model author can focus on the actual semantics of the model. 
The kernels use memory allocators backed by a thread-aware pooling library to reduce the performance impact of memory management as much as possible. 
A sequential simulation kernel can leverage the invariant that no allocated event will persist beyond a single time advance and use a arena-style allocator. Conservative and optimistic simulation need to use generic pool allocators. \\
An important observation to make is that intra-kernel events can be pooled more aggressively, whereas inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization. A simulation with high inter kernel events will suffer a performance hit, whereas the impact of high intra kernel events can be optimized using arena allocators.\\
The allocation interfaces are instrumented in dxex to trace memory usage in a simulation and can serve as a basic memory leak detector. Shielding the user from (de)allocation complexity comes at a cost in performance, as the comparison with adevs in simulations with a high event frequency \ref{fig:Interconnect_benchmark} of events shows.\\
Figure \ref{fig:memincrease} shows an experiment where we measure peak memory usage (maximum resident set size (kB)) for each synchronization protocol and determine how the memory usage differs with respect to sequential simulation. Both conservative kernels require only a small constant increase in memory usage. 
\\
% Implementation choices
Dxex uses Boost Pool\cite{boostpool} allocators in parallel simulation kernels, and can use arena-style allocators for sequential simulation. 
The latter can be faster, but at the cost of extra configuration. The allocators are supplemented by the library tcmalloc \cite{tcmalloc}. In Figure \ref{fig:memallocators} the Interconnect model is benchmarked under the different combinations of allocation strategies. \\
% Sequential
Interconnect is a perfect test case since it generates each round a set of events quadratic in size to the number of atomic models in the simulation. With each event dynamically allocated this makes it the perfect stress test for any allocation strategy. Both tcmalloc and Boost pools result in an advantage, but not necessarily combined. \\
% Paralell
We next investigate the effect of both strategies in an optimistic simulation. We have already demonstrated that conservative simulation differs little in memory usage from sequential, so we only benchmark the optimistic kernel. From the results it is clear that optimistic simulation benefits greatly from the use of tcmalloc, regardless of the allocator. Nonetheless the pool allocator still reduce the allocation overhead. \\
% Interpretation note
Both the pools and tcmalloc will try to keep memory allocated to them by the OS, this gives a somewhat misleading view in statistics collected from the OS. The dxex kernels themselves and the pool interfaces allow the user to instrument event allocation and deallocation exactly, useful for example in observing how efficient GVT-triggered deallocation is in practice.\\
%Conclude pgraph
In conclusion both techniques are required to reduce the overhead of memory allocations in dxex, and on by default. 
\paragraph*{Random Number Generators}
A non trivial amount of time in a simulation spent waiting for a random number generator can mask synchronization overhead. Several of our benchmarks use random number generators to test the effect of uncertainty in a simulation, either by random time advances or random destinations for events. \\
In a parallel simulation we also need to guarantee isolation between the random calls to avoid excessive synchronization on the rng object itself. 
Dxex uses the Tina random number generator collection (trng) \cite{PhysRevE.75.066701} as an alternative random number generator as it is designed with performance and multithreading usage in mind. 
We also wrote a new version of our models that stores the rng object in the state instead of calling a shared thread local object to isolate the synchronization overhead.\\
From Figure \ref{fig:Queuerngspeedup} we see that storing the rng in the state is a very expensive operation for the default STL random number generator. The size of the rng object in the state triggers memory overhead but it is interesting to see that adevs' conservative is also sensitive to this effect even though in theory in conservative synchronization no state copying/saving is required. 
Dxex's conservative kernel is insensitive to storing the rng object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation. 
The overhead of initially doing so is therefore constant and not a factor in long simulations.\\
Replacing the STL rng with the threading optimized trng we are more accurately able to isolate synchronization overhead. Figure \ref{fig:Queuerngspeedup} shows that both dxex and adevs in sequential simulation gain a ~ 3 speedup by using the faster random number generator. The parallel kernels have smaller speedup with adevs' conservative and dxex optimistic achieving an almost identical factor. Dxex conservative is almost insensitive to the changing of the rng. 
Dxex optimistic and adevs conservative are in the default configuration spend relatively more time waiting for random numbers than dxex conservative, demonstrated by the speedup gained when random number generation is accelerated.
The actual synchronization overhead in all parallel kernels still dominates the cost of random number generation which is seen from the speedup difference between parallel and sequential.\\
\begin{figure}
	\center

	\includegraphics[width=\modelfraction\columnwidth]{fig/rngspeedupeffectdevstone.eps}
	\caption{Speedup with different rng usage patterns in Queue model}
	\label{fig:Queuerngspeedup}

\end{figure}
% Analyse

### 2016-06-23 publications/Journal/sequential-performance.tex
Add journal paper draft.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

\subsection{Sequential Simulation}
% 2 things to show : baseline performance and contrast of dxex v adevs
\subsubsection{Queue}\label{4-seq-Queue}
% explain what we did: fixed number of models (400, 600 & 800), increasing depth
For the first benchmark, we tested the effect of hierarchical complexity of the model in the performance of the simulator. A set of three tests was performed, where each test has the same number of models but an increasing depth.
% compare dxex & adevs performance
The results can be seen in Figure~\ref{fig:Queue_benchmark_seq}. Since dxex symbolically flattens the model, there is no performance hit when the depth is increased. The overhead of running the directconnect algorithm is one time only and negligible when the end time of the simulation is sufficiently large. Adevs on the other hand does suffer from the increased depth. With every new hierarchical layer, routing an event from one atomic model to the next becomes more expensive, resulting in an increase in runtime.
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/queue_fixed_sequential.eps}
	\caption{Queue benchmark results for sequential simulation.}
	\label{fig:Queue_benchmark_seq}
\end{figure}
\subsubsection{Interconnect}\label{4-seq-Interconnect}
In the Interconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.
Analysis showed that this is caused by the high amount of exchanged events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools. Dxex reuses the same event type for all synchronization algorithms. To shield the user from threading and deallocation concerns dxex provides an event superclass from which the user can derive to create a specialized event type. Copying and deallocation semantics are solved by the kernels at a runtime cost in simulations where event frequency is very high. Profiling the diverging benchmarks clearly shows the increasing cost of output generation (event creation) and deallocation as the determining factor in the gap in performance. We refer the interested reader to the dxex \hyperref{https://bitbucket.org/bcardoen/devs-ex-machina}{}{repo}{repository} for the profiling call graphs for the different benchmarks.

\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_sequential.eps}
	\caption{Interconnect benchmark results for sequential simulation.}
	\label{fig:Interconnect_benchmark}
\end{figure}
% We want to show the effect of message storms, that dxex is sensitive to this (even with allocators)
\subsubsection{PHold}\label{4-seq-PHold}
% Phold : difference with interconnect, less messages == equal performance dxex adevs
The PHold model is very similar to the Interconnect model. The biggest difference is that the amount of messages sent is much lower. The number of events scales linear with the number of models, not quadratic. Figure~\ref{fig:Phold_benchmark} shows that the performance of dxex and adevs are very close to each-other, with adevs slightly outperforming dxex.
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/phold_sequential.eps}
	\caption{PHold benchmark results for sequential simulation.}
	\label{fig:Phold_benchmark}
\end{figure}
\subsubsection{PHoldTree}
The PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain as in Queue.
This topology is interesting since it highlights the effects of allocation which we have shown already in PHold and Interconnect to be of vital importance. This model allows us to investigate in depth the effects of non-cyclic allocation strategies and measure parallel speedup.\\
Since adevs does not use the directconnect algorithm, we expect some performance penalty between dxex and adevs. Conversely the messaging intensity in PHoldTree can be configured to highlight dxex's performance disadvantage in  benchmarks with a high event frequency as shown in the Interconnect benchmark.
\paragraph*{Event frequency}
Similarly when the probability on a priority message (p) increases the amount of messages generated will increase and thus the runtime. In Figure \ref{fig:PHoldtree_seq_p_benchmark} we observe that the messaging probability has a near linear impact on performance, with the shape of the curve due to rounding errors. The advantage adevs has in simulations where event frequency is high has been demonstrated by the interconnect benchmark. Dxex chooses to provide the user with an event type where the user need not consider deallocation or copy semantics, at a cost in performance where event frequency is high. Counterintuitively the probability of a high priority event in this model does not affect the performance difference between conservative and optimistic. The key parameter quickly becomes the load of a kernel in models. Our conservative implementation in an uncertain (trivial lookahead) simulation is very sensitive to a high load in models, whereas optimistic has no such limitation.
\paragraph*{Hierarchy}
To establish a performance baseline for the parallel simulation benchmarks we instrument how the PHoldTree benchmark performs sequential when we vary each of the model's parameters. 
When we increase the depth and fanout of the model, we expect to see an increase in runtime. In Figure \ref{fig:PHoldtree_seq_dn_benchmark} the n parameter (fanout) determines the performance penalty adevs suffers w.r.t dxex but we need to look at the profiling call graphs to determine what causes adevs' performance graph to behave non linear. \\
If we compare the instances d=2, n=2 with d=2, n=4 which show a clear deviation in performance between adevs and dxex for the latter, we conclude that the increase in width per subtree (n) leads to higher overhead for adevs. Dxex in contrast due to the directconnect algorithm has no such overhead. The intuition behind this effect, and especially why the d parameter does not have such an impact, is that an increase in n will rapidly increase the number of connections \ref{eq:pholdtreelinkcount} (and thus the routing problem adevs faces), whereas d will increase modelcount \ref{eq:pholdtreemodelcount} faster but have a lesser impact on connections. Both kernels scale linear in increasing number of atomic models. 
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/pholdtree_sequential_p.eps}
	\caption{PHoldTree for increasing probability of priority message.}
	\label{fig:PHoldtree_seq_p_benchmark}
\end{figure}
% Img for p increasing
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/pholdtree_sequential_dn.eps}
	\caption{PHoldTree : Effect of hierarchy.}
	\label{fig:PHoldtree_seq_dn_benchmark}
\end{figure}

### 2016-06-24 publications/Journal/3-features.tex
Re-add publication folder + draft journal paper.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

--- 2016-02-18 paper/3-features.tex
@@
+\newcommand{\plotfraction}{0.85}
 Historically, dxex is based on PythonPDEVS~\cite{PythonPDEVS}.
 Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
-Dxex is a C++11-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, while making some of its own additions.
+Dxex is a C++14-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, while making some of its own additions.
 So while the feature set is not too comparable, the architectural design, core simulation algorithm, and optimizations, are highly similar.
@@
 In our case, a user can pass the simulation kernel the synchronization protocol to use for this model.
-Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language.
+Our implementation in C++14 also allows for optimizations which were plainly impossible in an interpreted language.
 
 Since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
-This is due to dxex models being grafted on C++11, whereas PythonPDEVS models are grafted on Python.
+This is due to dxex models being grafted on C++14, whereas PythonPDEVS models are grafted on Python.
 
@@
 \subsection{Synchronization protocols}
-We previously explained the existence of different synchronization protocols exist, each optimized for a specific kind of model.
+We previously explained the existence of different synchronization protocols, each optimized for a specific kind of model.
 As no single synchronization protocol is ideal for all models, a general purpose simulation tool should support multiple protocols.
@@
 Finally, sequential simulation is still required for models where parallelism is bad, where all protocols actually slow down simulation.
-
-\vspace{0.3cm}
 
@@
     \includegraphics[width=\columnwidth]{fig/cores_class_diagram.eps}
-	\caption{Dxex kernel design.}
+	\caption{Dxex design.}
 	\label{fig:class_diagram}
@@
 
-Reading and writing to shared memory is done through the use of the new C++11 synchronization primitives.
+Reading and writing to shared memory is done through the use of the new C++14 synchronization primitives.
 Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
-C++11 further allows us to make the implementation portable, as well as more efficient: the compiler might know of optimizations specific to atomic variables.
+C++14 further allows us to make the implementation portable, as well as more efficient: the compiler might know of optimizations specific to atomic variables.
 
@@
 This made us switch to the use of thread-local (using \texttt{tcmalloc}) memory pools.
-Again, we made use of specific new features of C++11, that were not available in Python, or even previous versions of the C++ language standard.
+Again, we made use of specific new features of C++14, that were not available in Python, or even previous versions of the C++ language standard.
 
@@
 We define simulation kernel transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
-User should thus only provide one model, implemented in C++11, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.
+User should thus only provide one model, implemented in C++14, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.
 Switching between simulation kernels is as simple as altering the simulation termination time.
@@
 Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
+
+\subsection{Features}
+In this section we highlight some of the extra features dxex offers to modellers.
+\subsubsection{Tracing}
+Like PythonPDEVS dxex has full tracing capabilities supporting a wide number of output formats : json, XML, cell, console, ...
+.
+The user can implement his own outputpolicy and use it in our modular tracing design.
+The tracing functionality itself is implemented asynchronous and thread-safe.
+By overriding the output functions in the Message or State classes the user can generate model specific output for use with an existing tracing output policy.
+Tracing can be completely disabled to maximize simulation performance should this be desired (e.g.
+benchmarks).
+% Talk about adevs?
+\subsubsection{Memory Allocation}
+As we will demonstrate in \ref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel.
+Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation.
+This allows us by using a single compilation parameter to change the allocation strategy per kernel type.
+This interface allows instrumentation of memory usage.
+Not only useful to detect memory leaks should this be a concern, it can also be of use to a model author to trace event creation in large models.
+A model author can rely on the default implementation as the fastest common configuration verified by the benchmarks.
+\subsubsection{Model Allocation}
+A model author can specify which kernel a model should be allocated to, should such manual intervention be required.
+This is handled by the default model allocator.
+If no preference is given a simple striping scheme is followed.
+It should be obvious that this is not sufficient in most simulations.
+In \ref{sec:4-performance} we analyze allocation schemes and their performance impact.
+By overriding the default allocator a model author can tune the allocation scheme to maximize parallel speedup.
+This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.
+\subsubsection{Logging}
+A high performance asynchronous buffered threaded logging module is used in dxex which offers a complete overview of the entire simulation.
+The logging functionality can be used in any model, it only requires linking to the dxex library.
+Logging statements are written to file with a granular level system.
+In case of a severe crash the logging system will ensure persistence of the log file itself up until the last statement.
+Logging can be completely disabled to remove the performance impact.
+\subsubsection{Visualization and Statistics}
+Tracing offers a black box view of a DEVS model in a simulation, with transitions, events and states recorded at each state change.
+Dxex supplements this optionally with more in depth insights into the actual simulation.
+The allocation of models across kernels is visualized at the end of a simulation, with graphical representation of message frequency.
+This allows a user to fine tune future allocation schemes and the interface used by the visualization tool can also be used for more advanced purposes such as dynamic tracing of the simulation.\\
+Invaluable during the development of dxex were runtime statistics collected by a simulation kernel.
+These record any measurable event in the simulation such as : events created/destroyed, inter kernel event frequency, reverts, GVT calculations, lookahead and eot values and insights into the fairness between the kernels themselves.
+Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, or observe how balanced inter kernel message traffic actually is during a simulation.
+The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or allow activity tracking.
+Figures \ref{fig:pholdtree_visualize_seq},\ref{fig:pholdtree_visualize_parDFS},\ref{fig:pholdtree_visualize_parBFS} give an example of the modular visualization output generated by dxex after simulation the PholdTree model.
+These figures demonstrate the effect of allocation visually, the high inter kernels event count will degrade performance, as we will demonstrate in
+\ref{PholdTreeallocation}.
+Dotted lines represent connections between models that did not see events exchanged, directed edges represents events with frequency as weight and color to distinguis inter versus intra kernel events.
+% Add figure
+\begin{figure}
+	\center
+	\includegraphics[width=\plotfraction\columnwidth, height=8cm, keepaspectratio]{fig/pholdtreed1n3t5000.eps}
+	\caption{Visualization of PholdTree (d=1,n=3,t=5000) sequential simulation.}
+	\label{fig:pholdtree_visualize_seq}
+\end{figure}
+\begin{figure}
+	\center
+	\includegraphics[width=\plotfraction\columnwidth,  height=8cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
+	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
+	\label{fig:pholdtree_visualize_parBFS}
+\end{figure}
+\begin{figure}
+	\center
+	\includegraphics[width=\plotfraction\columnwidth,  height=8cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
+	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
+	\label{fig:pholdtree_visualize_parDFS}
+\end{figure}

### 2016-06-24 publications/Journal/4-performance.tex
Re-add publication folder + draft journal paper.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

% What we want to write

% DESIGN OF MODELS
\input{models.tex}

% SEQUENTIAL
\input{sequential-performance.tex}

% PARALLEL
\input{parallel-performance.tex}

% CONCLUSION
\input{conclusion.tex}

### 2016-06-24 publications/Journal/6-conclusions.tex
Re-add publication folder + draft journal paper.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

--- 2015-12-14 paper/6-conclusions.tex
@@
-In this paper, we introduced DEVS-Ex-Machina (``dxex''), a new C++11-based \textsf{Parallel DEVS} simulation tool.
-Our main contribution is the implementation of multiple synchronization protocols for parallel simulation.
+In this paper, we introduced DEVS-Ex-Machina (``dxex''), a new C++14-based \textsf{Parallel DEVS} simulation tool.
+Our main contribution is the implementation of multiple synchronization protocols for parallel multicore simulation.
 We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
@@
 Performance is measured both in elapsed time, and memory usage.
+Our empirical analysis shows that allocation of models over kernels is critical to enable a parallel speedup.
+Furthermore we have shown when and why optimistic synchronization can outperform conservative and vice versa.
 
@@
 Thirdly, activity algorithms, as already implemented by PythonPDEVS, can also be implemented in dxex, to determine how they influence simulation performance.
+Finally automatic allocation is possible by analysis of the connections between models.
+This information is already used in dxex to determine the blocking order for conservative synchronization and in the directconnect algorithm.
+A graph algorithm that distributes models while avoiding cycles in the resulting kernels topology could be used to maximize parallel speedup in either optimistic or conservative synchronization.

### 2016-06-24 publications/Journal/abstract.tex
Re-add publication folder + draft journal paper.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

--- 2015-12-11 paper/abstract.tex
@@
 Several synchronization protocols have been proposed, each with their distinct advantages and disadvantages.
-Due to the significantly different implementation of both protocols, most Parallel DEVS simulation tools are limited to only one such protocol.
-In this paper, we present a Parallel DEVS simulator, grafted on C++11, but based on PythonPDEVS, which supports both conservative and optimistic synchronization.
+Due to the significantly different implementation of these protocols, most Parallel DEVS simulation tools are limited to only one such protocol.
+In this paper, we present a Parallel DEVS simulator, grafted on C++14, but based on PythonPDEVS, which supports both conservative and optimistic synchronization.
 We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol.

### 2016-06-24 publications/Journal/conclusion.tex
Re-add publication folder + draft journal paper.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

\subsection{Conclusion}
We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
We will summarize the key aspects in achieving good parallel performance here. % reword aspects syn
\subsubsection{Allocation}
Allocation of atomic models over kernels is critical to obtain a speedup. In the case of the Interconnect model it is not possible to allocate models without introducing a cyclic dependency between kernels with severe performance degradation as a result. 
Even when no cycles exist in the topology between kernels a good allocation scheme will minimize the number of connections between kernels and aim for a star topology as in PHoldTree with depth first allocation or a chain topology as in the Queue model. 
Dxex can optionally offer the user a visualization of the behavior of the kernels under a user supplied allocation scheme to allow for more insight. 
The same instrumentation could in future be used to optimize an existing allocation scheme.
\subsubsection{Uncertainty}
A simulation where future behaviour cannot be predicted is typically not suited for conservative simulation, but we have demonstrated that dxex's conservative kernel can still offer a non trivial speedup in this context. 
Optimistic is not constrained by this uncertainty and can offer a good speedup regardless of lookahead. 
PHOLD and PholdTree demonstrate that a single model can benefit from different synchronization protocols depending on a single parameter.
\subsubsection{Limits}
Dxex's conservative kernel is constrained by the amount of models it controls, as shown in the PHoldTree benchmark. This is especially the case when the minimal lookahead is $\epsilon$ requiring near constant polling of models for lookahead values. 
Optimistic is not constrained by the number of models it manages, but can quickly consume all available memory in a simulation. While this effect can be reduced with a thread aware allocation scheme, the underlying problem remains.

### 2016-06-24 benchmarks/__init__.py

@@ -1,9 +1,0 @@
-#
-# This file is part of the DEVS Ex Machina project.
-# Copyright 2014 - 2015 University of Antwerp
-# https://www.uantwerpen.be/en/
-# Licensed under the EUPL V.1.1
-# A full copy of the license is in COPYING.txt, or can be found at
-# https://joinup.ec.europa.eu/community/eupl/og_page/eupl
-#      Author: Stijn Manhaeve
-#

### 2016-06-24 benchmarks/benchmarker.py

@@ -1,54 +1,0 @@
-#
-# This file is part of the DEVS Ex Machina project.
-# Copyright 2014 - 2015 University of Antwerp
-# https://www.uantwerpen.be/en/
-# Licensed under the EUPL V.1.1
-# A full copy of the license is in COPYING.txt, or can be found at
-# https://joinup.ec.europa.eu/community/eupl/og_page/eupl
-#      Author: Stijn Manhaeve
-#
-# Main file of the project. This file contains only the most high level functions
-
-
-def execBenchmark(bmark, driver, folders, compile=True):
-    """
-    Executes a benchmark and returns a list of output files.
-    params:
-      bmark: A Benchmark object
-      driver: A BenchmarkDriver object
-      folders: A Folders object. The data will be stored in this folder
-      compile: [default: True] Use the compilation function
-    returns:
-      an iterable over (Path, args)
-    """
-    # get real output folders
-    folders = driver.foldergen(bmark, folders)
-    # compile benchmark
-    if bmark.compile is not None and compile:
-        bmark.compile()
-    # use generator pattern to return a list with filenames
-    for args, gfile in zip(bmark.getExecList(), driver.filegen()):
-        rfile = folders.perf / gfile
-        driver.execute(args, rfile)
-        yield rfile, args
-
-
-def execAnalyze(bmark, driver, analyzer, folders, save=True):
-    """
-    Executes a benchmark and performs the analysis.
-    params:
-      bmark: A Benchmark object
-      driver: A BenchmarkDriver object
-      analyzer: A BenchmarkAnalyzer object
-      folders: A Folders object. The data will be stored in this folder
-      save: [optional, default is True] If this argument evaluates to True,
-                                        the gathered data will also be saved to a file.
-    returns:
-      An object that contains all the gathered data.
-    """
-    files = list(execBenchmark(bmark, driver, folders))
-    data = analyzer.open(bmark, folders)
-    data = analyzer.read(files, data)
-    if save:
-        analyzer.save(data, bmark, folders)
-    return data

### 2016-06-24 benchmarks/csv.py

@@ -1,290 +1,0 @@
-#
-# This file is part of the DEVS Ex Machina project.
-# Copyright 2014 - 2015 University of Antwerp
-# https://www.uantwerpen.be/en/
-# Licensed under the EUPL V.1.1
-# A full copy of the license is in COPYING.txt, or can be found at
-# https://joinup.ec.europa.eu/community/eupl/og_page/eupl
-#      Author: Stijn Manhaeve
-#
-# collection of functions for saving the data to Tim-readable csv
-from .misc import assertPath
-import traceback
-
-
-def _addDataLine(arg, res, f, delim=';', argparse=lambda x: "\"{}\"".format(" ".join(map(str, x)))):
-    """
-    Adds the data lines for a single argument and the individual results.
-    params:
-      arg: The given arguments for the benchmark
-      res: A dictionary containing all the results
-      f: a file object to which the data is written
-    """
-    # convert the arguments to something we can actually write to the file
-    argstr = argparse(arg)
-    if type(argstr) is not str:
-        argstr = delim.join(map(str, argstr))
-    # keep a list of lists that will contain all the data points
-    values = []
-    for _, val in sorted(res.items()):
-        if val is None:
-            continue
-        values.append(val["value"])
-        if "extra" in val.keys():
-            values.append(val["extra"]["value"])
-    # values now contains a list of lists of values.
-    # Just iterate over it and write it to the file
-    for vals in zip(*values):
-        f.write(argstr)
-        f.write(delim)
-        f.write(delim.join(map(str, vals)))
-        f.write('\n')
-
-
-def _addHeaderLine(res, f, delim=';', argColumns="\"command\""):
-    """
-    Adds a header line.
-    """
-    f.write(argColumns)
-    for name, val in sorted(res.items()):
-        if val is None:
-            continue
-        f.write(delim)
-        if val["unit"] is not None:
-            f.write("\"{} ({})\"".format(name, val["unit"]))
-        else:
-            f.write("\"{}\"".format(name))
-        if "extra" in val.keys():
-            val = val["extra"]
-            f.write(delim)
-            if val["unit"] is not None:
-                f.write("\"{} ({})\"".format(name, val["unit"]))
-            else:
-                f.write("\"{}\"".format(name))
[…]

### 2016-06-24 benchmarks/defaults.py

@@ -1,247 +1,0 @@
-#
-# This file is part of the DEVS Ex Machina project.
-# Copyright 2014 - 2015 University of Antwerp
-# https://www.uantwerpen.be/en/
-# Licensed under the EUPL V.1.1
-# A full copy of the license is in COPYING.txt, or can be found at
-# https://joinup.ec.europa.eu/community/eupl/og_page/eupl
-#      Author: Stijn Manhaeve
-#
-
-from .misc import assertPath, printVerbose
-from .structs import Benchmark, Folders, BenchmarkDriver, BenchmarkAnalyzer
-from itertools import count
-from datetime import datetime
-from pathlib import Path
-import base64
-import json
-import hashlib
-import platform
-import multiprocessing
-import locale
-from subprocess import call, TimeoutExpired, DEVNULL
-from collections import Iterable
-
-
-# global object for passing command line arguments.
-args = None
-
-
-# first, get user locale by forcing it to load it instead of the C locale
-locale.setlocale(locale.LC_ALL, '')
-# custom parsing function for floats
-def pfloat(f):
-    """
-    Tries to parse a float.
-    """
-    if len(f) == 0:
-        raise FloatingPointError("failed to parse")
-    try:
-        return locale.atof(f)
-    except:
-        try:
-            printVerbose(args.verbose, "note: failed to directly parse string: {}\tWill try again by removing last character".format(f))
-            if f.endswith('%'):
-                return pfloat(f[:-1])/100
-            return pfloat(f[:-1])
-        except:
-            raise ValueError("Failed to parse as float: {}".format(f))
-
-
-def defaultFilegen(format=None, **kwargs):
-    """
-    Generates an endless list of file names for perf output files.
-    You can optionally give a different format and some kwargs.
-    """
-    format = '{:04d}.perf' if format is None else format
-    for i in count():
-        yield format.format(i, **kwargs)
-
-
-def defaultFoldergen(benchmark, folders):
-    """
-    Opens the necesarry set of folders for a benchmark and returns these folders.
-    params:
-      benchmark: A Benchmark object. @see structs.py/Benchmark
-      folders: A Folders object, @see structs.py/Folders
-    """
-    fmt = '{0:%Y}.{0:%m}.{0:%d}_{0:%H}.{0:%M}.{0:%S}'
-    newpath = assertPath(folders.perf / benchmark.name / fmt.format(datetime.now()))
-    newpath.resolve()
[…]

### 2016-06-24 benchmarks/misc.py

@@ -1,33 +1,0 @@
-#
-# This file is part of the DEVS Ex Machina project.
-# Copyright 2014 - 2015 University of Antwerp
-# https://www.uantwerpen.be/en/
-# Licensed under the EUPL V.1.1
-# A full copy of the license is in COPYING.txt, or can be found at
-# https://joinup.ec.europa.eu/community/eupl/og_page/eupl
-#      Author: Stijn Manhaeve
-#
-
-
-def assertPath(path):
-    """
-    Creates the folder specified by path, if it doesn't exist already.
-    params:
-      path: a pathlib Path object denoting a folder.
-    returns
-      the same path object.
-      This allows using a more functional style of programming.
-    """
-    if not path.exists():
-        path.mkdir(parents=True)
-    return path
-
-
-def printVerbose(doPrint, *args, **kwargs):
-    """
-    prints only if verbose flag is used
-    params:
-      doPrint  if True, call the regular print with the remainder of the arguments.
-    """
-    if doPrint:
-        print(args, kwargs)

### 2016-06-24 benchmarks/structs.py

@@ -1,84 +1,0 @@
-#
-# This file is part of the DEVS Ex Machina project.
-# Copyright 2014 - 2015 University of Antwerp
-# https://www.uantwerpen.be/en/
-# Licensed under the EUPL V.1.1
-# A full copy of the license is in COPYING.txt, or can be found at
-# https://joinup.ec.europa.eu/community/eupl/og_page/eupl
-#      Author: Stijn Manhaeve
-#
-
-from collections import namedtuple
-
-Folders = namedtuple('Folders', 'root perf json plots')
-Folders.__doc__ = """
-Small container class for keeping track of some paths.
-members:
-  * root:  (field) The root folder of the benchmark system.
-  * perf:  (field) The folder that will contain the perf stat output.
-  * json:  (field) The folder that will contain json files for each benchmark.
-  * plots: (field) The folder where the plots are put.
-"""
-
-Benchmark = namedtuple('Benchmark', 'name compile getExecList shortName description')
-# evil witchcraft that changes the default arguments of the constructor.
-Benchmark.__new__.__defaults__ = ('', '')
-Benchmark.__doc__ = """
-Represents a single benchmark job.
-members:
-  * name: (field) A supposedly unique name for this benchmark.
-                  his name will be used as a folder name.
-  * compile: (function) This function will be executed once, before the actual benchmark starts.
-                        You can use this function to compile an executable.
-                        signature:
-                          void compile()
-  * getExecList: (function) Returns a (finite!) iterable that iterates over a list of strings or lists.
-                         These items should form the command that is executed with the perf stat calls.
-                         signature:
-                           iterable getExecList()
-  * shortName:   (field) [optional, default value: ""] A shorter name that need not be unique.
-  * description: (field) [optional, default value: ""] An optional description.
-"""
-
-BenchmarkDriver = namedtuple('BenchmarkDriver', 'filegen foldergen execute')
-BenchmarkDriver.__doc__ = """
-Collection of functions for executing an entire benchmark.
-members:
-  * filegen: (function) Generates new filenames.
-                        These filenames are used to write the perf stat output to disk.
-                        signature:
-                          iterable filegen()
-  * foldergen: (function) Creates a set of output folders for a specific Benchmark,
-                          based on a general folder set.
-                          signature:
-                            Folders foldergen(Benchmark, Folders)
-  * execute: (function) executes a single benchmark with a given list of arguments
-                        and a filepath. Returns True if the call was successful, otherwise False
-                        signature:
-                          bool execute(list, Path)
-"""
-
[…]

### 2016-06-24 benchmarks/test.py

@@ -1,38 +1,0 @@
-#
-# This file is part of the DEVS Ex Machina project.
-# Copyright 2014 - 2015 University of Antwerp
-# https://www.uantwerpen.be/en/
-# Licensed under the EUPL V.1.1
-# A full copy of the license is in COPYING.txt, or can be found at
-# https://joinup.ec.europa.eu/community/eupl/og_page/eupl
-#      Author: Stijn Manhaeve
-#
-
-import defaults
-import benchmarker
-from csv import toCSV
-
-
-def echotestGen():
-    for i in range(5):
-        yield ['dd', 'if=/dev/zero', 'of=/dev/null', 'count={}'.format(10**(i+2))]
-
-
-def noop():
-    """
-    literally does nothing (except for eating away precious cpu time).
-    """
-    pass
-
-if __name__ == '__main__':
-    bmark = defaults.Benchmark('basictest', None, echotestGen, 'Just a basic test that pipes some data from /dev/zero to /dev/null', 'basic test')
-    driver = defaults.defaultDriver
-    analyzer = defaults.perfAnalyzer
-    sysInfo = defaults.getSysInfo()
-    folders = defaults.getOFolders(sysInfo, "tests")
-    defaults.saveSysInfo(sysInfo, folders)
-    files = list(benchmarker.execBenchmark(bmark, driver, folders))
-    data = analyzer.open(bmark, folders)
-    data = analyzer.read(files, data)
-    analyzer.save(data, bmark, folders)
-    toCSV(data, folders.plots / "{}.csv".format(bmark.name))

### 2016-06-24 publications/Journal/1-introduction.tex
Re-add publication folder + draft journal paper.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
First set of revisions and fixes.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

--- 2016-02-15 paper/1-introduction.tex
@@
 Users who simulate a wide variety of models, with different ideal synchronization protocols, can simply run the same model with different synchronization protocols.
+We investigate in this paper how model allocation and uncertainty determine the choice between synchronization protocols.
+The synchronization overhead is demonstrated by reducing the computational load of a model to near zero.
+Dxex also offers a visualization of the simulation with in depth statistics on the synchronization algorithms.
+With all this information the modeller can make the best informed decision on which synchronization algorithm to use.
 
-Our tool is based on PythonPDEVS, but implemented in C++11 for increased performance.
+Our tool is based on PythonPDEVS, but implemented in C++11 for increased performance, using features from the new C++14 standard when possible.
 Unlike PythonPDEVS dxex only supports multicore parallelism.
@@
 We implemented a model that, depending on a single parameter, changes its ideal synchronization protocol.
+We demonstrate using several models the factors influencing the performance under a given synhronization protocol.
 Dxex, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol. With dxex users can always opt to use the fastest protocol available.
 To verify that our flexibility does not counter performance, we compare to adevs, currently one of the fastest \textsf{DEVS} simulation tools available~\cite{PythonPDEVS,DEVSSurvey}.
+
+% TODO uncomment if completed
+%Dxex offers online switching functionality that can be used by machine learning %techniques to automatically switch between the most optimal synchronzation protocol.
 

### 2016-06-24 publications/Journal/3-features.tex
First set of revisions and fixes.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
-Dxex is a C++14-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, while making some of its own additions.
+Dxex is a C++11-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, while making some of its own additions.
 So while the feature set is not too comparable, the architectural design, core simulation algorithm, and optimizations, are highly similar.
@@
 In our case, a user can pass the simulation kernel the synchronization protocol to use for this model.
-Our implementation in C++14 also allows for optimizations which were plainly impossible in an interpreted language.
+Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language. Dxex will use new optimizations from the C++14 standard when possible.
 
 Since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
-This is due to dxex models being grafted on C++14, whereas PythonPDEVS models are grafted on Python.
+This is due to dxex models being grafted on C++11, whereas PythonPDEVS models are grafted on Python.
 
@@
 
-Reading and writing to shared memory is done through the use of the new C++14 synchronization primitives.
+Reading and writing to shared memory is done through the use of the new C++11 synchronization primitives.
 Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
-C++14 further allows us to make the implementation portable, as well as more efficient: the compiler might know of optimizations specific to atomic variables.
+C++11 further allows us to make the implementation portable, as well as more efficient: the compiler might know of optimizations specific to atomic variables or constant expressions which are heavily used in dxex.
 
@@
 This made us switch to the use of thread-local (using \texttt{tcmalloc}) memory pools.
-Again, we made use of specific new features of C++14, that were not available in Python, or even previous versions of the C++ language standard.
+Again, we made use of specific new features of C++11, that were not available in Python, or even previous versions of the C++ language standard.
 
@@
 We define simulation kernel transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
-User should thus only provide one model, implemented in C++14, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.
+User should thus only provide one model, implemented in C++11, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.
 Switching between simulation kernels is as simple as altering the simulation termination time.
@@
 \subsection{Features}
-In this section we highlight some of the extra features dxex offers to modellers.
+We will enumerate some of the extra features dxex offers to modellers.
 \subsubsection{Tracing}
-Like PythonPDEVS dxex has full tracing capabilities supporting a wide number of output formats : json, XML, cell, console, ...
+Like PythonPDEVS dxex has full tracing capabilities supporting a wide number of output formats : Json, XML, Cell, console, ...
 .
@@
 The tracing functionality itself is implemented asynchronous and thread-safe.
-By overriding the output functions in the Message or State classes the user can generate model specific output for use with an existing tracing output policy.
+By overriding the output functions in the Message or State classes the user can generate model specific output for use with an existing tracing outputpolicy.
 Tracing can be completely disabled to maximize simulation performance should this be desired (e.g.
@@
 \subsubsection{Memory Allocation}
-As we will demonstrate in \ref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel.
+As we will demonstrate in Section~\ref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel.
 Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation.

### 2016-06-24 publications/Journal/conclusion.tex
First set of revisions and fixes.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 Even when no cycles exist in the topology between kernels a good allocation scheme will minimize the number of connections between kernels and aim for a star topology as in PHoldTree with depth first allocation or a chain topology as in the Queue model.
-Dxex can optionally offer the user a visualization of the behavior of the kernels under a user supplied allocation scheme to allow for more insight. 
+Dxex can optionally offer the user a visualization of the behaviour of the kernels under a user supplied allocation scheme to allow for more insight. 
 The same instrumentation could in future be used to optimize an existing allocation scheme.
@@
 Optimistic is not constrained by this uncertainty and can offer a good speedup regardless of lookahead. 
-PHOLD and PholdTree demonstrate that a single model can benefit from different synchronization protocols depending on a single parameter.
+Phold and PholdTree demonstrate that a single model can benefit from different synchronization protocols depending on a single parameter.
 \subsubsection{Limits}

### 2016-06-24 publications/Journal/parallel-performance.tex
First set of revisions and fixes.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 
-	\includegraphics[width=\modelfraction\columnwidth, height=8cm, keepaspectratio]{fig/queue_allocation.eps}
+	\includegraphics[width=\modelfraction\columnwidth, height=8cm, keepaspectratio, angle=-90 ]{fig/queue_allocation.eps}
 	\caption{Queue model (d=2, w=7, t=5000, random timeadvance) allocation and simulation trace across 4 kernels.}
@@
 \paragraph*{Allocation}\label{PholdTreeallocation}
-The PholdTree benchmark can be configured to use 2 allocation schemes : breadth first and depth first.
-The breadth first allocation scheme will result in kernels that will form a dependency chain between them with multiple branches, much like in the Queue model.
-Such a linear dependency chain can result in a parallel speedup as we demonstrate with the Queue model, but this need not be true for all models as we will demonstrate in this section.
-A single kernel that has an unbalanced number of atomic models, models with unequal computation load in transition functions or activity will slow down the remainder of the chain.
+The PholdTree benchmark can be configured to use 2 allocation schemes: breadth first and depth first. 
+The breadth first allocation scheme will result in kernels that will form a dependency chain with multiple branches, much like in the Queue model. 
+Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model, but this is not always true as we will demonstrate in this section. 
+A single kernel that has an unbalanced number of atomic models or models with unequal computation load in transition functions will slow down the remainder of the chain.
 This effect is also apparent if the OS unfairly schedules the thread a kernel runs on.
@@
 We do not implement this function in our PholdTree variant for adevs since this would force us to implement state saving in the model code and not use the provided state saving functionality in the kernel, as is done in dxex or PythonPDEVS's optimistic.\\
-In dxex state saving is never required for a conservative simulation, this is handled transparently for the user who need not implement this behavior. 
+In dxex state saving is never required for a conservative simulation, this is handled transparently for the user who need not implement this behaviour. 
 By using the state saving technique inside the model code we feel we would no longer compare identical models across different synchronization algorithms, although we do not doubt adevs' increased performance when these functions are overridden.
@@
 % Conservative
-Conservative has a more nuanced speedup behavior in this benchmark.
+Conservative has a more nuanced speedup behaviour in this benchmark.
 We have detailed how a conservative kernel in dxex scales linear in the number of atomic models (if lookahead is $\epsilon$), this effect is more clearly visible here.
@@
 \subsection{Measuring Synchronization Overhead}
-The simulation kernels have a non trivial overhead compared to a sequential kernel.
-In this section we will try to isolate this overhead and highlight the consequences this has for applying a synchronization protocol in real life models.
+The simulation kernels have a non trivial overhead compared to a sequential kernel. In this section we will try to isolate this overhead. 
 Profiling indicates that there are in the most trivial of models only 2 causes for computational load outside of the actual simulation : memory management and random number generation.
+By reducing the cost of both we will isolate the synchronization overhead.
+We use the Queue benchmark since it is highly parallel and its computational load in transitioning is only determined by random number generation and event/state allocation.
 \paragraph*{Memory management}\label{sec:4-subsec:overhead-pgraph:memory}
@@
 The kernels use memory allocators backed by a thread-aware pooling library to reduce the performance impact of memory management as much as possible. 
-A sequential simulation kernel can leverage the invariant that no allocated event will persist beyond a single time advance and use a arena-style allocator.
-Conservative and optimistic simulation need to use generic pool allocators.
+A sequential simulation kernel no allocated event will persist beyond a single time advance,  this enables the use of an arena-style allocator.
+Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.
 \\
@@
 \end{figure}
-% Analyse

### 2016-06-24 publications/Journal/abstract.tex
Slight modification of abstract.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

--- 2015-12-11 paper/abstract.tex
@@
 Several synchronization protocols have been proposed, each with their distinct advantages and disadvantages.
-Due to the significantly different implementation of both protocols, most Parallel DEVS simulation tools are limited to only one such protocol.
+Due to the significantly different implementation of these protocols, most Parallel DEVS simulation tools are limited to only one such protocol.
 In this paper, we present a Parallel DEVS simulator, grafted on C++11, but based on PythonPDEVS, which supports both conservative and optimistic synchronization.
-We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol.
+% TODO if completed, add online switching
+We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol and demonstrate the inlfuence model allocation and uncertainty has on this choice. 
 Performance results are compared to adevs, in terms of CPU time and memory usage.

### 2016-06-24 publications/Journal/1-introduction.tex
More fixes, resolve the unnumbered referencing of sections.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 
+%Usage of Hyperref with hardcoded label is hardly ideal, but ~\ref doesn't work if the layout disables numbered sections, so a hack is needed.
+The alternative is \nameref but that isn't all that readable.
 The remainder of this paper is organized as follows:
-Section~\ref{sec:2-background} introduces the necessary background on synchronization protocols.
-Section~\ref{sec:3-features} elaborates on our design that enables this flexibility.
-In Section~\ref{sec:4-performance}, we evaluate performance of our tool by comparing its different synchronization protocols, and by comparing to adevs.
-Related work is discussed in Section~\ref{sec:5-related-work}.
-Section~\ref{sec:6-conclusion} concludes the paper and gives future work.
+Section \hyperref[sec:2-background]{2} introduces the necessary background on synchronization protocols.
+Section~\hyperref[sec:3-features]{3} elaborates on our design that enables this flexibility.
+In Section~\hyperref[sec:4-performance]{4}, we evaluate performance of our tool by comparing its different synchronization protocols, and by comparing to adevs.
+Related work is discussed in Section~\hyperref[sec:5-related-work]{5}.
+Section~\hyperref[sec:6-conclusion]{6} concludes the paper and gives future work.

### 2016-06-24 publications/Journal/3-features.tex
More fixes, resolve the unnumbered referencing of sections.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 \subsubsection{Memory Allocation}
-As we will demonstrate in Section~\ref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel.
+As we will demonstrate in ~\nameref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel.
 Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation.
@@
 It should be obvious that this is not sufficient in most simulations.
-In \ref{sec:4-performance} we analyze allocation schemes and their performance impact.
+In Section \nameref{sec:4-performance} we analyze allocation schemes and their performance impact.
 By overriding the default allocator a model author can tune the allocation scheme to maximize parallel speedup.
@@
 These figures demonstrate the effect of allocation visually, the high inter kernels event count will degrade performance, as we will demonstrate in
-\ref{PholdTreeallocation}.
-Dotted lines represent connections between models that did not see events exchanged, directed edges represents events with frequency as weight and color to distinguis inter versus intra kernel events.
+\nameref{PholdTreeallocation} .
+Dotted lines represent connections between models that did not see events exchanged, directed edges represents events with frequency as weight and color to distinguish inter versus intra kernel events.
 % Add figure
@@
 	\center
-	\includegraphics[width=\plotfraction\columnwidth, height=8cm, keepaspectratio]{fig/pholdtreed1n3t5000.eps}
+	\includegraphics[width=\plotfraction\columnwidth, height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000.eps}
 	\caption{Visualization of PholdTree (d=1,n=3,t=5000) sequential simulation.}
@@
 	\center
-	\includegraphics[width=\plotfraction\columnwidth,  height=8cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
+	\includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
 	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
@@
 	\center
-	\includegraphics[width=\plotfraction\columnwidth,  height=8cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
+	\includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
 	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}

### 2016-06-24 publications/Journal/parallel-performance.tex
More fixes, resolve the unnumbered referencing of sections.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 In Figures \ref{fig:pholdtree_visualize_parBFS} and \ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for both allocation schemes highlighting the remarks made in this paragraph.
-
-\begin{figure}
-	\center
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeBFS.eps}
-	\caption{PholdTree model breadth first allocation with 3 kernels.}
-	\label{fig:PholdTree_model_bfs}
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeDFS.eps}
-	\caption{PholdTree model depth first allocation with 3 kernels.}
-	\label{fig:PholdTree_model_dfs}
-\end{figure}
+%
+%\begin{figure}
+%	\center
+%	
+%	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeBFS.eps}
+%	\caption{PholdTree model breadth first allocation with 3 kernels.}
+%	\label{fig:PholdTree_model_bfs}
+%	
+%	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeDFS.eps}
+%	\caption{PholdTree model depth first allocation with 3 kernels.}
+%	\label{fig:PholdTree_model_dfs}
+%\end{figure}
 
@@
 From \ref{eq:pholdtreemodelcount} we know that increasing d will lead to a very rapid increase in models, we want to observe what happens in a closer to linear increase in model count per kernel, so we vary n.
-\\
 
 % Explain why adevs is not plotted (its x100 slower)
-
 Adevs' conservative implementation cannot handle the uncertainty (lack of non $\epsilon$ lookahead) here and is omitted from the comparison.
@@
 Upon investigation we conclude that this slowdown is caused by not implementing the \{begin/end\}lookahead() functions, which when not overridden in a model trigger an exception on each invocation.
-\\
 This exception handling completely stalls the kernel.
-We do not implement this function in our PholdTree variant for adevs since this would force us to implement state saving in the model code and not use the provided state saving functionality in the kernel, as is done in dxex or PythonPDEVS's optimistic.\\
+We do not implement this function in our PholdTree variant for adevs since this would force us to implement state saving in the model code and not use the provided state saving functionality in the kernel, as is done in dxex or PythonPDEVS's optimistic.
 In dxex state saving is never required for a conservative simulation, this is handled transparently for the user who need not implement this behaviour. 
 By using the state saving technique inside the model code we feel we would no longer compare identical models across different synchronization algorithms, although we do not doubt adevs' increased performance when these functions are overridden.
-\\
 
@@
 The probability parameter is kept at 0.1 for the same reason, it only induces a linear increase in load.
-\\
 

### 2016-06-27 publications/Journal/1-introduction.tex
Update abstract and introduction
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 The synchronization overhead is demonstrated by reducing the computational load of a model to near zero.
-Dxex also offers a visualization of the simulation with in depth statistics on the synchronization algorithms.
-With all this information the modeller can make the best informed decision on which synchronization algorithm to use.
 
@@
 
-% TODO uncomment if completed
-%Dxex offers online switching functionality that can be used by machine learning %techniques to automatically switch between the most optimal synchronzation protocol.
+% Switching and statistics
+Dxex offers visualization of the simulation and in depth statistics.
+A modeller can then make a more informed decision on which synchronization protocol to use or even intervene during simulation and request a switch between protocols.
 

### 2016-06-27 publications/Journal/abstract.tex
Update abstract and introduction
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 Due to the significantly different implementation of these protocols, most Parallel DEVS simulation tools are limited to only one such protocol.
-In this paper, we present a Parallel DEVS simulator, grafted on C++11, but based on PythonPDEVS, which supports both conservative and optimistic synchronization.
-% TODO if completed, add online switching
+In this paper, we present a Parallel DEVS simulator, grafted on C++11, but based on PythonPDEVS, which supports both conservative and optimistic synchronization protocols with the capability to switch during simulation between those protocols.
 We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol and demonstrate the inlfuence model allocation and uncertainty has on this choice. 

### 2016-06-27 publications/Journal/3-features.tex
Update section 3.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
-In our case, a user can pass the simulation kernel the synchronization protocol to use for this model.
+In our case, a user can pass the simulation kernel the synchronization protocol to use for this model, or even switch the synchronization protocol during runtime.
 Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language. Dxex will use new optimizations from the C++14 standard when possible.
@@
 Within the returned time interval, the model promises not to raise an event.
-A node aggregates this information to computes its earliest output time (EOT).
+A node aggregates this information to compute its earliest output time (EOT).
 This value is written out in shared memory, where it can be read out by all other nodes.
@@
 We will enumerate some of the extra features dxex offers to modellers.
+\subsubsection{Runtime synchronization switching}
+A user can request a runtime switch between synchronization protocols.
+A typical use case is when a significant slowdown is observed or memory usage reaches a certain treshold.
+The simulation engine will swap out the current kernels with new instances while preserving the state of the models and the simulation itself, and resume using the new synchronization protocol.
+There is no real limit on the number of switches, but the replacing and synchronizing of the kernels comes at a runtime cost that is dependent on the number of present kernels and the actual hardware the simulation runs on.
+A test case requires for 2 kernels an overhead approaching a second.
+This is an experimental but promising feature that can be used in an automated framework driven by machine learning techniques.
 \subsubsection{Tracing}
-Like PythonPDEVS dxex has full tracing capabilities supporting a wide number of output formats : Json, XML, Cell, console, ...
-.
-The user can implement his own outputpolicy and use it in our modular tracing design.
-The tracing functionality itself is implemented asynchronous and thread-safe.
-By overriding the output functions in the Message or State classes the user can generate model specific output for use with an existing tracing outputpolicy.
+Like PythonPDEVS dxex has full tracing capabilities supporting a wide number of output formats : Json, XML, Cell, console.
+The user can implement a new outputpolicy and use it in our modular tracing design.
+The tracing functionality itself is implemented asynchronous and thread-safe, the user does not need to use synchronization primitives.
+By overriding the output functions in the Message and State classes the user can generate model specific output for use with an existing tracing outputpolicy.
 Tracing can be completely disabled to maximize simulation performance should this be desired (e.g.
 benchmarks).
-% Talk about adevs?
 \subsubsection{Memory Allocation}
-As we will demonstrate in ~\nameref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel.
+As we will demonstrate in ~\nameref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel with optimistic synchronization.
 Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation.
-This allows us by using a single compilation parameter to change the allocation strategy per kernel type.
+This allows us by using a single compilation parameter to change the allocation strategy.
 This interface allows instrumentation of memory usage.
 Not only useful to detect memory leaks should this be a concern, it can also be of use to a model author to trace event creation in large models.
-A model author can rely on the default implementation as the fastest common configuration verified by the benchmarks.
+The default implementation is the fastest common configuration verified by the benchmarks used in this publication.
 \subsubsection{Model Allocation}
@@
 This is handled by the default model allocator.
-If no preference is given a simple striping scheme is followed.
-It should be obvious that this is not sufficient in most simulations.
-In Section \nameref{sec:4-performance} we analyze allocation schemes and their performance impact.
-By overriding the default allocator a model author can tune the allocation scheme to maximize parallel speedup.
+If no preference is given a simple striping scheme is followed but this is not sufficient in most simulations to achieve a speedup in parallel.
+In Section \nameref{sec:4-performance} we analyse the performance impact of allocation schemes.
+By overriding the default allocator a model author can tune the allocation scheme for a specific model to maximize parallel speedup.
 This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.
@@
 Dxex supplements this optionally with more in depth insights into the actual simulation.
-The allocation of models across kernels is visualized at the end of a simulation, with graphical representation of message frequency.
-This allows a user to fine tune future allocation schemes and the interface used by the visualization tool can also be used for more advanced purposes such as dynamic tracing of the simulation.\\
+The allocation of models across kernels is visualized at the end of a simulation along with the event trace of all events exchanged, including reverted events in optimistic. 
+A user can then fine tune future allocation schemes.
+
 Invaluable during the development of dxex were runtime statistics collected by a simulation kernel.
 These record any measurable event in the simulation such as : events created/destroyed, inter kernel event frequency, reverts, GVT calculations, lookahead and eot values and insights into the fairness between the kernels themselves.
-Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, or observe how balanced inter kernel message traffic actually is during a simulation.
+Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, observe how balanced inter kernel message traffic actually is during a simulation or measure the reverted simulation transition in optimistic synchronization.
 The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or allow activity tracking.

### 2016-06-27 publications/Journal/3-features.tex
Updating section 3 complete.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 There is no real limit on the number of switches, but the replacing and synchronizing of the kernels comes at a runtime cost that is dependent on the number of present kernels and the actual hardware the simulation runs on.
-A test case requires for 2 kernels an overhead approaching a second.
+In a test case with 2 kernels the overhead of a double runtime switch is less than a second.
 This is an experimental but promising feature that can be used in an automated framework driven by machine learning techniques.
@@
 \subsubsection{Memory Allocation}
-As we will demonstrate in ~\nameref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel with optimistic synchronization.
+As we will demonstrate in ~\nameref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulations with high event frequency, especially in parallel with optimistic synchronization.
 Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation.
@@
 Not only useful to detect memory leaks should this be a concern, it can also be of use to a model author to trace event creation in large models.
-The default implementation is the fastest common configuration verified by the benchmarks used in this publication.
+The default implementation is the fastest common configuration verified by the benchmarks in this publication.
 \subsubsection{Model Allocation}
@@
 \subsubsection{Logging}
-A high performance asynchronous buffered threaded logging module is used in dxex which offers a complete overview of the entire simulation.
+A high performance asynchronous buffered logging module is used in dxex which offers a complete overview of the entire simulation.
 The logging functionality can be used in any model, it only requires linking to the dxex library.
@@
 
-Invaluable during the development of dxex were runtime statistics collected by a simulation kernel.
-These record any measurable event in the simulation such as : events created/destroyed, inter kernel event frequency, reverts, GVT calculations, lookahead and eot values and insights into the fairness between the kernels themselves.
-Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, observe how balanced inter kernel message traffic actually is during a simulation or measure the reverted simulation transition in optimistic synchronization.
-The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or allow activity tracking.
+Dxex optionally collects runtime statistics for each kernel.
+These record any measurable event in the simulation such as : events created/destroyed, inter/intra kernel event frequency, reverts, GVT calculations, lookahead and eot values and insights into the fairness between the kernels themselves.
+Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, observe how balanced inter kernel message traffic actually is during a simulation or measure reverted transitions in optimistic synchronization.
+The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or provide feedback to drive synchronization switching.
 Figures \ref{fig:pholdtree_visualize_seq},\ref{fig:pholdtree_visualize_parDFS},\ref{fig:pholdtree_visualize_parBFS} give an example of the modular visualization output generated by dxex after simulation the PholdTree model.
-These figures demonstrate the effect of allocation visually, the high inter kernels event count will degrade performance, as we will demonstrate in
+These figures demonstrate the effect of allocation visually and show the high inter kernels event count that degrades performance, as we will demonstrate in
 \nameref{PholdTreeallocation} .
-Dotted lines represent connections between models that did not see events exchanged, directed edges represents events with frequency as weight and color to distinguish inter versus intra kernel events.
+Dotted lines represent connections between models where no events were exchanged, directed edges represents events with frequency as weight, coloring is used to distinguish inter versus intra kernel events.
 % Add figure

### 2016-06-27 publications/Journal/models.tex
Update Section 4 : model description.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 	\item The \textit{Queue} model, based on the \textit{HI} model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
-	A single generator pushes events into the queue, which are processed by the processors after a fixed or random delay.
+	A single generator pushes events into the queue, which are consumed by the processors after a fixed or random delay.
 	It takes two parameters: width and depth, which determine the width and depth of the hierarchy.
@@
 The model consists of recursively defined trees connected to each other, similar to the structure of an AST.
-All links can be configured to be bidirectional.
-	Unlike the \textit{Queue} model the hierarchical nature is still present in the topology of the atomic models after the directconnect stage.
+Connections can be unidirectional or bidirectional.
+	Unlike the \textit{Queue} model the width of the hierarchy is still present in the topology of the atomic models after the directconnect stage.
 This model allows us to investigate parallel speedup in several key dimensions in combination or isolation : model allocation, message density, uncertainty and hierarchy.
-The model has a depth parameter, fanout (childcount), probability of priority message and linkage direction.
+The model has the following parameters : depth, fanout (width), probability of priority message and linkage direction.
 We demonstrate the importance of atomic to kernel allocation, in this model this can be configured as a bread-first versus a depth-first scheme.
-	\textit{PHOLDTree}, while a synthetic benchmark, can be used to model gossiping in social networks, hierarchical resource sharing or load balancing.
-Unlike \textit{Queue} this model does not lends itself easily to parallelism and thus highlights the difficulties model authors face in the real world when employing PDEVS solutions.
+	\textit{PHOLDTree} can be used to model gossiping in social networks, hierarchical resource sharing or load balancing.
+Unlike \textit{Queue} this model does not lend itself easily to parallelism and thus highlights the difficulties model authors face in when a parallel speedup is desired for real world models.
+	The lookahead of an atomic node is $\epsilon$, simulating uncertainty as will be often the case in real world models.
 	The number of atomic models in this benchmark is given by 

### 2016-06-27 publications/Journal/parallel-performance.tex
Update sequentials subsection of section 4.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 \subsubsection{PholdTree}
-%move to pholdtree
 We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization.
@@
 %\end{figure}
+
+%Counterintuitively the probability of a high priority event in this model does not affect the performance difference between conservative and optimistic.
+The key parameter quickly becomes the load of a kernel in models.
+Our conservative implementation in an uncertain (trivial lookahead) simulation is very sensitive to a high load in models, whereas optimistic has no such limitation.
 

### 2016-06-27 publications/Journal/sequential-performance.tex
Update sequentials subsection of section 4.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.
-Analysis showed that this is caused by the high amount of exchanged events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
-Dxex reuses the same event type for all synchronization algorithms.
+Analysis showed that this is caused by the high amount of events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
 To shield the user from threading and deallocation concerns dxex provides an event superclass from which the user can derive to create a specialized event type.
-Copying and deallocation semantics are solved by the kernels at a runtime cost in simulations where event frequency is very high.
-Profiling the diverging benchmarks clearly shows the increasing cost of output generation (event creation) and deallocation as the determining factor in the gap in performance.
+Copying and deallocation semantics and tracing are solved by the kernels at a runtime cost in simulations where event frequency is very high.
+Profiling the benchmarks clearly shows the increasing cost of output generation and deallocation as the determining factor in the gap in performance.
 We refer the interested reader to the dxex \hyperref{https://bitbucket.org/bcardoen/devs-ex-machina}{}{repo}{repository} for the profiling call graphs for the different benchmarks.
@@
 \subsubsection{PHoldTree}
-The PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain as in Queue.
+PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain as in Queue.
 This topology is interesting since it highlights the effects of allocation which we have shown already in PHold and Interconnect to be of vital importance.
@@
 Since adevs does not use the directconnect algorithm, we expect some performance penalty between dxex and adevs.
-Conversely the messaging intensity in PHoldTree can be configured to highlight dxex's performance disadvantage in  benchmarks with a high event frequency as shown in the Interconnect benchmark.
+The event frequency in PHoldTree can be configured to highlight dxex's performance disadvantage here as shown in the Interconnect benchmark.
 \paragraph*{Event frequency}
-Similarly when the probability on a priority message (p) increases the amount of messages generated will increase and thus the runtime.
-In Figure \ref{fig:PHoldtree_seq_p_benchmark} we observe that the messaging probability has a near linear impact on performance, with the shape of the curve due to rounding errors.
+Similarly when the probability on a priority event (p) increases the amount of event generated will increase and thus the runtime.
+In Figure \ref{fig:PHoldtree_seq_p_benchmark} we observe that p has a near linear impact on performance, with the shape of the curve due to rounding errors.
 The advantage adevs has in simulations where event frequency is high has been demonstrated by the interconnect benchmark.
-Dxex chooses to provide the user with an event type where the user need not consider deallocation or copy semantics, at a cost in performance where event frequency is high.
-Counterintuitively the probability of a high priority event in this model does not affect the performance difference between conservative and optimistic.
-The key parameter quickly becomes the load of a kernel in models.
-Our conservative implementation in an uncertain (trivial lookahead) simulation is very sensitive to a high load in models, whereas optimistic has no such limitation.
 \paragraph*{Hierarchy}
-To establish a performance baseline for the parallel simulation benchmarks we instrument how the PHoldTree benchmark performs sequential when we vary each of the model's parameters. 
+To establish a baseline for the parallel simulation benchmarks we measure how the PHoldTree benchmark performs sequentially when we vary each of the model's parameters. 
 When we increase the depth and fanout of the model, we expect to see an increase in runtime.
-In Figure \ref{fig:PHoldtree_seq_dn_benchmark} the n parameter (fanout) determines the performance penalty adevs suffers w.r.t dxex but we need to look at the profiling call graphs to determine what causes adevs' performance graph to behave non linear.
-\\
-If we compare the instances d=2, n=2 with d=2, n=4 which show a clear deviation in performance between adevs and dxex for the latter, we conclude that the increase in width per subtree (n) leads to higher overhead for adevs.
-Dxex in contrast due to the directconnect algorithm has no such overhead.
-The intuition behind this effect, and especially why the d parameter does not have such an impact, is that an increase in n will rapidly increase the number of connections \ref{eq:pholdtreelinkcount} (and thus the routing problem adevs faces), whereas d will increase modelcount \ref{eq:pholdtreemodelcount} faster but have a lesser impact on connections.
+In Figure \ref{fig:PHoldtree_seq_dn_benchmark} the n parameter (fanout) determines the performance penalty adevs suffers compared to dxex.\\
+If we compare the instances d=2, n=2 with d=2, n=4, which show a clear difference in performance between adevs and dxex, we conclude from the profiling callgraphs that the increase in width per subtree (n) leads to higher overhead for adevs.
+Dxex in contrast uses the directconnect algorithm and has no such overhead.
+An increase in n will rapidly increase the number of connections \ref{eq:pholdtreelinkcount} (and thus the routing problem adevs faces), whereas d will increase modelcount \ref{eq:pholdtreemodelcount} faster but have a lesser impact on connections.
 Both kernels scale linear in increasing number of atomic models.

### 2016-06-27 publications/Journal/parallel-performance.tex
Parallel section updated
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 Kernel 2 is dependent on events from kernels 0 and 1 but by definition of the optimistic synchronization protocol it will simulate ahead without waiting for events it depends on.
-Any simulation it performs is immediately void and will have to be reverted as soon as events from kernels 0 and 1 arrive.
-When kernel 2 reverts, kernel 3 will inevitably have to revert as well since any events it received from kernel 2 are invalidated and marked as such by antimessages.
+Any simulation it performs will have to be reverted as soon as events from kernels 0 and 1 arrive.
+When kernel 2 reverts, kernel 3 will inevitably have to revert as well since most of the events it received from kernel 2 are invalidated and marked as such by antimessages.
 This leads to severe performance degradation, with an increasing probability of cascading reverts as the number of kernels and models per kernels increases.
@@
 All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
-In interconnect there is no allocation scheme possible that avoids cyclic dependencies between simulation kernels, as shown in the trace \ref{fig:interconnect_allocation_parallel} of a simulation with 4 models.
+In Interconnect there is no allocation scheme possible that avoids cyclic dependencies between simulation kernels, as shown in the trace \ref{fig:interconnect_allocation_parallel} of a simulation with 4 models.
 Such a cycle forces sequential operation of the kernels with no speedup possible.
@@
 Adevs is significantly slower during conservative synchronization.
-Analysis of profiling results shows that exception handling in adevs is the main cause. 
+Analysis of profiling callgraphs shows that exception handling in adevs is the main cause. 
 To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.
-These functions require the user to implement a state saving mechanism in contrast to PythonPDEVS and dxex's optimistic kernels which handle this inside the kernel.
+These functions require the user to implement a state saving in contrast to PythonPDEVS and dxex's optimistic kernels which handle this inside the kernel.
 We feel this would lead to an unfair comparison as we would like to keep the models synchronization-agnostic across all benchmarks.
@@
 This situation closely mimics the case made in the comparison between both synchronization algorithms by~\cite{FujimotoBook}.
-In Figure \ref{fig:phold_allocation} it is clear that in Phold it is possible for dependency cycles to form between kernels which as we have shown for Interconnect degrades performance for both optimistic and conservative.
+In Figure \ref{fig:phold_allocation} it is clear that in Phold it is possible for dependency cycles to form between kernels which as we have shown in Interconnect degrades performance for both optimistic and conservative.
 This is also the cause of the sublinear speedup observed in our Phold benchmark.
@@
 We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization.
-We will demonstrate that allocation is critical to achieve a parallel speedup and that the parameters of the model will give either conservative or optimistic an advantage.
+We will demonstrate that allocation is critical to achieve a parallel speedup and that the configuration of the model will give either conservative or optimistic an advantage.
 \paragraph*{Allocation}\label{PholdTreeallocation}
@@
 Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model, but this is not always true as we will demonstrate in this section. 
-A single kernel that has an unbalanced number of atomic models or models with unequal computation load in transition functions will slow down the remainder of the chain.
-This effect is also apparent if the OS unfairly schedules the thread a kernel runs on.
-In conservative this will lead to excessive polling of the other kernels' eot values, whereas in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel.
+A single kernel that has an unbalanced number of atomic models or unequal computation load in transition functions will slow down the remainder of the chain.
+This effect is also apparent if the thread a kernel runs on is not fairly scheduled.
+In conservative this will lead to excessive polling of the other kernels' eot values, in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel.
 In Figures \ref{fig:pholdtree_visualize_parBFS} and \ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for both allocation schemes highlighting the remarks made in this paragraph.
@@
 
-%Counterintuitively the probability of a high priority event in this model does not affect the performance difference between conservative and optimistic.
-The key parameter quickly becomes the load of a kernel in models.
-Our conservative implementation in an uncertain (trivial lookahead) simulation is very sensitive to a high load in models, whereas optimistic has no such limitation.
-
 \paragraph*{Strong Scaling}\label{pholdtreestrongscale}
@@
 The probability of the priority message has no extra impact on performance other than that shown in sequential performance.
-The difference between depth first and breadth first allocated kernels is striking, the first results in sublinear speedup for both synchronization variants. %reword 
+
+The probability of a high priority event in this model does not affect the performance difference between conservative and optimistic.
+The key parameter quickly becomes the load of a kernel in models.
+Our conservative implementation in an uncertain simulation is very sensitive to a high load in models, whereas optimistic has no such limitation.
+
+The difference between depth first and breadth first allocated kernels is striking, the first results in sublinear speedup for both synchronization protocols. %reword 
 The breadth first allocation scheme will lead to a very high number of inter-kernel connections, which is detrimental for any parallel synchronization algorithm.
 An event exchanged between kernels cannot be securely deallocated without a GVT algorithm and all the complexity this entails.
-Even an asynchronous GVT algorithm will involve in a shared memory architecture some form of synchronization, and this will slow down the kernels.
-\\
-Similar to the Queue benchmark the breadth first allocation scheme leads to a chain topology for the kernels, but PholdTree differs in the number of links between the kernels in the chain.
+Even a fast asynchronous GVT algorithm will require some form of inter kernel synchronization and span a timeframe during which allocated memory cannot be reused, forcing new allocations.
+Similar to the Queue benchmark the breadth first allocation scheme leads to a topology for the kernels resembling a chain but with more branches in the chain.
 In Queue there is only a single model on the edge of a kernel exchanging messages to a single other model in a neighbouring kernel, this is not the case in the PholdTree under breadth first allocation.
@@
 Depth first allocation will converge to a star topology which in this benchmark leads to a significant speedup.\\
-The performance drop observed for 2 kernels for all allocation schemes and synchronization algorithms is in part due to the higher link count between kernels.
-With more kernels this effect will be spread across more kernels and result in a relative lower performance penalty.
+The performance drop observed for 2 kernels for all allocation schemes and synchronization protocols is in part due to the higher link count between kernels.
+With more kernels these connections will be spread across more kernels and result in a relative lower performance penalty.
 \\
@@
 In this paragraph we investigate dxex's parallel performance when the the number of atomic models distributed over a fixed number of kernels varies.
-From \ref{eq:pholdtreemodelcount} we know that increasing d will lead to a very rapid increase in models, we want to observe what happens in a closer to linear increase in model count per kernel, so we vary n.
+From \ref{eq:pholdtreemodelcount} we know that increasing d will lead to a very rapid increase in models.
+We want to observe what happens in a closer to linear increase in model count per kernel by varying n.
 
@@
 The simulation kernels have a non trivial overhead compared to a sequential kernel. In this section we will try to isolate this overhead.
+
 Profiling indicates that there are in the most trivial of models only 2 causes for computational load outside of the actual simulation : memory management and random number generation.
@@
 The kernels use memory allocators backed by a thread-aware pooling library to reduce the performance impact of memory management as much as possible. 
-A sequential simulation kernel no allocated event will persist beyond a single time advance,  this enables the use of an arena-style allocator.
+In a sequential simulation kernel no allocated event will persist beyond a single time advance,  this enables the use of an arena-style allocator.
 Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.
@@
 Interconnect is a perfect test case since it generates each round a set of events quadratic in size to the number of atomic models in the simulation.
-With each event dynamically allocated this makes it the perfect stress test for any allocation strategy.
+With each event dynamically allocated, this makes it the perfect stress test for any allocation strategy.
 Both tcmalloc and Boost pools result in an advantage, but not necessarily combined.
 \\
-% Paralell
+% Parallel
 We next investigate the effect of both strategies in an optimistic simulation.
@@
 From the results it is clear that optimistic simulation benefits greatly from the use of tcmalloc, regardless of the allocator.
-Nonetheless the pool allocator still reduce the allocation overhead.
+Nonetheless the pool allocator still reduces the allocation overhead.
 \\
@@
 Both the pools and tcmalloc will try to keep memory allocated to them by the OS, this gives a somewhat misleading view in statistics collected from the OS.
-The dxex kernels themselves and the pool interfaces allow the user to instrument event allocation and deallocation exactly, useful for example in observing how efficient GVT-triggered deallocation is in practice.\\
+The dxex kernels themselves and the pool interfaces allow the user to instrument event allocation and deallocation exactly which can be useful in measuring how efficient GVT-triggered deallocation is in practice.\\
 %Conclude pgraph
@@
 \paragraph*{Random Number Generators}
-A non trivial amount of time in a simulation spent waiting for a random number generator can mask synchronization overhead.
+A non trivial amount of time in a simulation spent waiting for a random number generator (rng) can mask synchronization overhead.
 Several of our benchmarks use random number generators to test the effect of uncertainty in a simulation, either by random time advances or random destinations for events.
@@
 The size of the rng object in the state triggers memory overhead but it is interesting to see that adevs' conservative is also sensitive to this effect even though in theory in conservative synchronization no state copying/saving is required.
-Dxex's conservative kernel is insensitive to storing the rng object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation. 
-The overhead of initially doing so is therefore constant and not a factor in long simulations.\\
+Dxex's conservative kernel is insensitive to storing the rng object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation. \\
 Replacing the STL rng with the threading optimized trng we are more accurately able to isolate synchronization overhead.
-Figure \ref{fig:Queuerngspeedup} shows that both dxex and adevs in sequential simulation gain a ~ 3 speedup by using the faster random number generator.
-The parallel kernels have smaller speedup with adevs' conservative and dxex optimistic achieving an almost identical factor.
+Figure \ref{fig:Queuerngspeedup} shows that both dxex and adevs in sequential simulation gain a factor ~3 speedup by using the faster random number generator.
+The parallel kernels have a smaller speedup with adevs' conservative and dxex optimistic achieving an almost identical factor.
 Dxex conservative is almost insensitive to the changing of the rng.
-Dxex optimistic and adevs conservative are in the default configuration spend relatively more time waiting for random numbers than dxex conservative, demonstrated by the speedup gained when random number generation is accelerated.
+Dxex optimistic and adevs conservative in the default configuration spend relatively more time waiting for random numbers than dxex conservative, demonstrated by the speedup gained when random number generation is accelerated.
 The actual synchronization overhead in all parallel kernels still dominates the cost of random number generation which is seen from the speedup difference between parallel and sequential.\\

### 2016-06-27 publications/Journal/conclusion.tex
Finalize revising section 4.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
-We will summarize the key aspects in achieving good parallel performance here. % reword aspects syn
+We will summarize the key factors in achieving good parallel performance here. % reword aspects syn
 \subsubsection{Allocation}
@@
 In the case of the Interconnect model it is not possible to allocate models without introducing a cyclic dependency between kernels with severe performance degradation as a result. 
-Even when no cycles exist in the topology between kernels a good allocation scheme will minimize the number of connections between kernels and aim for a star topology as in PHoldTree with depth first allocation or a chain topology as in the Queue model.
+Even when no cycles exist in the topology between kernels a good allocation scheme will minimize the number of connections between kernels, for example the star topology in PHoldTree with depth first allocation or the chain topology in the Queue model.
 Dxex can optionally offer the user a visualization of the behaviour of the kernels under a user supplied allocation scheme to allow for more insight. 
@@
 A simulation where future behaviour cannot be predicted is typically not suited for conservative simulation, but we have demonstrated that dxex's conservative kernel can still offer a non trivial speedup in this context.
-Optimistic is not constrained by this uncertainty and can offer a good speedup regardless of lookahead. 
+Optimistic is not limited by this uncertainty and can offer a good speedup regardless of lookahead. 
 Phold and PholdTree demonstrate that a single model can benefit from different synchronization protocols depending on a single parameter.

### 2016-06-27 publications/Journal/6-conclusions.tex
First set of revisions and fixes.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
Conclusion.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 Faster GVT implementations, such as those presented by~\cite{Fujimoto:1997:CGV:268403.268404} and~\cite{Bauer:2005:SND:1069810.1070159}, might further help to minimize this problem.
-Secondly, the idea of activity can be implemented for our simulation kernels, making it possible to dynamically switch between conservative and optimistic synchronization when behavioural changes are detected.
+Secondly, the runtime switching between synchronization protocols can be driven using machine learning techniques. The simulation engine is already capable of collecting data to inform such a process.
 Thirdly, activity algorithms, as already implemented by PythonPDEVS, can also be implemented in dxex, to determine how they influence simulation performance.
 Finally automatic allocation is possible by analysis of the connections between models.
-This information is already used in dxex to determine the blocking order for conservative synchronization and in the directconnect algorithm.
-A graph algorithm that distributes models while avoiding cycles in the resulting kernels topology could be used to maximize parallel speedup in either optimistic or conservative synchronization.
+This information is already used in dxex to construct the dependency graph in conservative synchronization.
+A graph algorithm that distributes models while avoiding cycles in the resulting kernel topology could be used to offer a parallel speedup in either optimistic or conservative synchronization, or serve as a default parallel allocation scheme that can be improved by the user.

### 2016-06-27 publications/Journal/abstract.tex
* fixed typo in abstract * some very minute changes.

@@
 In this paper, we present a Parallel DEVS simulator, grafted on C++11, but based on PythonPDEVS, which supports both conservative and optimistic synchronization protocols with the capability to switch during simulation between those protocols.
-We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol and demonstrate the inlfuence model allocation and uncertainty has on this choice. 
+We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol and demonstrate the influence model allocation and uncertainty has on this choice. 
 Performance results are compared to adevs, in terms of CPU time and memory usage.

### 2016-06-27 publications/Journal/parallel-performance.tex
* fixed typo in abstract * some very minute changes.

@@
 The kernels use memory allocators backed by a thread-aware pooling library to reduce the performance impact of memory management as much as possible. 
-In a sequential simulation kernel no allocated event will persist beyond a single time advance,  this enables the use of an arena-style allocator.
+In a sequential simulation kernel no allocated event will persist beyond a single time advance, allowing the use of an arena-style allocator.
 Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.
@@
 % Implementation choices
-Dxex uses Boost Pool\cite{boostpool} allocators in parallel simulation kernels, and can use arena-style allocators for sequential simulation.
+Dxex uses Boost Pool\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
 The latter can be faster, but at the cost of extra configuration.

### 2016-06-28 publications/Journal/3-features.tex
Shuffling around some sections

--- 2016-06-24 publications/Journal/3-features.tex
@@
 
-\subsection{Transparency}
-We define simulation kernel transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
+\subsection{Synchronization Protocol Transparency}
+We define synchronization protocol transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
 User should thus only provide one model, implemented in C++11, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.
@@
 Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
-
-\subsection{Features}
-We will enumerate some of the extra features dxex offers to modellers.
-\subsubsection{Tracing}
-Like PythonPDEVS dxex has full tracing capabilities supporting a wide number of output formats : Json, XML, Cell, console, ...
-.
-The user can implement his own outputpolicy and use it in our modular tracing design.
-The tracing functionality itself is implemented asynchronous and thread-safe.
-By overriding the output functions in the Message or State classes the user can generate model specific output for use with an existing tracing outputpolicy.
-Tracing can be completely disabled to maximize simulation performance should this be desired (e.g.
-benchmarks).
-% Talk about adevs?
-\subsubsection{Memory Allocation}
-As we will demonstrate in ~\nameref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel.
-Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation.
-This allows us by using a single compilation parameter to change the allocation strategy per kernel type.
-This interface allows instrumentation of memory usage.
-Not only useful to detect memory leaks should this be a concern, it can also be of use to a model author to trace event creation in large models.
-A model author can rely on the default implementation as the fastest common configuration verified by the benchmarks.
-\subsubsection{Model Allocation}
-A model author can specify which kernel a model should be allocated to, should such manual intervention be required.
-This is handled by the default model allocator.
-If no preference is given a simple striping scheme is followed.
-It should be obvious that this is not sufficient in most simulations.
-In Section \nameref{sec:4-performance} we analyze allocation schemes and their performance impact.
-By overriding the default allocator a model author can tune the allocation scheme to maximize parallel speedup.
-This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.
-\subsubsection{Logging}
-A high performance asynchronous buffered threaded logging module is used in dxex which offers a complete overview of the entire simulation.
-The logging functionality can be used in any model, it only requires linking to the dxex library.
-Logging statements are written to file with a granular level system.
-In case of a severe crash the logging system will ensure persistence of the log file itself up until the last statement.
-Logging can be completely disabled to remove the performance impact.
-\subsubsection{Visualization and Statistics}
-Tracing offers a black box view of a DEVS model in a simulation, with transitions, events and states recorded at each state change.
-Dxex supplements this optionally with more in depth insights into the actual simulation.
-The allocation of models across kernels is visualized at the end of a simulation, with graphical representation of message frequency.
-This allows a user to fine tune future allocation schemes and the interface used by the visualization tool can also be used for more advanced purposes such as dynamic tracing of the simulation.\\
-Invaluable during the development of dxex were runtime statistics collected by a simulation kernel.
-These record any measurable event in the simulation such as : events created/destroyed, inter kernel event frequency, reverts, GVT calculations, lookahead and eot values and insights into the fairness between the kernels themselves.
-Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, or observe how balanced inter kernel message traffic actually is during a simulation.
-The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or allow activity tracking.
-Figures \ref{fig:pholdtree_visualize_seq},\ref{fig:pholdtree_visualize_parDFS},\ref{fig:pholdtree_visualize_parBFS} give an example of the modular visualization output generated by dxex after simulation the PholdTree model.
-These figures demonstrate the effect of allocation visually, the high inter kernels event count will degrade performance, as we will demonstrate in
-\nameref{PholdTreeallocation} .
-Dotted lines represent connections between models that did not see events exchanged, directed edges represents events with frequency as weight and color to distinguish inter versus intra kernel events.
-% Add figure
-\begin{figure}
-	\center
-	\includegraphics[width=\plotfraction\columnwidth, height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000.eps}
-	\caption{Visualization of PholdTree (d=1,n=3,t=5000) sequential simulation.}
-	\label{fig:pholdtree_visualize_seq}
-\end{figure}
-\begin{figure}
-	\center
-	\includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
-	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
-	\label{fig:pholdtree_visualize_parBFS}
-\end{figure}
-\begin{figure}
-	\center
-	\includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
-	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
-	\label{fig:pholdtree_visualize_parDFS}
-\end{figure}

### 2016-06-28 publications/Journal/4a-allocation.tex
Shuffling around some sections

uthor can specify which kernel a model should be allocated to, should such manual intervention be required. This is handled by the default model allocator. If no preference is given a simple striping scheme is followed but this is not sufficient in most simulations to achieve a speedup in parallel. In Section \nameref{sec:4-performance} we analyse the performance impact of allocation schemes. By overriding the default allocator a model author can tune the allocation scheme for a specific model to maximize parallel speedup. This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.

### 2016-06-28 publications/Journal/4b-hotswap.tex
Shuffling around some sections

A user can request a runtime switch between synchronization protocols. A typical use case is when a significant slowdown is observed or memory usage reaches a certain treshold. The simulation engine will swap out the current kernels with new instances while preserving the state of the models and the simulation itself, and resume using the new synchronization protocol. There is no real limit on the number of switches, but the replacing and synchronizing of the kernels comes at a runtime cost that is dependent on the number of present kernels and the actual hardware the simulation runs on. In a test case with 2 kernels the overhead of a double runtime switch is less than a second.
This is an experimental but promising feature that can be used in an automated framework driven by machine learning techniques.

%TODO combination with statistics gathering
Dxex optionally collects runtime statistics for each kernel. These record any measurable event in the simulation such as : events created/destroyed, inter/intra kernel event frequency, reverts, GVT calculations, lookahead and eot values and insights into the fairness between the kernels themselves. Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, observe how balanced inter kernel message traffic actually is during a simulation or measure reverted transitions in optimistic synchronization.
The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or provide feedback to drive synchronization switching.

The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or provide feedback to drive synchronization switching.
Figures \ref{fig:pholdtree_visualize_seq},\ref{fig:pholdtree_visualize_parDFS},\ref{fig:pholdtree_visualize_parBFS} give an example of the modular visualization output generated by dxex after simulation the PholdTree model. These figures demonstrate the effect of allocation visually and show the high inter kernels event count that degrades performance, as we will demonstrate in
\nameref{PholdTreeallocation} . Dotted lines represent connections between models where no events were exchanged, directed edges represents events with frequency as weight, coloring is used to distinguish inter versus intra kernel events.
% Add figure
\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth, height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000.eps}
    \caption{Visualization of PholdTree (d=1,n=3,t=5000) sequential simulation.}
    \label{fig:pholdtree_visualize_seq}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
    \label{fig:pholdtree_visualize_parBFS}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
    \label{fig:pholdtree_visualize_parDFS}
\end{figure}

### 2016-06-28 publications/Journal/3-features.tex
New order of sections

@@
 
-In the remainder of this section, we will elaborate on our prominent new feature: support for multiple synchronization protocols within the same simulation tool, which are offered transparently to the model.
+In the remainder of this section, we will elaborate on our prominent new feature: the efficient implementation of multiple synchronization protocols within the same simulation tool, which are offered transparently to the model.
 
@@
 Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
+
+\subsection{Increasing Parallelism}
+The simulation kernels have a non trivial overhead compared to a sequential kernel. In this section we will try to isolate this overhead.
+
+Profiling indicates that there are in the most trivial of models only 2 causes for computational load outside of the actual simulation : memory management and random number generation.
+By reducing the cost of both we will isolate the synchronization overhead.
+We use the Queue benchmark since it is highly parallel and its computational load in transitioning is only determined by random number generation and event/state allocation.
+
+\subsubsection{Memory Management}
+\label{sec:4-subsec:overhead-pgraph:memory}
+\begin{figure}
+    \center
+
+    \includegraphics[width=\modelfraction\columnwidth]{fig/memory_allocators.eps}
+    \caption{Effect of memory allocators on sequential runtime.}
+    \label{fig:memallocators}
+
+    \includegraphics[width=\modelfraction\columnwidth]{fig/memory_allocators_parallel.eps}
+    \caption{Effect of memory allocators on parallel runtime.}
+    \label{fig:memallocators_parallel}
+\end{figure}
+In dxex a model author is given access to automatic memory management for events and states.
+A model written for a sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the model author) the (de)allocation semantics of events or states.
+\\
+In sequential and conservative simulation the kernel gives the guarantee that no state will be copied.
+Dxex's simulation control hides the implementation details of this management, a model author can focus on the actual semantics of the model.
+The kernels use memory allocators backed by a thread-aware pooling library to reduce the performance impact of memory management as much as possible.
+In a sequential simulation kernel no allocated event will persist beyond a single time advance, allowing the use of an arena-style allocator.
+Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.
+\\
+An important observation to make is that intra-kernel events can be pooled more aggressively, whereas inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
+A simulation with high inter kernel events will suffer a performance hit, whereas the impact of high intra kernel events can be optimized using arena allocators.\\
+The allocation interfaces are instrumented in dxex to trace memory usage in a simulation and can serve as a basic memory leak detector.
+Shielding the user from (de)allocation complexity comes at a cost in performance, as the comparison with adevs in simulations with a high event frequency \ref{fig:Interconnect_benchmark} of events shows.\\
+\\
+Dxex uses Boost Pool\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
+The latter can be faster, but at the cost of extra configuration.
+The allocators are supplemented by the library tcmalloc \cite{tcmalloc}.
+In Figure \ref{fig:memallocators} the Interconnect model is benchmarked under the different combinations of allocation strategies.
+\\
+Interconnect is a perfect test case since it generates each round a set of events quadratic in size to the number of atomic models in the simulation.
+With each event dynamically allocated, this makes it the perfect stress test for any allocation strategy.
+Both tcmalloc and Boost pools result in an advantage, but not necessarily combined.
+\\
+We next investigate the effect of both strategies in an optimistic simulation.
+We have already demonstrated that conservative simulation differs little in memory usage from sequential, so we only benchmark the optimistic kernel.
+From the results it is clear that optimistic simulation benefits greatly from the use of tcmalloc, regardless of the allocator.
+Nonetheless the pool allocator still reduces the allocation overhead.
+\\
+Both the pools and tcmalloc will try to keep memory allocated to them by the OS, this gives a somewhat misleading view in statistics collected from the OS.
+The dxex kernels themselves and the pool interfaces allow the user to instrument event allocation and deallocation exactly which can be useful in measuring how efficient GVT-triggered deallocation is in practice.\\
+In conclusion both techniques are required to reduce the overhead of memory allocations in dxex, and on by default.
+
+\subsubsection{Random Number Generators}
+A non trivial amount of time in a simulation spent waiting for a random number generator (rng) can mask synchronization overhead.
+Several of our benchmarks use random number generators to test the effect of uncertainty in a simulation, either by random time advances or random destinations for events.
+\\
+In a parallel simulation we also need to guarantee isolation between the random calls to avoid excessive synchronization on the rng object itself.
+Dxex uses the Tina random number generator collection (trng) \cite{PhysRevE.75.066701} as an alternative random number generator as it is designed with performance and multithreading usage in mind.
+We also wrote a new version of our models that stores the rng object in the state instead of calling a shared thread local object to isolate the synchronization overhead.\\
+From Figure \ref{fig:Queuerngspeedup} we see that storing the rng in the state is a very expensive operation for the default STL random number generator.
+The size of the rng object in the state triggers memory overhead but it is interesting to see that adevs' conservative is also sensitive to this effect even though in theory in conservative synchronization no state copying/saving is required.
+Dxex's conservative kernel is insensitive to storing the rng object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation. \\
+Replacing the STL rng with the threading optimized trng we are more accurately able to isolate synchronization overhead.
+Figure \ref{fig:Queuerngspeedup} shows that both dxex and adevs in sequential simulation gain a factor ~3 speedup by using the faster random number generator.
+The parallel kernels have a smaller speedup with adevs' conservative and dxex optimistic achieving an almost identical factor.
+Dxex conservative is almost insensitive to the changing of the rng.
+Dxex optimistic and adevs conservative in the default configuration spend relatively more time waiting for random numbers than dxex conservative, demonstrated by the speedup gained when random number generation is accelerated.
+The actual synchronization overhead in all parallel kernels still dominates the cost of random number generation which is seen from the speedup difference between parallel and sequential.\\
+\begin{figure}
+    \center
+
+    \includegraphics[width=\modelfraction\columnwidth]{fig/rngspeedupeffectdevstone.eps}
+    \caption{Speedup with different rng usage patterns in Queue model}
+    \label{fig:Queuerngspeedup}
+
+\end{figure}

### 2016-06-28 publications/Journal/parallel-performance.tex
New order of sections

@@
 \end{figure}
-
-\subsection{Measuring Synchronization Overhead}
-The simulation kernels have a non trivial overhead compared to a sequential kernel. In this section we will try to isolate this overhead.
-
-Profiling indicates that there are in the most trivial of models only 2 causes for computational load outside of the actual simulation : memory management and random number generation.
-By reducing the cost of both we will isolate the synchronization overhead.
-We use the Queue benchmark since it is highly parallel and its computational load in transitioning is only determined by random number generation and event/state allocation.
-\paragraph*{Memory management}\label{sec:4-subsec:overhead-pgraph:memory}
-\begin{figure}
-	\center
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/memincrease.eps}
-	\caption{Memory usage in parallel synchronization}
-	\label{fig:memincrease}
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/memory_allocators.eps}
-	\caption{Effect of memory allocators on sequential runtime.}
-	\label{fig:memallocators}
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/memory_allocators_parallel.eps}
-	\caption{Effect of memory allocators on parallel runtime.}
-	\label{fig:memallocators_parallel}
-\end{figure}
-In dxex a model author is given access to automatic memory management for events and states.
-A model written for a sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the model author) the (de)allocation semantics of events or states.
-\\
-In sequential and conservative simulation the kernel gives the guarantee that no state will be copied.
-Dxex's simulation control hides the implementation details of this management, a model author can focus on the actual semantics of the model.
-The kernels use memory allocators backed by a thread-aware pooling library to reduce the performance impact of memory management as much as possible. 
-In a sequential simulation kernel no allocated event will persist beyond a single time advance, allowing the use of an arena-style allocator.
-Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.
-\\
-An important observation to make is that intra-kernel events can be pooled more aggressively, whereas inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
-A simulation with high inter kernel events will suffer a performance hit, whereas the impact of high intra kernel events can be optimized using arena allocators.\\
-The allocation interfaces are instrumented in dxex to trace memory usage in a simulation and can serve as a basic memory leak detector.
-Shielding the user from (de)allocation complexity comes at a cost in performance, as the comparison with adevs in simulations with a high event frequency \ref{fig:Interconnect_benchmark} of events shows.\\
-Figure \ref{fig:memincrease} shows an experiment where we measure peak memory usage (maximum resident set size (kB)) for each synchronization protocol and determine how the memory usage differs with respect to sequential simulation.
-Both conservative kernels require only a small constant increase in memory usage.
-\\
-% Implementation choices
-Dxex uses Boost Pool\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
-The latter can be faster, but at the cost of extra configuration.
-The allocators are supplemented by the library tcmalloc \cite{tcmalloc}.
-In Figure \ref{fig:memallocators} the Interconnect model is benchmarked under the different combinations of allocation strategies.
-\\
-% Sequential
-Interconnect is a perfect test case since it generates each round a set of events quadratic in size to the number of atomic models in the simulation.
-With each event dynamically allocated, this makes it the perfect stress test for any allocation strategy.
-Both tcmalloc and Boost pools result in an advantage, but not necessarily combined.
-\\
-% Parallel
-We next investigate the effect of both strategies in an optimistic simulation.
-We have already demonstrated that conservative simulation differs little in memory usage from sequential, so we only benchmark the optimistic kernel.
-From the results it is clear that optimistic simulation benefits greatly from the use of tcmalloc, regardless of the allocator.
-Nonetheless the pool allocator still reduces the allocation overhead.
-\\
-% Interpretation note
-Both the pools and tcmalloc will try to keep memory allocated to them by the OS, this gives a somewhat misleading view in statistics collected from the OS.
-The dxex kernels themselves and the pool interfaces allow the user to instrument event allocation and deallocation exactly which can be useful in measuring how efficient GVT-triggered deallocation is in practice.\\
-%Conclude pgraph
-In conclusion both techniques are required to reduce the overhead of memory allocations in dxex, and on by default.
-\paragraph*{Random Number Generators}
-A non trivial amount of time in a simulation spent waiting for a random number generator (rng) can mask synchronization overhead.
-Several of our benchmarks use random number generators to test the effect of uncertainty in a simulation, either by random time advances or random destinations for events.
-\\
-In a parallel simulation we also need to guarantee isolation between the random calls to avoid excessive synchronization on the rng object itself. 
-Dxex uses the Tina random number generator collection (trng) \cite{PhysRevE.75.066701} as an alternative random number generator as it is designed with performance and multithreading usage in mind. 
-We also wrote a new version of our models that stores the rng object in the state instead of calling a shared thread local object to isolate the synchronization overhead.\\
-From Figure \ref{fig:Queuerngspeedup} we see that storing the rng in the state is a very expensive operation for the default STL random number generator.
-The size of the rng object in the state triggers memory overhead but it is interesting to see that adevs' conservative is also sensitive to this effect even though in theory in conservative synchronization no state copying/saving is required.
-Dxex's conservative kernel is insensitive to storing the rng object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation. \\
-Replacing the STL rng with the threading optimized trng we are more accurately able to isolate synchronization overhead.
-Figure \ref{fig:Queuerngspeedup} shows that both dxex and adevs in sequential simulation gain a factor ~3 speedup by using the faster random number generator.
-The parallel kernels have a smaller speedup with adevs' conservative and dxex optimistic achieving an almost identical factor.
-Dxex conservative is almost insensitive to the changing of the rng.
-Dxex optimistic and adevs conservative in the default configuration spend relatively more time waiting for random numbers than dxex conservative, demonstrated by the speedup gained when random number generation is accelerated.
-The actual synchronization overhead in all parallel kernels still dominates the cost of random number generation which is seen from the speedup difference between parallel and sequential.\\
-\begin{figure}
-	\center
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/rngspeedupeffectdevstone.eps}
-	\caption{Speedup with different rng usage patterns in Queue model}
-	\label{fig:Queuerngspeedup}
-
-\end{figure}

### 2016-06-28 publications/Journal/3-features.tex
New form for the explanation of core principles of dxex

--- 2016-06-28 publications/Journal/3-features.tex
@@
-\newcommand{\plotfraction}{0.85}
 Historically, dxex is based on PythonPDEVS~\cite{PythonPDEVS}.
@@
 Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
-This made us switch to the use of thread-local (using \texttt{tcmalloc}) memory pools.
+This made us switch to the use of thread-local (using \textit{tcmalloc}) memory pools.
 Again, we made use of specific new features of C++11, that were not available in Python, or even previous versions of the C++ language standard.
@@
 Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
+
+\subsection{Increasing Parallelism}
+The goal of our contribution is to increase simulation performance as much as possible, leveraging parallel computation in the process.
+Parallelizing the simulation kernel goes further, however, than merely implementing the different synchronization protocols.
+
+After implementing all synchronization protocols, performance was still not within acceptable levels.
+Profiling showed that the problem was caused by two issues: memory management and random number generation.
+For both, it is already known that they can have significantly impact on parallelizability of code: they introduce sequential blocks.
+In the remainder of this section, we explain our modifications to each and the difference in performance this gave us.
+
+\subsubsection{Memory Management}
+\label{sec:4-subsec:overhead-pgraph:memory}
+\begin{figure}
+    \center
+    \includegraphics[width=\modelfraction\columnwidth]{fig/memory_allocators_parallel.eps}
+    \caption{Effect of memory allocators on parallel execution time.}
+    \label{fig:memallocators_parallel}
+\end{figure}
+
+%TODO briefly explain this by refering to "what every computer programmer should know"
+
+In dxex a model author is given access to automatic memory management for events and states.
+A model written for a sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the model author) the (de)allocation semantics of events or states.
+
+In sequential and conservative simulation the kernel gives the guarantee that no state will be copied.
+Dxex's simulation control hides the implementation details of this management, a model author can focus on the actual semantics of the model.
+The kernels use memory allocators backed by a thread-aware pooling library to reduce the performance impact of memory management as much as possible.
+In a sequential simulation kernel no allocated event will persist beyond a single time advance, allowing the use of an arena-style allocator.
+Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.
+
+An important observation to make is that intra-kernel events can be pooled more aggressively, whereas inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
+A simulation with high inter kernel events will suffer a performance hit, whereas the impact of high intra kernel events can be optimized using arena allocators.
+
+The allocation interfaces are instrumented in dxex to trace memory usage in a simulation.
+Shielding the user from (de)allocation complexity comes at a cost in performance, as is later shown in the performance evaluation.
+
+Dxex uses Boost Pool~\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
+The latter can be faster, but at the cost of extra configuration.
+The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}.
+
+We primarily investigate this for optimistic simulation, as this is the most memory consuming mode of simulation~\cite{Fujimoto}.
+Simulation execution times with all four combinations are shown in Figure~\ref{fig:memallocators_parallel}.
+Optimistic simulation greatly benefits from the use of \textit{tcmalloc}, regardless of the allocator.
+Nonetheless the pool allocator also reduces the allocation overhead, though only by a relatively small fraction.
+
+Both pools and \textit{tcmalloc} try to keep memory allocated to them by the OS, this gives a somewhat misleading view in statistics collected from the OS.
+The dxex kernels themselves and the pool interfaces allow the user to instrument event allocation and deallocation exactly which can be useful in measuring how efficient GVT-triggered deallocation is in practice.
+
+In conclusion both techniques are required to reduce the overhead of memory allocations in dxex, and on by default.
+
+\subsubsection{Random Number Generators}
+%TODO explain about sequentialized random numbers
+A non-trivial amount of time in a simulation is spent waiting for a Random Number Generator (RNG) can mask synchronization overhead.
+Several of our benchmarks use random number generators to test the effect of uncertainty in a simulation, either by random time advances or random destinations for events.
+
+In a parallel simulation we also need to guarantee isolation between the random calls to avoid excessive synchronization on the RNG object itself.
+Dxex uses the Tina RNG collection (TRNG)~\cite{PhysRevE.75.066701} as an alternative random number generator as it is designed with performance and multithreading usage in mind.
+We also wrote a new version of our models that stores the RNG object in the state instead of calling a shared thread local object to isolate the synchronization overhead.
+
+From Figure~\ref{fig:Queuerngspeedup} we see that storing the RNG in the state is very expensive for the default STL random number generator.
+%TODO fill in values
+This is primarily caused by the significant difference in size: X bytes for the STL random number generator, and Y bytes for the Tina random number generator.
+Dxex's sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation.
+The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.
+
+Replacing the STL RNG with the threading optimized TRNG we are more accurately able to isolate synchronization overhead.
+Figure~\ref{fig:Queuerngspeedup} shows that dxex in sequential simulation gets three times faster by using TRNG.
+The parallel kernels have a smaller speedup.
+Dxex conservative is almost insensitive to the changing of the RNG, though a slight increase in performance can be noted.
+Dxex optimistic in the default configuration spends relatively more time waiting for random numbers than dxex conservative.
+%TODO why does it spend more time there?
+This is confirmed by the speedup gained when random number generation is accelerated.
+Synchronization overhead is still the limiting factor in performance, as seen by the big speedup gap between sequential and parallel simulation.
+
+\begin{figure}
+    \center
+
+    \includegraphics[width=\modelfraction\columnwidth]{fig/rngspeedupeffectdevstone.eps}
+    \caption{Speedup with different RNG usage patterns.}
+    \label{fig:Queuerngspeedup}
+\end{figure}

### 2016-06-28 publications/Journal/4a-allocation.tex
New order of sections
Moved PHOLDTree to a seperate section where it can better cover its contribution

@@
-uthor can specify which kernel a model should be allocated to, should such manual intervention be required.
+\section{Model Allocation}
+%TODO explain model allocation
+%TODO why model allocation
+Modeller can specify which kernel a model should be allocated to, should such manual intervention be required.
 This is handled by the default model allocator.
@@
 This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.
+
+\subsection{Performance Evaluation}
+%TODO do basic performance evaluation of allocation
+%TODO show model in statistics view
+
+The \textit{PHoldTree} model, a hierarchical model based on the semantics of PHOLD~\cite{PHOLD}.
+The model consists of recursively defined trees connected to each other, similar to the structure of an AST.
+Connections can be unidirectional or bidirectional.
+Unlike the \textit{Queue} model the width of the hierarchy is still present in the topology of the atomic models after the directconnect stage.
+This model allows us to investigate parallel speedup in several key dimensions in combination or isolation : model allocation, message density, uncertainty and hierarchy.
+The model has the following parameters : depth, fanout (width), probability of priority message and linkage direction.
+We demonstrate the importance of atomic to kernel allocation, in this model this can be configured as a bread-first versus a depth-first scheme.
+\textit{PHOLDTree} can be used to model gossiping in social networks, hierarchical resource sharing or load balancing.
+Unlike \textit{Queue} this model does not lend itself easily to parallelism and thus highlights the difficulties model authors face in when a parallel speedup is desired for real world models.
+The lookahead of an atomic node is $\epsilon$, simulating uncertainty as will be often the case in real world models.
+
+\begin{figure}
+\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree.pdf}
+\caption{PHOLDTree model for depth 1, width 2.}
+\label{fig:PHOLDTree_model}
+\end{figure}
+
+PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain as in Queue.
+This topology is interesting since it highlights the effects of allocation which we have shown already in PHold and Interconnect to be of vital importance.
+This model allows us to investigate in depth the effects of non-cyclic allocation strategies and measure parallel speedup.
+
+Since adevs does not use the directconnect algorithm, we expect some performance penalty between dxex and adevs.
+The event frequency in PHoldTree can be configured to highlight dxex's performance disadvantage here as shown in the Interconnect benchmark.
+
+\paragraph*{Event frequency}
+Similarly when the probability on a priority event (p) increases the amount of event generated will increase and thus the runtime.
+In Figure~\ref{fig:PHoldtree_seq_p_benchmark} we observe that p has a near linear impact on performance, with the shape of the curve due to rounding errors.
+The advantage adevs has in simulations where event frequency is high has been demonstrated by the interconnect benchmark.
+
+\paragraph*{Hierarchy}
+To establish a baseline for the parallel simulation benchmarks we measure how the PHoldTree benchmark performs sequentially when we vary each of the model's parameters.
+When we increase the depth and fanout of the model, we expect to see an increase in runtime.
+In Figure~\ref{fig:PHoldtree_seq_dn_benchmark} the n parameter (fanout) determines the performance penalty adevs suffers compared to dxex.
+
+If we compare the instances d=2, n=2 with d=2, n=4, which show a clear difference in performance between adevs and dxex, we conclude from the profiling callgraphs that the increase in width per subtree (n) leads to higher overhead for adevs.
+Dxex in contrast uses the directconnect algorithm and has no such overhead.
+An increase in n will rapidly increase the number of connections \ref{eq:pholdtreelinkcount} (and thus the routing problem adevs faces), whereas d will increase modelcount \ref{eq:pholdtreemodelcount} faster but have a lesser impact on connections.
+Both kernels scale linear in increasing number of atomic models.
+
+\begin{figure}
+    \center
+    \includegraphics[width=\plotfraction\columnwidth]{fig/pholdtree_sequential_p.eps}
+    \caption{PHoldTree for increasing probability of priority message.}
+    \label{fig:PHoldtree_seq_p_benchmark}
+\end{figure}
+% Img for p increasing
+\begin{figure}
+    \center
+    \includegraphics[width=\plotfraction\columnwidth]{fig/pholdtree_sequential_dn.eps}
+    \caption{PHoldTree : Effect of hierarchy.}
+    \label{fig:PHoldtree_seq_dn_benchmark}
+\end{figure}
+
+We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization.
+We will demonstrate that allocation is critical to achieve a parallel speedup and that the configuration of the model will give either conservative or optimistic an advantage.
+\paragraph*{Allocation}\label{PholdTreeallocation}
+The PholdTree benchmark can be configured to use 2 allocation schemes: breadth first and depth first.
+The breadth first allocation scheme will result in kernels that will form a dependency chain with multiple branches, much like in the Queue model.
+Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model, but this is not always true as we will demonstrate in this section.
+A single kernel that has an unbalanced number of atomic models or unequal computation load in transition functions will slow down the remainder of the chain.
+This effect is also apparent if the thread a kernel runs on is not fairly scheduled.
+In conservative this will lead to excessive polling of the other kernels' eot values, in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel.
+In Figures \ref{fig:pholdtree_visualize_parBFS} and \ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for both allocation schemes highlighting the remarks made in this paragraph.
+%
+%\begin{figure}
+%   \center
+%   
+%   \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeBFS.eps}
+%   \caption{PholdTree model breadth first allocation with 3 kernels.}
+%   \label{fig:PholdTree_model_bfs}
+%   
+%   \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeDFS.eps}
+%   \caption{PholdTree model depth first allocation with 3 kernels.}
+%   \label{fig:PholdTree_model_dfs}
+%\end{figure}
+
+\paragraph*{Strong Scaling}\label{pholdtreestrongscale}
+In Figure \ref{fig:PholdTree_plot_alloc_high} we see that the difference in performance for all 4 kernel configurations compared to that shown in Figure \ref{fig:PholdTree_plot_alloc_low} is a constant factor.
+The probability of the priority message has no extra impact on performance other than that shown in sequential performance.
+
+The probability of a high priority event in this model does not affect the performance difference between conservative and optimistic.
+The key parameter quickly becomes the load of a kernel in models.
+Our conservative implementation in an uncertain simulation is very sensitive to a high load in models, whereas optimistic has no such limitation.
+
+The difference between depth first and breadth first allocated kernels is striking, the first results in sublinear speedup for both synchronization protocols. %reword 
+The breadth first allocation scheme will lead to a very high number of inter-kernel connections, which is detrimental for any parallel synchronization algorithm.
+An event exchanged between kernels cannot be securely deallocated without a GVT algorithm and all the complexity this entails.
+Even a fast asynchronous GVT algorithm will require some form of inter kernel synchronization and span a timeframe during which allocated memory cannot be reused, forcing new allocations.
+Similar to the Queue benchmark the breadth first allocation scheme leads to a topology for the kernels resembling a chain but with more branches in the chain.
+In Queue there is only a single model on the edge of a kernel exchanging messages to a single other model in a neighbouring kernel, this is not the case in the PholdTree under breadth first allocation.
+This explains the difference in speedup between Queue and PholdTree despite the similarity in kernel topology.\\
+Depth first allocation still has a higher inter kernel connection count, but not on the same order as breadth first allocated PholdTree.
+Depth first allocation will converge to a star topology which in this benchmark leads to a significant speedup.\\
+The performance drop observed for 2 kernels for all allocation schemes and synchronization protocols is in part due to the higher link count between kernels.
+With more kernels these connections will be spread across more kernels and result in a relative lower performance penalty.
+\\
+Conservative synchronization suffers a further penalty in this benchmark when there are only a few kernels.
+The PholdTree model has a lookahead of $\epsilon$ leading the kernel to query each allocated atomic model for a lookahead value on each simulation round.
+As the atomic models are spread over more kernels this effect is reduced leading to an increase in performance.
+With 4 kernels conservative surpasses optimistic in speedup.
+This is clear evidence that conservative synchronization is a good candidate for parallel simulation even in simulations with uncertainty.
+\begin{figure}
+    \center
+
+    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreealloclowp.eps}
+    \caption{PholdTree model performance under different allocation schemes with low message probability}
+    \label{fig:PholdTree_plot_alloc_low}
+
+    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeallochighp.eps}
+    \caption{PholdTree model performance under different allocation schemes with high message probability}
+    \label{fig:PholdTree_plot_alloc_high}
+\end{figure}
+
+\paragraph*{Weak Scaling}
+%Intro
+In this paragraph we investigate dxex's parallel performance when the the number of atomic models distributed over a fixed number of kernels varies.
+From \ref{eq:pholdtreemodelcount} we know that increasing d will lead to a very rapid increase in models.
+We want to observe what happens in a closer to linear increase in model count per kernel by varying n.
+
+% Explain why adevs is not plotted (its x100 slower)
+Adevs' conservative implementation cannot handle the uncertainty (lack of non $\epsilon$ lookahead) here and is omitted from the comparison.
+Its parallel performance is two orders of magnitude slower than the sequential implementation.
+Upon investigation we conclude that this slowdown is caused by not implementing the \{begin/end\}lookahead() functions, which when not overridden in a model trigger an exception on each invocation.
+This exception handling completely stalls the kernel.
+We do not implement this function in our PholdTree variant for adevs since this would force us to implement state saving in the model code and not use the provided state saving functionality in the kernel, as is done in dxex or PythonPDEVS's optimistic.
+In dxex state saving is never required for a conservative simulation, this is handled transparently for the user who need not implement this behaviour.
+By using the state saving technique inside the model code we feel we would no longer compare identical models across different synchronization algorithms, although we do not doubt adevs' increased performance when these functions are overridden.
+
+% Optimistic
+In Figure \ref{fig:PholdTree_plot_weaknopt} we observe that dxex's optimistic kernels are slightly sensitive to the number of atomic models allocated to them.
+In the worst case with 2 kernels there is no real speedup observable, as soon as the number of kernels increases we see a converging speedup trend, that except for 32 kernels is almost constant despite the increase in n.
+Note that only depth first allocated kernels are measured here, breadth first as we can see in \ref{fig:PholdTree_plot_alloc_high} has no speedup advantage.
+The probability parameter is kept at 0.1 for the same reason, it only induces a linear increase in load.
+
+% Conservative
+Conservative has a more nuanced speedup behaviour in this benchmark.
+We have detailed how a conservative kernel in dxex scales linear in the number of atomic models (if lookahead is $\epsilon$), this effect is more clearly visible here.
+\\
+It is clear that as the fanout (n) of the model increases the kernel performance degrades rapidly.
+The intersection of the performance graph of each configuration is shifted to the right, the actual 1-speedup point increases as the kernelcount increases.
+For a d=4, n=5 configuration with 4 kernels each kernel has a load of ~1300 atomic models when it crosses the 1-speedup line.
+\\
+An 8 kernel configuration with d=4 n=6 can sustain a load of ~2700 atomic models before it reaches that point.
+From the results we see that the 32-kernel configuration is not yet slowing down with the current parameters.
+\\
+The modelcount is only a part of the explanation of this effect, as the kernelcount increases the amount of inter kernel messages will be split into more distinct sets which can be handled with more concurrency in dxex's architecture.
+That this effect is not always true can be observed from the d=4 n=3 32-kernel datapoint.
+Note that in this configuration 341 atomic models are distributed over 32 kernels, a relative low model load can more easily highlight synchronization overhead, even if allocation is optimal.\\
+As with optimistic we do not include breadth first allocated benchmarks in this speedup plot since none of those achieve a speedup higher than 1.
+\\
+If we combine both to determine which synchronization protocol is more optimal we see in Figure \ref{fig:PholdTree_plot_weakall} that for these configurations conservative with 8 kernels is an ideal configuration up until n=5, where the modelcount starts to degrade performance.
+Optimistic with the same kernelcount is almost insensitive to the increase in modelcount and is therefore a more robust choice.
+Note how optimistic and conservative with 32 kernels at this point still have a reasonable speedup which is surprising given the synchronization overhead in such a large set of kernels.
+\begin{figure}
+    \center
+
+    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingnopt.eps}
+    \caption{PholdTree model weak scaling under varying fanout and optimistic synchronization}
+    \label{fig:PholdTree_plot_weaknopt}
+
+    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingncon.eps}
+    \caption{PholdTree model weak scaling under varying fanout and conservative synchronization}
+    \label{fig:PholdTree_plot_weakncon}
+
+    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingall.eps}
+    \caption{PholdTree model weak scaling under varying fanout and different synchronization algorithms}
+    \label{fig:PholdTree_plot_weakall}
+\end{figure}
+
+\subsection{Linking Back}
+%TODO link back to initial performance evaluation

### 2016-06-28 publications/Journal/models.tex
New order of sections
Moved PHOLDTree to a seperate section where it can better cover its contribution

@@
 	An example for four models is shown in Figure~\ref{fig:interconnect_model}.
-
-	\item The \textit{PHoldTree} model, a hierarchical model based on the semantics of PHOLD~\cite{PHOLD}.
-The model consists of recursively defined trees connected to each other, similar to the structure of an AST.
-Connections can be unidirectional or bidirectional.
-	Unlike the \textit{Queue} model the width of the hierarchy is still present in the topology of the atomic models after the directconnect stage.
-This model allows us to investigate parallel speedup in several key dimensions in combination or isolation : model allocation, message density, uncertainty and hierarchy.
-The model has the following parameters : depth, fanout (width), probability of priority message and linkage direction.
-We demonstrate the importance of atomic to kernel allocation, in this model this can be configured as a bread-first versus a depth-first scheme.
-	\textit{PHOLDTree} can be used to model gossiping in social networks, hierarchical resource sharing or load balancing.
-Unlike \textit{Queue} this model does not lend itself easily to parallelism and thus highlights the difficulties model authors face in when a parallel speedup is desired for real world models.
-	The lookahead of an atomic node is $\epsilon$, simulating uncertainty as will be often the case in real world models.
-	The number of atomic models in this benchmark is given by 
-	\begin{equation} \label{eq:pholdtreemodelcount}
-	\frac{n^{d+2}-1}{n-1}
-	\end{equation}
-	Whereas the number of links between the models is given by
-	\begin{equation} \label{eq:pholdtreelinkcount}
-	(2n-1)\sum_{i=0}^{d}{n^i}
-	\end{equation}
-	%Expand on how this allows us to profile all parameters of d-stone/phold/iconnect.
 \end{enumerate}
-
-\newcommand{\modelfraction}{0.65}
-\newcommand{\betweenmodels}{0.8cm}
 
@@
 	\label{fig:PHOLD_model}
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree.pdf}
-	\caption{PHOLDTree model for depth 1, width 2.}
-	\label{fig:PHOLDTree_model}
 \end{figure}

### 2016-06-28 publications/Journal/parallel-performance.tex
Moved PHOLDTree to a seperate section where it can better cover its contribution

@@
 Results show that there is no single perfect synchronization algorithm for this model: depending on configuration, either synchronization protocol might be better.
-
-% The real life
-\subsubsection{PholdTree}
-We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization.
-We will demonstrate that allocation is critical to achieve a parallel speedup and that the configuration of the model will give either conservative or optimistic an advantage.
-\paragraph*{Allocation}\label{PholdTreeallocation}
-The PholdTree benchmark can be configured to use 2 allocation schemes: breadth first and depth first. 
-The breadth first allocation scheme will result in kernels that will form a dependency chain with multiple branches, much like in the Queue model. 
-Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model, but this is not always true as we will demonstrate in this section. 
-A single kernel that has an unbalanced number of atomic models or unequal computation load in transition functions will slow down the remainder of the chain.
-This effect is also apparent if the thread a kernel runs on is not fairly scheduled.
-In conservative this will lead to excessive polling of the other kernels' eot values, in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel.
-In Figures \ref{fig:pholdtree_visualize_parBFS} and \ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for both allocation schemes highlighting the remarks made in this paragraph.
-%
-%\begin{figure}
-%	\center
-%	
-%	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeBFS.eps}
-%	\caption{PholdTree model breadth first allocation with 3 kernels.}
-%	\label{fig:PholdTree_model_bfs}
-%	
-%	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeDFS.eps}
-%	\caption{PholdTree model depth first allocation with 3 kernels.}
-%	\label{fig:PholdTree_model_dfs}
-%\end{figure}
-
-\paragraph*{Strong Scaling}\label{pholdtreestrongscale}
-In Figure \ref{fig:PholdTree_plot_alloc_high} we see that the difference in performance for all 4 kernel configurations compared to that shown in Figure \ref{fig:PholdTree_plot_alloc_low} is a constant factor.
-The probability of the priority message has no extra impact on performance other than that shown in sequential performance.
-
-The probability of a high priority event in this model does not affect the performance difference between conservative and optimistic.
-The key parameter quickly becomes the load of a kernel in models.
-Our conservative implementation in an uncertain simulation is very sensitive to a high load in models, whereas optimistic has no such limitation.
-
-The difference between depth first and breadth first allocated kernels is striking, the first results in sublinear speedup for both synchronization protocols. %reword 
-The breadth first allocation scheme will lead to a very high number of inter-kernel connections, which is detrimental for any parallel synchronization algorithm.
-An event exchanged between kernels cannot be securely deallocated without a GVT algorithm and all the complexity this entails.
-Even a fast asynchronous GVT algorithm will require some form of inter kernel synchronization and span a timeframe during which allocated memory cannot be reused, forcing new allocations.
-Similar to the Queue benchmark the breadth first allocation scheme leads to a topology for the kernels resembling a chain but with more branches in the chain.
-In Queue there is only a single model on the edge of a kernel exchanging messages to a single other model in a neighbouring kernel, this is not the case in the PholdTree under breadth first allocation.
-This explains the difference in speedup between Queue and PholdTree despite the similarity in kernel topology.\\
-Depth first allocation still has a higher inter kernel connection count, but not on the same order as breadth first allocated PholdTree.
-Depth first allocation will converge to a star topology which in this benchmark leads to a significant speedup.\\
-The performance drop observed for 2 kernels for all allocation schemes and synchronization protocols is in part due to the higher link count between kernels.
-With more kernels these connections will be spread across more kernels and result in a relative lower performance penalty.
-\\
-Conservative synchronization suffers a further penalty in this benchmark when there are only a few kernels.
-The PholdTree model has a lookahead of $\epsilon$ leading the kernel to query each allocated atomic model for a lookahead value on each simulation round.
-As the atomic models are spread over more kernels this effect is reduced leading to an increase in performance.
-With 4 kernels conservative surpasses optimistic in speedup.
-This is clear evidence that conservative synchronization is a good candidate for parallel simulation even in simulations with uncertainty.
-\begin{figure}
-	\center
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreealloclowp.eps}
-	\caption{PholdTree model performance under different allocation schemes with low message probability}
-	\label{fig:PholdTree_plot_alloc_low}
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeallochighp.eps}
-	\caption{PholdTree model performance under different allocation schemes with high message probability}
-	\label{fig:PholdTree_plot_alloc_high}
-\end{figure}
-
-\paragraph*{Weak Scaling}
-%Intro
-In this paragraph we investigate dxex's parallel performance when the the number of atomic models distributed over a fixed number of kernels varies.
-From \ref{eq:pholdtreemodelcount} we know that increasing d will lead to a very rapid increase in models.
-We want to observe what happens in a closer to linear increase in model count per kernel by varying n.
-
-% Explain why adevs is not plotted (its x100 slower)
-Adevs' conservative implementation cannot handle the uncertainty (lack of non $\epsilon$ lookahead) here and is omitted from the comparison.
-Its parallel performance is two orders of magnitude slower than the sequential implementation.
-Upon investigation we conclude that this slowdown is caused by not implementing the \{begin/end\}lookahead() functions, which when not overridden in a model trigger an exception on each invocation.
-This exception handling completely stalls the kernel.
-We do not implement this function in our PholdTree variant for adevs since this would force us to implement state saving in the model code and not use the provided state saving functionality in the kernel, as is done in dxex or PythonPDEVS's optimistic.
-In dxex state saving is never required for a conservative simulation, this is handled transparently for the user who need not implement this behaviour. 
-By using the state saving technique inside the model code we feel we would no longer compare identical models across different synchronization algorithms, although we do not doubt adevs' increased performance when these functions are overridden.
-
-% Optimistic
-In Figure \ref{fig:PholdTree_plot_weaknopt} we observe that dxex's optimistic kernels are slightly sensitive to the number of atomic models allocated to them.
-In the worst case with 2 kernels there is no real speedup observable, as soon as the number of kernels increases we see a converging speedup trend, that except for 32 kernels is almost constant despite the increase in n.
-Note that only depth first allocated kernels are measured here, breadth first as we can see in \ref{fig:PholdTree_plot_alloc_high} has no speedup advantage.
-The probability parameter is kept at 0.1 for the same reason, it only induces a linear increase in load.
-
-% Conservative
-Conservative has a more nuanced speedup behaviour in this benchmark.
-We have detailed how a conservative kernel in dxex scales linear in the number of atomic models (if lookahead is $\epsilon$), this effect is more clearly visible here.
-\\
-It is clear that as the fanout (n) of the model increases the kernel performance degrades rapidly.
-The intersection of the performance graph of each configuration is shifted to the right, the actual 1-speedup point increases as the kernelcount increases.
-For a d=4, n=5 configuration with 4 kernels each kernel has a load of ~1300 atomic models when it crosses the 1-speedup line.
-\\
-An 8 kernel configuration with d=4 n=6 can sustain a load of ~2700 atomic models before it reaches that point.
-From the results we see that the 32-kernel configuration is not yet slowing down with the current parameters.
-\\
-The modelcount is only a part of the explanation of this effect, as the kernelcount increases the amount of inter kernel messages will be split into more distinct sets which can be handled with more concurrency in dxex's architecture.
-That this effect is not always true can be observed from the d=4 n=3 32-kernel datapoint.
-Note that in this configuration 341 atomic models are distributed over 32 kernels, a relative low model load can more easily highlight synchronization overhead, even if allocation is optimal.\\
-As with optimistic we do not include breadth first allocated benchmarks in this speedup plot since none of those achieve a speedup higher than 1.
-\\
-If we combine both to determine which synchronization protocol is more optimal we see in Figure \ref{fig:PholdTree_plot_weakall} that for these configurations conservative with 8 kernels is an ideal configuration up until n=5, where the modelcount starts to degrade performance.
-Optimistic with the same kernelcount is almost insensitive to the increase in modelcount and is therefore a more robust choice.
-Note how optimistic and conservative with 32 kernels at this point still have a reasonable speedup which is surprising given the synchronization overhead in such a large set of kernels.
-\begin{figure}
-	\center
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingnopt.eps}
-	\caption{PholdTree model weak scaling under varying fanout and optimistic synchronization}
-	\label{fig:PholdTree_plot_weaknopt}
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingncon.eps}
-	\caption{PholdTree model weak scaling under varying fanout and conservative synchronization}
-	\label{fig:PholdTree_plot_weakncon}
-
-	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingall.eps}
-	\caption{PholdTree model weak scaling under varying fanout and different synchronization algorithms}
-	\label{fig:PholdTree_plot_weakall}
-
-\end{figure}

### 2016-06-28 publications/Journal/3-features.tex
Finished restructure of section 3

--- 2016-06-28 publications/Journal/3-features.tex
@@
 Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
+
+\subsection{Increasing Parallelism}
+The goal of our contribution is to increase simulation performance as much as possible, leveraging parallel computation in the process.
+Parallelizing the simulation kernel goes further, however, than merely implementing the different synchronization protocols.
+
+We observed that after implementing all synchronization protocols, performance was still not within acceptable levels.
+Profiling revealed that most of the overhead was caused by two issues: memory management and random number generation.
+For both, it is already known that they can have significantly impact on parallelizability of code, since they introduce sequential blocks.
+Both were tackled using approaches that are in common use in the parallel programming world.
+We briefly mention how the application of these techniques influences our performance.
+
+\subsubsection{Memory Management}
+\label{sec:4-subsec:overhead-pgraph:memory}
+\begin{figure}
+    \center
+    \includegraphics[width=\modelfraction\columnwidth]{fig/memory_allocators_parallel.eps}
+    \caption{Effect of memory allocators on parallel execution time.}
+    \label{fig:memallocators_parallel}
+\end{figure}
+Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{WhatEveryComputerProgrammerShouldKnowAboutMemory}: memory bandwidth doesn't increase as fast as the number of cores using it.
+While this is always a problem, it is aggravated in dxex by providing automatic memory management for events and states.
+A model written for a sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the model author) the (de)allocation semantics of events or states.
+
+Furthermore, allocating and deallocating memory by making calls to the operating system, as typically done by calls such as \texttt{malloc}, happens sequentially.
+To counter this, our memory allocators are backed by a thread-aware pooling library.
+In a sequential simulation kernel no allocated event will persist beyond a single time advance, even allowing the use of an arena-style allocator.
+Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.
+
+%TODO what does it mean to "pool aggressively?"
+Intra-kernel events are pooled aggressively, whereas inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
+A simulation with many inter kernel events suffers a performance hit, whereas the impact of many intra kernel events can be minimized using arena allocators.
+
+Dxex uses Boost Pool~\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
+The latter can be faster, but at the cost of extra configuration.
+The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}, which reduces lock contention in \texttt{malloc} calls.
+
+We primarily investigate this for optimistic simulation, as this is the most memory consuming mode of simulation~\cite{Fujimoto}.
+Simulation execution times for all four combinations are shown in Figure~\ref{fig:memallocators_parallel}.
+Optimistic simulation greatly benefits from the use of \textit{tcmalloc}, regardless of the allocator.
+Nonetheless the pool allocator also reduces the allocation overhead, though only by a relatively small fraction.
+Both techniques are required to reduce the overhead of memory allocations in dxex, and on by default.
+
+Both pools and \textit{tcmalloc} try to keep memory allocated instead of returning it to the Operating System (OS).
+As a result, the OS will usually report memory consumption that is higher than the actual amount of stored data.
+
+\subsubsection{Random Number Generators}
+Random Number Generators (RNG) are another aspect of the program that results in sequentialization.
+All accesses to the RNG will result in the modification of a global (\textit{i.e.}, shared between threads) variable.
+This easily becomes a bottleneck in simulation, since random numbers are a common occurence in simulation~\cite{?}.
+As such, a non-trivial amount of time in a simulation is often spent waiting for an RNG.
+
+We of course still need to guarantee determinism and isolation between the calls to the RNG, as well as avoiding excessive synchronization.
+Dxex uses the Tina RNG collection (TRNG)~\cite{PhysRevE.75.066701} as an alternative random number generator with performance and multithreading in mind.
+Since the RNG is an implicit part of the state in the DEVS formalism, though often not implemented as such, we evaluated performance for both approaches: one global RNG per thread, and one RNG per atomic DEVS model.
+
+We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default STL random number generator.
+%TODO fill in values
+This is primarily caused by the significant difference in size: X bytes for the STL random number generator, and Y bytes for the Tina random number generator.
+Dxex's sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation.
+The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.
+
+Figure~\ref{fig:Queuerngspeedup} shows that dxex in sequential simulation gets three times faster by using TRNG, than when using the STL RNG.
+For parallel simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between sequential and parallel simulation.
+Conservative synchronization is almost insensitive to the changing of the RNG, though a slight increase in performance can be noted.
+Optimistic synchronization gets much slower when the RNG becomes part of the model state, since the state needs to be copied as well.
+This becomes a significant overhead when using the STL RNG, since performance plummets to a fraction of the original.
+Using TRNG avoids this problem completely, as the size of the RNG state is negligible.
+
+\begin{figure}
+    \center
+
+    \includegraphics[width=\modelfraction\columnwidth]{fig/rngspeedupeffectdevstone.eps}
+    \caption{Speedup with different RNG usage patterns.}
+    \label{fig:Queuerngspeedup}
+\end{figure}

### 2016-06-28 publications/Journal/3-features.tex
Added note for the figure
Performance conclusions to introduce the extra contribution of the paper

@@
 %TODO fill in values
-This is primarily caused by the significant difference in size: X bytes for the STL random number generator, and Y bytes for the Tina random number generator.
+This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
 Dxex's sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation.
@@
     \center
-
+    %TODO remove adevs results as not applicable here
     \includegraphics[width=\modelfraction\columnwidth]{fig/rngspeedupeffectdevstone.eps}

### 2016-06-28 publications/Journal/4b-hotswap.tex
New order of sections
New form for the explanation of core principles of dxex
Deduplicate the section title

@@
+%TODO explain hotswapping
+%TODO why hotswapping
+Simply because a synchronization protocol is ideal at the start of the simulation, does not mean that it will still be ideal during the simulation.
+It has been frequently shown, and repeated in the previous section, that model behaviour significantly influences the ideal synchronization protocol.
+Contrary to many modelling formalisms, the DEVS formalisms makes it possible to model basically any kind of discrete event model.
+As such, it is possible for the model to significantly change its behaviour throughout the simulation.
+
+Defining the ideal synchronization protocol at the start of the simulation, when information about future model behaviour might be scarce, might therefore not offer the best possible performance.
+
+\subsection{Extensibility}
+%TODO now purely manual
+%TODO framework provided to do this using a decider
+%TODO possibly using machine learning techniques
+
+\subsection{Statistics Gathering}
+%TODO traditionally only access to normal simulation state
+%TODO for machine learning might give access to simulator details
+%TODO such as ...
+
+\subsubsection{Visualization of Communication}
+%TODO explain simple visualizer
+%TODO can be used by hotswapper
+
 A user can request a runtime switch between synchronization protocols.

### 2016-06-28 publications/Journal/4a-allocation.tex
Deduplicate the section title
Restructure and write about hotswap component with statistics gathering

@@
-\section{Model Allocation}
 %TODO explain model allocation
@@
 %TODO link back to initial performance evaluation
+
+\begin{figure}
+    \center
+    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
+    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
+    \label{fig:pholdtree_visualize_parBFS}
+\end{figure}
+\begin{figure}
+    \center
+    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
+    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
+    \label{fig:pholdtree_visualize_parDFS}
+\end{figure}

### 2016-06-28 publications/Journal/4b-hotswap.tex
Restructure and write about hotswap component with statistics gathering

@@
 Defining the ideal synchronization protocol at the start of the simulation, when information about future model behaviour might be scarce, might therefore not offer the best possible performance.
+In dxex, we not only make it possible to define the synchronization protocol to use, but also to change this decision throughout simulation.
+To do this, all kernels are notified of the switch and they are forced to stop simulation using the current synchronization core.
+When stopped, each kernel instantiates a new core that is provided with the simulation state of the previous core.
+Simulation is then resumed with the new cores after the previous ones are destroyed.
 
-\subsection{Extensibility}
-%TODO now purely manual
-%TODO framework provided to do this using a decider
-%TODO possibly using machine learning techniques
+As usual, switching imposes an overhead and should thus only be done if the benefits outweigh the induced overhead.
+This overhead depends on the size of the model and the number of simulation cores.
+For a simple model and a few cores, the overhead is less than a single second.
+
+Although we currently only support manual switches between different synchronization protocols, this is not necessarily the case.
+Ideally, a new component is added to the simulation kernel, which monitors model behaviour and simulation performance, and toggles between them automatically.
+Our interface is thus augmented with the necessary bindings for such a component to be defined.
+Also, our interface is augmented with an interface for statistics gathering and model behaviour analysis.
+The implementation of such a component is currently left open: algorithms can be heavily based on machine learning or similar approaches.
 
 \subsection{Statistics Gathering}
-%TODO traditionally only access to normal simulation state
-%TODO for machine learning might give access to simulator details
-%TODO such as ...
+Traditionally, models are not exposed to simulation kernel details due to the wrong level of abstraction.
+Simulation models only care about being simulated, and not about how this is being done.
+This is different for a new simulation kernel component that has to monitor the behaviour of not only the model, but the simulator as well.
+
+We add performance metrics in the simulation kernel, which logs relevant performance metrics and processes them for use in other components.
+These metrics include the number of events created and destroyed, the number of inter and intra kernel events, the number of rollbacks, the measured lookahead, details of the Global Virtual Time (GVT) and Earliest Output Time (EOT) calculations, and basic information on the fairness between different simulation kernels.
+With all these metrics, a component can get a fair view on both model and simulation kernel behaviour.
+
+For example, if the actually seen lookahead is significantly higher than the defined lookahead, it might be interesting to switch to optimistic synchronization.
+When the number of rollbacks is very high, conservative synchronization might be ideal.
+And when neither of these two is an option, the simulation might just (temporarily) have to fall back to sequential simulation.
+
+For performance reasons, statistics gathering is optional due to the imposed overhead.
 
 \subsubsection{Visualization of Communication}
-%TODO explain simple visualizer
-%TODO can be used by hotswapper
+To provide some more insight in the models we used as benchmarks previously, we created a simple visualization of the simulation trace.
+This trace visualizes the allocation of the model and all defined connections.
+For each connection, the number of events transferred is annotated.
+Examples are shown for the three benchmark models used before: Figures~\ref{fig:Queue_allocation},~\ref{fig:interconnect_allocation_parallel}, and~\ref{fig:phold_allocation} shows traces for the Queue, Interconnect, and PHOLD models respectively.
 
-A user can request a runtime switch between synchronization protocols.
-A typical use case is when a significant slowdown is observed or memory usage reaches a certain treshold.
-The simulation engine will swap out the current kernels with new instances while preserving the state of the models and the simulation itself, and resume using the new synchronization protocol.
-There is no real limit on the number of switches, but the replacing and synchronizing of the kernels comes at a runtime cost that is dependent on the number of present kernels and the actual hardware the simulation runs on.
-In a test case with 2 kernels the overhead of a double runtime switch is less than a second.
-This is an experimental but promising feature that can be used in an automated framework driven by machine learning techniques.
-
-%TODO combination with statistics gathering
-Dxex optionally collects runtime statistics for each kernel.
-These record any measurable event in the simulation such as : events created/destroyed, inter/intra kernel event frequency, reverts, GVT calculations, lookahead and eot values and insights into the fairness between the kernels themselves.
-Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, observe how balanced inter kernel message traffic actually is during a simulation or measure reverted transitions in optimistic synchronization.
-The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or provide feedback to drive synchronization switching.
-
-The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or provide feedback to drive synchronization switching.
-Figures \ref{fig:pholdtree_visualize_seq},\ref{fig:pholdtree_visualize_parDFS},\ref{fig:pholdtree_visualize_parBFS} give an example of the modular visualization output generated by dxex after simulation the PholdTree model.
-These figures demonstrate the effect of allocation visually and show the high inter kernels event count that degrades performance, as we will demonstrate in
-\nameref{PholdTreeallocation} .
-Dotted lines represent connections between models where no events were exchanged, directed edges represents events with frequency as weight, coloring is used to distinguish inter versus intra kernel events.
-% Add figure
 \begin{figure}
     \center
-    \includegraphics[width=\plotfraction\columnwidth, height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000.eps}
-    \caption{Visualization of PholdTree (d=1,n=3,t=5000) sequential simulation.}
-    \label{fig:pholdtree_visualize_seq}
+    \includegraphics[width=\modelfraction\columnwidth, height=8cm, keepaspectratio, angle=-90 ]{fig/queue_allocation.eps}
+    \caption{Queue model (d=2, w=7, t=5000, random timeadvance) allocation and simulation trace across 4 kernels.}
+    \label{fig:Queue_allocation}
 \end{figure}
@@
     \center
-    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
-    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
-    \label{fig:pholdtree_visualize_parBFS}
+    \includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_parallel_allocation.eps}
+    \caption{Interconnect parallel simulation trace for 6 models on 3 kernels.}
+    \label{fig:interconnect_allocation_parallel}
 \end{figure}
@@
     \center
-    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
-    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
-    \label{fig:pholdtree_visualize_parDFS}
+    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_parallel_allocation.eps}
+    \caption{Phold benchmark trace for parallel simulation using three kernels.}
+    \label{fig:phold_allocation}
 \end{figure}
+
+Naturally, results similar to this are relevant information that can be used by the hotswapping component.

### 2016-06-28 publications/Journal/5-related-work.tex
Re-add publication folder + draft journal paper.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
Updated related work.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

--- 2016-02-15 paper/5-related-work.tex
@@
 In terms of performance, dxex is shown to be competitive with adevs.
-While a more extensive set of benchmarks is required to make accurate comparisons, initial results are promising.
 Despite its high performance, adevs does not support optimistic synchronization, which we have shown to be highly relevant.
@@
 
+% Paragraph runtime switching
+In the PDES community the problem of choosing between synchronization protocols is well known and documented \cite{Jha:1994:UFC:195291.182480}.
+In the work of \cite{Das:1996:APP:256562.256602} the challenges facing the implementation of runtime switching between protocols is explored.
+A PDES implementation of such a feature was presented in \cite{perumalla2005musik}. 
+Our contribution entails bringing this feature to the PDEVS community.
+
+% Paragraph allocation effects
+Our emphasis on model allocation as a key factor in parallel performance has been studied in detail by the PDES community \cite{PDESpartitioning}. 
+Referenced as partitioning of the simulation model, most studies distinguish between communication overhead and computational distribution (load balancing) as the two dimensions to partition over.
+Partitioning a simulation model is identified as an issue in the work of \cite{Scalability} to achieve scalability. 
+In \cite{PDEVSpartitioning} partitioning is studied in a PDEVS context with attention to load balancing and communication overhead.
+Our contribution studies the effect of partitioning with emphasis on the effect of communication between processes and in the presence of a flattened hierarchy. 
+We focus on static partitioning since this is a limiting factor for our conservative synchronization implementation which does not support model relocation.
+
 In summary, dxex tries to find the middle ground between the concepts of PythonPDEVS, the performance of adevs, and the multiple synchronization protocols of CD++.

### 2016-06-28 publications/Journal/4a-allocation.tex

@@
-%TODO explain model allocation
-%TODO why model allocation
-Modeller can specify which kernel a model should be allocated to, should such manual intervention be required.
+Although the synchronization protocol is one of the defining factors in simulation performance, model allocation has a significant impact on which protocol is ideal, and even whether or not parallelization would make sense.
+Indeed, if the model is distributed in such a way that frequent communication is necessary between cores, parallelism is naturally reduced.
+This thus brings us to the topic of model allocation.
+Model allocation was one of the features also implemented by PythonPDEVS~\cite{PythonPDEVS2}.
+
+The modeller can specify which kernel a model should be allocated to, should such manual intervention be required.
 This is handled by the default model allocator.
 If no preference is given a simple striping scheme is followed but this is not sufficient in most simulations to achieve a speedup in parallel.
-In Section \nameref{sec:4-performance} we analyse the performance impact of allocation schemes.
-By overriding the default allocator a model author can tune the allocation scheme for a specific model to maximize parallel speedup.
-This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.
+By overriding the default allocator a modeller tunes the allocation scheme for a specific model, maximizing parallel speedup.
+This interface can be used to employ graph algorithms for an automatic allocation scheme, for example avoiding cycles in the dependency graph.
 
@@
 %TODO show model in statistics view
+In order to evaluate the influence of model allocation, we define a new model, based on PHOLD~\cite{PHOLD}.
+The model structure resembles a tree: an atomic model can have a set of children, with children being connected to each other as well.
+Connections can be uni- or bidirectional.
 
-The \textit{PHoldTree} model, a hierarchical model based on the semantics of PHOLD~\cite{PHOLD}.
-The model consists of recursively defined trees connected to each other, similar to the structure of an AST.
-Connections can be unidirectional or bidirectional.
-Unlike the \textit{Queue} model the width of the hierarchy is still present in the topology of the atomic models after the directconnect stage.
-This model allows us to investigate parallel speedup in several key dimensions in combination or isolation : model allocation, message density, uncertainty and hierarchy.
-The model has the following parameters : depth, fanout (width), probability of priority message and linkage direction.
-We demonstrate the importance of atomic to kernel allocation, in this model this can be configured as a bread-first versus a depth-first scheme.
-\textit{PHOLDTree} can be used to model gossiping in social networks, hierarchical resource sharing or load balancing.
-Unlike \textit{Queue} this model does not lend itself easily to parallelism and thus highlights the difficulties model authors face in when a parallel speedup is desired for real world models.
+Unlike the Queue model, the width of the hierarchy is still present in the topology of the atomic models after the direct connection stage.
+The PHOLDTree model allows us to investigate parallel speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.
+
+PHOLDTree can be used to model gossiping in social networks, hierarchical resource sharing or load balancing.
 The lookahead of an atomic node is $\epsilon$, simulating uncertainty as will be often the case in real world models.
+We demonstrate the importance of allocation, by comparing a breadth-first versus a depth-first scheme.
 
@@
 \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree.pdf}
-\caption{PHOLDTree model for depth 1, width 2.}
+%TODO depth 1 but 3 levels deep?
+\caption{PHOLDTree model for depth 1 and width 2.}
 \label{fig:PHOLDTree_model}

### 2016-06-28 publications/Journal/conclusion.tex
New order of sections

@@
-\subsection{Conclusion}
+\subsection{Conclusions on Performance Evaluation}
 We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
-We will summarize the key factors in achieving good parallel performance here. % reword aspects syn
-\subsubsection{Allocation}
-Allocation of atomic models over kernels is critical to obtain a speedup. 
-In the case of the Interconnect model it is not possible to allocate models without introducing a cyclic dependency between kernels with severe performance degradation as a result. 
-Even when no cycles exist in the topology between kernels a good allocation scheme will minimize the number of connections between kernels, for example the star topology in PHoldTree with depth first allocation or the chain topology in the Queue model.
-Dxex can optionally offer the user a visualization of the behaviour of the kernels under a user supplied allocation scheme to allow for more insight. 
-The same instrumentation could in future be used to optimize an existing allocation scheme.
-\subsubsection{Uncertainty}
-A simulation where future behaviour cannot be predicted is typically not suited for conservative simulation, but we have demonstrated that dxex's conservative kernel can still offer a non trivial speedup in this context.
-Optimistic is not limited by this uncertainty and can offer a good speedup regardless of lookahead. 
-Phold and PholdTree demonstrate that a single model can benefit from different synchronization protocols depending on a single parameter.
-\subsubsection{Limits}
-Dxex's conservative kernel is constrained by the amount of models it controls, as shown in the PHoldTree benchmark.
-This is especially the case when the minimal lookahead is $\epsilon$ requiring near constant polling of models for lookahead values.
-Optimistic is not constrained by the number of models it manages, but can quickly consume all available memory in a simulation.
-While this effect can be reduced with a thread aware allocation scheme, the underlying problem remains.
+But even with the right synchronization protocol, we have seen that two problems remain.
+
+First, although one of both synchronization protocols might be ideally suited for specific model behaviour, nothing guarantees that the model will exhibit the same behaviour throughout the simulation.
+Similarly to the polymorphic scheduler~\cite{MasterThesis}, we wish to make it possible for the ideal option to be switched during simulation.
+When changes to the model behaviour are noticed, the used synchronization protocol is modified as well.
+
+Second, the allocation of models is tricky and has a significant impact on performance.
+While our parallel speedup for the Queue model, for example, was rather high, this is mostly due to characteristics of the model: the dependency graph does not contain any cycles.
+When cycles were introduced, as in the Interconnect model, performance became disastrous.
+
+In the next two sections, we elaborate on these two problems.

### 2016-06-28 publications/Journal/parallel-performance.tex

@@
 The difference between dxex conservative and adevs becoming smaller.
-The same effect can be seen for weak scaling in Figure~\ref{fig:Queue_plot_weak}.\\
-In Figure \ref{fig:Queue_allocation} the allocation of the Queue model is visualized.
+The same effect can be seen for weak scaling in Figure~\ref{fig:Queue_plot_weak}.
+
+%In Figure \ref{fig:Queue_allocation} the allocation of the Queue model is visualized.
 This trace allows us to demonstrate the benchmark results and explain why optimistic is not ideal for such a chain topology as present in Queue.
@@
 	\label{fig:Queue_plot_strong}
-
-\end{figure}
-
-\begin{figure}
-	\center
-
-	\includegraphics[width=\modelfraction\columnwidth, height=8cm, keepaspectratio, angle=-90 ]{fig/queue_allocation.eps}
-	\caption{Queue model (d=2, w=7, t=5000, random timeadvance) allocation and simulation trace across 4 kernels.}
-	\label{fig:Queue_allocation}
 
@@
 \end{figure}
-\begin{figure}
-	\center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/phold_parallel_allocation.eps}
-	\caption{Phold benchmark trace for parallel simulation using three kernels.}
-	\label{fig:phold_allocation}
-\end{figure}
 

### 2016-06-28 publications/Journal/sequential-performance.tex

@@
 % 2 things to show : baseline performance and contrast of dxex v adevs
-\subsubsection{Queue}\label{4-seq-Queue}
+
+\subsubsection{Queue}
+\label{4-seq-Queue}
 % explain what we did: fixed number of models (400, 600 & 800), increasing depth
@@
 A set of three tests was performed, where each test has the same number of models but an increasing depth.
-% compare dxex & adevs performance
 The results can be seen in Figure~\ref{fig:Queue_benchmark_seq}.
@@
 \end{figure}
-\subsubsection{Interconnect}\label{4-seq-Interconnect}
+
+\subsubsection{Interconnect}
+\label{4-seq-Interconnect}
 In the Interconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
@@
 \end{figure}
-% We want to show the effect of message storms, that dxex is sensitive to this (even with allocators)
-\subsubsection{PHold}\label{4-seq-PHold}
-% Phold : difference with interconnect, less messages == equal performance dxex adevs
+
+\subsubsection{PHold}
+\label{4-seq-PHold}
 The PHold model is very similar to the Interconnect model.
@@
 \end{figure}
-\subsubsection{PHoldTree}
-PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain as in Queue.
-This topology is interesting since it highlights the effects of allocation which we have shown already in PHold and Interconnect to be of vital importance.
-This model allows us to investigate in depth the effects of non-cyclic allocation strategies and measure parallel speedup.\\
-Since adevs does not use the directconnect algorithm, we expect some performance penalty between dxex and adevs.
-The event frequency in PHoldTree can be configured to highlight dxex's performance disadvantage here as shown in the Interconnect benchmark.
-\paragraph*{Event frequency}
-Similarly when the probability on a priority event (p) increases the amount of event generated will increase and thus the runtime.
-In Figure \ref{fig:PHoldtree_seq_p_benchmark} we observe that p has a near linear impact on performance, with the shape of the curve due to rounding errors.
-The advantage adevs has in simulations where event frequency is high has been demonstrated by the interconnect benchmark.
-\paragraph*{Hierarchy}
-To establish a baseline for the parallel simulation benchmarks we measure how the PHoldTree benchmark performs sequentially when we vary each of the model's parameters. 
-When we increase the depth and fanout of the model, we expect to see an increase in runtime.
-In Figure \ref{fig:PHoldtree_seq_dn_benchmark} the n parameter (fanout) determines the performance penalty adevs suffers compared to dxex.\\
-If we compare the instances d=2, n=2 with d=2, n=4, which show a clear difference in performance between adevs and dxex, we conclude from the profiling callgraphs that the increase in width per subtree (n) leads to higher overhead for adevs.
-Dxex in contrast uses the directconnect algorithm and has no such overhead.
-An increase in n will rapidly increase the number of connections \ref{eq:pholdtreelinkcount} (and thus the routing problem adevs faces), whereas d will increase modelcount \ref{eq:pholdtreemodelcount} faster but have a lesser impact on connections.
-Both kernels scale linear in increasing number of atomic models.
-\begin{figure}
-	\center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/pholdtree_sequential_p.eps}
-	\caption{PHoldTree for increasing probability of priority message.}
-	\label{fig:PHoldtree_seq_p_benchmark}
-\end{figure}
-% Img for p increasing
-\begin{figure}
-	\center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/pholdtree_sequential_dn.eps}
-	\caption{PHoldTree : Effect of hierarchy.}
-	\label{fig:PHoldtree_seq_dn_benchmark}
-\end{figure}

### 2016-06-28 publications/Journal/various.tex

As we will demonstrate in ~\nameref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulations with high event frequency, especially in parallel with optimistic synchronization. Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation. This allows us by using a single compilation parameter to change the allocation strategy. This interface allows instrumentation of memory usage. Not only useful to detect memory leaks should this be a concern, it can also be of use to a model author to trace event creation in large models. The default implementation is the fastest common configuration verified by the benchmarks in this publication.

### 2016-06-29 publications/Journal/4a-allocation.tex
Made performance evaluation part of allocation shorter and removed a figure

@@
 
-PHOLDTree can be used to model gossiping in social networks, hierarchical resource sharing or load balancing.
-The lookahead of an atomic node is $\epsilon$, simulating uncertainty as will be often the case in real world models.
+The PHOLDTree model is similar in structure to models of gossiping in social networks~\cite{?}, hierarchical resource sharing~\cite{?} or load balancing~\cite{?}.
+The lookahead of an atomic node is the minimally allowed $\epsilon$, indicating uncertainty, as is often the case in realistic models.
 We demonstrate the importance of allocation, by comparing a breadth-first versus a depth-first scheme.
@@
 PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain as in Queue.
-This topology is interesting since it highlights the effects of allocation which we have shown already in PHold and Interconnect to be of vital importance.
+This topology is interesting since it highlights the effects of allocation.
 This model allows us to investigate in depth the effects of non-cyclic allocation strategies and measure parallel speedup.
 
-Since adevs does not use the directconnect algorithm, we expect some performance penalty between dxex and adevs.
-The event frequency in PHoldTree can be configured to highlight dxex's performance disadvantage here as shown in the Interconnect benchmark.
-
-\paragraph*{Event frequency}
-Similarly when the probability on a priority event (p) increases the amount of event generated will increase and thus the runtime.
-In Figure~\ref{fig:PHoldtree_seq_p_benchmark} we observe that p has a near linear impact on performance, with the shape of the curve due to rounding errors.
-The advantage adevs has in simulations where event frequency is high has been demonstrated by the interconnect benchmark.
-
-\paragraph*{Hierarchy}
-To establish a baseline for the parallel simulation benchmarks we measure how the PHoldTree benchmark performs sequentially when we vary each of the model's parameters.
-When we increase the depth and fanout of the model, we expect to see an increase in runtime.
-In Figure~\ref{fig:PHoldtree_seq_dn_benchmark} the n parameter (fanout) determines the performance penalty adevs suffers compared to dxex.
-
-If we compare the instances d=2, n=2 with d=2, n=4, which show a clear difference in performance between adevs and dxex, we conclude from the profiling callgraphs that the increase in width per subtree (n) leads to higher overhead for adevs.
-Dxex in contrast uses the directconnect algorithm and has no such overhead.
-An increase in n will rapidly increase the number of connections \ref{eq:pholdtreelinkcount} (and thus the routing problem adevs faces), whereas d will increase modelcount \ref{eq:pholdtreemodelcount} faster but have a lesser impact on connections.
-Both kernels scale linear in increasing number of atomic models.
+Since adevs does not use direct connection, we expect a noticable performance difference between dxex and adevs.
+This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout determines the performance penalty adevs suffers compared to dxex.
+Profiling indeed indicated that an increase in width per subtree ($n$) leads to higher overheads in adevs due to the lack of direct connection.
+Dxex uses direct connection, making it independent on fanout, but only of the number of models.
+Slight deviations can still be seen, though, caused by the initialization overhead of direct connection.
+Both adevs and dxex scale linearly in the number of atomic models.
 
@@
     \center
-    \includegraphics[width=\plotfraction\columnwidth]{fig/pholdtree_sequential_p.eps}
-    \caption{PHoldTree for increasing probability of priority message.}
-    \label{fig:PHoldtree_seq_p_benchmark}
-\end{figure}
-% Img for p increasing
-\begin{figure}
-    \center
     \includegraphics[width=\plotfraction\columnwidth]{fig/pholdtree_sequential_dn.eps}
-    \caption{PHoldTree : Effect of hierarchy.}
+    \caption{Effect of hierarchy in sequential simulation.}
     \label{fig:PHoldtree_seq_dn_benchmark}

### 2016-06-29 publications/Journal/4a-allocation.tex
Add in new figures for allocation of BF and DF allocation

@@
 This model allows us to investigate in depth the effects of non-cyclic allocation strategies and measure parallel speedup.
+First, we evaluate the model in sequential simulation to provide a baseline for parallel simulation.
 
+\subsubsection{Sequential Simulation}
 Since adevs does not use direct connection, we expect a noticable performance difference between dxex and adevs.
@@
 Both adevs and dxex scale linearly in the number of atomic models.
+We expect to see a similar distinction in parallel simulation.
 
@@
 
-We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization.
-We will demonstrate that allocation is critical to achieve a parallel speedup and that the configuration of the model will give either conservative or optimistic an advantage.
-\paragraph*{Allocation}\label{PholdTreeallocation}
-The PholdTree benchmark can be configured to use 2 allocation schemes: breadth first and depth first.
+\subsubsection{Parallel Simulation}
+Next, we run the model using two different model allocation schemes: breadth-first and depth-first.
+But first, we explain what we mean by both allocation schemes.
+
+With breadth-first allocation, we traverse the tree in a breadth-first way, allocating subsequently visited atomic models to the same node.
+This means, intuitively, that atomic models at the same level in the tree, but not necessarily siblings, are frequently allocated to the same node.
+Since there is only infrequently some communication between 
+This allocation strategy is shown in Figure~\ref{fig:BF}.
+
+With depth-first allocation, we traverse the tree in a depth-first way, allocating subsequently visited atomic models to the same node.
+This means, intuitively, that subtreees are frequently allocated to the same node.
+This allocation strategy is shown in Figure~\ref{fig:DF}.
+
 The breadth first allocation scheme will result in kernels that will form a dependency chain with multiple branches, much like in the Queue model.
@@
 In Figures \ref{fig:pholdtree_visualize_parBFS} and \ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for both allocation schemes highlighting the remarks made in this paragraph.
-%
-%\begin{figure}
-%   \center
-%   
-%   \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeBFS.eps}
-%   \caption{PholdTree model breadth first allocation with 3 kernels.}
-%   \label{fig:PholdTree_model_bfs}
-%   
-%   \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeDFS.eps}
-%   \caption{PholdTree model depth first allocation with 3 kernels.}
-%   \label{fig:PholdTree_model_dfs}
-%\end{figure}
+\begin{figure}
+   \center
+
+   \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree_alloc_BF.pdf}
+   \caption{PholdTree model breadth first allocation with 4 kernels.}
+   \label{fig:PholdTree_model_bfs}
+
+   \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree_alloc_DF.pdf}
+   \caption{PholdTree model depth first allocation with 4 kernels.}
+   \label{fig:PholdTree_model_dfs}
+\end{figure}
 

### 2016-06-29 publications/Journal/4a-allocation.tex
Some analysis of results added

@@
    \label{fig:PholdTree_model_bfs}
-
    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree_alloc_DF.pdf}
@@
 \end{figure}
+
+After simulation, the traces can be visualized for both breadth-first and depth-first allocation.
+Using a breadth-first allocation scheme, as shown in Figure~\ref{fig:pholdtree_visualize_parBFS}, we notice that many events get exchanged between cores.
+This is caused by the high number of inter-core connections and the high number of events exchanged over these connections.
+The number of connections between nodes at the same simulation core is also rather low.
+Using a depth-first allocation scheme, however, as shown in Figure~\ref{fig:pholdtree_visualize_parDFS},
+
+\begin{figure}
+    \center
+    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
+    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
+    \label{fig:pholdtree_visualize_parBFS}
+\end{figure}
+\begin{figure}
+    \center
+    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
+    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
+    \label{fig:pholdtree_visualize_parDFS}
+\end{figure}
+
+Simulation results are shown in Figure~\ref{fig:?} for both allocation schemes in combination with both synchronization protocols.
+We see that for both synchronization protocols, the depth-first allocation is significantly better than breadth-first allocation.
+This is what we expected: depth-first allocation maintains locality better than breadth-first allocation.
+
+%TODO this paragraph is strange: is this really the case?
+Interestingly, we see that optimistic synchronization is less influenced by the allocation than conservative synchronization.
+This is likely caused due to the lower number of connections to take into account in conservative synchronization.
+Whereas conservative synchronization needs to take into account even scarcely used connections, optimistic synchronization does not.
+The same is true in the opposite direction, though, where optimistic synchronization is slower when a good allocation is chosen.
+Conservative synchronization will then be able to make better estimates, whereas optimistic synchronization does not make estimations.
 
@@
 %TODO link back to initial performance evaluation
-
-\begin{figure}
-    \center
-    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
-    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
-    \label{fig:pholdtree_visualize_parBFS}
-\end{figure}
-\begin{figure}
-    \center
-    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
-    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
-    \label{fig:pholdtree_visualize_parDFS}
-\end{figure}

### 2016-06-29 publications/Journal/4a-allocation.tex
Reworded conclusion about figure (some more might be necessary); removed everything about PHOLDTree that doesn't have to do with allocation

@@
 \begin{figure}
+    \center
     \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree.pdf}
@@
 
-Simulation results are shown in Figure~\ref{fig:?} for both allocation schemes in combination with both synchronization protocols.
+Simulation results are shown in Figure~\ref{fig:PholdTree_plot_alloc_high} for both allocation schemes in combination with both synchronization protocols.
 We see that for both synchronization protocols, the depth-first allocation is significantly better than breadth-first allocation.
 This is what we expected: depth-first allocation maintains locality better than breadth-first allocation.
+
+The most prominent aspect of these results is the low performance for conservative depth-first allocation for two cores.
+This is mostly caused by the difference between a sequential simulation and a parallel simulation: suddenly we need to take into account synchronization and passing around of lookahead values.
+And since the number of cores is low, the overhead dominates any further speedup.
+%TODO add some more explanation here maybe?
+%TODO why is this not the case for optimistic?
 
@@
 
-\paragraph*{Strong Scaling}\label{pholdtreestrongscale}
-In Figure \ref{fig:PholdTree_plot_alloc_high} we see that the difference in performance for all 4 kernel configurations compared to that shown in Figure \ref{fig:PholdTree_plot_alloc_low} is a constant factor.
-The probability of the priority message has no extra impact on performance other than that shown in sequential performance.
-
-The probability of a high priority event in this model does not affect the performance difference between conservative and optimistic.
-The key parameter quickly becomes the load of a kernel in models.
-Our conservative implementation in an uncertain simulation is very sensitive to a high load in models, whereas optimistic has no such limitation.
-
-The difference between depth first and breadth first allocated kernels is striking, the first results in sublinear speedup for both synchronization protocols. %reword 
-The breadth first allocation scheme will lead to a very high number of inter-kernel connections, which is detrimental for any parallel synchronization algorithm.
-An event exchanged between kernels cannot be securely deallocated without a GVT algorithm and all the complexity this entails.
-Even a fast asynchronous GVT algorithm will require some form of inter kernel synchronization and span a timeframe during which allocated memory cannot be reused, forcing new allocations.
-Similar to the Queue benchmark the breadth first allocation scheme leads to a topology for the kernels resembling a chain but with more branches in the chain.
-In Queue there is only a single model on the edge of a kernel exchanging messages to a single other model in a neighbouring kernel, this is not the case in the PholdTree under breadth first allocation.
-This explains the difference in speedup between Queue and PholdTree despite the similarity in kernel topology.\\
-Depth first allocation still has a higher inter kernel connection count, but not on the same order as breadth first allocated PholdTree.
-Depth first allocation will converge to a star topology which in this benchmark leads to a significant speedup.\\
-The performance drop observed for 2 kernels for all allocation schemes and synchronization protocols is in part due to the higher link count between kernels.
-With more kernels these connections will be spread across more kernels and result in a relative lower performance penalty.
-\\
-Conservative synchronization suffers a further penalty in this benchmark when there are only a few kernels.
-The PholdTree model has a lookahead of $\epsilon$ leading the kernel to query each allocated atomic model for a lookahead value on each simulation round.
-As the atomic models are spread over more kernels this effect is reduced leading to an increase in performance.
-With 4 kernels conservative surpasses optimistic in speedup.
-This is clear evidence that conservative synchronization is a good candidate for parallel simulation even in simulations with uncertainty.
 \begin{figure}
     \center
-
-    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreealloclowp.eps}
-    \caption{PholdTree model performance under different allocation schemes with low message probability}
-    \label{fig:PholdTree_plot_alloc_low}
-
     \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeallochighp.eps}
-    \caption{PholdTree model performance under different allocation schemes with high message probability}
+    \caption{PholdTree model performance using different allocation schemes.}
     \label{fig:PholdTree_plot_alloc_high}
-\end{figure}
-
-\paragraph*{Weak Scaling}
-%Intro
-In this paragraph we investigate dxex's parallel performance when the the number of atomic models distributed over a fixed number of kernels varies.
-From \ref{eq:pholdtreemodelcount} we know that increasing d will lead to a very rapid increase in models.
-We want to observe what happens in a closer to linear increase in model count per kernel by varying n.
-
-% Explain why adevs is not plotted (its x100 slower)
-Adevs' conservative implementation cannot handle the uncertainty (lack of non $\epsilon$ lookahead) here and is omitted from the comparison.
-Its parallel performance is two orders of magnitude slower than the sequential implementation.
-Upon investigation we conclude that this slowdown is caused by not implementing the \{begin/end\}lookahead() functions, which when not overridden in a model trigger an exception on each invocation.
-This exception handling completely stalls the kernel.
-We do not implement this function in our PholdTree variant for adevs since this would force us to implement state saving in the model code and not use the provided state saving functionality in the kernel, as is done in dxex or PythonPDEVS's optimistic.
-In dxex state saving is never required for a conservative simulation, this is handled transparently for the user who need not implement this behaviour.
-By using the state saving technique inside the model code we feel we would no longer compare identical models across different synchronization algorithms, although we do not doubt adevs' increased performance when these functions are overridden.
-
-% Optimistic
-In Figure \ref{fig:PholdTree_plot_weaknopt} we observe that dxex's optimistic kernels are slightly sensitive to the number of atomic models allocated to them.
-In the worst case with 2 kernels there is no real speedup observable, as soon as the number of kernels increases we see a converging speedup trend, that except for 32 kernels is almost constant despite the increase in n.
-Note that only depth first allocated kernels are measured here, breadth first as we can see in \ref{fig:PholdTree_plot_alloc_high} has no speedup advantage.
-The probability parameter is kept at 0.1 for the same reason, it only induces a linear increase in load.
-
-% Conservative
-Conservative has a more nuanced speedup behaviour in this benchmark.
-We have detailed how a conservative kernel in dxex scales linear in the number of atomic models (if lookahead is $\epsilon$), this effect is more clearly visible here.
-\\
-It is clear that as the fanout (n) of the model increases the kernel performance degrades rapidly.
-The intersection of the performance graph of each configuration is shifted to the right, the actual 1-speedup point increases as the kernelcount increases.
-For a d=4, n=5 configuration with 4 kernels each kernel has a load of ~1300 atomic models when it crosses the 1-speedup line.
-\\
-An 8 kernel configuration with d=4 n=6 can sustain a load of ~2700 atomic models before it reaches that point.
-From the results we see that the 32-kernel configuration is not yet slowing down with the current parameters.
-\\
-The modelcount is only a part of the explanation of this effect, as the kernelcount increases the amount of inter kernel messages will be split into more distinct sets which can be handled with more concurrency in dxex's architecture.
-That this effect is not always true can be observed from the d=4 n=3 32-kernel datapoint.
-Note that in this configuration 341 atomic models are distributed over 32 kernels, a relative low model load can more easily highlight synchronization overhead, even if allocation is optimal.\\
-As with optimistic we do not include breadth first allocated benchmarks in this speedup plot since none of those achieve a speedup higher than 1.
-\\
-If we combine both to determine which synchronization protocol is more optimal we see in Figure \ref{fig:PholdTree_plot_weakall} that for these configurations conservative with 8 kernels is an ideal configuration up until n=5, where the modelcount starts to degrade performance.
-Optimistic with the same kernelcount is almost insensitive to the increase in modelcount and is therefore a more robust choice.
-Note how optimistic and conservative with 32 kernels at this point still have a reasonable speedup which is surprising given the synchronization overhead in such a large set of kernels.
-\begin{figure}
-    \center
-
-    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingnopt.eps}
-    \caption{PholdTree model weak scaling under varying fanout and optimistic synchronization}
-    \label{fig:PholdTree_plot_weaknopt}
-
-    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingncon.eps}
-    \caption{PholdTree model weak scaling under varying fanout and conservative synchronization}
-    \label{fig:PholdTree_plot_weakncon}
-
-    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeweakscalingall.eps}
-    \caption{PholdTree model weak scaling under varying fanout and different synchronization algorithms}
-    \label{fig:PholdTree_plot_weakall}
 \end{figure}

### 2016-06-29 publications/Journal/4a-allocation.tex
Save some space by putting figures next to each other

@@
 
-   \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree_alloc_BF.pdf}
+   \includegraphics[width=\textwidth]{fig/pholdtree_alloc_BF.pdf}
    \caption{PholdTree model breadth first allocation with 4 kernels.}
    \label{fig:PholdTree_model_bfs}
-   \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree_alloc_DF.pdf}
+   \includegraphics[width=\textwidth]{fig/pholdtree_alloc_DF.pdf}
    \caption{PholdTree model depth first allocation with 4 kernels.}
@@
     \center
-    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
-    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
+    \begin{minipage}{0.45\textwidth}
+            \center
+            \includegraphics[width=\textwidth]{fig/pholdtreed1n3t5000c4BFS.eps}
+            \caption{Visualization of a PholdTree simulation with breadth first allocation and 4 kernels.}
             \label{fig:pholdtree_visualize_parBFS}
-\end{figure}
-\begin{figure}
+    \end{minipage}
+    \begin{minipage}{0.45\textwidth}
             \center
-    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
-    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
+            \includegraphics[width=\textwidth]{fig/pholdtreed1n3t5000c4DFS.eps}
+            \caption{Visualization of a PholdTree simulation with depth first allocation and 4 kernels.}
             \label{fig:pholdtree_visualize_parDFS}
+    \end{minipage}
 \end{figure}

### 2016-06-29 publications/Journal/5-related-work.tex
Slight rewrite of related work (style, LaTeX issues, ...)

--- 2016-02-13 paper/5-related-work.tex
@@
-Several similar \textsf{DEVS} simulation tools have already been implemented, though they differ in key aspects.
+Several similar \textsf{DEVS} simulation tools have already been implemented, though they differ in several key aspects.
 We discuss several dimensions of related work, as we try to compromise between different tools.
@@
 In terms of code design and philosophy, dxex is most related to PythonPDEVS~\cite{PythonPDEVS}.
-Performance of PythonPDEVS was still decent, through the use of simulation and activity hints from the modeler.
-This allowed the kernel to optimize its internal data structures and algorithms for the specific model being executed.
-All changes were completely transparent to the model, and were completely optional.
+Performance of PythonPDEVS was still decent through the use of ``hints'' from the modeler.
 In this spirit, we offer users the possibility to choose between different synchronization protocols.
 This allows users to choose the most appropriate synchronization protocol, depending on the model.
-Contrary to PythonPDEVS, however, dxex doesn't support distribution, model migrations~\cite{PythonPDEVS2}, or activity hints~\cite{PythonPDEVS_ACTIMS}.
+Contrary to PythonPDEVS, however, dxex doesn't support distributed simulation~\cite{JDF}, model migrations~\cite{PythonPDEVS2}, or activity hints~\cite{PythonPDEVS_ACTIMS}.
 
-While PythonPDEVS offers very fast turnaround cycles, due to the use of an interpreted language, simulation performance was easily outdone by compiled simulation kernels.
+Although PythonPDEVS offers very fast turnaround cycles, simulation performance was easily outdone by compiled simulation kernels.
 In terms of performance, adevs~\cite{adevs} offered much faster simulation, at the cost of a significant compilation time.
@@
 As a result, the complete simulation kernel has to be compiled again every time.
-Dxex compromises, as vle~\cite{vle} or PowerDEVS~\cite{PowerDEVS}, by separating the simulation kernel into a shared library.
-After the initial compilation of the simulation tool, only the model has to be compiled and linked to the library.
+Similarly to vle~\cite{vle} and PowerDEVS~\cite{PowerDEVS}, dxex compromises by separating the simulation kernel into a shared library.
+After the initial compilation of the simulation tool, only the model has to be compiled and linked to the shared library.
 This significantly shortens the turnaround cycle, while still offering good performance.
 In terms of performance, dxex is shown to be competitive with adevs.
-While a more extensive set of benchmarks is required to make accurate comparisons, initial results are promising.
 Despite its high performance, adevs does not support optimistic synchronization, which we have shown to be highly relevant.
@@
 Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though none have done it in a strictly modular way that allows straightforward protocol switching for a single given model.
-For example CD++~\cite{CD++} has both a conservative (CCD++~\cite{CCD++}) and optimistic (PCD++)~\cite{PCD++}) variant.
-Despite the implementation of both protocols, they are different projects entirely, and are incompatible with modern compilers.
-Dxex, on the other hand, is a single project, where switching between different synchronization protocols is as simple as switching any other configuration parameter.
+For example, adevs only supports conservative synchronization, and vle only supports experiment-level parallelism (\textit{i.e.}, run multiple experiments concurrently).
+Closest to our support for multiple synchronization protocols is CD++~\cite{CD++}.
+For CD++, both a conservative (CCD++~\cite{CCD++}) and optimistic (PCD++)~\cite{PCD++}) variant exist.
+Despite the implementation of both protocols, they are different projects entirely.
+Some features might therefore be implemented in CCD++ and not in PCD++, or vice versa.
+And while this might not yet be that significant a problem to this day, this problem will only get worse when each project follows its own course.
+Dxex, on the other hand, is a single project, where the choice of synchronization protocol is a simple configuration.
 CD++, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourselves to parallel simulation.
@@
 
+% Paragraph runtime switching
+In the PDES community, the problem of choosing between synchronization protocols is well known and documented~\cite{Jha:1994:UFC:195291.182480}.
+The challenges of implementing such runtime switching have previously been explored already~\cite{Das:1996:APP:256562.256602}, and an implementation was given by, for example~\cite{perumalla2005musik}.
+Our contribution entails bringing this same feature to the DEVS community, further expanding upon our unique feature of multiple synchronization protocols.
+
+% Paragraph allocation effects
+Model allocation and its impact on parallel performance has previously also been studied in the PDES community~\cite{PDESpartitioning}.
+Referenced as partitioning of the simulation model, most studies distinguish between communication overhead and computational distribution (load balancing) as the two dimensions to partition over.
+Partitioning a simulation model is identified as an issue to achieve scalability~\cite{Scalability}. 
+Some research in the context of Parallel DEVS has been done, where they turn their attention to load balancing and communication overhead~\cite{PDEVSpartitioning}.
+Our contribution studies the effect of partitioning with emphasis on the effect of communication between processes and in the presence of a flattened hierarchy. 
+We focus on static partitioning since this is a limiting factor for our conservative synchronization implementation which does not support model relocation.
+Model relocation, as implemented by PythonPDEVS~\cite{PythonPDEVS2}, might be an interesting addition to only model allocation at the start of simulation.
+
 In summary, dxex tries to find the middle ground between the concepts of PythonPDEVS, the performance of adevs, and the multiple synchronization protocols of CD++.
+To further profit from our multiple synchronization protocols in a single tool, we further added runtime switching between synchronization protocols and model allocation support.

### 2016-06-29 publications/Journal/6-conclusions.tex
Reword conclusions

@@
-In this paper, we introduced DEVS-Ex-Machina (``dxex''), a new C++14-based \textsf{Parallel DEVS} simulation tool.
+In this paper, we introduced DEVS-Ex-Machina (``dxex''), a new C++11-based \textsf{Parallel DEVS} simulation tool.
 Our main contribution is the implementation of multiple synchronization protocols for parallel multicore simulation.
 We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
-Dxex allows the user to choose between either conservative or optimistic synchronization as simple as any other configuration option.
-Notwithstanding this modularity, dxex achieves performance competitive to adevs, another very efficient \textsf{DEVS} simulation tool.
+Dxex allows the user to choose between either conservative or optimistic synchronization at the start of simulation.
+Depending on observed model behaviour and simulation performance, runtime switching between synchronization protocols can be used.
+
+Notwithstanding our modularity, dxex achieves performance competitive to adevs, another very efficient \textsf{DEVS} simulation tool.
 Performance is measured both in elapsed time, and memory usage.
@@
 Having simulation control, which can throttle the speed of nodes that use up too much memory, has been shown to work in these situations~\cite{FujimotoBook}.
-Faster GVT implementations, such as those presented by~\cite{Fujimoto:1997:CGV:268403.268404} and~\cite{Bauer:2005:SND:1069810.1070159}, might further help to minimize this problem.
+Faster GVT implementations~\cite{Fujimoto:1997:CGV:268403.268404,Bauer:2005:SND:1069810.1070159} might further help to minimize this problem.
 Secondly, the runtime switching between synchronization protocols can be driven using machine learning techniques. The simulation engine is already capable of collecting data to inform such a process.

### 2016-06-29 publications/Journal/abstract.tex
Updated abstract

--- 2016-06-24 publications/Journal/abstract.tex
@@
-With the ever increasing complexity of simulation models, parallel simulation
-becomes necessary to perform the simulation within reasonable time bounds.
+With the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform simulation within reasonable time bounds.
 The built-in parallelism of Parallel DEVS is often insufficient to tackle this problem on its own.
@@
 Due to the significantly different implementation of these protocols, most Parallel DEVS simulation tools are limited to only one such protocol.
-In this paper, we present a Parallel DEVS simulator, grafted on C++11, but based on PythonPDEVS, which supports both conservative and optimistic synchronization.
-% TODO if completed, add online switching
-We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol and demonstrate the inlfuence model allocation and uncertainty has on this choice. 
-Performance results are compared to adevs, in terms of CPU time and memory usage.
+In this paper, we present a Parallel DEVS simulator, grafted on C++11 and based on PythonPDEVS, supporting both conservative and optimistic synchronization protocols.
+We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol.
+A comparison is made to adevs in terms of CPU time and memory usage, to show that our modularity does not hinder performance.
+We further allow for an external component to gather simulation statistics, on which runtime switching between the different synchronization protocols can be based.
+Model allocation is also studied to see how our conservative and optimistic synchronization protocols are affected by good and bad allocations.

### 2016-06-29 publications/Journal/1-introduction.tex
Fixed figures for twocolumn

@@
 The remainder of this paper is organized as follows:
-Section \hyperref[sec:2-background]{2} introduces the necessary background on synchronization protocols.
+Section~\hyperref[sec:2-background]{2} introduces the necessary background on synchronization protocols.
 Section~\hyperref[sec:3-features]{3} elaborates on our design that enables this flexibility.

### 2016-06-29 publications/Journal/3-features.tex
Change rng plot.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
Fixed figures for twocolumn

@@
 
-\begin{figure}
-    \includegraphics[width=\columnwidth]{fig/cores_class_diagram.eps}
+\begin{figure*}
+    \includegraphics[width=\textwidth]{fig/cores_class_diagram.eps}
 	\caption{Dxex design.}
 	\label{fig:class_diagram}
-\end{figure}
+\end{figure*}
 
@@
     \center
-    \includegraphics[width=\modelfraction\columnwidth]{fig/memory_allocators_parallel.eps}
+    \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
     \caption{Effect of memory allocators on parallel execution time.}
@@
     \center
-    %TODO remove adevs results as not applicable here
-    \includegraphics[width=\modelfraction\columnwidth]{fig/rngspeedupeffectdevstone.eps}
-    \caption{Speedup with different RNG usage patterns.}
+    \includegraphics[width=\columnwidth]{fig/rngspeedupeffectdevstone.eps}
+    \caption{Speedup with different RNG usage patterns compared to STL random number generator.}
     \label{fig:Queuerngspeedup}

### 2016-06-29 publications/Journal/4a-allocation.tex
Removed "Linking back" subsection
Fixed figures for twocolumn

@@
     \center
-    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree.pdf}
+    \includegraphics[width=\columnwidth]{fig/pholdtree.pdf}
     %TODO depth 1 but 3 levels deep?
@@
     \center
-    \includegraphics[width=\plotfraction\columnwidth]{fig/pholdtree_sequential_dn.eps}
+    \includegraphics[width=\columnwidth]{fig/pholdtree_sequential_dn.eps}
     \caption{Effect of hierarchy in sequential simulation.}
@@
 
-   \includegraphics[width=\textwidth]{fig/pholdtree_alloc_BF.pdf}
+   \includegraphics[width=\columnwidth]{fig/pholdtree_alloc_BF.pdf}
    \caption{PholdTree model breadth first allocation with 4 kernels.}
    \label{fig:PholdTree_model_bfs}
-   \includegraphics[width=\textwidth]{fig/pholdtree_alloc_DF.pdf}
+   \includegraphics[width=\columnwidth]{fig/pholdtree_alloc_DF.pdf}
    \caption{PholdTree model depth first allocation with 4 kernels.}
@@
     \center
-    \begin{minipage}{0.45\textwidth}
-            \center
-            \includegraphics[width=\textwidth]{fig/pholdtreed1n3t5000c4BFS.eps}
+    \includegraphics[width=\columnwidth]{fig/pholdtreed1n3t5000c4BFS.eps}
     \caption{Visualization of a PholdTree simulation with breadth first allocation and 4 kernels.}
     \label{fig:pholdtree_visualize_parBFS}
-    \end{minipage}
-    \begin{minipage}{0.45\textwidth}
+\end{figure}
+\begin{figure}
     \center
-            \includegraphics[width=\textwidth]{fig/pholdtreed1n3t5000c4DFS.eps}
+    \includegraphics[width=\columnwidth]{fig/pholdtreed1n3t5000c4DFS.eps}
     \caption{Visualization of a PholdTree simulation with depth first allocation and 4 kernels.}
     \label{fig:pholdtree_visualize_parDFS}
-    \end{minipage}
 \end{figure}
@@
     \center
-    \includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeallochighp.eps}
+    \includegraphics[width=\columnwidth]{fig/pholdtreeallochighp.eps}
     \caption{PholdTree model performance using different allocation schemes.}
@@
 \end{figure}
-
-\subsection{Linking Back}
-%TODO link back to initial performance evaluation

### 2016-06-29 publications/Journal/4b-hotswap.tex
Save some space by putting figures next to each other
Fixed figures for twocolumn

@@
 
-\begin{figure}
+\begin{figure*}
     \center
-    \includegraphics[width=\modelfraction\columnwidth, height=8cm, keepaspectratio, angle=-90 ]{fig/queue_allocation.eps}
-    \caption{Queue model (d=2, w=7, t=5000, random timeadvance) allocation and simulation trace across 4 kernels.}
+    \includegraphics[height=\textwidth, angle=-90 ]{fig/queue_allocation.eps}
+    \caption{Queue model simulation trace across 4 kernels.}
     \label{fig:Queue_allocation}
-\end{figure}
+\end{figure*}
 \begin{figure}
@@
     \includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_parallel_allocation.eps}
-    \caption{Interconnect parallel simulation trace for 6 models on 3 kernels.}
+    \caption{Interconnect simulation trace for 6 models on 3 kernels.}
     \label{fig:interconnect_allocation_parallel}
@@
     \includegraphics[width=\plotfraction\columnwidth]{fig/phold_parallel_allocation.eps}
-    \caption{Phold benchmark trace for parallel simulation using three kernels.}
+    \caption{Phold simulation trace for parallel simulation using three kernels.}
     \label{fig:phold_allocation}

### 2016-06-29 publications/Journal/parallel-performance.tex
Fixed figures for twocolumn

@@
 
-	\includegraphics[width=\modelfraction\columnwidth]{fig/queue_fixed_weak_speedup.eps}
+	\includegraphics[width=\columnwidth]{fig/queue_fixed_weak_speedup.eps}
 	\caption{Queue model weak scaling speedup compared to dxex sequential.}
@@
 
-	\includegraphics[width=\modelfraction\columnwidth]{fig/queue_fixed_strong_speedup.eps}
+	\includegraphics[width=\columnwidth]{fig/queue_fixed_strong_speedup.eps}
 	\caption{Queue model strong scaling speedup compared to dxex sequential.}
@@
 	\center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_parallel.eps}
+	\includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
 	\caption{Interconnect benchmark results for parallel simulation.}
@@
 	\center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_parallel_allocation.eps}
+	\includegraphics[width=\columnwidth]{fig/interconnect_parallel_allocation.eps}
 	\caption{Interconnect parallel simulation trace for 6 models on 3 kernels.}
@@
     \center
-    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_remotes.eps}
+    \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
     \caption{Phold benchmark results for parallel simulation using four kernels, four atomics per node, with varying percentage of remote events.}
@@
 	\center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/phold_priority.eps}
+	\includegraphics[width=\columnwidth]{fig/phold_priority.eps}
 	\caption{Phold benchmark results for parallel simulation using four kernels, with varying amount of high-priority events.}

### 2016-06-29 publications/Journal/3-features.tex
Fixed the floating string, which was part of an eps for some reason

@@
 \end{figure}
-Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{WhatEveryComputerProgrammerShouldKnowAboutMemory}: memory bandwidth doesn't increase as fast as the number of cores using it.
+Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}: memory bandwidth doesn't increase as fast as the number of cores using it.
 While this is always a problem, it is aggravated in dxex by providing automatic memory management for events and states.

### 2016-06-29 publications/Journal/models.tex
Removed enumeration, which didn't look nice in two column

@@
 \subsection{Benchmarks}
-We use four different benchmarks, which cover different aspects of the simulation kernel:
-\begin{enumerate}
-	\item The \textit{Queue} model, based on the \textit{HI} model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
+We use three different benchmarks, which cover different aspects of the simulation kernel.
+
+First, the \textit{Queue} model, based on the \textit{HI} model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
 A single generator pushes events into the queue, which are consumed by the processors after a fixed or random delay.
@@
 
-	\item The \textit{PHOLD} model, presented by~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
+Second, the \textit{PHOLD} model, presented by~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
 Each atomic model is directly connected to every other atomic model.
@@
 
-	\item The \textit{Interconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
+Third, the \textit{Interconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
 Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event.
@@
 An example for four models is shown in Figure~\ref{fig:interconnect_model}.
-\end{enumerate}
 

### 2016-06-29 publications/Journal/1-introduction.tex
Changes to introduction, and fixed the outline of the paper

@@
 
-This paper introduces DEVS-Ex-Machina\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
-The selected synchronization protocol is transparent to the simulated model: users should merely determine, which protocol they wish to use.
+This paper introduces DEVS-Ex-Machina\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``dxex''): our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
+The selected synchronization protocol is transparent to the simulated model: users only determine which protocol they wish to use.
 Users who simulate a wide variety of models, with different ideal synchronization protocols, can simply run the same model with different synchronization protocols.
-We investigate in this paper how model allocation and uncertainty determine the choice between synchronization protocols. 
-The synchronization overhead is demonstrated by reducing the computational load of a model to near zero.
+Since the ideal synchronization protocol might change throughout the simulation, we also support runtime switching between these protocols.
+This runtime switching is based on performance metrics, which are logged during simulation.
+Information is made available to a seperate component, where a choice is made about which synchronization protocol to use next.
+Additionally, we investigate in this paper how model allocation influences the performance of our synchronization protocols.
+To this end, we have included an allocation component to our simulation kernel.
 
-Our tool is based on PythonPDEVS, but implemented in C++11 for increased performance, using features from the new C++14 standard when possible.
-Unlike PythonPDEVS dxex only supports multicore parallelism.
+Our tool is based on PythonPDEVS, but implemented in C++11 for increased performance, using features from the new C++14 standard when supported by the compiler.
+Unlike PythonPDEVS, dxex only supports multicore parallelism.
 
 We implemented a model that, depending on a single parameter, changes its ideal synchronization protocol.
-We demonstrate using several models the factors influencing the performance under a given synhronization protocol.
-Dxex, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol. With dxex users can always opt to use the fastest protocol available.
-To verify that our flexibility does not counter performance, we compare to adevs, currently one of the fastest \textsf{DEVS} simulation tools available~\cite{PythonPDEVS,DEVSSurvey}.
-
-% Switching and statistics
-Dxex offers visualization of the simulation and in depth statistics.
-A modeller can then make a more informed decision on which synchronization protocol to use or even intervene during simulation and request a switch between protocols.
+Using several models, we demonstrate the factors influencing the performance under a given synhronization protocol.
+Dxex, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol.
+With dxex users can always opt to use the fastest protocol available, and through its modularity, users could even implement their own.
+To verify that this modularity does not hinder performance, we compare to another Parallel DEVS simulation kernel, called adevs~\cite{adevs}.
 
@@
 The remainder of this paper is organized as follows:
-Section~\hyperref[sec:2-background]{2} introduces the necessary background on synchronization protocols.
-Section~\hyperref[sec:3-features]{3} elaborates on our design that enables this flexibility.
-In Section~\hyperref[sec:4-performance]{4}, we evaluate performance of our tool by comparing its different synchronization protocols, and by comparing to adevs.
-Related work is discussed in Section~\hyperref[sec:5-related-work]{5}.
-Section~\hyperref[sec:6-conclusion]{6} concludes the paper and gives future work.
+Section~\textsc{\nameref{sec:2-background}} introduces the necessary background on synchronization protocols.
+Section~\textsc{\nameref{sec:3-features}} elaborates on our design that enables this flexibility.
+In Section~\textsc{\nameref{sec:4-performance}}, we evaluate performance of our tool by comparing its different synchronization protocols, and compare to adevs.
+We continue by introducing runtime switching of synchronization protocols and different options for model allocation in Section~\textsc{\nameref{sec:4b-hotswap}} and Section~\textsc{\nameref{sec:4a-allocation}}, respectively.
+Related work is discussed in Section~\textsc{\nameref{sec:5-related-work}}.
+Section~\textsc{\nameref{sec:6-conclusion}} concludes the paper and gives future work.

### 2016-06-29 publications/Journal/parallel-performance.tex
Updated performance section and made consistent with new structure

@@
 \subsection{Parallel Simulation}
+We now continue by describing our parallel simulation performance for different synchronization protocols, and compared to adevs.
+This is done for our three benchmark models.
 
-% The good
 \subsubsection{Queue}
-\paragraph*{Allocation}
-The Queue model is one single chain of models. Each kernel gets one connected part of this chain. The result is that the kernels themselves also form a chain where events only travel in one direction.
+The Queue model is one single chain of models, resembling a pipeline.
+This structure can be exploited to prevent cyclic dependencies in the parallel simulation.
 
-\paragraph*{Strong and Weak Scaling}
 Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to dxex sequential for a fixed problem size.
@@
 
-%In Figure \ref{fig:Queue_allocation} the allocation of the Queue model is visualized.
-This trace allows us to demonstrate the benchmark results and explain why optimistic is not ideal for such a chain topology as present in Queue.
-Kernel 2 is dependent on events from kernels 0 and 1 but by definition of the optimistic synchronization protocol it will simulate ahead without waiting for events it depends on.
-Any simulation it performs will have to be reverted as soon as events from kernels 0 and 1 arrive.
-When kernel 2 reverts, kernel 3 will inevitably have to revert as well since most of the events it received from kernel 2 are invalidated and marked as such by antimessages.
-This leads to severe performance degradation, with an increasing probability of cascading reverts as the number of kernels and models per kernels increases.
 The speedup of adevs is always in comparison with the runtime of the corresponding dxex sequential benchmark.
@@
 
-% The really bad
 \subsubsection{Interconnect}\label{subsec:parallelinterconnect}
@@
 All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
-In Interconnect there is no allocation scheme possible that avoids cyclic dependencies between simulation kernels, as shown in the trace \ref{fig:interconnect_allocation_parallel} of a simulation with 4 models.
-Such a cycle forces sequential operation of the kernels with no speedup possible.
 
@@
 \end{figure}
-\begin{figure}
-	\center
-	\includegraphics[width=\columnwidth]{fig/interconnect_parallel_allocation.eps}
-	\caption{Interconnect parallel simulation trace for 6 models on 3 kernels.}
-	\label{fig:interconnect_allocation_parallel}
-\end{figure}
 
-% The okayish
 \subsubsection{Phold}
-% Same cause, different results
 In the Phold model, we first investigate the influence of the percentage of remote events on the speedup.
@@
 Even though a single synchronization protocol is always ideal in this case, it already shows that different synchronization protocols respond differently to a changing model.
+
 Adevs is significantly slower during conservative synchronization.
@@
 This situation closely mimics the case made in the comparison between both synchronization algorithms by~\cite{FujimotoBook}.
-In Figure \ref{fig:phold_allocation} it is clear that in Phold it is possible for dependency cycles to form between kernels which as we have shown in Interconnect degrades performance for both optimistic and conservative.
-This is also the cause of the sublinear speedup observed in our Phold benchmark.
 

### 2016-06-29 publications/Journal/sequential-performance.tex
Fixed figures for twocolumn
Updated performance section and made consistent with new structure

@@
 \subsection{Sequential Simulation}
-% 2 things to show : baseline performance and contrast of dxex v adevs
+We start by evaluating sequential simulation performance, in order to obtain a baseline for our comparison of parallel simulation performance.
 
@@
 	\center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/queue_fixed_sequential.eps}
+	\includegraphics[width=\columnwidth]{fig/queue_fixed_sequential.eps}
 	\caption{Queue benchmark results for sequential simulation.}
@@
 	\center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_sequential.eps}
+	\includegraphics[width=\columnwidth]{fig/interconnect_sequential.eps}
 	\caption{Interconnect benchmark results for sequential simulation.}
@@
 	\center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/phold_sequential.eps}
+	\includegraphics[width=\columnwidth]{fig/phold_sequential.eps}
 	\caption{PHold benchmark results for sequential simulation.}

### 2016-06-29 publications/Journal/models.tex
Added another reference

@@
 In this section, we evaluate the performance of different synchronization protocols in dxex.
-We also compare to adevs, currently one of the most efficient simulation kernels~\cite{DEVSSurvey}, to show that our modularity does not impede performance.
+We also compare to adevs~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVSSurvey,DEVStoneJournal}, to show that our modularity does not impede performance.
 CPU time and memory usage is compared for both sequential and parallel simulation.

### 2016-06-29 publications/Journal/3-features.tex
Added references

@@
 
-We primarily investigate this for optimistic simulation, as this is the most memory consuming mode of simulation~\cite{Fujimoto}.
+We primarily investigate this for optimistic simulation, as this is the most memory consuming mode of simulation~\cite{FujimotoBook}.
 Simulation execution times for all four combinations are shown in Figure~\ref{fig:memallocators_parallel}.
@@
 All accesses to the RNG will result in the modification of a global (\textit{i.e.}, shared between threads) variable.
-This easily becomes a bottleneck in simulation, since random numbers are a common occurence in simulation~\cite{?}.
+This easily becomes a bottleneck in simulation, since random numbers are a common occurence in simulation~\cite{Random}.
 As such, a non-trivial amount of time in a simulation is often spent waiting for an RNG.
@@
 We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default STL random number generator.
-%TODO fill in values
 This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.

### 2016-06-29 publications/Journal/4a-allocation.tex
Added references

@@
 
-The PHOLDTree model is similar in structure to models of gossiping in social networks~\cite{?}, hierarchical resource sharing~\cite{?} or load balancing~\cite{?}.
+The PHOLDTree model is similar in structure to models of gossiping in social networks~\cite{Gossip} and multi-granularity locking~\cite{MultiGranularityLocking}.
 The lookahead of an atomic node is the minimally allowed $\epsilon$, indicating uncertainty, as is often the case in realistic models.

### 2016-06-29 publications/Journal/4a-allocation.tex
Add another missing reference
Fixed missing references

@@
 \subsection{Performance Evaluation}
-%TODO do basic performance evaluation of allocation
-%TODO show model in statistics view
 In order to evaluate the influence of model allocation, we define a new model, based on PHOLD~\cite{PHOLD}.
@@
     \caption{Effect of hierarchy in sequential simulation.}
-    \label{fig:PHoldtree_seq_dn_benchmark}
+    \label{fig:PHOLDtree_seq_dn_benchmark}
 \end{figure}
@@
 This means, intuitively, that atomic models at the same level in the tree, but not necessarily siblings, are frequently allocated to the same node.
-Since there is only infrequently some communication between 
-This allocation strategy is shown in Figure~\ref{fig:BF}.
+Since there is only infrequently some communication between siblings, and even never between different subtrees, this does not sound an intuitive allocation.
+This allocation strategy is shown in Figure~\ref{fig:PholdTree_model_bfs}.
 
@@
 This means, intuitively, that subtreees are frequently allocated to the same node.
-This allocation strategy is shown in Figure~\ref{fig:DF}.
+This allocation strategy is shown in Figure~\ref{fig:PholdTree_model_dfs}.
 
@@
 This effect is also apparent if the thread a kernel runs on is not fairly scheduled.
-In conservative this will lead to excessive polling of the other kernels' eot values, in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel.
-In Figures \ref{fig:pholdtree_visualize_parBFS} and \ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for both allocation schemes highlighting the remarks made in this paragraph.
+In conservative this will lead to excessive polling of the other kernels' EOT values, in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel.
+In Figure~\ref{fig:pholdtree_visualize_parBFS} and Figure~\ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for breadth-first and depth-first, respectively, highlighting the remarks made in this paragraph.
+
 \begin{figure}

### 2016-06-29 publications/Journal/4b-hotswap.tex
Shorter title for section
Don't just iterate over figures

@@
-%TODO explain hotswapping
-%TODO why hotswapping
 Simply because a synchronization protocol is ideal at the start of the simulation, does not mean that it will still be ideal during the simulation.
@@
 For each connection, the number of events transferred is annotated.
-Examples are shown for the three benchmark models used before: Figures~\ref{fig:Queue_allocation},~\ref{fig:interconnect_allocation_parallel}, and~\ref{fig:phold_allocation} shows traces for the Queue, Interconnect, and PHOLD models respectively.
+Examples are shown for the three benchmark models used before: Figure~\ref{fig:Queue_allocation}, Figure~\ref{fig:interconnect_allocation_parallel}, and Figure~\ref{fig:phold_allocation} show traces for the Queue, Interconnect, and PHOLD models respectively.
 

### 2016-06-29 publications/Journal/memory.tex
Add missing file (subsection on memory)

--- 2016-02-15 paper/4-performance.tex
@@
-\newcommand{\plotfraction}{0.9}
-
-In this section, we evaluate the performance of different synchronization protocols in dxex.
-We also compare to adevs, currently one of the most efficient simulation kernels~\cite{DEVSSurvey}, to show that our modularity does not impede performance.
-CPU time and memory usage is compared for both sequential and parallel simulation.
-
-We start off with a comparison of sequential simulation, to show how adevs and dxex relate in this simple case.
-For the parallel simulation benchmarks, results are presented for both conservative and optimistic synchronization.
-
-For all benchmarks, results are well within a 5\% deviation of the average, such that only the average is used in the remainder of this section.
-The same compilation flags were used for both adevs and dxex benchmarks (``\texttt{-O3 -flto}'').
-To guarantee comparable results, no I/O was performed during benchmarks.
-Before benchmarking, simulation traces were used to verify that adevs and dxex return exactly the same simulation results.
-Benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.
-The exact parameters for each benchmark can be found in the repository, as well as the data used in this paper.
-
-\subsection{Benchmarks}
-We use three different benchmarks, which cover different aspects of the simulation kernel:
-\begin{enumerate}
-    \item The \textit{Queue} model, based on the \textit{HI} model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
-          A single generator pushes events into the queue, which are processed by the processors after a random delay.
-          It takes two parameters: width and depth, which determine the width and depth of the hierarchy.
-          This benchmark shows how the complexity of the simulation kernel behaves for an increasing amount of atomic models, and an increasingly deep hierarchy.
-          An example for width and depth 2 is shown in Figure~\ref{fig:queue_model}.
-
-    \item The \textit{PHOLD} model, presented by~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
-          Each atomic model is directly connected to every other atomic model.
-          After a random delay, an atomic model sends out an event to a randomly selected output port.
-          Output port selection happens in two phases: first it is decided whether the event should be sent to an atomic model at the same node.
-          Afterwards, a uniform selection is made between the remaining ports.
-          The model takes one parameter: the percentage of remote events, which determines the fraction of messages routed to other nodes.
-          This benchmark shows how the simulation kernel behaves in the presence of many local or remote events.
-          An example for four models, split over two nodes, is shown in Figure~\ref{fig:PHOLD_model}.
-
-    \item The \textit{HighInterconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
-          Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event.
-          The model takes one parameter: the number of models.
-          This benchmark investigates the complexity of event routing, and how the simulation kernel handles many simultaneous events.
-          An example for four models is shown in Figure~\ref{fig:interconnect_model}.
-\end{enumerate}
-
-\newcommand{\modelfraction}{0.65}
-
-\begin{figure}
-    \center
-    \includegraphics[width=\columnwidth]{fig/queue_model.pdf}
-    \caption{Queue model for depth and width 2.}
-    \label{fig:queue_model}
-
-    \includegraphics[width=\modelfraction\columnwidth]{fig/interconnect_model.pdf}
-    \caption{HighInterconnect model for three models.}
-    \label{fig:interconnect_model}
-
-    \includegraphics[width=\modelfraction\columnwidth]{fig/phold_model.pdf}
-    \caption{PHOLD model for three models.}
-    \label{fig:PHOLD_model}
-\end{figure}
-
-\subsection{Sequential Simulation Execution Time}
-Despite our core contribution being on parallel simulation, we still value a comparison of sequential simulation results.
-First, and foremost, parallel simulation results are tightly linked to sequential simulation results.
-Parallel simulation is achieved through synchronization of multiple sequential simulation kernels.
-Second, parallel simulation results are validated through the use of adevs.
-To provide a well-founded comparison to adevs in the parallel simulation benchmarks, sequential simulation results also need to be compared.
-
-Only the Queue and HighInterconnect models are relevant for sequential simulation, so we will not touch upon the PHold model yet.
-
-\subsubsection{Queue}
-In the Queue model, we increase both width and depth simultaneously.
-For example, the $400$ models configuration is obtained with a width and depth of $20$.
-As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
-Through profiling, we found that adevs spends much time routing the simulation messages (\textit{e.g.}, \textit{*}, \textit{@}) through the hierarchy, whereas dxex avoids this through its flattened model structure.
-Both simulation tools have similar complexities, though dxex is much faster thanks to its more efficient simulation control algorithms.
-
-\begin{figure}
-    \center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/queue_sequential.eps}
-	\caption{Queue benchmark results for sequential simulation.}
-	\label{fig:Queue_benchmark}
-\end{figure}
-
-\subsubsection{HighInterconnect}
-In the HighInterconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
-As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.
-Analysis showed that this is caused by the high amount of exchanged events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
-
-\begin{figure}
-    \center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_sequential.eps}
-	\caption{Interconnect benchmark results for sequential simulation.}
-	\label{fig:Interconnect_benchmark}
-\end{figure}
-
-\subsection{Parallel Simulation Execution Time}
-Next we analyse parallel simulation of our previously defined benchmarks.
-For dxex, we mention results of both conservative and optimistic synchronization.
-Since adevs supports only conservative synchronization, we don't mention optimistic synchronization results there.
-All experiments were performed using up to six simulation nodes, executed on a hexa-core machine.
-
-We highlight two main results:
-(1) dxex conservative synchronization is competitive with adevs;
-(2) dxex optimistic synchronization is sometimes more efficient than conservative synchronization.
-This shows that our contribution, offering both conservative and optimistic synchronization, is indeed beneficial for a general-purpose simulation tools.
-
-\subsubsection{Queue}
-In the Queue model, we allocate the chain of models such that each node is responsible for a series of connected models.
-This minimizes the number of inter-node messages.
-As the model is a queue, however, models further in the chain only activate later in the simulation.
-Since these are allocated to separate nodes, some nodes remain idle until simulation has progressed sufficiently far.
-
-Similar to the sequential benchmarks, Figure~\ref{fig:queue_benchmark_parallel} shows that dxex outperforms adevs, but now in terms of speedup.
-Results indicate that our implementation of conservative synchronization achieves much higher speedups than adevs.
-Approaching near linear speedup, this simulation is the ideal case for our conservative implementation, as there are no dependency cycles between models.
-Conservative synchronization also seems to be better than optimistic synchronization in this case, at the cost of providing the lookahead.
-
-\begin{figure}
-    \center
-	\includegraphics[width=\plotfraction\columnwidth]{fig/queue_parallel.eps}
-	\caption{Queue benchmark results for parallel simulation of a model of fixed size.}
-	\label{fig:queue_benchmark_parallel}
-\end{figure}
-
-\subsubsection{PHold}
-In the Phold model, we first investigate the influence of the fraction of remote events on the speedup.
-When remote events are rare, optimistic synchronization rarely has to roll back, thus increasing performance.
-With more common remote events, however, optimistic synchronization quickly slows down due to frequent rollbacks.
-Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block.
-Even though a single synchronization protocol is always ideal in this case, it already shows that different synchronization protocols respond differently to a changing model.
-Adevs is again significantly slower during conservative synchronization.
-Analysis of profiling results shows that exception handling in adevs is the main cause. 
-To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.
-
-\begin{figure}
-    \center
-    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_speedup_remotes.eps}
-    \caption{Phold benchmark results for parallel simulation using six cores, 4 atomics per node, with varying fraction of remote events.}
-\end{figure}
-
-We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization.
-We slightly modified the Phold benchmark, to include high-priority events.
-Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
-Even when normal events occur most often, conservative synchronization always blocks until it can make guarantees.
-Optimistic synchronization, however, simply goes forward in simulation time and rolls back when these high-priority events happen.
-This situation closely mimics the case made in the comparison between both synchronization algorithms by~\cite{FujimotoBook}.
-
-Figure~\ref{fig:phold_priority} shows how simulation performance is influenced by the fraction of these high-priority events.
-If barely any high-priority events occur, conservative synchronization is penalized due to its excessive blocking, which often turned out to be unnecessary.
-When many high-priority events occur, optimistic synchronization is penalized due to its mindless progression of simulation, which frequently needs to be rolled back.
-Results show that there is no single perfect synchronization algorithm for this model: depending on configuration, either synchronization protocol might be better.
-We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
-
-\begin{figure}
-    \center
-    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_speedup_priority.eps}
-    \caption{Phold benchmark results for parallel simulation using six cores, with varying amount of high-priority events.}
-    \label{fig:phold_priority}
-\end{figure}
-
-\subsubsection{Interconnect}
-In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
-The number of models is now kept constant at eight.
-Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
-When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing number of of rollbacks in optimistic simulation.
-All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
-
-\begin{figure}
-    \center
-    \includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_parallel.eps}
-    \caption{Interconnect benchmark results for parallel simulation.}
-    \label{fig:interconnect_benchmark_parallel}
-\end{figure}
-
 \subsection{Memory Usage}

### 2016-06-29 publications/Journal/1-introduction.tex
Rewording/rephrasing in first two sections

@@
 But with the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform the simulation within reasonable time bounds.
-And while \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, this is often insufficient.
-Several synchronization protocols from the discrete event simulation community~\cite{FujimotoBook} have been applied to \textsf{DEVS} simulation.
+Whereas \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, its inherent parallelism is often insufficient.
+Several synchronization protocols from the discrete event simulation community~\cite{FujimotoBook} have been applied to \textsf{DEVS} simulation~\cite{globaltimewarp}.
+Through synchronization protocols, different simulation cores can be at different points in simulated time, increasing parallelism significantly, at the cost of synchronization overhead.
 While several parallel \textsf{DEVS} simulation kernels exist, they are often limited to a single synchronization protocol.
 The reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other in specific models.
-The applicability of parallel simulation capabilities of current tools is therefore limited.
+The applicability of parallel simulation capabilities of current tools is therefore constrained to specific domains.
 
 This paper introduces DEVS-Ex-Machina\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``dxex''): our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
-The selected synchronization protocol is transparent to the simulated model: users only determine which protocol they wish to use.
-Users who simulate a wide variety of models, with different ideal synchronization protocols, can simply run the same model with different synchronization protocols.
+The selected synchronization protocol is transparent to the simulated model: users only determine which protocol to use.
+Users who simulate a wide variety of models, with different ideal synchronization protocols, can run the same model with different synchronization protocols.
 Since the ideal synchronization protocol might change throughout the simulation, we also support runtime switching between these protocols.
@@
 Our tool is based on PythonPDEVS, but implemented in C++11 for increased performance, using features from the new C++14 standard when supported by the compiler.
-Unlike PythonPDEVS, dxex only supports multicore parallelism.
+Unlike PythonPDEVS, dxex only supports multicore parallelism, thus no distributed simulation.
 

### 2016-06-29 publications/Journal/3-features.tex
Proofread section 3

@@
 Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
-Dxex is a C++11-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, while making some of its own additions.
-So while the feature set is not too comparable, the architectural design, core simulation algorithm, and optimizations, are highly similar.
+Dxex is a C++11-based \textsf{Parallel DEVS} simulator, based on the design of PythonPDEVS.
+Whereas the feature set is not too comparable, the architectural design, simulation algorithms, and optimizations, are highly similar.
 
 We will not make a detailed comparison with PythonPDEVS here, but only mention some supported features.
-Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
+Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions~\cite{JDF}, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
+We do not elaborate on these features in this paper.
 But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols (though only in parallel).
@@
 Currently, most parallel simulation tools choose only a single synchronization protocol due to the inherent differences between protocols.
-An uninformed choice on which one to implement is insufficient, as performance will likely be bad.
 We argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for dxex.
@@
 It shows that there is a simulation \texttt{Core}, which simulates the \texttt{AtomicModel}s connected to it.
-The superclass \texttt{Core} is merely the sequential simulation core, but can be used as-is.
+The superclass \texttt{Core} represents the sequential simulation core.
 Subclasses define specific variants, such as \texttt{ConservativeCore} (conservative synchronization), \texttt{OptimisticCore} (optimistic synchronization), and \texttt{DynamicCore} (\textsf{Dynamic Structure DEVS}).
@@
 Within the returned time interval, the model promises not to raise an event.
-A node aggregates this information to compute its earliest output time (EOT).
-This value is written out in shared memory, where it can be read out by all other nodes.
+A node aggregates this information to compute its Earliest Output Time (EOT).
+This value is communicated to other nodes through shared memory.
 
@@
 Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
-C++11 further allows us to make the implementation portable, as well as more efficient: the compiler might know of optimizations specific to atomic variables or constant expressions which are heavily used in dxex.
+C++11 further allows us to make the implementation portable, as well as more efficient: the compiler makes further optimizations to heavily used components.
 
@@
 This made us switch to the use of thread-local (using \textit{tcmalloc}) memory pools.
-Again, we made use of specific new features of C++11, that were not available in Python, or even previous versions of the C++ language standard.
+Again, we made use of specific new features of C++11, that are not available in Python, or even previous versions of the C++ language standard.
 
@@
 We define synchronization protocol transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
-User should thus only provide one model, implemented in C++11, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.
-Switching between simulation kernels is as simple as altering the simulation termination time.
+User should thus only provide one model, implemented in C++11, which can be simulated using sequential execution, conservative synchronization, or optimistic synchronization.
+The synchronization protocol to use is a simple configuration option.
 The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
@@
 
-Always defining a lookahead function might seem redundant, especially if users will never use conservative synchronization.
-Especially since defining the lookahead is often non-trivial and dependent on intricate model details.
+Always defining a lookahead function might seem redundant, especially should the user never want to use conservative synchronization.
+Definig the lookahead is difficult and can often be unpredictable.
 The more attractive option is for the simulation tool to provide a default lookahead function, such that simulation can run anyway, but likely not at peak performance.
@@
 While this is always a problem, it is aggravated in dxex by providing automatic memory management for events and states.
-A model written for a sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the model author) the (de)allocation semantics of events or states.
+A model written for sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.
 
@@
 To counter this, our memory allocators are backed by a thread-aware pooling library.
-In a sequential simulation kernel no allocated event will persist beyond a single time advance, even allowing the use of an arena-style allocator.
+In a sequential simulation kernel no allocated event persists beyond a single time advance, even allowing the use of an arena-style allocator.
 Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.
@@
 Intra-kernel events are pooled aggressively, whereas inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
-A simulation with many inter kernel events suffers a performance hit, whereas the impact of many intra kernel events can be minimized using arena allocators.
+A simulation with many inter kernel events suffers a performance hit, whereas the impact of many intra-kernel events can be minimized using arena allocators.
 

### 2016-06-29 publications/Journal/4a-allocation.tex
Phold and PHOLD consistently

@@
    \includegraphics[width=\columnwidth]{fig/pholdtree_alloc_BF.pdf}
-   \caption{PholdTree model breadth first allocation with 4 kernels.}
+   \caption{PHOLDTree model breadth first allocation with 4 kernels.}
    \label{fig:PholdTree_model_bfs}
    \includegraphics[width=\columnwidth]{fig/pholdtree_alloc_DF.pdf}
-   \caption{PholdTree model depth first allocation with 4 kernels.}
+   \caption{PHOLDTree model depth first allocation with 4 kernels.}
    \label{fig:PholdTree_model_dfs}
@@
     \includegraphics[width=\columnwidth]{fig/pholdtreed1n3t5000c4BFS.eps}
-    \caption{Visualization of a PholdTree simulation with breadth first allocation and 4 kernels.}
+    \caption{Visualization of a PHOLDTree simulation with breadth first allocation and 4 kernels.}
     \label{fig:pholdtree_visualize_parBFS}
@@
     \includegraphics[width=\columnwidth]{fig/pholdtreed1n3t5000c4DFS.eps}
-    \caption{Visualization of a PholdTree simulation with depth first allocation and 4 kernels.}
+    \caption{Visualization of a PHOLDTree simulation with depth first allocation and 4 kernels.}
     \label{fig:pholdtree_visualize_parDFS}
@@
     \includegraphics[width=\columnwidth]{fig/pholdtreeallochighp.eps}
-    \caption{PholdTree model performance using different allocation schemes.}
+    \caption{PHOLDTree model performance using different allocation schemes.}
     \label{fig:PholdTree_plot_alloc_high}

### 2016-06-29 publications/Journal/parallel-performance.tex
Phold and PHOLD consistently

@@
 We now continue by describing our parallel simulation performance for different synchronization protocols, and compared to adevs.
-This is done for our three benchmark models.
+The speedup of adevs is computed with the corresponding dxex sequential benchmark.
+This was done to take into account the performance difference observed in sequential simulation.
 
@@
 
-Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to dxex sequential for a fixed problem size.
-As the amount of kernels increases, the optimistic kernel quickly becomes the worst choice.
-The difference between dxex conservative and adevs becoming smaller.
+Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to sequential simulation for a fixed problem size.
+%TODO this is actually quite a strange choice...
+As the number of kernels increases, the optimistic kernel quickly becomes the worst choice.
+This is mainly caused by the pipeline structure of the model: the last models in the queue only respond to incoming messages and therefore have to be rolled back frequently.
+The difference between dxex conservative and adevs becoms smaller when more and more cores are used.
 The same effect can be seen for weak scaling in Figure~\ref{fig:Queue_plot_weak}.
 
-The speedup of adevs is always in comparison with the runtime of the corresponding dxex sequential benchmark.
 \begin{figure}
 	\center
-
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_weak_speedup.eps}
@@
 	\label{fig:Queue_plot_weak}
+\end{figure}
 
+\begin{figure}
+	\center
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_strong_speedup.eps}
@@
 	\label{fig:Queue_plot_strong}
-
 \end{figure}
@@
 Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
-When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing number of of rollbacks in optimistic simulation.
+When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and the increasing number of rollbacks in optimistic simulation.
 All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
@@
 
-\subsubsection{Phold}
-In the Phold model, we first investigate the influence of the percentage of remote events on the speedup.
+\subsubsection{PHOLD}
+In the PHOLD model, we first investigate the influence of the percentage of remote events on the speedup.
 A remote event in this context is an event that is sent from a model on one kernel to a model on another simulation kernel.
@@
 To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.
-These functions require the user to implement a state saving in contrast to PythonPDEVS and dxex's optimistic kernels which handle this inside the kernel.
-We feel this would lead to an unfair comparison as we would like to keep the models synchronization-agnostic across all benchmarks.
+These functions require the user to implement a state saving method.
+But in contrast to PythonPDEVS and dxex, which handle this inside the kernel, users need to manually define this.
+We feel this would lead to an unfair comparison as we would like to keep the models agnostic of the underlying protocols across all benchmarks.
 
@@
     \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
-    \caption{Phold benchmark results for parallel simulation using four kernels, four atomics per node, with varying percentage of remote events.}
+    \caption{PHOLD benchmark results for parallel simulation using four kernels, four atomics per node, with varying percentage of remote events.}
 \end{figure}
@@
 	\includegraphics[width=\columnwidth]{fig/phold_priority.eps}
-	\caption{Phold benchmark results for parallel simulation using four kernels, with varying amount of high-priority events.}
+	\caption{PHOLD benchmark results for parallel simulation using four kernels, with varying amount of high-priority events.}
 	\label{fig:phold_priority}
@@
 
-We slightly modified the Phold benchmark, to include high-priority events.
+We slightly modified the PHOLD benchmark, to include high-priority events.
 Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.

### 2016-06-29 publications/Journal/1-introduction.tex
Add reference to our SpringSim paper in the beginning

@@
 
-This paper introduces DEVS-Ex-Machina\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``dxex''): our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
+This paper introduces DEVS-Ex-Machina\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``dxex''): our simulation tool~\cite{dxex} which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
 The selected synchronization protocol is transparent to the simulated model: users only determine which protocol to use.

### 2016-06-29 publications/Journal/4a-allocation.tex
Proofread allocation

@@
-Although the synchronization protocol is one of the defining factors in simulation performance, model allocation has a significant impact on which protocol is ideal, and even whether or not parallelization would make sense.
-Indeed, if the model is distributed in such a way that frequent communication is necessary between cores, parallelism is naturally reduced.
-This thus brings us to the topic of model allocation.
-Model allocation was one of the features also implemented by PythonPDEVS~\cite{PythonPDEVS2}.
+Although the synchronization protocol is one of the defining factors in simulation performance, model allocation has a significant impact on which protocol is ideal.
+Depending on the model structure, and how it is mapped to the different cores, it might not even make sense to parallelize at all.
+Indeed, if the model is distributed such that frequent communication is necessary between cores, parallelism is naturally reduced.
+This brings us to the topic of model allocation, as also implemented in PythonPDEVS~\cite{PythonPDEVS2}.
 
@@
 This is handled by the default model allocator.
-If no preference is given a simple striping scheme is followed but this is not sufficient in most simulations to achieve a speedup in parallel.
-By overriding the default allocator a modeller tunes the allocation scheme for a specific model, maximizing parallel speedup.
-This interface can be used to employ graph algorithms for an automatic allocation scheme, for example avoiding cycles in the dependency graph.
+If no preference is given, a simple striping scheme is used but this is often insufficient.
+By overriding the default allocator, a modeller tunes the allocation scheme for a specific model, maximizing parallel speedup.
+This interface can be linked to graph algorithms for automatic allocation scheme generation, for example to avoid cycles in the dependency graph.
 
 \subsection{Performance Evaluation}
-In order to evaluate the influence of model allocation, we define a new model, based on PHOLD~\cite{PHOLD}.
-The model structure resembles a tree: an atomic model can have a set of children, with children being connected to each other as well.
-Connections can be uni- or bidirectional.
+To evaluate the influence of model allocation, we define a new model, based on PHOLD~\cite{PHOLD}.
+The model structure resembles a tree: an atomic model can have a set of children, with children being connected to each other recursively.
 
-Unlike the Queue model, the width of the hierarchy is still present in the topology of the atomic models after the direct connection stage.
+Unlike the Queue model, the width of the hierarchy is still present in the topology of the atomic models after direct connection.
 The PHOLDTree model allows us to investigate parallel speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.
@@
 The lookahead of an atomic node is the minimally allowed $\epsilon$, indicating uncertainty, as is often the case in realistic models.
-We demonstrate the importance of allocation, by comparing a breadth-first versus a depth-first scheme.
+We demonstrate the importance of allocation by comparing performance of a breadth-first versus a depth-first scheme.
+Both options are automated ways of allocation that are independent of the model.
 
@@
 
-PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain as in Queue.
+PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain, as was the case in the Queue model.
 This topology is interesting since it highlights the effects of allocation.
-This model allows us to investigate in depth the effects of non-cyclic allocation strategies and measure parallel speedup.
 First, we evaluate the model in sequential simulation to provide a baseline for parallel simulation.
@@
 Since adevs does not use direct connection, we expect a noticable performance difference between dxex and adevs.
-This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout determines the performance penalty adevs suffers compared to dxex.
-Profiling indeed indicated that an increase in width per subtree ($n$) leads to higher overheads in adevs due to the lack of direct connection.
-Dxex uses direct connection, making it independent on fanout, but only of the number of models.
+This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout ($n$) determines the performance penalty adevs suffers compared to dxex.
+Profiling indeed indicates that an increase in width per subtree ($n$) leads to higher overheads in adevs due to the lack of direct connection.
+Dxex uses direct connection, making it independent on fanout.
+Performance of dxex is, in this model, only dependent on the number of models.
 Slight deviations can still be seen, though, caused by the initialization overhead of direct connection.
 Both adevs and dxex scale linearly in the number of atomic models.
-We expect to see a similar distinction in parallel simulation.
 
@@
 
-The breadth first allocation scheme will result in kernels that will form a dependency chain with multiple branches, much like in the Queue model.
-Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model, but this is not always true as we will demonstrate in this section.
-A single kernel that has an unbalanced number of atomic models or unequal computation load in transition functions will slow down the remainder of the chain.
-This effect is also apparent if the thread a kernel runs on is not fairly scheduled.
-In conservative this will lead to excessive polling of the other kernels' EOT values, in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel.
-In Figure~\ref{fig:pholdtree_visualize_parBFS} and Figure~\ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for breadth-first and depth-first, respectively, highlighting the remarks made in this paragraph.
-
 \begin{figure}
    \center
-
    \includegraphics[width=\columnwidth]{fig/pholdtree_alloc_BF.pdf}
@@
    \label{fig:PholdTree_model_bfs}
+\end{figure}
+\begin{figure}
+   \center
    \includegraphics[width=\columnwidth]{fig/pholdtree_alloc_DF.pdf}
@@
 
-After simulation, the traces can be visualized for both breadth-first and depth-first allocation.
+The breadth first allocation scheme results in a dependency chain with multiple branches, much like in the Queue model.
+Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model.
+This is not always true though: a single kernel with an unbalanced number of atomic models or unequal computational load in transition functions slows down the remainder of the chain.
+This effect is also apparent if the thread a kernel runs on is not fairly scheduled.
+With conservative synchronization this leads to excessive polling of the EOT of the other kernels.
+With optimistic synchronization this leads to a cascade of rollbacks, since dependent kernels will simulate ahead of the slower kernel.
+
+After simulation the traces can be visualized for both breadth-first and depth-first allocation.
 Using a breadth-first allocation scheme, as shown in Figure~\ref{fig:pholdtree_visualize_parBFS}, we notice that many events get exchanged between cores.
@@
 We see that for both synchronization protocols, the depth-first allocation is significantly better than breadth-first allocation.
-This is what we expected: depth-first allocation maintains locality better than breadth-first allocation.
+This is what we expected for this model: depth-first allocation maintains locality better than breadth-first allocation.
+Whereas this is the case in this example, this is not true in general, as the ideal allocation depends on the model being simulated.
 

### 2016-06-29 publications/Journal/5-hotswap.tex
Rename files to fix the order again

--- 2016-06-29 publications/Journal/4b-hotswap.tex
@@
 Simply because a synchronization protocol is ideal at the start of the simulation, does not mean that it will still be ideal during the simulation.
-It has been frequently shown, and repeated in the previous section, that model behaviour significantly influences the ideal synchronization protocol.
+It is well known, and repeated in the previous section, that model behaviour significantly influences the ideal synchronization protocol.
 Contrary to many modelling formalisms, the DEVS formalisms makes it possible to model basically any kind of discrete event model.
@@
 
-Defining the ideal synchronization protocol at the start of the simulation, when information about future model behaviour might be scarce, might therefore not offer the best possible performance.
+Defining the ideal synchronization protocol at the start of the simulation, when information about future model behaviour is scarce, might therefore not offer the best possible performance.
 In dxex, we not only make it possible to define the synchronization protocol to use, but also to change this decision throughout simulation.
-To do this, all kernels are notified of the switch and they are forced to stop simulation using the current synchronization core.
-When stopped, each kernel instantiates a new core that is provided with the simulation state of the previous core.
+To do this, all kernels are notified of the switch and they are forced to stop simulation.
+When stopped, each kernel instantiates a new core, with the new synchronization protocol, that is provided with the simulation state of the previous core.
 Simulation is then resumed with the new cores after the previous ones are destroyed.
@@
 This overhead depends on the size of the model and the number of simulation cores.
-For a simple model and a few cores, the overhead is less than a single second.
+For a simple model and a few cores, the overhead is less than a second.
 
@@
 Ideally, a new component is added to the simulation kernel, which monitors model behaviour and simulation performance, and toggles between them automatically.
-Our interface is thus augmented with the necessary bindings for such a component to be defined.
+Our interface is augmented with the necessary bindings for such a decision component.
 Also, our interface is augmented with an interface for statistics gathering and model behaviour analysis.
-The implementation of such a component is currently left open: algorithms can be heavily based on machine learning or similar approaches.
+The implementation of such a component is currently left open, but we envision algorithms heavily based on machine learning.
 
@@
 We add performance metrics in the simulation kernel, which logs relevant performance metrics and processes them for use in other components.
-These metrics include the number of events created and destroyed, the number of inter and intra kernel events, the number of rollbacks, the measured lookahead, details of the Global Virtual Time (GVT) and Earliest Output Time (EOT) calculations, and basic information on the fairness between different simulation kernels.
+These metrics include the number of events created and destroyed, the number of inter- and intra-kernel events, the number of rollbacks, the measured lookahead, details of the Global Virtual Time (GVT) and Earliest Output Time (EOT) calculations, and information on the fairness between simulation kernels.
 With all these metrics, a component can get a fair view on both model and simulation kernel behaviour.
@@
 For example, if the actually seen lookahead is significantly higher than the defined lookahead, it might be interesting to switch to optimistic synchronization.
-When the number of rollbacks is very high, conservative synchronization might be ideal.
+When the number of rollbacks is excessively high, switching to conservative synchronization might be considered.
 And when neither of these two is an option, the simulation might just (temporarily) have to fall back to sequential simulation.
 
-For performance reasons, statistics gathering is optional due to the imposed overhead.
-
 \subsubsection{Visualization of Communication}
-To provide some more insight in the models we used as benchmarks previously, we created a simple visualization of the simulation trace.
+To provide more insight in our benchmark models, we created a simple visualization of their simulation trace.
 This trace visualizes the allocation of the model and all defined connections.
@@
 Examples are shown for the three benchmark models used before: Figure~\ref{fig:Queue_allocation}, Figure~\ref{fig:interconnect_allocation_parallel}, and Figure~\ref{fig:phold_allocation} show traces for the Queue, Interconnect, and PHOLD models respectively.
+Using this information, we notice that the Interconnect benchmark indeed has a lot of inter-kernel events.
+Despite the similar structure, the PHOLD model does not have as many inter-kernel events.
+These results are relevant information that can be used by the hotswapping component.
 
@@
 \end{figure}
-
-Naturally, results similar to this are relevant information that can be used by the hotswapping component.

### 2016-06-29 publications/Journal/8-conclusions.tex
Rename files to fix the order again

--- 2016-06-29 publications/Journal/6-conclusions.tex
@@
 Future work is possible in several directions.
-Firstly, our implementation of optimistic synchronization should be more tolerant to low-memory situations.
+First, our implementation of optimistic synchronization should be more tolerant to low-memory situations.
 In its current state, simulation will simply halt with an out-of-memory error.
@@
 Faster GVT implementations~\cite{Fujimoto:1997:CGV:268403.268404,Bauer:2005:SND:1069810.1070159} might further help to minimize this problem.
-Secondly, the runtime switching between synchronization protocols can be driven using machine learning techniques. The simulation engine is already capable of collecting data to inform such a process.
-Thirdly, activity algorithms, as already implemented by PythonPDEVS, can also be implemented in dxex, to determine how they influence simulation performance.
-Finally automatic allocation is possible by analysis of the connections between models.
+Second, the runtime switching between synchronization protocols can be driven using machine learning techniques.
+The simulation engine is already capable of collecting data to inform such a process, and is designed to listen for commands from an external component.
+Third, automatic allocation might be possible by analysis of the connections between models.
 This information is already used in dxex to construct the dependency graph in conservative synchronization.
-A graph algorithm that distributes models while avoiding cycles in the resulting kernel topology could be used to offer a parallel speedup in either optimistic or conservative synchronization, or serve as a default parallel allocation scheme that can be improved by the user.
+A graph algorithm that distributes models, while avoiding cycles, could be used to offer a parallel speedup in either optimistic or conservative synchronization.
+Similarly, it could serve as a default parallel allocation scheme that can be improved by the user.

### 2016-06-29 publications/Journal/1-introduction.tex
Added reference to another paper
Added some more references to papers

@@
 But with the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform the simulation within reasonable time bounds.
-Whereas \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, its inherent parallelism is often insufficient.
+Whereas \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, its inherent parallelism is often insufficient~\cite{Himmelspach}.
 Several synchronization protocols from the discrete event simulation community~\cite{FujimotoBook} have been applied to \textsf{DEVS} simulation~\cite{globaltimewarp}.
@@
 While several parallel \textsf{DEVS} simulation kernels exist, they are often limited to a single synchronization protocol.
-The reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other in specific models.
+The reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other in specific models~\cite{Jafer}.
 The applicability of parallel simulation capabilities of current tools is therefore constrained to specific domains.

### 2016-06-29 publications/Journal/1-introduction.tex
Revision, a few modifications in 1,5,6
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 Since the ideal synchronization protocol might change throughout the simulation, we also support runtime switching between these protocols.
-This runtime switching is based on performance metrics, which are logged during simulation.
-Information is made available to a seperate component, where a choice is made about which synchronization protocol to use next.
+This runtime switching can be based on performance metrics, which are logged during simulation.
+Information is made available to a separate component, where a choice can be made about which synchronization protocol to use next. % It's manual now, can be extended to full automatic.
 Additionally, we investigate in this paper how model allocation influences the performance of our synchronization protocols.
@@
 We implemented a model that, depending on a single parameter, changes its ideal synchronization protocol.
-Using several models, we demonstrate the factors influencing the performance under a given synhronization protocol.
+Using several models, we demonstrate the factors influencing the performance under a given synchronization protocol.
 Dxex, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol.

### 2016-06-29 publications/Journal/6-allocation.tex
Finished going through the citations
Revision, a few modifications in 1,5,6
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

--- 2016-06-29 publications/Journal/4a-allocation.tex
@@
 
-The PHOLDTree model is similar in structure to models of gossiping in social networks~\cite{Gossip} and multi-granularity locking~\cite{MultiGranularityLocking}.
+The PHOLDTree model is similar in structure to models of gossiping in social networks~\cite{Gossip}.
 The lookahead of an atomic node is the minimally allowed $\epsilon$, indicating uncertainty, as is often the case in realistic models.
@@
 The number of connections between nodes at the same simulation core is also rather low.
-Using a depth-first allocation scheme, however, as shown in Figure~\ref{fig:pholdtree_visualize_parDFS},
+Using a depth-first allocation scheme, however, as shown in Figure~\ref{fig:pholdtree_visualize_parDFS}, minimizes inter-core connections while maximizing intra-core connections.
 
@@
 And since the number of cores is low, the overhead dominates any further speedup.
+Optimistic is less sensitive to the number of models per core as it does not need to poll each model for a lookahead, this explains the lower runtime penalty observed for optimistic.
 %TODO add some more explanation here maybe?
@@
 %TODO this paragraph is strange: is this really the case? 
+%The effect is due to the topology of the kernels, in BFS the risk on reverts is higher. Optimistic is less influenced but will get a lower speedup in the good case.
 Interestingly, we see that optimistic synchronization is less influenced by the allocation than conservative synchronization.

### 2016-06-29 membenchvs.sh
Add missing memory script.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@ -1,0 +1,45 @@
+#!/bin/sh
+
+
+# @author Ben Cardoen - Stijn Manhaeve -- Devs Ex Machina project.
+DIR="build/Benchmark/"
+DXEXQ="dxexmachina_devstone"
+DXEXI="dxexmachina_interconnect"
+DXEXP="dxexmachina_phold"
+ADEVSQ="adevs_devstone"
+ADEVSI="adevs_interconnect"
+ADEVSP="adevs_phold"
+ARGSQ=" -d 10 -w 150 -r "
+ARGSI=" -w 20 "
+ARGSP=" -n 4 -s 2 -r 10 -p 0.1 "
+PARARGS=" -c 4 "
+OPT=" opdevs "
+CON=" cpdevs "
+
+declare -a arr=("$DIR$DXEXQ$ARGSQ"
+				"$DIR$DXEXQ$ARGSQ$PARARGS$CON"
+				"$DIR$DXEXQ$ARGSQ$PARARGS$OPT"
+				"$DIR$ADEVSQ$ARGSQ"
+				"$DIR$ADEVSQ$ARGSQ$PARARGS$CON"
+				"$DIR$DXEXI$ARGSI"
+				"$DIR$DXEXI$ARGSI$PARARGS$CON"
+				"$DIR$DXEXI$ARGSI$PARARGS$OPT"
+				"$DIR$ADEVSI$ARGSI"
+				"$DIR$ADEVSI$ARGSI$PARARGS$CON"
+				"$DIR$DXEXP$ARGSP"
+				"$DIR$DXEXP$ARGSP$PARARGS$CON"
+				"$DIR$DXEXP$ARGSP$PARARGS$OPT"
+				"$DIR$ADEVSP$ARGSP"
+				"$DIR$ADEVSP$ARGSP$PARARGS$CON"
+                )
+
+BASETIME=500000
+for i in "${arr[@]}";
+	do
+	for j in `seq 1 5`;
+		do
+		RTIME=$BASETIME
+		echo -n "memkb of $i $RTIME== "
+		eval "/usr/bin/time -v $i -t $RTIME" 2>&1 | grep Maximum | awk '{print $6}'
+	done
+done

### 2016-06-29 publications/Journal/2-background.tex
First set of revisions and fixes.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

--- 2016-02-15 paper/2-background.tex
@@
 The user is required to define the lookahead, using knowledge about the model's behaviour.
-Defining lookahead is not always a trivial task if there is no detailed knowledge of the model.
-Even slight changes in the model can change to the lookahead, and can therefore have a significant influence on simulation performance.
+Defining lookahead is far from a trivial task if there is no detailed knowledge of the model.
+Even slight changes in the model or its allocation can change the lookahead, and can therefore have a significant influence on simulation performance.
 
@@
 Cancelling events, however, can cause further rollbacks, as the receiving node might also have to roll back its state.
-In practice, a single causality error can have significant repercussions.
+In practice, a single causality error can have significant repercussions on performance.
 
 Further changes are required for unrecoverable operations, such as I/O and memory management.
-These are only executed after the lower bound of all simulation times, called \textit{Global Virtual Time} (GVT), has progressed beyond their execution time.
+These are only executed after the lower bound of all simulation times, called \textit{Global Virtual Time} (GVT)~\cite{FujimotoBook}, has progressed beyond their execution time.
 
@@
 If rollbacks occur frequently, state saving and rollback overhead can cause simulation to grind to a halt.
-Apart from overhead in CPU time, a significant memory overhead is present: all intermediate states are stored up to a point where it can be considered \textit{irreversible}.
-
+Apart from overhead in CPU time, a significant memory overhead is present: intermediate states are stored up to a point where it can be considered \textit{irreversible}.
 Note that, while optimistic synchronization does not explicitly depends on lookahead, performance still implicitly depends on lookahead.
+Instead of depending on the theoretically defined safe lookahead, performance is related to the actually perceived lookahead.

### 2016-06-29 publications/Journal/3-features.tex

@@
 
-Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
+Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $\epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
 

### 2016-06-29 publications/Journal/memory.tex

@@
 \texttt{tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
-Due to our motivation for memory usage analysis, we will only measure peak allocation.
-Profiling is done using Valgrind's massif tool~\cite{Nethercote:2007:VFH:1273442.1250746}.
+Due to our motivation for memory usage analysis, we will only measure peak allocation in maximum resident set size as reported by the OS.
 

### 2016-06-29 publications/Journal/models.tex

@@
 
-\subsection{Benchmarks}
-We use three different benchmarks, which cover different aspects of the simulation kernel.
+\subsection{Benchmark Models}
+We use three different benchmark models, covering different aspects of the simulation kernel.
 
 First, the \textit{Queue} model, based on the \textit{HI} model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
-A single generator pushes events into the queue, which are consumed by the processors after a fixed or random delay.
-It takes two parameters: width and depth, which determine the width and depth of the hierarchy.
+A single generator pushes events into the queue, which get consumed by the processors after a fixed or random delay.
+It takes two parameters: the width and depth of the hierarchy.
 This benchmark shows how the complexity of the simulation kernel behaves for an increasing amount of atomic models, and an increasingly deep hierarchy.
-An example for width and depth 2 is shown in Figure~\ref{fig:queue_model}.
+An example for width and depth equal to 2 is shown in Figure~\ref{fig:queue_model}.
 
-Second, the \textit{PHOLD} model, presented by~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
+Second, the \textit{PHOLD} model~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
 Each atomic model is directly connected to every other atomic model.
 After a random delay, an atomic model sends out an event to a randomly selected output port.
-Output port selection happens in two phases: first it is decided whether the event should be sent to an atomic model at the same node.
-Afterwards, a uniform selection is made between the remaining ports.
+Output port selection happens in two phases: first it is decided whether the event should be sent within the same node, or outside of the node.
+Afterwards, a uniform selection is made between the remaining models.
 The model takes two parameter: the percentage of remote events, which determines the fraction of messages routed to other nodes, and the percentage of priority events.
 Priority events are events generated in a very short time after the previous event.
-This benchmark shows how the simulation kernel behaves in the presence of many local or remote events and when the average time advance is larger than the lookahead in conservative simulation.
+This benchmark shows how the simulation kernel behaves in the presence of many local or remote events, in combination with a varying percentage of high-priority event.s
 An example for four models, split over two nodes, is shown in Figure~\ref{fig:PHOLD_model}.
@@
 The model takes one parameter: the number of models.
-This benchmark investigates the complexity of event routing, and how the simulation kernel handles many simultaneous events.
+This benchmark shows the complexity of event creation, event routing, and simultaneous event handling.
 An example for four models is shown in Figure~\ref{fig:interconnect_model}.
@@
 	\label{fig:queue_model}
+\end{figure}
 
-	\vspace{\betweenmodels}
-
+\begin{figure}
+    \center
 	\includegraphics[width=\modelfraction\columnwidth]{fig/interconnect_model.pdf}
@@
 	\label{fig:interconnect_model}
+\end{figure}
 
-	\vspace{\betweenmodels}
-
+\begin{figure}
+    \center
 	\includegraphics[width=\modelfraction\columnwidth]{fig/phold_model.pdf}

### 2016-06-29 publications/Journal/parallel-performance.tex

@@
 
-We slightly modified the PHOLD benchmark, to include high-priority events.
+We slightly modified the PHOLD benchmark to include high-priority events.
 Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
@@
 Optimistic synchronization, however, simply goes forward in simulation time and rolls back when these high-priority events happen.
-This situation closely mimics the case made in the comparison between both synchronization algorithms by~\cite{FujimotoBook}.
+This situation closely mimics the model typically used for comparing conservative and optimistic synchronization~\cite{FujimotoBook}.
 
@@
 If barely any high-priority events occur, conservative synchronization is penalized due to its excessive blocking, which often turned out to be unnecessary.
-When many high-priority events occur, optimistic synchronization is penalized due to its mindless progression of simulation, which frequently needs to be rolled back.
+When many high-priority events occur, optimistic synchronization is penalized due to its unconditional progression of simulation, which frequently needs to be rolled back.
 Results show that there is no single perfect synchronization algorithm for this model: depending on configuration, either synchronization protocol might be better.

### 2016-06-29 publications/Journal/sequential-performance.tex

@@
 The results can be seen in Figure~\ref{fig:Queue_benchmark_seq}.
-Since dxex symbolically flattens the model, there is no performance hit when the depth is increased.
-The overhead of running the directconnect algorithm is one time only and negligible when the end time of the simulation is sufficiently large.
-Adevs on the other hand does suffer from the increased depth.
+Since dxex performs direct connection on the model, there is no performance hit when the depth is increased.
+Direct connection only needs to be done at initialization, so it is a one time cost that is negligible when simulation termination end time is sufficiently large.
+Adevs, on the other hand, suffers from the increased depth, even though some similar (but not identical) optimization to event passing was made~\cite{adevs_opt}.
 With every new hierarchical layer, routing an event from one atomic model to the next becomes more expensive, resulting in an increase in runtime.
@@
 \label{4-seq-Interconnect}
-In the Interconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
-As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.
+In the Interconnect model, we increase the number of atomic models, quadratically increasing the number of couplings and the number of external transitions.
+As shown in Figure~\ref{fig:Interconnect_benchmark}, adevs now outperforms dxex by a fair margin.
 Analysis showed that this is caused by the high amount of events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
 To shield the user from threading and deallocation concerns dxex provides an event superclass from which the user can derive to create a specialized event type.
-Copying and deallocation semantics and tracing are solved by the kernels at a runtime cost in simulations where event frequency is very high.
+Copying, deallocation, and tracing are done at runtime, adding significant overhead when events happen frequently.
 Profiling the benchmarks clearly shows the increasing cost of output generation and deallocation as the determining factor in the gap in performance.
-We refer the interested reader to the dxex \hyperref{https://bitbucket.org/bcardoen/devs-ex-machina}{}{repo}{repository} for the profiling call graphs for the different benchmarks.
 

### 2016-06-30 publications/Journal/one_page_description.tex
Added one page description

\documentclass{article}
\begin{document}
\title{One page description}
\date{}
\maketitle

\textsf{Parallel DEVS}~\cite{ParallelDEVS} is a popular formalism for modelling complex dynamic systems using a discrete-event abstraction.
But whereas a number of tools have been constructed by academia and industry, parallel simulation is still not supported in most of them.
Even where it is supported, performance studies are often done ad-hoc, without any comparison to other tools.

In this paper, we introduce our tool called DEVS-Ex-Machina~\cite{dxex}: a \textsf{Parallel DEVS} simulation kernel supporting parallel simulation.
In contrast to the few \textsf{Parallel DEVS} simulation kernels that do support parallel simulation, we support multiple synchronization protocol.
It is through the synchronization algorithm that parallelism in the model becomes exploitable, and the ideal synchronization protocol is therefore dependent on the model~\cite{FujimotoBook}.

Not only do we support multiple synchronization protocols, but we offer these through a modular framework, making additional protocols possible.
Since model behaviour can change during simulation, runtime switching between synchronization protocols is also supported.
We compare simulation performance with adevs~\cite{adevs}, deemed one of the most efficient \textsf{Parallel DEVS} simulation kernels at this time.
Comparisons are done for sequential simulation performance, parallel simulation performance, and memory consumption.
For our tool, we also present results for the different synchronization protocols.
Results are compared in-depth, showing the effect of some fundamentally different choices in the design of our tool and adevs.

\bibliographystyle{SageV}
\bibliography{papers}
\end{document}

### 2016-06-30 publications/Journal/1-introduction.tex
Consistent italics and sans-serif fonts

@@
 
-This paper introduces DEVS-Ex-Machina\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``dxex''): our simulation tool~\cite{dxex} which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
+This paper introduces \textit{DEVS-Ex-Machina}\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``\textit{dxex}''): our simulation tool~\cite{dxex} which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
 The selected synchronization protocol is transparent to the simulated model: users only determine which protocol to use.
@@
 
-Our tool is based on PythonPDEVS, but implemented in C++11 for increased performance, using features from the new C++14 standard when supported by the compiler.
-Unlike PythonPDEVS, dxex only supports multicore parallelism, thus no distributed simulation.
+Our tool is based on \textit{PythonPDEVS}, but implemented in C++11 for increased performance, using features from the new C++14 standard when supported by the compiler.
+Unlike \textit{PythonPDEVS}, \textit{dxex} only supports multicore parallelism, thus no distributed simulation.
 
@@
 Using several models, we demonstrate the factors influencing the performance under a given synchronization protocol.
-Dxex, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol.
-With dxex users can always opt to use the fastest protocol available, and through its modularity, users could even implement their own.
-To verify that this modularity does not hinder performance, we compare to another Parallel DEVS simulation kernel, called adevs~\cite{adevs}.
+\textit{Dxex}, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol.
+With \textit{dxex} users can always opt to use the fastest protocol available, and through its modularity, users could even implement their own.
+To verify that this modularity does not hinder performance, we compare to another \textsf{Parallel DEVS} simulation kernel, called \textit{adevs}~\cite{adevs}.
 
@@
 Section~\textsc{\nameref{sec:3-features}} elaborates on our design that enables this flexibility.
-In Section~\textsc{\nameref{sec:4-performance}}, we evaluate performance of our tool by comparing its different synchronization protocols, and compare to adevs.
+In Section~\textsc{\nameref{sec:4-performance}}, we evaluate performance of our tool by comparing its different synchronization protocols, and compare to \textit{adevs}.
 We continue by introducing runtime switching of synchronization protocols and different options for model allocation in Section~\textsc{\nameref{sec:4b-hotswap}} and Section~\textsc{\nameref{sec:4a-allocation}}, respectively.

### 2016-06-30 publications/Journal/3-features.tex
Consistent italics and sans-serif fonts

@@
-Historically, dxex is based on PythonPDEVS~\cite{PythonPDEVS}.
+Historically, \textit{dxex} is based on \textit{PythonPDEVS}~\cite{PythonPDEVS}.
 Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
-Dxex is a C++11-based \textsf{Parallel DEVS} simulator, based on the design of PythonPDEVS.
+\textit{Dxex} is a C++11-based \textsf{Parallel DEVS} simulator, based on the design of \textit{PythonPDEVS}.
 Whereas the feature set is not too comparable, the architectural design, simulation algorithms, and optimizations, are highly similar.
 
-We will not make a detailed comparison with PythonPDEVS here, but only mention some supported features.
-Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions~\cite{JDF}, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
+We will not make a detailed comparison with \textit{PythonPDEVS} here, but only mention some supported features.
+\textit{Dxex} supports, similarly to \textit{PythonPDEVS}, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions~\cite{JDF}, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
 We do not elaborate on these features in this paper.
-But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols (though only in parallel).
-This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
+But whereas \textit{PythonPDEVS} only supports optimistic synchronization, \textit{dxex} support multiple synchronization protocols (though only in parallel).
+This is in line with the design principle of \textit{PythonPDEVS}: allow users to pass performance hints to the simulation kernel.
 In our case, a user can pass the simulation kernel the synchronization protocol to use for this model, or even switch the synchronization protocol during runtime.
-Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language. Dxex will use new optimizations from the C++14 standard when possible.
+Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language. \textit{Dxex} will use new optimizations from the C++14 standard when possible.
 
-Since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
-This is due to dxex models being grafted on C++11, whereas PythonPDEVS models are grafted on Python.
+Since there is no universal \textsf{DEVS} model standard, \textit{dxex} models are incompatible with \textit{PythonPDEVS} and vice versa.
+This is due to \textit{dxex} models being grafted on C++11, whereas \textit{PythonPDEVS} models are grafted on Python.
 
@@
 Currently, most parallel simulation tools choose only a single synchronization protocol due to the inherent differences between protocols.
-We argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for dxex.
+We argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for \textit{dxex}.
 
@@
 \subsubsection{Sequential}
-Our sequential simulation algorithm is very similar to the one found in PythonPDEVS, including many optimizations.
+Our sequential simulation algorithm is very similar to the one found in \textit{PythonPDEVS}, including many optimizations.
 Minor modifications were made, though, such that it can be overloaded by different synchronization protocol implementations.
@@
 
-An overview of dxex's design is given in Figure~\ref{fig:class_diagram}.
+An overview of \textit{dxex}'s design is given in Figure~\ref{fig:class_diagram}.
 It shows that there is a simulation \texttt{Core}, which simulates the \texttt{AtomicModel}s connected to it.
@@
     \includegraphics[width=\textwidth]{fig/cores_class_diagram.eps}
-	\caption{Dxex design.}
+	\caption{\textit{Dxex} design.}
 	\label{fig:class_diagram}
@@
 Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}: memory bandwidth doesn't increase as fast as the number of cores using it.
-While this is always a problem, it is aggravated in dxex by providing automatic memory management for events and states.
+While this is always a problem, it is aggravated in \textit{dxex} by providing automatic memory management for events and states.
 A model written for sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.
@@
 
-Dxex uses Boost Pool~\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
+\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
 The latter can be faster, but at the cost of extra configuration.
@@
 Nonetheless the pool allocator also reduces the allocation overhead, though only by a relatively small fraction.
-Both techniques are required to reduce the overhead of memory allocations in dxex, and on by default.
+Both techniques are required to reduce the overhead of memory allocations in \textit{dxex}, and are turned on by default.
 
@@
 We of course still need to guarantee determinism and isolation between the calls to the RNG, as well as avoiding excessive synchronization.
-Dxex uses the Tina RNG collection (TRNG)~\cite{PhysRevE.75.066701} as an alternative random number generator with performance and multithreading in mind.
-Since the RNG is an implicit part of the state in the DEVS formalism, though often not implemented as such, we evaluated performance for both approaches: one global RNG per thread, and one RNG per atomic DEVS model.
+\textit{Dxex} uses the Tina RNG collection (TRNG)~\cite{PhysRevE.75.066701} as an alternative random number generator with performance and multithreading in mind.
+Since the RNG is an implicit part of the state in the \textsf{Parallel DEVS} formalism, though often not implemented as such, we evaluated performance for both approaches: one global RNG per thread, and one RNG per atomic \textsf{DEVS} model.
 
@@
 This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
-Dxex's sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation.
+\textit{Dxex}'s sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
 The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.
 
-Figure~\ref{fig:Queuerngspeedup} shows that dxex in sequential simulation gets three times faster by using TRNG, than when using the STL RNG.
+Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in sequential simulation gets three times faster by using TRNG, than when using the STL RNG.
 For parallel simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between sequential and parallel simulation.

### 2016-06-30 publications/Journal/5-hotswap.tex
Consistent italics and sans-serif fonts

@@
 It is well known, and repeated in the previous section, that model behaviour significantly influences the ideal synchronization protocol.
-Contrary to many modelling formalisms, the DEVS formalisms makes it possible to model basically any kind of discrete event model.
+Contrary to many modelling formalisms, the \textsf{DEVS} formalisms makes it possible to model basically any kind of discrete event model.
 As such, it is possible for the model to significantly change its behaviour throughout the simulation.
@@
 Defining the ideal synchronization protocol at the start of the simulation, when information about future model behaviour is scarce, might therefore not offer the best possible performance.
-In dxex, we not only make it possible to define the synchronization protocol to use, but also to change this decision throughout simulation.
+In \textit{dxex}, we not only make it possible to define the synchronization protocol to use, but also to change this decision throughout simulation.
 To do this, all kernels are notified of the switch and they are forced to stop simulation.
@@
 When the number of rollbacks is excessively high, switching to conservative synchronization might be considered.
-And when neither of these two is an option, the simulation might just (temporarily) have to fall back to sequential simulation.
+%And when neither of these two is an option, the simulation might just (temporarily) have to fall back to sequential simulation. % Falling back to sequential is not supported atm.
 

### 2016-06-30 publications/Journal/6-allocation.tex
Consistent italics and sans-serif fonts

@@
 Indeed, if the model is distributed such that frequent communication is necessary between cores, parallelism is naturally reduced.
-This brings us to the topic of model allocation, as also implemented in PythonPDEVS~\cite{PythonPDEVS2}.
+This brings us to the topic of model allocation, as also implemented in \textit{PythonPDEVS}~\cite{PythonPDEVS2}.
 
@@
 \subsubsection{Sequential Simulation}
-Since adevs does not use direct connection, we expect a noticable performance difference between dxex and adevs.
-This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout ($n$) determines the performance penalty adevs suffers compared to dxex.
-Profiling indeed indicates that an increase in width per subtree ($n$) leads to higher overheads in adevs due to the lack of direct connection.
-Dxex uses direct connection, making it independent on fanout.
-Performance of dxex is, in this model, only dependent on the number of models.
+Since \textit{adevs} does not use direct connection, we expect a noticable performance difference between \textit{dxex} and \textit{adevs}.
+This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout ($n$) determines the performance penalty \textit{adevs} suffers compared to \textit{dxex}.
+Profiling indeed indicates that an increase in width per subtree ($n$) leads to higher overheads in \textit{adevs} due to the lack of direct connection.
+\textit{Dxex} uses direct connection, making it independent on fanout.
+Performance of \textit{dxex} is, in this model, only dependent on the number of models.
 Slight deviations can still be seen, though, caused by the initialization overhead of direct connection.
-Both adevs and dxex scale linearly in the number of atomic models.
+Both \textit{adevs} and \textit{dxex} scale linearly in the number of atomic models.
 

### 2016-06-30 publications/Journal/7-related-work.tex
Added some more references to papers
Consistent italics and sans-serif fonts

--- 2016-06-29 publications/Journal/5-related-work.tex
@@
 
-In terms of code design and philosophy, dxex is most related to PythonPDEVS~\cite{PythonPDEVS}.
-Performance of PythonPDEVS was still decent through the use of ``hints'' from the modeler.
+In terms of code design and philosophy, \textit{dxex} is most related to \textit{PythonPDEVS}~\cite{PythonPDEVS}.
+Performance of \textit{PythonPDEVS} was still decent through the use of ``hints'' from the modeler.
 In this spirit, we offer users the possibility to choose between different synchronization protocols.
 This allows users to choose the most appropriate synchronization protocol, depending on the model.
-Contrary to PythonPDEVS, however, dxex doesn't support distributed simulation~\cite{JDF}, model migrations~\cite{PythonPDEVS2}, or activity hints~\cite{PythonPDEVS_ACTIMS}.
+Contrary to \textit{PythonPDEVS}, however, \textit{dxex} doesn't support distributed simulation~\cite{JDF}, model migrations~\cite{PythonPDEVS2}, or activity hints~\cite{PythonPDEVS_ACTIMS}.
 
-Although PythonPDEVS offers very fast turnaround cycles, simulation performance was easily outdone by compiled simulation kernels.
-In terms of performance, adevs~\cite{adevs} offered much faster simulation, at the cost of a significant compilation time.
-The turnaround cycle in adevs is much slower though, specifically because the complete simulation kernel is implemented using templates in header files.
+Although \textit{PythonPDEVS} offers very fast turnaround cycles, simulation performance was easily outdone by compiled simulation kernels.
+In terms of performance, \textit{adevs}~\cite{adevs} offered much faster simulation, at the cost of a significant compilation time.
+The turnaround cycle in \textit{adevs} is much slower though, specifically because the complete simulation kernel is implemented using templates in header files.
 As a result, the complete simulation kernel has to be compiled again every time.
-Similarly to vle~\cite{vle} and PowerDEVS~\cite{PowerDEVS}, dxex compromises by separating the simulation kernel into a shared library.
+Similarly to \textit{vle}~\cite{vle} and \textit{PowerDEVS}~\cite{PowerDEVS}, \textit{dxex} compromises by separating the simulation kernel into a shared library.
 After the initial compilation of the simulation tool, only the model has to be compiled and linked to the shared library.
 This significantly shortens the turnaround cycle, while still offering good performance.
-In terms of performance, dxex is shown to be competitive with adevs.
-Despite its high performance, adevs does not support optimistic synchronization, which we have shown to be highly relevant.
+In terms of performance, \textit{dxex} is shown to be competitive with \textit{adevs}.
+Despite its high performance, \textit{adevs} does not support optimistic synchronization, which we have shown to be highly relevant.
 
 Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though none have done it in a strictly modular way that allows straightforward protocol switching for a single given model.
-For example, adevs only supports conservative synchronization, and vle only supports experiment-level parallelism (\textit{i.e.}, run multiple experiments concurrently).
-Closest to our support for multiple synchronization protocols is CD++~\cite{CD++}.
-For CD++, both a conservative (CCD++~\cite{CCD++}) and optimistic (PCD++)~\cite{PCD++}) variant exist.
+For example, \textit{adevs} only supports conservative synchronization, and \textit{vle} only supports experiment-level parallelism (\textit{i.e.}, run multiple experiments concurrently).
+Closest to our support for multiple synchronization protocols is \textit{CD++}~\cite{CD++}.
+For \textit{CD++}, both a conservative (\textit{CCD++}~\cite{CCD++}) and optimistic (\textit{PCD++})~\cite{PCD++}) variant exist.
 Despite the implementation of both protocols, they are different projects entirely.
-Some features might therefore be implemented in CCD++ and not in PCD++, or vice versa.
+Some features might therefore be implemented in \textit{CCD++} and not in \textit{PCD++}, or vice versa.
 And while this might not yet be that significant a problem to this day, this problem will only get worse when each project follows its own course.
-Dxex, on the other hand, is a single project, where the choice of synchronization protocol is a simple configuration.
-CD++, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourselves to parallel simulation.
+\textit{Dxex}, on the other hand, is a single project, where the choice of synchronization protocol is a simple configuration.
+\textit{CD++}, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourselves to parallel simulation.
 By limiting our approach to parallel simulation, we are able to achieve higher speedups through the use of shared memory communication.
 
-% Paragraph runtime switching
-In the PDES community, the problem of choosing between synchronization protocols is well known and documented~\cite{Jha:1994:UFC:195291.182480}.
+In the \textsf{PDES} community, the problem of choosing between synchronization protocols is well known and documented~\cite{Jha:1994:UFC:195291.182480}.
 The challenges of implementing such runtime switching have previously been explored already~\cite{Das:1996:APP:256562.256602}, and an implementation was given by, for example~\cite{perumalla2005musik}.
-Our contribution entails bringing this same feature to the DEVS community, further expanding upon our unique feature of multiple synchronization protocols.
+Our contribution entails bringing this same feature to the \textsf{Parallel DEVS} community, further expanding upon our support for multiple synchronization protocols.
 
-% Paragraph allocation effects
-Model allocation and its impact on parallel performance has previously also been studied in the PDES community~\cite{PDESpartitioning}.
+Model allocation and its impact on parallel performance has previously also been studied in the \textsf{PDES} community~\cite{PDESpartitioning}.
 Referenced as partitioning of the simulation model, most studies distinguish between communication overhead and computational distribution (load balancing) as the two dimensions to partition over.
 Partitioning a simulation model is identified as an issue to achieve scalability~\cite{Scalability}. 
-Some research in the context of Parallel DEVS has been done, where they turn their attention to load balancing and communication overhead~\cite{PDEVSpartitioning}.
+Some research in the context of \textsf{Parallel DEVS} has been done, where they turn their attention to load balancing and communication overhead~\cite{PDEVSpartitioning}.
 Our contribution studies the effect of partitioning with emphasis on the effect of communication between processes and in the presence of a flattened hierarchy. 
 We focus on static partitioning since this is a limiting factor for our conservative synchronization implementation which does not support model relocation.
-Model relocation, as implemented by PythonPDEVS~\cite{PythonPDEVS2}, might be an interesting addition to only model allocation at the start of simulation.
+Model relocation, as implemented by \textit{PythonPDEVS}~\cite{PythonPDEVS2}, might be an interesting addition to only model allocation at the start of simulation.
 
-In summary, dxex tries to find the middle ground between the concepts of PythonPDEVS, the performance of adevs, and the multiple synchronization protocols of CD++.
+In summary, \textit{dxex} tries to find the middle ground between the concepts of \textit{PythonPDEVS}, the performance of \textit{adevs}, and the multiple synchronization protocols of \textit{CD++}.
 To further profit from our multiple synchronization protocols in a single tool, we further added runtime switching between synchronization protocols and model allocation support.

### 2016-06-30 publications/Journal/8-conclusions.tex
Consistent italics and sans-serif fonts

@@
-In this paper, we introduced DEVS-Ex-Machina (``dxex''), a new C++11-based \textsf{Parallel DEVS} simulation tool.
+In this paper, we introduced \textit{DEVS-Ex-Machina} (``\textit{dxex}''), a new C++11-based \textsf{Parallel DEVS} simulation tool.
 Our main contribution is the implementation of multiple synchronization protocols for parallel multicore simulation.
 We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
-Dxex allows the user to choose between either conservative or optimistic synchronization at the start of simulation.
+\textit{Dxex} allows the user to choose between either conservative or optimistic synchronization at the start of simulation.
 Depending on observed model behaviour and simulation performance, runtime switching between synchronization protocols can be used.
 
-Notwithstanding our modularity, dxex achieves performance competitive to adevs, another very efficient \textsf{DEVS} simulation tool.
+Notwithstanding our modularity, \textit{dxex} achieves performance competitive to \textit{adevs}, another very efficient \textsf{DEVS} simulation tool.
 Performance is measured both in elapsed time, and memory usage.
@@
 Third, automatic allocation might be possible by analysis of the connections between models.
-This information is already used in dxex to construct the dependency graph in conservative synchronization.
+This information is already used in \textit{dxex} to construct the dependency graph in conservative synchronization.
 A graph algorithm that distributes models, while avoiding cycles, could be used to offer a parallel speedup in either optimistic or conservative synchronization.

### 2016-06-30 publications/Journal/memory.tex
Removed note about adevs having no results for PHOLD
Consistent italics and sans-serif fonts

@@
 While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation unfeasible.
-We therefore also investigate memory usage of different synchronization protocols, and again compare to adevs.
+We therefore also investigate memory usage of different synchronization protocols, and again compare to \textit{adevs}.
 
 We do not tackle the problem of states that become too large for a single machine to hold.
-This problem can be mitigated by distribution over multiple machines, which neither dxex or adevs support.
+This problem can be mitigated by distribution over multiple machines, which neither \textit{dxex} or \textit{adevs} support.
 
 \subsubsection{Remarks}
-Both dxex and adevs use \texttt{tcmalloc} as memory allocator, allowing for thread-local allocation.
+Both \textit{dxex} and \textit{adevs} use \textit{tcmalloc} as memory allocator, allowing for thread-local allocation.
 Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
-\texttt{tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
+\textit{Tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
 Due to our motivation for memory usage analysis, we will only measure peak allocation in maximum resident set size as reported by the OS.
@@
 Results are in megabytes, and show the total memory footprint of the running application (\textit{i.e.}, text, stack, and heap).
+Note the logarithmic scale due to the high memory consumption of optimistic synchronization.
 
@@
 Optimistic synchronization results vary heavily depending on thread scheduling by the operating system, as this influences the drift between nodes.
-Comparing similar approaches, we notice that dxex and adevs have very similar memory use.
+Comparing similar approaches, we notice that \textit{dxex} and \textit{adevs} have very similar memory use.
 
@@
 Additional memory is required for the multiple threads, but also to store all events that are processed simultaneously.
-
-For the Phold benchmark, adevs using conservative synchronization took too long using our profiling tool, and was therefore aborted.
-Therefore, no results are shown for adevs.
 

### 2016-06-30 publications/Journal/models.tex
Consistent italics and sans-serif fonts

@@
 
-In this section, we evaluate the performance of different synchronization protocols in dxex.
-We also compare to adevs~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVSSurvey,DEVStoneJournal}, to show that our modularity does not impede performance.
+In this section, we evaluate the performance of different synchronization protocols in \textit{dxex}.
+We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVSSurvey,DEVStoneJournal}, to show that our modularity does not impede performance.
 CPU time and memory usage is compared for both sequential and parallel simulation.
 
-We start off with a comparison of sequential simulation, to show how adevs and dxex relate in this simple case.
+We start off with a comparison of sequential simulation, to show how \textit{adevs} and \textit{dxex} relate in this simple case.
 For the parallel simulation benchmarks, results are presented for both conservative and optimistic synchronization.
@@
 For all benchmarks, results are well within a 5\% deviation of the average, such that only the average is used in the remainder of this section.
-The same compilation flags were used for both adevs and dxex benchmarks (``\texttt{-O3 -flto}'').
+The same compilation flags were used for both \textit{adevs} and \textit{dxex} benchmarks (``\texttt{-O3 -flto}'').
 To guarantee comparable results, no I/O was performed during benchmarks.
-Before benchmarking, simulation traces were used to verify that adevs and dxex return exactly the same simulation results.
+Before benchmarking, simulation traces were used to verify that \textit{adevs} and \textit{dxex} return exactly the same simulation results.
 Benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.

### 2016-06-30 publications/Journal/parallel-performance.tex
Consistent italics and sans-serif fonts

@@
 \subsection{Parallel Simulation}
-We now continue by describing our parallel simulation performance for different synchronization protocols, and compared to adevs.
-The speedup of adevs is computed with the corresponding dxex sequential benchmark.
+We now continue by describing our parallel simulation performance for different synchronization protocols, and compared to \textit{adevs}.
+The speedup of \textit{adevs} is computed with the corresponding \textit{dxex} sequential benchmark.
 This was done to take into account the performance difference observed in sequential simulation.
@@
 This is mainly caused by the pipeline structure of the model: the last models in the queue only respond to incoming messages and therefore have to be rolled back frequently.
-The difference between dxex conservative and adevs becoms smaller when more and more cores are used.
+The difference between \textit{dxex} conservative and \textit{adevs} becoms smaller when more and more cores are used.
 The same effect can be seen for weak scaling in Figure~\ref{fig:Queue_plot_weak}.
@@
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_weak_speedup.eps}
-	\caption{Queue model weak scaling speedup compared to dxex sequential.}
+	\caption{Queue model weak scaling speedup compared to \textit{dxex} sequential.}
 	\label{fig:Queue_plot_weak}
@@
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_strong_speedup.eps}
-	\caption{Queue model strong scaling speedup compared to dxex sequential.}
+	\caption{Queue model strong scaling speedup compared to \textit{dxex} sequential.}
 	\label{fig:Queue_plot_strong}
@@
 
-Adevs is significantly slower during conservative synchronization.
-Analysis of profiling callgraphs shows that exception handling in adevs is the main cause. 
-To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.
+\textit{Adevs} is significantly slower during conservative synchronization.
+Analysis of profiling callgraphs shows that exception handling in \textit{adevs} is the main cause. 
+To keep the models equivalent, the \textit{adevs} version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.
 These functions require the user to implement a state saving method.
-But in contrast to PythonPDEVS and dxex, which handle this inside the kernel, users need to manually define this.
+But in contrast to \textit{PythonPDEVS} and \textit{dxex}, which handle this inside the kernel, users need to manually define this.
 We feel this would lead to an unfair comparison as we would like to keep the models agnostic of the underlying protocols across all benchmarks.

### 2016-06-30 publications/Journal/sequential-performance.tex
Consistent italics and sans-serif fonts

@@
 The results can be seen in Figure~\ref{fig:Queue_benchmark_seq}.
-Since dxex performs direct connection on the model, there is no performance hit when the depth is increased.
+Since \textit{dxex} performs direct connection on the model, there is no performance hit when the depth is increased.
 Direct connection only needs to be done at initialization, so it is a one time cost that is negligible when simulation termination end time is sufficiently large.
-Adevs, on the other hand, suffers from the increased depth, even though some similar (but not identical) optimization to event passing was made~\cite{adevs_opt}.
+\textit{Adevs}, on the other hand, suffers from the increased depth, even though some similar (but not identical) optimization to event passing was made~\cite{adevs_opt}.
 With every new hierarchical layer, routing an event from one atomic model to the next becomes more expensive, resulting in an increase in runtime.
@@
 In the Interconnect model, we increase the number of atomic models, quadratically increasing the number of couplings and the number of external transitions.
-As shown in Figure~\ref{fig:Interconnect_benchmark}, adevs now outperforms dxex by a fair margin.
-Analysis showed that this is caused by the high amount of events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
-To shield the user from threading and deallocation concerns dxex provides an event superclass from which the user can derive to create a specialized event type.
+As shown in Figure~\ref{fig:Interconnect_benchmark}, \textit{adevs} now outperforms \textit{dxex} by a fair margin.
+Analysis showed that this is caused by the high amount of events: event creation is much slower in \textit{dxex} than it is in \textit{adevs}, despite \textit{dxex}'s use of memory pools.
+To shield the user from threading and deallocation concerns \textit{dxex} provides an event superclass from which the user can derive to create a specialized event type.
 Copying, deallocation, and tracing are done at runtime, adding significant overhead when events happen frequently.
@@
 The number of events scales linear with the number of models, not quadratic.
-Figure~\ref{fig:Phold_benchmark} shows that the performance of dxex and adevs are very close to each-other, with adevs slightly outperforming dxex.
+Figure~\ref{fig:Phold_benchmark} shows that the performance of \textit{dxex} and \textit{adevs} are very close to each-other, with \textit{adevs} slightly outperforming \textit{dxex}.
 \begin{figure}

### 2016-09-23 publications/Journal/models.tex
* removed reference 22

@@
 In this section, we evaluate the performance of different synchronization protocols in \textit{dxex}.
-We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVSSurvey,DEVStoneJournal}, to show that our modularity does not impede performance.
+We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVStoneJournal}, to show that our modularity does not impede performance.
 CPU time and memory usage is compared for both sequential and parallel simulation.

### 2016-09-23 publications/Journal/3-features.tex
Address first referee's comments (p1)
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 Always defining a lookahead function might seem redundant, especially should the user never want to use conservative synchronization.
-Definig the lookahead is difficult and can often be unpredictable.
+Defining the lookahead is difficult and can often be unpredictable.
 The more attractive option is for the simulation tool to provide a default lookahead function, such that simulation can run anyway, but likely not at peak performance.
@@
 Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $\epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
+Providing this default implementation has no impact in sequential or optimistic simulation, since the function is never called the compiler will optimize it out.
+By providing this default implementation in the model base class we ensure that a model can run in sequential, optimistic and conservative simulation by default.
 

### 2016-09-23 publications/Journal/3-features.tex
Referee 1's comments p2
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 
-Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in sequential simulation gets three times faster by using TRNG, than when using the STL RNG.
+Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in sequential simulation is three times faster using TRNG compared to using the STL RNG.
 For parallel simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between sequential and parallel simulation.
 Conservative synchronization is almost insensitive to the changing of the RNG, though a slight increase in performance can be noted.
-Optimistic synchronization gets much slower when the RNG becomes part of the model state, since the state needs to be copied as well.
+Optimistic synchronization slows down significantly when the RNG becomes part of the model state, since the state needs to be copied as well.
 This becomes a significant overhead when using the STL RNG, since performance plummets to a fraction of the original.

### 2016-09-23 publications/Journal/8-conclusions.tex
Address first referee's comments (p1)
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
Referee 1's comments p2
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 Furthermore we have shown when and why optimistic synchronization can outperform conservative and vice versa.
+Finally we investigated the effect of memory (de)allocation on parallel simulation.
 
@@
 In its current state, simulation will simply halt with an out-of-memory error.
+The use of transactional memory can offer several advantages in this project.
+If it becomes part of the new C++17 standard it would be of great interest to see if it can help reduce the performance effects of memory allocation and synchronization.
 Having simulation control, which can throttle the speed of nodes that use up too much memory, has been shown to work in these situations~\cite{FujimotoBook}.

### 2016-09-27 publications/Journal/7-related-work.tex
Update Related Work with reviewer comments.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 \textit{CD++}, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourselves to parallel simulation. 
-By limiting our approach to parallel simulation, we are able to achieve higher speedups through the use of shared memory communication.
+A new architecture for sequential PDEVS simulation has been introduced in \cite{SeqPDEVSArch} with promising performance results.
+This algorithm achieves a speedup by using the inherent parallelism in PDEVS without the need for synchronization.
+By limiting our approach to a pure parallel simulation, we are able to achieve higher speedups through the use of shared memory communication.
 
@@
 Partitioning a simulation model is identified as an issue to achieve scalability~\cite{Scalability}. 
-Some research in the context of \textsf{Parallel DEVS} has been done, where they turn their attention to load balancing and communication overhead~\cite{PDEVSpartitioning}.
+Some research in the context of \textsf{Parallel DEVS} has been done, where they turn their attention to load balancing and communication overhead~\cite{PDEVSpartitioning, NonFragmenting}.
 Our contribution studies the effect of partitioning with emphasis on the effect of communication between processes and in the presence of a flattened hierarchy. 

### 2016-09-30 publications/Journal/2-background.tex
Consistent italics and sans-serif fonts
First 3 sections review for style and acronyms.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
-This section briefly introduces the synchronization protocols used by dxex: conservative and optimistic synchronization.
+This section briefly introduces the synchronization protocols used by \textit{dxex}: conservative and optimistic synchronization.
 
@@
 Outgoing events are cancelled after a rollback, through the use of anti-messages, as potentially different output events have to be generated.
-Cancelling events, however, can cause further rollbacks, as the receiving node might also have to roll back its state.
-In practice, a single causality error can have significant repercussions on performance.
+Cancelling events can cause further rollbacks as the receiving node might also have to roll back its state.
+In practice a single causality error can have significant repercussions on performance.
 
@@
 
-The main advantage is that performance is not limited by a small lookahead, caused by a very infrequent event.
+The main advantage is that a small lookahead, caused by a very infrequent event, will not limit performance.
+%The main advantage is that performance is not limited by a small lookahead caused by a very infrequent event.
 If an (almost) instantaneous event rarely occurs, performance is only impacted when it occurs, and not at every simulation step.

### 2016-09-30 publications/Journal/3-features.tex
First 3 sections review for style and acronyms.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
-This made us switch to the use of thread-local (using \textit{tcmalloc}) memory pools.
+This made us switch to the use of thread-local %(using \textit{tcmalloc})% see page 6 for introducing tcmalloc
+ memory pools.
 Again, we made use of specific new features of C++11, that are not available in Python, or even previous versions of the C++ language standard.
@@
 
-Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $\epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
+Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default value $\epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
 Providing this default implementation has no impact in sequential or optimistic simulation, since the function is never called the compiler will optimize it out.
@@
 
-Furthermore, allocating and deallocating memory by making calls to the operating system, as typically done by calls such as \texttt{malloc}, happens sequentially.
+Furthermore, allocating and deallocating memory by making calls to the operating system, as is typically done by calls such as \texttt{malloc}, happens sequentially.
 To counter this, our memory allocators are backed by a thread-aware pooling library.
@@
 
-%TODO what does it mean to "pool aggressively?"
-Intra-kernel events are pooled aggressively, whereas inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
-A simulation with many inter kernel events suffers a performance hit, whereas the impact of many intra-kernel events can be minimized using arena allocators.
+Intra-kernel events can be (de)allocated without requiring synchronization with the other kernels. 
+They can be returned immediately to the memory allocator as the lifetime of these objects is known at creation.
+In contrast inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
+As a result simulations with many inter-kernel events suffer a performance hit, whereas the impact of many intra-kernel events can be minimized using arena ~\cite{Arena} allocators.
 
@@
 \subsubsection{Random Number Generators}
-Random Number Generators (RNG) are another aspect of the program that results in sequentialization.
+Random Number Generators (RNG) are another aspect of the program that can limit parallelization.
 All accesses to the RNG will result in the modification of a global (\textit{i.e.}, shared between threads) variable.
@@
 
-We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default STL random number generator.
+We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default Standard Template Library (STL) random number generator.
 This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.

### 2016-09-30 publications/Journal/3-features.tex
Further review of style, acronyms.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 It shows that there is a simulation \texttt{Core}, which simulates the \texttt{AtomicModel}s connected to it.
-The superclass \texttt{Core} represents the sequential simulation core.
+The superclass \texttt{Core} represents the sequential simulation kernel.
 Subclasses define specific variants, such as \texttt{ConservativeCore} (conservative synchronization), \texttt{OptimisticCore} (optimistic synchronization), and \texttt{DynamicCore} (\textsf{Dynamic Structure DEVS}).
@@
 This easily becomes a bottleneck in simulation, since random numbers are a common occurence in simulation~\cite{Random}.
-As such, a non-trivial amount of time in a simulation is often spent waiting for an RNG.
+As such, a nontrivial amount of time in a simulation is often spent waiting for an RNG.
 

### 2016-09-30 publications/Journal/5-hotswap.tex
Further review of style, acronyms.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 We add performance metrics in the simulation kernel, which logs relevant performance metrics and processes them for use in other components.
-These metrics include the number of events created and destroyed, the number of inter- and intra-kernel events, the number of rollbacks, the measured lookahead, details of the Global Virtual Time (GVT) and Earliest Output Time (EOT) calculations, and information on the fairness between simulation kernels.
+These metrics include the number of events created and destroyed, the number of inter- and intra-kernel events, the number of rollbacks, the measured lookahead, details of the GVT and EOT calculations, and information on the fairness between simulation kernels.
 With all these metrics, a component can get a fair view on both model and simulation kernel behaviour.

### 2016-09-30 publications/Journal/conclusion.tex
Further review of style, acronyms.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 Similarly to the polymorphic scheduler~\cite{MasterThesis}, we wish to make it possible for the ideal option to be switched during simulation.
-When changes to the model behaviour are noticed, the used synchronization protocol is modified as well.
+When changes to the model behaviour are noticed, the used synchronization protocol can be switched as well.
 
-Second, the allocation of models is tricky and has a significant impact on performance.
+Second, the allocation of models is nontrivial and has a significant impact on performance.
 While our parallel speedup for the Queue model, for example, was rather high, this is mostly due to characteristics of the model: the dependency graph does not contain any cycles.

### 2016-09-30 publications/Journal/conclusion.tex
Initial section of pdevs.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
+PDEVS is compared with the synchronziation protocols either in isolation or complementing each other. We have shown that % continue results
 But even with the right synchronization protocol, we have seen that two problems remain.

### 2016-09-30 publications/Journal/pdevs.tex
First draft of pdevs + figs.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

\subsection{Comparison with PDEVS}
In this section we contrast the speedup between the original PDEVS algorithm and the synchronization protocols. 
Note that the PDEVS algorithm and the synchronization protocols are complementary in \textit{dxex}, not exclusive. 
Following the theoretical analysis published in ~\cite{amdahlpdevs} a comparison between PDEVS and the synchronization protocols was warranted. 
We compare the ideal scenario for PDEVS where all events are concurrent with a scenario where events are random.
In this work we do not consider the distribution of events, introducing a random number generator as we will show in this work can mask performance results. 
By limiting ourselves to the best and worst case scenario we demonstrate the capabilities and limitations of both approaches.
The influence of the computational load of the transition functions is a second factor we investigate in this section.
We are interested in scenarios where PDEVS and the synchronization protocols can complement each other in addition to observing their speedup in isolation.
The benchmarks are run on the same hardware as our previous results, a 8 x AMD Opteron(TM) Processor 6274 machine with 8 cores per cpu (64 cores) and 192 GB RAM. 
The benchmarks use a maximum of 16 threads to be divided between PDEVS , conservative or optimistic.

\paragraph{Model}
As shown in the previous section, when most models are dependent on each other across kernels there is no speedup obtainable by either optimistic or conservative. 
We therefore opted to use the \textit{Queue} model in this benchmark, with depth 4 and width 300.
The effect of combining 4 synchronized kernels with each 4 threads for PDEVS is contrasted with 16 threads for PDEVS and 16 kernels for conservative and optimistic.  
This allows us to observe which is more efficient in obtaining a speedup.
We simulate a computational load by a call to sleep in the transition functions set at 5ms.

\paragraph{Concurrent events}
First we create the ideal scenario for PDEVS, all events are concurrent with a significant computational load in the transition functions.

In figure \ref{fig:pdevs_plot_fixed_sleep} we observe that PDEVS obtains a speedup approaching 8. 
Conservative and optimistic are run first with 16 kernels, then with 4 kernels having 4  PDEVS threads each. 
The resulting speedup is half of that obtained of the sequential kernel with 16 PDEVS threads.

\paragraph{Random events}
Random events allow us to demonstrate the overhead of PDEVS. The probability that events are concurrent is strongly reduced.
The transition function has the same computational load as in the previous configuration. 
The results in figure \ref{fig:pdevs_plot_random_sleep} demonstrate that the overhead in the sequential DPEVS kernel is negligible. In contrast both conservative and optimistic suffer a significant slowdown when combined with PDEVS.
We clearly see that conservative and optimistic outperform the sequential kernels in this scenario, complementing the synchronized kernels with PDEVS is not warranted here.

\paragraph{Computational load}
When we remove the computational load we isolate the overhead of the parallelism. 
In this paper we are mainly interested in this overhead which is otherwise masked by adding a computational cost to the transition functions. 
All the benchmarks in this work are run with a near negligible computational load in the transition functions for this purpose.
If a parallel speedup can be obtained in a simulation model without any load in the transition function this will translate in a higher speedup with an increased load, as this load can then be spread over kernels or threads.
Although this configuration has only concurrent events the overhead of the synchronization is severe. 
The runtime cost of this overhead is extreme in the sequential kernel. 
The synchronized kernels are impacted less by the overhead of the PDEVS algorithm.

\paragraph{Discussion}
In \textit{dxex} PDEVS and the two synchronization protocols are complementary. The user is not forced to choose between either but can divide the capabilities of his machine over both approaches.
While PDEVS can offer a significant speedup we see that to offer this speedup the concurrency of events must be significant in combination with a high computational load in the transition functions. 
We measure the overhead introduced by PDEVS to give practitioners insight into which model can benefit from which paradigm.
The overhead of the synchronization protocols in specific scenarios  is shown in the previous section.
For a given model it is difficult to predict the concurrency of events, although the computational load of a transition function can be approximated.
In these benchmarks we show which configurations can lead to a speedup for PDEVS or the synchronization protocols in isolation or combination.

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/pdevs_fixed_sleep.eps}
	\caption{Queue speedup benchmark for PDEVS and synchronization with maximal concurrent events and significant computational load}
	\label{fig:pdevs_plot_fixed_sleep}
\end{figure}

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/pdevs_random_sleep.eps}
	\caption{Queue speedup benchmark for PDEVS and synchronization with randomized events and significant computational load}
	\label{fig:pdevs_plot_random_sleep}
\end{figure}

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/pdevs_no_sleep.eps}
	\caption{Queue speedup benchmark for PDEVS and synchronization with maximal concurrent events and trivial computational load}
	\label{fig:pdevs_plot_no_sleep}
\end{figure}

### 2016-09-30 pdevsbench.sh
Modify benchmark parameters to fix idle chain problem.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@ -75,7 +75,7 @@ if [ ! -f "$EXEC_16THREADSLEEP" ]; then
 fi
 
 ARGS_NOSLEEP=" -d 4 -w 400 -t 5000000 "
-ARGS_SLEEP=" -d 4 -w 400 -t 5000 "
+ARGS_SLEEP=" -d 4 -w 50 -t 10000 "
 REPEAT=5
 mkdir -p $OUT_DIR
 echo "Executing benchmarks"

### 2016-09-30 pdevsbench.sh
Disable random/nosleep
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@ -92,15 +92,15 @@ perf stat -r $REPEAT -o "$OUT_DIR/pdevs_sleep_seq.txt" $EXEC_16THREADSLEEP $ARGS
 perf stat -r $REPEAT -o "$OUT_DIR/pdevs_sleep_con.txt" $EXEC_4THREADSLEEP $ARGS_SLEEP cpdevs -c 4
 perf stat -r $REPEAT -o "$OUT_DIR/pdevs_sleep_opt.txt" $EXEC_4THREADSLEEP $ARGS_SLEEP opdevs -c 4
 
-perf stat -r $REPEAT -o "$OUT_DIR/normal_seq_r.txt" $EXEC_NORM $ARGS_NOSLEEP -r
-perf stat -r $REPEAT -o "$OUT_DIR/normal_con_r.txt" $EXEC_NORM $ARGS_NOSLEEP cpdevs -c 16 -r
-perf stat -r $REPEAT -o "$OUT_DIR/normal_opt_r.txt" $EXEC_NORM $ARGS_NOSLEEP opdevs -c 16 -r
+#perf stat -r $REPEAT -o "$OUT_DIR/normal_seq_r.txt" $EXEC_NORM $ARGS_NOSLEEP -r
+#perf stat -r $REPEAT -o "$OUT_DIR/normal_con_r.txt" $EXEC_NORM $ARGS_NOSLEEP cpdevs -c 16 -r
+#perf stat -r $REPEAT -o "$OUT_DIR/normal_opt_r.txt" $EXEC_NORM $ARGS_NOSLEEP opdevs -c 16 -r
 perf stat -r $REPEAT -o "$OUT_DIR/npdevs_sleep_seq_r.txt" $EXEC_1THREADSLEEP $ARGS_SLEEP -r
 perf stat -r $REPEAT -o "$OUT_DIR/npdevs_sleep_con_r.txt" $EXEC_1THREADSLEEP $ARGS_SLEEP cpdevs -c 16 -r
 perf stat -r $REPEAT -o "$OUT_DIR/npdevs_sleep_opt_r.txt" $EXEC_1THREADSLEEP $ARGS_SLEEP opdevs -c 16 -r
-perf stat -r $REPEAT -o "$OUT_DIR/pdevs_seq_r.txt" $EXEC_16THREAD $ARGS_NOSLEEP -r
[…]

### 2016-09-30 publications/Journal/pdevs.tex
Staging commit.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 As shown in the previous section, when most models are dependent on each other across kernels there is no speedup obtainable by either optimistic or conservative. 
-We therefore opted to use the \textit{Queue} model in this benchmark, with depth 4 and width 300.
+We therefore opted to use the \textit{Queue} model in this benchmark.
 The effect of combining 4 synchronized kernels with each 4 threads for PDEVS is contrasted with 16 threads for PDEVS and 16 kernels for conservative and optimistic.  
 This allows us to observe which is more efficient in obtaining a speedup.
-We simulate a computational load by a call to sleep in the transition functions set at 5ms.
+We simulate a computational load by an optional call to sleep in the transition functions set at 5ms. 
+The model is configured with depth 4 and width 300 if the transition function has no load, and width 50 if the computational load is active.
+In our configuration an imminent model will generate output and lead to another model becoming imminent. With respect to the 
+theoretical analysis in \cite{amdahlpdevs} this gives a parameter value q = 1. 
+The models are equally distributed over the kernels and between threads in PDEVS. 
+The coefficient of variation of our results is less than 1\%. 
+Communication between the kernels and threads, defined as communication overhead in the theoretical work is hard to to estimate here but given our coefficient of variation results we can assume this to be constant.
 
@@
 First we create the ideal scenario for PDEVS, all events are concurrent with a significant computational load in the transition functions.
-
+All models are imminent at the same time in this configuration allowing for maximum parallelism.
 In figure \ref{fig:pdevs_plot_fixed_sleep} we observe that PDEVS obtains a speedup approaching 8. 
@@
 The transition function has the same computational load as in the previous configuration.
-The results in figure \ref{fig:pdevs_plot_random_sleep} demonstrate that the overhead in the sequential DPEVS kernel is negligible.
-In contrast both conservative and optimistic suffer a significant slowdown when combined with PDEVS.
-We clearly see that conservative and optimistic outperform the sequential kernels in this scenario, complementing the synchronized kernels with PDEVS is not warranted here.
 
@@
 In this paper we are mainly interested in this overhead which is otherwise masked by adding a computational cost to the transition functions.
-All the benchmarks in this work are run with a near negligible computational load in the transition functions for this purpose.
-If a parallel speedup can be obtained in a simulation model without any load in the transition function this will translate in a higher speedup with an increased load, as this load can then be spread over kernels or threads.
-Although this configuration has only concurrent events the overhead of the synchronization is severe. 
-The runtime cost of this overhead is extreme in the sequential kernel. 
-The synchronized kernels are impacted less by the overhead of the PDEVS algorithm.
 
@@
 In these benchmarks we show which configurations can lead to a speedup for PDEVS or the synchronization protocols in isolation or combination.
+In \citep{amdahlpdevs} a theoretical bound is defined for the maximum speedup obtained by PDEVS. 
+Our tool flattens the hierarchy of the coupled models
 

### 2016-09-30 publications/Journal/pdevs.tex
Draft version of paper + pdevs results corresponding with reviewer 3 comments.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 All models are imminent at the same time in this configuration allowing for maximum parallelism.
-In figure \ref{fig:pdevs_plot_fixed_sleep} we observe that PDEVS obtains a speedup approaching 8. 
+In figure \ref{fig:pdevs_plot_fixed_sleep} we observe that PDEVS obtains a speedup approaching 10.  
 Conservative and optimistic are run first with 16 kernels, then with 4 kernels having 4  PDEVS threads each. 
-The resulting speedup is half of that obtained of the sequential kernel with 16 PDEVS threads.
+Both protocols achieve a similar speedup compared to PDEVS. Combining the protocols with PDEVS gives a near identical speedup.
+In this scenario there is no real advantage between the different parallel configurations.
 
@@
 The transition function has the same computational load as in the previous configuration. 
+The results in figure \ref{fig:pdevs_plot_random_sleep} show that PDEVS adds little overhead in sequential simulation. 
+In combination with either synchronization protocol DPEVS can achieve a reasonable speedup, but both synchronization protocols achieve a far higher speedup without PDEVS.
 
@@
 In this paper we are mainly interested in this overhead which is otherwise masked by adding a computational cost to the transition functions. 
+Figure \ref{fig:pdevs_plot_no_sleep} shows that in sequential simulation the PDEVS overhead is crippling.
+Conservative combined with PDEVS reduces performance to that of sequential without PDEVS with worse results for optimistic in combination in PDEVS.
+Both synchronization protocols achieve a reasonable speedup without PDEVS.
 
 \paragraph{Discussion}
-In \textit{dxex} PDEVS and the two synchronization protocols are complementary. The user is not forced to choose between either but can divide the capabilities of his machine over both approaches.
-While PDEVS can offer a significant speedup we see that to offer this speedup the concurrency of events must be significant in combination with a high computational load in the transition functions. 
+In \textit{dxex} PDEVS and the two synchronization protocols can be complementary. 
+The user is not forced to choose between either but can divide the capabilities of his machine over both approaches.
+While PDEVS can offer a significant speedup, we see that to offer this speedup the concurrency of events must be significant in combination with a high computational load in the transition functions. 
 We measure the overhead introduced by PDEVS to give practitioners insight into which model can benefit from which paradigm.
@@
 For a given model it is difficult to predict the concurrency of events, although the computational load of a transition function can be approximated.
-In these benchmarks we show which configurations can lead to a speedup for PDEVS or the synchronization protocols in isolation or combination.
-In \citep{amdahlpdevs} a theoretical bound is defined for the maximum speedup obtained by PDEVS. 
-Our tool flattens the hierarchy of the coupled models
+In these benchmarks we show which configurations can lead to a speedup for PDEVS or the synchronization protocols in either isolation or combination.
+In a model with high concurrency in simultaneous events and a high computational load PDEVS can offer a significant speedup.
+This speedup is not exclusive to PDEVS. Conservative and optimistic synchronization protocols can achieve the same speedup given a non cyclic model. 
+From the theoretical model we know that in a cyclic model PDEVS can achieve a higher speedup.
+% Discuss theoretical bound
 

### 2016-09-30 publications/Journal/6-allocation.tex
Address first referee's comments (p1)
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 Although the synchronization protocol is one of the defining factors in simulation performance, model allocation has a significant impact on which protocol is ideal.
-Depending on the model structure, and how it is mapped to the different cores, it might not even make sense to parallelize at all.
-Indeed, if the model is distributed such that frequent communication is necessary between cores, parallelism is naturally reduced.
+Depending on the model structure, and how it is mapped to the different kernels, it might not even make sense to parallelize at all.
+Indeed, if the model is distributed such that frequent communication is necessary between kernels, parallelism is naturally reduced.
 This brings us to the topic of model allocation, as also implemented in \textit{PythonPDEVS}~\cite{PythonPDEVS2}.
@@
 
+Both allocators will try to divide models evenly over kernels.
+The effects of varying the number of models per kernel are already evaluated in the previous section on scaling.
+Here we want to highlight the overhead of communication and inter-kernel dependence.
+
 \begin{figure}
@@
 After simulation the traces can be visualized for both breadth-first and depth-first allocation.
-Using a breadth-first allocation scheme, as shown in Figure~\ref{fig:pholdtree_visualize_parBFS}, we notice that many events get exchanged between cores.
-This is caused by the high number of inter-core connections and the high number of events exchanged over these connections.
-The number of connections between nodes at the same simulation core is also rather low.
-Using a depth-first allocation scheme, however, as shown in Figure~\ref{fig:pholdtree_visualize_parDFS}, minimizes inter-core connections while maximizing intra-core connections.
+Using a breadth-first allocation scheme, as shown in Figure~\ref{fig:pholdtree_visualize_parBFS}, we notice that many events get exchanged between kernels.
+This is caused by the high number of inter-kernel connections and the high number of events exchanged over these connections.
+The number of connections between nodes at the same simulation kernel is also rather low.
+Using a depth-first allocation scheme, however, as shown in Figure~\ref{fig:pholdtree_visualize_parDFS}, minimizes inter-kernel connections while maximizing intra-kernel connections.
 
@@
 
-The most prominent aspect of these results is the low performance for conservative depth-first allocation for two cores.
+The most prominent aspect of these results is the low performance for conservative depth-first allocation for two kernels.
 This is mostly caused by the difference between a sequential simulation and a parallel simulation: suddenly we need to take into account synchronization and passing around of lookahead values.
-And since the number of cores is low, the overhead dominates any further speedup.
-Optimistic is less sensitive to the number of models per core as it does not need to poll each model for a lookahead, this explains the lower runtime penalty observed for optimistic.
-%TODO add some more explanation here maybe?
-%TODO why is this not the case for optimistic?
+And since the number of kernels is low, the overhead dominates any further speedup.
+Optimistic is less sensitive to the number of models per kernel as it does not need to poll each model for a lookahead, this explains the lower runtime penalty observed for optimistic.
 
-%TODO this paragraph is strange: is this really the case? 
 %The effect is due to the topology of the kernels, in BFS the risk on reverts is higher. Optimistic is less influenced but will get a lower speedup in the good case.

### 2016-09-30 publications/Journal/memory.tex

@@
 Apart from simulation execution time, memory usage during simulation is also of great importance.
-While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation unfeasible.
+While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation infeasible.
 We therefore also investigate memory usage of different synchronization protocols, and again compare to \textit{adevs}.
@@
 Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
-\textit{Tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
+\textit{tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
 Due to our motivation for memory usage analysis, we will only measure peak allocation in maximum resident set size as reported by the OS.

### 2016-09-30 publications/Journal/models.tex

@@
 
-Second, the \textit{PHOLD} model~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
+Second, the \textit{Interconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
+Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event.
+The model takes one parameter: the number of models.
+This benchmark shows the complexity of event creation, event routing, and simultaneous event handling.
+An example for three models is shown in Figure~\ref{fig:interconnect_model}.
+
+Third, the \textit{PHOLD} model~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
 Each atomic model is directly connected to every other atomic model.
@@
 An example for four models, split over two nodes, is shown in Figure~\ref{fig:PHOLD_model}.
-
-Third, the \textit{Interconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
-Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event.
-The model takes one parameter: the number of models.
-This benchmark shows the complexity of event creation, event routing, and simultaneous event handling.
-An example for four models is shown in Figure~\ref{fig:interconnect_model}.
 

### 2016-10-03 publications/Journal/1-introduction.tex
Small changes to section 1

@@
 Whereas \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, its inherent parallelism is often insufficient~\cite{Himmelspach}.
-Several synchronization protocols from the discrete event simulation community~\cite{FujimotoBook} have been applied to \textsf{DEVS} simulation~\cite{globaltimewarp}.
-Through synchronization protocols, different simulation cores can be at different points in simulated time, increasing parallelism significantly, at the cost of synchronization overhead.
+Several synchronization protocols from the discrete event simulation community~\cite{FujimotoBook} have been applied to (\textsf{Parallel}) \textsf{DEVS} simulation~\cite{globaltimewarp}.
+With synchronization protocols, different simulation cores can be at different points in simulated time, significantly increasing parallelism, at the cost of synchronization overhead.
 While several parallel \textsf{DEVS} simulation kernels exist, they are often limited to a single synchronization protocol.
-The reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other in specific models~\cite{Jafer}.
-The applicability of parallel simulation capabilities of current tools is therefore constrained to specific domains.
+The reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other for specific models~\cite{Jafer}.
+The parallel simulation capabilities of current tools are therefore limited to specific domains.
 
 This paper introduces \textit{DEVS-Ex-Machina}\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``\textit{dxex}''): our simulation tool~\cite{dxex} which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
-The selected synchronization protocol is transparent to the simulated model: users only determine which protocol to use.
-Users who simulate a wide variety of models, with different ideal synchronization protocols, can run the same model with different synchronization protocols.
-Since the ideal synchronization protocol might change throughout the simulation, we also support runtime switching between these protocols.
+The selected synchronization protocol is transparent to the simulated model: users only determine the protocol to use.
+Users who simulate a wide variety of models, with different ideal synchronization protocols, can run the exact same model using the exact same tool, but with different synchronization protocols.
+As model behaviour, and thus the ideal synchronization protocol, might change during simulation, runtime switching of synchronization protocols is also supported.
 This runtime switching can be based on performance metrics, which are logged during simulation.
-Information is made available to a separate component, where a choice can be made about which synchronization protocol to use next. % It's manual now, can be extended to full automatic.
-Additionally, we investigate in this paper how model allocation influences the performance of our synchronization protocols.
-To this end, we have included an allocation component to our simulation kernel.
+Information is made available to a separate component, where a choice can be made about which synchronization protocol to use. % It's manual now, can be extended to full automatic.
+Additionally, we investigate how model allocation influences the performance of our synchronization protocols.
+To this end, we have included an allocation component in our simulation kernel.
 
-Our tool is based on \textit{PythonPDEVS}, but implemented in C++11 for increased performance, using features from the new C++14 standard when supported by the compiler.
+Our tool is based on \textit{PythonPDEVS}~\cite{PythonPDEVS}, but implemented in C++11 for increased performance, using features from the new C++14 standard when supported by the compiler.
 Unlike \textit{PythonPDEVS}, \textit{dxex} only supports multicore parallelism, thus no distributed simulation.
 
-We implemented a model that, depending on a single parameter, changes its ideal synchronization protocol.
-Using several models, we demonstrate the factors influencing the performance under a given synchronization protocol.
+Using several benchmark models, we demonstrate the factors influencing the performance under a given synchronization protocol.
+Additionally, we investigate a model which changes its behaviour (and ideal synchronization protocol) during simulation.
 \textit{Dxex}, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol.
-With \textit{dxex} users can always opt to use the fastest protocol available, and through its modularity, users could even implement their own.
-To verify that this modularity does not hinder performance, we compare to another \textsf{Parallel DEVS} simulation kernel, called \textit{adevs}~\cite{adevs}.
+With \textit{dxex} users can always opt to use the fastest protocol available, and through its modularity, users could even implement their own, or variants of existing ones.
+To verify that this modularity does not hamper performance, we compare to \textit{adevs}~\cite{adevs}, another \textsf{Parallel DEVS} simulation tool.
 
-%Usage of Hyperref with hardcoded label is hardly ideal, but ~\ref doesn't work if the layout disables numbered sections, so a hack is needed.
-The alternative is \nameref but that isn't all that readable.
 The remainder of this paper is organized as follows:

### 2016-10-03 publications/Journal/2-background.tex
Small changes to section 2

@@
-This section briefly introduces the synchronization protocols used by \textit{dxex}: conservative and optimistic synchronization.
+This section briefly introduces the two most prominent synchronization protocols: conservative and optimistic synchronization.
+Both algorithms are supported by Dxex.
 
@@
 Deadlocks can occur due to a dependency cycle of models.
-Multiple algorithms are defined in the literature to handle both the core protocol, as well as resolution schemes to handle or avoid the deadlocks~\cite{FujimotoBook}.
+Multiple algorithms are defined in the literature to handle both the base protocol, as well as resolution schemes to handle or avoid the deadlocks~\cite{FujimotoBook}.
 
-The main advantage of conservative synchronization is its low overhead if the lookahead is high.
-Each node then simulates in parallel, and sporadically notifies other nodes about its local simulation time.
+The main advantage of conservative synchronization is its low overhead if lookahead is high.
+Each node then simulates in parallel, notifying other nodes about its local simulation time.
 The disadvantage, however, is that the amount of parallelism is explicitly limited by the lookahead.
-If a node can influence another (almost) instantaneously, no matter how rarely it occurs, the amount of parallelism is severely reduced.
+If a node can influence another (almost) instantaneously, no matter how rarely it occurs, the amount of parallelism is severely reduced throughout the complete simulation run.
 The user is required to define the lookahead, using knowledge about the model's behaviour.
-Defining lookahead is far from a trivial task if there is no detailed knowledge of the model.
+Defining an accurate and high lookahead is far from a trivial task if there is no detailed knowledge of the model.
 Even slight changes in the model or its allocation can change the lookahead, and can therefore have a significant influence on simulation performance.
@@
 A completely different synchronization protocol is \textit{optimistic synchronization}~\cite{TimeWarp}.
-Whereas conservative synchronization prevents causality errors at all costs, optimistic synchronization allows them to happen, but corrects them.
-Each node simulates as fast as possible, without taking note of any other node.
+Whereas conservative synchronization prevents causality errors at all costs, optimistic synchronization allows them to happen, but corrects them as soon as they are detected.
+Each node simulates as fast as possible, independent of other nodes.
 It assumes that no events occur from other nodes, unless it has explicitly received one at that time.
-When this assumption is violated, the node rolls back its simulation time and state to right before the moment when the event has to be processed.
-As simulation is rolled back to a time prior to the event must be processed, the event can then be processed as if no causality error ever occurred.
+When this assumption is violated (\textit{i.e.}, an event arrives that should have been processed in the past), the node rolls back its simulation time and state to right before the moment when the event has to be processed.
+As simulation is rolled back to a point in time before the event has to be processed, the event can be processed as if no causality error ever occurred.
 
@@
 
-The main advantage is that a small lookahead, caused by a very infrequent event, will not limit performance.
-%The main advantage is that performance is not limited by a small lookahead caused by a very infrequent event.
+The main advantage is that a small lookahead, caused by infrequent events, does not limit performance.
 If an (almost) instantaneous event rarely occurs, performance is only impacted when it occurs, and not at every simulation step.
 The disadvantage is unpredictable performance due to the arbitrary cost of rollbacks and their propagation.
+Even the occurence of rollbacks is non-deterministic, as it is caused by the interleaving of different simulation nodes and their communication.
 If rollbacks occur frequently, state saving and rollback overhead can cause simulation to grind to a halt.
-Apart from overhead in CPU time, a significant memory overhead is present: intermediate states are stored up to a point where it can be considered \textit{irreversible}.
+Nonetheless, it can be proven that simulation always progresses, and eventually always terminates.
+Apart from overhead in CPU time, a significant memory overhead is present: intermediate states are stored up to a point where they can be considered \textit{irreversible}.
 Note that, while optimistic synchronization does not explicitly depends on lookahead, performance still implicitly depends on lookahead.

### 2016-10-03 publications/Journal/3-features.tex
Small changes to section 3

@@
 Historically, \textit{dxex} is based on \textit{PythonPDEVS}~\cite{PythonPDEVS}.
-Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
+Python is a good language for prototypes, but performance has proven insufficient to compete with C++-based simulation kernels~\cite{MasterThesis}.
 \textit{Dxex} is a C++11-based \textsf{Parallel DEVS} simulator, based on the design of \textit{PythonPDEVS}.
-Whereas the feature set is not too comparable, the architectural design, simulation algorithms, and optimizations, are highly similar.
+Whereas the feature set is not too comparable, the architectural design, simulation algorithms, and optimizations are highly similar.
 
@@
 But whereas \textit{PythonPDEVS} only supports optimistic synchronization, \textit{dxex} support multiple synchronization protocols (though only in parallel).
-This is in line with the design principle of \textit{PythonPDEVS}: allow users to pass performance hints to the simulation kernel.
-In our case, a user can pass the simulation kernel the synchronization protocol to use for this model, or even switch the synchronization protocol during runtime.
-Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language. \textit{Dxex} will use new optimizations from the C++14 standard when possible.
+This is in line with the design principle of \textit{PythonPDEVS}: allow users to pass performance hints~\cite{PythonPDEVS_ACTIMS} to the simulation kernel.
+In our case, a user can configure the simulation kernel with the synchronization protocol to use, or even switch the synchronization protocol during runtime.
+Our implementation in C++11 also allows for (compiled) optimizations which were plainly impossible in an interpreted language.
+\textit{Dxex} uses new optimizations from the C++14 standard when possible.
+The C++11 standard threading primitives are used to run the different simulation cores.
+Within a single simulation core, OpenMP~\cite{openmp4} is used to parallelize the transition functions, as is usual in \textsf{Parallel DEVS}.
 
@@
 
-In the remainder of this section, we will elaborate on our prominent new feature: the efficient implementation of multiple synchronization protocols within the same simulation tool, which are offered transparently to the model.
+In the remainder of this section, we elaborate on our prominent new feature: the efficient implementation of multiple synchronization protocols within the same simulation tool, offered transparently to the model.
 
@@
 We previously explained the existence of different synchronization protocols, each optimized for a specific kind of model.
-As no single synchronization protocol is ideal for all models, a general purpose simulation tool should support multiple protocols.
-Currently, most parallel simulation tools choose only a single synchronization protocol due to the inherent differences between protocols.
-We argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for \textit{dxex}.
+As no single synchronization protocol is ideal for all domains, a general purpose simulation tool should support multiple protocols.
+We argue that a general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for \textit{dxex}.
 
@@
 Optimistic synchronization for when lookahead is unpredictable, or there are rare (almost) instantaneous events.
-Finally, sequential simulation is still required for models where parallelism is bad, where all protocols actually slow down simulation.
+Finally, sequential simulation is still required for models where parallelism is bad. 
+It is possible for the synchronization overhead to become larger than the achieved parallelism, resulting in slower simulation than sequential.
+
+Data exchange between different simulation cores happens through shared memory, using the new C++11 synchronization primitives.
+Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
+C++11 further allows us to make the implementation portable, as well as more efficient: the compiler makes further optimizations to heavily used components.
 
@@
 Within the returned time interval, the model promises not to raise an event.
-A node aggregates this information to compute its Earliest Output Time (EOT).
-This value is communicated to other nodes through shared memory.
-
-Reading and writing to shared memory is done through the use of the new C++11 synchronization primitives.
-Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
-C++11 further allows us to make the implementation portable, as well as more efficient: the compiler makes further optimizations to heavily used components.
 
@@
 Mattern's algorithm~\cite{mattern} is used to determine the GVT, as it runs asynchronously and uses only $2n$ synchronization messages.
-Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed, and irreversible actions are committed.
+Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed, and irreversible actions (\textit{e.g.}, printing) are committed.
 
@@
 \subsection{Synchronization Protocol Transparency}
-We define synchronization protocol transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
+We define synchronization protocol transparency as having a single model, which can be executed on each supported synchronization kernel, without any modifications.
 User should thus only provide one model, implemented in C++11, which can be simulated using sequential execution, conservative synchronization, or optimistic synchronization.
@@
 \label{sec:4-subsec:overhead-pgraph:memory}
-\begin{figure}
-    \center
-    \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
-    \caption{Effect of memory allocators on parallel execution time.}
-    \label{fig:memallocators_parallel}
-\end{figure}
-Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}: memory bandwidth doesn't increase as fast as the number of cores using it.
+Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}, since memory bandwidth doesn't increase as fast as the number of cores using it.
 While this is always a problem, it is aggravated in \textit{dxex} by providing automatic memory management for events and states.
@@
 They can be returned immediately to the memory allocator as the lifetime of these objects is known at creation.
-In contrast inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
-As a result simulations with many inter-kernel events suffer a performance hit, whereas the impact of many intra-kernel events can be minimized using arena ~\cite{Arena} allocators.
+In contrast, inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
+As a result, simulations with many inter-kernel events suffer a performance hit, whereas the impact of many intra-kernel events can be minimized using arena allocators~\cite{Arena}.
 
 \textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
-The latter can be faster, but at the cost of extra configuration.
-The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}, which reduces lock contention in \texttt{malloc} calls.
+The latter can be faster, but at the cost of additional configuration.
+The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}, reducing lock contention in \texttt{malloc} calls.
+
+\begin{figure}
+    \center
+    \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
+    %TODO: Add value on top of bars
+    \caption{Effect of memory allocators on parallel execution time, using optimistic synchronization.}
+    \label{fig:memallocators_parallel}
+\end{figure}
 
@@
 
-We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default Standard Template Library (STL) random number generator.
+We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default Standard Template Library (STL) random number generator with optimistic synchronization.
 This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
@@
 
-Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in sequential simulation is three times faster using TRNG compared to using the STL RNG.
+Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in sequential simulation is three times faster using any of the three alternatives, only the STL RNG is slow.
 For parallel simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between sequential and parallel simulation.
-Conservative synchronization is almost insensitive to the changing of the RNG, though a slight increase in performance can be noted.
-Optimistic synchronization slows down significantly when the RNG becomes part of the model state, since the state needs to be copied as well.
-This becomes a significant overhead when using the STL RNG, since performance plummets to a fraction of the original.
-Using TRNG avoids this problem completely, as the size of the RNG state is negligible.
+Conservative synchronization is almost unaffected by changing the RNG.
 
@@
     \includegraphics[width=\columnwidth]{fig/rngspeedupeffectdevstone.eps}
+    %TODO: why is STL RNG in state just as fast in sequential? Does storing it in the state really do that much?
+    %TODO: consistency: TRNG, STL RNG in state, and TRNG in state
     \caption{Speedup with different RNG usage patterns compared to STL random number generator.}

### 2016-10-03 publications/Journal/4-performance.tex
Update Section 4 : model description.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
Shuffling around some sections
New order of sections
Reword conclusions
Initial section of pdevs.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
Merged performance section in one tex file

@@
-% What we want to write
+In this section, we evaluate the performance of different synchronization protocols in \textit{dxex}.
+We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVStoneJournal}, to show that our modularity does not impede performance.
+CPU time and memory usage is compared for both sequential and parallel simulation.
 
-% DESIGN OF MODELS
-\input{models.tex}
+We start off with a comparison of sequential simulation, to show how \textit{adevs} and \textit{dxex} relate in this simple case.
+For the parallel simulation benchmarks, results are presented for both conservative and optimistic synchronization.
 
-% SEQUENTIAL
-\input{sequential-performance.tex}
+For all benchmarks, results are well within a 5\% deviation of the average, such that only the average is used in the remainder of this section.
+The same compilation flags were used for both \textit{adevs} and \textit{dxex} benchmarks (``\texttt{-O3 -flto}'').
+To guarantee comparable results, no I/O was performed during benchmarks.
+Before benchmarking, simulation traces were used to verify that \textit{adevs} and \textit{dxex} return exactly the same simulation results.
+Benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.
+The exact parameters for each benchmark can be found in the repository, as well as the data used in this paper.
 
-% PARALLEL
-\input{parallel-performance.tex}
+\subsection{Benchmark Models}
+We use three different benchmark models, covering different aspects of the simulation kernel.
 
-% CONCLUSION
-\input{conclusion.tex}
+First, the \textit{Queue} model, based on the \textit{HI} model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
+A single generator pushes events into the queue, which get consumed by the processors after a fixed or random delay.
+It takes two parameters: the width and depth of the hierarchy.
+This benchmark shows how the complexity of the simulation kernel behaves for an increasing amount of atomic models, and an increasingly deep hierarchy.
+An example for width and depth equal to 2 is shown in Figure~\ref{fig:queue_model}.
+
+Second, the \textit{Interconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
+Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event.
+The model takes one parameter: the number of models.
+This benchmark shows the complexity of event creation, event routing, and simultaneous event handling.
+An example for three models is shown in Figure~\ref{fig:interconnect_model}.
+
+Third, the \textit{PHOLD} model~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
+Each atomic model is directly connected to every other atomic model.
+After a random delay, an atomic model sends out an event to a randomly selected output port.
+Output port selection happens in two phases: first it is decided whether the event should be sent within the same node, or outside of the node.
+Afterwards, a uniform selection is made between the remaining models.
+The model takes two parameter: the percentage of remote events, which determines the fraction of messages routed to other nodes, and the percentage of priority events.
+Priority events are events generated in a very short time after the previous event.
+This benchmark shows how the simulation kernel behaves in the presence of many local or remote events, in combination with a varying percentage of high-priority event.s
+An example for four models, split over two nodes, is shown in Figure~\ref{fig:PHOLD_model}.
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/queue_model_fixed.pdf}
+	\caption{Queue model for depth and width 2.}
+	\label{fig:queue_model}
+\end{figure}
+
+\begin{figure}
+    \center
+	\includegraphics[width=\modelfraction\columnwidth]{fig/interconnect_model.pdf}
+	\caption{Interconnect model for three models.}
+	\label{fig:interconnect_model}
+\end{figure}
+
+\begin{figure}
+    \center
+	\includegraphics[width=\modelfraction\columnwidth]{fig/phold_model.pdf}
+	\caption{PHOLD model for three models.}
+	\label{fig:PHOLD_model}
+\end{figure}
+
+\subsection{Sequential Simulation}
+We start by evaluating sequential simulation performance, in order to obtain a baseline for our comparison of parallel simulation performance.
+
+\subsubsection{Queue}
+\label{4-seq-Queue}
+% explain what we did: fixed number of models (400, 600 & 800), increasing depth
+For the first benchmark, we tested the effect of hierarchical complexity of the model in the performance of the simulator.
+A set of three tests was performed, where each test has the same number of models but an increasing depth.
+The results can be seen in Figure~\ref{fig:Queue_benchmark_seq}.
+Since \textit{dxex} performs direct connection on the model, there is no performance hit when the depth is increased.
+Direct connection only needs to be done at initialization, so it is a one time cost that is negligible when simulation termination end time is sufficiently large.
+\textit{Adevs}, on the other hand, suffers from the increased depth, even though some similar (but not identical) optimization to event passing was made~\cite{adevs_opt}.
+With every new hierarchical layer, routing an event from one atomic model to the next becomes more expensive, resulting in an increase in runtime.
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/queue_fixed_sequential.eps}
+	\caption{Queue benchmark results for sequential simulation.}
+	\label{fig:Queue_benchmark_seq}
+\end{figure}
+
+\subsubsection{Interconnect}
+\label{4-seq-Interconnect}
+In the Interconnect model, we increase the number of atomic models, quadratically increasing the number of couplings and the number of external transitions.
+As shown in Figure~\ref{fig:Interconnect_benchmark}, \textit{adevs} now outperforms \textit{dxex} by a fair margin.
+Analysis showed that this is caused by the high amount of events: event creation is much slower in \textit{dxex} than it is in \textit{adevs}, despite \textit{dxex}'s use of memory pools.
+To shield the user from threading and deallocation concerns \textit{dxex} provides an event superclass from which the user can derive to create a specialized event type.
+Copying, deallocation, and tracing are done at runtime, adding significant overhead when events happen frequently.
+Profiling the benchmarks clearly shows the increasing cost of output generation and deallocation as the determining factor in the gap in performance.
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/interconnect_sequential.eps}
+	\caption{Interconnect benchmark results for sequential simulation.}
+	\label{fig:Interconnect_benchmark}
+\end{figure}
+
+\subsubsection{PHold}
+\label{4-seq-PHold}
+The PHold model is very similar to the Interconnect model.
+The biggest difference is that the amount of messages sent is much lower.
+The number of events scales linear with the number of models, not quadratic.
+Figure~\ref{fig:Phold_benchmark} shows that the performance of \textit{dxex} and \textit{adevs} are very close to each-other, with \textit{adevs} slightly outperforming \textit{dxex}.
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/phold_sequential.eps}
+	\caption{PHold benchmark results for sequential simulation.}
+	\label{fig:Phold_benchmark}
+\end{figure}
+
+\subsection{Parallel Simulation}
+We now continue by describing our parallel simulation performance for different synchronization protocols, and compared to \textit{adevs}.
+The speedup of \textit{adevs} is computed with the corresponding \textit{dxex} sequential benchmark.
+This was done to take into account the performance difference observed in sequential simulation.
+
+\subsubsection{Queue}
+The Queue model is one single chain of models, resembling a pipeline.
+This structure can be exploited to prevent cyclic dependencies in the parallel simulation.
+
+Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to sequential simulation for a fixed problem size.
+%TODO this is actually quite a strange choice...
+As the number of kernels increases, the optimistic kernel quickly becomes the worst choice.
+This is mainly caused by the pipeline structure of the model: the last models in the queue only respond to incoming messages and therefore have to be rolled back frequently.
+The difference between \textit{dxex} conservative and \textit{adevs} becoms smaller when more and more cores are used.
+The same effect can be seen for weak scaling in Figure~\ref{fig:Queue_plot_weak}.
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/queue_fixed_weak_speedup.eps}
+	\caption{Queue model weak scaling speedup compared to \textit{dxex} sequential.}
+	\label{fig:Queue_plot_weak}
+\end{figure}
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/queue_fixed_strong_speedup.eps}
+	\caption{Queue model strong scaling speedup compared to \textit{dxex} sequential.}
+	\label{fig:Queue_plot_strong}
+\end{figure}
+
+\subsubsection{Interconnect}\label{subsec:parallelinterconnect}
+In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
+The number of models is now kept constant at eight.
+Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
+When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and the increasing number of rollbacks in optimistic simulation.
+All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
+	\caption{Interconnect benchmark results for parallel simulation.}
+	\label{fig:interconnect_benchmark_parallel}
+\end{figure}
+
+\subsubsection{PHOLD}
+In the PHOLD model, we first investigate the influence of the percentage of remote events on the speedup.
+A remote event in this context is an event that is sent from a model on one kernel to a model on another simulation kernel.
+When remote events are rare, optimistic synchronization rarely has to roll back, thus increasing performance.
+With more common remote events, however, optimistic synchronization quickly slows down due to frequent rollbacks.
+Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block and wait.
+Even though a single synchronization protocol is always ideal in this case, it already shows that different synchronization protocols respond differently to a changing model.
+
+\textit{Adevs} is significantly slower during conservative synchronization.
+Analysis of profiling callgraphs shows that exception handling in \textit{adevs} is the main cause. 
+To keep the models equivalent, the \textit{adevs} version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.
+These functions require the user to implement a state saving method.
+But in contrast to \textit{PythonPDEVS} and \textit{dxex}, which handle this inside the kernel, users need to manually define this.
+We feel this would lead to an unfair comparison as we would like to keep the models agnostic of the underlying protocols across all benchmarks.
+
+\begin{figure}
+    \center
+    \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
+    \caption{PHOLD benchmark results for parallel simulation using four kernels, four atomics per node, with varying percentage of remote events.}
+\end{figure}
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/phold_priority.eps}
+	\caption{PHOLD benchmark results for parallel simulation using four kernels, with varying amount of high-priority events.}
+	\label{fig:phold_priority}
+\end{figure}
+
+We slightly modified the PHOLD benchmark to include high-priority events.
+Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
+Even when normal events occur most often, conservative synchronization always blocks until it can make guarantees.
+Optimistic synchronization, however, simply goes forward in simulation time and rolls back when these high-priority events happen.
+This situation closely mimics the model typically used for comparing conservative and optimistic synchronization~\cite{FujimotoBook}.
+
+Figure~\ref{fig:phold_priority} shows how simulation performance is influenced by the fraction of these high-priority events.
+If barely any high-priority events occur, conservative synchronization is penalized due to its excessive blocking, which often turned out to be unnecessary.
+When many high-priority events occur, optimistic synchronization is penalized due to its unconditional progression of simulation, which frequently needs to be rolled back.
+Results show that there is no single perfect synchronization algorithm for this model: depending on configuration, either synchronization protocol might be better.
+\subsection{Comparison with PDEVS}
+In this section we contrast the speedup between the original PDEVS algorithm and the synchronization protocols. 
+Note that the PDEVS algorithm and the synchronization protocols are complementary in \textit{dxex}, not exclusive. 
+Following the theoretical analysis published in ~\cite{amdahlpdevs} a comparison between PDEVS and the synchronization protocols was warranted. 
+We compare the ideal scenario for PDEVS where all events are concurrent with a scenario where events are random.
+In this work we do not consider the distribution of events, introducing a random number generator as we will show in this work can mask performance results. 
+By limiting ourselves to the best and worst case scenario we demonstrate the capabilities and limitations of both approaches.
+The influence of the computational load of the transition functions is a second factor we investigate in this section.
+We are interested in scenarios where PDEVS and the synchronization protocols can complement each other in addition to observing their speedup in isolation.
+The benchmarks are run on the same hardware as our previous results, a 8 x AMD Opteron(TM) Processor 6274 machine with 8 cores per cpu (64 cores) and 192 GB RAM. 
+The benchmarks use a maximum of 16 threads to be divided between PDEVS , conservative or optimistic.
+
+\paragraph{Model}
+As shown in the previous section, when most models are dependent on each other across kernels there is no speedup obtainable by either optimistic or conservative. 
+We therefore opted to use the \textit{Queue} model in this benchmark.
+The effect of combining 4 synchronized kernels with each 4 threads for PDEVS is contrasted with 16 threads for PDEVS and 16 kernels for conservative and optimistic.  
+This allows us to observe which is more efficient in obtaining a speedup.
+We simulate a computational load by an optional call to sleep in the transition functions set at 5ms. 
+The model is configured with depth 4 and width 300 if the transition function has no load, and width 50 if the computational load is active.
+In our configuration an imminent model will generate output and lead to another model becoming imminent. With respect to the 
+theoretical analysis in~\cite{amdahlpdevs} this gives a parameter value q = 1. 
+The models are equally distributed over the kernels and between threads in PDEVS. 
+The coefficient of variation of our results is less than 1\%. 
+Communication between the kernels and threads, defined as communication overhead in the theoretical work is hard to to estimate here but given our coefficient of variation results we can assume this to be constant.
+
+\paragraph{Concurrent events}
+First we create the ideal scenario for PDEVS, all events are concurrent with a significant computational load in the transition functions.
+All models are imminent at the same time in this configuration allowing for maximum parallelism.
+In figure \ref{fig:pdevs_plot_fixed_sleep} we observe that PDEVS obtains a speedup approaching 10.  
+Conservative and optimistic are run first with 16 kernels, then with 4 kernels having 4  PDEVS threads each. 
+Both protocols achieve a similar speedup compared to PDEVS. Combining the protocols with PDEVS gives a near identical speedup.
+In this scenario there is no real advantage between the different parallel configurations.
+
+\paragraph{Random events}
+Random events allow us to demonstrate the overhead of PDEVS. The probability that events are concurrent is strongly reduced.
+The transition function has the same computational load as in the previous configuration. 
+The results in figure \ref{fig:pdevs_plot_random_sleep} show that PDEVS adds little overhead in sequential simulation. 
+In combination with either synchronization protocol DPEVS can achieve a reasonable speedup, but both synchronization protocols achieve a far higher speedup without PDEVS.
+
+\paragraph{Computational load}
+When we remove the computational load we isolate the overhead of the parallelism. 
+In this paper we are mainly interested in this overhead which is otherwise masked by adding a computational cost to the transition functions. 
+Figure \ref{fig:pdevs_plot_no_sleep} shows that in sequential simulation the PDEVS overhead is crippling.
+Conservative combined with PDEVS reduces performance to that of sequential without PDEVS with worse results for optimistic in combination in PDEVS.
+Both synchronization protocols achieve a reasonable speedup without PDEVS.
+
+\paragraph{Discussion}
+In \textit{dxex} PDEVS and the two synchronization protocols can be complementary. 
+The user is not forced to choose between either but can divide the capabilities of his machine over both approaches.
+While PDEVS can offer a significant speedup, we see that to offer this speedup the concurrency of events must be significant in combination with a high computational load in the transition functions. 
+We measure the overhead introduced by PDEVS to give practitioners insight into which model can benefit from which paradigm.
+The overhead of the synchronization protocols in specific scenarios  is shown in the previous section.
+For a given model it is difficult to predict the concurrency of events, although the computational load of a transition function can be approximated.
+In these benchmarks we show which configurations can lead to a speedup for PDEVS or the synchronization protocols in either isolation or combination.
+In a model with high concurrency in simultaneous events and a high computational load PDEVS can offer a significant speedup.
+This speedup is not exclusive to PDEVS. Conservative and optimistic synchronization protocols can achieve the same speedup given a non cyclic model. 
+From the theoretical model we know that in a cyclic model PDEVS can achieve a higher speedup.
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/pdevs_fixed_sleep.eps}
+	\caption{Queue speedup benchmark for PDEVS and synchronization with maximal concurrent events and significant computational load}
+	\label{fig:pdevs_plot_fixed_sleep}
+\end{figure}
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/pdevs_random_sleep.eps}
+	\caption{Queue speedup benchmark for PDEVS and synchronization with randomized events and significant computational load}
+	\label{fig:pdevs_plot_random_sleep}
+\end{figure}
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/pdevs_no_sleep.eps}
+	\caption{Queue speedup benchmark for PDEVS and synchronization with maximal concurrent events and trivial computational load}
+	\label{fig:pdevs_plot_no_sleep}
+\end{figure}
+
+\subsection{Memory Usage}
+Apart from simulation execution time, memory usage during simulation is also of great importance.
+While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation infeasible.
+We therefore also investigate memory usage of different synchronization protocols, and again compare to \textit{adevs}.
+
+We do not tackle the problem of states that become too large for a single machine to hold.
+This problem can be mitigated by distribution over multiple machines, which neither \textit{dxex} or \textit{adevs} support.
+
+\subsubsection{Remarks}
+Both \textit{dxex} and \textit{adevs} use \textit{tcmalloc} as memory allocator, allowing for thread-local allocation.
+Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
+\textit{tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
+Due to our motivation for memory usage analysis, we will only measure peak allocation in maximum resident set size as reported by the OS.
+
+\subsubsection{Results}
+Figure~\ref{fig:memory} shows the memory used by the different benchmarks.
+Results are in megabytes, and show the total memory footprint of the running application (\textit{i.e.}, text, stack, and heap).
+Note the logarithmic scale due to the high memory consumption of optimistic synchronization.
+
+Unsurprisingly, optimistic synchronization results show very high memory usage due to the saved states.
+Note the logarithmic scale that was used for this reason.
+Optimistic synchronization results vary heavily depending on thread scheduling by the operating system, as this influences the drift between nodes. 
+Comparing similar approaches, we notice that \textit{dxex} and \textit{adevs} have very similar memory use.
+
+Conservative simulation always uses more memory than sequential simulation, as is to be expected.
+Additional memory is required for the multiple threads, but also to store all events that are processed simultaneously.
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{fig/memory_voorlopig.eps}
+    \caption{Memory usage results.}
+    \label{fig:memory}
+\end{figure}
+
+\subsection{Conclusions on Performance Evaluation}
+We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
+PDEVS is compared with the synchronization protocols either in isolation or complementing each other. We have shown that both approaches are useful in specific model configurations.
+But even with the right synchronization protocol, we have seen that two problems remain.
+
+First, although one of both synchronization protocols might be ideally suited for specific model behaviour, nothing guarantees that the model will exhibit the same behaviour throughout the simulation.
+Similarly to the polymorphic scheduler~\cite{MasterThesis}, we wish to make it possible for the ideal option to be switched during simulation.
+When changes to the model behaviour are noticed, the used synchronization protocol can be switched as well.
+
+Second, the allocation of models is nontrivial and has a significant impact on performance.
+While our parallel speedup for the Queue model, for example, was rather high, this is mostly due to characteristics of the model: the dependency graph does not contain any cycles.
+When cycles were introduced, as in the Interconnect model, performance became disastrous.
+
+In the next two sections, we elaborate on these two problems.

### 2016-10-03 publications/Journal/4-performance.tex
Several changes to section 4, mainly terminology about the inherent parallelism of PDEVS

@@
 Benchmarks were performed using Linux, but our simulation tool works equally well on Windows and Mac.
-The exact parameters for each benchmark can be found in the repository, as well as the data used in this paper.
+The exact parameters for each benchmark can be found in our repository.
+
+The benchmarks are ran on a machine with 8 x AMD Opteron(TM) Processor 6274 with 8 cores per CPU (for a total of 64 cores) and 192 GB RAM.
 
@@
 This benchmark shows how the complexity of the simulation kernel behaves for an increasing amount of atomic models, and an increasingly deep hierarchy.
-An example for width and depth equal to 2 is shown in Figure~\ref{fig:queue_model}.
+An example for a width and depth of 2 is shown in Figure~\ref{fig:queue_model}.
 
 Second, the \textit{Interconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
-Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event.
+Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event (\textit{i.e.}, the event is broadcasted).
 The model takes one parameter: the number of models.
@@
 Afterwards, a uniform selection is made between the remaining models.
-The model takes two parameter: the percentage of remote events, which determines the fraction of messages routed to other nodes, and the percentage of priority events.
-Priority events are events generated in a very short time after the previous event.
-This benchmark shows how the simulation kernel behaves in the presence of many local or remote events, in combination with a varying percentage of high-priority event.s
+The model takes two parameters: the percentage of remote events (determining the percentage of messages routed to other nodes), and the percentage of high-priority events.
+High-priority events are events generated in a very short time after the previous event.
+This benchmark shows how the simulation kernel behaves in the presence of many local or remote events, in combination with a varying percentage of high-priority events.
 An example for four models, split over two nodes, is shown in Figure~\ref{fig:PHOLD_model}.
@@
 \label{4-seq-Queue}
-% explain what we did: fixed number of models (400, 600 & 800), increasing depth
 For the first benchmark, we tested the effect of hierarchical complexity of the model in the performance of the simulator.
@@
 The results can be seen in Figure~\ref{fig:Queue_benchmark_seq}.
-Since \textit{dxex} performs direct connection on the model, there is no performance hit when the depth is increased.
-Direct connection only needs to be done at initialization, so it is a one time cost that is negligible when simulation termination end time is sufficiently large.
+Since \textit{dxex} performs direct connection~\cite{SymbolicFlattening} on the model, there is no performance hit when the depth is increased.
+Direct connection only needs to be done at initialization, so it is a one time cost that is negligible for long running simulations.
 \textit{Adevs}, on the other hand, suffers from the increased depth, even though some similar (but not identical) optimization to event passing was made~\cite{adevs_opt}.
@@
 Copying, deallocation, and tracing are done at runtime, adding significant overhead when events happen frequently.
-Profiling the benchmarks clearly shows the increasing cost of output generation and deallocation as the determining factor in the gap in performance.
+Profiling the benchmarks revealed the increased cost of output generation and deallocation as the determining factor.
 
@@
 This was done to take into account the performance difference observed in sequential simulation.
+As such, the highest speedup indicates the fastest results among all tools, independent of sequential simulation results.
 
@@
 
-Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to sequential simulation for a fixed problem size.
-%TODO this is actually quite a strange choice...
+Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to sequential simulation for a fixed problem size (\textit{i.e.}, strong scaling).
 As the number of kernels increases, the optimistic kernel quickly becomes the worst choice.
@@
 The difference between \textit{dxex} conservative and \textit{adevs} becoms smaller when more and more cores are used.
-The same effect can be seen for weak scaling in Figure~\ref{fig:Queue_plot_weak}.
+The same effect can be seen when the problem size is increased in tandem with the number of used cores (\textit{i.e.}, weak scaling) in Figure~\ref{fig:Queue_plot_weak}.
+
+\begin{figure}
+	\center
+	\includegraphics[width=\columnwidth]{fig/queue_fixed_strong_speedup.eps}
+	\caption{Queue model strong scaling speedup compared to \textit{dxex} sequential.}
+	\label{fig:Queue_plot_strong}
+\end{figure}
 
@@
 
-\begin{figure}
-	\center
-	\includegraphics[width=\columnwidth]{fig/queue_fixed_strong_speedup.eps}
-	\caption{Queue model strong scaling speedup compared to \textit{dxex} sequential.}
-	\label{fig:Queue_plot_strong}
-\end{figure}
-
-\subsubsection{Interconnect}\label{subsec:parallelinterconnect}
+\subsubsection{Interconnect}
+\label{subsec:parallelinterconnect}
 In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
@@
 	\includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
+    %TODO cyclic allocation in title?
 	\caption{Interconnect benchmark results for parallel simulation.}
@@
 When remote events are rare, optimistic synchronization rarely has to roll back, thus increasing performance.
-With more common remote events, however, optimistic synchronization quickly slows down due to frequent rollbacks.
+With more frequent remote events, however, optimistic synchronization quickly slows down due to frequent rollbacks.
 Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block and wait.
@@
     \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
-    \caption{PHOLD benchmark results for parallel simulation using four kernels, four atomics per node, with varying percentage of remote events.}
+    \caption{PHOLD benchmark results for parallel simulation using four kernels, with varying percentage of remote events.}
 \end{figure}
@@
 
-We slightly modified the PHOLD benchmark to include high-priority events.
 Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
@@
 When many high-priority events occur, optimistic synchronization is penalized due to its unconditional progression of simulation, which frequently needs to be rolled back.
-Results show that there is no single perfect synchronization algorithm for this model: depending on configuration, either synchronization protocol might be better.
-\subsection{Comparison with PDEVS}
-In this section we contrast the speedup between the original PDEVS algorithm and the synchronization protocols. 
-Note that the PDEVS algorithm and the synchronization protocols are complementary in \textit{dxex}, not exclusive. 
-Following the theoretical analysis published in ~\cite{amdahlpdevs} a comparison between PDEVS and the synchronization protocols was warranted. 
-We compare the ideal scenario for PDEVS where all events are concurrent with a scenario where events are random.
-In this work we do not consider the distribution of events, introducing a random number generator as we will show in this work can mask performance results. 
-By limiting ourselves to the best and worst case scenario we demonstrate the capabilities and limitations of both approaches.
-The influence of the computational load of the transition functions is a second factor we investigate in this section.
-We are interested in scenarios where PDEVS and the synchronization protocols can complement each other in addition to observing their speedup in isolation.
-The benchmarks are run on the same hardware as our previous results, a 8 x AMD Opteron(TM) Processor 6274 machine with 8 cores per cpu (64 cores) and 192 GB RAM. 
-The benchmarks use a maximum of 16 threads to be divided between PDEVS , conservative or optimistic.
+Results show that there is no single perfect synchronization algorithm for this kind of models: depending on configuration, either synchronization protocol might be better.
+
+\subsection{Comparison to the abstract simulator}
+The abstract simulator of \textsf{Parallel DEVS} included its own notion of parallelism: transition functions that happen at the same point in simulated time can be trivially parallelized.
+This parallelization is trivial, in the sense that there is no need for synchronization protocols.
+The parallelization of the transition functions (from now on called the ``inherent parallelism'') is, at least in \textit{dxex}, independent of the use of a synchronization protocol.
+That is, both can be used simultaneously.
+Following the theoretical analysis published in~\cite{amdahlpdevs}, a comparison between this inherent parallelism and the use of synchronization protocols is warranted.
 
 \paragraph{Model}
-As shown in the previous section, when most models are dependent on each other across kernels there is no speedup obtainable by either optimistic or conservative. 
-We therefore opted to use the \textit{Queue} model in this benchmark.
-The effect of combining 4 synchronized kernels with each 4 threads for PDEVS is contrasted with 16 threads for PDEVS and 16 kernels for conservative and optimistic.  
+We have opted to use the \textit{Queue} model for this comparison, as it allows for many interesting configurations.
+%TODO what is meant with this allocation of threads and synchronized kernels?
+The effect of combining 4 synchronized kernels, with each 4 threads for PDEVS, is contrasted with 16 threads for PDEVS and 16 kernels for conservative and optimistic.
 This allows us to observe which is more efficient in obtaining a speedup.
-We simulate a computational load by an optional call to sleep in the transition functions set at 5ms. 
+We simulate a computational load by an optional call to sleep in the transition functions set at 5 milliseconds.
 The model is configured with depth 4 and width 300 if the transition function has no load, and width 50 if the computational load is active.
-In our configuration an imminent model will generate output and lead to another model becoming imminent. With respect to the 
-theoretical analysis in~\cite{amdahlpdevs} this gives a parameter value q = 1. 
-The models are equally distributed over the kernels and between threads in PDEVS. 
+In our configuration, an imminent model generates output, resulting in another model becoming imminent.
+With respect to the theoretical analysis in~\cite{amdahlpdevs} this gives a parameter value $q = 1$.
+Models are equally distributed over the kernels and between threads in PDEVS.
 The coefficient of variation of our results is less than 1\%.
-Communication between the kernels and threads, defined as communication overhead in the theoretical work is hard to to estimate here but given our coefficient of variation results we can assume this to be constant.
+The communication overhead is hard to estimate, but given our coefficient of variation, we can expect this overhead to be constant.
+
+We consider two cases: one where all transition functions happen simultaneously, and one where transition functions never happen simultaneously.
+We defer the discussion on which of these two is the most realistic, as this depends on the problem domain.
+For example, in a simulation with a fixed timestep (\textit{e.g.}, cell-based models, discretization of continuous model), transition functions will often occur simultaneously.
+Conversely, simulations with an arbitrary timestep (\textit{e.g.}, many independent systems communicating together) have very few simultaneous events.
 
 \paragraph{Concurrent events}
-First we create the ideal scenario for PDEVS, all events are concurrent with a significant computational load in the transition functions.
-All models are imminent at the same time in this configuration allowing for maximum parallelism.
-In figure \ref{fig:pdevs_plot_fixed_sleep} we observe that PDEVS obtains a speedup approaching 10.  
-Conservative and optimistic are run first with 16 kernels, then with 4 kernels having 4  PDEVS threads each. 
-Both protocols achieve a similar speedup compared to PDEVS. Combining the protocols with PDEVS gives a near identical speedup.
+First we create a model where all transitions happen simultaneously, with a significant computational load in the transition functions.
+In Figure~\ref{fig:pdevs_plot_fixed_sleep}, we observe that exploiting only the inherent parallelism, results in a speedup of about 10.  
+%TODO again: what is this?
+Simulation with conservative and optimistic synchronization is run first with 16 kernels, then with 4 kernels having 4 PDEVS threads each. 
+Both synchronization protocols achieve a speedup similar to only exploiting the inherent parallelism.
+Combining the protocols with the inherent parallelism gives near identical results.
 In this scenario there is no real advantage between the different parallel configurations.
-
-\paragraph{Random events}
-Random events allow us to demonstrate the overhead of PDEVS. The probability that events are concurrent is strongly reduced.
-The transition function has the same computational load as in the previous configuration. 
-The results in figure \ref{fig:pdevs_plot_random_sleep} show that PDEVS adds little overhead in sequential simulation. 
-In combination with either synchronization protocol DPEVS can achieve a reasonable speedup, but both synchronization protocols achieve a far higher speedup without PDEVS.
-
-\paragraph{Computational load}
-When we remove the computational load we isolate the overhead of the parallelism. 
-In this paper we are mainly interested in this overhead which is otherwise masked by adding a computational cost to the transition functions. 
-Figure \ref{fig:pdevs_plot_no_sleep} shows that in sequential simulation the PDEVS overhead is crippling.
-Conservative combined with PDEVS reduces performance to that of sequential without PDEVS with worse results for optimistic in combination in PDEVS.
-Both synchronization protocols achieve a reasonable speedup without PDEVS.
-
-\paragraph{Discussion}
-In \textit{dxex} PDEVS and the two synchronization protocols can be complementary. 
-The user is not forced to choose between either but can divide the capabilities of his machine over both approaches.
-While PDEVS can offer a significant speedup, we see that to offer this speedup the concurrency of events must be significant in combination with a high computational load in the transition functions. 
-We measure the overhead introduced by PDEVS to give practitioners insight into which model can benefit from which paradigm.
-The overhead of the synchronization protocols in specific scenarios  is shown in the previous section.
-For a given model it is difficult to predict the concurrency of events, although the computational load of a transition function can be approximated.
-In these benchmarks we show which configurations can lead to a speedup for PDEVS or the synchronization protocols in either isolation or combination.
-In a model with high concurrency in simultaneous events and a high computational load PDEVS can offer a significant speedup.
-This speedup is not exclusive to PDEVS. Conservative and optimistic synchronization protocols can achieve the same speedup given a non cyclic model. 
-From the theoretical model we know that in a cyclic model PDEVS can achieve a higher speedup.
 
@@
 	\includegraphics[width=\columnwidth]{fig/pdevs_fixed_sleep.eps}
+    %TODO fix key in figure: pdevs --> inherent parallelism
 	\caption{Queue speedup benchmark for PDEVS and synchronization with maximal concurrent events and significant computational load}
@@
 
+\paragraph{Random events}
+Now we randomize the time advance in the models, resulting in very few transition functions that happen simultaneously.
+Even when two transitions are only minimally apart in simulated time, they cannot be executed in parallel, as there might otherwise be a causality error.
+The transition function has the same computational load as in the previous configuration. 
+The results in Figure~\ref{fig:pdevs_plot_random_sleep} show that parallelizing transition functions only, adds little overhead. 
+Nonetheless, performance is not increased at all: the chances of executing a few transition functions in parallel does not make up for the overhead of checking.
+In combination with either synchronization protocol, a minor improvement can still be seen, but this is primarily caused by the synchronization protocol, and not by parallelizing the transition functions.
+
 \begin{figure}
@@
 	\includegraphics[width=\columnwidth]{fig/pdevs_random_sleep.eps}
+    %TODO fix key in figure: pdevs --> inherent parallelism
 	\caption{Queue speedup benchmark for PDEVS and synchronization with randomized events and significant computational load}
@@
 
+\paragraph{Computational load}
+Finally, we remove the computation load from the transition function, in combination with many concurrent events.
+Figure~\ref{fig:pdevs_plot_no_sleep} shows that in sequential simulation, the overhead of thread management and shared memory communication is crippling for performance.
+Even though many events occur concurrently, the actual computation being done in the transition function does not outweight the overhead, resulting in much slower simulation than sequential execution of transition functions.
+When combined with synchronization protocols, performance is still reasonable, though the use of the inherent parallelism is not beneficial either way.
+Both synchronization protocols achieve a reasonable speedup without PDEVS, as the overhead of thread management (creation and destruction) is avoided.
+These results are similar to~\cite{Himmelspach}.
+
 \begin{figure}
@@
 	\includegraphics[width=\columnwidth]{fig/pdevs_no_sleep.eps}
+    %TODO fix key in figure: pdevs --> inherent parallelism
 	\caption{Queue speedup benchmark for PDEVS and synchronization with maximal concurrent events and trivial computational load}
@@
 
+\paragraph{Discussion}
+In \textit{dxex}, use of the inherent parallelism found in \textsf{Parallel DEVS} can be combined with any of the supported synchronization protocols.
+The user is not forced to choose between either but can divide the capabilities of his machine over both approaches.
+While exploitation of the inherent parallelism can offer a significant speedup, a high number of transitions must be triggered simultaneously throughout the whole simulation.
+Additionally, the computational load of the transition functions should be high enough to warrant the thread management overhead.
+We conclude that both approaches have their distinct advantages and disadvantages.
+The inherent parallelism of \textsf{Parallel DEVS} avoids synchronization overhead and is trivial to implement.
+It is not influenced by the coupling of models, and only by the number of simultaneous transition functions and the computational load in the transition function.
+In contrast, either of the two synchronization protocols is heavily influenced by the coupling of the models, but not by the computational load or number of simultaneous transitions.
+
 \subsection{Memory Usage}
@@
 While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation infeasible.
-We therefore also investigate memory usage of different synchronization protocols, and again compare to \textit{adevs}.
+We therefore also investigate the memory usage of different synchronization protocols, and again compare to \textit{adevs}.
 
@@
 We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
-PDEVS is compared with the synchronization protocols either in isolation or complementing each other. We have shown that both approaches are useful in specific model configurations.
+The use of synchronization protocols does also not prevent the exploitation of the inherent parallelism found in \textsf{ParallelPDEVS}.
+Both approaches are useful in specific model configurations.
 But even with the right synchronization protocol, we have seen that two problems remain.

### 2016-10-03 publications/Journal/5-hotswap.tex
Small changes to section 5

@@
-Simply because a synchronization protocol is ideal at the start of the simulation, does not mean that it will still be ideal during the simulation.
+Simply because a synchronization protocol is ideal at the start of the simulation, does not mean that it stays ideal during the simulation.
 It is well known, and repeated in the previous section, that model behaviour significantly influences the ideal synchronization protocol.
@@
 Also, our interface is augmented with an interface for statistics gathering and model behaviour analysis.
-The implementation of such a component is currently left open, but we envision algorithms heavily based on machine learning.
+With all interfaces implemented and tested, we only leave open the actual switching logic.
+Machine learning is a possible direction for future implementations of this decision component.
 
 \subsection{Statistics Gathering}
-Traditionally, models are not exposed to simulation kernel details due to the wrong level of abstraction.
-Simulation models only care about being simulated, and not about how this is being done.
+Traditionally, models are not exposed to simulation kernel details as they work at a different level of abstraction:
+simulation models only care about simulation results, and not about how these results are obtained (\textit{e.g.}, through parallel or distributed simulation).
 This is different for a new simulation kernel component that has to monitor the behaviour of not only the model, but the simulator as well.
@@
 These metrics include the number of events created and destroyed, the number of inter- and intra-kernel events, the number of rollbacks, the measured lookahead, details of the GVT and EOT calculations, and information on the fairness between simulation kernels.
-With all these metrics, a component can get a fair view on both model and simulation kernel behaviour.
+With all these metrics, the decision component can get a global view on both model and simulation kernel behaviour.
 
@@
 When the number of rollbacks is excessively high, switching to conservative synchronization might be considered.
-%And when neither of these two is an option, the simulation might just (temporarily) have to fall back to sequential simulation. % Falling back to sequential is not supported atm.
 
@@
 For each connection, the number of events transferred is annotated.
-Examples are shown for the three benchmark models used before: Figure~\ref{fig:Queue_allocation}, Figure~\ref{fig:interconnect_allocation_parallel}, and Figure~\ref{fig:phold_allocation} show traces for the Queue, Interconnect, and PHOLD models respectively.
+Examples are shown for the three benchmark models used before: Figure~\ref{fig:Queue_allocation}, Figure~\ref{fig:interconnect_allocation_parallel}, and Figure~\ref{fig:phold_allocation} show traces for the Queue, Interconnect, and PHOLD models, respectively.
 Using this information, we notice that the Interconnect benchmark indeed has a lot of inter-kernel events.
@@
 \end{figure*}
+
 \begin{figure}
@@
 \end{figure}
+
 \begin{figure}

### 2016-10-03 publications/Journal/6-allocation.tex
Small changes to section 6

@@
 Depending on the model structure, and how it is mapped to the different kernels, it might not even make sense to parallelize at all.
-Indeed, if the model is distributed such that frequent communication is necessary between kernels, parallelism is naturally reduced.
+Indeed, if the model is allocated such that frequent communication is necessary between kernels, parallelism is naturally reduced.
 This brings us to the topic of model allocation, as also implemented in \textit{PythonPDEVS}~\cite{PythonPDEVS2}.
@@
 The PHOLDTree model is similar in structure to models of gossiping in social networks~\cite{Gossip}.
-The lookahead of an atomic node is the minimally allowed $\epsilon$, indicating uncertainty, as is often the case in realistic models.
+The lookahead of an atomic node is the minimally allowed $\epsilon$, as is often the case in realistic models (\textit{e.g.}, if it is unknown what the lookahead might be).
 We demonstrate the importance of allocation by comparing performance of a breadth-first versus a depth-first scheme.
@@
     \includegraphics[width=\columnwidth]{fig/pholdtree.pdf}
-    %TODO depth 1 but 3 levels deep?
-    \caption{PHOLDTree model for depth 1 and width 2.}
+    \caption{PHOLDTree model for depth 3 and width 2.}
     \label{fig:PHOLDTree_model}
@@
 \subsubsection{Sequential Simulation}
-Since \textit{adevs} does not use direct connection, we expect a noticable performance difference between \textit{dxex} and \textit{adevs}.
+Since \textit{adevs} does not use direct connection, we expect a noticeable performance difference between \textit{dxex} and \textit{adevs}.
 This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout ($n$) determines the performance penalty \textit{adevs} suffers compared to \textit{dxex}.
 Profiling indeed indicates that an increase in width per subtree ($n$) leads to higher overheads in \textit{adevs} due to the lack of direct connection.
-\textit{Dxex} uses direct connection, making it independent on fanout.
+\textit{Dxex} uses direct connection, making it independent of fanout.
 Performance of \textit{dxex} is, in this model, only dependent on the number of models.
@@
 
-With depth-first allocation, we traverse the tree in a depth-first way, allocating subsequently visited atomic models to the same node.
-This means, intuitively, that subtreees are frequently allocated to the same node.
-This allocation strategy is shown in Figure~\ref{fig:PholdTree_model_dfs}.
-
-Both allocators will try to divide models evenly over kernels.
-The effects of varying the number of models per kernel are already evaluated in the previous section on scaling.
-Here we want to highlight the overhead of communication and inter-kernel dependence.
-
 \begin{figure}
@@
 \end{figure}
+
+With depth-first allocation, we traverse the tree in a depth-first way, allocating subsequently visited atomic models to the same node.
+This means, intuitively, that subtreees are frequently allocated to the same node.
+This allocation strategy is shown in Figure~\ref{fig:PholdTree_model_dfs}.
+
 \begin{figure}
@@
 \end{figure}
+
+Both allocators will try to divide models evenly over kernels.
+The effects of varying the number of models per kernel are already evaluated in the previous section on scaling.
+Here we want to highlight the overhead of communication and inter-kernel dependence.
 
@@
     \includegraphics[width=\columnwidth]{fig/pholdtreed1n3t5000c4BFS.eps}
-    \caption{Visualization of a PHOLDTree simulation with breadth first allocation and 4 kernels.}
+    \caption{Visualization of a simulation of the model in Figure~\ref{fig:PholdTree_model_bfs}.}
     \label{fig:pholdtree_visualize_parBFS}
 \end{figure}
+
 \begin{figure}
@@
     \includegraphics[width=\columnwidth]{fig/pholdtreed1n3t5000c4DFS.eps}
-    \caption{Visualization of a PHOLDTree simulation with depth first allocation and 4 kernels.}
+    \caption{Visualization of a simulation of the model in Figure~\ref{fig:PholdTree_model_dfs}.}
     \label{fig:pholdtree_visualize_parDFS}
@@
 Interestingly, we see that optimistic synchronization is less influenced by the allocation than conservative synchronization.
-This is likely caused due to the lower number of connections to take into account in conservative synchronization.
+This is likely caused by the lower number of connections to take into account in conservative synchronization.
 Whereas conservative synchronization needs to take into account even scarcely used connections, optimistic synchronization does not.

### 2016-10-03 publications/Journal/7-related-work.tex
Further review of style, acronyms.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
Small changes to section 7

@@
 Although \textit{PythonPDEVS} offers very fast turnaround cycles, simulation performance was easily outdone by compiled simulation kernels.
-In terms of performance, \textit{adevs}~\cite{adevs} offered much faster simulation, at the cost of a significant compilation time.
-The turnaround cycle in \textit{adevs} is much slower though, specifically because the complete simulation kernel is implemented using templates in header files.
+In terms of performance, \textit{adevs}~\cite{adevs} offered much faster simulation, at the cost of compilation time.
+While this is beneficial for long running simulations, small simulations are therefore negatively impacted.
+The turnaround cycle in \textit{adevs} is much slower, specifically because the complete simulation kernel is implemented using templates in header files.
 As a result, the complete simulation kernel has to be compiled again every time.
@@
 In terms of performance, \textit{dxex} is shown to be competitive with \textit{adevs}.
-Despite its high performance, \textit{adevs} does not support optimistic synchronization, which we have shown to be highly relevant.
+Despite its high performance, \textit{adevs} does not support optimistic synchronization, which we have shown to be highly relevant for certain kinds of models.
 
-Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though none have done it in a strictly modular way that allows straightforward protocol switching for a single given model.
-For example, \textit{adevs} only supports conservative synchronization, and \textit{vle} only supports experiment-level parallelism (\textit{i.e.}, run multiple experiments concurrently).
+Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though none have done so in a strictly modular way that allows straightforward protocol switching for a single given model.
+For example, \textit{adevs} only supports conservative synchronization, and \textit{vle} only supports experiment-level parallelism (\textit{i.e.}, run multiple experiments in parallel).
 Closest to our support for multiple synchronization protocols is \textit{CD++}~\cite{CD++}.
-For \textit{CD++}, both a conservative (\textit{CCD++}~\cite{CCD++}) and optimistic (\textit{PCD++})~\cite{PCD++}) variant exist.
-Despite the implementation of both protocols, they are different projects entirely.
+For \textit{CD++}, both a conservative (\textit{CCD++}~\cite{CCD++}) and optimistic (\textit{PCD++}~\cite{PCD++}) variant exist.
+Despite the implementation of both protocols, they are entirely different projects.
 Some features might therefore be implemented in \textit{CCD++} and not in \textit{PCD++}, or vice versa.
-And while this might not yet be that significant a problem to this day, this problem will only get worse when each project follows its own course.
-\textit{Dxex}, on the other hand, is a single project, where the choice of synchronization protocol is a simple configuration.
+And while this might not be a problem at this time, the problem will only get worse when each project follows its own course.
+\textit{Dxex}, on the other hand, is a single project, where the choice of synchronization protocol is a simple configuration option.
 \textit{CD++}, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourselves to parallel simulation. 
@@
 By limiting our approach to a pure parallel simulation, we are able to achieve higher speedups through the use of shared memory communication. 
+Recent work on parallel speedup in DEVS investigates theoretical limits of PDEVS~\cite{activitypdevs, amdahlpdevs}
 
 In the \textsf{PDES} community, the problem of choosing between synchronization protocols is well known and documented~\cite{Jha:1994:UFC:195291.182480}.
-The challenges of implementing such runtime switching have previously been explored already~\cite{Das:1996:APP:256562.256602}, and an implementation was given by, for example~\cite{perumalla2005musik}.
+The challenges of implementing such runtime switching have previously been explored already~\cite{Das:1996:APP:256562.256602}, and implemented~\cite{perumalla2005musik}.
 Our contribution entails bringing this same feature to the \textsf{Parallel DEVS} community, further expanding upon our support for multiple synchronization protocols.
@@
 Model allocation and its impact on parallel performance has previously also been studied in the \textsf{PDES} community~\cite{PDESpartitioning}.
-Referenced as partitioning of the simulation model, most studies distinguish between communication overhead and computational distribution (load balancing) as the two dimensions to partition over.
+Referenced as partitioning of the simulation model, most studies distinguish between communication and computation as the two dimensions to partition over.
 Partitioning a simulation model is identified as an issue to achieve scalability~\cite{Scalability}. 
-Some research in the context of \textsf{Parallel DEVS} has been done, where they turn their attention to load balancing and communication overhead~\cite{PDEVSpartitioning, NonFragmenting}.
+Some research in this context has been done for \textsf{Parallel DEVS}~\cite{PDEVSpartitioning, NonFragmenting}.
 Our contribution studies the effect of partitioning with emphasis on the effect of communication between processes and in the presence of a flattened hierarchy. 
-We focus on static partitioning since this is a limiting factor for our conservative synchronization implementation which does not support model relocation.
-Model relocation, as implemented by \textit{PythonPDEVS}~\cite{PythonPDEVS2}, might be an interesting addition to only model allocation at the start of simulation.
+We focus on static partitioning since this is a limiting factor for our conservative synchronization implementation which does not support model migration.
+Model migration, as implemented by \textit{PythonPDEVS}, might be an interesting addition to model allocation.
 

### 2016-10-05 publications/Journal/reply_to_reviewers.txt
Use plain text reply instead of odt

Referees' comments:

Referee: 1

Recommendation: Accept If Certain Minor Revisions Are Made

Comments:
The authors propose a new architecture for a PDEVS simulator. This simulator allows to dynamically change the simulation synchronization protocols for single machine parallelism.

An implementation in C++11 is described implementing sequential, conservative and optimistic protocols. The tool implemented is evaluated using synthetic models based in DEVStone and PHOLD. Based on the results obtained, the authors discuss the challenges of memory allocation and core association. These challenges are different for different protocols, and motivate the switching approach.

Notes:
- Expand first use of Acronyms.
We have reviewed the text to expand acronyms where this was not the case.

-Page 7 (5 in the paper), lines 46-54. Authors mention, the inconvenience for modelers to be required to define lookahead in every model. An alternative is using the inheritance mechanisms of C++.
Model could be defined with no look-ahead awareness as in classic simulators. Then, a "model-with-lookahead" can be defined as specialization of model. Then, a model can be specialized with look-ahead information when simulated in conservative protocol. That way, both can be passed to the "SequentialCore", but classic model will not be accepted by the "ConservativeCore" if passed.
The current default implementation of the lookahead function in the base model class has no performance impact in sequential or optimistic simulation, the compiler can deduce that this function is never invoked.
By providing it depending on the simulation type (by way of specialization) we would break one of our design aims, namely reusing the same model across the different synchronization protocols. This would further complicate the switching between optimistic and conservative at runtime, as it would involve a casting operation which may or may not fail at runtime.
Furthermore, a model designed for sequential simulation can be run in conservative simulation in the current design without providing a lookahead. Using our statistics module the author can then use the gathered information to set the actual lookahead.
We have modified the relevant paragraph to clarify this point.

 - Page 8 (6 in the paper) lines 12 to 17, second column. Have you considered the use of transactional memory?
We have considered this, but unfortunately transactional memory is still not in the C++ standard. In our design we opted for maximum portability and using a non standard implementation would hinder this. There is ongoing work to include transactional memory in the C++17 standard but the current prototype implementation in GCC is too experimental to use, especially since the implementation uses a single lock which would constrain performance. We have modified our future work section to add this information.

 - Figure 6 is explained before Figure 5
Corrected

- Figure 7, line dxex w 800 is invisible in B&W, line adevs w800 is hard to see in B&W
- Figures 8, 9, 10, 11, 12, 13, 14, 20, and 25 are plotted in different style than previous ones, also most lines cannot be distinguished, or are hard to read in B&W.
- Figure 21 and 22 are hard to identify the light colours differences in B&W.
Corrected.
The graphs have been modified to use better distinguishable colors and line styles. The use of color has also been made more consistent.

- Page 10 (8 in paper), line 41.  "An example for 4 models is shown in Figure 5", the Figure showed has 3 models and the interconnection is not HI.
Corrected.
Note that the model is indeed not HI, as the text mentions it is a merge between PHOLD and HI.

- Page 17 (15 in paper), have you tried to assign different quentities of models per core? All examples seem to use equal quantity per core (with exception of the last core, taking less in some cases).
The scaling effects on performance have been investigated in the performance section. In this scenario (bfs vs dfs allocation) we specifically wanted to balance the model load over kernels. This will not always result in an equal load as is indeed shown in the allocation plot, but the main effect we want to demonstrate here is the effect of inter kernel communication and resulting dependencies between kernels this introduces. We have added a paragraph to section 6 to explain our reasoning.

- Page 18 (16 in paper), lines 16 a 33, second column. CD++ has an effort of merging multiple approaches in a single simulator. However, the proposed approach is static, once the simulation starts, there is no dynamic protocol change. The new architecture for this approach is showed in [1], and a initial implementation in C++11 is evaluated.
We’ve updated the related work section :
“A new architecture for sequential PDEVS simulation has been introduced in \cite{SeqPDEVSArch} with promising performance results. This algorithm achieves a speedup by using the inherent parallelism in PDEVS without the need for synchronization.”

- In conclusions, it is mentioned that the main contribution is the implementation of the simulator. I believe the discussion over protocol switching, memory and core assignation is the core is more important.
We have included memory allocation result in the conclusion. The other topics are already mentioned in the conclusion.

Additional Questions:
1. Are the application of modeling and simulation methods clearly motivated?: Yes

If no, please identify the points where the motivation lacks clarity.:

2. How would you rate the impact of the results achieved for the application area?: Noticeable

Could you please shortly explain your opinion?:

3. Is the paper logically and technically correct?: Appears to be, but didn't check completely.

If no, could you please describe the incorrect parts of the paper and make suggestions how these parts can be corrected.:

4. How would you describe the technical depth of the paper?: Suitable only for an expert

5. Is the coverage of the topic sufficiently comprehensive and balanced?: Yes

If no, please identify which parts of the topic are missing or superficially treated:

If certain parts are greatly overstressed, please identify which ones:

1. Are the title and abstract appropriate?: Yes

If no, could you please describe why the title and / or abstract appear to be misleading?:

2. Does the introduction clearly state the background and / or motivation in terms understandable to the non-specialist?: Yes

If no, could you please identify where the introduction should be clarified?:

3. Are the references or bibliography appropriate and complete?: No

If no, please provide references or research areas with which the present work should be included and compared with.: There are recent work on similar algorithms by the group of Lin Uhrmacher at the University of Rostock, and papers by D. Vicino published in the Symposium on Theory of Modeling & Simulation that the authors should revise, discuss and cite in their related work section.
We did not find relevant work from D Vicino on partitioning, but updated the related work section with a review of a paper from Lin Uhrmacher regarding partitioning.

4. How would you rate the overall organization of the paper?: Satisfactory

If it can be improved, how?:

5. Is the English satisfactory?: No, but can be easily polished.
We have made several style changes to the text.

Referee: 2

Recommendation: Accept If Certain Minor Revisions Are Made

Comments:
(There are no comments.)

Additional Questions:
1. Are the application of modeling and simulation methods clearly motivated?: Yes

If no, please identify the points where the motivation lacks clarity.:

2. How would you rate the impact of the results achieved for the application area?: Very High

Could you please shortly explain your opinion?: The paper presents important results concerning the performance of parallel simulation synchronization protocols.

3. Is the paper logically and technically correct?: Yes

If no, could you please describe the incorrect parts of the paper and make suggestions how these parts can be corrected.:

4. How would you describe the technical depth of the paper?: Appropriate for a worker in the general field of the paper

5. Is the coverage of the topic sufficiently comprehensive and balanced?: Yes

If no, please identify which parts of the topic are missing or superficially treated:

If certain parts are greatly overstressed, please identify which ones:

1. Are the title and abstract appropriate?: Yes

If no, could you please describe why the title and / or abstract appear to be misleading?:

2. Does the introduction clearly state the background and / or motivation in terms understandable to the non-specialist?: Yes

If no, could you please identify where the introduction should be clarified?:

3. Are the references or bibliography appropriate and complete?: No

If no, please provide references or research areas with which the present work should be included and compared with.: I think that two recent  references concerning PDEVS are missing :
Zeigler, B.P., Nutaro J.J., Seo, C., What’s the Best Possible Speedup Achievable in Distributed Simulation: Amdahl’s Law Reconstructed, DEVS TMS, SpringSim (2015), 930-937.

PDEVS Protocol Performance Prediction using Activity Patterns with Finite Probabilistic DEVS (Bernard P. Zeigler, Laurent Capocchi and Jean Francois Santucci) in SpringSim 2016.
A new subsection was added with extended results and discussion regarding these two papers, the related work section was updated with these references.

Furthermore I think that it is worthwhile to remove the reference number 22 (the article has been published in a very bad workshop without any review and paper is not pertinent).
Thank you for notifying us about the quality of this reference. We have consequently removed it from our list of references.

In addition please verify the references (at least  in 15  references the year of publication is missing).
The missing years are due to the Latex style file. Some @inproceedings don’t include the year in their title. We have added the year field to every reference, so it isn’t our fault. We changed the items to @articles to work around this issue.

4. How would you rate the overall organization of the paper?: Satisfactory

If it can be improved, how?:

5. Is the English satisfactory?: Yes

Referee: 3

Recommendation: Author(s) Should Prepare A Major Revision For A Second Review

Comments:
(There are no comments.)

Additional Questions:
1. Are the application of modeling and simulation methods clearly motivated?: Yes

If no, please identify the points where the motivation lacks clarity.:

2. How would you rate the impact of the results achieved for the application area?: Significant

Could you please shortly explain your opinion?:

3. Is the paper logically and technically correct?: No

If no, could you please describe the incorrect parts of the paper and make suggestions how these parts can be corrected.: The paper doesn't consider including the standard DEVS simulation protocol as one of the alternatives supported by the software
The authors completely ignore recent results on the standard parallel DEVS simulation protocol that shows that on the average this protocal is not far from the
time of the best possible method conservative or optimistic  method. Moreover, the DEVS protocol converges to the  best possible speedup for models with moderate to high  coupling.  See the reference given later.

Three aspects need to be addressed:
The test models need to be extended with those that incorporate intrinsic parallelism.
	We have benchmarked 3 configurations that demonstrate the different aspects highlighted in the
	theoretical work and analyzed their results.
The standard parallel DEVS simulation protocol should be implemented as an alternative and compare with the others.
	We have implemented this. PDEVS can operate with or without the synchronization protocols allowing for a in depth comparison.
The results of the standard, conservative and optimistic methods need to be compared with the theoretical predictions
	In a new subsection we analyze the performance differences between PDEVS, standard, optimistic and conservative.

4. How would you describe the technical depth of the paper?: Appropriate for a worker in the general field of the paper

5. Is the coverage of the topic sufficiently comprehensive and balanced?: No

If no, please identify which parts of the topic are missing or superficially treated: Missing is discussion on the Parallel DEVS Simulation Protocal as mentioned in other sections of this review.
	See item 3.

If certain parts are greatly overstressed, please identify which ones:

1. Are the title and abstract appropriate?: No

If no, could you please describe why the title and / or abstract appear to be misleading?: The title is not appropriate for the limited perspective adopted as the paper stands.  If it were to be extended to consider the standard parallel DEVS simulation protocol the title would be warranted.
We have implemented the requested changes.

2. Does the introduction clearly state the background and / or motivation in terms understandable to the non-specialist?: Yes

If no, could you please identify where the introduction should be clarified?:

3. Are the references or bibliography appropriate and complete?: No

If no, please provide references or research areas with which the present work should be included and compared with.: Bernard P. Zeigler James J. Nutaro, and C. Seo. What’s the Best Possible Speedup Achievable in Distributed Simulation: Amdahl’s Law Reconstructed, DEVS TMS, SpringSim 2015.
This reference was included and discussed with the empirical results.

4. How would you rate the overall organization of the paper?: Could be improved

If it can be improved, how?: Often terms are used before they are defined or linked to references later. Looks like sections of the paper were re-arranged without checking for dependencies.
We have revised the paper and made changes where this issue was apparent.

5. Is the English satisfactory?: Yes

### 2016-10-05 publications/Journal/reply_to_reviewers.txt
Made the plain text reply somewhat readable

@@
-Referees' comments:
-
-Referee: 1
-
-Recommendation: Accept If Certain Minor Revisions Are Made
-
-Comments:
-The authors propose a new architecture for a PDEVS simulator. This simulator allows to dynamically change the simulation synchronization protocols for single machine parallelism.
-
-An implementation in C++11 is described implementing sequential, conservative and optimistic protocols.
+> Referees' comments:
+> 
+> Referee: 1
+> 
+> Recommendation: Accept If Certain Minor Revisions Are Made
+> 
+> Comments:
+> The authors propose a new architecture for a PDEVS simulator. This simulator allows to dynamically change the simulation synchronization protocols for single machine parallelism.
+> An implementation in C++11 is described implementing sequential, conservative and optimistic protocols.
 The tool implemented is evaluated using synthetic models based in DEVStone and PHOLD.
@@
 These challenges are different for different protocols, and motivate the switching approach.
+> 
+> Notes:
+> - Expand first use of Acronyms.
 
-Notes:
-- Expand first use of Acronyms.
 We have reviewed the text to expand acronyms where this was not the case.
 
--Page 7 (5 in the paper), lines 46-54. Authors mention, the inconvenience for modelers to be required to define lookahead in every model. An alternative is using the inheritance mechanisms of C++.
-Model could be defined with no look-ahead awareness as in classic simulators.
+> -Page 7 (5 in the paper), lines 46-54. Authors mention, the inconvenience for modelers to be required to define lookahead in every model. An alternative is using the inheritance mechanisms of C++.
+> Model could be defined with no look-ahead awareness as in classic simulators.
 Then, a "model-with-lookahead" can be defined as specialization of model.
@@
 That way, both can be passed to the "SequentialCore", but classic model will not be accepted by the "ConservativeCore" if passed.
+
 The current default implementation of the lookahead function in the base model class has no performance impact in sequential or optimistic simulation, the compiler can deduce that this function is never invoked.
@@
 
- - Page 8 (6 in the paper) lines 12 to 17, second column. Have you considered the use of transactional memory?
+>  - Page 8 (6 in the paper) lines 12 to 17, second column. Have you considered the use of transactional memory?
+
 We have considered this, but unfortunately transactional memory is still not in the C++ standard.
@@
 
- - Figure 6 is explained before Figure 5
+>  - Figure 6 is explained before Figure 5
+
 Corrected
 
-- Figure 7, line dxex w 800 is invisible in B&W, line adevs w800 is hard to see in B&W
-- Figures 8, 9, 10, 11, 12, 13, 14, 20, and 25 are plotted in different style than previous ones, also most lines cannot be distinguished, or are hard to read in B&W.
-- Figure 21 and 22 are hard to identify the light colours differences in B&W.
-Corrected.
+> - Figure 7, line dxex w 800 is invisible in B&W, line adevs w800 is hard to see in B&W
+> - Figures 8, 9, 10, 11, 12, 13, 14, 20, and 25 are plotted in different style than previous ones, also most lines cannot be distinguished, or are hard to read in B&W.
+> - Figure 21 and 22 are hard to identify the light colours differences in B&W.
+
 The graphs have been modified to use better distinguishable colors and line styles. The use of color has also been made more consistent.
 
-- Page 10 (8 in paper), line 41.  "An example for 4 models is shown in Figure 5", the Figure showed has 3 models and the interconnection is not HI.
-Corrected.
+> - Page 10 (8 in paper), line 41.  "An example for 4 models is shown in Figure 5", the Figure showed has 3 models and the interconnection is not HI.
+
 Note that the model is indeed not HI, as the text mentions it is a merge between PHOLD and HI.
 
-- Page 17 (15 in paper), have you tried to assign different quentities of models per core?
+> - Page 17 (15 in paper), have you tried to assign different quentities of models per core?
 All examples seem to use equal quantity per core (with exception of the last core, taking less in some cases).
+
 The scaling effects on performance have been investigated in the performance section.
@@
 
-- Page 18 (16 in paper), lines 16 a 33, second column.
+> - Page 18 (16 in paper), lines 16 a 33, second column.
 CD++ has an effort of merging multiple approaches in a single simulator.
@@
 The new architecture for this approach is showed in [1], and a initial implementation in C++11 is evaluated.
-We’ve updated the related work section :
-“A new architecture for sequential PDEVS simulation has been introduced in \cite{SeqPDEVSArch} with promising performance results.
-This algorithm achieves a speedup by using the inherent parallelism in PDEVS without the need for synchronization.”
 
-- In conclusions, it is mentioned that the main contribution is the implementation of the simulator.
+We have updated the related work section.
+
+> - In conclusions, it is mentioned that the main contribution is the implementation of the simulator.
 I believe the discussion over protocol switching, memory and core assignation is the core is more important.
+
 We have included memory allocation result in the conclusion. The other topics are already mentioned in the conclusion.
 
-Additional Questions:
-1. Are the application of modeling and simulation methods clearly motivated?: Yes
+> Additional Questions:
+> 1. Are the application of modeling and simulation methods clearly motivated?: Yes
+> 
+> If no, please identify the points where the motivation lacks clarity.:
+> 
+> 2. How would you rate the impact of the results achieved for the application area?: Noticeable
+> 
+> Could you please shortly explain your opinion?:
+> 
+> 3. Is the paper logically and technically correct?: Appears to be, but didn't check completely.
+> 
+> If no, could you please describe the incorrect parts of the paper and make suggestions how these parts can be corrected.:
+> 
+> 4. How would you describe the technical depth of the paper?: Suitable only for an expert
+> 
+> 5. Is the coverage of the topic sufficiently comprehensive and balanced?: Yes
+> 
+> If no, please identify which parts of the topic are missing or superficially treated:
+> 
+> If certain parts are greatly overstressed, please identify which ones:
+> 
+> 1. Are the title and abstract appropriate?: Yes
+> 
+> If no, could you please describe why the title and / or abstract appear to be misleading?:
+> 
+> 2. Does the introduction clearly state the background and / or motivation in terms understandable to the non-specialist?: Yes
+> 
+> If no, could you please identify where the introduction should be clarified?:
+> 
+> 3. Are the references or bibliography appropriate and complete?: No
+> 
+> If no, please provide references or research areas with which the present work should be included and compared with.: There are recent work on similar algorithms by the group of Lin Uhrmacher at the University of Rostock, and papers by D.
+Vicino published in the Symposium on Theory of Modeling & Simulation that the authors should revise, discuss and cite in their related work section.
 
-If no, please identify the points where the motivation lacks clarity.:
-
-2. How would you rate the impact of the results achieved for the application area?: Noticeable
-
-Could you please shortly explain your opinion?:
-
-3. Is the paper logically and technically correct?: Appears to be, but didn't check completely.
-
-If no, could you please describe the incorrect parts of the paper and make suggestions how these parts can be corrected.:
-
-4. How would you describe the technical depth of the paper?: Suitable only for an expert
-
-5. Is the coverage of the topic sufficiently comprehensive and balanced?: Yes
-
-If no, please identify which parts of the topic are missing or superficially treated:
-
-If certain parts are greatly overstressed, please identify which ones:
-
-1. Are the title and abstract appropriate?: Yes
-
-If no, could you please describe why the title and / or abstract appear to be misleading?:
-
-2. Does the introduction clearly state the background and / or motivation in terms understandable to the non-specialist?: Yes
-
-If no, could you please identify where the introduction should be clarified?:
-
-3. Are the references or bibliography appropriate and complete?: No
-
-If no, please provide references or research areas with which the present work should be included and compared with.: There are recent work on similar algorithms by the group of Lin Uhrmacher at the University of Rostock, and papers by D.
-Vicino published in the Symposium on Theory of Modeling & Simulation that the authors should revise, discuss and cite in their related work section.
 We did not find relevant work from D Vicino on partitioning, but updated the related work section with a review of a paper from Lin Uhrmacher regarding partitioning.
 
-4. How would you rate the overall organization of the paper?: Satisfactory
+> 4. How would you rate the overall organization of the paper?: Satisfactory
+> 
+> If it can be improved, how?:
+> 
+> 5. Is the English satisfactory?: No, but can be easily polished.
 
-If it can be improved, how?:
-
-5. Is the English satisfactory?: No, but can be easily polished.
 We have made several style changes to the text.
 
-Referee: 2
+> Referee: 2
+> 
+> Recommendation: Accept If Certain Minor Revisions Are Made
+> 
+> Comments:
+> (There are no comments.)
+> 
+> Additional Questions:
+> 1. Are the application of modeling and simulation methods clearly motivated?: Yes
+> 
+> If no, please identify the points where the motivation lacks clarity.:
+> 
+> 2. How would you rate the impact of the results achieved for the application area?: Very High
+> 
+> Could you please shortly explain your opinion?: The paper presents important results concerning the performance of parallel simulation synchronization protocols.
+> 
+> 3. Is the paper logically and technically correct?: Yes
+> 
+> If no, could you please describe the incorrect parts of the paper and make suggestions how these parts can be corrected.:
+> 
+> 4. How would you describe the technical depth of the paper?: Appropriate for a worker in the general field of the paper
+> 
+> 5. Is the coverage of the topic sufficiently comprehensive and balanced?: Yes
+> 
+> If no, please identify which parts of the topic are missing or superficially treated:
+> 
+> If certain parts are greatly overstressed, please identify which ones:
+> 
+> 1. Are the title and abstract appropriate?: Yes
+> 
+> If no, could you please describe why the title and / or abstract appear to be misleading?:
+> 
+> 2. Does the introduction clearly state the background and / or motivation in terms understandable to the non-specialist?: Yes
+> 
+> If no, could you please identify where the introduction should be clarified?:
+> 
+> 3. Are the references or bibliography appropriate and complete?: No
+> 
+> If no, please provide references or research areas with which the present work should be included and compared with.: I think that two recent references concerning PDEVS are missing :
+> Zeigler, B.P., Nutaro J.J., Seo, C., What's the Best Possible Speedup Achievable in Distributed Simulation: Amdahl's Law Reconstructed, DEVS TMS, SpringSim (2015), 930-937.
+> PDEVS Protocol Performance Prediction using Activity Patterns with Finite Probabilistic DEVS (Bernard P. Zeigler, Laurent Capocchi and Jean Francois Santucci) in SpringSim 2016.
 
-Recommendation: Accept If Certain Minor Revisions Are Made
-
-Comments:
-(There are no comments.)
-
-Additional Questions:
-1. Are the application of modeling and simulation methods clearly motivated?: Yes
-
-If no, please identify the points where the motivation lacks clarity.:
-
-2. How would you rate the impact of the results achieved for the application area?: Very High
-
-Could you please shortly explain your opinion?: The paper presents important results concerning the performance of parallel simulation synchronization protocols.
-
-3. Is the paper logically and technically correct?: Yes
-
-If no, could you please describe the incorrect parts of the paper and make suggestions how these parts can be corrected.:
-
-4. How would you describe the technical depth of the paper?: Appropriate for a worker in the general field of the paper
-
-5. Is the coverage of the topic sufficiently comprehensive and balanced?: Yes
-
-If no, please identify which parts of the topic are missing or superficially treated:
-
-If certain parts are greatly overstressed, please identify which ones:
-
-1. Are the title and abstract appropriate?: Yes
-
-If no, could you please describe why the title and / or abstract appear to be misleading?:
-
-2. Does the introduction clearly state the background and / or motivation in terms understandable to the non-specialist?: Yes
-
-If no, could you please identify where the introduction should be clarified?:
-
-3. Are the references or bibliography appropriate and complete?: No
-
-If no, please provide references or research areas with which the present work should be included and compared with.: I think that two recent  references concerning PDEVS are missing :
-Zeigler, B.P., Nutaro J.J., Seo, C., What’s the Best Possible Speedup Achievable in Distributed Simulation: Amdahl’s Law Reconstructed, DEVS TMS, SpringSim (2015), 930-937.
-
-PDEVS Protocol Performance Prediction using Activity Patterns with Finite Probabilistic DEVS (Bernard P. Zeigler, Laurent Capocchi and Jean Francois Santucci) in SpringSim 2016.
 A new subsection was added with extended results and discussion regarding these two papers, the related work section was updated with these references.
 
-Furthermore I think that it is worthwhile to remove the reference number 22 (the article has been published in a very bad workshop without any review and paper is not pertinent).
+> Furthermore I think that it is worthwhile to remove the reference number 22 (the article has been published in a very bad workshop without any review and paper is not pertinent).
+
 Thank you for notifying us about the quality of this reference. We have consequently removed it from our list of references.
 
-In addition please verify the references (at least  in 15  references the year of publication is missing).
+> In addition please verify the references (at least  in 15  references the year of publication is missing).
+
 The missing years are due to the Latex style file.
-Some @inproceedings don’t include the year in their title.
+Some @inproceedings don't include the year in their title.
 We have added the year field to every reference, so it isn’t our fault.
@@
 
-4. How would you rate the overall organization of the paper?: Satisfactory
-
-If it can be improved, how?:
-
-5. Is the English satisfactory?: Yes
-
-Referee: 3
-
-Recommendation: Author(s) Should Prepare A Major Revision For A Second Review
-
-Comments:
-(There are no comments.)
-
-Additional Questions:
-1. Are the application of modeling and simulation methods clearly motivated?: Yes
-
-If no, please identify the points where the motivation lacks clarity.:
-
-2. How would you rate the impact of the results achieved for the application area?: Significant
-
-Could you please shortly explain your opinion?:
-
-3. Is the paper logically and technically correct?: No
-
-If no, could you please describe the incorrect parts of the paper and make suggestions how these parts can be corrected.: The paper doesn't consider including the standard DEVS simulation protocol as one of the alternatives supported by the software
-The authors completely ignore recent results on the standard parallel DEVS simulation protocol that shows that on the average this protocal is not far from the
-time of the best possible method conservative or optimistic  method.
+> 4. How would you rate the overall organization of the paper?: Satisfactory
+> 
+> If it can be improved, how?:
+> 
+> 5. Is the English satisfactory?: Yes
+> 
+> Referee: 3
+> 
+> Recommendation: Author(s) Should Prepare A Major Revision For A Second Review
+> 
+> Comments:
+> (There are no comments.)
+> 
+> Additional Questions:
+> 1. Are the application of modeling and simulation methods clearly motivated?: Yes
+> 
+> If no, please identify the points where the motivation lacks clarity.:
+> 
+> 2. How would you rate the impact of the results achieved for the application area?: Significant
+> 
+> Could you please shortly explain your opinion?:
+> 
+> 3. Is the paper logically and technically correct?: No
+> 
+> If no, could you please describe the incorrect parts of the paper and make suggestions how these parts can be corrected.: The paper doesn't consider including the standard DEVS simulation protocol as one of the alternatives supported by the software
+> The authors completely ignore recent results on the standard parallel DEVS simulation protocol that shows that on the average this protocal is not far from the
+> time of the best possible method conservative or optimistic  method.
 Moreover, the DEVS protocol converges to the  best possible speedup for models with moderate to high  coupling.
 See the reference given later.
+> 
+> Three aspects need to be addressed:
+> The test models need to be extended with those that incorporate intrinsic parallelism.
 
-Three aspects need to be addressed:
-The test models need to be extended with those that incorporate intrinsic parallelism.
-	We have benchmarked 3 configurations that demonstrate the different aspects highlighted in the
-	theoretical work and analyzed their results.
-The standard parallel DEVS simulation protocol should be implemented as an alternative and compare with the others.
+We have benchmarked 3 configurations that demonstrate the different aspects highlighted in the theoretical work and analyzed their results.
+
+> The standard parallel DEVS simulation protocol should be implemented as an alternative and compare with the others.
+
 We have implemented this. PDEVS can operate with or without the synchronization protocols allowing for a in depth comparison.
-The results of the standard, conservative and optimistic methods need to be compared with the theoretical predictions
+
+> The results of the standard, conservative and optimistic methods need to be compared with the theoretical predictions
+
 In a new subsection we analyze the performance differences between PDEVS, standard, optimistic and conservative.
 
-4. How would you describe the technical depth of the paper?: Appropriate for a worker in the general field of the paper
+> 4. How would you describe the technical depth of the paper?: Appropriate for a worker in the general field of the paper
+> 
+> 5. Is the coverage of the topic sufficiently comprehensive and balanced?: No
+> 
+> If no, please identify which parts of the topic are missing or superficially treated: Missing is discussion on the Parallel DEVS Simulation Protocal as mentioned in other sections of this review.
 
-5. Is the coverage of the topic sufficiently comprehensive and balanced?: No
-
-If no, please identify which parts of the topic are missing or superficially treated: Missing is discussion on the Parallel DEVS Simulation Protocal as mentioned in other sections of this review.
 See item 3.
 
-If certain parts are greatly overstressed, please identify which ones:
+> If certain parts are greatly overstressed, please identify which ones:
+> 
+> 1. Are the title and abstract appropriate?: No
+> 
+> If no, could you please describe why the title and / or abstract appear to be misleading?: The title is not appropriate for the limited perspective adopted as the paper stands.
+If it were to be extended to consider the standard parallel DEVS simulation protocol the title would be warranted.
 
-1. Are the title and abstract appropriate?: No
-
-If no, could you please describe why the title and / or abstract appear to be misleading?: The title is not appropriate for the limited perspective adopted as the paper stands.
-If it were to be extended to consider the standard parallel DEVS simulation protocol the title would be warranted.
 We have implemented the requested changes.
 
-2. Does the introduction clearly state the background and / or motivation in terms understandable to the non-specialist?: Yes
-
-If no, could you please identify where the introduction should be clarified?:
-
-3. Are the references or bibliography appropriate and complete?: No
-
-If no, please provide references or research areas with which the present work should be included and compared with.: Bernard P.
+> 2. Does the introduction clearly state the background and / or motivation in terms understandable to the non-specialist?: Yes
+> 
+> If no, could you please identify where the introduction should be clarified?:
+> 
+> 3. Are the references or bibliography appropriate and complete?: No
+> 
+> If no, please provide references or research areas with which the present work should be included and compared with.: Bernard P.
 Zeigler James J.
@@
 Seo.
-What’s the Best Possible Speedup Achievable in Distributed Simulation: Amdahl’s Law Reconstructed, DEVS TMS, SpringSim 2015.
+What's the Best Possible Speedup Achievable in Distributed Simulation: Amdahl's Law Reconstructed, DEVS TMS, SpringSim 2015.
+
 This reference was included and discussed with the empirical results.
 
-4. How would you rate the overall organization of the paper?: Could be improved
+> 4. How would you rate the overall organization of the paper?: Could be improved
+> 
+> If it can be improved, how?: Often terms are used before they are defined or linked to references later. Looks like sections of the paper were re-arranged without checking for dependencies.
 
-If it can be improved, how?: Often terms are used before they are defined or linked to references later. Looks like sections of the paper were re-arranged without checking for dependencies.
 We have revised the paper and made changes where this issue was apparent.
 
-5. Is the English satisfactory?: Yes
+> 5. Is the English satisfactory?: Yes

### 2016-10-05 publications/Journal/reply_to_reviewers.txt
Elaborated some replies

@@
 
-The current default implementation of the lookahead function in the base model class has no performance impact in sequential or optimistic simulation, the compiler can deduce that this function is never invoked.
-By providing it depending on the simulation type (by way of specialization) we would break one of our design aims, namely reusing the same model across the different synchronization protocols.
-This would further complicate the switching between optimistic and conservative at runtime, as it would involve a casting operation which may or may not fail at runtime.
-Furthermore, a model designed for sequential simulation can be run in conservative simulation in the current design without providing a lookahead.
-Using our statistics module the author can then use the gathered information to set the actual lookahead.
-We have modified the relevant paragraph to clarify this point.
+The proposed solution would indeed lift out the lookahead for sequential and
+optimistic models. This is, however, in conflict with our intended design:
+have a single model that can be simulated using all possible synchronization
+protocols. While it is indeed true that sequential and optimistic models will
+also have to define a lookahead, this is not too much of a burden to the user:
+a default lookahead (of epsilon) is provided automatically. If the compiler
+sees that this value is unused, it can even be optimized away. Additionally,
+this would complicate runtime switching to conservative, as some models might
+not be simulatable this way.
+
+We have modified the relevant paragraph to include this motivation as to why
+we don't use inheritance.
 
@@
 
-We have considered this, but unfortunately transactional memory is still not in the C++ standard.
-In our design we opted for maximum portability and using a non standard implementation would hinder this.
-There is ongoing work to include transactional memory in the C++17 standard but the current prototype implementation in GCC is too experimental to use, especially since the implementation uses a single lock which would constrain performance.
-We have modified our future work section to add this information.
+We agree that transactional memory could be beneficial to our simulator.
+Unfortunately, transactional memory is still not in the C++ standard. As we
+try to strive for maximum portability, using non-standard implementations goes
+against our design philosophy. When included in the new C++ standard, it would
+certainly be of great interest to look at the influence of transactional
+memory. We have updated our future work accordingly.
 
@@
 
-The graphs have been modified to use better distinguishable colors and line styles. The use of color has also been made more consistent.
+The graphs have been modified to use better distinguishable colors and line
+styles. The use of color has also been made more consistent.
 
@@
 
-Note that the model is indeed not HI, as the text mentions it is a merge between PHOLD and HI.
+This model is indeed not HI, but rather a combination of PHOLD and HI, as
+explained in the text. The number of models has been corrected.
 
@@
 
-The scaling effects on performance have been investigated in the performance section.
-In this scenario (bfs vs dfs allocation) we specifically wanted to balance the model load over kernels.
-This will not always result in an equal load as is indeed shown in the allocation plot, but the main effect we want to demonstrate here is the effect of inter kernel communication and resulting dependencies between kernels this introduces.
-We have added a paragraph to section 6 to explain our reasoning.
+The scaling effects on performance have been investigated in the performance
+section. In this scenario (bfs vs dfs allocation) we specifically wanted to
+balance the model load over kernels. This will not always result in an equal
+load as is indeed shown in the allocation plot. The main effect we want to
+demonstrate here is the effect of inter-kernel communication and the
+dependencies between kernels this introduces. We have added a paragraph to
+section 6 to explain our reasoning.
 
@@
 
-We have updated the related work section.
+We could not find where [1] references to, but we believe it to be:
+Vicino D., Niyonkuru D., Wainer G. et al. Sequential PDEVS Architecture. In
+proceedings of the 2015 Symposium on Theory of Modeling and Simulation - DEVS.
+2015. pp. 165-172.
+
+We have updated the related work section with a reference to this work.
 
@@
 
-We have included memory allocation result in the conclusion. The other topics are already mentioned in the conclusion.
+We have included memory allocation result in the conclusion. The other topics
+are already mentioned in the conclusion.
 
@@
 
-We did not find relevant work from D Vicino on partitioning, but updated the related work section with a review of a paper from Lin Uhrmacher regarding partitioning.
+We have added the work by Vicino on CD++ and the work by Uhrmacher on
+partitioning.
 
@@
 
-A new subsection was added with extended results and discussion regarding these two papers, the related work section was updated with these references.
+A new subsection was added with extended results and discussion regarding
+these two papers, the related work section was updated with these references.
 
@@
 
-Thank you for notifying us about the quality of this reference. We have consequently removed it from our list of references.
+Thank you for notifying us about the quality of this reference. We have
+consequently removed it from our list of references.
 
@@
 
-The missing years are due to the Latex style file.
-Some @inproceedings don't include the year in their title.
-We have added the year field to every reference, so it isn’t our fault.
-We changed the items to @articles to work around this issue.
+The missing year of publication seems to be due to the bibliography style
+included in the Simulation LaTeX style files: @inproceedings did not display
+the year, even though it was defined in our bibtex files. We have manually
+modified this bibtex style for now, to include the years. The final style,
+however, will be up to the publisher.
 
@@
 
-We have benchmarked 3 configurations that demonstrate the different aspects highlighted in the theoretical work and analyzed their results.
+Both our Queue and Interconnect model could be configured to have high
+intrinsic parallelism. We have included results for this configuration in a
+new subsection (as discussed below).
 
@@
 
-We have implemented this. PDEVS can operate with or without the synchronization protocols allowing for a in depth comparison.
+Our simulation kernel has been extended with the option to parallelize the
+transition functions, exploiting the inherent parallelism found in Parallel
+DEVS. This can be done in combination with other synchronization protocols,
+allowing for an in-depth comparison.
 
@@
 
-In a new subsection we analyze the performance differences between PDEVS, standard, optimistic and conservative.
+A new subsection was added to elaborate on the inherent parallelism that can
+be found in Parallel DEVS, and our implementation exploiting this. A
+comparison was made between all options. These results clearly show that using
+the inherent parallelism is indeed a valid option when there are many
+simultaneous transition functions and high coupling. For situations with low
+(or none) simultaneous transition functions and low coupling, synchronization
+protocols still seem to be the better option. This makes us believe that both
+approaches are complementary. Ideally, the allocation could be optimized to
+place highly-coupled models with simultaneous transition functions together at
+a node, and have these low-coupled nodes with barely any simultaneous
+transition functions be synchronized using a synchronization protocol.
 
@@
 
-See item 3.
+This discussion was added (see our previous comment).
 
@@
 
-We have implemented the requested changes.
+As per our previous comment, we have implemented this protocol and added a
+discussion about it.
 

### 2016-10-10 publications/Journal/1-introduction.tex
Minor corrections.

@@
 Section~\textsc{\nameref{sec:2-background}} introduces the necessary background on synchronization protocols.
-Section~\textsc{\nameref{sec:3-features}} elaborates on our design that enables this flexibility.
-In Section~\textsc{\nameref{sec:4-performance}}, we evaluate performance of our tool by comparing its different synchronization protocols, and compare to \textit{adevs}.
+Section~\textsc{\nameref{sec:3-features}} elaborates on our design that enables the flexibility to switch protocols.
+In Section~\textsc{\nameref{sec:4-performance}}, we evaluate the performance of our tool using the different synchronization protocols, and we also compare with \textit{adevs}'s performance.
 We continue by introducing runtime switching of synchronization protocols and different options for model allocation in Section~\textsc{\nameref{sec:4b-hotswap}} and Section~\textsc{\nameref{sec:4a-allocation}}, respectively.
 Related work is discussed in Section~\textsc{\nameref{sec:5-related-work}}.
-Section~\textsc{\nameref{sec:6-conclusion}} concludes the paper and gives future work.
+Section~\textsc{\nameref{sec:6-conclusion}} concludes the paper and presents ideas for future work.

### 2016-10-10 publications/Journal/2-background.tex
Minor corrections.

@@
 In conservative synchronization, a node progresses independent of all other nodes, up to the point in time where it can guarantee that no causality errors happen.
-When simulation reaches this point, the node blocks until it can guarantee a new time until which no causality errors occur.
+When the simulation reaches this point, the node blocks until it can guarantee a new time until which no causality errors occur.
 In practice, this means that all nodes are aware of the current simulation time of all other nodes, and the time it takes an event to propagate (called \textit{lookahead}).
@@
 
-Rolling back simulation time requires the node to store past model states, such that they can be restored later.
+Rolling back simulation time requires the node to store past model states, so that they can be restored later.
 All incoming and outgoing events need to be stored as well.

### 2016-10-10 publications/Journal/3-features.tex
Minor corrections.

@@
 
-Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default value $\epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
-Providing this default implementation has no impact in sequential or optimistic simulation, since the function is never called the compiler will optimize it out.
+Defining a lookahead function is therefore recommended in combination with conservative synchronization, but it is not a necessity, as a default value $\epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
+Providing this default implementation has no impact in sequential or optimistic simulation; as the function is never called, the compiler will optimize it out.
 By providing this default implementation in the model base class we ensure that a model can run in sequential, optimistic and conservative simulation by default.

### 2016-10-10 publications/Journal/4-performance.tex
Minor corrections.

@@
 Analysis showed that this is caused by the high amount of events: event creation is much slower in \textit{dxex} than it is in \textit{adevs}, despite \textit{dxex}'s use of memory pools.
-To shield the user from threading and deallocation concerns \textit{dxex} provides an event superclass from which the user can derive to create a specialized event type.
+To shield the user from threading and deallocation concerns, \textit{dxex} provides an event superclass from which the user can derive to create a specialized event type.
 Copying, deallocation, and tracing are done at runtime, adding significant overhead when events happen frequently.
@@
 The number of events scales linear with the number of models, not quadratic.
-Figure~\ref{fig:Phold_benchmark} shows that the performance of \textit{dxex} and \textit{adevs} are very close to each-other, with \textit{adevs} slightly outperforming \textit{dxex}.
+Figure~\ref{fig:Phold_benchmark} shows that in terms of performance \textit{dxex} and \textit{adevs} are very close to each other, with \textit{adevs} slightly outperforming \textit{dxex}.
 
@@
 Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to sequential simulation for a fixed problem size (\textit{i.e.}, strong scaling).
-As the number of kernels increases, the optimistic kernel quickly becomes the worst choice.
+As the number of kernels increases, the optimistic protocol quickly becomes the worst choice.
 This is mainly caused by the pipeline structure of the model: the last models in the queue only respond to incoming messages and therefore have to be rolled back frequently.
-The difference between \textit{dxex} conservative and \textit{adevs} becoms smaller when more and more cores are used.
+The difference between \textit{dxex} conservative protocol and \textit{adevs} becomes smaller when more and more cores are used.
 The same effect can be seen when the problem size is increased in tandem with the number of used cores (\textit{i.e.}, weak scaling) in Figure~\ref{fig:Queue_plot_weak}.
@@
 Figure~\ref{fig:pdevs_plot_no_sleep} shows that in sequential simulation, the overhead of thread management and shared memory communication is crippling for performance.
-Even though many events occur concurrently, the actual computation being done in the transition function does not outweight the overhead, resulting in much slower simulation than sequential execution of transition functions.
+Even though many events occur concurrently, the actual computation being done in the transition function does not outweigh the overhead, resulting in much slower simulation than sequential execution of transition functions.
 When combined with synchronization protocols, performance is still reasonable, though the use of the inherent parallelism is not beneficial either way.
@@
 In \textit{dxex}, use of the inherent parallelism found in \textsf{Parallel DEVS} can be combined with any of the supported synchronization protocols.
-The user is not forced to choose between either but can divide the capabilities of his machine over both approaches.
+The user is not forced to choose between either: the capabilities of the computer can be divided over both approaches.
 While exploitation of the inherent parallelism can offer a significant speedup, a high number of transitions must be triggered simultaneously throughout the whole simulation.
@@
 The inherent parallelism of \textsf{Parallel DEVS} avoids synchronization overhead and is trivial to implement.
-It is not influenced by the coupling of models, and only by the number of simultaneous transition functions and the computational load in the transition function.
+It is not influenced by the coupling of models, only by the number of simultaneous transition functions and the computational load in the transition function.
 In contrast, either of the two synchronization protocols is heavily influenced by the coupling of the models, but not by the computational load or number of simultaneous transitions.
@@
 Apart from simulation execution time, memory usage during simulation is also of great importance.
-While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation infeasible.
+Having insufficient memory may cause sudden deterioration in performance, even to the point of making the simulation infeasible.
 We therefore also investigate the memory usage of different synchronization protocols, and again compare to \textit{adevs}.
@@
 We do not tackle the problem of states that become too large for a single machine to hold.
-This problem can be mitigated by distribution over multiple machines, which neither \textit{dxex} or \textit{adevs} support.
+This problem can be mitigated by distribution over multiple machines, which neither \textit{dxex} nor \textit{adevs} support.
 

### 2016-11-28 publications/Journal/3-features.tex
Summary of changes: - add proposed reply to reviewers - update abstract, introduction to mention PDEVS explicitly as a protocol compl$ - detail in a subsection implementation of PDEVS - detail our usage of parameter q Complete todo from previous review.

@@
 Again, we made use of specific new features of C++11, that are not available in Python, or even previous versions of the C++ language standard.
+
+% protocol or inherent parallelism?
+\subsubsection{PDEVS}
+In our tool each synchronization kernel is capable of executing its transitions in parallel.
+A configurable number of threads is allocated to optimally divide the processing load of concurrent transitions.
+The parallel execution of transitions introduces some overhead in thread pooling and locking, and disallows some optimizations in optimally rescheduling models.
+The concurrency of events and the combined computational load of the transitions has to outweigh this overhead to obtain a significant speedup, as we will demonstrate in Section~\textsc{\nameref{sec:4-performance}}.
 

### 2016-11-28 publications/Journal/4-performance.tex
Summary of changes: - add proposed reply to reviewers - update abstract, introduction to mention PDEVS explicitly as a protocol compl$ - detail in a subsection implementation of PDEVS - detail our usage of parameter q Complete todo from previous review.

@@
 %TODO what is meant with this allocation of threads and synchronized kernels?
+% 4 Kernels running on a dedicated thread, with per kernel 4 threads allocated to use for PDEVS.
 The effect of combining 4 synchronized kernels, with each 4 threads for PDEVS, is contrasted with 16 threads for PDEVS and 16 kernels for conservative and optimistic.
+The first scenario runs each kernel on a dedicated thread synchronized with the other 3 kernels.
+Each kernel can use up to 4 threads for PDEVS.
+In the second scenario PDEVS can use up to 16 threads while running in a sequential kernel.
+Finally 16 synchronized kernels are used with one thread per kernel.
 This allows us to observe which is more efficient in obtaining a speedup.
@@
 With respect to the theoretical analysis in~\cite{amdahlpdevs} this gives a parameter value $q = 1$.
+Q is the probability that a receiving model becomes immediately imminent (i.e.
+scheduled in the next timestep).
+Without random scheduling the receiving model becomes immediately imminent.
+With random scheduling the receiving model becomes imminent with a scheduled time in [1, timestep].
+This would result in a probability $q = \frac{n}{timestep}$ with n the receiving models at that point in time.
+The benchmarks are run sufficiently long enough to guarantee that the frequency of internal and external transitions is equal between random and non-random benchmarks.
+This results in n equal to all models in the benchmark.
+Only the concurrency of events differs.
 Models are equally distributed over the kernels and between threads in PDEVS.

### 2016-11-28 publications/Journal/abstract.tex
Update title
Summary of changes: - add proposed reply to reviewers - update abstract, introduction to mention PDEVS explicitly as a protocol compl$ - detail in a subsection implementation of PDEVS - detail our usage of parameter q Complete todo from previous review.

@@
 In this paper, we present a Parallel DEVS simulator, grafted on C++11 and based on PythonPDEVS, supporting both conservative and optimistic synchronization protocols.
+The simulator not only supports both protocols but also has the capability to switch between them at runtime.
+The simulator complements each of the synchronization protocols with the PDEVS protocol.
 We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol.
 A comparison is made to adevs in terms of CPU time and memory usage, to show that our modularity does not hinder performance.
+We compare the speedup obtained by synchronization with that of the inherent parallelism of PDEVS in isolation and combination, and contrast the results with the theoretical limits.
 We further allow for an external component to gather simulation statistics, on which runtime switching between the different synchronization protocols can be based.

### 2016-12-01 publications/Journal/4-performance.tex
Changes to comments and applied feedback from @stijn_manhaeve
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 The model is configured with depth 4 and width 300 if the transition function has no load, and width 50 if the computational load is active.
-In our configuration, an imminent model generates output, resulting in another model becoming imminent.
-With respect to the theoretical analysis in~\cite{amdahlpdevs} this gives a parameter value $q = 1$.
-Q is the probability that a receiving model becomes immediately imminent (i.e.
-scheduled in the next timestep).
-Without random scheduling the receiving model becomes immediately imminent.
-With random scheduling the receiving model becomes imminent with a scheduled time in [1, timestep].
-This would result in a probability $q = \frac{n}{timestep}$ with n the receiving models at that point in time.
+In our configuration, an imminent model generates output, resulting in the receiving model becoming imminent.
+With respect to the theoretical analysis ~\cite{amdahlpdevs} this gives a parameter value $q = 1$ for a fixed timeadvance and $q=0.5$ for a random timeadvance.
+Q is the probability that a receiving model becomes immediately imminent.
+Without a fixed timeadvance the receiving model becomes immediately imminent. 
+With random scheduling the receiving model becomes imminent with a scheduled time in the current timeslot or the next, chosen with a uniform distribution, therefore $q = \frac{1}{2}$.
  The benchmarks are run sufficiently long enough to guarantee that the frequency of internal and external transitions is equal between random and non-random benchmarks.
-This results in n equal to all models in the benchmark.
-Only the concurrency of events differs.
+The key difference is that although the event frequency is the same, their concurrency is not.
 Models are equally distributed over the kernels and between threads in PDEVS.
@@
 In Figure~\ref{fig:pdevs_plot_fixed_sleep}, we observe that exploiting only the inherent parallelism, results in a speedup of about 10.  
-%TODO again: what is this?
 Simulation with conservative and optimistic synchronization is run first with 16 kernels, then with 4 kernels having 4 PDEVS threads each. 

### 2016-12-01 publications/Journal/1-introduction.tex
Summary of changes: - add proposed reply to reviewers - update abstract, introduction to mention PDEVS explicitly as a protocol compl$ - detail in a subsection implementation of PDEVS - detail our usage of parameter q Complete todo from previous review.
First changes after review
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 
-This paper introduces \textit{DEVS-Ex-Machina}\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``\textit{dxex}''): our simulation tool~\cite{dxex} which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
+This paper introduces \textit{DEVS-Ex-Machina}\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``\textit{dxex}''): our simulation tool~\cite{dxex} which offers multiple synchronization protocols: no synchronization (single core execution), conservative synchronization, or optimistic synchronization.
+Each inter core synchronization protocol can be complemented with the parallel execution of concurrent events within a simulation core.
+We refer to this intra core protocol as pSim for the remainder of this paper.
+\\
+This results in the following list of available parallelization approaches:\\
+\begin{itemize}
+\item No inter core synchronization (single core), no intra core synchronization : S
+\item Conservative inter core synchronization, no intra core synchronization : C
+\item Optimistic inter core synchronization, no intra core synchronization : O
+\item No inter core synchronization (single core), pSim intra core synchronization : SP
+\item Conservative inter core synchronization, pSim intra core synchronization : CP
+\item Optimistic inter core synchronization, pSim intra core synchronization : OP
+\end{itemize}
 The selected synchronization protocol is transparent to the simulated model: users only determine the protocol to use.

### 2016-12-01 publications/Journal/3-features.tex
First changes after review
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 
-\subsubsection{Sequential}
-Our sequential simulation algorithm is very similar to the one found in \textit{PythonPDEVS}, including many optimizations.
+\subsubsection{Single Core}
+Our single core simulation algorithm is very similar to the one found in \textit{PythonPDEVS}, including many optimizations.
 Minor modifications were made, though, such that it can be overloaded by different synchronization protocol implementations.
@@
 It shows that there is a simulation \texttt{Core}, which simulates the \texttt{AtomicModel}s connected to it.
-The superclass \texttt{Core} represents the sequential simulation kernel.
+The superclass \texttt{Core} represents the single core simulation kernel.
 Subclasses define specific variants, such as \texttt{ConservativeCore} (conservative synchronization), \texttt{OptimisticCore} (optimistic synchronization), and \texttt{DynamicCore} (\textsf{Dynamic Structure DEVS}).
@@
 % protocol or inherent parallelism?
-\subsubsection{PDEVS}
-In our tool each synchronization kernel is capable of executing its transitions in parallel.
+\subsubsection{pSim}
+In our tool each simulation core, regardless of inter core synchronization, is capable of executing concurrent transitions in parallel.
 A configurable number of threads is allocated to optimally divide the processing load of concurrent transitions.

### 2016-12-01 publications/Journal/4-performance.tex
First changes after review
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 \subsection{Comparison to the abstract simulator}
+%reference pSim
 The abstract simulator of \textsf{Parallel DEVS} included its own notion of parallelism: transition functions that happen at the same point in simulated time can be trivially parallelized.
+In this work we refer to this approach as pSim.
 This parallelization is trivial, in the sense that there is no need for synchronization protocols.
@@
 In our configuration, an imminent model generates output, resulting in the receiving model becoming imminent.
-With respect to the theoretical analysis ~\cite{amdahlpdevs} this gives a parameter value $q = 1$ for a fixed timeadvance and $q=0.5$ for a random timeadvance.
-Q is the probability that a receiving model becomes immediately imminent.
-Without a fixed timeadvance the receiving model becomes immediately imminent. 
-With random scheduling the receiving model becomes imminent with a scheduled time in the current timeslot or the next, chosen with a uniform distribution, therefore $q = \frac{1}{2}$.
+The probability that an internal transition in a model generates output and is connected to a receiving model here is 1.
+With respect to the theoretical analysis ~\cite{amdahlpdevs} this gives a parameter value $q = 1$.
+The probability that a model becomes imminent (i.e.
+is scheduled to execute an internal transition at a given point in time) in this model is equal to 1 when a fixed timeadvance is used,  %todo p calculation
+$\frac{1}{timestep}$ when a random time advance is used, this corresponds with $p = 1$ and $p = \frac{1}{100}$ respectively.
+
 The benchmarks are run sufficiently long enough to guarantee that the frequency of internal and external transitions is equal between random and non-random benchmarks.
+In our model each model will be executing internal and external transitions, creating an ideal use case to evaluate the speedup inherent parallelism can obtain.
 The key difference is that although the event frequency is the same, their concurrency is not. 

### 2016-12-01 publications/Journal/4-performance.tex
Added citation to another paper comparing simulators, which was just accepted.

--- 2016-11-28 publications/Journal/4-performance.tex
@@
 In this section, we evaluate the performance of different synchronization protocols in \textit{dxex}.
-We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVStoneJournal}, to show that our modularity does not impede performance.
+We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVStoneJournal,DEVSSurvey}, to show that our modularity does not impede performance.
 CPU time and memory usage is compared for both sequential and parallel simulation.
@@
 %TODO what is meant with this allocation of threads and synchronized kernels?
-% 4 Kernels running on a dedicated thread, with per kernel 4 threads allocated to use for PDEVS.
 The effect of combining 4 synchronized kernels, with each 4 threads for PDEVS, is contrasted with 16 threads for PDEVS and 16 kernels for conservative and optimistic.
-The first scenario runs each kernel on a dedicated thread synchronized with the other 3 kernels.
-Each kernel can use up to 4 threads for PDEVS.
-In the second scenario PDEVS can use up to 16 threads while running in a sequential kernel.
-Finally 16 synchronized kernels are used with one thread per kernel.
 This allows us to observe which is more efficient in obtaining a speedup.
@@
 With respect to the theoretical analysis in~\cite{amdahlpdevs} this gives a parameter value $q = 1$.
-Q is the probability that a receiving model becomes immediately imminent (i.e.
-scheduled in the next timestep).
-Without random scheduling the receiving model becomes immediately imminent.
-With random scheduling the receiving model becomes imminent with a scheduled time in [1, timestep].
-This would result in a probability $q = \frac{n}{timestep}$ with n the receiving models at that point in time.
-The benchmarks are run sufficiently long enough to guarantee that the frequency of internal and external transitions is equal between random and non-random benchmarks.
-This results in n equal to all models in the benchmark.
-Only the concurrency of events differs.
 Models are equally distributed over the kernels and between threads in PDEVS.

### 2016-12-03 publications/Journal/3-features.tex
* turned the pSim into a macro (\pSim) in case we might want to change the name. * replaced a few occurrences of PDEVS with \pSim * changed mentions of "sequential" and "parallel" to single core and multi core where applicable. - - -> might need review to check if I didn't miss anything

@@
 Optimistic synchronization for when lookahead is unpredictable, or there are rare (almost) instantaneous events.
-Finally, sequential simulation is still required for models where parallelism is bad. 
-It is possible for the synchronization overhead to become larger than the achieved parallelism, resulting in slower simulation than sequential.
+Finally, single core simulation is still required for models where parallelism is bad. 
+It is possible for the synchronization overhead to become larger than the achieved parallelism, resulting in slower simulation than single core.
 
@@
 % protocol or inherent parallelism?
-\subsubsection{pSim}
+\subsubsection{\pSim}
 In our tool each simulation core, regardless of inter core synchronization, is capable of executing concurrent transitions in parallel.
@@
 We define synchronization protocol transparency as having a single model, which can be executed on each supported synchronization kernel, without any modifications.
-User should thus only provide one model, implemented in C++11, which can be simulated using sequential execution, conservative synchronization, or optimistic synchronization.
+User should thus only provide one model, implemented in C++11, which can be simulated using single core execution, conservative synchronization, or optimistic synchronization.
 The synchronization protocol to use is a simple configuration option.
@@
 While this is always a problem, it is aggravated in \textit{dxex} by providing automatic memory management for events and states.
-A model written for sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.
+A model written for single core simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.
 
@@
 To counter this, our memory allocators are backed by a thread-aware pooling library.
-In a sequential simulation kernel no allocated event persists beyond a single time advance, even allowing the use of an arena-style allocator.
+In a single core simulation kernel no allocated event persists beyond a single time advance, even allowing the use of an arena-style allocator.
 Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.
@@
 
-\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
+\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators in multi core simulation kernels and arena-style allocators for single core simulation.
 The latter can be faster, but at the cost of additional configuration.
@@
     %TODO: Add value on top of bars
-    \caption{Effect of memory allocators on parallel execution time, using optimistic synchronization.}
+    \caption{Effect of memory allocators on multi core execution time, using optimistic synchronization.}
     \label{fig:memallocators_parallel}
@@
 All accesses to the RNG will result in the modification of a global (\textit{i.e.}, shared between threads) variable.
-This easily becomes a bottleneck in simulation, since random numbers are a common occurence in simulation~\cite{Random}.
+This easily becomes a bottleneck in simulation, since random numbers are a common occurrence in simulation~\cite{Random}.
 As such, a nontrivial amount of time in a simulation is often spent waiting for an RNG.
@@
 This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
-\textit{Dxex}'s sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
+\textit{Dxex}'s single core and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
 The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.
 
-Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in sequential simulation is three times faster using any of the three alternatives, only the STL RNG is slow.
-For parallel simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between sequential and parallel simulation.
+Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in single core simulation is three times faster using any of the three alternatives, only the STL RNG is slow.
+For multi core simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between single core and multi core simulation.
 Conservative synchronization is almost unaffected by changing the RNG.

### 2016-12-03 publications/Journal/4-performance.tex
* turned the pSim into a macro (\pSim) in case we might want to change the name. * replaced a few occurrences of PDEVS with \pSim * changed mentions of "sequential" and "parallel" to single core and multi core where applicable. - - -> might need review to check if I didn't miss anything

--- 2016-12-01 publications/Journal/4-performance.tex
@@
 We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVStoneJournal}, to show that our modularity does not impede performance.
-CPU time and memory usage is compared for both sequential and parallel simulation.
+CPU time and memory usage is compared for both single core and multi core simulation.
 
-We start off with a comparison of sequential simulation, to show how \textit{adevs} and \textit{dxex} relate in this simple case.
-For the parallel simulation benchmarks, results are presented for both conservative and optimistic synchronization.
+We start off with a comparison of single core simulation, to show how \textit{adevs} and \textit{dxex} relate in this simple case.
+For the multi core simulation benchmarks, results are presented for both conservative and optimistic synchronization.
 
@@
 
-\subsection{Sequential Simulation}
-We start by evaluating sequential simulation performance, in order to obtain a baseline for our comparison of parallel simulation performance.
+\subsection{Signle Core Simulation}
+We start by evaluating single core simulation performance, in order to obtain a baseline for our comparison of multi core simulation performance.
 
@@
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_sequential.eps}
-	\caption{Queue benchmark results for sequential simulation.}
+	\caption{Queue benchmark results for single core simulation.}
 	\label{fig:Queue_benchmark_seq}
@@
 	\includegraphics[width=\columnwidth]{fig/interconnect_sequential.eps}
-	\caption{Interconnect benchmark results for sequential simulation.}
+	\caption{Interconnect benchmark results for single core simulation.}
 	\label{fig:Interconnect_benchmark}
@@
 	\includegraphics[width=\columnwidth]{fig/phold_sequential.eps}
-	\caption{PHold benchmark results for sequential simulation.}
+	\caption{PHold benchmark results for single core simulation.}
 	\label{fig:Phold_benchmark}
@@
 \subsection{Parallel Simulation}
-We now continue by describing our parallel simulation performance for different synchronization protocols, and compared to \textit{adevs}.
-The speedup of \textit{adevs} is computed with the corresponding \textit{dxex} sequential benchmark.
-This was done to take into account the performance difference observed in sequential simulation.
-As such, the highest speedup indicates the fastest results among all tools, independent of sequential simulation results.
+We now continue by describing our multi core simulation performance for different synchronization protocols, and compared to \textit{adevs}.
+The speedup of \textit{adevs} is computed with the corresponding \textit{dxex} single core benchmark.
+This was done to take into account the performance difference observed in single core simulation.
+As such, the highest speedup indicates the fastest results among all tools, independent of single core simulation results.
 
@@
 The Queue model is one single chain of models, resembling a pipeline.
-This structure can be exploited to prevent cyclic dependencies in the parallel simulation.
+This structure can be exploited to prevent cyclic dependencies in the multi core simulation.
 
-Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to sequential simulation for a fixed problem size (\textit{i.e.}, strong scaling).
+Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to single core simulation for a fixed problem size (\textit{i.e.}, strong scaling).
 As the number of kernels increases, the optimistic protocol quickly becomes the worst choice.
@@
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_strong_speedup.eps}
-	\caption{Queue model strong scaling speedup compared to \textit{dxex} sequential.}
+	\caption{Queue model strong scaling speedup compared to \textit{dxex} single core.}
 	\label{fig:Queue_plot_strong}
@@
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_weak_speedup.eps}
-	\caption{Queue model weak scaling speedup compared to \textit{dxex} sequential.}
+	\caption{Queue model weak scaling speedup compared to \textit{dxex} single core.}
 	\label{fig:Queue_plot_weak}
@@
 When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and the increasing number of rollbacks in optimistic simulation.
-All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
+All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in a multi core setting.
 
@@
     %TODO cyclic allocation in title?
-	\caption{Interconnect benchmark results for parallel simulation.}
+	\caption{Interconnect benchmark results for multi core simulation.}
 	\label{fig:interconnect_benchmark_parallel}
@@
     \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
-    \caption{PHOLD benchmark results for parallel simulation using four kernels, with varying percentage of remote events.}
+    \caption{PHOLD benchmark results for multi core simulation using four kernels, with varying percentage of remote events.}
 \end{figure}
@@
 	\includegraphics[width=\columnwidth]{fig/phold_priority.eps}
-	\caption{PHOLD benchmark results for parallel simulation using four kernels, with varying amount of high-priority events.}
+	\caption{PHOLD benchmark results for multi core simulation using four kernels, with varying amount of high-priority events.}
 	\label{fig:phold_priority}
@@
 The abstract simulator of \textsf{Parallel DEVS} included its own notion of parallelism: transition functions that happen at the same point in simulated time can be trivially parallelized.
-In this work we refer to this approach as pSim.
+In this work we refer to this approach as \pSim so as not to confuse with the PDEVS formalism.
 This parallelization is trivial, in the sense that there is no need for synchronization protocols.
@@
 % 4 Kernels running on a dedicated thread, with per kernel 4 threads allocated to use for PDEVS.
-The effect of combining 4 synchronized kernels, with each 4 threads for PDEVS, is contrasted with 16 threads for PDEVS and 16 kernels for conservative and optimistic.
+The effect of combining 4 synchronized kernels, with each 4 threads for \pSim, is contrasted with 16 threads for \pSim and 16 kernels for conservative and optimistic.
 The first scenario runs each kernel on a dedicated thread synchronized with the other 3 kernels.
-Each kernel can use up to 4 threads for PDEVS.
-In the second scenario PDEVS can use up to 16 threads while running in a sequential kernel.
+Each kernel can use up to 4 threads for \pSim.
+In the second scenario \pSim can use up to 16 threads while running in a single core kernel.
 Finally 16 synchronized kernels are used with one thread per kernel.
@@
 The model is configured with depth 4 and width 300 if the transition function has no load, and width 50 if the computational load is active.
-In our configuration, an imminent model generates output, resulting in the receiving model becoming imminent.
-The probability that an internal transition in a model generates output and is connected to a receiving model here is 1.
-With respect to the theoretical analysis ~\cite{amdahlpdevs} this gives a parameter value $q = 1$.
-The probability that a model becomes imminent (i.e.
-is scheduled to execute an internal transition at a given point in time) in this model is equal to 1 when a fixed timeadvance is used,  %todo p calculation
-$\frac{1}{timestep}$ when a random time advance is used, this corresponds with $p = 1$ and $p = \frac{1}{100}$ respectively.
+In our configuration, an imminent model always generates output, resulting in the receiving model becoming imminent.
+The probability that an internal transition in a model generates output and is connected to a receiving model here is $1$.
+The probability that a model becomes immediately imminent depends on whether fixed or random timeadvance is used.
+In the case of fixed timeadvance, a model will always become immediately imminent.
+When random timeadvance is used, this probability is $\frac{1}{n}$, with $n$ denoting the total amount of atomic models.
+These probabilities correspond with the $q$ parameter in the theoretical analysis ~\cite{amdahlpdevs}.
+The probability that a model is imminent (i.e.
+is scheduled to execute an internal transition) at any given point in time in this model is equal to $1$ when a fixed timeadvance is used and $\frac{1}{n}$ when a random time advance is used.
+These probabilities correspond with the $p$ parameter in the theoretical analysis.
 
-The benchmarks are run sufficiently long enough to guarantee that the frequency of internal and external transitions is equal between random and non-random benchmarks.
+The benchmarks are run sufficiently long enough to guarantee that the frequency of internal and external transitions is equal within a benchmark, regardless of the randomness of the timeadvance.
 In our model each model will be executing internal and external transitions, creating an ideal use case to evaluate the speedup inherent parallelism can obtain.
 The key difference is that although the event frequency is the same, their concurrency is not. 
-Models are equally distributed over the kernels and between threads in PDEVS.
-The coefficient of variation of our results is less than 1\%.
+Models are equally distributed over the kernels and between threads in \pSim.
+The coefficient of variation of our results is less than $1\%$.
 The communication overhead is hard to estimate, but given our coefficient of variation, we can expect this overhead to be constant.
@@
 In Figure~\ref{fig:pdevs_plot_fixed_sleep}, we observe that exploiting only the inherent parallelism, results in a speedup of about 10.  
-Simulation with conservative and optimistic synchronization is run first with 16 kernels, then with 4 kernels having 4 PDEVS threads each. 
+Simulation with conservative and optimistic synchronization is run first with 16 kernels, then with 4 kernels having 4 \pSim threads each. 
 Both synchronization protocols achieve a speedup similar to only exploiting the inherent parallelism.
@@
 Finally, we remove the computation load from the transition function, in combination with many concurrent events.
-Figure~\ref{fig:pdevs_plot_no_sleep} shows that in sequential simulation, the overhead of thread management and shared memory communication is crippling for performance.
-Even though many events occur concurrently, the actual computation being done in the transition function does not outweigh the overhead, resulting in much slower simulation than sequential execution of transition functions.
+Figure~\ref{fig:pdevs_plot_no_sleep} shows that in single core simulation, the overhead of thread management and shared memory communication is crippling for performance.
+Even though many events occur concurrently, the actual computation being done in the transition function does not outweigh the overhead, resulting in much slower simulation than single core execution of transition functions.
 When combined with synchronization protocols, performance is still reasonable, though the use of the inherent parallelism is not beneficial either way.
-Both synchronization protocols achieve a reasonable speedup without PDEVS, as the overhead of thread management (creation and destruction) is avoided.
+Both synchronization protocols achieve a reasonable speedup without \pSim, as the overhead of thread management (creation and destruction) is avoided.
 These results are similar to~\cite{Himmelspach}.
@@
 
-Conservative simulation always uses more memory than sequential simulation, as is to be expected.
+Conservative simulation always uses more memory than single core simulation, as is to be expected.
 Additional memory is required for the multiple threads, but also to store all events that are processed simultaneously.
@@
 Second, the allocation of models is nontrivial and has a significant impact on performance.
-While our parallel speedup for the Queue model, for example, was rather high, this is mostly due to characteristics of the model: the dependency graph does not contain any cycles.
+While our multi core speedup for the Queue model, for example, was rather high, this is mostly due to characteristics of the model: the dependency graph does not contain any cycles.
 When cycles were introduced, as in the Interconnect model, performance became disastrous.

### 2016-12-03 publications/Journal/6-allocation.tex
Minor corrections.
* turned the pSim into a macro (\pSim) in case we might want to change the name. * replaced a few occurrences of PDEVS with \pSim * changed mentions of "sequential" and "parallel" to single core and multi core where applicable. - - -> might need review to check if I didn't miss anything

@@
 If no preference is given, a simple striping scheme is used but this is often insufficient.
-By overriding the default allocator, a modeller tunes the allocation scheme for a specific model, maximizing parallel speedup.
+By overriding the default allocator, a modeller tunes the allocation scheme for a specific model, maximizing multi core speedup.
 This interface can be linked to graph algorithms for automatic allocation scheme generation, for example to avoid cycles in the dependency graph.
@@
 Unlike the Queue model, the width of the hierarchy is still present in the topology of the atomic models after direct connection.
-The PHOLDTree model allows us to investigate parallel speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.
+The PHOLDTree model allows us to investigate multi core speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.
 
@@
 This topology is interesting since it highlights the effects of allocation.
-First, we evaluate the model in sequential simulation to provide a baseline for parallel simulation.
+First, we evaluate the model in single core simulation to provide a baseline for parallel simulation.
 
-\subsubsection{Sequential Simulation}
+\subsubsection{Single Core Simulation}
 Since \textit{adevs} does not use direct connection, we expect a noticeable performance difference between \textit{dxex} and \textit{adevs}.
@@
     \includegraphics[width=\columnwidth]{fig/pholdtree_sequential_dn.eps}
-    \caption{Effect of hierarchy in sequential simulation.}
+    \caption{Effect of hierarchy in single core simulation.}
     \label{fig:PHOLDtree_seq_dn_benchmark}
@@
 
-\subsubsection{Parallel Simulation}
+\subsubsection{Multi Core Simulation}
 Next, we run the model using two different model allocation schemes: breadth-first and depth-first.
@@
 With depth-first allocation, we traverse the tree in a depth-first way, allocating subsequently visited atomic models to the same node.
-This means, intuitively, that subtreees are frequently allocated to the same node.
+This means, intuitively, that subtrees are frequently allocated to the same node.
 This allocation strategy is shown in Figure~\ref{fig:PholdTree_model_dfs}.
@@
 The breadth first allocation scheme results in a dependency chain with multiple branches, much like in the Queue model.
-Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model.
+Such a linear dependency chain can result in a multi core speedup as we demonstrated with the Queue model.
 This is not always true though: a single kernel with an unbalanced number of atomic models or unequal computational load in transition functions slows down the remainder of the chain.
@@
 The most prominent aspect of these results is the low performance for conservative depth-first allocation for two kernels.
-This is mostly caused by the difference between a sequential simulation and a parallel simulation: suddenly we need to take into account synchronization and passing around of lookahead values.
+This is mostly caused by the difference between a single core simulation and a multi core simulation: suddenly we need to take into account synchronization and passing around of lookahead values.
 And since the number of kernels is low, the overhead dominates any further speedup.

### 2016-12-06 publications/Journal/3-features.tex
Review features section.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 As no single synchronization protocol is ideal for all domains, a general purpose simulation tool should support multiple protocols.
-We argue that a general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for \textit{dxex}.
+We argue that a general purpose simulation tool should support single core, conservative, and optimistic synchronization, as is the case for \textit{dxex}.
+Futhermore it should be possible to combine each inter core synchronization protocol with the parallel execution of events defined by \pSim.
 
@@
 The more attractive option is for the simulation tool to provide a default lookahead function, such that simulation can run anyway, but likely not at peak performance.
-Depending on the model, simulation performance might still be faster than sequential simulation.
+Depending on the model, simulation performance might still be faster than single core simulation.
 
@@
 They can be returned immediately to the memory allocator as the lifetime of these objects is known at creation. 
+\pSim therefore has an advantage in lower overhead in comparison with inter kernel synchronization protocols.
 In contrast, inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
@@
     \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
-    %TODO: Add value on top of bars
     \caption{Effect of memory allocators on multi core execution time, using optimistic synchronization.}
@@
     %TODO: why is STL RNG in state just as fast in sequential? Does storing it in the state really do that much?
-    %TODO: consistency: TRNG, STL RNG in state, and TRNG in state
+    % The pattern is identical, the only difference is in optimistic where the copying of a larger structure
+    % (stl rng in state) becomes expensive enough. Another possible reason for a speedup w.r.t not storing
+    % is the dependency between calls to the random number generator, the compiler can determine that
+    % these calculations are local to the object.
     \caption{Speedup with different RNG usage patterns compared to STL random number generator.}

### 2016-12-06 publications/Journal/4-performance.tex
Review section 4
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 For the multi core simulation benchmarks, results are presented for both conservative and optimistic synchronization.
+We conclude with a comparison between inter and intra kernel parallelization and give use cases where these protocols can complement each other in the context of recent theoretical work ~\cite{amdahlpdevs}.
 
@@
 \subsection{Comparison to the abstract simulator}
-%reference pSim
 The abstract simulator of \textsf{Parallel DEVS} included its own notion of parallelism: transition functions that happen at the same point in simulated time can be trivially parallelized.
@@
 First we create a model where all transitions happen simultaneously, with a significant computational load in the transition functions.
-In Figure~\ref{fig:pdevs_plot_fixed_sleep}, we observe that exploiting only the inherent parallelism, results in a speedup of about 10.  
+In Figure~\ref{fig:psim_plot_fixed_sleep}, we observe that exploiting only the inherent parallelism, results in a speedup of about 10.  
 Simulation with conservative and optimistic synchronization is run first with 16 kernels, then with 4 kernels having 4 \pSim threads each. 
@@
 	\includegraphics[width=\columnwidth]{fig/pdevs_fixed_sleep.eps}
-    %TODO fix key in figure: pdevs --> inherent parallelism
 	\caption{Queue speedup benchmark for PDEVS and synchronization with maximal concurrent events and significant computational load}
-	\label{fig:pdevs_plot_fixed_sleep}
+	\label{fig:psim_plot_fixed_sleep}
 \end{figure}
@@
 The transition function has the same computational load as in the previous configuration. 
-The results in Figure~\ref{fig:pdevs_plot_random_sleep} show that parallelizing transition functions only, adds little overhead. 
+The results in Figure~\ref{fig:psim_plot_random_sleep} show that parallelizing transition functions only, adds little overhead. 
 Nonetheless, performance is not increased at all: the chances of executing a few transition functions in parallel does not make up for the overhead of checking.
@@
 	\includegraphics[width=\columnwidth]{fig/pdevs_random_sleep.eps}
-    %TODO fix key in figure: pdevs --> inherent parallelism
 	\caption{Queue speedup benchmark for PDEVS and synchronization with randomized events and significant computational load}
-	\label{fig:pdevs_plot_random_sleep}
+	\label{fig:psim_plot_random_sleep}
 \end{figure}
@@
 Finally, we remove the computation load from the transition function, in combination with many concurrent events.
-Figure~\ref{fig:pdevs_plot_no_sleep} shows that in single core simulation, the overhead of thread management and shared memory communication is crippling for performance.
+Figure~\ref{fig:psim_plot_no_sleep} shows that in single core simulation, the overhead of thread management and shared memory communication is crippling for performance.
 Even though many events occur concurrently, the actual computation being done in the transition function does not outweigh the overhead, resulting in much slower simulation than single core execution of transition functions.
@@
 	\includegraphics[width=\columnwidth]{fig/pdevs_no_sleep.eps}
-    %TODO fix key in figure: pdevs --> inherent parallelism
 	\caption{Queue speedup benchmark for PDEVS and synchronization with maximal concurrent events and trivial computational load}
-	\label{fig:pdevs_plot_no_sleep}
+	\label{fig:psim_plot_no_sleep}
 \end{figure}

### 2016-12-06 publications/Journal/5-hotswap.tex
* fixed image 3 capitalization of abbreviations * fixed image 12 title * fixed images 15-17 mentioning pdevs, using "inherent parallelism" instead * fixed images 19-21 - - -> also changed the rotation and scaling of image 19 back to what it should be had the .eps to .pdf converter of Latex not messed up earlier. * fixed images 26-27 --- note: on image 21, I haven't changed the style of the 0 lines into the regular red+dashed+arrow at the end because this was a quick Inkscape job.
* turned the pSim into a macro (\pSim) in case we might want to change the name. * replaced a few occurrences of PD

@@
 For a simple model and a few cores, the overhead is less than a second. 
+Creating new cores and moving the simulation models has a constant overhead linear in the amount of cores and models.
+The time required to synchronize and halt the existing cores is variable, especially if the old synchronization protocol is optimistic since there is no real limit on the virtual time difference between cores.
+Given that the existing cores are equally loaded, this time difference will in practice scale linearly with the number of cores.
 
@@
     \center
-    \includegraphics[height=\textwidth, angle=-90 ]{fig/queue_allocation.eps}
+    \includegraphics[width=\textwidth ]{fig/queue_allocation.eps}
     \caption{Queue model simulation trace across 4 kernels.}
@@
     \includegraphics[width=\plotfraction\columnwidth]{fig/phold_parallel_allocation.eps}
-    \caption{Phold simulation trace for parallel simulation using three kernels.}
+    \caption{Phold simulation trace for multi core simulation using three kernels.}
     \label{fig:phold_allocation}

### 2016-12-06 publications/Journal/4-performance.tex
Final changes.
- update plots to use pSim reference in comparison section 4
- list all changes in reply to reviewers file
- A few small changes in section 4 regarding 2nd source of parallelism as mentioned by reviewer.

Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 In this section, we evaluate the performance of different synchronization protocols in \textit{dxex}.
-We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVStoneJournal}, to show that our modularity does not impede performance.
+We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVStoneJournal,DEVSSurvey}, to show that our modularity does not impede performance.
 CPU time and memory usage is compared for both single core and multi core simulation.
@@
 The key difference is that although the event frequency is the same, their concurrency is not. 
+This benchmark combines both possible sources of parallelization : concurrent internal and external transitions. 
 Models are equally distributed over the kernels and between threads in \pSim.

### 2016-12-07 publications/Journal/3-features.tex
Update design section and UML to switch from core to kernel if referring to the simulator instance.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

@@
 \textit{Dxex} uses new optimizations from the C++14 standard when possible.
-The C++11 standard threading primitives are used to run the different simulation cores.
-Within a single simulation core, OpenMP~\cite{openmp4} is used to parallelize the transition functions, as is usual in \textsf{Parallel DEVS}.
+The C++11 standard threading primitives are used to run the different simulation kernels.
+Within a single simulation kernel, OpenMP~\cite{openmp4} is used to parallelize the transition functions, as is usual in \textsf{Parallel DEVS}.
 
@@
 We argue that a general purpose simulation tool should support single core, conservative, and optimistic synchronization, as is the case for \textit{dxex}.
-Futhermore it should be possible to combine each inter core synchronization protocol with the parallel execution of events defined by \pSim.
+Futhermore it should be possible to combine each inter kernel synchronization protocol with the parallel execution of events defined by \pSim.
 
@@
 
-Data exchange between different simulation cores happens through shared memory, using the new C++11 synchronization primitives.
+Data exchange between different simulation kernels happens through shared memory, using the new C++11 synchronization primitives.
 Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
@@
 An overview of \textit{dxex}'s design is given in Figure~\ref{fig:class_diagram}.
-It shows that there is a simulation \texttt{Core}, which simulates the \texttt{AtomicModel}s connected to it.
-The superclass \texttt{Core} represents the single core simulation kernel.
-Subclasses define specific variants, such as \texttt{ConservativeCore} (conservative synchronization), \texttt{OptimisticCore} (optimistic synchronization), and \texttt{DynamicCore} (\textsf{Dynamic Structure DEVS}).
+It shows that there is a simulation \texttt{Kernel}, which simulates the \texttt{AtomicModel}s connected to it.
+The superclass \texttt{Kernel} represents the single core simulation kernel.
+Subclasses define specific variants, such as \texttt{ConservativeKernel} (conservative synchronization), \texttt{OptimisticKernel} (optimistic synchronization), and \texttt{DynamicKernel} (\textsf{Dynamic Structure DEVS}).
 
@@
 \subsubsection{\pSim}
-In our tool each simulation core, regardless of inter core synchronization, is capable of executing concurrent transitions in parallel.
+In our tool each simulation kernel, regardless of inter kernel synchronization, is capable of executing concurrent transitions in parallel.
 A configurable number of threads is allocated to optimally divide the processing load of concurrent transitions.

### 2016-12-07 publications/Journal/4-performance.tex

@@
 	\includegraphics[width=\modelfraction\columnwidth]{fig/phold_model.pdf}
-	\caption{PHOLD model for three models.}
+	\caption{PHOLD model for four models. Color denotes the two nodes.}
 	\label{fig:PHOLD_model}
@@
 
-\subsection{Signle Core Simulation}
+\subsection{Single Core Simulation}
 We start by evaluating single core simulation performance, in order to obtain a baseline for our comparison of multi core simulation performance.

### 2016-12-07 publications/Journal/1-introduction.tex
* turned the pSim into a macro (\pSim) in case we might want to change the name. * replaced a few occurrences of PDEVS with \pSim * changed mentions of "sequential" and "parallel" to single core and multi core where applicable. - - -> might need review to check if I didn't miss anything
Rewrote first part of the paper to the new synchronization terminology

--- 2016-10-10 publications/Journal/1-introduction.tex
@@
 Several synchronization protocols from the discrete event simulation community~\cite{FujimotoBook} have been applied to (\textsf{Parallel}) \textsf{DEVS} simulation~\cite{globaltimewarp}.
-With synchronization protocols, different simulation cores can be at different points in simulated time, significantly increasing parallelism, at the cost of synchronization overhead.
-While several parallel \textsf{DEVS} simulation kernels exist, they are often limited to a single synchronization protocol.
+With synchronization protocols, different simulation kernels can be at different points in simulated time, significantly increasing parallelism at the cost of synchronization overhead.
+While several parallel \textsf{DEVS} simulation tools exist, they are often limited to a single synchronization protocol.
 The reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other for specific models~\cite{Jafer}.
@@
 
-This paper introduces \textit{DEVS-Ex-Machina}\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``\textit{dxex}''): our simulation tool~\cite{dxex} which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
+This paper introduces \textit{DEVS-Ex-Machina}\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``\textit{dxex}''): our simulation tool~\cite{dxex} which offers 6 synchronization protocols.
+The simulation tool can spawn multiple kernels, each of which has its own simulation time and simulation control, though they may exchange timestamped events with each other.
+Each kernel can spawn multiple threads, each of which will be mapped to a physical CPU core by the operating system.
+We note the presence of two kinds of synchronization: inter-kernel and intra-kernel.
+Inter-kernel synchronization uses either no synchronization, conservative synchronization, or optimistic synchronization.
+Intra-kernel synchronization uses either no synchronization or exploits the inherent parallelism of Parallel DEVS to allocate concurrent transitions to separate threads.
+The inherent parallelism in the intra-kernel synchronization protocol is used \pSim in the remainder of this paper.
+
+This results in the following set of synchronization protocol combinations:
+\begin{itemize}
+\item No inter-kernel synchronization, no intra-kernel synchronization (\textit{NN})
+\item Conservative inter-kernel synchronization, no intra-kernel synchronization (\textit{CN})
+\item Optimistic inter-kernel synchronization, no intra-kernel synchronization (\textit{ON})
+\item No inter-kernel synchronization, \pSim intra-kernel synchronization (\textit{NP})
+\item Conservative inter-kernel synchronization, \pSim intra-kernel synchronization (\textit{CP})
+\item Optimistic inter-kernel synchronization, \pSim intra-kernels synchronization (\textit{OP})
+\end{itemize}
+
+When refering to a set of them, we use the notation \textit{x} to denote any possible algorithm.
+For example, \textit{xP} refers to NP, CP, and OP.
+
 The selected synchronization protocol is transparent to the simulated model: users only determine the protocol to use.
-Users who simulate a wide variety of models, with different ideal synchronization protocols, can run the exact same model using the exact same tool, but with different synchronization protocols.
+Users who simulate a wide variety of models, with different ideal synchronization protocols, can keep using the exact same tool, but with different synchronization protocols.
 As model behaviour, and thus the ideal synchronization protocol, might change during simulation, runtime switching of synchronization protocols is also supported.
@@
 Using several benchmark models, we demonstrate the factors influencing the performance under a given synchronization protocol.
-Additionally, we investigate a model which changes its behaviour (and ideal synchronization protocol) during simulation.
+Additionally, we investigate a model which changes its behaviour (and corresponding ideal synchronization protocol) during simulation.
 \textit{Dxex}, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol.

### 2016-12-07 publications/Journal/2-background.tex
First changes after review
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
* turned the pSim into a macro (\pSim) in case we might want to change the name. * replaced a few occurrences of PDEVS with \pSim * changed mentions of "sequential" and "parallel" to single core and multi core where applicable. - - -> might need review to check if I didn't miss anything
Review abstract, intro, background
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
Rewrote first part of the paper to the new synchronization terminology

@@
 This section briefly introduces the two most prominent synchronization protocols: conservative and optimistic synchronization.
-Both algorithms are supported by Dxex.
+Both algorithms are supported by dxex. 
+These protocols synchronize between kernels, but do not specify the parallelization of events within a single kernel.
+For example, \pSim is orthogonal to this as it decides upon the parallelization of events within a single kernel.
 

### 2016-12-07 publications/Journal/3-features.tex
Rewrote first part of the paper to the new synchronization terminology

@@
 We do not elaborate on these features in this paper.
-But whereas \textit{PythonPDEVS} only supports optimistic synchronization, \textit{dxex} support multiple synchronization protocols (though only in parallel).
+But whereas \textit{PythonPDEVS} only supports optimistic synchronization, \textit{dxex} support multiple synchronization protocols (though not distributed).
 This is in line with the design principle of \textit{PythonPDEVS}: allow users to pass performance hints~\cite{PythonPDEVS_ACTIMS} to the simulation kernel.
 In our case, a user can configure the simulation kernel with the synchronization protocol to use, or even switch the synchronization protocol during runtime.
-Our implementation in C++11 also allows for (compiled) optimizations which were plainly impossible in an interpreted language.
+Our implementation in C++11 also allows for (compiled) optimizations which were plainly impossible in an interpreted language, such as Python.
 \textit{Dxex} uses new optimizations from the C++14 standard when possible.
@@
 As no single synchronization protocol is ideal for all domains, a general purpose simulation tool should support multiple protocols.
-We argue that a general purpose simulation tool should support single core, conservative, and optimistic synchronization, as is the case for \textit{dxex}.
-Futhermore it should be possible to combine each inter kernel synchronization protocol with the parallel execution of events defined by \pSim.
+We argue that a general purpose simulation tool should support all six standard synchronization protocols, as is the case for \textit{dxex}.
 
-These different protocols relate to three different model characteristics.
-Conservative synchronization for when high lookahead exists between different nodes, and barely any blocking is necessary.
-Optimistic synchronization for when lookahead is unpredictable, or there are rare (almost) instantaneous events.
-Finally, single core simulation is still required for models where parallelism is bad. 
-It is possible for the synchronization overhead to become larger than the achieved parallelism, resulting in slower simulation than single core.
+Different protocols relate to different model characteristics.
+For example, \textit{Cx} is for when high lookahead exists between different nodes, whereas \textit{Ox} is for when lookahead is unpredictable.
+It is possible for the synchronization overhead to become larger than the achieved parallelism, resulting in slower simulation than fully sequential execution (\textit{NN}).
 
 Data exchange between different simulation kernels happens through shared memory, using the new C++11 synchronization primitives.
-Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
+This was also possible in previous versions of the C++ standard by falling back to non-portable C functions.
 C++11 further allows us to make the implementation portable, as well as more efficient: the compiler makes further optimizations to heavily used components.
 
-\subsubsection{Single Core}
-Our single core simulation algorithm is very similar to the one found in \textit{PythonPDEVS}, including many optimizations.
-Minor modifications were made, though, such that it can be overloaded by different synchronization protocol implementations.
+\subsubsection{Inter-Kernel Synchronization}
+Our core simulation algorithm is very similar to the one found in \textit{PythonPDEVS}, including many optimizations.
+Minor modifications were made though, such that it can be overloaded by different synchronization protocol implementations.
 This way, the \textsf{DEVS} simulation algorithm is implemented once, but parts can be overridden as needed.
-In theory, more synchronization protocols (\textit{e.g.}, other algorithms for conservative synchronization) can be added without altering our design.
 
@@
 It shows that there is a simulation \texttt{Kernel}, which simulates the \texttt{AtomicModel}s connected to it.
-The superclass \texttt{Kernel} represents the single core simulation kernel.
+The superclass \texttt{Kernel} represents a standalone simulation kernel (\textit{Nx}).
 Subclasses define specific variants, such as \texttt{ConservativeKernel} (conservative synchronization), \texttt{OptimisticKernel} (optimistic synchronization), and \texttt{DynamicKernel} (\textsf{Dynamic Structure DEVS}).
+In theory, more synchronization protocols (\textit{e.g.}, other algorithms for conservative synchronization) can be added without altering our design.
 
@@
 
-\subsubsection{Conservative}
-For conservative synchronization, each node must determine the nodes it is influenced by.
+The following inter-kernel synchronization protocols are implemented.
+
+\paragraph{None (\textit{Nx})}
+No inter-kernel synchronization is the base case, implemented in the \texttt{Kernel}.
+It can be overloaded by any of the other simulation kernels, which augment it with inter-kernel synchronization methods.
+
+\paragraph{Conservative (\textit{Cx})}
+For conservative synchronization, each kernel determines the kernels it is influenced by.
 Each model needs to provide a lookahead function, which determines the lookahead depending on the current simulation state.
@@
 
-\subsubsection{Optimistic}
+\paragraph{Optimistic (\textit{Ox})}
 For optimistic synchronization, each node must be able to roll back to a previous point in time.
-This is often implemented through the use of state saving.
+This is implemented with state saving.
 This needs to be done carefully in order to avoid unnecessary copies, and minimize the overhead.
@@
 Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed, and irreversible actions (\textit{e.g.}, printing) are committed.
-
 The main problem we encountered in our implementation is the aggressive use of memory.
@@
 
-% protocol or inherent parallelism?
-\subsubsection{\pSim}
-In our tool each simulation kernel, regardless of inter kernel synchronization, is capable of executing concurrent transitions in parallel.
-A configurable number of threads is allocated to optimally divide the processing load of concurrent transitions.
+\subsubsection{Intra-Kernel Synchronization}
+In our tool, each simulation kernel is capable of executing concurrent transitions in parallel, whether they are external, internal, or confluent.
+
+\paragraph{None (\textit{xN})}
+No intra-kernel parallelism is used, meaning that all concurrent transitions are processed sequentially.
+Note that the order in which they are processed is non-deterministic.
+
+\paragraph{\pSim (\textit{xP})}
+With intra-kernel parallelism, a configurable number of threads is allocated to optimally divide the processing load of concurrent transitions.
 The parallel execution of transitions introduces some overhead in thread pooling and locking, and disallows some optimizations in optimally rescheduling models.
@@
 \subsection{Synchronization Protocol Transparency}
-We define synchronization protocol transparency as having a single model, which can be executed on each supported synchronization kernel, without any modifications.
-User should thus only provide one model, implemented in C++11, which can be simulated using single core execution, conservative synchronization, or optimistic synchronization.
+We define synchronization protocol transparency as having a single model, which can be executed on any kind of kernel, without any modifications.
+User should thus only provide one model, implemented in C++11, which can be simulated using any of the six synchronization protocols.
 The synchronization protocol to use is a simple configuration option.
-The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
+The exception is conservative synchronization (\textit{Cx}), where a lookahead function is required, which is not used in other synchronization kernels.
 Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.
@@
 The more attractive option is for the simulation tool to provide a default lookahead function, such that simulation can run anyway, but likely not at peak performance.
-Depending on the model, simulation performance might still be faster than single core simulation.
+Depending on the model, simulation performance might still be faster than \textit{Nx}.
 
@@
 Providing this default implementation has no impact in sequential or optimistic simulation; as the function is never called, the compiler will optimize it out.
-By providing this default implementation in the model base class we ensure that a model can run in sequential, optimistic and conservative simulation by default.
+By providing this default implementation in the model base class we ensure that a model can run with any synchronization protocol.
 
@@
 The goal of our contribution is to increase simulation performance as much as possible, leveraging parallel computation in the process.
-Parallelizing the simulation kernel goes further, however, than merely implementing the different synchronization protocols.
+Parallelizing goes further, however, than merely implementing the different synchronization protocols.
 
@@
 \label{sec:4-subsec:overhead-pgraph:memory}
-Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}, since memory bandwidth doesn't increase as fast as the number of cores using it.
+Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}, since memory bandwidth doesn't increase as fast as the number of CPU cores using it.
 While this is always a problem, it is aggravated in \textit{dxex} by providing automatic memory management for events and states.
-A model written for single core simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.
+A model written for \textit{Nx} synchronization will run correctly with conservative (\textit{Cx}) or optimistic (\textit{Ox}) synchronization without altering, from the point of view of the modeller, the (de)allocation semantics of events or states.
 
@@
 To counter this, our memory allocators are backed by a thread-aware pooling library.
-In a single core simulation kernel no allocated event persists beyond a single time advance, even allowing the use of an arena-style allocator.
+With \textit{Nx} synchronization, no allocated event persists beyond a single time advance, even allowing the use of an arena-style allocator.
 Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.
 
-Intra-kernel events can be (de)allocated without requiring synchronization with the other kernels. 
+Intra-kernel events can be (de)allocated without synchronization with the other kernels. 
 They can be returned immediately to the memory allocator as the lifetime of these objects is known at creation. 
-\pSim therefore has an advantage in lower overhead in comparison with inter kernel synchronization protocols.
-In contrast, inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
-As a result, simulations with many inter-kernel events suffer a performance hit, whereas the impact of many intra-kernel events can be minimized using arena allocators~\cite{Arena}.
+In contrast, inter-kernel events need a GVT algorithm to determine when safe deallocation can occur.
+Intra-kernel synchronization protocols therefore have a lower overhead than inter-kernel synchronization protocols.
+Simulations with many inter-kernel events suffer a performance hit, whereas the impact of many intra-kernel events can be minimized using arena allocators~\cite{Arena}.
 
-\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators in multi core simulation kernels and arena-style allocators for single core simulation.
+\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators for \textit{Cx} and \textit{Ox} synchronization, and arena-style allocators for \textit{Nx} synchronization.
 The latter can be faster, but at the cost of additional configuration.
@@
     \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
-    \caption{Effect of memory allocators on multi core execution time, using optimistic synchronization.}
+    \caption{Effect of memory allocators with \textit{ON} synchronization.}
     \label{fig:memallocators_parallel}
@@
 
-We of course still need to guarantee determinism and isolation between the calls to the RNG, as well as avoiding excessive synchronization.
+We still need to guarantee determinism and isolation between the calls to the RNG, as well as avoiding excessive synchronization.
 \textit{Dxex} uses the Tina RNG collection (TRNG)~\cite{PhysRevE.75.066701} as an alternative random number generator with performance and multithreading in mind.
+%TODO one RNG per kernel or thread in new terminology?
 Since the RNG is an implicit part of the state in the \textsf{Parallel DEVS} formalism, though often not implemented as such, we evaluated performance for both approaches: one global RNG per thread, and one RNG per atomic \textsf{DEVS} model.
@@
 This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
-\textit{Dxex}'s single core and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
-The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.
+Only \textit{Ox} synchronization seems affected, as it needs to copy more bytes in every transition due to state saving.
+No additional copies need to happen in \textit{Nx} or \textit{Cx} synchronization.
 
-Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in single core simulation is three times faster using any of the three alternatives, only the STL RNG is slow.
-For multi core simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between single core and multi core simulation.
-Conservative synchronization is almost unaffected by changing the RNG.
+Figure~\ref{fig:Queuerngspeedup} shows that, for \textit{NN} synchronization, any of the three alternatives is three times faster than STL RNG.
+For \textit{Cx} and \textit{Ox} synchronization, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap with \textit{Nx} synchronization.
+\textit{Cx} synchronization is almost unaffected by changing the RNG.
 

### 2016-12-07 publications/Journal/4-performance.tex
Rewrote first part of the paper to the new synchronization terminology

--- 2016-12-06 publications/Journal/4-performance.tex
@@
 In this section, we evaluate the performance of different synchronization protocols in \textit{dxex}.
-We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVStoneJournal,DEVSSurvey}, to show that our modularity does not impede performance.
-CPU time and memory usage is compared for both single core and multi core simulation.
+We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation tools~\cite{DEVStoneJournal,DEVSSurvey}, to show that our modularity does not impede performance.
+CPU time and memory usage is compared for different synchronization protocols.
 
-We start off with a comparison of single core simulation, to show how \textit{adevs} and \textit{dxex} relate in this simple case.
-For the multi core simulation benchmarks, results are presented for both conservative and optimistic synchronization.
-We conclude with a comparison between inter and intra kernel parallelization and give use cases where these protocols can complement each other in the context of recent theoretical work ~\cite{amdahlpdevs}.
+We start with a comparison of \textit{NN} synchronization, to show how \textit{adevs} and \textit{dxex} relate in this simple case.
+Afterwards, we compare inter-kernel synchronization protocols, including a comparison with \textit{adevs} again.
+Inter-kernel synchronization protocols are then compared in combination with intra-kernel synchronization protocols, in the context of recent theoretical work~\cite{amdahlpdevs}.
 
@@
 \subsection{Benchmark Models}
-We use three different benchmark models, covering different aspects of the simulation kernel.
+We use three different benchmark models, covering different aspects of the simulation tool.
 
@@
 It takes two parameters: the width and depth of the hierarchy.
-This benchmark shows how the complexity of the simulation kernel behaves for an increasing amount of atomic models, and an increasingly deep hierarchy.
+This benchmark reveals the complexity of the simulation kernel for an increasing amount of atomic models and an increasingly deep hierarchy.
 An example for a width and depth of 2 is shown in Figure~\ref{fig:queue_model}.
@@
 Second, the \textit{Interconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
-Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event (\textit{i.e.}, the event is broadcasted).
-The model takes one parameter: the number of models.
+Similar to PHOLD, all models are connected to one another, but all through the same port: every atomic model receives each generated event (\textit{i.e.}, the event is broadcasted).
+The model takes one parameter: the number of atomic models.
 This benchmark shows the complexity of event creation, event routing, and simultaneous event handling.
-An example for three models is shown in Figure~\ref{fig:interconnect_model}.
+An example for three atomic models is shown in Figure~\ref{fig:interconnect_model}.
 
-Third, the \textit{PHOLD} model~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
+Third, the \textit{PHOLD} model~\cite{PHOLD}, creates $n$ atomic models, where each atomic model has exactly $n-1$ output ports.
 Each atomic model is directly connected to every other atomic model.
 After a random delay, an atomic model sends out an event to a randomly selected output port.
-Output port selection happens in two phases: first it is decided whether the event should be sent within the same node, or outside of the node.
-Afterwards, a uniform selection is made between the remaining models.
-The model takes two parameters: the percentage of remote events (determining the percentage of messages routed to other nodes), and the percentage of high-priority events.
+Output port selection happens in two phases: first it is decided whether the event should be sent within the same kernel, or outside of the kernel.
+Afterwards, a uniform selection is made between the possible atomic models.
+The model takes two parameters: the percentage of remote events (determining the percentage of events routed to other kernels), and the percentage of high-priority events.
 High-priority events are events generated in a very short time after the previous event.
 This benchmark shows how the simulation kernel behaves in the presence of many local or remote events, in combination with a varying percentage of high-priority events.
-An example for four models, split over two nodes, is shown in Figure~\ref{fig:PHOLD_model}.
+An example for four atomic models, split over two kernels, is shown in Figure~\ref{fig:PHOLD_model}.
 
@@
 
-\subsection{Signle Core Simulation}
-We start by evaluating single core simulation performance, in order to obtain a baseline for our comparison of multi core simulation performance.
+\subsection{Single kernel (\textit{NN} Synchronization)}
+We start by evaluating \textit{NN} synchronization performance, in order to obtain a baseline for our comparison of other synchronization protocols.
 
@@
 For the first benchmark, we tested the effect of hierarchical complexity of the model in the performance of the simulator.
-A set of three tests was performed, where each test has the same number of models but an increasing depth.
+A set of three tests was performed, where each test has the same number of atomic models but an increasing depth.
 The results can be seen in Figure~\ref{fig:Queue_benchmark_seq}.
 Since \textit{dxex} performs direct connection~\cite{SymbolicFlattening} on the model, there is no performance hit when the depth is increased.
-Direct connection only needs to be done at initialization, so it is a one time cost that is negligible for long running simulations.
+Direct connection only needs to be done at initialization, making it a neglible one time cost for long running simulations.
 \textit{Adevs}, on the other hand, suffers from the increased depth, even though some similar (but not identical) optimization to event passing was made~\cite{adevs_opt}.
@@
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_sequential.eps}
-	\caption{Queue benchmark results for single core simulation.}
+	\caption{Queue benchmark results with \textit{NN} synchronization.}
 	\label{fig:Queue_benchmark_seq}
@@
 	\includegraphics[width=\columnwidth]{fig/interconnect_sequential.eps}
-	\caption{Interconnect benchmark results for single core simulation.}
+	\caption{Interconnect benchmark results with \textit{NN} synchronization.}
 	\label{fig:Interconnect_benchmark}
@@
 The biggest difference is that the amount of messages sent is much lower.
-The number of events scales linear with the number of models, not quadratic.
+The number of events scales linear with the number of atomic models, not quadratic.
 Figure~\ref{fig:Phold_benchmark} shows that in terms of performance \textit{dxex} and \textit{adevs} are very close to each other, with \textit{adevs} slightly outperforming \textit{dxex}.
@@
 	\includegraphics[width=\columnwidth]{fig/phold_sequential.eps}
-	\caption{PHold benchmark results for single core simulation.}
+	\caption{PHold benchmark results with \textit{NN} synchronization.}
 	\label{fig:Phold_benchmark}
@@
 
-\subsection{Parallel Simulation}
-We now continue by describing our multi core simulation performance for different synchronization protocols, and compared to \textit{adevs}.
-The speedup of \textit{adevs} is computed with the corresponding \textit{dxex} single core benchmark.
-This was done to take into account the performance difference observed in single core simulation.
-As such, the highest speedup indicates the fastest results among all tools, independent of single core simulation results.
+\subsection{Inter-Kernel Parallelism (\textit{CN} and \textit{ON} synchronization)}
+We now continue by describing our inter-kernel parallelims performance for different synchronization protocols, and compared to \textit{adevs}.
+The speedup of \textit{adevs} is computed with the corresponding \textit{dxex} \textit{NN} synchronization benchmark.
+This was done to take into account the performance difference observed in \textit{NN} synchronization.
+As such, the highest speedup indicates the fastest results among all tools, independent of \textit{NN} synchronization results for \textit{adevs}.
 
 \subsubsection{Queue}
-The Queue model is one single chain of models, resembling a pipeline.
-This structure can be exploited to prevent cyclic dependencies in the multi core simulation.
+The Queue model is one single chain of atomic models, resembling a pipeline.
+This structure can be exploited to prevent cyclic dependencies in \textit{CN} and \textit{ON} synchronization.
 
-Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to single core simulation for a fixed problem size (\textit{i.e.}, strong scaling).
-As the number of kernels increases, the optimistic protocol quickly becomes the worst choice.
-This is mainly caused by the pipeline structure of the model: the last models in the queue only respond to incoming messages and therefore have to be rolled back frequently.
-The difference between \textit{dxex} conservative protocol and \textit{adevs} becomes smaller when more and more cores are used.
-The same effect can be seen when the problem size is increased in tandem with the number of used cores (\textit{i.e.}, weak scaling) in Figure~\ref{fig:Queue_plot_weak}.
+Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to \textit{NN} synchronization for a fixed problem size, with an increasing number of used CPU cores (\textit{i.e.}, strong scaling).
+As the number of cores increases, \textit{ON} quickly becomes the worst choice.
+This is mainly caused by the pipeline structure of the model: the last atomic models in the queue only respond to incoming messages and therefore have to be rolled back frequently.
+The difference between \textit{dxex} \textit{CN} and \textit{adevs} \textit{CN} becomes smaller when more and more cores are used.
+The same effect can be seen when the problem size is increased in tandem with the number of used CPU cores (\textit{i.e.}, weak scaling) in Figure~\ref{fig:Queue_plot_weak}.
 
@@
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_strong_speedup.eps}
-	\caption{Queue model strong scaling speedup compared to \textit{dxex} single core.}
+	\caption{Queue model strong scaling speedup compared to \textit{dxex} \textit{NN} synchronization.}
 	\label{fig:Queue_plot_strong}
@@
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_weak_speedup.eps}
-	\caption{Queue model weak scaling speedup compared to \textit{dxex} single core.}
+	\caption{Queue model weak scaling speedup compared to \textit{dxex} \textit{NN} synchronization.}
 	\label{fig:Queue_plot_weak}
@@
 In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
-The number of models is now kept constant at eight.
+The number of atomic models is now kept constant at eight.
 Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.

### 2016-12-07 publications/Journal/abstract.tex
Rewrote first part of the paper to the new synchronization terminology

@@
 The simulator not only supports both protocols but also has the capability to switch between them at runtime.
-The simulator complements each of the synchronization protocols with the PDEVS protocol.
+The simulator can combine each synchronization protocols with either a threaded or sequential implementation of the PDEVS protocol.
 We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol.
@@
 We further allow for an external component to gather simulation statistics, on which runtime switching between the different synchronization protocols can be based.
-Model allocation is also studied to see how our conservative and optimistic synchronization protocols are affected by good and bad allocations.
+The effects of allocation on our synchronization protocols is also studied.

### 2016-12-08 publications/Journal/4-performance.tex
Made whole paper (more) consistent

@@
 \subsection{Inter-Kernel Parallelism (\textit{CN} and \textit{ON} synchronization)}
-We now continue by describing our inter-kernel parallelims performance for different synchronization protocols, and compared to \textit{adevs}.
+We now continue by describing our inter-kernel parallelims performance for different synchronization protocols, compared to \textit{adevs}.
 The speedup of \textit{adevs} is computed with the corresponding \textit{dxex} \textit{NN} synchronization benchmark.
@@
 As such, the highest speedup indicates the fastest results among all tools, independent of \textit{NN} synchronization results for \textit{adevs}.
+We only compare \textit{xN} results, as we are now only interested in the differences between the various inter-kernel synchronization protocols, and have thus disabled all other forms of parallelism.
+These results are generalized to conservative and optimistic synchronization in general.
+The comparison between \textit{xN} and \textit{xP} is made later on.
 
@@
 \label{subsec:parallelinterconnect}
-In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
+In the Interconnect model, we determine how broadcast communication is supported across kernels.
 The number of atomic models is now kept constant at eight.
 Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
-When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and the increasing number of rollbacks in optimistic simulation.
-All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in a multi core setting.
+When the number of kernels increases, performance decreases due to increasing contention in conservative simulation and the increasing number of rollbacks in optimistic simulation.
+All atomic models depend on each other and have no computational load whatsoever, negating any possible performance gain by splitting up the work over multiple kernels.
 
@@
 	\includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
-    %TODO cyclic allocation in title?
-	\caption{Interconnect benchmark results for multi core simulation.}
+	\caption{Interconnect benchmark results for \textit{ON} and \textit{CN} synchronization.}
 	\label{fig:interconnect_benchmark_parallel}
@@
 In the PHOLD model, we first investigate the influence of the percentage of remote events on the speedup.
-A remote event in this context is an event that is sent from a model on one kernel to a model on another simulation kernel.
+A remote event in this context is an event that is sent from an atomic model on one kernel to an atomic model on another kernel.
 When remote events are rare, optimistic synchronization rarely has to roll back, thus increasing performance.
@@
 Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block and wait.
-Even though a single synchronization protocol is always ideal in this case, it already shows that different synchronization protocols respond differently to a changing model.
+Even though a single synchronization protocol is ideal throughout the whole simulation run, it shows that different synchronization protocols respond very differently to a changing model.
 
-\textit{Adevs} is significantly slower during conservative synchronization.
+\textit{Adevs} is significantly slower for \textit{CN} synchronization.
 Analysis of profiling callgraphs shows that exception handling in \textit{adevs} is the main cause. 
@@
 When many high-priority events occur, optimistic synchronization is penalized due to its unconditional progression of simulation, which frequently needs to be rolled back.
-Results show that there is no single perfect synchronization algorithm for this kind of models: depending on configuration, either synchronization protocol might be better.
+Results show that there is no single perfect synchronization algorithm for this kind of model: depending on configuration, either synchronization protocol might be better.
 
-\subsection{Comparison to the abstract simulator}
-The abstract simulator of \textsf{Parallel DEVS} included its own notion of parallelism: transition functions that happen at the same point in simulated time can be trivially parallelized.
-In this work we refer to this approach as \pSim so as not to confuse with the PDEVS formalism.
-This parallelization is trivial, in the sense that there is no need for synchronization protocols.
-The parallelization of the transition functions (from now on called the ``inherent parallelism'') is, at least in \textit{dxex}, independent of the use of a synchronization protocol.
-That is, both can be used simultaneously.
-Following the theoretical analysis published in~\cite{amdahlpdevs}, a comparison between this inherent parallelism and the use of synchronization protocols is warranted.
+\subsection{Intra-Kernel Parallelism (\textit{xP} synchronization)}
+The abstract simulator of \textsf{Parallel DEVS} included its own notion of parallelism, which we refered to as \pSim, and have implemented as \textit{xP} synchronization.
+The \textit{xP} synchronization protocol is trivial, as it merely executes a set of independent functions concurrently.
+As indicated before, intra-kernel synchronization is independent of inter-kernel synchronization in \textit{dxex}.
+That is, all six combinations are possible.
+Following the theoretical analysis published in~\cite{amdahlpdevs}, a comparison is warranted between \textit{xN} and \textit{xP} synchronization.
 
@@
 We have opted to use the \textit{Queue} model for this comparison, as it allows for many interesting configurations.
-%TODO what is meant with this allocation of threads and synchronized kernels?
-% 4 Kernels running on a dedicated thread, with per kernel 4 threads allocated to use for PDEVS.
-The effect of combining 4 synchronized kernels, with each 4 threads for \pSim, is contrasted with 16 threads for \pSim and 16 kernels for conservative and optimistic.
-The first scenario runs each kernel on a dedicated thread synchronized with the other 3 kernels.
-Each kernel can use up to 4 threads for \pSim.
-In the second scenario \pSim can use up to 16 threads while running in a single core kernel.
-Finally 16 synchronized kernels are used with one thread per kernel.
-This allows us to observe which is more efficient in obtaining a speedup.
-We simulate a computational load by an optional call to sleep in the transition functions set at 5 milliseconds.
-The model is configured with depth 4 and width 300 if the transition function has no load, and width 50 if the computational load is active.
-In our configuration, an imminent model always generates output, resulting in the receiving model becoming imminent.
-The probability that an internal transition in a model generates output and is connected to a receiving model here is $1$.
-The probability that a model becomes immediately imminent depends on whether fixed or random timeadvance is used.
-In the case of fixed timeadvance, a model will always become immediately imminent.
-When random timeadvance is used, this probability is $\frac{1}{n}$, with $n$ denoting the total amount of atomic models.
-These probabilities correspond with the $q$ parameter in the theoretical analysis ~\cite{amdahlpdevs}.
-The probability that a model is imminent (i.e.
-is scheduled to execute an internal transition) at any given point in time in this model is equal to $1$ when a fixed timeadvance is used and $\frac{1}{n}$ when a random time advance is used.
-These probabilities correspond with the $p$ parameter in the theoretical analysis.
+We have three different configurations, each using 16 threads in total, to be ran on a CPU with 16 CPU cores:
+\textit{NP} synchronization uses one kernel with 16 threads each;
+\textit{OP} and \textit{CP} synchronization use 4 kernels with 4 threads each; and
+\textit{ON} and \textit{CN} synchronization use 16 kernels with 1 thread each.
+Additionaly, \textit{NN} synchronization only uses one kernel and one thread, as it is completely sequential.
+All speedup results are shown in comparison to \textit{NN} synchronization.
+Atomic models are always equally distributed over the available kernels.
 
-The benchmarks are run sufficiently long enough to guarantee that the frequency of internal and external transitions is equal within a benchmark, regardless of the randomness of the timeadvance.
-In our model each model will be executing internal and external transitions, creating an ideal use case to evaluate the speedup inherent parallelism can obtain.
+This allows us to observe which is more efficient in obtaining a speedup, each with the same number of CPU cores available.
+We simulate a computational load by a sleep of 5 milliseconds.
+The model is configured with depth 4 and width 300 if the transition function has no load (\textit{i.e.}, no sleep), and width 50 if the computational load is active (\textit{i.e.}, sleeps for 5 milliseconds).
+In our configuration, an imminent atomic model always generates output, resulting in the receiving atomic model becoming imminent.
+The probability that an internal transition in an atomic model generates output and is connected to a receiving atomic model is $1$ ($q = 1$~\cite{amdahlpdevs}).
+The probability that an atomic model becomes immediately imminent depends on whether fixed or random time advance is used ($p$~\cite{amdahlpdevs}).
+In the case of a fixed time advance, a model will always become immediately imminent ($p = 1$).
+When random time advance is used, this probability is $\frac{1}{n}$, with $n$ denoting the total amount of atomic models, as only one atomic model becomes active ($p = \frac{1}{n}$).
+
+The benchmarks are run sufficiently long enough to guarantee that the frequency of internal and external transitions is equal within a benchmark, regardless of the randomness of the time advance.
+In our model each atomic model executes internal and external transitions, creating an ideal use case to evaluate the speedup \textit{xP} synchronization can obtain. 
 The key difference is that although the event frequency is the same, their concurrency is not. 
-This benchmark combines both possible sources of parallelization : concurrent internal and external transitions. 
-Models are equally distributed over the kernels and between threads in \pSim.
 The coefficient of variation of our results is less than $1\%$.
@@
 We defer the discussion on which of these two is the most realistic, as this depends on the problem domain.
-For example, in a simulation with a fixed timestep (\textit{e.g.}, cell-based models, discretization of continuous model), transition functions will often occur simultaneously.
+For example, in a simulation with a fixed timestep (\textit{e.g.}, cell-based models, discretization of continuous model), transition functions often occur simultaneously.
 Conversely, simulations with an arbitrary timestep (\textit{e.g.}, many independent systems communicating together) have very few simultaneous events.
 
-\paragraph{Concurrent events}
+\paragraph{Concurrent events ($q = 1; p = 1$)}
 First we create a model where all transitions happen simultaneously, with a significant computational load in the transition functions.
-In Figure~\ref{fig:psim_plot_fixed_sleep}, we observe that exploiting only the inherent parallelism, results in a speedup of about 10.  
-Simulation with conservative and optimistic synchronization is run first with 16 kernels, then with 4 kernels having 4 \pSim threads each. 
-Both synchronization protocols achieve a speedup similar to only exploiting the inherent parallelism.
-Combining the protocols with the inherent parallelism gives near identical results.
+In Figure~\ref{fig:psim_plot_fixed_sleep}, we observe that all synchronization protocols result in a speedup of about 10.  
 In this scenario there is no real advantage between the different parallel configurations.
+Note, however, that \textit{NP} synchronization is trivial to implement, whereas \textit{Ox} and \textit{Cx} synchronization are much more difficult to implement.
 
@@
 	\includegraphics[width=\columnwidth]{fig/pdevs_fixed_sleep.eps}
-	\caption{Queue speedup benchmark for PDEVS and synchronization with maximal concurrent events and significant computational load}
+	\caption{Queue speedup benchmark with $p = 1$, $q = 1$, and significant computational load}
 	\label{fig:psim_plot_fixed_sleep}
@@
 
-\paragraph{Random events}
-Now we randomize the time advance in the models, resulting in very few transition functions that happen simultaneously.
+\paragraph{Random events ($q = 1; p = \frac{1}{n}$)}
+Now we randomize the time advance in the atomic models, resulting in very few transition functions that happen simultaneously.
 Even when two transitions are only minimally apart in simulated time, they cannot be executed in parallel, as there might otherwise be a causality error.
 The transition function has the same computational load as in the previous configuration. 
-The results in Figure~\ref{fig:psim_plot_random_sleep} show that parallelizing transition functions only, adds little overhead. 
-Nonetheless, performance is not increased at all: the chances of executing a few transition functions in parallel does not make up for the overhead of checking.
-In combination with either synchronization protocol, a minor improvement can still be seen, but this is primarily caused by the synchronization protocol, and not by parallelizing the transition functions.
+Results are shown in Figure~\ref{fig:psim_plot_random_sleep}.
+We observe that \textit{NP} synchronization adds little overhead, but doesn't achieve any significant speedup either.
+With \textit{ON} and \textit{CN} synchronization, we again achieve high speedups.
+This is not the case with \textit{OP} and \textit{CP} synchronization, as we only have four kernels available for inter-kernel synchronization.
+The four threads per kernel, in this case, are not used fully as only two transitions occur simultaneously at all times.
 
@@
 	\includegraphics[width=\columnwidth]{fig/pdevs_random_sleep.eps}
-	\caption{Queue speedup benchmark for PDEVS and synchronization with randomized events and significant computational load}
+	\caption{Queue speedup benchmark with $p = \frac{1}{n}$, $q = 1$, and significant computational load}
 	\label{fig:psim_plot_random_sleep}
@@
 Finally, we remove the computation load from the transition function, in combination with many concurrent events.
-Figure~\ref{fig:psim_plot_no_sleep} shows that in single core simulation, the overhead of thread management and shared memory communication is crippling for performance.
-Even though many events occur concurrently, the actual computation being done in the transition function does not outweigh the overhead, resulting in much slower simulation than single core execution of transition functions.
-When combined with synchronization protocols, performance is still reasonable, though the use of the inherent parallelism is not beneficial either way.
-Both synchronization protocols achieve a reasonable speedup without \pSim, as the overhead of thread management (creation and destruction) is avoided.
-These results are similar to~\cite{Himmelspach}.
+Figure~\ref{fig:psim_plot_no_sleep} shows that for \textit{NP} synchronization, the overhead of thread management and shared memory communication is crippling for performance.
+Even though many events occur concurrently, the computation in the transition function does not outweigh the overhead.
+This results in much slower simulation than \textit{NN} synchronization.
+Even \textit{OP} and \textit{CP} synchronization are unable to achieve any performance increase, whereas \textit{ON} and \textit{CN} synchronization increase performance marginally.
+This is because the overhead of thread management is avoided, similar to the results obtained by Himmelspach \textit{et al.}~\cite{Himmelspach}
 
@@
 \paragraph{Discussion}
-In \textit{dxex}, use of the inherent parallelism found in \textsf{Parallel DEVS} can be combined with any of the supported synchronization protocols.
-The user is not forced to choose between either: the capabilities of the computer can be divided over both approaches.
-While exploitation of the inherent parallelism can offer a significant speedup, a high number of transitions must be triggered simultaneously throughout the whole simulation.
+In \textit{dxex}, any inter-kernel synchronization protocol can be combined with any intra-kernel synchronization protocol.
+While \textit{xP} synchronization can offer a significant speedup at a trivial implementation cost, $p$ must be high throughout the whole simulation.
 Additionally, the computational load of the transition functions should be high enough to warrant the thread management overhead.
-We conclude that both approaches have their distinct advantages and disadvantages.
-The inherent parallelism of \textsf{Parallel DEVS} avoids synchronization overhead and is trivial to implement.
-It is not influenced by the coupling of models, only by the number of simultaneous transition functions and the computational load in the transition function.
-In contrast, either of the two synchronization protocols is heavily influenced by the coupling of the models, but not by the computational load or number of simultaneous transitions.
+We conclude that each synchronization protocol has its distinct advantages and disadvantages:
+inter-kernel synchronization protocols depend on the coupling of models and are difficult to implement, whereas intra-kernel synchronization protocols depend on the number of concurrent transitions and are trivial to implement.
 
@@
 Having insufficient memory may cause sudden deterioration in performance, even to the point of making the simulation infeasible.
-We therefore also investigate the memory usage of different synchronization protocols, and again compare to \textit{adevs}.
+We therefore also investigate the memory usage of different synchronization protocols, again comparing to \textit{adevs}.
 
 We do not tackle the problem of states that become too large for a single machine to hold.
-This problem can be mitigated by distribution over multiple machines, which neither \textit{dxex} nor \textit{adevs} support.
+This problem can be mitigated by distributing the state over multiple machines, which neither \textit{dxex} nor \textit{adevs} support.
 
@@
 Due to our motivation for memory usage analysis, we will only measure peak allocation in maximum resident set size as reported by the OS.
+We only show results for \textit{xN} synchronization, as \textit{xP} synchronization has no significant additional memory requirements for a reasonable number of threads.
 
@@
 
-Unsurprisingly, optimistic synchronization results show very high memory usage due to the saved states.
+Unsurprisingly, \textit{Ox} synchronization results show very high memory usage due to the saved states.
 Note the logarithmic scale that was used for this reason.
@@
 
-Conservative simulation always uses more memory than single core simulation, as is to be expected.
-Additional memory is required for the multiple threads, but also to store all events that are processed simultaneously.
+\textit{Cx} synchronization always uses more memory than \textit{Nx} synchronization, as is to be expected.
+Additional memory is required for the multiple kernels, but also to store all events that are processed simultaneously.
 
@@
 We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
-The use of synchronization protocols does also not prevent the exploitation of the inherent parallelism found in \textsf{ParallelPDEVS}.
-Both approaches are useful in specific model configurations.
+Each synchronization protocol has its own specific kind of models for which it is the best one.
 But even with the right synchronization protocol, we have seen that two problems remain.
 
-First, although one of both synchronization protocols might be ideally suited for specific model behaviour, nothing guarantees that the model will exhibit the same behaviour throughout the simulation.
-Similarly to the polymorphic scheduler~\cite{MasterThesis}, we wish to make it possible for the ideal option to be switched during simulation.
+First, although a synchronization protocols might be ideally suited for specific model behaviour, nothing guarantees that the model will exhibit the same behaviour throughout the simulation.
+Similarly to the polymorphic scheduler~\cite{MasterThesis}, we wish to make it possible for the ideal synchronization protocol to be switched during simulation.
 When changes to the model behaviour are noticed, the used synchronization protocol can be switched as well.
@@
 Second, the allocation of models is nontrivial and has a significant impact on performance.
-While our multi core speedup for the Queue model, for example, was rather high, this is mostly due to characteristics of the model: the dependency graph does not contain any cycles.
+While speedup for the Queue model, for example, was rather high in most cases, this is mostly due to characteristics of the model: the dependency graph does not contain any cycles.
 When cycles were introduced, as in the Interconnect model, performance became disastrous.

### 2016-12-08 publications/Journal/5-hotswap.tex
Made whole paper (more) consistent

--- 2016-10-03 publications/Journal/5-hotswap.tex
@@
 In \textit{dxex}, we not only make it possible to define the synchronization protocol to use, but also to change this decision throughout simulation.
-To do this, all kernels are notified of the switch and they are forced to stop simulation.
-When stopped, each kernel instantiates a new core, with the new synchronization protocol, that is provided with the simulation state of the previous core.
-Simulation is then resumed with the new cores after the previous ones are destroyed.
+To switch between intra-kernel synchronization protocols, we only have to execute all concurrent transitions sequentially.
+This case is trivial, as it just flips a boolean, and is thus not further considered in this section.
+To switch between inter-kernel synchronization protocols, all kernels are notified of the switch and they are forced to stop simulation.
+When stopped, each kernel instantiates a new kernel, with the new synchronization protocol, that is provided with the simulation state of the previous kernel.
+Simulation is then resumed with the new kernels after the previous ones are destroyed.
 
 As usual, switching imposes an overhead and should thus only be done if the benefits outweigh the induced overhead.
-This overhead depends on the size of the model and the number of simulation cores.
-For a simple model and a few cores, the overhead is less than a second.
+This overhead depends on the size of the model and the number of kernels.
+For a simple model and a few kernels, the overhead is less than a second. 
+Creating new kernels and moving the simulation models has an overhead linear in the amount of kernels and atomic models.
+The time required to synchronize and halt the existing kernels is variable, especially if the old synchronization protocol is optimistic since there is no real limit on the virtual time difference between kernels.
+Given that the existing kernels are equally loaded, this time difference will in practice scale linearly with the number of kernels.
 
 Although we currently only support manual switches between different synchronization protocols, this is not necessarily the case.
-Ideally, a new component is added to the simulation kernel, which monitors model behaviour and simulation performance, and toggles between them automatically.
+Ideally, a new component is added to the kernel, which monitors model behaviour and simulation performance, and toggles between them automatically.
 Our interface is augmented with the necessary bindings for such a decision component.
@@
 Traditionally, models are not exposed to simulation kernel details as they work at a different level of abstraction:
-simulation models only care about simulation results, and not about how these results are obtained (\textit{e.g.}, through parallel or distributed simulation).
-This is different for a new simulation kernel component that has to monitor the behaviour of not only the model, but the simulator as well.
+modellers only care about simulation results, and not about how these results are obtained (\textit{e.g.}, through parallel or distributed simulation).
+This is different for a new kernel component that has to monitor the behaviour of not only the model, but the simulator as well.
 
-We add performance metrics in the simulation kernel, which logs relevant performance metrics and processes them for use in other components.
-These metrics include the number of events created and destroyed, the number of inter- and intra-kernel events, the number of rollbacks, the measured lookahead, details of the GVT and EOT calculations, and information on the fairness between simulation kernels.
-With all these metrics, the decision component can get a global view on both model and simulation kernel behaviour.
+We add performance metrics in the kernel, which logs relevant performance metrics and processes them for use in other components.
+These metrics include the number of events created and destroyed, the number of inter- and intra-kernel events, the number of rollbacks, the measured lookahead, details of the GVT and EOT calculations, and information on the fairness between kernels.
+With all these metrics, the decision component can get a global view on both model and kernel behaviour.
 

### 2016-12-08 publications/Journal/6-allocation.tex
Review section 6
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
Made whole paper (more) consistent

--- 2016-10-03 publications/Journal/6-allocation.tex
@@
 This brings us to the topic of model allocation, as also implemented in \textit{PythonPDEVS}~\cite{PythonPDEVS2}.
+In this section we focus on the inter kernel synchronization protocols, as intra kernel synchronization is unaffected by the inter kernel topology.
 
@@
 If no preference is given, a simple striping scheme is used but this is often insufficient.
-By overriding the default allocator, a modeller tunes the allocation scheme for a specific model, maximizing parallel speedup.
+By overriding the default allocator, a modeller tunes the allocation scheme for a specific model.
 This interface can be linked to graph algorithms for automatic allocation scheme generation, for example to avoid cycles in the dependency graph.
@@
 Unlike the Queue model, the width of the hierarchy is still present in the topology of the atomic models after direct connection.
-The PHOLDTree model allows us to investigate parallel speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.
+The PHOLDTree model allows us to investigate speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.
 
@@
 
-PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain, as was the case in the Queue model.
+PHoldtree, like Queue, is highly hierarchical, but its flattened structure cannot be partitioned into a chain, as was the case in the Queue model.
 This topology is interesting since it highlights the effects of allocation.
-First, we evaluate the model in sequential simulation to provide a baseline for parallel simulation.
+First, we evaluate the model with \textit{NN} synchronization to provide a baseline for further results.
 
-\subsubsection{Sequential Simulation}
+\subsubsection{No Inter-Kernel Synchronization (\textit{NN})}
 Since \textit{adevs} does not use direct connection, we expect a noticeable performance difference between \textit{dxex} and \textit{adevs}.
@@
 \textit{Dxex} uses direct connection, making it independent of fanout.
-Performance of \textit{dxex} is, in this model, only dependent on the number of models.
+Performance of \textit{dxex} is, for this model, only dependent on the number of models.
 Slight deviations can still be seen, though, caused by the initialization overhead of direct connection.
@@
     \includegraphics[width=\columnwidth]{fig/pholdtree_sequential_dn.eps}
-    \caption{Effect of hierarchy in sequential simulation.}
+    \caption{Effect of hierarchy with \textit{NN} synchronization.}
     \label{fig:PHOLDtree_seq_dn_benchmark}
@@
 
-\subsubsection{Parallel Simulation}
-Next, we run the model using two different model allocation schemes: breadth-first and depth-first.
-But first, we explain what we mean by both allocation schemes.
+\subsubsection{Inter-Kernel Synchronization (\textit{CN} and \textit{ON})}
+Next, we run the model using two different allocation schemes: breadth-first and depth-first.
+First, we explain both allocation schemes.
 
-With breadth-first allocation, we traverse the tree in a breadth-first way, allocating subsequently visited atomic models to the same node.
-This means, intuitively, that atomic models at the same level in the tree, but not necessarily siblings, are frequently allocated to the same node.
+Breadth-first allocation traverses the tree in a breadth-first way, allocating subsequently visited atomic models to the same node.
+Intuitively, atomic models at the same level in the tree, but not necessarily siblings, are frequently allocated to the same node.
 Since there is only infrequently some communication between siblings, and even never between different subtrees, this does not sound an intuitive allocation.
@@
 
-With depth-first allocation, we traverse the tree in a depth-first way, allocating subsequently visited atomic models to the same node.
-This means, intuitively, that subtreees are frequently allocated to the same node.
-This allocation strategy is shown in Figure~\ref{fig:PholdTree_model_dfs}.
+Depth-first allocation traverses the tree in a depth-first way, allocating subsequently visited atomic models to the same node.
+Intuitively, subtrees are frequently allocated to the same node, as shown in Figure~\ref{fig:PholdTree_model_dfs}.
 
@@
 Both allocators will try to divide models evenly over kernels.
-The effects of varying the number of models per kernel are already evaluated in the previous section on scaling.
+The effects of varying the number of atomic models per kernel are already evaluated in the previous section on scaling.
 Here we want to highlight the overhead of communication and inter-kernel dependence.
@@
 The breadth first allocation scheme results in a dependency chain with multiple branches, much like in the Queue model.
-Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model.
-This is not always true though: a single kernel with an unbalanced number of atomic models or unequal computational load in transition functions slows down the remainder of the chain.
-This effect is also apparent if the thread a kernel runs on is not fairly scheduled.
+Such a linear dependency chain can result in a higher speedup, as demonstrated with the Queue model.
+This is not always true though: a single kernel with an unbalanced computational load slows down the remainder of the chain.
+This effect is also apparent if the thread a kernel runs on is not fairly scheduled by the operating system.
 With conservative synchronization this leads to excessive polling of the EOT of the other kernels.
@@
 This is caused by the high number of inter-kernel connections and the high number of events exchanged over these connections.
-The number of connections between nodes at the same simulation kernel is also rather low.
+The number of connections between atomic models at the same simulation kernel is also rather low.
 Using a depth-first allocation scheme, however, as shown in Figure~\ref{fig:pholdtree_visualize_parDFS}, minimizes inter-kernel connections while maximizing intra-kernel connections.
@@
 We see that for both synchronization protocols, the depth-first allocation is significantly better than breadth-first allocation.
-This is what we expected for this model: depth-first allocation maintains locality better than breadth-first allocation.
+This is what we expected for this model: depth-first allocation preserves locality better than breadth-first allocation.
 Whereas this is the case in this example, this is not true in general, as the ideal allocation depends on the model being simulated.
@@
 The most prominent aspect of these results is the low performance for conservative depth-first allocation for two kernels.
-This is mostly caused by the difference between a sequential simulation and a parallel simulation: suddenly we need to take into account synchronization and passing around of lookahead values.
-And since the number of kernels is low, the overhead dominates any further speedup.
-Optimistic is less sensitive to the number of models per kernel as it does not need to poll each model for a lookahead, this explains the lower runtime penalty observed for optimistic.
+This is mostly caused by the introduction of synchronization protocols: suddenly we need to take into account other kernels and passing around of lookahead values.
+And since the number of kernels is low, the overhead dominates.
+Optimistic is less sensitive to the number of atomic models per kernel as it does not need to poll each model for a lookahead, this explains the lower runtime penalty observed for optimistic.
 
-%The effect is due to the topology of the kernels, in BFS the risk on reverts is higher. Optimistic is less influenced but will get a lower speedup in the good case.
 Interestingly, we see that optimistic synchronization is less influenced by the allocation than conservative synchronization.

### 2016-12-08 publications/Journal/7-related-work.tex
Made whole paper (more) consistent

@@
 In terms of code design and philosophy, \textit{dxex} is most related to \textit{PythonPDEVS}~\cite{PythonPDEVS}.
-Performance of \textit{PythonPDEVS} was still decent through the use of ``hints'' from the modeler.
+Performance of \textit{PythonPDEVS} was decent through the use of ``hints'' from the modeler.
 In this spirit, we offer users the possibility to choose between different synchronization protocols.
@@
 Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though none have done so in a strictly modular way that allows straightforward protocol switching for a single given model.
-For example, \textit{adevs} only supports conservative synchronization, and \textit{vle} only supports experiment-level parallelism (\textit{i.e.}, run multiple experiments in parallel).
+For example, \textit{adevs} only supports conservative synchronization, and \textit{vle} only supports experiment-level parallelism (\textit{i.e.}, run independent experiments in parallel).
 Closest to our support for multiple synchronization protocols is \textit{CD++}~\cite{CD++}.
@@
 A new architecture for sequential PDEVS simulation has been introduced in \cite{SeqPDEVSArch} with promising performance results.
-This algorithm achieves a speedup by using the inherent parallelism in PDEVS without the need for synchronization.
-By limiting our approach to a pure parallel simulation, we are able to achieve higher speedups through the use of shared memory communication. 
+This algorithm achieves a speedup by using \textit{xP} synchronization.
+By limiting our approach to pure parallel simulation (\textit{i.e.}, no distributed simulation), we are able to achieve higher speedups through the use of shared memory communication. 
 Recent work on parallel speedup in DEVS investigates theoretical limits of PDEVS~\cite{activitypdevs, amdahlpdevs}
@@
 Referenced as partitioning of the simulation model, most studies distinguish between communication and computation as the two dimensions to partition over.
-Partitioning a simulation model is identified as an issue to achieve scalability~\cite{Scalability}. 
+Partitioning a model is identified as an issue to achieve scalability~\cite{Scalability}. 
 Some research in this context has been done for \textsf{Parallel DEVS}~\cite{PDEVSpartitioning, NonFragmenting}.
-Our contribution studies the effect of partitioning with emphasis on the effect of communication between processes and in the presence of a flattened hierarchy. 
+Our contribution studies the effect of partitioning with emphasis on the effect of communication between kernels and in the presence of a flattened hierarchy. 
 We focus on static partitioning since this is a limiting factor for our conservative synchronization implementation which does not support model migration.

### 2016-12-08 publications/Journal/8-conclusions.tex
Small changes to section 8
Made whole paper (more) consistent

@@
 In this paper, we introduced \textit{DEVS-Ex-Machina} (``\textit{dxex}''), a new C++11-based \textsf{Parallel DEVS} simulation tool.
-Our main contribution is the implementation of multiple synchronization protocols for parallel multicore simulation.
+Our main contribution is the implementation of multiple synchronization protocols for parallel simulation.
 We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
-\textit{Dxex} allows the user to choose between either conservative or optimistic synchronization at the start of simulation.
+\textit{Dxex} allows the user to choose between any of the six supported synchronization protocols.
+We distinguish between inter-kernel (none, conservative, and optimistic synchronization) and intra-kernel (sequential or parallel concurrent transitions) synchronization protocols, which are orthogonal to one another.
 Depending on observed model behaviour and simulation performance, runtime switching between synchronization protocols can be used.
@@
 Furthermore we have shown when and why optimistic synchronization can outperform conservative and vice versa.
+We have also shown the influence of using the parallelism inherent in the Parallel DEVS formalism, and its interaction with inter-kernel synchronization algorithms.
 Finally we investigated the effect of memory (de)allocation on parallel simulation.
@@
 In its current state, simulation will simply halt with an out-of-memory error.
-The use of transactional memory can offer several advantages in this project.
-If it becomes part of the new C++17 standard it would be of great interest to see if it can help reduce the performance effects of memory allocation and synchronization.
 Having simulation control, which can throttle the speed of nodes that use up too much memory, has been shown to work in these situations~\cite{FujimotoBook}.
@@
 Similarly, it could serve as a default parallel allocation scheme that can be improved by the user.
+Finally, the use of transactional memory can offer several advantages in this project.
+If it becomes part of the new C++17 standard it would be of great interest to see if it can help reduce the performance effects of memory allocation and synchronization.

### 2016-12-08 publications/Journal/4-performance.tex
Updated captions of figures as well; PHOLD -> PHold consistently

@@
 
-Second, the \textit{Interconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
-Similar to PHOLD, all models are connected to one another, but all through the same port: every atomic model receives each generated event (\textit{i.e.}, the event is broadcasted).
+Second, the \textit{Interconnect} model, a merge of PHold~\cite{PHold} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
+Similar to PHold, all models are connected to one another, but all through the same port: every atomic model receives each generated event (\textit{i.e.}, the event is broadcasted).
 The model takes one parameter: the number of atomic models.
@@
 
-Third, the \textit{PHOLD} model~\cite{PHOLD}, creates $n$ atomic models, where each atomic model has exactly $n-1$ output ports.
+Third, the \textit{PHold} model~\cite{PHold}, creates $n$ atomic models, where each atomic model has exactly $n-1$ output ports.
 Each atomic model is directly connected to every other atomic model.
@@
 This benchmark shows how the simulation kernel behaves in the presence of many local or remote events, in combination with a varying percentage of high-priority events.
-An example for four atomic models, split over two kernels, is shown in Figure~\ref{fig:PHOLD_model}.
+An example for four atomic models, split over two kernels, is shown in Figure~\ref{fig:PHold_model}.
 
@@
 	\includegraphics[width=\modelfraction\columnwidth]{fig/phold_model.pdf}
-	\caption{PHOLD model for four models. Color denotes the two nodes.}
-	\label{fig:PHOLD_model}
+	\caption{PHold model for four models. Color denotes the two nodes.}
+	\label{fig:PHold_model}
 \end{figure}
@@
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_strong_speedup.eps}
-	\caption{Queue model strong scaling speedup compared to \textit{dxex} \textit{NN} synchronization.}
+	\caption{Queue model strong scaling speedup compared to \textit{dxex} \textit{NN} synchronization.
+             Each kernel uses one thread, and has one physical CPU core allocated to it.}
 	\label{fig:Queue_plot_strong}
@@
 	\includegraphics[width=\columnwidth]{fig/queue_fixed_weak_speedup.eps}
-	\caption{Queue model weak scaling speedup compared to \textit{dxex} \textit{NN} synchronization.}
+	\caption{Queue model weak scaling speedup compared to \textit{dxex} \textit{NN} synchronization.
+             Each kernel uses one thread, and has one physical CPU core allocated to it.}
 	\label{fig:Queue_plot_weak}
@@
 	\includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
-	\caption{Interconnect benchmark results for \textit{ON} and \textit{CN} synchronization.}
+	\caption{Interconnect benchmark results for \textit{ON} and \textit{CN} synchronization, compared to \textit{dxex} \textit{NN} synchronization.
+             Each kernel uses one thread, and has one physical CPU core allocated to it.}
 	\label{fig:interconnect_benchmark_parallel}
@@
 
-\subsubsection{PHOLD}
-In the PHOLD model, we first investigate the influence of the percentage of remote events on the speedup.
+\subsubsection{PHold}
+In the PHold model, we first investigate the influence of the percentage of remote events on the speedup.
 A remote event in this context is an event that is sent from an atomic model on one kernel to an atomic model on another kernel.
@@
     \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
-    \caption{PHOLD benchmark results for multi core simulation using four kernels, with varying percentage of remote events.}
+    \caption{PHold benchmark results using four kernels, with varying percentage of remote events.}
 \end{figure}
@@
 	\includegraphics[width=\columnwidth]{fig/phold_priority.eps}
-	\caption{PHOLD benchmark results for multi core simulation using four kernels, with varying amount of high-priority events.}
+	\caption{PHold benchmark results using four kernels, with varying amount of high-priority events.}
 	\label{fig:phold_priority}
@@
 	\includegraphics[width=\columnwidth]{fig/pdevs_no_sleep.eps}
-	\caption{Queue speedup benchmark for PDEVS and synchronization with maximal concurrent events and trivial computational load}
+	\caption{Queue speedup benchmark with $p = 1$, $q = 1$, and trivial computational load}
 	\label{fig:psim_plot_no_sleep}

### 2016-12-08 publications/Journal/5-hotswap.tex
Updated captions of figures as well; PHOLD -> PHold consistently

@@
 For each connection, the number of events transferred is annotated.
-Examples are shown for the three benchmark models used before: Figure~\ref{fig:Queue_allocation}, Figure~\ref{fig:interconnect_allocation_parallel}, and Figure~\ref{fig:phold_allocation} show traces for the Queue, Interconnect, and PHOLD models, respectively.
+Examples are shown for the three benchmark models used before: Figure~\ref{fig:Queue_allocation}, Figure~\ref{fig:interconnect_allocation_parallel}, and Figure~\ref{fig:phold_allocation} show traces for the Queue, Interconnect, and PHold models, respectively.
 Using this information, we notice that the Interconnect benchmark indeed has a lot of inter-kernel events.
-Despite the similar structure, the PHOLD model does not have as many inter-kernel events.
+Despite the similar structure, the PHold model does not have as many inter-kernel events.
 These results are relevant information that can be used by the hotswapping component.
@@
     \includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_parallel_allocation.eps}
-    \caption{Interconnect simulation trace for 6 models on 3 kernels.}
+    \caption{Interconnect simulation trace for 6 atomic models on 3 kernels.}
     \label{fig:interconnect_allocation_parallel}
@@
     \includegraphics[width=\plotfraction\columnwidth]{fig/phold_parallel_allocation.eps}
-    \caption{Phold simulation trace for multi core simulation using three kernels.}
+    \caption{PHold simulation trace for 6 atomic models on 3 kernels.}
     \label{fig:phold_allocation}

### 2016-12-08 publications/Journal/6-allocation.tex
Updated captions of figures as well; PHOLD -> PHold consistently

@@
 \subsection{Performance Evaluation}
-To evaluate the influence of model allocation, we define a new model, based on PHOLD~\cite{PHOLD}.
+To evaluate the influence of model allocation, we define a new model, based on PHold~\cite{PHOLD}.
 The model structure resembles a tree: an atomic model can have a set of children, with children being connected to each other recursively.
@@
 Unlike the Queue model, the width of the hierarchy is still present in the topology of the atomic models after direct connection.
-The PHOLDTree model allows us to investigate speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.
+The PHoldTree model allows us to investigate speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.
 
-The PHOLDTree model is similar in structure to models of gossiping in social networks~\cite{Gossip}.
+The PHoldTree model is similar in structure to models of gossiping in social networks~\cite{Gossip}.
 The lookahead of an atomic node is the minimally allowed $\epsilon$, as is often the case in realistic models (\textit{e.g.}, if it is unknown what the lookahead might be).
@@
     \includegraphics[width=\columnwidth]{fig/pholdtree.pdf}
-    \caption{PHOLDTree model for depth 3 and width 2.}
+    \caption{PHoldTree model for depth 3 and width 2.}
     \label{fig:PHOLDTree_model}
@@
 \subsubsection{Inter-Kernel Synchronization (\textit{CN} and \textit{ON})}
-Next, we run the model using two different allocation schemes: breadth-first and depth-first.
+Next, we run the model using two different allocation schemes, based on Breadth-First Search (BFS) and Depth-First Search (DFS).
 First, we explain both allocation schemes.
@@
    \includegraphics[width=\columnwidth]{fig/pholdtree_alloc_BF.pdf}
-   \caption{PHOLDTree model breadth first allocation with 4 kernels.}
+   \caption{PHoldTree model breadth first allocation with 4 kernels.}
    \label{fig:PholdTree_model_bfs}
@@
    \includegraphics[width=\columnwidth]{fig/pholdtree_alloc_DF.pdf}
-   \caption{PHOLDTree model depth first allocation with 4 kernels.}
+   \caption{PHoldTree model depth first allocation with 4 kernels.}
    \label{fig:PholdTree_model_dfs}
@@
     \includegraphics[width=\columnwidth]{fig/pholdtreeallochighp.eps}
-    \caption{PHOLDTree model performance using different allocation schemes.}
+    \caption{PHoldTree model performance using different allocation schemes.}
     \label{fig:PholdTree_plot_alloc_high}

### 2016-12-08 publications/Journal/4-performance.tex
* updated all images * fixed broken citations with incorrect capitalization

@@
 
-Second, the \textit{Interconnect} model, a merge of PHold~\cite{PHold} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
+Second, the \textit{Interconnect} model, a merge of PHold~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
 Similar to PHold, all models are connected to one another, but all through the same port: every atomic model receives each generated event (\textit{i.e.}, the event is broadcasted).
@@
 
-Third, the \textit{PHold} model~\cite{PHold}, creates $n$ atomic models, where each atomic model has exactly $n-1$ output ports.
+Third, the \textit{PHold} model~\cite{PHOLD}, creates $n$ atomic models, where each atomic model has exactly $n-1$ output ports.
 Each atomic model is directly connected to every other atomic model.
